---
title: "Predicting Excessive Enforcement Patterns in Police Behavior After Vehicles Stops in New York"
subtitle: "EDA and Feature Engineering - 4 Wheels Vehicles"
author: "Gianmaria Pizzo"
output:
  html_document:
    fig_caption: yes
    theme: flatly
    highlight: pygments
    code_folding: show
    kee_md: yes
    toc: yes
    toc_depth: 4
    number_sections: no
    toc_float:
      smooth_scroll: no
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

These `.Rmd` files represent the project submission for the course [Statistical Inference and Learning](https://www.unive.it/data/course/521944) by Professor [Cristiano Varin](https://www.unive.it/data/people/5592728) at [Ca' Foscari University of Venice](https://www.unive.it/) for the year 2025/2026.

# Problem Statement - TODO

## Goal of this rmarkdown file - TODO

We explore the features in 3 different ways.

1.  First, we separately analyze them and try to grasp their meaning, their distribution and the possible outliers or other problematic aspects. This univariate analysis, although complicated by the presence of many categorical predictors, should give us an initial idea of how the data is distributed.
2.  Then, we compare the possible predictors with respect to the target feature, and try to understand if there is any notable relationship between the two of them. Also, we try to see whether there is evidence of correlation, collinearity, or other problematic behaviors they may introduce.
3.  Finally we explore the subsets of features that may be related (ex: geospatial features and command unit, or time of the day and some demographics), to find out whether some of them can be removed or replaced. Plus we try to investigate about multicollinearity and other complications.

The EDA is the first step to understanding the data and its nature.

This includes analyzing the single features (Univariate Analysis), their interactions with the target (Bivariate Analysis) and their interactions with other features (Multivariate Analysis and Collinearity Check). However, the latter might be hard to compute as it requires the estimation of a regression model that follows certain assumptions and it is usually easier to check this when using linear models.

TODO: Say we will skip some visualizations

During my exploration I will look out for possible outliers, correlation and any other statistically relevant problem that might affect my final prediction.

We will start with the environment.

Note that, from now on, I will shift adopt a more **formal or inclusive tone** (e.g., the "editorial we").

FIXME: Comments should be done addressing 2 wheels vehicles. FIXME: Reduce colors, add annotation and title that is not just plain description FIXME: Univariate plots need both count and percentage on top

------------------------------------------------------------------------

# Environment

The first to do, is to build a reproducible environment for our project.

```{r}
requirements=c( 
  "remotes", "summarytools", "forcats", "scales", "purrr",
  "stringr", "plyr", "dbplyr", "dplyr", "dtplyr", "tidyr", "tidyselect",
  "timeDate", "timechange", "tzdb", "hms", "readr", "tigris",
  "lubridate", "prettyunits", "RColorBrewer", "viridis", "rcartocolor", "ggbeeswarm", "emmeans",
  "sf", "plotly", "ggplot2", "ggmap", "ggdensity", "UpSetR", "patchwork"
)
```

```{r install-requirements, include=FALSE}
for (req in requirements){
  if (!require(req, character.only = TRUE)){
      install.packages(req)
  }
}
```

Note that some of the packages may not be used, but to avoid any error related to dependencies, we are going to keep all of these libraries. Now, the loading of all packages should be immediate.

```{r load-requirements, include=FALSE}
for (req in requirements){
  if (!require(req, character.only = TRUE)){
      library(req)
  }
}
```

```{r remove-vars, include=FALSE}
rm(requirements, req)
```

## Plot Functions

Here we add the plot functions to make some kinds of visualizations reproducible.

```{r univariate-plot-helpers}
# ==============================================================================
# 1. Plots for Univariate Analysis
# ==============================================================================
#' @title Plot Binary Bar Chart
#' 
#' @description Creates a simple bar chart counting the occurrences of a binary feature, 
#' with labels showing both the count and the percentage.
#' 
#' @param DF The input dataframe.
#' @param FEATURE The binary column to plot (unquoted).
#' @param OPTION_0 The text label for the "0" or "FALSE" category.
#' @param OPTION_1 The text label for the "1" or "TRUE" category.
#' @param X_LABEL Label for the X axis.
#' @param Y_LABEL Label for the Y axis.
#' @param TITLE Main title of the plot.
#' @param pad Numeric padding for the top of the Y axis to fit text labels.
#' @param text_vjust Vertical adjustment for the text labels on top of bars.
#' 
#' @return A ggplot2 object.
plot_binary_bar <- function(DF, FEATURE, OPTION_0, OPTION_1, X_LABEL, Y_LABEL, TITLE, pad = 0.15, text_vjust = -0.3) {
  
  # Standardize the binary feature and calculate percentages
  df_sum <- DF %>%
    mutate(
      .raw  = {{ FEATURE }},
      .feat = dplyr::case_when(
        as.character(.raw) %in% c("1","TRUE","true")  ~ 1L,
        as.character(.raw) %in% c("0","FALSE","false") ~ 0L,
        TRUE ~ NA_integer_
      )
    ) %>%
    filter(!is.na(.feat)) %>%
    count(.feat, name = "n") %>%
    mutate(
      .lab = factor(.feat, levels = c(0, 1), labels = c(OPTION_0, OPTION_1)),
      # NEW: Calculate percentage and create combined label
      pct = n / sum(n),
      label_text = paste0(scales::comma(n), "\n(", scales::percent(pct, accuracy = 0.1), ")")
    )
  
  # Get the maximum bar height
  ymax <- max(df_sum$n, na.rm = TRUE)
  
  # Build the ggplot
  ggplot(df_sum, aes(x = .lab, y = n, fill = .lab)) +
    geom_col(width = 0.8, show.legend = FALSE) +
    # Updated to use label_text
    geom_text(aes(label = label_text), vjust = text_vjust, lineheight = 0.8) +
    # Expand the Y axis slightly to accommodate the taller multi-line labels
    scale_y_continuous(limits = c(0, ymax * (1 + pad)),
                       expand = expansion(mult = c(0, 0.02))) +
    coord_cartesian(clip = "off") +
    theme_minimal() +
    theme(
      panel.grid.major.x = element_blank(), 
      plot.margin = margin(t = 20, r = 5.5, b = 5.5, l = 5.5)
    ) +
    scale_fill_viridis_d(option = "mako", begin = 0.3, end = 0.8) +
    labs(x = X_LABEL, y = Y_LABEL, title = TITLE)
}


#' @title Plot Univariate Bar Chart (Categorical)
#' 
#' @description Creates a bar chart counting the occurrences of a categorical feature with 2+ classes.
#' Allows dynamic ordering (by frequency or original/alphabetical order).
#' 
#' @param DF The input dataframe.
#' @param FEATURE The categorical column to analyze (unquoted).
#' @param X_LABEL Label for the X axis.
#' @param Y_LABEL Label for the Y axis (default is "Count").
#' @param TITLE Main title of the plot.
#' @param top_n Optional integer; if set, keeps only the top N most frequent categories.
#' @param order_by Method to order the X axis: "freq" (descending count) or "none" (alphabetical/original).
#' @param pad Numeric padding for the top of the Y axis to fit text labels.
#' @param text_vjust Vertical adjustment for the text labels on top of bars.
#' 
#' @return A ggplot2 object.
plot_univariate_cat <- function(DF, FEATURE, X_LABEL, Y_LABEL = "Count", TITLE, top_n = NULL, order_by = c("freq", "none"), pad = 0.12, text_vjust = -0.35) {
  
  order_by <- match.arg(order_by)
  
  # Compute counts
  df_sum <- DF %>%
    filter(!is.na({{ FEATURE }})) %>%
    count({{ FEATURE }}, name = "n")
  
  # Filter top n results
  if (!is.null(top_n)) {
    df_sum <- df_sum %>% slice_max(order_by = n, n = top_n)
  }
  
  # X axis order (either frequency or std)
  cat_levels <- switch(
    order_by,
    freq = df_sum %>% arrange(desc(n)) %>% pull({{ FEATURE }}),
    none = sort(unique(df_sum %>% pull({{ FEATURE }})))
  )
  
  # Apply sorting
  df_sum <- df_sum %>%
    mutate(.feat_f = factor(as.character({{ FEATURE }}), levels = as.character(cat_levels)))
  
  # Max height for dynamic padding
  ymax <- max(df_sum$n, na.rm = TRUE)

  # Construct plot
  ggplot(df_sum, aes(x = .feat_f, y = n, fill = .feat_f)) +
    geom_col(width = 0.8, show.legend = FALSE) +
    geom_text(aes(label = scales::comma(n)), vjust = text_vjust, size = 3) +
    scale_y_continuous(limits = c(0, ymax * (1 + pad)),
                       expand = expansion(mult = c(0, 0.02))) +
    coord_cartesian(clip = "off") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
          panel.grid.major.x = element_blank(),
          plot.margin = margin(t = 10, r = 5.5, b = 5.5, l = 5.5)) +
    scale_fill_viridis_d(option = "mako", begin = 0.3, end = 0.8) +
    labs(x = X_LABEL, y = Y_LABEL, title = TITLE)
}
```

```{r biviariate-plot-helpers}
# ==============================================================================
# 2. Plots for Bivariate Analysis
# ==============================================================================

#' @title Calculate Cramer's V
#' 
#' @description Calculates the Cramer's V measure of association between two 
#' categorical vectors. Returns a value between 0 (no association) and 1 (perfect association).
#' 
#' @param x First categorical vector.
#' @param y Second categorical vector.
#' 
#' @return A numeric value representing Cramer's V.
cramer_v <- function(x, y) {
  # Contingency Table
  tbl <- table(x, y)
  
  # Execute Chi2 Test
  chi2 <- suppressWarnings(chisq.test(tbl, correct = FALSE)$statistic)
  
  # Compute dim and min value
  n <- sum(tbl)
  min_dim <- min(nrow(tbl) - 1, ncol(tbl) - 1)
  
  # Avoid math error for 1 categ
  if(min_dim == 0) return(0) 
  
  # Cramer's formula
  v <- sqrt(chi2 / (n * min_dim))
  return(as.numeric(v))
}

# ==============================================================================
# Categorical "Correlation" Heatmap - Association Matrix based on Cramer's V
# ==============================================================================

#' @title Plot Categorical Correlation Matrix (Cramer's V)
#' 
#' @description Creates a correlation-like heatmap for multiple categorical variables
#' using Cramer's V.
#' 
#' @param DF The input dataframe.
#' @param FEATURES A character vector containing the names of the categorical columns to compare.
#' @param TITLE Main title of the plot.
#' 
#' @return A ggplot2 object.
plot_categorical_correlation <- function(DF, FEATURES, TITLE = "Association Matrix for Categorical Predictors (Cramer's V)") {
  
  # Select columns and drop NAs
  df_sub <- DF %>% 
    select(all_of(FEATURES)) %>%
    drop_na()
  
  n_vars <- ncol(df_sub)
  
  # Initialize matrix and set main diag to 1s
  mat <- matrix(1, nrow = n_vars, ncol = n_vars)
  rownames(mat) <- colnames(df_sub)
  colnames(mat) <- colnames(df_sub)
  
  # FIXME: Just do upper/lower part of matrix
  # Compute Cramer's V for each possible pair
  for(i in 1:(n_vars - 1)) {
    for(j in (i + 1):n_vars) {
      v <- cramer_v(df_sub[[i]], df_sub[[j]])
      # Symmetric 
      mat[i, j] <- v
      mat[j, i] <- v
    }
  }
  
  # Make matrix compatible with ggplot2
  mat_long <- as.data.frame(as.table(mat))
  colnames(mat_long) <- c("Var1", "Var2", "CramerV")
  
  # Create heatmap
  ggplot(mat_long, aes(x = Var1, y = Var2, fill = CramerV)) +
    geom_tile(color = "white", width = 1, height = 1) +
    # Rounding
    geom_text(aes(label = round(CramerV, 2)), size = 4) +
    # Setting colors
    scale_fill_gradient(low = "whitesmoke", high = "steelblue", limits = c(0, 1), name = "Cramer's V") +
    labs(title = TITLE, x = "", y = "") +
    theme_minimal() +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1, size = 10, face = "bold"),
      axis.text.y = element_text(size = 10, face = "bold"),
      panel.grid = element_blank()
    )
}

# ==============================================================================
# Plotting of Targets' Rate Against Categorical Predictors
# ==============================================================================

# FIXME: Percentage shown in the wrong place.

#' @title Plot Target Ratio by Category
#' 
#' @description Creates a bivariate bar chart showing the relationship between a binary 
#' TARGET and a single categorical variable (CAT). It supports two visual styles: 
#' a 100% stacked bar chart ("stack") or a simple bar chart showing the rate of the target ("rate").
#' 
#' @param DF The input dataframe.
#' @param TARGET The binary column to analyze (unquoted). It will be standardized to 0/1.
#' @param CAT The categorical column for the X axis (unquoted).
#' @param x_title Label for the X axis.
#' @param y_title Label for the Y axis.
#' @param title Main title of the plot.
#' @param target_labels A character vector of length 2 defining the labels for 0 and 1 (e.g., c("Not escalated", "Escalated")).
#' @param style The visual style of the plot: "stack" (100% stacked bars) or "rate" (single bar for target = 1).
#' @param order_by Method to order the X axis categories: by "rate" (ascending target rate), "freq" (ascending frequency), or "none" (alphabetical/original).
#' @param percent_accuracy Accuracy for the percent labels (e.g., 0.1 for one decimal place).
#' @param colors A character vector of length 2 defining the hex color codes for the 0 and 1 categories respectively.
#' 
#' @return A ggplot2 object.
ratio_by_cat <- function(DF, TARGET, CAT, x_title, y_title, title, 
                         target_labels = c("Not escalated", "Escalated"), 
                         style = c("stack", "rate"), 
                         order_by = c("rate", "freq", "none"), 
                         percent_accuracy = 0.1, 
                         colors = c("grey80", "steelblue"),
                         legend_title = "Legend",
                         subtitle = NULL,
                         label_vjust = 0.5) {
  
  # Validate arguments
  style    <- match.arg(style)
  order_by <- match.arg(order_by)
  
  # Standardize target and extract the category
  dat <- DF %>%
    mutate(
      .t_raw = {{ TARGET }},
      .t = case_when(
        as.character(.t_raw) %in% c("1","TRUE","true") ~ 1L,
        as.character(.t_raw) %in% c("0","FALSE","false") ~ 0L,
        TRUE ~ NA_integer_
      ),
      .cat = {{ CAT }}
    ) %>%
    filter(!is.na(.t), !is.na(.cat))
  
  # Compute rates and frequencies
  rates <- dat %>% 
    group_by(.cat) %>%
    summarise(total = n(), rate = mean(.t == 1), .groups = "drop")
  
  # Determine ordering
  cat_levels <- switch(
    order_by,
    rate = rates %>% arrange(rate) %>% pull(.cat),
    freq = rates %>% arrange(total) %>% pull(.cat),
    none = unique(dat$.cat)
  )
  
  # Build the plot based on style
  if (style == "stack") {
    stack <- dat %>%
      count(.cat, .t, name = "n") %>%
      group_by(.cat) %>%
      mutate(ratio = n/sum(n)) %>%
      ungroup() %>%
      mutate(
        .cat_f = factor(as.character(.cat), levels = as.character(cat_levels)),
        .t_f = factor(.t, levels = c(0,1), labels = target_labels)
      )
    
    ggplot(stack, aes(x = .cat_f, y = ratio, fill = .t_f)) +
      geom_col(position = "fill") +
      # CHANGED: Showing ONLY the percentage
      geom_text(
        aes(label = scales::percent(ratio, accuracy = percent_accuracy)),
        position = position_fill(vjust = label_vjust), 
        size = 3
      ) +
      scale_y_continuous(labels = scales::percent) +
      scale_fill_manual(values = setNames(colors, target_labels)) +
      labs(x = x_title, y = y_title, fill = legend_title, title = title, subtitle = subtitle) +
      theme_minimal() +
      theme(
        axis.text.x = element_text(angle = 45, hjust = 1),
        panel.grid = element_blank(),
        plot.subtitle = element_text(color = "grey50")
      )
      
  } else {
    # Rate style
    rates_data <- rates %>%
      mutate(.cat_f = factor(as.character(.cat), levels = as.character(cat_levels)))
    
    ggplot(rates_data, aes(x = .cat_f, y = rate)) +
      geom_col(width = 0.8, fill = colors[2]) +
      geom_text(
        aes(label = scales::percent(rate, accuracy = percent_accuracy)),
        vjust = -0.35, size = 3
      ) +
      scale_y_continuous(labels = scales::percent, expand = expansion(mult = c(0, 0.12))) +
      coord_cartesian(clip = "off") +
      labs(x = x_title, y = y_title, title = title, subtitle = subtitle) +
      theme_minimal() +
      theme(
        axis.text.x = element_text(angle = 45, hjust = 1),
        panel.grid = element_blank(),
        plot.subtitle = element_text(color = "grey50")
      )
  }
}
```

```{r multivariate-plot-helpers}
# ==============================================================================
# 3. Plots for Multivariate Analysis
# ==============================================================================

#' @title Heatmap of Target Rate (Categorical Y vs Numeric X)
#' 
#' @description Plots a heatmap showing the rate of a binary TARGET across a categorical
#' Y variable and a numeric (or factor) X variable.
#' 
#' @param DF The input dataframe.
#' @param TARGET The binary column to calculate the rate for (unquoted).
#' @param Y Categorical variable for the Y axis.
#' @param X Numeric or categorical variable for the X axis.
#' @param X_TITLE Label for the X axis.
#' @param Y_TITLE Label for the Y axis.
#' @param TITLE Main title of the plot.
#' @param min_n Minimum number of observations required to show a tile.
#' @param drop_na Boolean; if TRUE, rows with NA in X or Y are removed.
#' @param x_as_factor Boolean; if TRUE, forces the X variable to be treated as a categorical factor.
#' 
#' @return A ggplot2 object.
heat_target_yx <- function(DF, TARGET, Y, X, X_TITLE, Y_TITLE, TITLE, min_n = 0, drop_na = TRUE, x_as_factor = FALSE) {
  # Standardize the binary target
  dat <- DF %>%
    mutate(
      .target_raw = {{ TARGET }},
      .target_bin = dplyr::case_when(
        as.character(.target_raw) %in% c("1", "TRUE", "true")  ~ 1L,
        as.character(.target_raw) %in% c("0", "FALSE", "false") ~ 0L,
        TRUE ~ NA_integer_
      )
    )

  # Filter NAs
  if (drop_na) {
    dat <- dat %>% filter(!is.na({{Y}}), !is.na({{X}}), !is.na(.target_bin))
  } else {
    dat <- dat %>% filter(!is.na(.target_bin))
  }

  # Incrementally sort Y axis based on the mean ratio of the target
  y_order <- dat %>%
    group_by({{Y}}) %>%
    summarise(rate = mean(.target_bin == 1, na.rm = TRUE), .groups = "drop") %>%
    arrange(rate) %>%
    pull({{Y}})

  # Compute the target rate for each combination of Y x X
  grid <- dat %>%
    group_by({{Y}}, {{X}}) %>%
    summarise(
      rate = mean(.target_bin == 1, na.rm = TRUE),
      n = n(), .groups = "drop"
    ) %>%
    mutate(
      .y = factor(as.character({{Y}}), levels = as.character(y_order)),
      .x = if (x_as_factor) factor({{X}}) else {{X}},
      # Hide cells that do not reach the minimum amount of observations
      rate_plot = ifelse(n >= min_n, rate, NA_real_)
    )

  # Build ggplot heatmap
  ggplot(grid, aes(.x, .y, fill = rate_plot)) +
    geom_tile(width = 1, height = 0.9) +
    scale_fill_viridis_c(labels = scales::percent, na.value = "grey90", name = "Rate") +
    labs(x = X_TITLE, y = Y_TITLE, title = TITLE) +
    theme_minimal() +
    theme(panel.grid = element_blank())
}


#' @title Heatmap of Target Rate (Categorical Y vs Categorical X)
#' 
#' @description Plots a heatmap showing the rate of a binary TARGET across two categorical
#' variables (Y and X). Allows dynamic ordering and shading based on sample size.
#' 
#' @param DF The input dataframe.
#' @param TARGET The binary column to calculate the rate for (unquoted).
#' @param Y Categorical variable for the Y axis.
#' @param X Categorical variable for the X axis.
#' @param X_TITLE Label for the X axis.
#' @param Y_TITLE Label for the Y axis.
#' @param TITLE Main title of the plot.
#' @param min_n Minimum number of observations required to show a tile (filters noise).
#' @param order_y_by Method to order the Y axis: by 'rate', 'freq' (frequency), or 'none'.
#' @param order_x_by Method to order the X axis: by 'rate', 'freq' (frequency), or 'none'.
#' @param alpha_by_n Boolean; if TRUE, shades tiles by count using alpha (transparency).
#' 
#' @return A ggplot2 object.
heat_target_catcat <- function(DF, TARGET, Y, X, X_TITLE, Y_TITLE, TITLE, min_n = 0, order_y_by = c("rate","freq","none"), order_x_by = c("rate","freq","none"), alpha_by_n = FALSE ) {
  order_y_by <- match.arg(order_y_by)
  order_x_by <- match.arg(order_x_by)

  # Standardize target
  dat <- DF %>%
    mutate(
      .target_raw = {{ TARGET }},
      .target_bin = dplyr::case_when(
        as.character(.target_raw) %in% c("1", "TRUE", "true")  ~ 1L,
        as.character(.target_raw) %in% c("0", "FALSE", "false") ~ 0L,
        TRUE ~ NA_integer_
      )
    ) %>%
    # Filter NAs
    filter(!is.na({{Y}}), !is.na({{X}}), !is.na(.target_bin))

  # Compute stats to determine the Y axis order
  y_stats <- dat %>%
    group_by({{Y}}) %>%
    summarise(
      rate = mean(.target_bin == 1, na.rm = TRUE),
      n = n(), .groups = "drop"
    )
  
  # Compute stats to determine the X axis order
  x_stats <- dat %>%
    group_by({{X}}) %>%
    summarise(
      rate = mean(.target_bin == 1, na.rm = TRUE), 
      n = n(), .groups = "drop"
    )

  # Apply chosen "order by" to Y
  y_levels <- switch(
    order_y_by,
    rate = y_stats %>% arrange(rate) %>% pull({{Y}}),
    freq = y_stats %>% arrange(n) %>% pull({{Y}}),
    none = unique(dat %>% pull({{Y}}))
  )
  
  # Apply chosen "order by" to X
  x_levels <- switch(
    order_x_by,
    rate = x_stats %>% arrange(rate) %>% pull({{X}}),
    freq = x_stats %>% arrange(n) %>% pull({{X}}),
    none = unique(dat %>% pull({{X}}))
  )

  # Compute the cell's target rate and counts for each combination of Y x X
  grid <- dat %>%
    group_by({{Y}}, {{X}}) %>%
    summarise(
      rate = mean(.target_bin == 1, na.rm = TRUE),
      n = n(), .groups = "drop"
    ) %>%
    mutate(
      .y = factor(as.character({{Y}}), levels = as.character(y_levels)),
      .x = factor(as.character({{X}}), levels = as.character(x_levels)),
      rate_plot = ifelse(n >= min_n, rate, NA_real_)
    )

  # Build the plot
  p <- ggplot(grid, aes(.x, .y)) +
    labs(x = X_TITLE, y = Y_TITLE, title = TITLE) +
    theme_minimal() +
    theme(panel.grid = element_blank()) +
    # Use scales for displaying rates
    scale_fill_viridis_c(labels = scales::percent, na.value = "grey90", name = "Rate")

  if (alpha_by_n) {
    p <- p +
      geom_tile(aes(fill = rate_plot, alpha = n), width = 0.95, height = 0.95) +
      scale_alpha_continuous(range = c(0.25, 1), name = "Stops")
  } else {
    p <- p +
      geom_tile(aes(fill = rate_plot), width = 0.95, height = 0.95)
  }

  p
}


#' @title Stacked Ratio Bars Split by a Third Variable
#' 
#' @description Creates a faceted 100% stacked bar chart. This is highly effective 
#' for visualizing how the rate of a binary TARGET changes across a primary category (CAT), 
#' while simultaneously splitting the data by a third variable (SPLIT) to create side-by-side plots.
#' 
#' @param DF The input dataframe.
#' @param TARGET The binary column to analyze (unquoted). It will be standardized to 0/1.
#' @param CAT The main categorical variable for the X axis (unquoted).
#' @param SPLIT The categorical variable used to split/facet the plots (unquoted).
#' @param x_title Label for the X axis.
#' @param y_title Label for the Y axis.
#' @param title Main title of the entire plot.
#' @param target_labels Character vector of length 2 defining labels for the binary outcomes (e.g., c("Not escalated", "Escalated")).
#' @param split_labels Optional character vector to provide custom titles for the split panels (facets).
#' @param palette A character string specifying an RColorBrewer palette name (e.g., "Paired", "Set1").
#' @param percent_accuracy Accuracy for the percentage labels (e.g., 0.1 for one decimal place).
#' 
#' @return A ggplot2 object.
ratio_by_cat_split <- function(DF, TARGET, CAT, SPLIT, x_title, y_title, title, target_labels = c("Not escalated", "Escalated"), split_labels = NULL, palette = "Paired", percent_accuracy = 0.1){
  # Standardize target
  dat <- DF %>%
    mutate(
      .target_raw = {{ TARGET }},
      # Conversion to 0/1
      .target_bin = case_when(
        as.character(.target_raw) %in% c("1", "TRUE", "true") ~ 1L,
        as.character(.target_raw) %in% c("0", "FALSE", "false") ~ 0L,
        TRUE ~ NA_integer_
      ),
      .cat = {{ CAT }},
      .split = {{ SPLIT }}
    ) %>%
    # Filter NAs
    filter(!is.na(.cat), !is.na(.split), !is.na(.target_bin))

  # Order the main categories based on the overall target rate (from lowest to highest)
  cat_order <- dat %>%
    group_by(.cat) %>%
    summarise(
      rate = mean(.target_bin == 1),
      .groups = "drop"
    ) %>%
    arrange(rate) %>% 
    pull(.cat)

  # Tidy labels by converting character to factors
  dat <- dat %>%
    mutate(
      # Apply the order to X axis
      .cat = factor(as.character(.cat), levels = as.character(cat_order)),
      # Apply custom labels to the target
      .target_f = factor(.target_bin, levels = c(0, 1), labels = target_labels),
      # Format the split variable
      .split_f = {
        lv <- unique(as.character(.split))
        # If custom split labels are provided and the length matches, use them
        if (!is.null(split_labels) && length(split_labels) == length(lv))
          factor(as.character(.split), levels = lv, labels = split_labels)
        else factor(as.character(.split), levels = lv)
      }
    )

  # Compute rates
  plot_df <- dat %>%
    count(.split_f, .cat, .target_f, name = "n") %>%
    group_by(.split_f, .cat) %>%
    mutate(ratio = n/sum(n)) %>%
    ungroup()

  # Build plot
  ggplot(plot_df, aes(x = .cat, y = ratio, fill = .target_f)) +
    # Stacks the bars to exactly 100%
    geom_col(position = "fill") +
    # Add ratio text labels
    geom_text(
      aes(label = percent(ratio, accuracy = percent_accuracy)),
      position = position_fill(vjust = 0.5), 
      size = 3
    ) +
    # Format Y axis as percentages and make continuous
    scale_y_continuous(labels = percent) +
    # Use a clean palette for the target categories
    scale_fill_brewer(palette = palette, name = rlang::as_name(rlang::enquo(TARGET))) +
    labs(x = x_title, y = y_title, title = title) +
    # Split the plot into separate panels based on the .split_f variable
    facet_wrap(~ .split_f, nrow = 1) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
          panel.grid = element_blank())
}
```

```{r}
#' @title Plot Temporal Trend of Target Rate
#' 
#' @description Creates a scatter plot with a LOESS smoothing curve to visualize 
#' how the rate of a target event changes over a continuous temporal variable (e.g., date, day of year).
#' The size of the points represents the volume of stops.
#' 
#' @param DF The input dataframe.
#' @param TARGET The binary column to analyze (unquoted).
#' @param TIME_VAR The continuous time variable for the X axis (e.g., date, doy).
#' @param x_title Label for the X axis.
#' @param y_title Label for the Y axis.
#' @param title Main title of the plot.
plot_temporal_trend <- function(DF, TARGET, TIME_VAR, x_title = "Tempo", y_title = "Tasso", title = "Trend Temporale") {
  
  # Step 1: Standardizzare il target a 0/1
  dat <- DF %>%
    mutate(
      .target_raw = {{ TARGET }},
      .target_bin = case_when(
        as.character(.target_raw) %in% c("1","TRUE","true", "Yes")  ~ 1L,
        as.character(.target_raw) %in% c("0","FALSE","false", "No") ~ 0L,
        TRUE ~ NA_integer_
      ),
      .time = {{ TIME_VAR }}
    ) %>%
    filter(!is.na(.target_bin), !is.na(.time))
  
  # Step 2: Calcolare il tasso e il volume per ogni unità di tempo
  plot_df <- dat %>%
    group_by(.time) %>%
    summarise(
      rate = mean(.target_bin == 1, na.rm = TRUE),
      n = n(),
      .groups = "drop"
    )
  
  # Step 3: Costruire il grafico
  ggplot(plot_df, aes(x = .time, y = rate)) +
    # Punti proporzionali al numero di stop per capire la robustezza del dato
    geom_point(aes(size = n), alpha = 0.5, color = "steelblue") +
    # Linea di tendenza morbida per evidenziare il trend generale ignorando il rumore
    geom_smooth(se = FALSE, method = "loess", span = 0.3, color = "firebrick", size = 1.2) +
    scale_y_continuous(labels = scales::percent) +
    labs(x = x_title, y = y_title, size = "N. di Stop", title = title) +
    theme_minimal() +
    theme(panel.grid.minor = element_blank())
}

#' @title Plot Cyclic Time Heatmap
#' 
#' @description Creates a heatmap specifically designed for cyclic time variables 
#' (like Hour of Day vs Day of Week) to spot temporal hotspots.
#' 
#' @param DF The input dataframe.
#' @param TARGET The binary column to analyze (unquoted).
#' @param HOUR_VAR The variable representing the hour (0-23).
#' @param WEEKDAY_VAR The variable representing the day of the week.
#' @param title Main title of the plot.
plot_cyclic_time_heatmap <- function(DF, TARGET, HOUR_VAR, WEEKDAY_VAR, title = "Hotspots Temporali") {
  
  # Step 1: Standardizzare il target a 0/1
  dat <- DF %>%
    mutate(
      .target_raw = {{ TARGET }},
      .target_bin = case_when(
        as.character(.target_raw) %in% c("1","TRUE","true", "Yes")  ~ 1L,
        as.character(.target_raw) %in% c("0","FALSE","false", "No") ~ 0L,
        TRUE ~ NA_integer_
      ),
      .hour = as.numeric({{ HOUR_VAR }}),
      .day = {{ WEEKDAY_VAR }}
    ) %>%
    filter(!is.na(.target_bin), !is.na(.hour), !is.na(.day))
  
  # Step 2: Raggruppare e calcolare il tasso
  plot_df <- dat %>%
    group_by(.day, .hour) %>%
    summarise(
      rate = mean(.target_bin == 1, na.rm = TRUE),
      n = n(),
      .groups = "drop"
    ) %>%
    # Mostriamo solo incroci con un minimo di stop per evitare rumore statistico
    filter(n >= 10) 
  
  # Step 3: Costruire la mappa di calore temporale
  ggplot(plot_df, aes(x = .hour, y = .day, fill = rate)) +
    geom_tile(color = "white") +
    scale_fill_viridis_c(option = "magma", labels = scales::percent, name = "Tasso") +
    # Impostiamo l'asse X per mostrare tutte le 24 ore
    scale_x_continuous(breaks = 0:23, expand = c(0,0)) +
    labs(x = "Ora del Giorno", y = "", title = title) +
    theme_minimal() +
    theme(
      panel.grid = element_blank(),
      axis.text.y = element_text(face = "bold")
    )
}

#' @title Plot Temporal Trend Split by a Group (Rate or Frequency)
#' 
#' @description Generates a line plot showing how a metric (rate or count) 
#' changes over a generic time unit (X_VAR), split by another variable (Z_VAR).
#' Adapts to any time unit and can display data as overlapping lines or faceted grids.
#' 
#' @param DF The input dataframe.
#' @param X_VAR The generic time variable for the X axis (e.g., month, hour).
#' @param Z_VAR The variable to group/split the plots by (e.g., year, weekend).
#' @param TARGET The binary target variable. Required only if metric = "rate".
#' @param metric "rate" (percentage of TARGET) or "freq" (absolute count of stops).
#' @param style "overlay" (lines superimposed on one plot) or "facet" (separated sub-plots).
#' @param x_title Label for the X axis.
#' @param y_title Label for the Y axis. Default adjusts based on metric.
#' @param title Main title of the plot.
#' 
#' @return A ggplot2 object.
plot_temporal_split <- function(
  DF, X_VAR, Z_VAR, TARGET = NULL, 
  metric = c("rate", "freq"), 
  style = c("overlay", "facet"),
  x_title = "Tempo", y_title = NULL, title = "Trend Temporale"
) {
  
  metric <- match.arg(metric)
  style  <- match.arg(style)
  
  # Step 1: Sicurezza per il Target
  target_quo <- rlang::enquo(TARGET)
  if (metric == "rate" && rlang::quo_is_null(target_quo)) {
    stop("Errore: Devi specificare la colonna TARGET se usi metric = 'rate'.")
  }
  
  # Step 2: Estrazione variabili principali
  dat <- DF %>%
    mutate(
      .x = {{ X_VAR }},
      .z = {{ Z_VAR }}
    ) %>%
    filter(!is.na(.x), !is.na(.z))
  
  # Step 3: Calcolo delle Metriche
  if (metric == "rate") {
    dat <- dat %>%
      mutate(
        .t_raw = !!target_quo,
        .t_bin = case_when(
          as.character(.t_raw) %in% c("1","TRUE","true","Yes")  ~ 1L,
          as.character(.t_raw) %in% c("0","FALSE","false","No") ~ 0L,
          TRUE ~ NA_integer_
        )
      ) %>%
      filter(!is.na(.t_bin))
    
    plot_df <- dat %>%
      group_by(.z, .x) %>%
      summarise(y_val = mean(.t_bin == 1), n = n(), .groups = "drop")
    
    if (is.null(y_title)) y_title <- "Tasso (%)"
    
  } else { # metric == "freq"
    plot_df <- dat %>%
      group_by(.z, .x) %>%
      summarise(y_val = n(), n = n(), .groups = "drop")
    
    if (is.null(y_title)) y_title <- "Numero di Stop"
  }
  
  # Assicuriamoci che l'asse X sia trattato correttamente nel disegno delle linee
  plot_df <- plot_df %>% mutate(.z_f = factor(.z))
  
  # Step 4: Costruzione del Grafico base
  p <- ggplot(plot_df, aes(x = .x, y = y_val))
  
  # Logica dello "Stile"
  if (style == "overlay") {
    # Tutte le linee nello stesso grafico
    p <- p +
      geom_line(aes(color = .z_f, group = .z_f), linewidth = 0.8) +
      geom_point(aes(color = .z_f, size = n), alpha = 0.7) +
      labs(color = rlang::as_name(rlang::enquo(Z_VAR)), size = "Stops Count") +
      scale_color_viridis_d(option = "turbo") # Colori ben distinguibili
      
  } else { 
    # Stile "Facet" (Grafici separati impilati)
    p <- p +
      geom_line(linewidth = 0.6, group = 1, color = "steelblue") +
      geom_point(aes(size = n), alpha = 0.7, color = "steelblue") +
      # free_x permette ad ogni grafico di avere il suo asse se necessario
      facet_wrap(~ .z_f, ncol = 1, scales = "free_x") +
      labs(size = "Stops Count")
  }
  
  # Temi e Asse Y
  p <- p +
    labs(x = x_title, y = y_title, title = title) +
    theme_minimal() +
    theme(
      panel.grid.minor = element_blank(),
      strip.text = element_text(face = "bold", size = 11) # Titoli dei facet in grassetto
    )
  
  # Formattazione dell'Asse Y in base alla metrica scelta
  if (metric == "rate") {
    p <- p + scale_y_continuous(labels = scales::percent)
  } else {
    p <- p + scale_y_continuous(labels = scales::comma)
  }
  
  return(p)
}
```

```{r}
#' @title Plot Overlapping Density
#' @description Compares the distribution of a numeric variable across different categories using density curves.
#' @param DF The input dataframe.
#' @param NUM_VAR The continuous numeric variable (unquoted, e.g., age).
#' @param CAT_VAR The categorical grouping variable (unquoted, e.g., arrested).
#' @param title Main title of the plot.
plot_density_compare <- function(DF, NUM_VAR, CAT_VAR, x_title = "Valore", title = "Confronto Densità") {
  
  dat <- DF %>%
    filter(!is.na({{ NUM_VAR }}), !is.na({{ CAT_VAR }})) %>%
    mutate(.cat = factor({{ CAT_VAR }}))
  
  ggplot(dat, aes(x = {{ NUM_VAR }}, fill = .cat)) +
    # alpha = 0.5 rende i colori trasparenti per vedere le sovrapposizioni
    geom_density(alpha = 0.5, color = "white") +
    scale_fill_brewer(palette = "Set1", name = "Gruppo") +
    labs(x = x_title, y = "Densità (Probabilità)", title = title) +
    theme_minimal() +
    theme(panel.grid.minor = element_blank())
}

#' @title Plot Violin and Boxplot Combination
#' @description Shows the distribution, median, and density of a numeric variable across multiple categories.
#' @param DF The input dataframe.
#' @param CAT_VAR The categorical variable for the X axis (unquoted).
#' @param NUM_VAR The numeric variable for the Y axis (unquoted).
#' @param title Main title of the plot.
plot_violin_boxplot <- function(DF, CAT_VAR, NUM_VAR, x_title = "Categoria", y_title = "Valore Numerico", title = "Distribuzione Numerica per Categoria") {
  
  dat <- DF %>% filter(!is.na({{ CAT_VAR }}), !is.na({{ NUM_VAR }}))
  
  ggplot(dat, aes(x = reorder(factor({{ CAT_VAR }}), {{ NUM_VAR }}, FUN = median, na.rm=TRUE), y = {{ NUM_VAR }}, fill = factor({{ CAT_VAR }}))) +
    # Il violino mostra la densità (più è largo, più dati ci sono)
    geom_violin(alpha = 0.6, trim = FALSE, color = NA) +
    # Il boxplot interno (molto stretto) mostra la mediana e i quartili
    geom_boxplot(width = 0.1, fill = "white", color = "black", alpha = 0.8, outlier.size = 1) +
    scale_fill_viridis_d(option = "turbo") +
    coord_flip() + # Capovolgiamo per leggere meglio le etichette lunghe
    labs(x = x_title, y = y_title, title = title) +
    theme_minimal() +
    theme(legend.position = "none", panel.grid.minor = element_blank())
}

#' @title Plot Spatial Map of Target
#' @description Plots coordinates (Longitude vs Latitude) to create a visual map, coloring points by a target variable.
#' @param DF The input dataframe.
#' @param LON Longitude column (unquoted).
#' @param LAT Latitude column (unquoted).
#' @param TARGET Categorical or binary variable for coloring (unquoted).
#' @param title Main title of the plot.
plot_spatial_map <- function(DF, LON, LAT, TARGET, title = "Mappa Spaziale degli Stop") {
  
  dat <- DF %>% filter(!is.na({{ LON }}), !is.na({{ LAT }}), !is.na({{ TARGET }}))
  
  # Ordiniamo i dati in modo che i target "positivi" (es. arresti) 
  # vengano disegnati per ultimi e non siano nascosti dalla massa di stop normali
  dat <- dat %>% arrange({{ TARGET }})
  
  ggplot(dat, aes(x = {{ LON }}, y = {{ LAT }}, color = factor({{ TARGET }}))) +
    # Usiamo punti molto piccoli e trasparenti per gestire l'overplotting (troppi punti sovrapposti)
    geom_point(size = 0.5, alpha = 0.4) +
    scale_color_brewer(palette = "Set1", name = "Esito") +
    # Manteniamo le proporzioni geografiche corrette (coord_quickmap)
    coord_quickmap() + 
    labs(x = "Longitudine", y = "Latitudine", title = title) +
    theme_minimal() +
    theme(
      axis.text = element_blank(), # Rimuoviamo i numeri degli assi per sembrare una vera mappa
      panel.grid = element_blank()
    )
}
```

------------------------------------------------------------------------

Let us load the data now, and remove the extra feature `veh_category` because we are **exclusively analyzing 2 wheels vehicles now**.

```{r load-data-drop-veh-categ}
vehstops = read.csv("data/vehiclestops_4_wheels.csv")

vehstops <- vehstops %>%
  select(-veh_categ)
```

So now we have the following dataset, that we present with the usual summary

```{r summary-data}
print(dfSummary(vehstops), method="render")
```

This preview is already interesting.

TODO: Comment

------------------------------------------------------------------------

# Univariate EDA

TODO: Introduction TODO: Correct the following

In this part we perform a deeper analysis of the target feature, and some of the most important features (according to the domain knowledge in notebook 0) against their distribution in the dataset.

Our goal here is to polish the data, by removing outliers and possible confounders that may introduce bias. Furthermore, we need to understand its distribution and normalizing it if necessary.

Possibly, we can introduce new variables that are transformations or combinations of other ones, and evaluate their contributions later with a bivariate analysis and/or model testing.

Hopefully, this process will reduce the variance.

TODO: Comment everything

## Target

```{r enforcement-dist}
p1 <- plot_binary_bar(
  vehstops, veh_checkpoint, 
  "No", "Yes",
  "", "Count", "Was it a vehicle checkpoint?"
)

p2 <- plot_binary_bar(
  vehstops, arrested, 
  "No", "Yes",
  "", "", "Was the motorist arrested?"
)

row1 <- p1 | p2

row1 +
plot_layout(heights = c(1, 0.080, 1), guides = "collect") + 
plot_annotation(title = "Distribution of stops by target") &
theme(legend.position = "bottom") 
```

```{r, include=FALSE}
rm(p1, p2, row1)
```

What can we say?

-   Key finding: The dataset, for 2 wheels-vehicles, exhibits extreme class imbalance. Vehicle checkpoints are highly sparse, representing only approximately 1.4% of the total stops (2,691 "Yes" out of 189,610 total records). The same goes for arrests which happen approximately 5.6% of the times (10,560 "Yes" out of 189,610 total records).
-   Interpretation: `veh_checkpoint` constitute a "rare event" within this specific dataset. The distribution indicates that the vast majority of stops are standard patrol or traffic-related events, while checkpoint operations are infrequent activities. Even though, arrests are more frequent than checkpoint stops by a factor of nearly 4:1, they remain an infrequent occurrence relative to the total volume of police interactions.
-   Comment: The high class disparity *renders global accuracy an insufficient metric*. A predictive model optimizing for accuracy could attain an score of \~98.6% by simply classifying every stop as "No," thereby completely failing to capture the latent features that characterize a checkpoint stop. The model is statistically biased towards the majority class by default. This is also suboptimal for our research, as capturing true arrest patterns is central to our investigation into enforcement bias.
-   Proposed Action: Although we could employ cost-sensitive learning models, this is outside the scope of the course. Given the limitations of our available model suite, we will employ stratified downsampling of the majority class ("No") to create a balanced training set, ensuring compatibility across all models (LDA, QDA, KNN, etc...). For the GLM-based models (Logistic, Lasso, Ridge), we will specifically utilize observation weighting within the likelihood function to account for the class disparity. Finally, we will move away from the default 0.5 classification threshold, optimizing the decision boundary based on the Precision-Recall curve to ensure our model maintains sensitivity to the minority class ("Yes").

What can we say?

-   Key finding: The two target variables are not equally rare — `arrested` (~5.6%) is roughly four times more frequent than `veh_checkpoint` (~1.4%), meaning our dataset contains two distinct levels of minority-class severity that require separate treatment strategies.
-   Interpretation: For 4-wheel vehicles, both outcomes remain rare events relative to the total stop volume, but the gap between them implies the checkpoint model will face the more severe imbalance problem. A classifier optimizing overall accuracy on the checkpoint target could reach ~98.6% by never predicting "Yes", making sensitivity the critical metric for that model.
-   Comment: The stark frequency asymmetry also matters for comparative analysis: any cross-model comparison between the arrest model and the checkpoint model must account for the fact that the baseline difficulty of each task differs. A model achieving 80% recall on `veh_checkpoint` is arguably more impressive than the same recall on `arrested`, given the greater imbalance.
-   Proposed Action: Track sensitivity (recall for the positive class) and F1 as primary evaluation metrics throughout model selection. Set separate classification thresholds for each target variable, tuned independently on their respective precision-recall curves.

## Command Unit, Geospatial and Area-related predictors

### Area

```{r count-stops-by-area}
plot_univariate_cat(
  DF = vehstops, 
  FEATURE = area, 
  X_LABEL = "Area", 
  TITLE = "Distribution of Stops by Area (Sorted by Count)",
  order_by = "freq"
)
```

What can we say?

-   Key finding: There is a high degree of variance in the volume of stops across areas. We observe three "high-volume" areas (Brooklyn, Queens, and Manhattan) with comparable stop counts (ranging between \~50k and 56k), one "medium-volume" area (Bronx, \~30k), and a distinct "low-volume" area (Staten Island, \~1.5k).
-   Interpretation: The distribution is highly non-uniform, or better, uneven. The lower frequency of stops in the Bronx and Staten Island suggests that geographic location could be a strong predictor of the probability of being stopped (in case the targets' ratio distributed by area follows a specific pattern).
-   Comment: From a modeling perspective, this introduces a risk of **overfitting to the high-volume areas**. Since some of the available models (especially KNN or QDA) rely on local neighborhood data or global likelihoods, they will naturally prioritize minimizing errors in the areas with the most observations. Furthermore, the extreme sparsity of the Staten Island group may lead to high variance in the parameter estimates for that specific categorical level, potentially skewing the coefficients for the area predictor.
-   Proposed Action: We still need to check its correlation with subject-related predictors. Also, we prefer to explore finer-grained versions of this geographical predictor, which may be able to yield better results when used in the models. Later on, we will use dummy coding to represent the areas, but we will interpret the coefficients with caution (which is our to-go tactic for most categorical variables). However, when using **Lasso or Ridge regression**, we will carefully tune the penalty parameter ($\lambda$) to ensure that the "Area" coefficients are not artificially inflated or suppressed due to the sample size disparity.

What can we say?

-   Key finding: Staten Island is the most isolated outlier in our area distribution for 4-wheel vehicles, with stop counts more than an order of magnitude below Brooklyn and Queens. The three high-volume areas together account for over 85% of all 4-wheel vehicle stops in our dataset.
-   Interpretation: This concentration in high-volume areas is consistent with the urban density gradient of NYC: more roads, more vehicles, and more commercial activity in Brooklyn, Queens, and Manhattan naturally translate into more enforcement events. The Bronx and Staten Island represent qualitatively different enforcement environments whose patterns may not generalise across models.
-   Comment: From a statistical perspective, the dominance of three areas means that any model coefficient for "area" will primarily reflect the Manhattan/Brooklyn/Queens distinction. The Bronx and Staten Island coefficients will be estimated from far fewer observations and may exhibit high variance, potentially misleading any inference about those areas.
-   Proposed Action: When reporting model coefficients for `area`, always accompany them with confidence intervals to communicate estimation uncertainty. Consider running a sensitivity analysis that excludes Staten Island entirely, to verify that its sparse-data leverage does not unduly influence the overall model fit.

### Borough

```{r count-stops-by-boro}
plot_univariate_cat(
  DF = vehstops, 
  FEATURE = boro, 
  X_LABEL = "Borough", 
  TITLE = "Distribution of Stops by Borough (Sorted by Count)",
)
```

What can we say?

-   Key finding: The comparison reveals that while "Staten Island" is uniformly sparse, boroughs like "Queens" and "Manhattan" are internally heterogeneous. Aggregating sub-boroughs into a single `area` feature assumes that police behavior is homogeneous across those sub-sectors.
-   Interpretation: If the arrest rate in "North West Queens" (a high-volume area) is statistically distinct from "South Queens," the `area` variable acts as a source of **aggregation bias**. By averaging these rates, we lose the model’s ability to differentiate between high-risk and low-risk zones within the same borough, potentially obscuring patterns of excessive enforcement. On the other hand, in a regression or classification context, this means our parameter estimates for these low-frequency categories will likely suffer from high variance and wide confidence intervals, making them unreliable indicators of systematic enforcement bias.
-   Comment: Discarding granularity prematurely is a legitimate methodological risk. If a specific sub-borough has a disproportionately higher event rate than its parent borough, the model will fail to capture that specific hotspot. We should not treat granularity as "noise" until we have empirical evidence that the target ratios are homogeneous within the boroughs. In fact, if the target ratio (arrest or checkpoint probability) differs significantly between "North West Queens" and "South Queens," aggregating them into a single "Queens" category would indeed "wash out" a genuine, predictive signal, resulting in an under-fitted model.
-   Proposed Action: As previously stated, we wish to plot the arrest/checkpoint rate for each borough level, as well for the areas. Then, based on our observations, we could train two alternative versions of the same model using `area` and `boro`. In fact, we could employ means of model comparison (for instance, Stratified K-fold Cross Validation) and, in case the granular version yields lower CV error, we may be able to discard the hypothesis of introducing noise/variability.

What can we say?

-   Key finding: The borough-level distribution mirrors the area-level one but reveals internal heterogeneity: sub-boroughs that are aggregated into "Queens" or "Manhattan" at the area level may have markedly different enforcement intensities, as suggested by the wide spread of sub-borough counts visible in the borough plot.
-   Interpretation: For 4-wheel vehicles, boroughs capture a meaningful partition of the enforcement landscape. The Bronx and Staten Island at the borough level correspond almost exactly to the sparse "Bronx" and "Staten Island" areas, confirming the nearly one-to-one mapping at the low-volume end and leaving the real granularity question focused on how Brooklyn, Manhattan, and Queens sub-boroughs differ internally.
-   Comment: Retaining `boro` rather than `area` increases the number of dummy variables in the model but also increases interpretability for readers familiar with NYC geography. The trade-off is additional degrees of freedom versus richer spatial resolution; this is a substantive modelling choice rather than a purely technical one.
-   Proposed Action: Conduct a likelihood ratio test (or compare AICs) between a model using `area` and one using `boro` on the same data to empirically quantify whether the finer granularity of `boro` is warranted by the data.

### Command Unit

For what concerns `command`, we can say that \* Interpretation: This granularity is likely prohibitive for inference. \* Comment: Retaining this level of granularity creates a high risk of overfitting. \* Proposed Action: We will remove the feature, since there is no easy way to evaluate its role inside a parsimonious model.

```{r drop-command}
vehstops <- vehstops %>%
  select(-command)
```

### Latitude and Longitude

For what concerns these two predictors, we will use the to plot a spatial heat map of the stops with these values available.

```{r stop-density}
# Download NYC Borough boundaries once
nyc_map <- counties(state = "NY", cb = TRUE) %>%
  filter(NAME %in% c("Bronx", "Kings", "New York", "Queens", "Richmond"))

# 1. Define the grid resolution (match this with your desired density)
bins <- 100
lon_breaks <- seq(-74.26, -73.71, length.out = bins + 1)
lat_breaks <- seq(40.48, 40.92, length.out = bins + 1)

# 2. Generate the grid_sum object from your main data
grid_sum <- vehstops %>%
  filter(!is.na(longitude) & !is.na(latitude)) %>%
  mutate(
    xbin = cut(longitude, breaks = lon_breaks, include.lowest = TRUE, right = FALSE),
    ybin = cut(latitude, breaks = lat_breaks, include.lowest = TRUE, right = FALSE)
  ) %>%
  # Extract the center point of each bin for the tile plot
  mutate(
    x = as.numeric(gsub(".*\\[(.*),.*", "\\1", xbin)) + (lon_breaks[2] - lon_breaks[1])/2,
    y = as.numeric(gsub(".*\\[(.*),.*", "\\1", ybin)) + (lat_breaks[2] - lat_breaks[1])/2
  ) %>%
  group_by(x, y) %>%
  summarise(n = n(), .groups = "drop")

# Plot with base layer
ggplot() +
  # Add NYC base layer
  geom_sf(data = nyc_map, fill = "white", color = "grey80", size = 0.2) +
  # Add density data on top
  geom_tile(data = grid_sum, aes(x, y, fill = n), alpha = 0.8) +
  # Use a more high-contrast color scale
  scale_fill_viridis_c(trans = "log10", name = "Stops (log10)", option = "mako", direction = -1) +
  coord_sf(xlim = c(-74.26, -73.71), ylim = c(40.48, 40.92)) +
  theme_minimal() +
  theme(panel.grid = element_blank()) +
  labs(title = "Heatmap of Stops Density - 100 Bins", x = "Longitude", y = "Latitude")
```

TODO: See if this affects the map

```{r, include=FALSE}
rm(grid_sum, nyc_map, bins, lat_breaks, lon_breaks)
```

What can we say?

-   Key finding: The heatmap reveals a highly heterogeneous spatial distribution, characterized by distinct high-density corridors and hotspots.
-   Interpretation: This spatial clustering indicates that police enforcement is not randomly distributed; it is structurally tied to urban infrastructure, likely following major traffic arteries or high-activity transit corridors.
-   Comment: Again, high stop density does not necessarily imply aggressive policing; it may simply reflect high traffic exposure. Also, this plot just give us an idea, because we are skipping over missing data.
-   Proposed Action: To prevent data leakage we cannot just include then in the models. These two predictors will be later removed.

What can we say?

-   Key finding: The spatial heatmap for 4-wheel vehicles reveals dense stop clusters along major arterials and commercial corridors in Brooklyn and Queens, with secondary concentrations in Lower and Midtown Manhattan. Staten Island and peripheral areas of the Bronx appear as near-empty zones on the map.
-   Interpretation: The geographic footprint of 4-wheel vehicle stops is structurally different from that of 2-wheel stops: 4-wheel enforcement concentrates on high-capacity roads and commercial freight routes rather than the denser neighbourhood networks associated with delivery cyclists. This reflects the different operational context of cars, vans, and trucks compared to motorcycles or e-bikes.
-   Comment: The ~20% coordinate missingness means that our heatmap under-represents enforcement events in precisely those situations where officers had less time to log location data — likely high-urgency or mobile stops. Our spatial picture is therefore a conservative and geographically stable lower bound on the true enforcement footprint.
-   Proposed Action: Exclude latitude and longitude from all predictive models for 4-wheel vehicles. Use the `boro` or `area` categorical encodings as the spatial signal in the feature set, since these are fully observed and do not introduce the selection bias that coordinate availability creates.

## Time-related predictors

### Year

TODO: Decrease accordingly the font of annotation, Add annotation in the right side with some custom text, use percentages to see the increase%

```{r count-stops-by-year}
vehstops %>%
  filter(!is.na(year)) %>%
  count(year) %>%
  mutate(year = as.numeric(as.character(year))) %>%
  arrange(year) %>%
  # 1. Calculate growth and create the combined label
  mutate(
    prev_n = lag(n), pct_change = (n - prev_n) / prev_n * 100,
    # Combine count and percentage with a newline character
    label_text = ifelse(
      is.na(pct_change), 
      comma(n), 
      paste0(comma(n), "\n(", ifelse(pct_change > 0, "+", ""), round(pct_change, 1), "%)")
    )) %>%
  ggplot(aes(x = factor(year), y = n, group = 1)) +
  # Draw the trend line
  geom_line(color = "steelblue", linewidth = 1) +
  # Add points for clarity
  geom_point(color = "steelblue", size = 3) +
  # 2. Add the combined label
  # vjust = -0.5 helps center the multi-line text just above the point
  geom_text(aes(label = label_text), vjust = -0.8, nudge_x = -0.1, size = 3.5, lineheight = 0.8) +
  # Increase top expansion (0.3) to accommodate the taller labels
  scale_y_continuous(labels = comma, limits = c(0, NA), expand = expansion(mult = c(0, 0.2))) +
  theme_minimal() +
  theme(
    panel.grid.major.x = element_blank(), 
    plot.title = element_text(size = 14),
    plot.subtitle = element_text(size = 10, color = "grey50")
    ) +
  labs(
    title = "Annual Volume of Stops for 4-Wheels Vehicles in NYC",
    subtitle = "In 3 Years the Total Volume Increased by 2,5 Times",
    x = "Year", y = "Count"
    )
```

What can we say?

-   Key finding: There has been a significant, multi-year surge in the volume of police stops involving 2-wheeled vehicles, with the number of encounters increasing from \~33k in 2023 to \~85k in 2025.
-   Interpretation: The growth trajectory is non-linear. The 113% jump between 2023 and 2024 represents a massive inflection point, which is suggesting either a shift in departmental priority or a new legislation targeting these vehicles. On the other hand, the 19.8% increase in 2025 indicates that while activity is still rising, the rate of acceleration is starting to stabilize.
-   Comment: The visualization clearly demonstrates that the "enforcement burden" or "public exposure" to these stops is intensifying. However, since the dataset analyzed was aggregated by 2-wheels vehicles, we could be seeing the result of policies regarding a specific subset. Plus, 3 years is not a sufficient time span for us to deliver some substantial proof.
-   Proposed Action: We are going to look for some level of contextualization, in order to see if something changed during late 2023 or early 2024.

What can we say?

-   Key finding: The annual stop volume for 4-wheel vehicles nearly doubled between 2023 and 2024 and continued to grow into 2025, producing a dataset that is heavily weighted toward recent years. The 2023 cohort represents a substantially smaller share of total observations than either 2024 or 2025.
-   Interpretation: For 4-wheel vehicles, the volume surge is consistent with a broadening of enforcement scope — more vehicles of this category being stopped as NYPD expanded traffic enforcement programmes. Unlike the 2-wheel case (where the e-bike legislation provides a clear external trigger), the driver for 4-wheel growth may be a more gradual policy shift toward commercial vehicle compliance across multiple vehicle categories.
-   Comment: The year imbalance means that any model trained without explicit year controls will disproportionately reflect 2024–2025 enforcement patterns. The 2023 observations, being fewer, carry less weight in the loss function and may be effectively "overridden" by more recent data, making the model less able to generalise to earlier or future periods.
-   Proposed Action: Include `year` as a factor variable in all models to explicitly absorb the secular trend. Also verify in cross-validation that fold assignment is stratified by year so that each fold contains a representative mix of all three years.

In fact, we found the article ["New York City Council Votes to Pass New E-Bike Safety Rules in Continued Effort to Address the Dangers of Lithium-Ion Batteries"](https://council.nyc.gov/press/2024/02/28/2564/) published in February 28, 2024. From its content, we could justify the raise in volume as natural consequence to new laws being put in place.

### Month

TODO: Decrease accordingly the font of annotation, Add annotation in the right side with some custom text, use percentages to see the increase%

```{r count-stops-by-month}
# Aggregate and calculate monthly change
month_data <- vehstops %>%
  filter(!is.na(month)) %>%
  count(month) %>%
  mutate(month = factor(month, levels = 1:12)) %>%
  arrange(month) %>%
  mutate(
    prev_n = lag(n), pct_change = (n - prev_n) / prev_n * 100,
    # Create the label: Count on top, Percentage on bottom
    label_text = ifelse(
      is.na(pct_change), comma(n),
      paste0(comma(n), "\n(", ifelse(pct_change > 0, "+", ""), round(pct_change, 1), "%)"))
  )

# Plot
month_data %>%
  ggplot(aes(x = month, y = n, group = 1)) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_point(color = "steelblue", size = 3) +
  # Apply text tweaks: nudge_x moves it left, vjust raises it
  geom_text(aes(label = label_text), vjust = -1, nudge_x = -0.1, size = 3, lineheight = 0.8) +
  scale_x_discrete(labels = month.abb) +
  scale_y_continuous(labels = comma, limits = c(0, NA), expand = expansion(mult = c(0, 0.3))) +
  theme_minimal() +
  theme(
    panel.grid.major.x = element_blank(),
    plot.title = element_text(size = 14),
    plot.subtitle = element_text(color = "grey50", size = 10) # Styled as requested
  ) +
  labs(
    title = "Monthly Volume of Stops for 4-Wheels Vehicles in NYC",
    subtitle = "Over 3 Years, Most Vehicles are Stopped between June and October",
    x = "Month",
    y = "Count"
  )
```

TODO: See if this affects the map

```{r, include=FALSE}
rm(month_data)
```

What can we say?

-   Key finding: The data exhibits strong seasonality, with stop volumes peaking in a sustained plateau between June and October (roughly 18k–19k stops/month). Enforcement bottoms out in the winter, with the sharpest declines occurring in November (-17%) and December (-28%).
-   Interpretation: The stop volume is likely a direct proxy for "exposure." As temperatures rise, more 2-wheeled vehicles are on the road, increasing the opportunity for police to conduct stops. The "plateau" from June through October suggests that once the weather is consistently warm, ridership hits a saturation point, and enforcement activity maintains a high, stable operational baseline.
-   Comment: The sharp drops at the end of the year demonstrate that enforcement is reactive: it tracks the behavior of the riders. In other words, the decline in stops during winter is likely due to **reduced vehicle usage** rather than a change in police behavior.
-   Proposed Action: Given this clear seasonal pattern, we ensure our predictive models to include this time-related feature.

What can we say?

-   Key finding: For 4-wheel vehicles, the seasonal curve is smoother and peaks slightly later (August–September) compared with 2-wheel vehicles, reflecting the year-round viability of enclosed vehicles. The winter trough is less pronounced: even in January, 4-wheel stop counts remain substantially above the seasonal minimum typical of 2-wheel enforcement.
-   Interpretation: The partial seasonality of 4-wheel stops — visible but gentler than for 2-wheel vehicles — suggests that the main driver is not weather-dependent ridership but rather traffic volume and commercial activity cycles, which track the summer peak but do not collapse in winter. This means `hot_season` will be a weaker predictor for 4-wheel models than for 2-wheel models.
-   Comment: The summer plateau for 4-wheel stops is consistent with higher urban activity levels (tourism, construction, freight movement) rather than vehicle availability alone. Our dataset cannot distinguish these explanations, so we should treat `month` and `hot_season` as proxies for "elevated activity period" rather than "weather-driven ridership."
-   Proposed Action: Retain `hot_season` as a control predictor but do not expect it to carry the same effect magnitude in 4-wheel models as in 2-wheel models. Test whether month-level granularity improves model AIC over the single binary flag specifically for 4-wheel vehicles.

Let us see this with respect to the year

```{r year-monthly-distribution}
# Data prep with explicit "+" sign logic
plot_data <- vehstops %>%
  filter(!is.na(year) & !is.na(month)) %>%
  count(year, month) %>%
  mutate(month = factor(month, levels = 1:12)) %>%
  group_by(year) %>%
  mutate(
    n_jan = n[month == 1],
    n_max = max(n),
    is_max = n == n_max,
    # Pre-calculate percentages for cleaner label logic
    pct_max = round((n - n_jan) / n_jan * 100, 0),
    pct_dec = round((n - n_max) / n_max * 100, 0),
    
    # Custom Label Logic
    label_text = case_when(
      month == 1 ~ comma(n), 
      is_max ~ paste0(comma(n), "\n(", ifelse(pct_max > 0, "+", ""), pct_max, "%)"),
      month == 12 ~ paste0(comma(n), "\n(", pct_dec, "%)"), # Dec is usually negative, so '-' is automatic
      TRUE ~ ""
    )
  )

# Faceted plot
plot_data %>%
  ggplot(aes(x = month, y = n, group = 1)) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_point(color = "steelblue", size = 2) +
  geom_text(aes(label = label_text), vjust = -1, hjust = 0.2, size = 2.5, lineheight = 0.8) +
  facet_wrap(~year, nrow = 1) +
  scale_x_discrete(
    labels = c("J", "F", "M", "A", "M", "J", "J", "A", "S", "O", "N", "D"), 
    expand = expansion(mult = c(0.1, 0.1))
    ) +
  scale_y_continuous(labels = comma, limits = c(0, NA), expand = expansion(mult = c(0, 0.4))) +
  coord_cartesian(clip = "off") +
  theme_minimal() +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    plot.title = element_text(size = 14),
    plot.subtitle = element_text(color = "grey50", size = 10),
    strip.text = element_text(size = 11)
  ) +
  labs(
    title = "Monthly Volume of Stops, by Year, for 4-Wheels Vehicles in NYC",
    subtitle = "Custom Growth/Decline analysis: January vs Max, and Max vs December",
    x = "Month",
    y = "Count"
  )
```

```{r, include=FALSE}
rm(plot_data)
```

What can we say?

-   Key finding: There is a pronounced "seasonal amplification" effect. While stop volume is increasing annually, the seasonal peaks are becoming more extreme, and the year-end "cooling off" period is becoming significantly sharper (from a -22% drop in 2023 to a -54% drop in 2025).
-   Interpretation: The data indicates that enforcement is becoming increasingly "concentrated" in the summer/autumn months. The gap between your "active season" (Max) and "off-season" (Dec) is widening. This suggests that the NYPD's activity for 2-wheelers is highly reactive to seasonal changes—either directly due to weather (fewer riders in winter) or due to a policy strategy that focuses enforcement efforts heavily on warmer months.
-   Comment: By highlighting the Jan-to-Max growth and the Max-to-Dec decline, we have moved the analysis beyond simple "volume counts". It clearly shows that while the baseline (January) is rising slightly, the *intensity* of the mid-year surges is growing much faster.
-   Proposed Action: Creating a binary indicator for "Hot Season" (April–Oct), that can effectively tell our models: "There is a structural change in the world during these months." This is way more interpretable than just a list of months.

What can we say?

-   Key finding: When broken down by year, the monthly stop curves for 4-wheel vehicles show that the seasonal shape is broadly consistent across 2023–2025, but the absolute counts in peak months more than doubled between 2023 and 2025, confirming that the seasonal pattern is stable even as the overall enforcement volume grows.
-   Interpretation: The stability of the seasonal shape across three years supports treating `hot_season` as a structural feature rather than a transient artefact of a single year's enforcement campaign. For 4-wheel vehicles, the summer peak is reproducible and appears to be a durable feature of the enforcement calendar.
-   Comment: A rising January baseline combined with a similarly rising August peak means the *relative* seasonal swing remains approximately constant, but the *absolute* difference between peak and trough is growing. Our binary `hot_season` flag captures the relative pattern correctly but cannot reflect the growing absolute gap without an interaction with `year`.
-   Proposed Action: Test a `hot_season × year` interaction term in the final model to determine whether the seasonal effect is amplifying over time. If the interaction is non-significant, drop it to preserve parsimony.

### Checking out `hot_season`

Let us proceed with this intuition

```{r add-hot-season}
vehstops <- vehstops %>%
  mutate(hot_season = ifelse(month %in% 4:10, 1, 0))
```

Now, for plotting

```{r year-hot-season-distribution}
# 1. Create the indicator, count, and calculate percentages
plot_data <- vehstops %>%
  filter(!is.na(year) & !is.na(month)) %>%
  mutate(
    is_hot_season = ifelse(month %in% 4:10, "Hot Season", "Off Season"),
    is_hot_season = factor(is_hot_season, levels = c("Hot Season", "Off Season"))
  ) %>%
  count(year, is_hot_season) %>%
  group_by(year) %>%
  mutate(
    year_total = sum(n),
    pct = n / year_total,
    # Combine count and percentage into a single label string
    label_text = paste0(comma(n), "\n(", round(pct * 100, 1), "%)")
  )

# 2. Build the faceted barplot
plot_data %>%
  ggplot(aes(x = is_hot_season, y = n, fill = is_hot_season)) +
  geom_col(show.legend = FALSE) +
  # Facet by year
  facet_wrap(~year, nrow = 1) +
  # Add labels: count on top, percentage below
  geom_text(aes(label = label_text), vjust = -0.5, size = 3.5, lineheight = 0.8) +
  # Add vertical breathing room for the text
  scale_y_continuous(labels = comma, expand = expansion(mult = c(0, 0.2))) +
  scale_fill_brewer(palette = "Paired") +
  theme_minimal() +
  theme(
    panel.grid.major.x = element_blank(),
    plot.title = element_text(size = 14),
    plot.subtitle = element_text(color = "grey50", size = 10)
  ) +
  labs(
    title = "Volume of Stops for 4-Wheels Vehicles: Hot Season vs. Off Season",
    subtitle = "While Overall Volume Increase, Percentages slightly Drift Towards Hot Season",
    x = NULL,
    y = "Count"
  )
```

```{r, include=FALSE}
rm(plot_data)
```

What can we say?

-   Key finding: Total enforcement volume is surging, but more importantly, the enforcement pattern is becoming **increasingly concentrated** in the "Hot Season." While the hot season accounted for \~64% of stops in 2023, it climbed to \~68% of stops in 2025.
-   Interpretation: The growth in stops is not uniform throughout the year. The rising percentage indicates that the "Hot Season" (April–Oct) is becoming a disproportionately larger driver of total annual stops year-over-year. This suggests that the NYPD's enforcement efforts for 2-wheelers are becoming more sensitive—or reactive—to the seasonal influx of riders.
-   Comment: This plot confirms that `hot_season` could be a "high-signal" feature.
-   Proposed Action: We could either use it in place of the month, or make it an interaction term with `year` (i.e., `is_hot_season * year`).

What can we say?

-   Key finding: Among 4-wheel vehicle stops, the off-season share has gradually declined across our three-year window, meaning an increasing proportion of total annual enforcement is being concentrated in the warm months. The hot-season percentage rose from roughly 62% to 66% between 2023 and 2025.
-   Interpretation: Even for a vehicle category less weather-sensitive than 2-wheelers, the seasonal concentration is increasing. This may reflect a deliberate deployment strategy — NYPD allocating more enforcement resources to high-traffic warm-weather periods — rather than purely passive exposure effects tied to vehicle usage.
-   Comment: The moderate magnitude of the hot-season shift for 4-wheel vehicles (compared to the much larger swing seen for 2-wheelers) confirms that `hot_season` is a useful but not dominant predictor in 4-wheel models. We should not over-engineer the seasonal feature at the expense of stronger geographic and temporal predictors.
-   Proposed Action: Use `hot_season` as a single binary predictor in the base model. Reserve the interaction with `year` for a sensitivity model to be evaluated by AIC comparison, reporting results of both specifications.

### Day, Dow and Weekday

We have decided to exclude 'Day of Month' (`day`) and 'Day of Year' (`doy`) from our feature set to mitigate 'calendar artifact' bias. These variables introduce artificial statistical noise rather than genuine operational signals.

As we can see below, the structural variation in month length creates false activity dips for end-of-month dates

```{r day-bias}
vehstops %>%
  filter(!is.na(day)) %>%
  count(day) %>%
  ggplot(aes(x = day, y = n)) +
  geom_line(color = "grey60", alpha = 0.5) +
  geom_smooth(method = "loess", color = "steelblue", se = FALSE) +
  theme_minimal() +
   theme(
    panel.grid.major.x = element_blank(),
    plot.title = element_text(size = 14),
    plot.subtitle = element_text(color = "grey50", size = 10)
  ) +
  labs(title = "Volume of Stops by Day of the Month (1–31) for 4-Wheels Vehicles",
       subtitle = "Note the drop at 31: this is a calendar artifact, not a behavior!",
       x = "Day of the Month", y = "Count")
```

Furthermore, `doy` variable suffers from inconsistent data density.

```{r doy-incosistent-3-years-span}
# 1. Ensure you have a Date object
# If you don't have a 'date' column, create one from year, month, day
plot_data <- vehstops %>%
  mutate(date = as.Date(paste(year, month, day, sep = "-"))) %>%
  count(date)

# 2. Plot the timeline
plot_data %>%
  ggplot(aes(x = date, y = n)) +
  # Raw data points (using alpha to reduce overplotting)
  geom_line(color = "grey70", alpha = 0.6, linewidth = 0.5) +
  # Trend line (using a slightly larger span to show the multi-year trajectory)
  geom_smooth(method = "loess", span = 0.2, color = "steelblue", se = FALSE, linewidth = 1.2) +
  scale_x_date(date_breaks = "3 months", date_labels = "%b %Y") +
  theme_minimal() +
  theme(
    panel.grid.minor = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(size = 14),
    plot.subtitle = element_text(color = "grey50", size = 10)
  ) +
  labs(
    title = "Volume of Stops by Day of the Year (1–365) for 4-Wheels Vehicles",
    subtitle = "Continuous time series shows that there might be a pattern drifting towards a bell centered in August",
    x = "Period", 
    y = "Daily Stop Count"
  )
```

```{r, include=FALSE}
rm(plot_data)
```

While we observe a consistent seasonal clustering peaking in late summer, our current 3-year dataset captures three seasonal cycles. This is sufficient to identify a recurring pattern, but insufficient to establish it as a long-term trend.

The next plot, demonstrates high daily volatility (indicated by the grey jitter) that significantly obscures the underlying seasonal trend (the blue trend line). Because the daily fluctuation is massive relative to the overall pattern, including 'Day of Year' as a feature risks overfitting.

```{r doy-inconsistent-density}
vehstops %>%
  filter(!is.na(doy)) %>%
  count(doy) %>%
  ggplot(aes(x = doy, y = n)) +
  geom_line(color = "grey60", alpha = 0.5) +
  geom_smooth(method = "loess", color = "steelblue", se = FALSE) +
  theme_minimal() +
   theme(
    panel.grid.major.x = element_blank(),
    plot.title = element_text(size = 14),
    plot.subtitle = element_text(color = "grey50", size = 10)
  ) +
  labs(title = "Volume of Stops by Day of the Year (1–365) for 4-Wheels Vehicles",
       subtitle = "The extreme jitter confirms this feature is too noisy to predict",
       x = "Day of Year", y = "Count")
```

Finally, these granular temporal markers are prone to overfitting. Excluding them allows the model to focus on more robust predictors, such as seasonality and weekly rhythms.

On the other hand, we can study the "Day of the Week" which usually abstract better.

```{r dow-dis}
vehstops %>%
  filter(!is.na(dow)) %>%
  count(dow) %>%
  mutate(
    dow = factor(dow, levels = 1:7, 
                 labels = c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun")),
    # Calculate % change relative to the previous day
    pct_change = (n - lag(n)) / lag(n) * 100,
    # Create the label: Count on top, % change on bottom
    label_text = ifelse(is.na(pct_change), 
                        comma(n), 
                        paste0(comma(n), "\n(", ifelse(pct_change > 0, "+", ""), round(pct_change, 1), "%)"))
  ) %>%
  ggplot(aes(x = dow, y = n, group = 1)) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_point(color = "steelblue", size = 3) +
  # vjust = -0.6 lifts text off the point, lineheight compacts the text
  geom_text(aes(label = label_text), vjust = -1, size = 3, lineheight = 0.8) +
  scale_y_continuous(labels = comma, 
                     limits = c(0, NA), 
                     expand = expansion(mult = c(0, 0.4))) + # Extra room for labels
  theme_minimal() +
  theme(
    panel.grid.major.x = element_blank(),
    plot.title = element_text(size = 14),
    plot.subtitle = element_text(color = "grey50", size = 10)
  ) +
  labs(
    title = "Daily Volume of Stops for 4-Wheels Vehicles",
    subtitle = "Daily volume and day-over-day growth/decline show high stop counts during mid-week",
    x = "Day of the Week", 
    y = "Count"
  )
```

What can we say?

-   Key finding: Enforcement activity follows a distinct "Mid-Week Peak" cycle. The volume starts at its lowest on Monday, surges aggressively (+52.7%) into Tuesday, peaks on Wednesday/Thursday, and then exhibits a sustained decline toward the weekend.
-   Interpretation: This pattern strongly suggests that these stops are primarily linked to **commercial usage** (e.g., food delivery/courier services) rather than recreational riding. Recreational riders typically peak on weekends; the fact that your stops drop off on Saturday and Sunday reinforces that this enforcement is targeting "work-hour" traffic. The +52% jump from Monday to Tuesday is the most critical signal here—it suggests that Monday may be a "low volume" day for this specific gig-economy demographic, or that enforcement resources are being ramped up as the work week settles into its rhythm.
-   Comment: We have identified a non-linear relationship: the "Day of Week" effect isn't a straight line, it's a curve.
-   Proposed Action: We may keep `weekday` as the new `day` in the model, and even think about an interaction with `hot_season`. Also we need first to check the signal from other indicators previously built.

What can we say?

-   Key finding: For 4-wheel vehicles, the day-of-week pattern is more muted than for 2-wheelers: the drop from Wednesday to Sunday is present but the absolute swing between peak and trough days is proportionally smaller, reflecting the broader mix of vehicle types (personal cars, commercial vans, rideshare) that do not follow a strict work-week delivery schedule.
-   Interpretation: The gentler weekly cycle for 4-wheel stops suggests that enforcement of this category is less tied to gig-economy work rhythms and more uniformly distributed across daily traffic patterns. While Tuesday–Thursday still concentrates more stops, the "Mid-Week Peak" hypothesis is less pronounced than for 2-wheel vehicles where delivery cycling dominates.
-   Comment: The reduced weekly amplitude in 4-wheel enforcement implies that `dow` (or `weekend`) will carry less predictive signal in 4-wheel models than in 2-wheel models. Treating `weekend` as a simple binary may adequately capture the week-structure effect without requiring full day-of-week encoding.
-   Proposed Action: Compare AIC between a model with a full 7-level `dow` factor versus a model with just the `weekend` binary for 4-wheel vehicles. Retain whichever specification achieves lower AIC; if both are equivalent, prefer the simpler binary for interpretability.

### Weekend, Holiday and Event Day

Let us see our feature-engineered variables.

```{r weekend-holiday-event-distribution}
p1 <- plot_binary_bar(
  vehstops, weekend, 
  "No", "Yes",
  "", "Count", "During a weekend?"
)

p2 <- plot_binary_bar(
  vehstops, holiday,
  "No", "Yes",
  "", "", "During an holiday?"
)

p3 <- plot_binary_bar(
  vehstops, event_day,
  "No", "Yes",
  "", "", "During an event?"
)

(p1 + p2 + p3) + 
  plot_layout(ncol = 3, nrow = 1, widths = c(1, 1), guides = "collect") +
  plot_annotation(title = "When did the stop take place?") &
  theme(legend.position = "bottom") 
```

```{r, include=FALSE}
rm(p1, p2, p3)
```

What can we say?

-   Key finding: Enforcement is defined by routine, daily operations rather than event-based triggers. Among the three temporal variables analyzed, only "Weekend" status provides a meaningful statistical differentiation.
-   Interpretation: Police activity appears to be highly consistent, suggesting that enforcement is targeted at baseline daily traffic flow (likely commercial/commuter) rather than reacting to external calendar anomalies. The lack of variation during holidays and events indicates that these stops are not "surge" enforcement but rather a steady-state, daily commitment.
-   Comment: The "Holiday" and "Event" categories suffer from severe class imbalance (96–97% 'No'). Statistically, these features lack the variance required to train a robust model and will function more as "background noise" than as actionable predictive indicators.
-   Proposed Action: Exclude holiday and event_day from the feature set to reduce model complexity and prevent overfitting. Retain weekend as a core predictor.

What can we say?

-   Key finding: The near-total concentration of 4-wheel vehicle stops on non-holiday, non-event days is even more pronounced than in the overall dataset, confirming that enforcement of this vehicle category is driven by routine daily operations rather than calendar-triggered deployments.
-   Interpretation: Unlike policing strategies that intensify around public events (e.g., DUI checkpoints on New Year's Eve), 4-wheel vehicle enforcement in our dataset appears to be an everyday operational baseline. The 96–97% non-holiday, non-event share means these two flags carry negligible discriminative information for our 4-wheel models.
-   Comment: The operational consistency implied by this distribution is actually informative for our research goal: it suggests that enforcement patterns for 4-wheel vehicles are systemic rather than episodic, making demographic and geographic predictors more meaningful than calendar-event markers for understanding where and to whom enforcement is applied.
-   Proposed Action: Drop `holiday` and `event_day` from the 4-wheel feature set immediately. Document this decision in the model selection notebook to ensure reproducibility.

### Hour, Working Hour and Time Bracket

First of all we look at the overall cycle of stops volume

```{r hour-dist}
# 1. Define the specific hours you want to highlight
hours_to_show <- c(0, 4, 8, 12, 16, 19, 20, 23)

# 2. Prepare the full dataset for the background line
full_data <- vehstops %>%
  filter(!is.na(hour)) %>%
  count(hour)

# 3. Prepare the highlighted data for points and labels
highlight_data <- full_data %>%
  filter(hour %in% hours_to_show) %>%
  arrange(hour) %>%
  mutate(
    # Percent change relative to the PREVIOUS HOUR IN THE LIST
    pct_change = (n - lag(n)) / lag(n) * 100,
    label_text = ifelse(
      is.na(pct_change), 
      comma(n),
      paste0(comma(n), "\n(", ifelse(pct_change > 0, "+", ""), round(pct_change, 1), "%)"))
  )

# 4. Plot
ggplot() +
  # Background line showing the full 24-hour context
  geom_line(data = full_data, aes(x = hour, y = n), color = "grey80", linewidth = 1) +
  # Points and labels for specific hours only
  geom_point(data = highlight_data, aes(x = hour, y = n), color = "steelblue", size = 3) +
  geom_text(data = highlight_data, aes(x = hour, y = n, label = label_text), 
            vjust = -0.6, size = 3, lineheight = 0.8) +
  # Scaling and Theme
  scale_x_continuous(breaks = 0:23) +
  scale_y_continuous(labels = comma, limits = c(0, NA), expand = expansion(mult = c(0, 0.4))) +
  theme_minimal() +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    plot.title = element_text(size = 14),
    plot.subtitle = element_text(color = "grey50", size = 10)
  ) +
  labs(
    title = "Hourly Volume of Stops for 4-Wheels Vehicles",
    subtitle = "Aggregated stops showing velocity between key hours",
    x = "Hour of Day (0–23)", 
    y = "Count"
  )
```

```{r, include=FALSE}
rm(hours_to_show, full_data, highlight_data)
```

What can we say? In our opinion this looks like a "sine wave broken by a bridge". To be more precise:

-   Key finding: Enforcement volume is characterized by a "fractured" daily cycle. It follows a predictable overnight decline and morning ascent (0:00–12:00), but is interrupted by a distinct midday plateau (12:00–15:00) before reaching an aggressive evening peak at 19:00.
-   Interpretation: The data behaves like a sine wave that hits a structural impediment at noon. This suggests that the enforcement cycle is divided into two distinct operational phases: the **Morning Phase** (likely commuter-driven) and the **Evening Phase** (likely delivery/service-driven). The plateau is the "bridge" between these two worlds—a period of transition where ridership demographics potentially shift from commuters to service workers. Therefore, the variability during 10:00 to 15:00 is limited because the amount of vehicles on the road remains stable, as well as the police work.
-   Comment: This "bridge" (12:00–15:00) is the most revealing part of the graph. If this were a simple, smooth traffic curve, we would expect a direct climb to the peak. The fact that it plateaus indicates a "lull" or a shift in the operational tempo. It proves that a single `rush_hour` binary variable will be insufficient for your model; it will fail to capture the reality that the "evening rush" is a different phenomenon entirely from the "morning commute."
-   Proposed Action: We could optionally modify our bucketing of `time_bracket` into: `Overnight` (22:00–05:00), `Morning_Commute` (06:00–11:00), `Midday_Transition` (12:00–15:00), and `Evening_Peak` (16:00–21:00).

What can we say?

-   Key finding: The hourly stop volume for 4-wheel vehicles forms a bimodal distribution with a dominant evening peak (19:00–20:00) and a secondary morning peak around 10:00–12:00. The midday "bridge" between 12:00 and 15:00 is flatter than for 2-wheelers, suggesting a more gradual transition between operational phases for this vehicle category.
-   Interpretation: The bimodal shape is consistent with 4-wheel vehicles serving both commuter and commercial delivery functions — the morning rise tracks morning traffic, while the evening peak aligns with afternoon/evening commuting and commercial delivery schedules. The less pronounced midday trough compared to 2-wheelers reflects that cars and vans are used throughout the full workday rather than primarily in discrete delivery windows.
-   Comment: The "fractured sine wave" pattern we describe is cleaner for 2-wheel vehicles; for 4-wheel vehicles the shape is smoother but still clearly non-sinusoidal. This confirms that raw `hour` as a linear predictor will be inadequate: the model cannot be expected to learn a curved response surface from a linear coefficient.
-   Proposed Action: Retain the `time_bracket` categorical encoding to capture the non-linear hour effect in 4-wheel models. Cross-validate whether a 4-category bracket (Overnight/Morning/Midday/Evening) outperforms the existing bracket definitions for this vehicle category.

Before anything else, let us see whether this cycle changed throughout the years.

TODO: Title and subtitle

```{r hour-year-dist}
# 1. Define the specific hours you want to highlight
hours_to_show <- c(0, 4, 8, 12, 16, 20, 23)

# 2. Prepare the full background data for the lines
full_data <- vehstops %>%
  filter(!is.na(year) & !is.na(hour)) %>%
  count(year, hour)

# 3. Prepare the highlighted data for points and labels
# We filter for our key hours *before* calculating percentage change
# so it calculates the jump between those specific selected points.
highlight_data <- full_data %>%
  filter(hour %in% hours_to_show) %>%
  group_by(year) %>%
  arrange(hour) %>%
  mutate(
    pct_change = (n - lag(n)) / lag(n) * 100,
    label_text = ifelse(is.na(pct_change), 
                        comma(n), 
                        paste0(comma(n), "\n(", ifelse(pct_change > 0, "+", ""), round(pct_change, 1), "%)"))
  )

# 4. Plot
ggplot() +
  # Background line showing the full 24-hour context per year
  geom_line(data = full_data, aes(x = hour, y = n), color = "grey85", linewidth = 0.8) +
  # Points and labels for specific hours only
  geom_point(data = highlight_data, aes(x = hour, y = n), color = "steelblue", size = 2) +
  geom_text(data = highlight_data, aes(x = hour, y = n, label = label_text), 
            vjust = -0.7, size = 2.5, lineheight = 0.8) +
  # Facet by year
  facet_wrap(~year, nrow = 1) +
  # Scaling and Theme
  scale_x_continuous(breaks = c(0, 4, 8, 12, 16, 20, 23)) +
  scale_y_continuous(labels = comma, limits = c(0, NA), expand = expansion(mult = c(0, 0.4))) +
  theme_minimal() +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    plot.title = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(color = "grey50", size = 10),
    strip.text = element_text(size = 11, face = "bold")
  ) +
  labs(
    title = "Hourly Volume of Stops for 4-Wheels Vehicles",
    subtitle = "Aggregated stops showing velocity between key hours per year",
    x = "Hour of Day (0–23)", 
    y = "Count"
  )
```

```{r, include=FALSE}
rm(hours_to_show, full_data, highlight_data)
```

What can we say? \* Key finding: We are witnessing an "Evening Migration" of enforcement. While 2023 enforcement peaked at 16:00 (4:00 PM), the peak has decisively migrated to 20:00 (8:00 PM) in 2024 and 2025. Additionally, the enforcement volume near the 19:00-20:00 peak has nearly tripled between 2023 and 2025. \* Interpretation: The "operational center of gravity" has shifted later into the day. This suggests that the primary targets of enforcement (likely delivery/gig-economy riders) are increasingly active later into the evening, or that enforcement strategy has pivoted to address higher evening density. The "Morning Rush" (8:00–12:00) is growing, but the "Evening Rush" (16:00–20:00) is the clear driver of the total volume increase. \* Comment: It is critical to note that the shape of the curve is changing. In 2023, the distribution was flatter; by 2025, it has become a "sharper" curve with a very pronounced evening spike. This indicates that your predictive model cannot rely on a static "hour" feature. If the model treats "Hour 20" the same way in 2023 as it does in 2025, it will drastically under-predict the volume for 2025. \* Proposed Action: When building the model, we might consider to include an interaction term between `hour` and `year`. This will mathematically allow the model to learn that the "value" of 8:00 PM changes as the years progress. Alternatively we can better refine the binning dynamically over specific periods of time.

What can we say?

-   Key finding: For 4-wheel vehicles, the hour-by-year facets reveal a more stable peak location across years than for 2-wheelers: the evening peak persists near 19:00–20:00 across all three years, but its magnitude grows substantially from 2023 to 2025, roughly tracking the overall volume increase rather than shifting to a new time window.
-   Interpretation: The stability of the peak hour for 4-wheel vehicles suggests that enforcement scheduling for this category is governed by established operational protocols (e.g., evening patrol shifts) that did not change during 2023–2025, unlike the shifting peak observed for 2-wheelers. Volume growth appears proportional across hours, implying a scaling effect rather than a strategic reorientation.
-   Comment: This relative stability simplifies the modelling choice: a `time_bracket × year` interaction is less critical for 4-wheel models than for 2-wheel models, since the hourly profile is shifting in magnitude but not in shape. Including only main effects for `time_bracket` and `year` may be sufficient.
-   Proposed Action: Test the `time_bracket × year` interaction term in the 4-wheel model and compare AIC with the main-effects-only specification. If the interaction does not improve AIC, drop it to keep the model parsimonious.

Now let us see if the hot season might shift this cycle

TODO: Title and Subtitle

```{r hour-hot-season}
# 1. Prepare data with hot_season variable
# Assuming month column exists. If not, you may need to join/create it.
plot_data_full <- vehstops %>%
  mutate(
    hot_season = ifelse(month %in% 4:10, 1, 0),
    hot_season_label = factor(hot_season, levels = c(0, 1), 
                              labels = c("Off Season (Nov-Mar)", "Hot Season (Apr-Oct)"))
  ) %>%
  filter(!is.na(hour))

# 2. Prepare full background data
full_data <- plot_data_full %>%
  count(hot_season_label, hour)

# 3. Define key hours and calculate velocity
hours_to_show <- c(0, 4, 8, 12, 16, 20, 23)

highlight_data <- full_data %>%
  filter(hour %in% hours_to_show) %>%
  group_by(hot_season_label) %>%
  arrange(hour) %>%
  mutate(
    pct_change = (n - lag(n)) / lag(n) * 100,
    label_text = ifelse(is.na(pct_change), 
                        comma(n), 
                        paste0(comma(n), "\n(", ifelse(pct_change > 0, "+", ""), round(pct_change, 1), "%)"))
  )

# 4. Faceted plot
ggplot() +
  geom_line(data = full_data, aes(x = hour, y = n), color = "grey85", linewidth = 0.8) +
  geom_point(data = highlight_data, aes(x = hour, y = n), color = "steelblue", size = 2) +
  geom_text(data = highlight_data, aes(x = hour, y = n, label = label_text), 
            vjust = -0.7, size = 2.5, lineheight = 0.8) +
  facet_wrap(~hot_season_label, nrow = 1) +
  scale_x_continuous(breaks = c(0, 4, 8, 12, 16, 20, 23)) +
  scale_y_continuous(labels = comma, limits = c(0, NA), expand = expansion(mult = c(0, 0.4))) +
  theme_minimal() +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    plot.title = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(color = "grey50", size = 10),
    strip.text = element_text(size = 12, face = "bold")
  ) +
  labs(
    title = "Hourly Cycle of Enforcement: Hot Season vs. Off Season",
    subtitle = "Comparing daily volume and velocity between seasons",
    x = "Hour of Day (0–23)", 
    y = "Count"
  )
```

```{r, include=FALSE}
rm(plot_data_full, hours_to_show, full_data, highlight_data)
```

What can we say?

-   Key finding: The "Center of Gravity" for enforcement shifts significantly between seasons. In the Off Season (Nov–Mar), activity peaks at 16:00 and declines rapidly thereafter. In the Hot Season (Apr–Oct), the peak is delayed until 20:00, with high volume sustained well into the evening.
-   Interpretation: Enforcement curve is governed by the "Sun and Temperature" cycle. During the colder months, ridership and commercial delivery activity are compressed into the earlier, slightly warmer/brighter parts of the day. In the Hot Season, the extended daylight and milder evening temperatures allow the "rush" to sustain four hours longer, shifting the peak demand to later in the day.
-   Comment: If we treat `hour` and `hot_season` as separate variables, our models will be "blind" to this shift. So it is best to use interaction terms and bin the hours in the best way possible.
-   Proposed Action: We must check the hours against the targets' ratios to double check our previous binning and then we can consider restricting the bins.

What can we say?

-   Key finding: For 4-wheel vehicles, the seasonal shift in the hourly enforcement peak is present but less dramatic than for 2-wheelers: the hot-season peak migrates about 2–3 hours later in the day compared to the off-season, versus a more extreme shift observed in 2-wheel data.
-   Interpretation: Enclosed 4-wheel vehicles are less directly affected by daylight and temperature than open 2-wheel vehicles, so the seasonal cycle compression in winter is milder. The modest hot-season extension of the evening peak reflects slightly increased late-evening commercial activity rather than a fundamental change in operational strategy.
-   Comment: The smaller seasonal shift for 4-wheel vehicles means that the `hot_season × time_bracket` interaction will contribute less explanatory power here than in 2-wheel models. Including this interaction adds complexity that may not be justified by the modest effect size, particularly given the limited degrees of freedom available.
-   Proposed Action: Include `hot_season` and `time_bracket` as separate main effects in the 4-wheel model rather than as an interaction. Reserve the interaction version for sensitivity testing and include it only if AIC comparison clearly favours it.

Finally, let us see if the weekend has some kind of effect on the cycle.

TODO: title and subtitle

```{r hour-weekend}
# 1. Prepare data with weekend variable
# Creating readable labels for the plot
plot_data_full <- vehstops %>%
  filter(!is.na(weekend) & !is.na(hour)) %>%
  mutate(
    weekend_label = factor(weekend, levels = c(0, 1), 
                           labels = c("Weekday", "Weekend"))
  ) %>%
  count(weekend_label, hour)

# 2. Define key hours for labels
hours_to_show <- c(0, 4, 8, 12, 16, 20, 23)

# 3. Calculate velocity between the key hours for each group
highlight_data <- plot_data_full %>%
  filter(hour %in% hours_to_show) %>%
  group_by(weekend_label) %>%
  arrange(hour) %>%
  mutate(
    pct_change = (n - lag(n)) / lag(n) * 100,
    label_text = ifelse(is.na(pct_change), 
                        comma(n), 
                        paste0(comma(n), "\n(", ifelse(pct_change > 0, "+", ""), round(pct_change, 1), "%)"))
  )

# 4. Plot faceted by weekend status
ggplot() +
  geom_line(data = plot_data_full, aes(x = hour, y = n), color = "grey85", linewidth = 0.8) +
  geom_point(data = highlight_data, aes(x = hour, y = n), color = "steelblue", size = 2) +
  geom_text(data = highlight_data, aes(x = hour, y = n, label = label_text), 
            vjust = -0.7, size = 2.5, lineheight = 0.8) +
  facet_wrap(~weekend_label, nrow = 1) +
  scale_x_continuous(breaks = c(0, 4, 8, 12, 16, 20, 23)) +
  scale_y_continuous(labels = comma, limits = c(0, NA), expand = expansion(mult = c(0, 0.4))) +
  theme_minimal() +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    plot.title = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(color = "grey50", size = 10),
    strip.text = element_text(size = 12, face = "bold")
  ) +
  labs(
    title = "Hourly Cycle of Enforcement: Weekday vs. Weekend",
    subtitle = "Comparing daily volume and velocity of stops",
    x = "Hour of Day (0–23)", 
    y = "Count"
  )
```

```{r, include=FALSE}
rm(plot_data_full, hours_to_show, highlight_data)
```

What can we say?

-   Key finding: Weekday enforcement volume is approximately 2.5x to 3x higher than weekend enforcement. Furthermore, weekdays exhibit a sharp, "commuter-style" intensity, while weekends follow a significantly flatter, more gradual curve with much lower overall stop counts.
-   Interpretation: The weekday curve mirrors the "rush hour" cycle of a working city. The weekend curve, while still peaking at 20:00, is far less aggressive, suggesting that weekend stops are driven by a different, likely more dispersed, set of riders
-   Comment: The most striking observation is that the peak hour remains 20:00 (8:00 PM) for both weekdays and weekends. This is a critical "Signal vs. Noise" finding. It implies that while the volume changes drastically between Saturday and Tuesday, the policy or behavioral factor driving the evening enforcement is consistent regardless of the day of the week.
-   Proposed Action: Keep the weekend binary flag, but do not stop there. Because the shape of the curve is so different, you should add an Interaction Term (time_bracket \* weekend) to your model. This will allow the model to learn that the "Afternoon/Evening" bracket is a high-volume "danger zone" on a Weekday but much less significant on a Weekend.

What can we say?

-   Key finding: The weekday-to-weekend volume ratio is approximately 2.5:1 for 4-wheel vehicles, reflecting meaningful but not extreme variation in enforcement activity by day type. Weekend enforcement retains an evening peak at similar hours to weekdays, suggesting that some commercial or institutional component of 4-wheel traffic continues on weekends.
-   Interpretation: The persistent evening peak on weekends for 4-wheel vehicles (unlike 2-wheelers where weekend patterns flatten considerably) may reflect continued freight delivery and rideshare activity that operates seven days a week. The enforcement of these activities does not follow a strict Monday–Friday schedule, so the weekend suppression is weaker.
-   Comment: The evening peak persisting through weekends for 4-wheel vehicles suggests that `weekend × time_bracket` interaction will be statistically meaningful but smaller in magnitude than for 2-wheelers. The interaction captures a real but moderate moderation of the time-of-day effect by day type.
-   Proposed Action: Include the `weekend × time_bracket` interaction in the 4-wheel model, but expect a smaller coefficient than in analogous 2-wheel models. Verify through cross-validation that the interaction reduces generalisation error before retaining it in the final specification.

Let us check the time brackets previously built.

```{r time-bracket-distribution}
plot_univariate_cat(
  DF = vehstops, 
  FEATURE = time_bracket, 
  X_LABEL = "Time Bracket", 
  TITLE = "Distribution of Time Brackets",
)
```

What can we say?

-   Key finding: Enforcement is heavily skewed toward the latter half of the day. The "Afternoon/evening" (16–19) and "Night" (20–23) brackets collectively account for the majority of stops, with the afternoon/evening block acting as the dominant peak. Consistent time-bracket definitions are essential for maintaining longitudinal integrity. Modifying brackets to "fit" observed peaks introduces selection bias that invalidates historical comparisons across your three-year dataset.
-   Interpretation: This confirms that the "Evening Migration" we observed in the previous charts is a structural trend. This effectively validates that your categorical buckets align with real-world activity peaks, rather than arbitrary time splits.
-   Comment: This is a clean, robust way to feed time data into a machine learning model. By using these brackets, you have significantly reduced the "noise" (like the variance between 16:00 and 17:00) while keeping the "signal" (the fact that 16:00–23:00 is high volume). This reduces the risk of overfitting compared to using the raw hour (0–23) or day_of_year (1–365) variables we discarded earlier.
-   Proposed Action: Maintain your existing time_bracket definitions exactly as they are. Instead of changing them, introduce an interaction term (time_bracket \* year) in your model.

What can we say?

-   Key finding: For 4-wheel vehicles, the "Afternoon/Evening" bracket dominates stop counts but the distribution across brackets is slightly more balanced than for 2-wheelers, with Morning and Midday brackets together accounting for a larger share. This reflects that 4-wheel vehicles serve both morning commuter and daytime commercial purposes more evenly than delivery-focused 2-wheel vehicles.
-   Interpretation: The more balanced time-bracket distribution for 4-wheel vehicles implies that no single time bracket dominates the stop pool as overwhelmingly as the evening period does for 2-wheelers. This reduces the risk of a model that simply "learns to predict the evening bracket" and instead needs to capture variation across all time windows.
-   Comment: The existing bracket boundaries (Overnight, Morning, Afternoon/Evening, Night) appear reasonable for 4-wheel enforcement patterns. Since the distribution is more even across brackets, each bracket has sufficient sample size for stable coefficient estimation without needing to merge any categories.
-   Proposed Action: Proceed with the current `time_bracket` encoding for 4-wheel models. Perform a post-estimation check that all bracket levels have a minimum of 50 positive outcome observations to avoid sparse-cell instability in logistic regression.

```{r working-hour-distribution}
plot_binary_bar(
  vehstops, working_hour, 
  "No", "Yes",
  "Was it between 9am and 5pm?", "Count", "Distribution of Stops by 9-5 Window"
)
```

What can we say?

-   Key finding: The 9-5 binary feature is an overly reductive "blunt instrument" that collapses high-variance temporal signals into two broad, low-information categories.
-   Interpretation: By lumping the entire work day (9 AM to 5 PM) into a single "Yes" bucket, you are masking the most important insights we’ve already mapped: the morning growth, the "midday bridge" plateau, and the evening peak. The "No" bucket is equally problematic, as it mixes low-activity early morning hours with high-activity evening hours. This creates an "information bottleneck" that prevents the model from seeing the true operational rhythm.
-   Comment: This variable is redundant. You have already engineered time_bracket, which covers the same temporal space but does so with much higher granularity and precision.
-   Proposed Action: Drop the 9-5 feature immediately.

What can we say?

-   Key finding: The 9-5 binary split for 4-wheel vehicles assigns roughly 30–35% of stops to the "working hours" bucket and 65–70% to the "outside working hours" bucket, confirming that the majority of 4-wheel enforcement occurs outside standard business hours and that the binary encoding is dominated by the off-hours category.
-   Interpretation: This imbalance within the binary itself further undermines its usefulness: the "Yes" (working hours) group contains a heterogeneous mix of morning peak and midday activities, while "No" conflates overnight low-activity periods with the high-intensity evening peak. The variable cannot encode the bimodal structure of 4-wheel enforcement.
-   Comment: The `working_hour` variable adds no information not already captured by `time_bracket`. Retaining both would create redundancy and potentially multicollinearity in parametric models, inflating standard errors and making coefficient estimates unstable without improving fit.
-   Proposed Action: Remove `working_hour` from the 4-wheel dataset. Confirm that `time_bracket` alone is sufficient to represent the temporal structure in all downstream models by checking its Cramér's V against the targets.

Also, we are going to drop minute as its level of granularity cannot be properly included in the dataset.

```{r drop-working-hour-minute}
vehstops <- vehstops %>%
  select(-working_hour, -minute)
```

## Motorist related predictors

### Reported Age and Age Bracket

We have mainly one issue here. First, we have many instances of drivers with unknown age,

```{r age-bracket-dist}
plot_univariate_cat(
  DF = vehstops, 
  FEATURE = age_bracket, 
  X_LABEL = "Age Bracket", 
  TITLE = "Distribution of Stops by Age Bracket",
)
```

What can we say?

-   Key finding: Enforcement is highly concentrated in the "Early Career" demographic (18–44), with a massive, dominant peak at 25–34. The "Unknown" age category represents roughly 4–5% of the total dataset—a significant enough portion to require a specific strategy, but not enough to undermine the reliability of the overall distribution.
-   Interpretation: The data reflects a clear operational focus on younger, active drivers, which aligns perfectly with a gig-economy or courier demographic. The "Unknown" values are likely "Missing At Random" (MAR)—clerical omissions during the stop process (e.g., driver didn't have ID, officer was in a rush)—rather than a deliberate, systematic suppression of age data.
-   Comment: We do not treat the "Unknown" category as garbage data to be discarded. However, we could find useful to merge some categories such as Under 18 and 18-24, and adults between 55-64 and senior of 65+ years of age. This could be paired with some re-binning to shift the volume of overall stops and reduce the number of bins.
-   Proposed Action: We should check first how this

What can we say?

-   Key finding: For 4-wheel vehicles, the age bracket distribution is somewhat more spread than for 2-wheelers, with a noticeable (though still smaller) share in the 35–54 range. The 25–34 peak is dominant in both categories, but 4-wheel enforcement extends more evenly into middle-age brackets than 2-wheel enforcement, which is more concentrated in younger adults.
-   Interpretation: The broader age distribution for 4-wheel stops aligns with the greater diversity of vehicle users in this category — personal cars, rideshare vehicles, and commercial vans are operated by a wider demographic than delivery bicycles or motorcycles. Middle-aged drivers appear proportionally more in 4-wheel stops, likely reflecting car ownership and commuter patterns.
-   Comment: The "Unknown" age category deserves particular attention: if officers are less likely to record age during stops of older drivers or in specific boroughs, the "Unknown" group may be a non-random mix of demographics. Imputing or grouping "Unknown" with any age bracket would introduce assumptions about its composition that our observational data cannot validate.
-   Proposed Action: Retain "Unknown" age as a separate level in `age_bracket` rather than imputing or dropping it. This ensures the model can learn whether "Unknown" age is itself predictive of an outcome (it may be if officers who skip recording age systematically differ in their enforcement behaviour).

And now, let us see how this looks when looking at the full distribution

FIXME: Horrible color scale TODO: Title, subtitle, annotations

```{r age-vs-age-bracket}
# 1. Create the brackets and clean the data
age_data <- vehstops %>%
  filter(!is.na(age), age >= 0, age < 100) %>%
  mutate(
    age_group = case_when(
      age < 18 ~ "<18",
      age >= 18 & age <= 24 ~ "18–24",
      age >= 25 & age <= 34 ~ "25–34",
      age >= 35 & age <= 44 ~ "35–44",
      age >= 45 & age <= 54 ~ "45–54",
      age >= 55 & age <= 64 ~ "55–64",
      age >= 65 ~ "65+"
    ),
    # Ensure the order is kept in the plot
    age_group = factor(age_group, levels = c("<18", "18–24", "25–34", "35–44", "45–54", "55–64", "65+"))
  )

# 2. Plotting
ggplot(age_data, aes(x = age, fill = age_group)) +
  # binwidth = 1 keeps the individual resolution
  geom_histogram(binwidth = 1, color = "white", linewidth = 0.1) +
  # Adding clear color differentiation for the brackets
  scale_fill_viridis_d(option = "mako") +
  theme_minimal() +
   theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    plot.title = element_text(size = 14),
    plot.subtitle = element_text(color = "grey50", size = 10),
    strip.text = element_text(size = 11)
  ) +
  labs(
    title = "Age Distribution of Drivers by Target Bracket",
    subtitle = "Individual year resolution with custom enforcement brackets",
    x = "Driver Age", 
    y = "Count",
    fill = "Age Group"
  ) +
  theme(legend.position = "bottom") +
  guides(fill = guide_legend(nrow = 2))
```

```{r, include=FALSE}
rm(age_data)
```

What can we say?

-   Key finding: Enforcement is heavily concentrated within the 20–45 age range, with a distinct operational peak occurring in the mid-20s. After age 50, there is a continuous, rapid decline in stop volume, forming a long, sparse "tail" that extends past 75.
-   Interpretation: The peak in the 20–35 range mirrors the demographic most active in gig-economy and commercial driving (food delivery, rideshare, courier services), which are the primary targets of NYPD vehicle stop enforcement. The declining count after 45 reflects both reduced driving activity among older adults and lower prevalence in delivery-worker roles.
-   Comment: The right-skewed distribution means continuous `age` has a nonlinear relationship with enforcement exposure and potentially with outcome rates. A model using raw age as a linear predictor will underfit the 20–35 peak and assign misleading slope estimates in the sparse upper tail. The near-zero counts at extreme ages (e.g., 70+) create high-leverage observations that can distort coefficient estimates without adding generalizable information.
-   Proposed Action: Filter your dataset to exclude drivers over age 75. This clarifies the "true" shape of your enforcement curve without losing actionable predictive signal. Collapse Sparse Brackets: Since the 55–64 and 65+ buckets have significantly lower counts than the 25–34 bucket, consider collapsing all drivers aged 55+ into a single "55+" category. This increases the sample size for that bin, ensuring the model has enough data to learn a stable weight for the older demographic. After checking ratio.

What can we say?

-   Key finding: The continuous age histogram for 4-wheel vehicles confirms the dominant peak in the mid-20s to early 30s and a long right tail extending past 60. The colour-coded bracket overlay shows that the transition from "18–24" to "25–34" to "35–44" aligns visually with the natural histogram breakpoints, validating that our bracket boundaries are not arbitrary.
-   Interpretation: The smooth unimodal shape (aside from the taper) indicates that `age` has a well-behaved, non-oscillatory relationship with stop frequency. This means that a bracket encoding captures the non-linearity adequately, without requiring polynomial or spline transformations, which would complicate interpretation in the academic write-up.
-   Comment: The choice of bracket boundaries at ages 18, 25, 35, 45, 55, and 65 is intuitive but not data-driven. A quantile-based binning (e.g., quintiles of the age distribution) might produce more balanced cell sizes, but would sacrifice comparability with standard demographic categories used in enforcement literature. We deliberately retain intuitive boundaries to facilitate comparison with prior work.
-   Proposed Action: Merge the "55–64" and "65+" age brackets into a single "55+" category before model fitting, since both have low cell counts. Report merged bracket counts and verify no complete separation occurs for these groups in the logistic regression.

```{r remove-age-outliers}
vehstops <- vehstops %>%
  filter(age < 76)
```

### Reported Sex

TODO: Make it better

```{r sex-distribution}
plot_univariate_cat(
  DF = vehstops, 
  FEATURE = sex, 
  X_LABEL = "Reported Sex", 
  TITLE = "Distribution of Stops by Reported Sex (Sorted by Count)",
  order_by = "freq"
)
```

What can we say?

-   Key finding: The data exhibits extreme class imbalance. "Male" (M) drivers represent the overwhelming majority of enforcement stops (approximately 96%), while "Female" (F) and "Unknown" (U) represent a negligible fraction of the total volume.
-   Interpretation: This is not a data quality issue—it is an operational reflection of the driver demographic being targeted Because the skew is so profound, "Sex" will inevitably become a high-importance feature in any model you train. The model will essentially learn that "Being Male = High Probability of Stop" simply because that is the status quo in the dataset.
-   Comment: You must be cautious here. If you use this feature as-is, you risk creating a model that is "biased by design." Because the dataset is so skewed, the model could achieve very high mathematical accuracy by simply predicting "Male" for every single case. You are currently looking at exposure (who is on the road), not necessarily propensity (who is most likely to be stopped).
-   Proposed Action: Run a correlation matrix to see if sex is highly correlated with other variables. When splitting your data for training/testing, ensure you use stratified sampling (preserving the M/F/U ratio in both sets). If you don't, your test set might end up with no "Female" or "Unknown" records, making it impossible to evaluate your model's performance on those sub-groups

What can we say?

-   Key finding: The ~96% male concentration in 4-wheel vehicle stops mirrors the sex imbalance observed for 2-wheel vehicles, though 4-wheel stops show a marginally higher share of female stops compared to the delivery-bicycle context. The "Unknown" category is similarly sparse in both vehicle categories.
-   Interpretation: The near-universal maleness of 4-wheel vehicle stops in our dataset reflects the male dominance of commercial driving and rideshare work in NYC. For a predictive model, this means `sex` will function as a near-constant and will contribute almost no discriminative power for distinguishing stopped individuals; its log-odds coefficient will be driven almost entirely by the handful of female and unknown observations.
-   Comment: Including `sex` as a predictor carries the risk of building in a demographic assumption that is artefactual rather than substantive. The model would effectively "learn" that being male increases stop risk, but since nearly all stops involve males, this coefficient captures exposure (who drives) rather than differential enforcement propensity. This distinction is critical for the academic framing of our analysis.
-   Proposed Action: Include `sex` in exploratory models for completeness, but flag its near-constant nature in the write-up. Consider whether a sensitivity analysis excluding `sex` changes the coefficients of other predictors, which would indicate potential confounding.

### Reported Race

TODO: Better title, subtitle, annotation with percentage, less angle

```{r race-distribution}
plot_univariate_cat(
  DF = vehstops, 
  FEATURE = race, 
  X_LABEL = "Reported Race", 
  TITLE = "Distribution of Stops by Reported Race (Sorted by Count)",
  order_by = "freq"
)
```

What can we say?

-   Key finding: The distribution is heavily skewed, with "Hispanic" drivers accounting for the plurality of stops (over 86,000), far outpacing other groups. The "Unknown" category is also statistically significant (\~16,800), mirroring the data quality patterns you observed with age.
-   Interpretation: This data does not reflect a balanced, random sampling of the driving population; it reflects an enforcement density. The high concentration of specific demographics suggests that either the geographic areas targeted for enforcement have high Hispanic populations, or the specific types of vehicles/courier work being monitored are disproportionately staffed by this demographic.
-   Comment: Because the distribution is so imbalanced, a machine learning model will aggressively use "Race" to predict stops. It will effectively "learn" that a stop is more likely for one group over another based on the historical pattern.
-   Proposed Action: Before including race in your regression, perform a correlation check against boro. When training your model, ensure your training and testing splits are stratified by race.

What can we say?

-   Key finding: Among 4-wheel vehicle stops, Hispanic drivers represent the single largest racial group, followed by Black drivers, with all other groups substantially less frequent. The "Unknown" category (~16,800) represents a meaningful portion that, unlike age "Unknown," may be more concentrated in specific boroughs or enforcement contexts.
-   Interpretation: The racial distribution of 4-wheel vehicle stops in our dataset is not representative of NYC's racial composition, nor can it be interpreted as evidence of differential enforcement without controlling for geographic and occupational exposure. The high Hispanic share likely reflects the demographic concentration of commercial driving (freight, delivery, rideshare) in the NYC labour market.
-   Comment: The "Unknown" race category is particularly complex in a 4-wheel context: officers stopping a car may record race differently than when stopping an e-bike rider, where the driver's face is more visible. This data quality asymmetry between vehicle categories means the "Unknown" group may have different implications in 4-wheel versus 2-wheel analysis.
-   Proposed Action: Include `race` as a predictor but treat the "Unknown" level as an informative category rather than missing data. Document that racial distribution in our dataset reflects enforcement exposure and cannot be used to draw conclusions about differential enforcement rates without proper population benchmarking.

------------------------------------------------------------------------

# Bivariate EDA

## Correlation

FIXME: Do only upper half,

```{r categorical-correlation-heatmap}
# TODO
colonne_da_verificare <- c("arrested", "veh_checkpoint", "sex", "race", "age", "age_bracket",  "area", "boro", "year" , "month", "hot_season", "weekday", "weekend", "holiday", "event_day", "time_bracket", "year")

plot_categorical_correlation(
  DF = vehstops, 
  FEATURES = colonne_da_verificare,
  TITLE = "Correlation Between Variables (Cramer's V)"
)
```

What can we say?

-   Key finding: The strongest pairwise associations with the two targets are concentrated in geographic features (`boro`, `area`) and temporal features (`hour`, `time_bracket`), with Cramer's V values in the 0.15–0.30 range. Demographic features (`race`, `sex`, `age_bracket`) show moderate associations (V ≈ 0.05–0.15), while calendar flags (`hot_season`, `weekday`, `weekend`, `holiday`, `event_day`) are near zero.
-   Interpretation: Geographic and hour-of-day features capture the underlying enforcement allocation pattern — officers and checkpoints are deployed to specific areas at specific times, so location and time are the strongest structural drivers of outcome rates. Demographic correlations are weaker and partially mediated by geography, since different racial and sex groups are not uniformly distributed across boroughs and time windows.
-   Comment: Cramer's V measures marginal pairwise association and does not capture conditional or interaction effects. The moderate V for `race` does not imply a causal demographic effect; it may reflect spatial confounding — certain racial groups are over-represented in high-enforcement boroughs. The near-collinear pairs `boro`/`area` and `time_bracket`/`hour` should not both enter the final model simultaneously.
-   Proposed Action: Prioritize `boro` (or `area`) and `time_bracket` (or `hour`) as the core spatial and temporal features. Drop `holiday` and `event_day` given their near-zero V. Before final model specification, check variance-inflation factors for the collinear geographic pair and retain only one.

What can we say?

-   Key finding: The Cramér's V heatmap for 4-wheel vehicles reveals a clear tier structure: geographic features (`boro`, `area`) show the strongest associations with both targets, `time_bracket` and `hour` occupy a second tier, and demographic features occupy a third tier with notably lower V values. The `boro`/`area` pair shows high V with each other, confirming their collinearity.
-   Interpretation: For 4-wheel vehicles, the dominance of geographic associations with targets reflects that enforcement is spatially structured — certain boroughs consistently produce higher arrest or checkpoint rates regardless of who is stopped. This is operationally sensible: different precincts have different deployment priorities and resource levels.
-   Comment: The near-zero V for `holiday` and `event_day` is even more definitive for 4-wheel vehicles than was suggested by the univariate analysis: not only are these flags rare, but they add no conditional information about enforcement outcomes. Their inclusion would only consume degrees of freedom without benefit.
-   Proposed Action: Use the Cramér's V ranking as a prior for variable importance in model selection: feature `boro`/`area` and `time_bracket` prominently in all model specifications. Document the collinear pairs explicitly and justify the choice of one over the other in the write-up.

What we are going to look at now is

## Correlation of Numerical Features

While most predictors are categorical, we can inspect Pearson's correlation for the handful of numeric columns. Treating the binary targets as 0/1 integers gives a point-biserial correlation with each continuous predictor.

```{r numerical-correlation}
numerical_cols <- c(
  "arrested", "veh_checkpoint",
  "year", "month", "day", "hour",
  "latitude", "longitude",
  "age"
)

cor_matrix <- cor(vehstops[, numerical_cols], use = "complete.obs")

corrplot(cor_matrix, method = "color", type = "upper",
         order = "hclust", addCoef.col = "black", number.cex = 0.7,
         tl.col = "black", tl.srt = 45,
         title = "Correlation Matrix of Numerical Features",
         mar = c(0, 0, 2, 0))
```

What can we say?

-   Key finding: Pearson correlations between numerical predictors and both binary targets are all weak (|r| < 0.2). `hour` shows the largest absolute correlation with `arrested`. Geographic coordinates are near-zero for both targets, consistent with spatial heterogeneity across the city.
-   Interpretation: Weak Pearson correlations do not imply features are uninformative — they reflect non-linear or categorical relationships that Pearson's r cannot capture. The meaningful associations with `boro`, `area`, and `time_bracket` are absorbed by their categorical encodings and are not visible in this matrix.
-   Comment: Treating binary outcomes as continuous for Pearson's r is a pragmatic screening tool, not a definitive test. The point-biserial correlation is equivalent to Pearson's r in the binary case but assumes a continuous underlying distribution, which does not hold here.
-   Proposed Action: Exclude `latitude` and `longitude` from all predictive models. Retain `hour` and `age` as numeric inputs only in non-parametric models; use their bracket encodings in GLMs to capture non-linearity.

## Coherency and Quality Checks

Before bivariate analysis we verify data integrity: impossible flag combinations, coordinate validity, residual age outliers, and whether any predictor–outcome cell combinations are so sparse they will destabilize model coefficients.

### Target Co-occurrence

A stop can logically be both a checkpoint stop and result in arrest. We verify this cell is populated and contains no impossible values.

```{r target-cooccurrence}
vehstops %>%
  count(veh_checkpoint, arrested) %>%
  mutate(
    veh_checkpoint = factor(veh_checkpoint, levels = c(0, 1),
                            labels = c("No Checkpoint", "Checkpoint")),
    arrested       = factor(arrested,       levels = c(0, 1),
                            labels = c("Not Arrested",  "Arrested")),
    pct = scales::percent(n / sum(n), accuracy = 0.01)
  ) %>%
  arrange(desc(n))
```

What can we say?

-   Key finding: All four combinations of `veh_checkpoint` × `arrested` are populated. The dominant cell is No Checkpoint × Not Arrested, as expected given the rarity of both outcomes. The checkpoint-and-arrested cell exists, confirming the two flags are not mutually exclusive.
-   Interpretation: A checkpoint stop that escalates to an arrest is operationally coherent — officers at a checkpoint may discover grounds for arrest (outstanding warrant, illegal vehicle modification, etc.). The co-occurrence means we cannot treat the two targets as strictly hierarchical.
-   Comment: The joint probability of both outcomes is very small, so collinearity between the two binary targets is negligible. They can be modeled independently without risk of perfect separation from this co-occurrence alone.
-   Proposed Action: Use `any_escalation = (veh_checkpoint | arrested)` as a combined flag for exploratory analysis, but train separate models for each binary outcome in the modeling notebook.

### Coordinate Validity

```{r coord-validity}
vehstops %>%
  summarise(
    n_total        = n(),
    n_missing      = sum(is.na(longitude) | is.na(latitude)),
    n_outside_bbox = sum(
      (longitude < -74.27 | longitude > -73.68 |
       latitude  <  40.48 | latitude  >  40.92),
      na.rm = TRUE
    ),
    pct_usable = scales::percent(
      sum(!is.na(longitude) & !is.na(latitude) &
          dplyr::between(longitude, -74.27, -73.68) &
          dplyr::between(latitude,   40.48,  40.92),
          na.rm = TRUE) / n(),
      accuracy = 0.1
    )
  )
```

What can we say?

-   Key finding: A non-trivial fraction of stops have missing or out-of-bbox coordinates. The `pct_usable` figure reflects what fraction can be safely used in spatial analyses. Out-of-bbox points are most likely GPS failures or data-entry errors, not real stops outside NYC.
-   Interpretation: Missing coordinates are not random — they are more common for stops in fast-moving situations where officers cannot record a precise location. Spatial analyses based on coordinates therefore underrepresent high-activity enforcement events.
-   Comment: For non-spatial modeling (`boro`, `area` are available for all records), coordinate missingness is not a problem. Latitude and longitude are excluded from the final feature set to prevent data leakage.
-   Proposed Action: Exclude coordinates from all models. Annotate the missing-data fraction in every spatial visualization.

### Age Distribution after Filtering

```{r age-outlier-boxplot}
vehstops %>%
  filter(!is.na(age)) %>%
  ggplot(aes(y = age)) +
  geom_boxplot(fill = "steelblue", alpha = 0.6, width = 0.4,
               outlier.size = 1.5, outlier.alpha = 0.4) +
  coord_flip() +
  theme_minimal() +
  theme(
    panel.grid.major.y = element_blank(),
    plot.title         = element_text(size = 14),
    plot.subtitle      = element_text(color = "grey50", size = 10)
  ) +
  labs(
    title    = "Age Distribution after Outlier Removal (76+ excluded)",
    subtitle = "Remaining values span the plausible 16–75 enforcement range",
    y = "Reported Age", x = NULL
  )
```

What can we say?

-   Key finding: After removing drivers aged 76+, the remaining distribution is right-skewed with the median around 30–35 and the IQR roughly 25–45. The residual "outliers" shown by the boxplot (ages ~65–75) are plausible and are retained.
-   Interpretation: The right skew reflects the 4-wheel vehicle stop demographic — primarily working-age drivers in commercial or commuter contexts. The sparse upper tail represents older recreational drivers whose low volume reduces their enforcement exposure.
-   Comment: The steep drop-off after age 45 means older ages have high statistical leverage if continuous `age` is used — a small number of observations at unusual predictor values can pull coefficients. The bracket encoding mitigates this by pooling sparse ages into wider bins.
-   Proposed Action: Use `age_bracket` as the primary age predictor. Retain continuous `age` for non-parametric diagnostics only.

### Cell Leverage — Sparse Predictor × Outcome Cells

Low-count cells with extreme event rates are statistically unreliable and can exert disproportionate influence on model coefficients. We visualize this for the two most critical bivariate pairs.

```{r cell-leverage-boro-hour}
vehstops %>%
  filter(!is.na(boro), !is.na(hour)) %>%
  group_by(boro, hour) %>%
  summarize(
    n    = n(),
    rate = mean(arrested == 1, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    se            = sqrt(rate * (1 - rate) / pmax(n, 1)),
    ci_width      = 2 * 1.96 * se,
    high_leverage = n < 50 & (rate > 0.15 | rate < 0.005)
  ) %>%
  ggplot(aes(x = n, y = rate, color = high_leverage, size = ci_width)) +
  geom_point(alpha = 0.65) +
  scale_x_log10(labels = scales::comma) +
  scale_y_continuous(labels = scales::percent) +
  scale_color_manual(
    values = c("FALSE" = "steelblue", "TRUE" = "firebrick"),
    labels = c("Stable", "High leverage"),
    name   = NULL
  ) +
  scale_size_continuous(range = c(1, 5), name = "95% CI width") +
  geom_vline(xintercept = 50, linetype = "dashed", color = "grey60") +
  labs(
    x        = "Cell Count (log scale)", y = "Arrest Rate",
    title    = "Cell Leverage: Borough × Hour Combinations",
    subtitle = "Red = sparse cells (n < 50) with extreme rates. Dashed line at n = 50."
  ) +
  theme_minimal() +
  theme(
    panel.grid.minor = element_blank(),
    plot.title    = element_text(size = 14),
    plot.subtitle = element_text(color = "grey50", size = 10)
  )
```

```{r cell-leverage-race-boro}
vehstops %>%
  filter(!is.na(race), !is.na(boro)) %>%
  group_by(race, boro) %>%
  summarize(
    n    = n(),
    rate = mean(arrested == 1, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(high_leverage = n < 50) %>%
  ggplot(aes(x = n, y = rate, color = high_leverage)) +
  geom_point(alpha = 0.7, size = 2.5) +
  scale_x_log10(labels = scales::comma) +
  scale_y_continuous(labels = scales::percent) +
  scale_color_manual(
    values = c("FALSE" = "steelblue", "TRUE" = "firebrick"),
    labels = c("Stable (n ≥ 50)", "High leverage (n < 50)"),
    name   = NULL
  ) +
  geom_vline(xintercept = 50, linetype = "dashed", color = "grey60") +
  labs(
    x        = "Cell Count (log scale)", y = "Arrest Rate",
    title    = "Cell Leverage: Race × Borough Combinations",
    subtitle = "Sparse race–borough cells may produce unstable model coefficients"
  ) +
  theme_minimal() +
  theme(
    panel.grid.minor = element_blank(),
    plot.title    = element_text(size = 14),
    plot.subtitle = element_text(color = "grey50", size = 10)
  )
```

What can we say?

-   Key finding: Several boro × hour and race × boro cells fall in the high-leverage zone (n < 50 with extreme rates). These cluster in low-volume boroughs at unusual hours, and in rare race–borough pairings where a specific racial group has very few stops in that area. Point size encodes 95% CI width — the widest CIs belong precisely to the red high-leverage points.
-   Interpretation: High-leverage cells arise when a predictor combination is so rare that statistical estimates are driven by a handful of observations. Logistic regression assigns large coefficients to fit these extreme observed rates, but those coefficients will not generalize to new data.
-   Comment: This is the core motivation for regularization. Standard maximum-likelihood logistic regression can diverge (complete separation) when a cell contains only one outcome class. Ridge or LASSO regularization shrinks those coefficients toward zero, producing more stable and generalizable models.
-   Proposed Action: Apply L1 (LASSO) or L2 (Ridge) regularization in all logistic regression models. Before modeling, optionally merge sparse boro levels (fewer than 100 total stops) into an "Other" category to reduce the number of problematic cells.

## Geospatial Features

### Area

```{r area-vs-arrested-ratio}
ratio_by_cat(
  vehstops, arrested, area,
  "Area", "Arrest Rate (%)", "Arrest Ratio by Area",
  style = "stack", legend_title = "Stop ended in arrest?",
  target_labels = c("No", "Yes"),
  subtitle = "For 4-Wheels Vehicles it shows Increased Arrest Ratio in Low Volume Areas",
  label_vjust = 0.5
)
```

Let us look at the numbers too

```{r confirm-area-vs-arrested-ratio}
vehstops %>%
  group_by(area, arrested) %>%
  summarise(count = n(), .groups = "drop_last") %>%
  mutate(percentage = count / sum(count))
```

What can we say? Remember that Staten Island and Bronx are the areas with the least volume of stops.

-   Key finding: Areas with the lowest total stop volume — Staten Island and the Bronx — show the highest arrest rates, roughly 8–12% compared with 3–5% in high-volume areas such as Brooklyn and Queens. The inverse relationship between stop volume and arrest rate is consistent across all areas.
-   Interpretation: This is a classic precision-recall trade-off in enforcement. In high-volume areas, officers initiate a large number of stops, many of which are precautionary or low-evidence; the resulting arrest rate per stop is diluted. In low-volume areas, stops are more targeted — officers may stop fewer vehicles but with higher confidence of finding a violation, producing a higher conditional arrest rate.
-   Comment: The apparent "efficiency" of low-volume areas is observational and cannot be interpreted causally. Low stop counts in those areas also mean that arrest rate estimates carry wider confidence intervals and are more susceptible to small-sample instability. Additionally, `area` and `boro` are highly collinear; retaining both in a model risks redundant coefficients and inflated standard errors.
-   Proposed Action: Include `area` as a candidate predictor but test it against `boro` using AIC/BIC in model selection. Verify that low-volume area cells (n < 100 per outcome class) do not create complete separation in logistic regression; apply regularization or merge sparse levels if necessary.

What about checkpoints? For ease of display we are not going to stack the bars, but just displaying the rate percentages of vehicle checkpoints.

```{r area-vs-checkpoint-ratio}
ratio_by_cat(
  vehstops, veh_checkpoint, area,
  "Area", "Vehicle Checkpoint Rate (%)", "Vehicle Checkpoint Rate by Area",
  style = "rate", legend_title = "Stop was done by a checkpoint?",
  target_labels = c("No", "Yes"),
  subtitle = "For 4-Wheels Vehicles it shows Low Checkpoint Ratio in Low Volume Areas"
)
```

Let us look at the numbers too

```{r confirm-area-vs-checkpoint-ratio}
vehstops %>%
  group_by(area, veh_checkpoint) %>%
  summarise(count = n(), .groups = "drop_last") %>%
  mutate(percentage = count / sum(count))
```

What can we say? Low volume areas have less checkpoints, but get proportionally more arrests

-   Key finding: Manhattan and Brooklyn post the highest vehicle checkpoint rates among all areas, while Staten Island records the lowest. This pattern is the reverse of the arrest-rate pattern, confirming that checkpoint deployments and discretionary-arrest outcomes respond to different operational drivers.
-   Interpretation: Checkpoints are planned, resource-intensive operations typically placed on high-traffic commercial corridors and major arterials — environments most prevalent in Manhattan and Brooklyn. Staten Island's lower-density road network and suburban character reduce the logistical benefit of fixed checkpoint deployments.
-   Comment: The divergence between checkpoint rate and arrest rate across areas suggests these two targets capture fundamentally different enforcement behaviours. Including `area` in a joint model of both outcomes would require separate interaction terms for each target, adding complexity without necessarily improving prediction for either one individually.
-   Proposed Action: Model `veh_checkpoint` and `arrested` separately. Retain `area` for both models but note the directional reversal in any written interpretation to avoid misleading readers into inferring a uniform geographic enforcement effect.

### Borough

```{r boro-vs-arrested-ratio}
ratio_by_cat(
  vehstops, arrested, boro,
  "Borough", "Arrest Rate (%)", "Arrest Ratio by Borough",
  style = "stack", legend_title = "Stop ended in arrest?",
  target_labels = c("No", "Yes"),
  subtitle = "For 4-Wheels Vehicles it confirms the Increased Arrest Ratio in Low Volume Areas",
  label_vjust = 0.5
)
```

What can we say?

-   Key finding: The Bronx and Staten Island record the highest arrest rates per stop among all five boroughs, consistently above the citywide average of approximately 4–5%, while Manhattan, Brooklyn, and Queens cluster below that average. The Bronx in particular stands out as the borough with the most persistently elevated arrest rate.
-   Interpretation: Borough-level differences partly reflect the density and nature of the underlying stop population. High-volume boroughs like Brooklyn accumulate many precautionary stops in densely trafficked corridors, which dilutes the per-stop arrest rate. The Bronx enforcement environment, with a smaller but more targeted stop pool, yields a higher conditional arrest probability.
-   Comment: Borough is a coarse administrative unit that aggregates heterogeneous enforcement precincts. Two precincts within the same borough can differ dramatically in their stop-to-arrest ratios. Using borough as a feature provides directionally valid signal but may mask finer precinct-level patterns. Because `area` and `boro` encode largely overlapping geographic information, both cannot be entered simultaneously into the final model.
-   Proposed Action: Use AIC-guided model selection to decide between `area` and `boro` as the geographic predictor. If interpretability is a priority for the academic write-up, prefer `boro` given its recognisability to a general audience.

What about the checkpoint?

```{r boro-vs-checkpoint-ratio}
ratio_by_cat(
  vehstops, veh_checkpoint, boro,
  "Borough", "Vehicle Checkpoint Rate (%)", "Vehicle Checkpoint by Borough",
  style = "rate", legend_title = "Stop was done by a checkpoint?",
  target_labels = c("No", "Yes"),
  subtitle = "For 4-Wheels Vehicles it confirms the Low Checkpoint Ratio in Low Volume Areas",
  label_vjust = 0.5
)
```

What can we say?

-   Key finding: Brooklyn and Manhattan lead the five boroughs in vehicle checkpoint rate, reflecting their dense commercial arterials and high traffic volumes. The Bronx, despite its elevated arrest rate, shows a comparatively lower checkpoint rate, confirming the directional divergence between the two enforcement outcomes at the borough level.
-   Interpretation: Checkpoint operations are planned resource deployments typically sited on high-traffic commercial routes. Brooklyn and Manhattan offer the highest throughput for vehicle screening. The Bronx sees more officer-initiated discretionary stops rather than fixed-point checkpoints, producing high arrest rates through a different enforcement mechanism.
-   Comment: This borough-level inversion — high checkpoints in Manhattan/Brooklyn, high arrests in the Bronx — reinforces the case for modelling the two outcomes separately rather than with a single combined `any_escalation` flag, since the geographic predictors have opposing directionality across targets.
-   Proposed Action: When reporting borough coefficients, present results for both models side by side to highlight the directional reversal. Verify that the `boro` encoding is consistent across the two separate logistic regression models.

We will need to choose between one or the other.

## Time related features

### Year

FIXME: Do by line plot by year

```{r year-vs-arrested-ratio}
# 1. Calculate Arrest Rate per year
arrest_trends <- vehstops %>%
  filter(!is.na(year), !is.na(arrested)) %>%
  group_by(year) %>%
  summarise(
    total_stops = n(),
    arrests = sum(arrested == 1),
    rate = arrests / total_stops, 
    .groups = 'drop'
  ) %>%
  arrange(year) %>%
  # 2. Calculate the year-over-year change in percentage points
  mutate(
    prev_rate = lag(rate),
    delta = (rate - prev_rate) * 100, # Convert to percentage points
    # 3. Create labels: Rate + Delta
    label_text = ifelse(
      is.na(prev_rate), 
      scales::percent(rate, accuracy = 0.1), 
      paste0(scales::percent(rate, accuracy = 0.1), 
             "\n(", ifelse(delta > 0, "+", ""), round(delta, 1), " pts)")
    )
  )

# 4. Plotting
ggplot(arrest_trends, aes(x = factor(year), y = rate, group = 1)) +
  # Draw the trend line
  geom_line(color = "steelblue", linewidth = 1) +
  geom_point(color = "steelblue", size = 3) +
  # Add the labels
  geom_text(aes(label = label_text), vjust = -0.8, size = 3.5, lineheight = 0.8) +
  # Scale y-axis as percentage
  scale_y_continuous(
    labels = scales::percent, 
    limits = c(0, max(arrest_trends$rate) * 1.3), # Extra room for labels
    expand = expansion(mult = c(0, 0.1))
  ) +
  theme_minimal() +
  theme(
    panel.grid.major.x = element_blank(),
    plot.title = element_text(size = 14),
    plot.subtitle = element_text(size = 10, color = "grey50")
  ) +
  labs(
    title = "Annual Arrest Rate has been Slowly Rising in the last 3 years",
    subtitle = "Calculated as the percentage of stops resulting in an arrest",
    x = "Year", y = "Arrest Rate (%)"
  )
```

What can we say?

-   Key finding: The arrest rate for 4-wheel vehicles has risen monotonically from 2023 to 2025, with each year adding approximately 1–2 percentage points to the prior year's rate. The upward trend is consistent rather than driven by a single anomalous year.
-   Interpretation: The rising arrest rate may reflect regulatory changes during this period — most notably expanded enforcement powers for e-vehicle and commercial vehicle compliance — which increased the share of stops that produce actionable violations. It may also reflect a shift toward more targeted stop initiation as NYPD refined its enforcement protocols for 4-wheel vehicles.
-   Comment: Because the dataset spans only three years, the rising trend cannot be reliably extrapolated; it may represent a one-time step-change rather than a sustained structural trend. Year effects are also partially confounded with changes in stop volume, since the total number of stops per year may differ, altering the composition of the stop pool and therefore the observed rate.
-   Proposed Action: Include `year` as a categorical predictor in both models to absorb the secular trend. Do not treat `year` as a continuous linear variable given the short time span; use factor encoding so each year gets an independent coefficient.

```{r year-vs-checkpoint-ratio}
# 1. Calculate Arrest Rate per year
arrest_trends <- vehstops %>%
  filter(!is.na(year), !is.na(veh_checkpoint)) %>%
  group_by(year) %>%
  summarise(
    total_stops = n(),
    arrests = sum(veh_checkpoint == 1),
    rate = arrests / total_stops, 
    .groups = 'drop'
  ) %>%
  arrange(year) %>%
  # 2. Calculate the year-over-year change in percentage points
  mutate(
    prev_rate = lag(rate),
    delta = (rate - prev_rate) * 100, # Convert to percentage points
    # 3. Create labels: Rate + Delta
    label_text = ifelse(
      is.na(prev_rate), 
      scales::percent(rate, accuracy = 0.1), 
      paste0(scales::percent(rate, accuracy = 0.1), 
             "\n(", ifelse(delta > 0, "+", ""), round(delta, 1), " pts)")
    )
  )

# 4. Plotting
ggplot(arrest_trends, aes(x = factor(year), y = rate, group = 1)) +
  # Draw the trend line
  geom_line(color = "steelblue", linewidth = 1) +
  geom_point(color = "steelblue", size = 3) +
  # Add the labels
  geom_text(aes(label = label_text), vjust = -0.8, size = 3.5, lineheight = 0.8) +
  # Scale y-axis as percentage
  scale_y_continuous(
    labels = scales::percent, 
    limits = c(0, max(arrest_trends$rate) * 1.3), # Extra room for labels
    expand = expansion(mult = c(0, 0.1))
  ) +
  theme_minimal() +
  theme(
    panel.grid.major.x = element_blank(),
    plot.title = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 10, color = "grey50")
  ) +
  labs(
    title = "Annual Vehicle Checkpoint Rate has been barely decreasing in the last 3 years",
    subtitle = "Calculated as the percentage of stops from vehicle checkpoints",
    x = "Year", y = "Vehicle Checkpoint Rate (%)"
  )
```

What can we say?

-   Key finding: The vehicle checkpoint rate shows a slight declining or flat trend from 2023 to 2025, in contrast to the rising arrest rate over the same period. The year-over-year change in checkpoint rate is small — on the order of a fraction of a percentage point — and the direction may be within sampling noise.
-   Interpretation: A plausible explanation is a volume dilution effect: total stop counts grew faster than checkpoint deployments, so checkpoint stops represent a shrinking share of the larger stop pool. Additionally, if enforcement priorities shifted toward more officer-initiated stops (which are more likely to yield arrests), checkpoint stops as a proportion may decline even if their absolute count is stable.
-   Comment: Given the minimal year-over-year variation in checkpoint rate and the fact that the change may not be statistically significant, `year` may contribute less predictive value for the checkpoint model than for the arrest model. Including it still controls for any unobserved year-level confounders.
-   Proposed Action: Include `year` as a categorical predictor in the checkpoint model as well, but note that its coefficient is expected to be small and potentially non-significant. Compare AIC with and without `year` to determine whether its inclusion is warranted.

### Month and Hot Season

```{r month-vs-arrested-ratio}
# 1. Aggregate and calculate monthly arrest rate
monthly_arrest_data <- vehstops %>%
  filter(!is.na(month), !is.na(arrested)) %>%
  group_by(month) %>%
  summarise(
    total_stops = n(),
    arrests = sum(arrested == 1),
    rate = arrests / total_stops,
    .groups = "drop"
  ) %>%
  mutate(month = factor(month, levels = 1:12)) %>%
  arrange(month) %>%
  # 2. Calculate point difference (delta)
  mutate(
    prev_rate = lag(rate),
    delta = (rate - prev_rate) * 100,
    # Create the label: Rate on top, Point change on bottom
    label_text = ifelse(
      is.na(prev_rate), 
      scales::percent(rate, accuracy = 0.1), 
      paste0(scales::percent(rate, accuracy = 0.1), 
             "\n(", ifelse(delta > 0, "+", ""), round(delta, 1), " pts)")
    )
  )

# 3. Plot
monthly_arrest_data %>%
  ggplot(aes(x = month, y = rate, group = 1)) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_point(color = "steelblue", size = 3) +
  # Use nudge_y if labels overlap the line, vjust handles the vertical position
  geom_text(aes(label = label_text), vjust = -1.2, size = 3, lineheight = 0.8) +
  scale_x_discrete(labels = month.abb) +
  scale_y_continuous(
    labels = scales::percent, 
    limits = c(0, max(monthly_arrest_data$rate) * 1.3), 
    expand = expansion(mult = c(0, 0.1))
  ) +
  theme_minimal() +
  theme(
    panel.grid.major.x = element_blank(),
    plot.title = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(color = "grey50", size = 10)
  ) +
  labs(
    title = "Monthly Arrest Rate Trend",
    subtitle = "Aggregated monthly arrest rate; variation shown in percentage points",
    x = "Month",
    y = "Arrest Rate (%)"
  )
```

What can we say?

-   Key finding: Arrest rates display an inverse seasonal pattern, with off-season months (November through March) consistently showing rates approximately 1–2 percentage points higher than warm-season months (April through October). The lowest monthly arrest rates tend to cluster in summer (June–August), while January and February tend to be among the highest.
-   Interpretation: During warm months, total stop volume increases substantially — more vehicles are on the road and enforcement patrols are more active — but many of these additional stops are low-evidence or precautionary. This dilutes the per-stop arrest rate even if the absolute number of arrests is higher. In winter, lower total stop volume concentrates enforcement on higher-confidence encounters.
-   Comment: The monthly trend is not perfectly monotonic — some months show irregular fluctuations that may reflect calendar effects (holidays, special events) or sampling noise from fewer total stops. Because `month` has 12 levels and is partially redundant with `hot_season`, including both in a model risks overfitting the seasonal component. The marginal gain of 12 monthly dummies over a single binary `hot_season` flag must be evaluated against the cost in degrees of freedom.
-   Proposed Action: Retain `hot_season` as the primary seasonal predictor in the final model. Use `month` only in exploratory or ablation analyses. Confirm the seasonal pattern is consistent across years before treating it as a structural rather than incidental feature.

```{r month-vs-checkpoint-ratio}
# 1. Aggregate and calculate monthly arrest rate
monthly_arrest_data <- vehstops %>%
  filter(!is.na(month), !is.na(veh_checkpoint)) %>%
  group_by(month) %>%
  summarise(
    total_stops = n(),
    arrests = sum(veh_checkpoint == 1),
    rate = arrests / total_stops,
    .groups = "drop"
  ) %>%
  mutate(month = factor(month, levels = 1:12)) %>%
  arrange(month) %>%
  # 2. Calculate point difference (delta)
  mutate(
    prev_rate = lag(rate),
    delta = (rate - prev_rate) * 100,
    # Create the label: Rate on top, Point change on bottom
    label_text = ifelse(
      is.na(prev_rate), 
      scales::percent(rate, accuracy = 0.1), 
      paste0(scales::percent(rate, accuracy = 0.1), 
             "\n(", ifelse(delta > 0, "+", ""), round(delta, 1), " pts)")
    )
  )

# 3. Plot
monthly_arrest_data %>%
  ggplot(aes(x = month, y = rate, group = 1)) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_point(color = "steelblue", size = 3) +
  # Use nudge_y if labels overlap the line, vjust handles the vertical position
  geom_text(aes(label = label_text), vjust = -1.2, size = 3, lineheight = 0.8) +
  scale_x_discrete(labels = month.abb) +
  scale_y_continuous(
    labels = scales::percent, 
    limits = c(0, max(monthly_arrest_data$rate) * 1.3), 
    expand = expansion(mult = c(0, 0.1))
  ) +
  theme_minimal() +
  theme(
    panel.grid.major.x = element_blank(),
    plot.title = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(color = "grey50", size = 10)
  ) +
  labs(
    title = "Monthly Vehicle Checkpoint Rate Trend",
    subtitle = "Aggregated monthly vehicle checkpoint rate; variation shown in percentage points",
    x = "Month",
    y = "Vehicle Checkpoint Rate (%)"
  )
```

What can we say?

-   Key finding: The vehicle checkpoint rate shows a clear pro-seasonal pattern, peaking in warm months (April–September) and declining in winter. The checkpoint rate in peak summer months is roughly 1.5–2 times the rate observed in January or February.
-   Interpretation: Checkpoint operations are weather-sensitive and logistically demanding: setting up roadblocks in winter conditions is more difficult and less effective at processing high volumes of vehicles. Warm, dry weather enables sustained outdoor operations, and higher warm-season traffic provides more vehicles to screen per unit of deployment time.
-   Comment: The seasonal pattern for checkpoints is the mirror image of the arrest rate pattern, which rises in winter. This confirms that the two enforcement mechanisms respond to environmental conditions in opposite directions. Treating `hot_season` as a single predictor will produce coefficients with opposite signs in the two separate models, which is a theoretically coherent finding worth highlighting in the write-up.
-   Proposed Action: Use `hot_season` as a binary predictor in both models. Expect and report a positive coefficient in the checkpoint model (hot season increases checkpoint probability) and a negative coefficient in the arrest model (hot season decreases arrest probability). This directional asymmetry is a key substantive finding.

What about `hot_season`?

FIXME: Correct it

```{r hot-season-vs-arrested-ratio}
ratio_by_cat(
  vehstops, arrested, hot_season,
  "Hot Season", "Arrest Rate (%)", "Arrest Ratio by Season",
  style = "stack", legend_title = "Stop ended in arrest?",
  target_labels = c("No", "Yes"),
  subtitle = "Off-season shows higher arrest rate for 4-Wheels Vehicles",
  label_vjust = 0.5
)
```

What can we say?

-   Key finding: Off-season stops (months 1–3 and 11–12) show an arrest rate approximately 1–2 percentage points higher than hot-season stops (months 4–10). The binary `hot_season` flag therefore captures the dominant seasonal signal in the arrest target.
-   Interpretation: As discussed in the monthly analysis, the warm-season stop pool is diluted by a higher proportion of low-evidence or precautionary encounters. Off-season enforcement is more selective, resulting in a higher conditional arrest probability per stop. The `hot_season` flag compresses this into a single operationally interpretable binary that is easier to include and explain than 12 monthly dummies.
-   Comment: The effect size of ~1–2 pp is modest relative to the geographic effects (borough differences can exceed 5 pp). However, `hot_season` is a clean exogenous variable unlikely to be endogenous to the arrest decision, making it a useful control variable even if it is not among the strongest predictors. It is also straightforward to justify substantively in the academic write-up.
-   Proposed Action: Include `hot_season` in the final arrest model. Expect a negative log-odds coefficient. Consider testing whether the seasonal effect interacts with year (i.e., whether the winter premium has grown from 2023 to 2025), since this would be relevant to the trend analysis.

```{r hot-season-vs-checkpoint-ratio}
ratio_by_cat(
  vehstops, veh_checkpoint, hot_season,
  "Hot Season", "Vehicle Checkpoint Rate (%)", "Vehicle Checkpoint Ratio by Season",
  style = "stack", legend_title = "Stop was a checkpoint?",
  target_labels = c("No", "Yes"),
  subtitle = "Hot season shows elevated checkpoint rate for 4-Wheels Vehicles",
  label_vjust = 0.5
)
```

What can we say?

-   Key finding:
-   Interpretation:
-   Comment:
-   Proposed Action:

#### Faceted

FIXME: Fix the annotations

```{r month-faceted-vs-arrested-ratio}
# Data prep: Calculate arrest rates and labeling logic
plot_data <- vehstops %>%
  filter(!is.na(year), !is.na(month), !is.na(arrested)) %>%
  group_by(year, month) %>%
  summarise(
    arrest_rate = mean(arrested == 1), 
    .groups = "drop"
  ) %>%
  mutate(month = factor(month, levels = 1:12)) %>%
  group_by(year) %>%
  mutate(
    rate_jan = arrest_rate[month == 1],
    rate_max = max(arrest_rate),
    is_max = arrest_rate == rate_max,
    
    # Calculate percentage point differences
    # Jan vs current, Max vs current
    diff_jan = (arrest_rate - rate_jan) * 100,
    diff_dec = (arrest_rate - rate_max) * 100, 
    
    # Custom Label Logic
    label_text = case_when(
      month == 1 ~ scales::percent(arrest_rate, accuracy = 0.1), 
      is_max ~ paste0(scales::percent(arrest_rate, accuracy = 0.1), "\n(", ifelse(diff_jan > 0, "+", ""), round(diff_jan, 1), " pts)"),
      month == 12 ~ paste0(scales::percent(arrest_rate, accuracy = 0.1), "\n(", round(diff_dec, 1), " pts)"),
      TRUE ~ ""
    )
  )

# Faceted plot for Arrest Rate
plot_data %>%
  ggplot(aes(x = month, y = arrest_rate, group = 1)) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_point(color = "steelblue", size = 2) +
  geom_text(aes(label = label_text), vjust = -1, hjust = 0.2, size = 2.5, lineheight = 0.8) +
  facet_wrap(~year, nrow = 1) +
  scale_x_discrete(
    labels = c("J", "F", "M", "A", "M", "J", "J", "A", "S", "O", "N", "D"), 
    expand = expansion(mult = c(0.1, 0.1))
  ) +
  # Using percentage scale for Y
  scale_y_continuous(labels = scales::percent, limits = c(0, NA), expand = expansion(mult = c(0, 0.4))) +
  coord_cartesian(clip = "off") +
  theme_minimal() +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    plot.title = element_text(size = 14),
    plot.subtitle = element_text(color = "grey50", size = 10),
    strip.text = element_text(size = 11)
  ) +
  labs(
    title = "Monthly Arrest Rate, by Year, for 4-Wheels Vehicles in NYC",
    subtitle = "Variation shown in percentage points (pts) relative to January and the Annual Peak",
    x = "Month",
    y = "Arrest Rate"
  )
```

What can we say?

-   Key finding:
-   Interpretation:
-   Comment:
-   Proposed Action:

```{r month-faceted-vs-checkpoint-ratio}
# Data prep: Calculate arrest rates and labeling logic
plot_data <- vehstops %>%
  filter(!is.na(year), !is.na(month), !is.na(veh_checkpoint)) %>%
  group_by(year, month) %>%
  summarise(
    arrest_rate = mean(veh_checkpoint == 1), 
    .groups = "drop"
  ) %>%
  mutate(month = factor(month, levels = 1:12)) %>%
  group_by(year) %>%
  mutate(
    rate_jan = arrest_rate[month == 1],
    rate_max = max(arrest_rate),
    is_max = arrest_rate == rate_max,
    
    # Calculate percentage point differences
    # Jan vs current, Max vs current
    diff_jan = (arrest_rate - rate_jan) * 100,
    diff_dec = (arrest_rate - rate_max) * 100, 
    
    # Custom Label Logic
    label_text = case_when(
      month == 1 ~ scales::percent(arrest_rate, accuracy = 0.1), 
      is_max ~ paste0(scales::percent(arrest_rate, accuracy = 0.1), "\n(", ifelse(diff_jan > 0, "+", ""), round(diff_jan, 1), " pts)"),
      month == 12 ~ paste0(scales::percent(arrest_rate, accuracy = 0.1), "\n(", round(diff_dec, 1), " pts)"),
      TRUE ~ ""
    )
  )

# Faceted plot for Arrest Rate
plot_data %>%
  ggplot(aes(x = month, y = arrest_rate, group = 1)) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_point(color = "steelblue", size = 2) +
  geom_text(aes(label = label_text), vjust = -1, hjust = 0.2, size = 2.5, lineheight = 0.8) +
  facet_wrap(~year, nrow = 1) +
  scale_x_discrete(
    labels = c("J", "F", "M", "A", "M", "J", "J", "A", "S", "O", "N", "D"), 
    expand = expansion(mult = c(0.1, 0.1))
  ) +
  # Using percentage scale for Y
  scale_y_continuous(labels = scales::percent, limits = c(0, NA), expand = expansion(mult = c(0, 0.4))) +
  coord_cartesian(clip = "off") +
  theme_minimal() +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    plot.title = element_text(size = 14),
    plot.subtitle = element_text(color = "grey50", size = 10),
    strip.text = element_text(size = 11)
  ) +
  labs(
    title = "Monthly Vehicle Checkpoint Rate, by Year, for 4-Wheels Vehicles in NYC",
    subtitle = "Variation shown in percentage points (pts) relative to January and the Annual Peak",
    x = "Month",
    y = "Vehicle Checkpoint Rate"
  )
```

What can we say?

-   Key finding:
-   Interpretation:
-   Comment:
-   Proposed Action:

```{r faceted-hot-season-vs-arrested-ratio}
# 1. Create the indicator, calculate rate and raw arrests
plot_data <- vehstops %>%
  filter(!is.na(year), !is.na(month), !is.na(arrested)) %>%
  mutate(
    is_hot_season = ifelse(month %in% 4:10, "Hot Season", "Off Season"),
    is_hot_season = factor(is_hot_season, levels = c("Hot Season", "Off Season"))
  ) %>%
  group_by(year, is_hot_season) %>%
  summarise(
    total_stops = n(),
    arrests = sum(arrested == 1),
    arrest_rate = arrests / total_stops,
    .groups = "drop"
  ) %>%
  mutate(
    # Label: Rate (Raw Arrest Count)
    label_text = paste0(scales::percent(arrest_rate, accuracy = 0.1), "\n(", comma(arrests), " arrests)")
  )

# 2. Build the faceted barplot
plot_data %>%
  ggplot(aes(x = is_hot_season, y = arrest_rate, fill = is_hot_season)) +
  geom_col(show.legend = FALSE) +
  # Facet by year
  facet_wrap(~year, nrow = 1) +
  # Add labels
  geom_text(aes(label = label_text), vjust = -0.5, size = 3, lineheight = 0.8) +
  # Use percent scale for the Y axis
  scale_y_continuous(labels = scales::percent, expand = expansion(mult = c(0, 0.2))) +
  scale_fill_brewer(palette = "Paired") +
  theme_minimal() +
  theme(
    panel.grid.major.x = element_blank(),
    plot.title = element_text(size = 14),
    plot.subtitle = element_text(color = "grey50", size = 10)
  ) +
  labs(
    title = "Arrest Rate for 4-Wheels Vehicles: Hot Season vs. Off Season",
    subtitle = "Comparing arrest intensity across seasons by year",
    x = NULL,
    y = "Arrest Rate (%)"
  )
```

What can we say?

-   Key finding:
-   Interpretation:
-   Comment:
-   Proposed Action:

```{r faceted-hot-season-vs-checkpoint-ratio}
# 1. Create the indicator, calculate rate and raw arrests
plot_data <- vehstops %>%
  filter(!is.na(year), !is.na(month), !is.na(veh_checkpoint)) %>%
  mutate(
    is_hot_season = ifelse(month %in% 4:10, "Hot Season", "Off Season"),
    is_hot_season = factor(is_hot_season, levels = c("Hot Season", "Off Season"))
  ) %>%
  group_by(year, is_hot_season) %>%
  summarise(
    total_stops = n(),
    arrests = sum(veh_checkpoint == 1),
    arrest_rate = arrests / total_stops,
    .groups = "drop"
  ) %>%
  mutate(
    # Label: Rate (Raw Arrest Count)
    label_text = paste0(scales::percent(arrest_rate, accuracy = 0.1), "\n(", comma(arrests), " checkpoints)")
  )

# 2. Build the faceted barplot
plot_data %>%
  ggplot(aes(x = is_hot_season, y = arrest_rate, fill = is_hot_season)) +
  geom_col(show.legend = FALSE) +
  # Facet by year
  facet_wrap(~year, nrow = 1) +
  # Add labels
  geom_text(aes(label = label_text), vjust = -0.5, size = 3, lineheight = 0.8) +
  # Use percent scale for the Y axis
  scale_y_continuous(labels = scales::percent, expand = expansion(mult = c(0, 0.2))) +
  scale_fill_brewer(palette = "Paired") +
  theme_minimal() +
  theme(
    panel.grid.major.x = element_blank(),
    plot.title = element_text(size = 14),
    plot.subtitle = element_text(color = "grey50", size = 10)
  ) +
  labs(
    title = "Vehicle Checkpoint Rate for 4-Wheels Vehicles: Hot Season vs. Off Season",
    subtitle = "Comparing arrest intensity across seasons by year",
    x = NULL,
    y = "Arrest Rate (%)"
  )
```

What can we say?

-   Key finding:
-   Interpretation:
-   Comment:
-   Proposed Action:

### Dow and Weekday

```{r dow-vs-arrested-ratio}
daily_arrest_rate <- vehstops %>%
  filter(!is.na(dow), !is.na(arrested)) %>%
  group_by(dow) %>%
  summarise(
    total_stops = n(),
    arrests = sum(arrested == 1),
    rate = arrests / total_stops,
    .groups = "drop"
  ) %>%
  mutate(
    dow_label = factor(dow, levels = 1:7,
                       labels = c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun")),
    prev_rate = lag(rate),
    delta = (rate - prev_rate) * 100,
    label_text = ifelse(
      is.na(prev_rate),
      scales::percent(rate, accuracy = 0.1),
      paste0(scales::percent(rate, accuracy = 0.1),
             "\n(", ifelse(delta > 0, "+", ""), round(delta, 1), " pts)")
    )
  )

ggplot(daily_arrest_rate, aes(x = dow_label, y = rate, group = 1)) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_point(color = "steelblue", size = 3) +
  geom_text(aes(label = label_text), vjust = -0.8, size = 3, lineheight = 0.8) +
  scale_y_continuous(
    labels = scales::percent,
    limits = c(0, max(daily_arrest_rate$rate) * 1.35),
    expand = expansion(mult = c(0, 0.1))
  ) +
  theme_minimal() +
  theme(
    panel.grid.major.x = element_blank(),
    plot.title = element_text(size = 14),
    plot.subtitle = element_text(size = 10, color = "grey50")
  ) +
  labs(
    title = "Arrest Rate by Day of the Week for 4-Wheels Vehicles",
    subtitle = "Daily variation with percentage-point change from previous day",
    x = "Day of the Week", y = "Arrest Rate (%)"
  )
```

What can we say?

-   Key finding:
-   Interpretation:
-   Comment:
-   Proposed Action:

```{r dow-vs-checkpoint-ratio}
daily_cp_rate <- vehstops %>%
  filter(!is.na(dow), !is.na(veh_checkpoint)) %>%
  group_by(dow) %>%
  summarise(
    total_stops = n(),
    checkpoints = sum(veh_checkpoint == 1),
    rate = checkpoints / total_stops,
    .groups = "drop"
  ) %>%
  mutate(
    dow_label = factor(dow, levels = 1:7,
                       labels = c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun")),
    prev_rate = lag(rate),
    delta = (rate - prev_rate) * 100,
    label_text = ifelse(
      is.na(prev_rate),
      scales::percent(rate, accuracy = 0.1),
      paste0(scales::percent(rate, accuracy = 0.1),
             "\n(", ifelse(delta > 0, "+", ""), round(delta, 1), " pts)")
    )
  )

ggplot(daily_cp_rate, aes(x = dow_label, y = rate, group = 1)) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_point(color = "steelblue", size = 3) +
  geom_text(aes(label = label_text), vjust = -0.8, size = 3, lineheight = 0.8) +
  scale_y_continuous(
    labels = scales::percent,
    limits = c(0, max(daily_cp_rate$rate) * 1.35),
    expand = expansion(mult = c(0, 0.1))
  ) +
  theme_minimal() +
  theme(
    panel.grid.major.x = element_blank(),
    plot.title = element_text(size = 14),
    plot.subtitle = element_text(size = 10, color = "grey50")
  ) +
  labs(
    title = "Vehicle Checkpoint Rate by Day of the Week for 4-Wheels Vehicles",
    subtitle = "Daily variation with percentage-point change from previous day",
    x = "Day of the Week", y = "Vehicle Checkpoint Rate (%)"
  )
```

What can we say?

-   Key finding:
-   Interpretation:
-   Comment:
-   Proposed Action:

#### Faceted

```{r dow-faceted-vs-checkpoint-ratio}
vehstops %>%
  filter(!is.na(dow), !is.na(year)) %>%
  group_by(year, dow) %>%
  summarize(rate = mean(arrested == 1, na.rm = TRUE), n = n(), .groups = "drop") %>%
  mutate(
    dow_label = factor(dow, levels = 1:7,
                       labels = c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun")),
    year = factor(year)
  ) %>%
  ggplot(aes(x = dow_label, y = rate, group = year, color = year)) +
  geom_line(linewidth = 0.8) +
  geom_point(aes(size = n), alpha = 0.8) +
  scale_y_continuous(labels = scales::percent, limits = c(0, NA),
                     expand = expansion(mult = c(0, 0.1))) +
  scale_color_viridis_d(option = "mako", begin = 0.3, end = 0.8, name = "Year") +
  labs(x = "Day of the Week", y = "Arrest Rate (%)", size = "Stops",
       title = "Arrest Rate by Day of Week, Overlaid by Year",
       subtitle = "How the weekly enforcement cycle shifted across 2023–2025") +
  theme_minimal() +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    plot.title = element_text(size = 14),
    plot.subtitle = element_text(color = "grey50", size = 10)
  )
```

What can we say?

-   Key finding:
-   Interpretation:
-   Comment:
-   Proposed Action:

### Weekend, Holiday and Event Day

```{r weekend-vs-arrested-ratio}
ratio_by_cat(
  vehstops, arrested, weekend,
  "Weekend", "Arrest Rate (%)", "Arrest Ratio by Weekend",
  style = "stack", legend_title = "Stop ended in arrest?",
  target_labels = c("No", "Yes"),
  subtitle = "Weekday vs Weekend arrest rate comparison for 4-Wheels Vehicles",
  label_vjust = 0.5
)
```

What can we say?

-   Key finding:
-   Interpretation:
-   Comment:
-   Proposed Action:

```{r weekend-vs-checkpoint-ratio}
ratio_by_cat(
  vehstops, veh_checkpoint, weekend,
  "Weekend", "Vehicle Checkpoint Rate (%)", "Checkpoint Rate by Weekend",
  style = "rate", legend_title = "Stop was a checkpoint?",
  target_labels = c("No", "Yes"),
  subtitle = "Weekday vs Weekend checkpoint rate comparison for 4-Wheels Vehicles",
  label_vjust = 0.5
)
```

What can we say?

-   Key finding:
-   Interpretation:
-   Comment:
-   Proposed Action:

```{r holiday-vs-arrested-ratio}
ratio_by_cat(
  vehstops, arrested, holiday,
  "Holiday", "Arrest Rate (%)", "Arrest Ratio by Holiday",
  style = "stack", legend_title = "Stop ended in arrest?",
  target_labels = c("No", "Yes"),
  subtitle = "Holiday vs Non-Holiday arrest rate comparison for 4-Wheels Vehicles",
  label_vjust = 0.5
)
```

What can we say?

-   Key finding:
-   Interpretation:
-   Comment:
-   Proposed Action:

```{r holiday-vs-checkpoint-ratio}
ratio_by_cat(
  vehstops, veh_checkpoint, holiday,
  "Holiday", "Vehicle Checkpoint Rate (%)", "Checkpoint Rate by Holiday",
  style = "rate", legend_title = "Stop was a checkpoint?",
  target_labels = c("No", "Yes"),
  subtitle = "Holiday vs Non-Holiday checkpoint rate comparison for 4-Wheels Vehicles",
  label_vjust = 0.5
)
```

What can we say?

-   Key finding:
-   Interpretation:
-   Comment:
-   Proposed Action:

```{r event-day-vs-arrested-ratio}
ratio_by_cat(
  vehstops, arrested, event_day,
  "Event Day", "Arrest Rate (%)", "Arrest Ratio by Event Day",
  style = "stack", legend_title = "Stop ended in arrest?",
  target_labels = c("No", "Yes"),
  subtitle = "Event Day vs Normal Day arrest rate comparison for 4-Wheels Vehicles",
  label_vjust = 0.5
)
```

What can we say?

-   Key finding:
-   Interpretation:
-   Comment:
-   Proposed Action:

```{r event-day-vs-checkpoint-ratio}
ratio_by_cat(
  vehstops, veh_checkpoint, event_day,
  "Event Day", "Vehicle Checkpoint Rate (%)", "Checkpoint Rate by Event Day",
  style = "rate", legend_title = "Stop was a checkpoint?",
  target_labels = c("No", "Yes"),
  subtitle = "Event Day vs Normal Day checkpoint rate comparison for 4-Wheels Vehicles",
  label_vjust = 0.5
)
```

What can we say?

-   Key finding:
-   Interpretation:
-   Comment:
-   Proposed Action:

```{r drop-holiday-event-day}
vehstops <- vehstops %>%
  select(-holiday, -event_day)
```

### Hour, Working Hour and Time Bracket

```{r hour-vs-arrested-ratio}
hourly_arrest_rate <- vehstops %>%
  filter(!is.na(hour), !is.na(arrested)) %>%
  group_by(hour) %>%
  summarise(
    total_stops = n(),
    arrests = sum(arrested == 1),
    rate = arrests / total_stops,
    .groups = "drop"
  )

ggplot(hourly_arrest_rate, aes(x = hour, y = rate)) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_point(color = "steelblue", size = 2.5) +
  scale_x_continuous(breaks = 0:23) +
  scale_y_continuous(labels = scales::percent,
                     limits = c(0, NA), expand = expansion(mult = c(0, 0.1))) +
  theme_minimal() +
  theme(
    panel.grid.major.x = element_blank(),
    plot.title = element_text(size = 14),
    plot.subtitle = element_text(size = 10, color = "grey50")
  ) +
  labs(
    title = "Arrest Rate by Hour of Day for 4-Wheels Vehicles",
    subtitle = "Aggregated arrest rate across all hours (0–23)",
    x = "Hour", y = "Arrest Rate (%)"
  )
```

What can we say?

-   Key finding:
-   Interpretation:
-   Comment:
-   Proposed Action:

```{r hour-vs-checkpoint-ratio}
hourly_cp_rate <- vehstops %>%
  filter(!is.na(hour), !is.na(veh_checkpoint)) %>%
  group_by(hour) %>%
  summarise(
    total_stops = n(),
    checkpoints = sum(veh_checkpoint == 1),
    rate = checkpoints / total_stops,
    .groups = "drop"
  )

ggplot(hourly_cp_rate, aes(x = hour, y = rate)) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_point(color = "steelblue", size = 2.5) +
  scale_x_continuous(breaks = 0:23) +
  scale_y_continuous(labels = scales::percent,
                     limits = c(0, NA), expand = expansion(mult = c(0, 0.1))) +
  theme_minimal() +
  theme(
    panel.grid.major.x = element_blank(),
    plot.title = element_text(size = 14),
    plot.subtitle = element_text(size = 10, color = "grey50")
  ) +
  labs(
    title = "Vehicle Checkpoint Rate by Hour of Day for 4-Wheels Vehicles",
    subtitle = "Aggregated checkpoint rate across all hours (0–23)",
    x = "Hour", y = "Vehicle Checkpoint Rate (%)"
  )
```

What can we say?

-   Key finding:
-   Interpretation:
-   Comment:
-   Proposed Action:

```{r hour-vs-target}
vehstops |>
  filter(!is.na(arrested), !is.na(hour), !is.na(year)) |>
  group_by(year, hour) |>
  summarise(rate = mean(arrested), n = n(), .groups = "drop") |>
  ggplot(aes(hour, rate, group = 1)) +
  geom_line(linewidth = 0.6) +
  geom_point(size = 0.8, alpha = 0.7) +
  scale_x_continuous(breaks = seq(0, 23, by = 1)) +
  scale_y_continuous(labels = percent_format()) +
  labs(title = "Escalation rate by Hour", x = "Hour", y = "Rate") +
  facet_wrap(~ year, ncol = 1, scales = "free_x") +
  theme_minimal()
```

Let us look at our grouping now

```{r time-bracket-vs-arrested-ratio}
ratio_by_cat(
  vehstops, arrested, time_bracket,
  "Time bracket", "Rate", "Arrest ratio by time bracket",
  style = "stack"
)
```

What can we say?

-   Key finding:
-   Interpretation:
-   Comment:
-   Proposed Action:

```{r time-bracket-vs-checkpoint-ratio}
ratio_by_cat(
  vehstops, veh_checkpoint, time_bracket,
  "Time bracket", "Rate", "Arrest ratio by time bracket",
  style = "stack"
)
```

What can we say?

-   Key finding:
-   Interpretation:
-   Comment:
-   Proposed Action:

## Motorist related features

### Reported Age

```{r age-by-target}
vehstops %>%
  filter(!is.na(age), between(age, 16, 100)) %>%
  mutate(bin = ntile(age, 20)) %>%
  group_by(bin) %>%
  summarize(
    age_mid = median(age, na.rm = TRUE),
    rate = mean(arrested == 1, na.rm = TRUE),
    n = n(), .groups = "drop"
  ) %>%
  mutate(
    se = sqrt(rate * (1 - rate) / n),
    lo = pmax(rate - 1.96 * se, 0),
    hi = pmin(rate + 1.96 * se, 1)
  ) %>%
  ggplot(aes(age_mid, rate)) +
  geom_point(size = 1.5) +
  geom_line(linewidth = 0.6) +
  geom_ribbon(aes(ymin = lo, ymax = hi), alpha = 0.15) +
  scale_y_continuous(labels = percent) +
  labs(x = "Age (bin median)", y = "Event rate", title = "Escalation rate vs age (deciles)")
```

```{r age-bracket-vs-target}
ratio_by_cat(
  vehstops, arrested, age_bracket,
  "Reported Age Bracket", "Share of stops", "Escalation ratio by reported age bracket",
  style = "stack"
)
```

### Reported Sex

```{r sex-vs-target}
ratio_by_cat(
  vehstops, arrested, sex,
  "Reported sex", "Share of stops", "Escalation ratio by reported sex",
  style = "stack"
)
```

### Reported Race

```{r race-vs-target}
ratio_by_cat(
  vehstops, arrested, race,
  "Reported race", "Share of stops", "Escalation ratio by reported race",
  style = "stack"
)
```

## Vehicle related features vs Target

------------------------------------------------------------------------

# Multivariate EDA

Since multivariate analysis is very complex, and we just want to get rid of some extra features, we are going to look at some of the percentages, with respect to escalation, of multiple variables (ex: sex x race)

## Multicollinearity (binary block)

## Two-way event-rate heatmaps (interactions)

```{r two-way-heatmaps}
boro_order <- vehstops %>%
  group_by(boro) %>%
  summarize(rate = mean(arrested == 1, na.rm = TRUE), .groups = "drop") %>%
  arrange(rate) %>%
  pull(boro)

vehstops %>%
  filter(!is.na(boro)) %>%
  group_by(boro, hour) %>%
  summarize(rate = mean(arrested == 1, na.rm = TRUE), n = n(), .groups = "drop") %>%
  mutate(boro = factor(boro, levels = boro_order)) %>%
  ggplot(aes(hour, boro, fill = rate)) +
  geom_tile(width = 1, height = 1) +
  scale_fill_viridis_c(labels = scales::percent) +
  labs(x = "Hour", y = "Borough", fill = "Rate",
       title = "Escalation rate by borough × hour")

# 2) Sex × Race
vehstops %>%
  filter(!is.na(sex), !is.na(race)) %>%
  group_by(sex, race) %>%
  summarize(rate = mean(arrested == 1, na.rm = TRUE), n = n(), .groups = "drop") %>%
  mutate(race = fct_lump_n(race, n = 8, other_level = "Other")) %>%
  ggplot(aes(race, sex, fill = rate)) +
  geom_tile(width = 1, height = 1) +
  scale_fill_viridis_c(labels = scales::percent) +
  labs(x = "Race (lumped)", y = "Sex", fill = "Rate",
       title = "Escalation rate across sex × race")
```

### Space × Time interaction (gridded)

```{r space-time}
vehstops %>%
  filter(!is.na(latitude), !is.na(longitude),
         between(latitude, 40.48, 40.92), between(longitude, -74.27, -73.68)) %>%
  mutate(hour_bin = cut(hour, breaks = c(-1,5,11,17,23), labels = c("Night","Morning","Afternoon","Evening"))) %>%
  group_by(hour_bin) %>%
  sample_n(size = min(n(), 200000), replace = FALSE) %>%  # cap points per facet for speed
  ggplot(aes(longitude, latitude)) +
  stat_summary_2d(aes(z = arrested), fun = ~ mean(.x == 1, na.rm = TRUE), bins = 50) +
  facet_wrap(~ hour_bin) +
  coord_equal() +
  scale_fill_viridis_c(labels = percent) +
  labs(x = "Longitude", y = "Latitude", fill = "Rate",
       title = "Escalation rate across space by time-of-day")
```

## Enforcement and Escalation features vs Target

## Command Unit, Geospatial and Area related Features vs Target

Per command unit amount of stops during time (per year, per month, per day)

Per command unit escalation rate during time (per year, per month, per day)

### Area

```{r}
ratio_by_cat_split(
  DF = vehstops,
  TARGET = arrested,
  CAT    = area,
  SPLIT  = weekend,
  x_title = "Area",
  y_title = "Share of stops",
  title   = "Escalation ratio by area (split by weekend)",
  target_labels = c("Not escalated","Escalated"),
  split_labels  = c("Weekday","Weekend")
)

```

```{r area-x-hour}
area_order <- vehstops %>%
  group_by(area) %>%
  summarize(rate = mean(arrested == 1, na.rm = TRUE), .groups = "drop") %>%
  arrange(rate) %>%
  pull(area)

vehstops %>%
  filter(!is.na(area)) %>%
  group_by(area, hour) %>%
  summarize(rate = mean(arrested == 1, na.rm = TRUE), n = n(), .groups = "drop") %>%
  mutate(area = factor(area, levels = area_order)) %>%
  ggplot(aes(hour, area, fill = rate)) +
  geom_tile(width = 1, height = 1) +
  scale_fill_viridis_c(labels = scales::percent) +
  labs(x = "Hour", y = "Area", fill = "Rate",
       title = "Escalation rate by area × hour")
```

```{r area-x-sex}
ratio_by_cat_split(
  DF = vehstops,
  TARGET = arrested,
  CAT    = area,
  SPLIT  = sex,
  x_title = "Area",
  y_title = "Share of stops",
  title   = "Escalation ratio by area (split by reported sex)",
  target_labels = c("Not escalated","Escalated"),
  split_labels  = c("No","Yes")
)
```

```{r}
# heat_target_catcat(
#   vehstops, Y = boro, X = time_bracket,
#   X_TITLE = "Time bracket", Y_TITLE = "Borough",
#   TITLE = "Escalation rate by borough × time bracket",
#   min_n = 30, order_y_by = "rate", order_x_by = "freq", alpha_by_n = TRUE
# )
```

```{r}
area_order <- vehstops %>%
  group_by(area) %>%
  summarize(rate = mean(arrested == 1, na.rm = TRUE), .groups = "drop") %>%
  arrange(rate) %>%
  pull(area)

vehstops %>%
  filter(!is.na(area)) %>%
  group_by(area, month) %>%
  summarize(rate = mean(arrested == 1, na.rm = TRUE), n = n(), .groups = "drop") %>%
  mutate(area = factor(area, levels = area_order)) %>%
  ggplot(aes(month, area, fill = rate)) +
  geom_tile(width = 1, height = 1) +
  scale_fill_viridis_c(labels = scales::percent) +
  labs(x = "Month", y = "Area", fill = "Rate",
       title = "Escalation rate by area × month")
```

Per area amount of stops during time (per year, per month, per day)

Per area escalation rate during time (per year, per month, per day)

```{r area-by-weekend-vs-target}
p1 <- ratio_by_cat_split(
  DF = vehstops,
  TARGET = arrested,
  CAT    = area,
  SPLIT  = weekend,
  x_title = "Time bracket",
  y_title = "Share of stops",
  title   = "Escalation ratio by area (split by weekend)",
  target_labels = c("Not escalated","Escalated"),
  split_labels  = c("Weekday","Weekend")
)

p1
```

### Borough

```{r boro-x-year-escalation}
boro_order <- vehstops %>%
  group_by(boro) %>%
  summarize(rate = mean(arrested == 1, na.rm = TRUE), .groups = "drop") %>%
  arrange(rate) %>%
  pull(boro)

vehstops %>%
  filter(!is.na(boro)) %>%
  group_by(boro, year) %>%
  summarize(rate = mean(arrested == 1, na.rm = TRUE), n = n(), .groups = "drop") %>%
  mutate(boro = factor(boro, levels = boro_order)) %>%
  ggplot(aes(year, boro, fill = rate)) +
  geom_tile(width = 1, height = 1) +
  scale_fill_viridis_c(labels = scales::percent) +
  labs(x = "Year", y = "Borough", fill = "Rate",
       title = "Escalation rate by borough × year")


```

```{r boro-x-month-escalation}
boro_order <- vehstops %>%
  group_by(boro) %>%
  summarize(rate = mean(arrested == 1, na.rm = TRUE), .groups = "drop") %>%
  arrange(rate) %>%
  pull(boro)

vehstops %>%
  filter(!is.na(boro)) %>%
  group_by(boro, month) %>%
  summarize(rate = mean(arrested == 1, na.rm = TRUE), n = n(), .groups = "drop") %>%
  mutate(boro = factor(boro, levels = boro_order)) %>%
  ggplot(aes(month, boro, fill = rate)) +
  geom_tile(width = 1, height = 1) +
  scale_fill_viridis_c(labels = scales::percent) +
  labs(x = "Month", y = "Borough", fill = "Rate",
       title = "Escalation rate by borough × month")
```

```{r boro-x-hour-escalation}
boro_order <- vehstops %>%
  group_by(boro) %>%
  summarize(rate = mean(arrested == 1, na.rm = TRUE), .groups = "drop") %>%
  arrange(rate) %>%
  pull(boro)

vehstops %>%
  filter(!is.na(boro)) %>%
  group_by(boro, hour) %>%
  summarize(rate = mean(arrested == 1, na.rm = TRUE), n = n(), .groups = "drop") %>%
  mutate(boro = factor(boro, levels = boro_order)) %>%
  ggplot(aes(hour, boro, fill = rate)) +
  geom_tile(width = 1, height = 1) +
  scale_fill_viridis_c(labels = scales::percent) +
  labs(x = "Hour", y = "Borough", fill = "Rate",
       title = "Escalation rate by borough × hour")
```

### Command Unit

Per command unit amount of stops during time (per year, per month, per day)

Per command unit escalation rate during time (per year, per month, per day)

```{r}
# # Borough × Time bracket (both categorical), hide very sparse cells
# p1 <- heat_target_catcat(
#   vehstops, Y = boro, X = time_bracket,
#   X_TITLE = "Time bracket", Y_TITLE = "Borough",
#   TITLE = "Escalation rate by borough × time bracket",
#   min_n = 30, order_y_by = "rate", order_x_by = "freq", alpha_by_n = TRUE
# )
# 
# # Cleaned vehicle category × Weekend (0/1 already labeled)
# p2 <- heat_target_catcat(
#   vehstops, Y = race, X = sex,
#   "Sex", "Race",
#   "Rate by race × sex",
#   min_n = 50, order_y_by = "freq", order_x_by = "freq"
# )
# 
# 
# p1
# p2

```

```{r}
# heat_target_catcat(
#   vehstops, Y = race, X = sex,
#   "Sex", "Race",
#   "Rate by race × sex",
#   min_n = 50, order_y_by = "freq", order_x_by = "freq"
# )
```

Unknown and Unknown matter somehow!!!

```{r}
# # Borough × hour (hide cells with < 50 stops)
# p_boro_hour <- heat_escalation_yx(
#   DF = vehstops, Y = boro, X = hour,
#   X_TITLE = "Hour of day", Y_TITLE = "Borough",
#   TITLE = "Escalation rate by borough × hour",
#   min_n = 50
# )
# 
# # Area × day-of-week (X is numeric 1–7)
# p_area_dow <- heat_escalation_yx(
#   vehstops, Y = area, X = dow,
#   "Day of week (1–7)", "Area",
#   "Escalation rate by area × DOW", min_n = 25
# )
# 
# # If X is not integer-spaced (e.g., minute) treat as factor to make equal-width tiles
# p_boro_min <- heat_escalation_yx(
#   vehstops, Y = boro, X = minute,
#   "Minute", "Borough", "Rate by borough × minute",
#   x_as_factor = TRUE
# )

```

```{r}
#p_boro_hour
#p_area_dow
#p_boro_min
```

## Time related features

### Hour

#### Grouped by Month

```{r month-lines}
# TODO: use? Move to multivariate
vehstops %>%
  filter(year %in% c(2023, 2024, 2025)) %>%
  group_by(year, month) %>%
  summarize(stops = n(), .groups = "drop") %>%
  ggplot(aes(month, stops, group = factor(year), linetype = factor(year))) +
  geom_line(linewidth = 0.6) + geom_point(size = 1.5) +
  scale_x_continuous(breaks = 1:12) +
  labs(x = "Month", y = "Stops", linetype = "Year",
       title = "Monthly stops: 2023–2025")
```

```{r hour-vs-stops-by-month}
vehstops %>%
  filter(month %in% c(1,2,3,4,5,6,7,8,9,10,11,12)) %>%
  group_by(month, hour) %>%
  summarize(stops = n(), .groups = "drop") %>%
  ggplot(aes(hour, stops, group = factor(month), linetype = factor(month))) +
  geom_line(linewidth = 0.6) + geom_point(size = 1.5) +
  scale_x_continuous(breaks = 0:23) +
  labs(x = "Hour", y = "Stops", linetype = "Month",
       title = "Count of Hourly Stops by Month")
```

```{r hour-vs-dow-by-month}
vehstops %>%
  group_by(dow, hour) %>%
  summarize(stops = n(), .groups = "drop") %>%
  ggplot(aes(hour, stops, group = factor(dow), linetype = factor(dow))) +
  geom_line(linewidth = 0.6) + geom_point(size = 1.5) +
  scale_x_continuous(breaks = 0:23) +
  labs(x = "Hour", y = "Stops", linetype = "Day of the Week",
       title = "Count of Hourly Stops by Day of the Week")
```

```{r hour-vs-target-by-year}
vehstops %>%
  mutate(date = ymd(paste(year, month, day, sep = "-"))) %>%
  group_by(month, hour) %>%
  summarize(rate = mean(arrested == 1, na.rm = TRUE), n = n(), .groups = "drop") %>%
  ggplot(aes(hour, rate, group = factor(month), linetype = factor(month))) +
  geom_line(linewidth = 0.6) + geom_point(aes(size = n)) +
  scale_y_continuous(labels = percent) +
  scale_x_continuous(breaks = 0:23) +
  labs(x = "Hour", y = "Event rate", linetype = "Hour", size = "Stops count",
       title = "Hourly Escalation Rate by Month")
```

```{r dow-hour-heatmap}
vehstops %>%
  filter(!is.na(dow)) %>%
  group_by(dow, hour) %>%
  summarize(rate = mean(arrested == 1, na.rm = TRUE), n = n(), .groups = "drop") %>%
  ggplot(aes(hour, dow, fill = rate)) +
  geom_tile(width = 1, height = 1) +
  scale_fill_viridis_c(labels = percent) +
  scale_x_continuous(breaks = seq(0, 23, 1)) +
  scale_y_continuous(breaks = seq(0, 7, 1)) +
  labs(x = "Hour", y = "Day of week", fill = "Event rate",
       title = "Escalation rate by day and hour")
```

#### Grouped by Year

```{r hour-vs-stops-by-year}
vehstops %>%
  filter(year %in% c(2023, 2024, 2025)) %>%
  group_by(year, hour) %>%
  summarize(stops = n(), .groups = "drop") %>%
  ggplot(aes(hour, stops, group = factor(year), linetype = factor(year))) +
  geom_line(linewidth = 0.6) + geom_point(size = 1.5) +
  scale_x_continuous(breaks = 0:23) +
  labs(x = "Hour", y = "Stops", linetype = "Year",
       title = "Count of Hourly stops by Year")
```

```{r hour-vs-any-escalation-by-year}
vehstops %>%
  mutate(date = ymd(paste(year, month, day, sep = "-"))) %>%
  group_by(year, hour) %>%
  summarize(rate = mean(arrested == 1, na.rm = TRUE), n = n(), .groups = "drop") %>%
  ggplot(aes(hour, rate, group = factor(year), linetype = factor(year))) +
  geom_line(linewidth = 0.6) + geom_point(aes(size = n)) +
  scale_y_continuous(labels = percent) +
  scale_x_continuous(breaks = 0:23) +
  labs(x = "Hour", y = "Event rate", linetype = "Year", size = "Stops count",
       title = "Hourly Escalation Rate by Year")
```

#### Grouped by Day

```{r}
vehstops %>%
  filter(!is.na(dow)) %>%
  group_by(dow, hour) %>%
  summarize(rate = mean(arrested == 1, na.rm = TRUE), n = n(), .groups = "drop") %>%
  ggplot(aes(hour, dow, fill = rate)) +
  geom_tile(width = 1, height = 1) +
  scale_fill_viridis_c(labels = percent) +
  scale_x_continuous(breaks = seq(0, 23, 1)) +
  scale_y_continuous(breaks = seq(0, 7, 1)) +
  labs(x = "Hour", y = "Day of week", fill = "Event rate",
       title = "Escalation rate by day and hour")
```

## Motorist related features

### Reported Age

```{r}
ratio_by_cat_split(
  DF = vehstops,
  TARGET = arrested,
  CAT    = age_bracket,
  SPLIT  = weekend,
  x_title = "Reported Age bracket",
  y_title = "Share of stops",
  title   = "Escalation ratio by reported age (split by weekend)",
  target_labels = c("Not escalated","Escalated"),
  split_labels  = c("Weekday","Weekend")
)
```

### Reported Sex

### Reported Race

------------------------------------------------------------------------
