---
title: "Predicting Excessive Enforcement Patterns in Police Behavior After Vehicles Stops in New York"
subtitle: "EDA and Feature Engineering - 4 Wheels Vehicles"
author: "Gianmaria Pizzo"
output:
  html_document:
    fig_caption: yes
    theme: flatly
    highlight: pygments
    code_folding: show
    kee_md: yes
    toc: yes
    toc_depth: 4
    number_sections: no
    toc_float:
      smooth_scroll: no
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

These `.Rmd` files represent the project submission for the course [Statistical Inference and Learning](https://www.unive.it/data/course/521944) by Professor [Cristiano Varin](https://www.unive.it/data/people/5592728) at [Ca' Foscari University of Venice](https://www.unive.it/) for the year 2025/2026.

# Problem Statement - TODO

## Goal of this rmarkdown file - TODO

We explore the features in 3 different ways.

1.  First, we separately analyze them and try to grasp their meaning, their distribution and the possible outliers or other problematic aspects. This univariate analysis, although complicated by the presence of many categorical predictors, should give us an initial idea of how the data is distributed.
2.  Then, we compare the possible predictors with respect to the target feature, and try to understand if there is any notable relationship between the two of them. Also, we try to see whether there is evidence of correlation, collinearity, or other problematic behaviors they may introduce.
3.  Finally we explore the subsets of features that may be related (ex: geospatial features and command unit, or time of the day and some demographics), to find out whether some of them can be removed or replaced. Plus we try to investigate about multicollinearity and other complications.

The EDA is the first step to understanding the data and its nature.

This includes analyzing the single features (Univariate Analysis), their interactions with the target (Bivariate Analysis) and their interactions with other features (Multivariate Analysis and Collinearity Check). However, the latter might be hard to compute as it requires the estimation of a regression model that follows certain assumptions and it is usually easier to check this when using linear models.

TODO: Say we will skip some visualizations

During my exploration I will look out for possible outliers, correlation and any other statistically relevant problem that might affect my final prediction.

We will start with the environment.

Note that, from now on, I will shift adopt a more **formal or inclusive tone** (e.g., the "editorial we").

FIXME: Comments should be done addressing 2 wheels vehicles. FIXME: Reduce colors, add annotation and title that is not just plain description FIXME: Univariate plots need both count and percentage on top

------------------------------------------------------------------------

# Environment

The first to do, is to build a reproducible environment for our project.

```{r}
requirements=c( 
  "remotes", "summarytools", "forcats", "scales", "purrr",
  "stringr", "plyr", "dbplyr", "dplyr", "dtplyr", "tidyr", "tidyselect",
  "timeDate", "timechange", "tzdb", "hms", "readr", "tigris",
  "lubridate", "prettyunits", "RColorBrewer", "viridis", "rcartocolor", "ggbeeswarm", "emmeans",
  "sf", "plotly", "ggplot2", "ggmap", "ggdensity", "UpSetR", "patchwork"
)
```

```{r install-requirements, include=FALSE}
for (req in requirements){
  if (!require(req, character.only = TRUE)){
      install.packages(req)
  }
}
```

Note that some of the packages may not be used, but to avoid any error related to dependencies, we are going to keep all of these libraries. Now, the loading of all packages should be immediate.

```{r load-requirements, include=FALSE}
for (req in requirements){
  if (!require(req, character.only = TRUE)){
      library(req)
  }
}
```

```{r remove-vars, include=FALSE}
rm(requirements, req)
```

## Plot Functions

Here we add the plot functions to make some kinds of visualizations reproducible.

```{r univariate-plot-helpers}
# ==============================================================================
# 1. Plots for Univariate Analysis
# ==============================================================================
#' @title Plot Binary Bar Chart
#' 
#' @description Creates a simple bar chart counting the occurrences of a binary feature, 
#' with labels showing both the count and the percentage.
#' 
#' @param DF The input dataframe.
#' @param FEATURE The binary column to plot (unquoted).
#' @param OPTION_0 The text label for the "0" or "FALSE" category.
#' @param OPTION_1 The text label for the "1" or "TRUE" category.
#' @param X_LABEL Label for the X axis.
#' @param Y_LABEL Label for the Y axis.
#' @param TITLE Main title of the plot.
#' @param pad Numeric padding for the top of the Y axis to fit text labels.
#' @param text_vjust Vertical adjustment for the text labels on top of bars.
#' 
#' @return A ggplot2 object.
plot_binary_bar <- function(DF, FEATURE, OPTION_0, OPTION_1, X_LABEL, Y_LABEL, TITLE, pad = 0.15, text_vjust = -0.3) {
  
  # Standardize the binary feature and calculate percentages
  df_sum <- DF %>%
    mutate(
      .raw  = {{ FEATURE }},
      .feat = dplyr::case_when(
        as.character(.raw) %in% c("1","TRUE","true")  ~ 1L,
        as.character(.raw) %in% c("0","FALSE","false") ~ 0L,
        TRUE ~ NA_integer_
      )
    ) %>%
    filter(!is.na(.feat)) %>%
    count(.feat, name = "n") %>%
    mutate(
      .lab = factor(.feat, levels = c(0, 1), labels = c(OPTION_0, OPTION_1)),
      # NEW: Calculate percentage and create combined label
      pct = n / sum(n),
      label_text = paste0(scales::comma(n), "\n(", scales::percent(pct, accuracy = 0.1), ")")
    )
  
  # Get the maximum bar height
  ymax <- max(df_sum$n, na.rm = TRUE)
  
  # Build the ggplot
  ggplot(df_sum, aes(x = .lab, y = n, fill = .lab)) +
    geom_col(width = 0.8, show.legend = FALSE) +
    # Updated to use label_text
    geom_text(aes(label = label_text), vjust = text_vjust, lineheight = 0.8) +
    # Expand the Y axis slightly to accommodate the taller multi-line labels
    scale_y_continuous(limits = c(0, ymax * (1 + pad)),
                       expand = expansion(mult = c(0, 0.02))) +
    coord_cartesian(clip = "off") +
    theme_minimal() +
    theme(
      panel.grid.major.x = element_blank(), 
      plot.margin = margin(t = 20, r = 5.5, b = 5.5, l = 5.5)
    ) +
    scale_fill_viridis_d(option = "mako", begin = 0.3, end = 0.8) +
    labs(x = X_LABEL, y = Y_LABEL, title = TITLE)
}


#' @title Plot Univariate Bar Chart (Categorical)
#' 
#' @description Creates a bar chart counting the occurrences of a categorical feature with 2+ classes.
#' Allows dynamic ordering (by frequency or original/alphabetical order).
#' 
#' @param DF The input dataframe.
#' @param FEATURE The categorical column to analyze (unquoted).
#' @param X_LABEL Label for the X axis.
#' @param Y_LABEL Label for the Y axis (default is "Count").
#' @param TITLE Main title of the plot.
#' @param top_n Optional integer; if set, keeps only the top N most frequent categories.
#' @param order_by Method to order the X axis: "freq" (descending count) or "none" (alphabetical/original).
#' @param pad Numeric padding for the top of the Y axis to fit text labels.
#' @param text_vjust Vertical adjustment for the text labels on top of bars.
#' 
#' @return A ggplot2 object.
plot_univariate_cat <- function(DF, FEATURE, X_LABEL, Y_LABEL = "Count", TITLE, top_n = NULL, order_by = c("freq", "none"), pad = 0.12, text_vjust = -0.35) {
  
  order_by <- match.arg(order_by)
  
  # Compute counts
  df_sum <- DF %>%
    filter(!is.na({{ FEATURE }})) %>%
    count({{ FEATURE }}, name = "n")
  
  # Filter top n results
  if (!is.null(top_n)) {
    df_sum <- df_sum %>% slice_max(order_by = n, n = top_n)
  }
  
  # X axis order (either frequency or std)
  cat_levels <- switch(
    order_by,
    freq = df_sum %>% arrange(desc(n)) %>% pull({{ FEATURE }}),
    none = sort(unique(df_sum %>% pull({{ FEATURE }})))
  )
  
  # Apply sorting
  df_sum <- df_sum %>%
    mutate(.feat_f = factor(as.character({{ FEATURE }}), levels = as.character(cat_levels)))
  
  # Max height for dynamic padding
  ymax <- max(df_sum$n, na.rm = TRUE)

  # Construct plot
  ggplot(df_sum, aes(x = .feat_f, y = n, fill = .feat_f)) +
    geom_col(width = 0.8, show.legend = FALSE) +
    geom_text(aes(label = scales::comma(n)), vjust = text_vjust, size = 3) +
    scale_y_continuous(limits = c(0, ymax * (1 + pad)),
                       expand = expansion(mult = c(0, 0.02))) +
    coord_cartesian(clip = "off") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
          panel.grid.major.x = element_blank(),
          plot.margin = margin(t = 10, r = 5.5, b = 5.5, l = 5.5)) +
    scale_fill_viridis_d(option = "mako", begin = 0.3, end = 0.8) +
    labs(x = X_LABEL, y = Y_LABEL, title = TITLE)
}
```

```{r biviariate-plot-helpers}
# ==============================================================================
# 2. Plots for Bivariate Analysis
# ==============================================================================

#' @title Calculate Cramer's V
#' 
#' @description Calculates the Cramer's V measure of association between two 
#' categorical vectors. Returns a value between 0 (no association) and 1 (perfect association).
#' 
#' @param x First categorical vector.
#' @param y Second categorical vector.
#' 
#' @return A numeric value representing Cramer's V.
cramer_v <- function(x, y) {
  # Contingency Table
  tbl <- table(x, y)
  
  # Execute Chi2 Test
  chi2 <- suppressWarnings(chisq.test(tbl, correct = FALSE)$statistic)
  
  # Compute dim and min value
  n <- sum(tbl)
  min_dim <- min(nrow(tbl) - 1, ncol(tbl) - 1)
  
  # Avoid math error for 1 categ
  if(min_dim == 0) return(0) 
  
  # Cramer's formula
  v <- sqrt(chi2 / (n * min_dim))
  return(as.numeric(v))
}

# ==============================================================================
# Categorical "Correlation" Heatmap - Association Matrix based on Cramer's V
# ==============================================================================

#' @title Plot Categorical Correlation Matrix (Cramer's V)
#' 
#' @description Creates a correlation-like heatmap for multiple categorical variables
#' using Cramer's V.
#' 
#' @param DF The input dataframe.
#' @param FEATURES A character vector containing the names of the categorical columns to compare.
#' @param TITLE Main title of the plot.
#' 
#' @return A ggplot2 object.
plot_categorical_correlation <- function(DF, FEATURES, TITLE = "Association Matrix for Categorical Predictors (Cramer's V)") {
  
  # Select columns and drop NAs
  df_sub <- DF %>% 
    select(all_of(FEATURES)) %>%
    drop_na()
  
  n_vars <- ncol(df_sub)
  
  # Initialize matrix and set main diag to 1s
  mat <- matrix(1, nrow = n_vars, ncol = n_vars)
  rownames(mat) <- colnames(df_sub)
  colnames(mat) <- colnames(df_sub)
  
  # FIXME: Just do upper/lower part of matrix
  # Compute Cramer's V for each possible pair
  for(i in 1:(n_vars - 1)) {
    for(j in (i + 1):n_vars) {
      v <- cramer_v(df_sub[[i]], df_sub[[j]])
      # Symmetric 
      mat[i, j] <- v
      mat[j, i] <- v
    }
  }
  
  # Make matrix compatible with ggplot2
  mat_long <- as.data.frame(as.table(mat))
  colnames(mat_long) <- c("Var1", "Var2", "CramerV")
  
  # Create heatmap
  ggplot(mat_long, aes(x = Var1, y = Var2, fill = CramerV)) +
    geom_tile(color = "white", width = 1, height = 1) +
    # Rounding
    geom_text(aes(label = round(CramerV, 2)), size = 4) +
    # Setting colors
    scale_fill_gradient(low = "whitesmoke", high = "steelblue", limits = c(0, 1), name = "Cramer's V") +
    labs(title = TITLE, x = "", y = "") +
    theme_minimal() +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1, size = 10, face = "bold"),
      axis.text.y = element_text(size = 10, face = "bold"),
      panel.grid = element_blank()
    )
}

# ==============================================================================
# Plotting of Targets' Rate Against Categorical Predictors
# ==============================================================================

# FIXME: Percentage shown in the wrong place.

#' @title Plot Target Ratio by Category
#' 
#' @description Creates a bivariate bar chart showing the relationship between a binary 
#' TARGET and a single categorical variable (CAT). It supports two visual styles: 
#' a 100% stacked bar chart ("stack") or a simple bar chart showing the rate of the target ("rate").
#' 
#' @param DF The input dataframe.
#' @param TARGET The binary column to analyze (unquoted). It will be standardized to 0/1.
#' @param CAT The categorical column for the X axis (unquoted).
#' @param x_title Label for the X axis.
#' @param y_title Label for the Y axis.
#' @param title Main title of the plot.
#' @param target_labels A character vector of length 2 defining the labels for 0 and 1 (e.g., c("Not escalated", "Escalated")).
#' @param style The visual style of the plot: "stack" (100% stacked bars) or "rate" (single bar for target = 1).
#' @param order_by Method to order the X axis categories: by "rate" (ascending target rate), "freq" (ascending frequency), or "none" (alphabetical/original).
#' @param percent_accuracy Accuracy for the percent labels (e.g., 0.1 for one decimal place).
#' @param colors A character vector of length 2 defining the hex color codes for the 0 and 1 categories respectively.
#' 
#' @return A ggplot2 object.
ratio_by_cat <- function(DF, TARGET, CAT, x_title, y_title, title, 
                         target_labels = c("Not escalated", "Escalated"), 
                         style = c("stack", "rate"), 
                         order_by = c("rate", "freq", "none"), 
                         percent_accuracy = 0.1, 
                         colors = c("grey80", "steelblue"),
                         legend_title = "Legend",
                         subtitle = NULL,
                         label_vjust = 0.5) {
  
  # Validate arguments
  style    <- match.arg(style)
  order_by <- match.arg(order_by)
  
  # Standardize target and extract the category
  dat <- DF %>%
    mutate(
      .t_raw = {{ TARGET }},
      .t = case_when(
        as.character(.t_raw) %in% c("1","TRUE","true") ~ 1L,
        as.character(.t_raw) %in% c("0","FALSE","false") ~ 0L,
        TRUE ~ NA_integer_
      ),
      .cat = {{ CAT }}
    ) %>%
    filter(!is.na(.t), !is.na(.cat))
  
  # Compute rates and frequencies
  rates <- dat %>% 
    group_by(.cat) %>%
    summarise(total = n(), rate = mean(.t == 1), .groups = "drop")
  
  # Determine ordering
  cat_levels <- switch(
    order_by,
    rate = rates %>% arrange(rate) %>% pull(.cat),
    freq = rates %>% arrange(total) %>% pull(.cat),
    none = unique(dat$.cat)
  )
  
  # Build the plot based on style
  if (style == "stack") {
    stack <- dat %>%
      count(.cat, .t, name = "n") %>%
      group_by(.cat) %>%
      mutate(ratio = n/sum(n)) %>%
      ungroup() %>%
      mutate(
        .cat_f = factor(as.character(.cat), levels = as.character(cat_levels)),
        .t_f = factor(.t, levels = c(0,1), labels = target_labels)
      )
    
    ggplot(stack, aes(x = .cat_f, y = ratio, fill = .t_f)) +
      geom_col(position = "fill") +
      # CHANGED: Showing ONLY the percentage
      geom_text(
        aes(label = scales::percent(ratio, accuracy = percent_accuracy)),
        position = position_fill(vjust = label_vjust), 
        size = 3
      ) +
      scale_y_continuous(labels = scales::percent) +
      scale_fill_manual(values = setNames(colors, target_labels)) +
      labs(x = x_title, y = y_title, fill = legend_title, title = title, subtitle = subtitle) +
      theme_minimal() +
      theme(
        axis.text.x = element_text(angle = 45, hjust = 1),
        panel.grid = element_blank(),
        plot.subtitle = element_text(color = "grey50")
      )
      
  } else {
    # Rate style
    rates_data <- rates %>%
      mutate(.cat_f = factor(as.character(.cat), levels = as.character(cat_levels)))
    
    ggplot(rates_data, aes(x = .cat_f, y = rate)) +
      geom_col(width = 0.8, fill = colors[2]) +
      geom_text(
        aes(label = scales::percent(rate, accuracy = percent_accuracy)),
        vjust = -0.35, size = 3
      ) +
      scale_y_continuous(labels = scales::percent, expand = expansion(mult = c(0, 0.12))) +
      coord_cartesian(clip = "off") +
      labs(x = x_title, y = y_title, title = title, subtitle = subtitle) +
      theme_minimal() +
      theme(
        axis.text.x = element_text(angle = 45, hjust = 1),
        panel.grid = element_blank(),
        plot.subtitle = element_text(color = "grey50")
      )
  }
}
```

```{r multivariate-plot-helpers}
# ==============================================================================
# 3. Plots for Multivariate Analysis
# ==============================================================================

#' @title Heatmap of Target Rate (Categorical Y vs Numeric X)
#' 
#' @description Plots a heatmap showing the rate of a binary TARGET across a categorical
#' Y variable and a numeric (or factor) X variable.
#' 
#' @param DF The input dataframe.
#' @param TARGET The binary column to calculate the rate for (unquoted).
#' @param Y Categorical variable for the Y axis.
#' @param X Numeric or categorical variable for the X axis.
#' @param X_TITLE Label for the X axis.
#' @param Y_TITLE Label for the Y axis.
#' @param TITLE Main title of the plot.
#' @param min_n Minimum number of observations required to show a tile.
#' @param drop_na Boolean; if TRUE, rows with NA in X or Y are removed.
#' @param x_as_factor Boolean; if TRUE, forces the X variable to be treated as a categorical factor.
#' 
#' @return A ggplot2 object.
heat_target_yx <- function(DF, TARGET, Y, X, X_TITLE, Y_TITLE, TITLE, min_n = 0, drop_na = TRUE, x_as_factor = FALSE) {
  # Standardize the binary target
  dat <- DF %>%
    mutate(
      .target_raw = {{ TARGET }},
      .target_bin = dplyr::case_when(
        as.character(.target_raw) %in% c("1", "TRUE", "true")  ~ 1L,
        as.character(.target_raw) %in% c("0", "FALSE", "false") ~ 0L,
        TRUE ~ NA_integer_
      )
    )

  # Filter NAs
  if (drop_na) {
    dat <- dat %>% filter(!is.na({{Y}}), !is.na({{X}}), !is.na(.target_bin))
  } else {
    dat <- dat %>% filter(!is.na(.target_bin))
  }

  # Incrementally sort Y axis based on the mean ratio of the target
  y_order <- dat %>%
    group_by({{Y}}) %>%
    summarise(rate = mean(.target_bin == 1, na.rm = TRUE), .groups = "drop") %>%
    arrange(rate) %>%
    pull({{Y}})

  # Compute the target rate for each combination of Y x X
  grid <- dat %>%
    group_by({{Y}}, {{X}}) %>%
    summarise(
      rate = mean(.target_bin == 1, na.rm = TRUE),
      n = n(), .groups = "drop"
    ) %>%
    mutate(
      .y = factor(as.character({{Y}}), levels = as.character(y_order)),
      .x = if (x_as_factor) factor({{X}}) else {{X}},
      # Hide cells that do not reach the minimum amount of observations
      rate_plot = ifelse(n >= min_n, rate, NA_real_)
    )

  # Build ggplot heatmap
  ggplot(grid, aes(.x, .y, fill = rate_plot)) +
    geom_tile(width = 1, height = 0.9) +
    scale_fill_viridis_c(labels = scales::percent, na.value = "grey90", name = "Rate") +
    labs(x = X_TITLE, y = Y_TITLE, title = TITLE) +
    theme_minimal() +
    theme(panel.grid = element_blank())
}


#' @title Heatmap of Target Rate (Categorical Y vs Categorical X)
#' 
#' @description Plots a heatmap showing the rate of a binary TARGET across two categorical
#' variables (Y and X). Allows dynamic ordering and shading based on sample size.
#' 
#' @param DF The input dataframe.
#' @param TARGET The binary column to calculate the rate for (unquoted).
#' @param Y Categorical variable for the Y axis.
#' @param X Categorical variable for the X axis.
#' @param X_TITLE Label for the X axis.
#' @param Y_TITLE Label for the Y axis.
#' @param TITLE Main title of the plot.
#' @param min_n Minimum number of observations required to show a tile (filters noise).
#' @param order_y_by Method to order the Y axis: by 'rate', 'freq' (frequency), or 'none'.
#' @param order_x_by Method to order the X axis: by 'rate', 'freq' (frequency), or 'none'.
#' @param alpha_by_n Boolean; if TRUE, shades tiles by count using alpha (transparency).
#' 
#' @return A ggplot2 object.
heat_target_catcat <- function(DF, TARGET, Y, X, X_TITLE, Y_TITLE, TITLE, min_n = 0, order_y_by = c("rate","freq","none"), order_x_by = c("rate","freq","none"), alpha_by_n = FALSE ) {
  order_y_by <- match.arg(order_y_by)
  order_x_by <- match.arg(order_x_by)

  # Standardize target
  dat <- DF %>%
    mutate(
      .target_raw = {{ TARGET }},
      .target_bin = dplyr::case_when(
        as.character(.target_raw) %in% c("1", "TRUE", "true")  ~ 1L,
        as.character(.target_raw) %in% c("0", "FALSE", "false") ~ 0L,
        TRUE ~ NA_integer_
      )
    ) %>%
    # Filter NAs
    filter(!is.na({{Y}}), !is.na({{X}}), !is.na(.target_bin))

  # Compute stats to determine the Y axis order
  y_stats <- dat %>%
    group_by({{Y}}) %>%
    summarise(
      rate = mean(.target_bin == 1, na.rm = TRUE),
      n = n(), .groups = "drop"
    )
  
  # Compute stats to determine the X axis order
  x_stats <- dat %>%
    group_by({{X}}) %>%
    summarise(
      rate = mean(.target_bin == 1, na.rm = TRUE), 
      n = n(), .groups = "drop"
    )

  # Apply chosen "order by" to Y
  y_levels <- switch(
    order_y_by,
    rate = y_stats %>% arrange(rate) %>% pull({{Y}}),
    freq = y_stats %>% arrange(n) %>% pull({{Y}}),
    none = unique(dat %>% pull({{Y}}))
  )
  
  # Apply chosen "order by" to X
  x_levels <- switch(
    order_x_by,
    rate = x_stats %>% arrange(rate) %>% pull({{X}}),
    freq = x_stats %>% arrange(n) %>% pull({{X}}),
    none = unique(dat %>% pull({{X}}))
  )

  # Compute the cell's target rate and counts for each combination of Y x X
  grid <- dat %>%
    group_by({{Y}}, {{X}}) %>%
    summarise(
      rate = mean(.target_bin == 1, na.rm = TRUE),
      n = n(), .groups = "drop"
    ) %>%
    mutate(
      .y = factor(as.character({{Y}}), levels = as.character(y_levels)),
      .x = factor(as.character({{X}}), levels = as.character(x_levels)),
      rate_plot = ifelse(n >= min_n, rate, NA_real_)
    )

  # Build the plot
  p <- ggplot(grid, aes(.x, .y)) +
    labs(x = X_TITLE, y = Y_TITLE, title = TITLE) +
    theme_minimal() +
    theme(panel.grid = element_blank()) +
    # Use scales for displaying rates
    scale_fill_viridis_c(labels = scales::percent, na.value = "grey90", name = "Rate")

  if (alpha_by_n) {
    p <- p +
      geom_tile(aes(fill = rate_plot, alpha = n), width = 0.95, height = 0.95) +
      scale_alpha_continuous(range = c(0.25, 1), name = "Stops")
  } else {
    p <- p +
      geom_tile(aes(fill = rate_plot), width = 0.95, height = 0.95)
  }

  p
}


#' @title Stacked Ratio Bars Split by a Third Variable
#' 
#' @description Creates a faceted 100% stacked bar chart. This is highly effective 
#' for visualizing how the rate of a binary TARGET changes across a primary category (CAT), 
#' while simultaneously splitting the data by a third variable (SPLIT) to create side-by-side plots.
#' 
#' @param DF The input dataframe.
#' @param TARGET The binary column to analyze (unquoted). It will be standardized to 0/1.
#' @param CAT The main categorical variable for the X axis (unquoted).
#' @param SPLIT The categorical variable used to split/facet the plots (unquoted).
#' @param x_title Label for the X axis.
#' @param y_title Label for the Y axis.
#' @param title Main title of the entire plot.
#' @param target_labels Character vector of length 2 defining labels for the binary outcomes (e.g., c("Not escalated", "Escalated")).
#' @param split_labels Optional character vector to provide custom titles for the split panels (facets).
#' @param palette A character string specifying an RColorBrewer palette name (e.g., "Paired", "Set1").
#' @param percent_accuracy Accuracy for the percentage labels (e.g., 0.1 for one decimal place).
#' 
#' @return A ggplot2 object.
ratio_by_cat_split <- function(DF, TARGET, CAT, SPLIT, x_title, y_title, title, target_labels = c("Not escalated", "Escalated"), split_labels = NULL, palette = "Paired", percent_accuracy = 0.1){
  # Standardize target
  dat <- DF %>%
    mutate(
      .target_raw = {{ TARGET }},
      # Conversion to 0/1
      .target_bin = case_when(
        as.character(.target_raw) %in% c("1", "TRUE", "true") ~ 1L,
        as.character(.target_raw) %in% c("0", "FALSE", "false") ~ 0L,
        TRUE ~ NA_integer_
      ),
      .cat = {{ CAT }},
      .split = {{ SPLIT }}
    ) %>%
    # Filter NAs
    filter(!is.na(.cat), !is.na(.split), !is.na(.target_bin))

  # Order the main categories based on the overall target rate (from lowest to highest)
  cat_order <- dat %>%
    group_by(.cat) %>%
    summarise(
      rate = mean(.target_bin == 1),
      .groups = "drop"
    ) %>%
    arrange(rate) %>% 
    pull(.cat)

  # Tidy labels by converting character to factors
  dat <- dat %>%
    mutate(
      # Apply the order to X axis
      .cat = factor(as.character(.cat), levels = as.character(cat_order)),
      # Apply custom labels to the target
      .target_f = factor(.target_bin, levels = c(0, 1), labels = target_labels),
      # Format the split variable
      .split_f = {
        lv <- unique(as.character(.split))
        # If custom split labels are provided and the length matches, use them
        if (!is.null(split_labels) && length(split_labels) == length(lv))
          factor(as.character(.split), levels = lv, labels = split_labels)
        else factor(as.character(.split), levels = lv)
      }
    )

  # Compute rates
  plot_df <- dat %>%
    count(.split_f, .cat, .target_f, name = "n") %>%
    group_by(.split_f, .cat) %>%
    mutate(ratio = n/sum(n)) %>%
    ungroup()

  # Build plot
  ggplot(plot_df, aes(x = .cat, y = ratio, fill = .target_f)) +
    # Stacks the bars to exactly 100%
    geom_col(position = "fill") +
    # Add ratio text labels
    geom_text(
      aes(label = percent(ratio, accuracy = percent_accuracy)),
      position = position_fill(vjust = 0.5), 
      size = 3
    ) +
    # Format Y axis as percentages and make continuous
    scale_y_continuous(labels = percent) +
    # Use a clean palette for the target categories
    scale_fill_brewer(palette = palette, name = rlang::as_name(rlang::enquo(TARGET))) +
    labs(x = x_title, y = y_title, title = title) +
    # Split the plot into separate panels based on the .split_f variable
    facet_wrap(~ .split_f, nrow = 1) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
          panel.grid = element_blank())
}
```

```{r}
#' @title Plot Temporal Trend of Target Rate
#' 
#' @description Creates a scatter plot with a LOESS smoothing curve to visualize 
#' how the rate of a target event changes over a continuous temporal variable (e.g., date, day of year).
#' The size of the points represents the volume of stops.
#' 
#' @param DF The input dataframe.
#' @param TARGET The binary column to analyze (unquoted).
#' @param TIME_VAR The continuous time variable for the X axis (e.g., date, doy).
#' @param x_title Label for the X axis.
#' @param y_title Label for the Y axis.
#' @param title Main title of the plot.
plot_temporal_trend <- function(DF, TARGET, TIME_VAR, x_title = "Tempo", y_title = "Tasso", title = "Trend Temporale") {
  
  # Step 1: Standardizzare il target a 0/1
  dat <- DF %>%
    mutate(
      .target_raw = {{ TARGET }},
      .target_bin = case_when(
        as.character(.target_raw) %in% c("1","TRUE","true", "Yes")  ~ 1L,
        as.character(.target_raw) %in% c("0","FALSE","false", "No") ~ 0L,
        TRUE ~ NA_integer_
      ),
      .time = {{ TIME_VAR }}
    ) %>%
    filter(!is.na(.target_bin), !is.na(.time))
  
  # Step 2: Calcolare il tasso e il volume per ogni unità di tempo
  plot_df <- dat %>%
    group_by(.time) %>%
    summarise(
      rate = mean(.target_bin == 1, na.rm = TRUE),
      n = n(),
      .groups = "drop"
    )
  
  # Step 3: Costruire il grafico
  ggplot(plot_df, aes(x = .time, y = rate)) +
    # Punti proporzionali al numero di stop per capire la robustezza del dato
    geom_point(aes(size = n), alpha = 0.5, color = "steelblue") +
    # Linea di tendenza morbida per evidenziare il trend generale ignorando il rumore
    geom_smooth(se = FALSE, method = "loess", span = 0.3, color = "firebrick", size = 1.2) +
    scale_y_continuous(labels = scales::percent) +
    labs(x = x_title, y = y_title, size = "N. di Stop", title = title) +
    theme_minimal() +
    theme(panel.grid.minor = element_blank())
}

#' @title Plot Cyclic Time Heatmap
#' 
#' @description Creates a heatmap specifically designed for cyclic time variables 
#' (like Hour of Day vs Day of Week) to spot temporal hotspots.
#' 
#' @param DF The input dataframe.
#' @param TARGET The binary column to analyze (unquoted).
#' @param HOUR_VAR The variable representing the hour (0-23).
#' @param WEEKDAY_VAR The variable representing the day of the week.
#' @param title Main title of the plot.
plot_cyclic_time_heatmap <- function(DF, TARGET, HOUR_VAR, WEEKDAY_VAR, title = "Hotspots Temporali") {
  
  # Step 1: Standardizzare il target a 0/1
  dat <- DF %>%
    mutate(
      .target_raw = {{ TARGET }},
      .target_bin = case_when(
        as.character(.target_raw) %in% c("1","TRUE","true", "Yes")  ~ 1L,
        as.character(.target_raw) %in% c("0","FALSE","false", "No") ~ 0L,
        TRUE ~ NA_integer_
      ),
      .hour = as.numeric({{ HOUR_VAR }}),
      .day = {{ WEEKDAY_VAR }}
    ) %>%
    filter(!is.na(.target_bin), !is.na(.hour), !is.na(.day))
  
  # Step 2: Raggruppare e calcolare il tasso
  plot_df <- dat %>%
    group_by(.day, .hour) %>%
    summarise(
      rate = mean(.target_bin == 1, na.rm = TRUE),
      n = n(),
      .groups = "drop"
    ) %>%
    # Mostriamo solo incroci con un minimo di stop per evitare rumore statistico
    filter(n >= 10) 
  
  # Step 3: Costruire la mappa di calore temporale
  ggplot(plot_df, aes(x = .hour, y = .day, fill = rate)) +
    geom_tile(color = "white") +
    scale_fill_viridis_c(option = "magma", labels = scales::percent, name = "Tasso") +
    # Impostiamo l'asse X per mostrare tutte le 24 ore
    scale_x_continuous(breaks = 0:23, expand = c(0,0)) +
    labs(x = "Ora del Giorno", y = "", title = title) +
    theme_minimal() +
    theme(
      panel.grid = element_blank(),
      axis.text.y = element_text(face = "bold")
    )
}

#' @title Plot Temporal Trend Split by a Group (Rate or Frequency)
#' 
#' @description Generates a line plot showing how a metric (rate or count) 
#' changes over a generic time unit (X_VAR), split by another variable (Z_VAR).
#' Adapts to any time unit and can display data as overlapping lines or faceted grids.
#' 
#' @param DF The input dataframe.
#' @param X_VAR The generic time variable for the X axis (e.g., month, hour).
#' @param Z_VAR The variable to group/split the plots by (e.g., year, weekend).
#' @param TARGET The binary target variable. Required only if metric = "rate".
#' @param metric "rate" (percentage of TARGET) or "freq" (absolute count of stops).
#' @param style "overlay" (lines superimposed on one plot) or "facet" (separated sub-plots).
#' @param x_title Label for the X axis.
#' @param y_title Label for the Y axis. Default adjusts based on metric.
#' @param title Main title of the plot.
#' 
#' @return A ggplot2 object.
plot_temporal_split <- function(
  DF, X_VAR, Z_VAR, TARGET = NULL, 
  metric = c("rate", "freq"), 
  style = c("overlay", "facet"),
  x_title = "Tempo", y_title = NULL, title = "Trend Temporale"
) {
  
  metric <- match.arg(metric)
  style  <- match.arg(style)
  
  # Step 1: Sicurezza per il Target
  target_quo <- rlang::enquo(TARGET)
  if (metric == "rate" && rlang::quo_is_null(target_quo)) {
    stop("Errore: Devi specificare la colonna TARGET se usi metric = 'rate'.")
  }
  
  # Step 2: Estrazione variabili principali
  dat <- DF %>%
    mutate(
      .x = {{ X_VAR }},
      .z = {{ Z_VAR }}
    ) %>%
    filter(!is.na(.x), !is.na(.z))
  
  # Step 3: Calcolo delle Metriche
  if (metric == "rate") {
    dat <- dat %>%
      mutate(
        .t_raw = !!target_quo,
        .t_bin = case_when(
          as.character(.t_raw) %in% c("1","TRUE","true","Yes")  ~ 1L,
          as.character(.t_raw) %in% c("0","FALSE","false","No") ~ 0L,
          TRUE ~ NA_integer_
        )
      ) %>%
      filter(!is.na(.t_bin))
    
    plot_df <- dat %>%
      group_by(.z, .x) %>%
      summarise(y_val = mean(.t_bin == 1), n = n(), .groups = "drop")
    
    if (is.null(y_title)) y_title <- "Tasso (%)"
    
  } else { # metric == "freq"
    plot_df <- dat %>%
      group_by(.z, .x) %>%
      summarise(y_val = n(), n = n(), .groups = "drop")
    
    if (is.null(y_title)) y_title <- "Numero di Stop"
  }
  
  # Assicuriamoci che l'asse X sia trattato correttamente nel disegno delle linee
  plot_df <- plot_df %>% mutate(.z_f = factor(.z))
  
  # Step 4: Costruzione del Grafico base
  p <- ggplot(plot_df, aes(x = .x, y = y_val))
  
  # Logica dello "Stile"
  if (style == "overlay") {
    # Tutte le linee nello stesso grafico
    p <- p +
      geom_line(aes(color = .z_f, group = .z_f), linewidth = 0.8) +
      geom_point(aes(color = .z_f, size = n), alpha = 0.7) +
      labs(color = rlang::as_name(rlang::enquo(Z_VAR)), size = "Stops Count") +
      scale_color_viridis_d(option = "turbo") # Colori ben distinguibili
      
  } else { 
    # Stile "Facet" (Grafici separati impilati)
    p <- p +
      geom_line(linewidth = 0.6, group = 1, color = "steelblue") +
      geom_point(aes(size = n), alpha = 0.7, color = "steelblue") +
      # free_x permette ad ogni grafico di avere il suo asse se necessario
      facet_wrap(~ .z_f, ncol = 1, scales = "free_x") +
      labs(size = "Stops Count")
  }
  
  # Temi e Asse Y
  p <- p +
    labs(x = x_title, y = y_title, title = title) +
    theme_minimal() +
    theme(
      panel.grid.minor = element_blank(),
      strip.text = element_text(face = "bold", size = 11) # Titoli dei facet in grassetto
    )
  
  # Formattazione dell'Asse Y in base alla metrica scelta
  if (metric == "rate") {
    p <- p + scale_y_continuous(labels = scales::percent)
  } else {
    p <- p + scale_y_continuous(labels = scales::comma)
  }
  
  return(p)
}
```

```{r}
#' @title Plot Overlapping Density
#' @description Compares the distribution of a numeric variable across different categories using density curves.
#' @param DF The input dataframe.
#' @param NUM_VAR The continuous numeric variable (unquoted, e.g., age).
#' @param CAT_VAR The categorical grouping variable (unquoted, e.g., arrested).
#' @param title Main title of the plot.
plot_density_compare <- function(DF, NUM_VAR, CAT_VAR, x_title = "Valore", title = "Confronto Densità") {
  
  dat <- DF %>%
    filter(!is.na({{ NUM_VAR }}), !is.na({{ CAT_VAR }})) %>%
    mutate(.cat = factor({{ CAT_VAR }}))
  
  ggplot(dat, aes(x = {{ NUM_VAR }}, fill = .cat)) +
    # alpha = 0.5 rende i colori trasparenti per vedere le sovrapposizioni
    geom_density(alpha = 0.5, color = "white") +
    scale_fill_brewer(palette = "Set1", name = "Gruppo") +
    labs(x = x_title, y = "Densità (Probabilità)", title = title) +
    theme_minimal() +
    theme(panel.grid.minor = element_blank())
}

#' @title Plot Violin and Boxplot Combination
#' @description Shows the distribution, median, and density of a numeric variable across multiple categories.
#' @param DF The input dataframe.
#' @param CAT_VAR The categorical variable for the X axis (unquoted).
#' @param NUM_VAR The numeric variable for the Y axis (unquoted).
#' @param title Main title of the plot.
plot_violin_boxplot <- function(DF, CAT_VAR, NUM_VAR, x_title = "Categoria", y_title = "Valore Numerico", title = "Distribuzione Numerica per Categoria") {
  
  dat <- DF %>% filter(!is.na({{ CAT_VAR }}), !is.na({{ NUM_VAR }}))
  
  ggplot(dat, aes(x = reorder(factor({{ CAT_VAR }}), {{ NUM_VAR }}, FUN = median, na.rm=TRUE), y = {{ NUM_VAR }}, fill = factor({{ CAT_VAR }}))) +
    # Il violino mostra la densità (più è largo, più dati ci sono)
    geom_violin(alpha = 0.6, trim = FALSE, color = NA) +
    # Il boxplot interno (molto stretto) mostra la mediana e i quartili
    geom_boxplot(width = 0.1, fill = "white", color = "black", alpha = 0.8, outlier.size = 1) +
    scale_fill_viridis_d(option = "turbo") +
    coord_flip() + # Capovolgiamo per leggere meglio le etichette lunghe
    labs(x = x_title, y = y_title, title = title) +
    theme_minimal() +
    theme(legend.position = "none", panel.grid.minor = element_blank())
}

#' @title Plot Spatial Map of Target
#' @description Plots coordinates (Longitude vs Latitude) to create a visual map, coloring points by a target variable.
#' @param DF The input dataframe.
#' @param LON Longitude column (unquoted).
#' @param LAT Latitude column (unquoted).
#' @param TARGET Categorical or binary variable for coloring (unquoted).
#' @param title Main title of the plot.
plot_spatial_map <- function(DF, LON, LAT, TARGET, title = "Mappa Spaziale degli Stop") {
  
  dat <- DF %>% filter(!is.na({{ LON }}), !is.na({{ LAT }}), !is.na({{ TARGET }}))
  
  # Ordiniamo i dati in modo che i target "positivi" (es. arresti) 
  # vengano disegnati per ultimi e non siano nascosti dalla massa di stop normali
  dat <- dat %>% arrange({{ TARGET }})
  
  ggplot(dat, aes(x = {{ LON }}, y = {{ LAT }}, color = factor({{ TARGET }}))) +
    # Usiamo punti molto piccoli e trasparenti per gestire l'overplotting (troppi punti sovrapposti)
    geom_point(size = 0.5, alpha = 0.4) +
    scale_color_brewer(palette = "Set1", name = "Esito") +
    # Manteniamo le proporzioni geografiche corrette (coord_quickmap)
    coord_quickmap() + 
    labs(x = "Longitudine", y = "Latitudine", title = title) +
    theme_minimal() +
    theme(
      axis.text = element_blank(), # Rimuoviamo i numeri degli assi per sembrare una vera mappa
      panel.grid = element_blank()
    )
}
```

------------------------------------------------------------------------

Let us load the data now, and remove the extra feature `veh_category` because we are **exclusively analyzing 2 wheels vehicles now**.

```{r load-data-drop-veh-categ}
vehstops = read.csv("data/vehiclestops_4_wheels.csv")

vehstops <- vehstops %>%
  select(-veh_categ)
```

So now we have the following dataset, that we present with the usual summary

```{r summary-data}
print(dfSummary(vehstops), method="render")
```

This preview is already interesting.

TODO: Comment

------------------------------------------------------------------------

# Univariate EDA

TODO: Introduction TODO: Correct the following

In this part we perform a deeper analysis of the target feature, and some of the most important features (according to the domain knowledge in notebook 0) against their distribution in the dataset.

Our goal here is to polish the data, by removing outliers and possible confounders that may introduce bias. Furthermore, we need to understand its distribution and normalizing it if necessary.

Possibly, we can introduce new variables that are transformations or combinations of other ones, and evaluate their contributions later with a bivariate analysis and/or model testing.

Hopefully, this process will reduce the variance.

TODO: Comment everything

## Target

```{r enforcement-dist}
p1 <- plot_binary_bar(
  vehstops, veh_checkpoint, 
  "No", "Yes",
  "", "Count", "Was it a vehicle checkpoint?"
)

p2 <- plot_binary_bar(
  vehstops, arrested, 
  "No", "Yes",
  "", "", "Was the motorist arrested?"
)

row1 <- p1 | p2

row1 +
plot_layout(heights = c(1, 0.080, 1), guides = "collect") + 
plot_annotation(title = "Distribution of stops by target") &
theme(legend.position = "bottom") 
```

```{r, include=FALSE}
rm(p1, p2, row1)
```

What can we say?

-   Key finding: The dataset, for 2 wheels-vehicles, exhibits extreme class imbalance. Vehicle checkpoints are highly sparse, representing only approximately 1.4% of the total stops (2,691 "Yes" out of 189,610 total records). The same goes for arrests which happen approximately 5.6% of the times (10,560 "Yes" out of 189,610 total records).
-   Interpretation: `veh_checkpoint` constitute a "rare event" within this specific dataset. The distribution indicates that the vast majority of stops are standard patrol or traffic-related events, while checkpoint operations are infrequent activities. Even though, arrests are more frequent than checkpoint stops by a factor of nearly 4:1, they remain an infrequent occurrence relative to the total volume of police interactions.
-   Comment: The high class disparity *renders global accuracy an insufficient metric*. A predictive model optimizing for accuracy could attain an score of \~98.6% by simply classifying every stop as "No," thereby completely failing to capture the latent features that characterize a checkpoint stop. The model is statistically biased towards the majority class by default. This is also suboptimal for our research, as capturing true arrest patterns is central to our investigation into enforcement bias.
-   Proposed Action: Although we could employ cost-sensitive learning models, this is outside the scope of the course. Given the limitations of our available model suite, we will employ stratified downsampling of the majority class ("No") to create a balanced training set, ensuring compatibility across all models (LDA, QDA, KNN, etc...). For the GLM-based models (Logistic, Lasso, Ridge), we will specifically utilize observation weighting within the likelihood function to account for the class disparity. Finally, we will move away from the default 0.5 classification threshold, optimizing the decision boundary based on the Precision-Recall curve to ensure our model maintains sensitivity to the minority class ("Yes").

## Command Unit, Geospatial and Area-related predictors

### Area

```{r count-stops-by-area}
plot_univariate_cat(
  DF = vehstops, 
  FEATURE = area, 
  X_LABEL = "Area", 
  TITLE = "Distribution of Stops by Area (Sorted by Count)",
  order_by = "freq"
)
```

What can we say?

-   Key finding: There is a high degree of variance in the volume of stops across areas. We observe three "high-volume" areas (Brooklyn, Queens, and Manhattan) with comparable stop counts (ranging between \~50k and 56k), one "medium-volume" area (Bronx, \~30k), and a distinct "low-volume" area (Staten Island, \~1.5k).
-   Interpretation: The distribution is highly non-uniform, or better, uneven. The lower frequency of stops in the Bronx and Staten Island suggests that geographic location could be a strong predictor of the probability of being stopped (in case the targets' ratio distributed by area follows a specific pattern).
-   Comment: From a modeling perspective, this introduces a risk of **overfitting to the high-volume areas**. Since some of the available models (especially KNN or QDA) rely on local neighborhood data or global likelihoods, they will naturally prioritize minimizing errors in the areas with the most observations. Furthermore, the extreme sparsity of the Staten Island group may lead to high variance in the parameter estimates for that specific categorical level, potentially skewing the coefficients for the area predictor.
-   Proposed Action: We still need to check its correlation with subject-related predictors. Also, we prefer to explore finer-grained versions of this geographical predictor, which may be able to yield better results when used in the models. Later on, we will use dummy coding to represent the areas, but we will interpret the coefficients with caution (which is our to-go tactic for most categorical variables). However, when using **Lasso or Ridge regression**, we will carefully tune the penalty parameter ($\lambda$) to ensure that the "Area" coefficients are not artificially inflated or suppressed due to the sample size disparity.

### Borough

```{r count-stops-by-boro}
plot_univariate_cat(
  DF = vehstops, 
  FEATURE = boro, 
  X_LABEL = "Borough", 
  TITLE = "Distribution of Stops by Borough (Sorted by Count)",
)
```

What can we say?

-   Key finding: The comparison reveals that while "Staten Island" is uniformly sparse, boroughs like "Queens" and "Manhattan" are internally heterogeneous. Aggregating sub-boroughs into a single `area` feature assumes that police behavior is homogeneous across those sub-sectors.
-   Interpretation: If the arrest rate in "North West Queens" (a high-volume area) is statistically distinct from "South Queens," the `area` variable acts as a source of **aggregation bias**. By averaging these rates, we lose the model’s ability to differentiate between high-risk and low-risk zones within the same borough, potentially obscuring patterns of excessive enforcement. On the other hand, in a regression or classification context, this means our parameter estimates for these low-frequency categories will likely suffer from high variance and wide confidence intervals, making them unreliable indicators of systematic enforcement bias.
-   Comment: Discarding granularity prematurely is a legitimate methodological risk. If a specific sub-borough has a disproportionately higher event rate than its parent borough, the model will fail to capture that specific hotspot. We should not treat granularity as "noise" until we have empirical evidence that the target ratios are homogeneous within the boroughs. In fact, if the target ratio (arrest or checkpoint probability) differs significantly between "North West Queens" and "South Queens," aggregating them into a single "Queens" category would indeed "wash out" a genuine, predictive signal, resulting in an under-fitted model.
-   Proposed Action: As previously stated, we wish to plot the arrest/checkpoint rate for each borough level, as well for the areas. Then, based on our observations, we could train two alternative versions of the same model using `area` and `boro`. In fact, we could employ means of model comparison (for instance, Stratified K-fold Cross Validation) and, in case the granular version yields lower CV error, we may be able to discard the hypothesis of introducing noise/variability.

### Command Unit

For what concerns `command`, we can say that \* Interpretation: This granularity is likely prohibitive for inference. \* Comment: Retaining this level of granularity creates a high risk of overfitting. \* Proposed Action: We will remove the feature, since there is no easy way to evaluate its role inside a parsimonious model.

```{r drop-command}
vehstops <- vehstops %>%
  select(-command)
```

### Latitude and Longitude

For what concerns these two predictors, we will use the to plot a spatial heat map of the stops with these values available.

```{r stop-density}
# Download NYC Borough boundaries once
nyc_map <- counties(state = "NY", cb = TRUE) %>%
  filter(NAME %in% c("Bronx", "Kings", "New York", "Queens", "Richmond"))

# 1. Define the grid resolution (match this with your desired density)
bins <- 100
lon_breaks <- seq(-74.26, -73.71, length.out = bins + 1)
lat_breaks <- seq(40.48, 40.92, length.out = bins + 1)

# 2. Generate the grid_sum object from your main data
grid_sum <- vehstops %>%
  filter(!is.na(longitude) & !is.na(latitude)) %>%
  mutate(
    xbin = cut(longitude, breaks = lon_breaks, include.lowest = TRUE, right = FALSE),
    ybin = cut(latitude, breaks = lat_breaks, include.lowest = TRUE, right = FALSE)
  ) %>%
  # Extract the center point of each bin for the tile plot
  mutate(
    x = as.numeric(gsub(".*\\[(.*),.*", "\\1", xbin)) + (lon_breaks[2] - lon_breaks[1])/2,
    y = as.numeric(gsub(".*\\[(.*),.*", "\\1", ybin)) + (lat_breaks[2] - lat_breaks[1])/2
  ) %>%
  group_by(x, y) %>%
  summarise(n = n(), .groups = "drop")

# Plot with base layer
ggplot() +
  # Add NYC base layer
  geom_sf(data = nyc_map, fill = "white", color = "grey80", size = 0.2) +
  # Add density data on top
  geom_tile(data = grid_sum, aes(x, y, fill = n), alpha = 0.8) +
  # Use a more high-contrast color scale
  scale_fill_viridis_c(trans = "log10", name = "Stops (log10)", option = "mako", direction = -1) +
  coord_sf(xlim = c(-74.26, -73.71), ylim = c(40.48, 40.92)) +
  theme_minimal() +
  theme(panel.grid = element_blank()) +
  labs(title = "Heatmap of Stops Density - 100 Bins", x = "Longitude", y = "Latitude")
```

TODO: See if this affects the map

```{r, include=FALSE}
rm(grid_sum, nyc_map, bins, lat_breaks, lon_breaks)
```

What can we say?

-   Key finding: The heatmap reveals a highly heterogeneous spatial distribution, characterized by distinct high-density corridors and hotspots.
-   Interpretation: This spatial clustering indicates that police enforcement is not randomly distributed; it is structurally tied to urban infrastructure, likely following major traffic arteries or high-activity transit corridors.
-   Comment: Again, high stop density does not necessarily imply aggressive policing; it may simply reflect high traffic exposure. Also, this plot just give us an idea, because we are skipping over missing data.
-   Proposed Action: To prevent data leakage we cannot just include then in the models. These two predictors will be later removed.

## Time-related predictors

### Year

TODO: Decrease accordingly the font of annotation, Add annotation in the right side with some custom text, use percentages to see the increase%

```{r count-stops-by-year}
vehstops %>%
  filter(!is.na(year)) %>%
  count(year) %>%
  mutate(year = as.numeric(as.character(year))) %>%
  arrange(year) %>%
  # 1. Calculate growth and create the combined label
  mutate(
    prev_n = lag(n), pct_change = (n - prev_n) / prev_n * 100,
    # Combine count and percentage with a newline character
    label_text = ifelse(
      is.na(pct_change), 
      comma(n), 
      paste0(comma(n), "\n(", ifelse(pct_change > 0, "+", ""), round(pct_change, 1), "%)")
    )) %>%
  ggplot(aes(x = factor(year), y = n, group = 1)) +
  # Draw the trend line
  geom_line(color = "steelblue", linewidth = 1) +
  # Add points for clarity
  geom_point(color = "steelblue", size = 3) +
  # 2. Add the combined label
  # vjust = -0.5 helps center the multi-line text just above the point
  geom_text(aes(label = label_text), vjust = -0.8, nudge_x = -0.1, size = 3.5, lineheight = 0.8) +
  # Increase top expansion (0.3) to accommodate the taller labels
  scale_y_continuous(labels = comma, limits = c(0, NA), expand = expansion(mult = c(0, 0.2))) +
  theme_minimal() +
  theme(
    panel.grid.major.x = element_blank(), 
    plot.title = element_text(size = 14),
    plot.subtitle = element_text(size = 10, color = "grey50")
    ) +
  labs(
    title = "Annual Volume of Stops for 2-Wheels Vehicles in NYC",
    subtitle = "In 3 Years the Total Volume Increased by 2,5 Times",
    x = "Year", y = "Count"
    )
```

What can we say?

-   Key finding: There has been a significant, multi-year surge in the volume of police stops involving 2-wheeled vehicles, with the number of encounters increasing from \~33k in 2023 to \~85k in 2025.
-   Interpretation: The growth trajectory is non-linear. The 113% jump between 2023 and 2024 represents a massive inflection point, which is suggesting either a shift in departmental priority or a new legislation targeting these vehicles. On the other hand, the 19.8% increase in 2025 indicates that while activity is still rising, the rate of acceleration is starting to stabilize.
-   Comment: The visualization clearly demonstrates that the "enforcement burden" or "public exposure" to these stops is intensifying. However, since the dataset analyzed was aggregated by 2-wheels vehicles, we could be seeing the result of policies regarding a specific subset. Plus, 3 years is not a sufficient time span for us to deliver some substantial proof.
-   Proposed Action: We are going to look for some level of contextualization, in order to see if something changed during late 2023 or early 2024.

In fact, we found the article ["New York City Council Votes to Pass New E-Bike Safety Rules in Continued Effort to Address the Dangers of Lithium-Ion Batteries"](https://council.nyc.gov/press/2024/02/28/2564/) published in February 28, 2024. From its content, we could justify the raise in volume as natural consequence to new laws being put in place.

### Month

TODO: Decrease accordingly the font of annotation, Add annotation in the right side with some custom text, use percentages to see the increase%

```{r count-stops-by-month}
# Aggregate and calculate monthly change
month_data <- vehstops %>%
  filter(!is.na(month)) %>%
  count(month) %>%
  mutate(month = factor(month, levels = 1:12)) %>%
  arrange(month) %>%
  mutate(
    prev_n = lag(n), pct_change = (n - prev_n) / prev_n * 100,
    # Create the label: Count on top, Percentage on bottom
    label_text = ifelse(
      is.na(pct_change), comma(n),
      paste0(comma(n), "\n(", ifelse(pct_change > 0, "+", ""), round(pct_change, 1), "%)"))
  )

# Plot
month_data %>%
  ggplot(aes(x = month, y = n, group = 1)) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_point(color = "steelblue", size = 3) +
  # Apply text tweaks: nudge_x moves it left, vjust raises it
  geom_text(aes(label = label_text), vjust = -1, nudge_x = -0.1, size = 3, lineheight = 0.8) +
  scale_x_discrete(labels = month.abb) +
  scale_y_continuous(labels = comma, limits = c(0, NA), expand = expansion(mult = c(0, 0.3))) +
  theme_minimal() +
  theme(
    panel.grid.major.x = element_blank(),
    plot.title = element_text(size = 14),
    plot.subtitle = element_text(color = "grey50", size = 10) # Styled as requested
  ) +
  labs(
    title = "Monthly Volume of Stops for 2-Wheels Vehicles in NYC",
    subtitle = "Over 3 Years, Most Vehichles are Stopped between June and October",
    x = "Month",
    y = "Count"
  )
```

TODO: See if this affects the map

```{r, include=FALSE}
rm(month_data)
```

What can we say?

-   Key finding: The data exhibits strong seasonality, with stop volumes peaking in a sustained plateau between June and October (roughly 18k–19k stops/month). Enforcement bottoms out in the winter, with the sharpest declines occurring in November (-17%) and December (-28%).
-   Interpretation: The stop volume is likely a direct proxy for "exposure." As temperatures rise, more 2-wheeled vehicles are on the road, increasing the opportunity for police to conduct stops. The "plateau" from June through October suggests that once the weather is consistently warm, ridership hits a saturation point, and enforcement activity maintains a high, stable operational baseline.
-   Comment: The sharp drops at the end of the year demonstrate that enforcement is reactive: it tracks the behavior of the riders. In other words, the decline in stops during winter is likely due to **reduced vehicle usage** rather than a change in police behavior.
-   Proposed Action: Given this clear seasonal pattern, we ensure our predictive models to include this time-related feature.

Let us see this with respect to the year

```{r year-monthly-distribution}
# Data prep with explicit "+" sign logic
plot_data <- vehstops %>%
  filter(!is.na(year) & !is.na(month)) %>%
  count(year, month) %>%
  mutate(month = factor(month, levels = 1:12)) %>%
  group_by(year) %>%
  mutate(
    n_jan = n[month == 1],
    n_max = max(n),
    is_max = n == n_max,
    # Pre-calculate percentages for cleaner label logic
    pct_max = round((n - n_jan) / n_jan * 100, 0),
    pct_dec = round((n - n_max) / n_max * 100, 0),
    
    # Custom Label Logic
    label_text = case_when(
      month == 1 ~ comma(n), 
      is_max ~ paste0(comma(n), "\n(", ifelse(pct_max > 0, "+", ""), pct_max, "%)"),
      month == 12 ~ paste0(comma(n), "\n(", pct_dec, "%)"), # Dec is usually negative, so '-' is automatic
      TRUE ~ ""
    )
  )

# Faceted plot
plot_data %>%
  ggplot(aes(x = month, y = n, group = 1)) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_point(color = "steelblue", size = 2) +
  geom_text(aes(label = label_text), vjust = -1, hjust = 0.2, size = 2.5, lineheight = 0.8) +
  facet_wrap(~year, nrow = 1) +
  scale_x_discrete(
    labels = c("J", "F", "M", "A", "M", "J", "J", "A", "S", "O", "N", "D"), 
    expand = expansion(mult = c(0.1, 0.1))
    ) +
  scale_y_continuous(labels = comma, limits = c(0, NA), expand = expansion(mult = c(0, 0.4))) +
  coord_cartesian(clip = "off") +
  theme_minimal() +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    plot.title = element_text(size = 14),
    plot.subtitle = element_text(color = "grey50", size = 10),
    strip.text = element_text(size = 11)
  ) +
  labs(
    title = "Monthly Volume of Stops, by Year, for 2-Wheels Vehicles in NYC",
    subtitle = "Custom Growth/Decline analysis: January vs Max, and Max vs December",
    x = "Month",
    y = "Count"
  )
```

```{r, include=FALSE}
rm(plot_data)
```

What can we say?

-   Key finding: There is a pronounced "seasonal amplification" effect. While stop volume is increasing annually, the seasonal peaks are becoming more extreme, and the year-end "cooling off" period is becoming significantly sharper (from a -22% drop in 2023 to a -54% drop in 2025).
-   Interpretation: The data indicates that enforcement is becoming increasingly "concentrated" in the summer/autumn months. The gap between your "active season" (Max) and "off-season" (Dec) is widening. This suggests that the NYPD's activity for 2-wheelers is highly reactive to seasonal changes—either directly due to weather (fewer riders in winter) or due to a policy strategy that focuses enforcement efforts heavily on warmer months.
-   Comment: By highlighting the Jan-to-Max growth and the Max-to-Dec decline, we have moved the analysis beyond simple "volume counts". It clearly shows that while the baseline (January) is rising slightly, the *intensity* of the mid-year surges is growing much faster.
-   Proposed Action: Creating a binary indicator for "Hot Season" (April–Oct), that can effectively tell our models: "There is a structural change in the world during these months." This is way more interpretable than just a list of months.

### Checking out `hot_season`

Let us proceed with this intuition

```{r add-hot-season}
vehstops <- vehstops %>%
  mutate(hot_season = ifelse(month %in% 4:10, 1, 0))
```

Now, for plotting

```{r year-hot-season-distribution}
# 1. Create the indicator, count, and calculate percentages
plot_data <- vehstops %>%
  filter(!is.na(year) & !is.na(month)) %>%
  mutate(
    is_hot_season = ifelse(month %in% 4:10, "Hot Season", "Off Season"),
    is_hot_season = factor(is_hot_season, levels = c("Hot Season", "Off Season"))
  ) %>%
  count(year, is_hot_season) %>%
  group_by(year) %>%
  mutate(
    year_total = sum(n),
    pct = n / year_total,
    # Combine count and percentage into a single label string
    label_text = paste0(comma(n), "\n(", round(pct * 100, 1), "%)")
  )

# 2. Build the faceted barplot
plot_data %>%
  ggplot(aes(x = is_hot_season, y = n, fill = is_hot_season)) +
  geom_col(show.legend = FALSE) +
  # Facet by year
  facet_wrap(~year, nrow = 1) +
  # Add labels: count on top, percentage below
  geom_text(aes(label = label_text), vjust = -0.5, size = 3.5, lineheight = 0.8) +
  # Add vertical breathing room for the text
  scale_y_continuous(labels = comma, expand = expansion(mult = c(0, 0.2))) +
  scale_fill_brewer(palette = "Paired") +
  theme_minimal() +
  theme(
    panel.grid.major.x = element_blank(),
    plot.title = element_text(size = 14),
    plot.subtitle = element_text(color = "grey50", size = 10)
  ) +
  labs(
    title = "Volume of Stops for 2-Wheels Vehicles: Hot Season vs. Off Season",
    subtitle = "While Overall Volume Increase, Percentages slightly Drift Towards Hot Season",
    x = NULL,
    y = "Count"
  )
```

```{r, include=FALSE}
rm(plot_data)
```

What can we say?

-   Key finding: Total enforcement volume is surging, but more importantly, the enforcement pattern is becoming **increasingly concentrated** in the "Hot Season." While the hot season accounted for \~64% of stops in 2023, it climbed to \~68% of stops in 2025.
-   Interpretation: The growth in stops is not uniform throughout the year. The rising percentage indicates that the "Hot Season" (April–Oct) is becoming a disproportionately larger driver of total annual stops year-over-year. This suggests that the NYPD's enforcement efforts for 2-wheelers are becoming more sensitive—or reactive—to the seasonal influx of riders.
-   Comment: This plot confirms that `hot_season` could be a "high-signal" feature.
-   Proposed Action: We could either use it in place of the month, or make it an interaction term with `year` (i.e., `is_hot_season * year`).

### Day, Dow and Weekday

We have decided to exclude 'Day of Month' (`day`) and 'Day of Year' (`doy`) from our feature set to mitigate 'calendar artifact' bias. These variables introduce artificial statistical noise rather than genuine operational signals.

As we can see below, the structural variation in month length creates false activity dips for end-of-month dates

```{r day-bias}
vehstops %>%
  filter(!is.na(day)) %>%
  count(day) %>%
  ggplot(aes(x = day, y = n)) +
  geom_line(color = "grey60", alpha = 0.5) +
  geom_smooth(method = "loess", color = "steelblue", se = FALSE) +
  theme_minimal() +
   theme(
    panel.grid.major.x = element_blank(),
    plot.title = element_text(size = 14),
    plot.subtitle = element_text(color = "grey50", size = 10)
  ) +
  labs(title = "Volume of Stops by Day of the Month (1–31) for 2-Wheels Vehicles",
       subtitle = "Note the drop at 31: this is a calendar artifact, not a behavior!",
       x = "Day of the Month", y = "Count")
```

Furthermore, `doy` variable suffers from inconsistent data density.

```{r doy-incosistent-3-years-span}
# 1. Ensure you have a Date object
# If you don't have a 'date' column, create one from year, month, day
plot_data <- vehstops %>%
  mutate(date = as.Date(paste(year, month, day, sep = "-"))) %>%
  count(date)

# 2. Plot the timeline
plot_data %>%
  ggplot(aes(x = date, y = n)) +
  # Raw data points (using alpha to reduce overplotting)
  geom_line(color = "grey70", alpha = 0.6, linewidth = 0.5) +
  # Trend line (using a slightly larger span to show the multi-year trajectory)
  geom_smooth(method = "loess", span = 0.2, color = "steelblue", se = FALSE, linewidth = 1.2) +
  scale_x_date(date_breaks = "3 months", date_labels = "%b %Y") +
  theme_minimal() +
  theme(
    panel.grid.minor = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(size = 14),
    plot.subtitle = element_text(color = "grey50", size = 10)
  ) +
  labs(
    title = "Volume of Stops by Day of the Year (1–365) for 2-Wheels Vehicles",
    subtitle = "Continuous time series shows that there might be a pattern drifting towards a bell centered in August",
    x = "Period", 
    y = "Daily Stop Count"
  )
```

```{r, include=FALSE}
rm(plot_data)
```

While we observe a consistent seasonal clustering peaking in late summer, our current 3-year dataset captures three seasonal cycles. This is sufficient to identify a recurring pattern, but insufficient to establish it as a long-term trend.

The next plot, demonstrates high daily volatility (indicated by the grey jitter) that significantly obscures the underlying seasonal trend (the blue trend line). Because the daily fluctuation is massive relative to the overall pattern, including 'Day of Year' as a feature risks overfitting.

```{r doy-inconsistent-density}
vehstops %>%
  filter(!is.na(doy)) %>%
  count(doy) %>%
  ggplot(aes(x = doy, y = n)) +
  geom_line(color = "grey60", alpha = 0.5) +
  geom_smooth(method = "loess", color = "steelblue", se = FALSE) +
  theme_minimal() +
   theme(
    panel.grid.major.x = element_blank(),
    plot.title = element_text(size = 14),
    plot.subtitle = element_text(color = "grey50", size = 10)
  ) +
  labs(title = "Volume of Stops by Day of the Year (1–365) for 2-Wheels Vehicles",
       subtitle = "The extreme jitter confirms this feature is too noisy to predict",
       x = "Day of Year", y = "Count")
```

Finally, these granular temporal markers are prone to overfitting. Excluding them allows the model to focus on more robust predictors, such as seasonality and weekly rhythms.

On the other hand, we can study the "Day of the Week" which usually abstract better.

```{r dow-dis}
vehstops %>%
  filter(!is.na(dow)) %>%
  count(dow) %>%
  mutate(
    dow = factor(dow, levels = 1:7, 
                 labels = c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun")),
    # Calculate % change relative to the previous day
    pct_change = (n - lag(n)) / lag(n) * 100,
    # Create the label: Count on top, % change on bottom
    label_text = ifelse(is.na(pct_change), 
                        comma(n), 
                        paste0(comma(n), "\n(", ifelse(pct_change > 0, "+", ""), round(pct_change, 1), "%)"))
  ) %>%
  ggplot(aes(x = dow, y = n, group = 1)) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_point(color = "steelblue", size = 3) +
  # vjust = -0.6 lifts text off the point, lineheight compacts the text
  geom_text(aes(label = label_text), vjust = -1, size = 3, lineheight = 0.8) +
  scale_y_continuous(labels = comma, 
                     limits = c(0, NA), 
                     expand = expansion(mult = c(0, 0.4))) + # Extra room for labels
  theme_minimal() +
  theme(
    panel.grid.major.x = element_blank(),
    plot.title = element_text(size = 14),
    plot.subtitle = element_text(color = "grey50", size = 10)
  ) +
  labs(
    title = "Daily Volume of Stops for 2-Wheels Vehicles",
    subtitle = "Daily volume and day-over-day growth/decline show high stop counts during mid-week",
    x = "Day of the Week", 
    y = "Count"
  )
```

What can we say?

-   Key finding: Enforcement activity follows a distinct "Mid-Week Peak" cycle. The volume starts at its lowest on Monday, surges aggressively (+52.7%) into Tuesday, peaks on Wednesday/Thursday, and then exhibits a sustained decline toward the weekend.
-   Interpretation: This pattern strongly suggests that these stops are primarily linked to **commercial usage** (e.g., food delivery/courier services) rather than recreational riding. Recreational riders typically peak on weekends; the fact that your stops drop off on Saturday and Sunday reinforces that this enforcement is targeting "work-hour" traffic. The +52% jump from Monday to Tuesday is the most critical signal here—it suggests that Monday may be a "low volume" day for this specific gig-economy demographic, or that enforcement resources are being ramped up as the work week settles into its rhythm.
-   Comment: We have identified a non-linear relationship: the "Day of Week" effect isn't a straight line, it's a curve.
-   Proposed Action: We may keep `weekday` as the new `day` in the model, and even think about an interaction with `hot_season`. Also we need first to check the signal from other indicators previously built.

### Weekend, Holiday and Event Day

Let us see our feature-engineered variables.

```{r weekend-holiday-event-distribution}
p1 <- plot_binary_bar(
  vehstops, weekend, 
  "No", "Yes",
  "", "Count", "During a weekend?"
)

p2 <- plot_binary_bar(
  vehstops, holiday,
  "No", "Yes",
  "", "", "During an holiday?"
)

p3 <- plot_binary_bar(
  vehstops, event_day,
  "No", "Yes",
  "", "", "During an event?"
)

(p1 + p2 + p3) + 
  plot_layout(ncol = 3, nrow = 1, widths = c(1, 1), guides = "collect") +
  plot_annotation(title = "When did the stop take place?") &
  theme(legend.position = "bottom") 
```

```{r, include=FALSE}
rm(p1, p2, p3)
```

What can we say?

-   Key finding: Enforcement is defined by routine, daily operations rather than event-based triggers. Among the three temporal variables analyzed, only "Weekend" status provides a meaningful statistical differentiation.
-   Interpretation: Police activity appears to be highly consistent, suggesting that enforcement is targeted at baseline daily traffic flow (likely commercial/commuter) rather than reacting to external calendar anomalies. The lack of variation during holidays and events indicates that these stops are not "surge" enforcement but rather a steady-state, daily commitment.
-   Comment: The "Holiday" and "Event" categories suffer from severe class imbalance (96–97% 'No'). Statistically, these features lack the variance required to train a robust model and will function more as "background noise" than as actionable predictive indicators.
-   Proposed Action: Exclude holiday and event_day from the feature set to reduce model complexity and prevent overfitting. Retain weekend as a core predictor.

### Hour, Working Hour and Time Bracket

First of all we look at the overall cycle of stops volume

```{r hour-dist}
# 1. Define the specific hours you want to highlight
hours_to_show <- c(0, 4, 8, 12, 16, 19, 20, 23)

# 2. Prepare the full dataset for the background line
full_data <- vehstops %>%
  filter(!is.na(hour)) %>%
  count(hour)

# 3. Prepare the highlighted data for points and labels
highlight_data <- full_data %>%
  filter(hour %in% hours_to_show) %>%
  arrange(hour) %>%
  mutate(
    # Percent change relative to the PREVIOUS HOUR IN THE LIST
    pct_change = (n - lag(n)) / lag(n) * 100,
    label_text = ifelse(
      is.na(pct_change), 
      comma(n),
      paste0(comma(n), "\n(", ifelse(pct_change > 0, "+", ""), round(pct_change, 1), "%)"))
  )

# 4. Plot
ggplot() +
  # Background line showing the full 24-hour context
  geom_line(data = full_data, aes(x = hour, y = n), color = "grey80", linewidth = 1) +
  # Points and labels for specific hours only
  geom_point(data = highlight_data, aes(x = hour, y = n), color = "steelblue", size = 3) +
  geom_text(data = highlight_data, aes(x = hour, y = n, label = label_text), 
            vjust = -0.6, size = 3, lineheight = 0.8) +
  # Scaling and Theme
  scale_x_continuous(breaks = 0:23) +
  scale_y_continuous(labels = comma, limits = c(0, NA), expand = expansion(mult = c(0, 0.4))) +
  theme_minimal() +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    plot.title = element_text(size = 14),
    plot.subtitle = element_text(color = "grey50", size = 10)
  ) +
  labs(
    title = "Hourly Volume of Stops for 2-Wheels Vehicles",
    subtitle = "Aggregated stops showing velocity between key hours",
    x = "Hour of Day (0–23)", 
    y = "Count"
  )
```

```{r, include=FALSE}
rm(hours_to_show, full_data, highlight_data)
```

What can we say? In our opinion this looks like a "sine wave broken by a bridge". To be more precise:

-   Key finding: Enforcement volume is characterized by a "fractured" daily cycle. It follows a predictable overnight decline and morning ascent (0:00–12:00), but is interrupted by a distinct midday plateau (12:00–15:00) before reaching an aggressive evening peak at 19:00.
-   Interpretation: The data behaves like a sine wave that hits a structural impediment at noon. This suggests that the enforcement cycle is divided into two distinct operational phases: the **Morning Phase** (likely commuter-driven) and the **Evening Phase** (likely delivery/service-driven). The plateau is the "bridge" between these two worlds—a period of transition where ridership demographics potentially shift from commuters to service workers. Therefore, the variability during 10:00 to 15:00 is limited because the amount of vehicles on the road remains stable, as well as the police work.
-   Comment: This "bridge" (12:00–15:00) is the most revealing part of the graph. If this were a simple, smooth traffic curve, we would expect a direct climb to the peak. The fact that it plateaus indicates a "lull" or a shift in the operational tempo. It proves that a single `rush_hour` binary variable will be insufficient for your model; it will fail to capture the reality that the "evening rush" is a different phenomenon entirely from the "morning commute."
-   Proposed Action: We could optionally modify our bucketing of `time_bracket` into: `Overnight` (22:00–05:00), `Morning_Commute` (06:00–11:00), `Midday_Transition` (12:00–15:00), and `Evening_Peak` (16:00–21:00).

Before anything else, let us see whether this cycle changed throughout the years.

TODO: Title and subtitle

```{r hour-year-dist}
# 1. Define the specific hours you want to highlight
hours_to_show <- c(0, 4, 8, 12, 16, 20, 23)

# 2. Prepare the full background data for the lines
full_data <- vehstops %>%
  filter(!is.na(year) & !is.na(hour)) %>%
  count(year, hour)

# 3. Prepare the highlighted data for points and labels
# We filter for our key hours *before* calculating percentage change
# so it calculates the jump between those specific selected points.
highlight_data <- full_data %>%
  filter(hour %in% hours_to_show) %>%
  group_by(year) %>%
  arrange(hour) %>%
  mutate(
    pct_change = (n - lag(n)) / lag(n) * 100,
    label_text = ifelse(is.na(pct_change), 
                        comma(n), 
                        paste0(comma(n), "\n(", ifelse(pct_change > 0, "+", ""), round(pct_change, 1), "%)"))
  )

# 4. Plot
ggplot() +
  # Background line showing the full 24-hour context per year
  geom_line(data = full_data, aes(x = hour, y = n), color = "grey85", linewidth = 0.8) +
  # Points and labels for specific hours only
  geom_point(data = highlight_data, aes(x = hour, y = n), color = "steelblue", size = 2) +
  geom_text(data = highlight_data, aes(x = hour, y = n, label = label_text), 
            vjust = -0.7, size = 2.5, lineheight = 0.8) +
  # Facet by year
  facet_wrap(~year, nrow = 1) +
  # Scaling and Theme
  scale_x_continuous(breaks = c(0, 4, 8, 12, 16, 20, 23)) +
  scale_y_continuous(labels = comma, limits = c(0, NA), expand = expansion(mult = c(0, 0.4))) +
  theme_minimal() +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    plot.title = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(color = "grey50", size = 10),
    strip.text = element_text(size = 11, face = "bold")
  ) +
  labs(
    title = "Hourly Volume of Stops for 2-Wheels Vehicles",
    subtitle = "Aggregated stops showing velocity between key hours per year",
    x = "Hour of Day (0–23)", 
    y = "Count"
  )
```

```{r, include=FALSE}
rm(hours_to_show, full_data, highlight_data)
```

What can we say? \* Key finding: We are witnessing an "Evening Migration" of enforcement. While 2023 enforcement peaked at 16:00 (4:00 PM), the peak has decisively migrated to 20:00 (8:00 PM) in 2024 and 2025. Additionally, the enforcement volume near the 19:00-20:00 peak has nearly tripled between 2023 and 2025. \* Interpretation: The "operational center of gravity" has shifted later into the day. This suggests that the primary targets of enforcement (likely delivery/gig-economy riders) are increasingly active later into the evening, or that enforcement strategy has pivoted to address higher evening density. The "Morning Rush" (8:00–12:00) is growing, but the "Evening Rush" (16:00–20:00) is the clear driver of the total volume increase. \* Comment: It is critical to note that the shape of the curve is changing. In 2023, the distribution was flatter; by 2025, it has become a "sharper" curve with a very pronounced evening spike. This indicates that your predictive model cannot rely on a static "hour" feature. If the model treats "Hour 20" the same way in 2023 as it does in 2025, it will drastically under-predict the volume for 2025. \* Proposed Action: When building the model, we might consider to include an interaction term between `hour` and `year`. This will mathematically allow the model to learn that the "value" of 8:00 PM changes as the years progress. Alternatively we can better refine the binning dynamically over specific periods of time.

Now let us see if the hot season might shift this cycle

TODO: Title and Subtitle

```{r hour-hot-season}
# 1. Prepare data with hot_season variable
# Assuming month column exists. If not, you may need to join/create it.
plot_data_full <- vehstops %>%
  mutate(
    hot_season = ifelse(month %in% 4:10, 1, 0),
    hot_season_label = factor(hot_season, levels = c(0, 1), 
                              labels = c("Off Season (Nov-Mar)", "Hot Season (Apr-Oct)"))
  ) %>%
  filter(!is.na(hour))

# 2. Prepare full background data
full_data <- plot_data_full %>%
  count(hot_season_label, hour)

# 3. Define key hours and calculate velocity
hours_to_show <- c(0, 4, 8, 12, 16, 20, 23)

highlight_data <- full_data %>%
  filter(hour %in% hours_to_show) %>%
  group_by(hot_season_label) %>%
  arrange(hour) %>%
  mutate(
    pct_change = (n - lag(n)) / lag(n) * 100,
    label_text = ifelse(is.na(pct_change), 
                        comma(n), 
                        paste0(comma(n), "\n(", ifelse(pct_change > 0, "+", ""), round(pct_change, 1), "%)"))
  )

# 4. Faceted plot
ggplot() +
  geom_line(data = full_data, aes(x = hour, y = n), color = "grey85", linewidth = 0.8) +
  geom_point(data = highlight_data, aes(x = hour, y = n), color = "steelblue", size = 2) +
  geom_text(data = highlight_data, aes(x = hour, y = n, label = label_text), 
            vjust = -0.7, size = 2.5, lineheight = 0.8) +
  facet_wrap(~hot_season_label, nrow = 1) +
  scale_x_continuous(breaks = c(0, 4, 8, 12, 16, 20, 23)) +
  scale_y_continuous(labels = comma, limits = c(0, NA), expand = expansion(mult = c(0, 0.4))) +
  theme_minimal() +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    plot.title = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(color = "grey50", size = 10),
    strip.text = element_text(size = 12, face = "bold")
  ) +
  labs(
    title = "Hourly Cycle of Enforcement: Hot Season vs. Off Season",
    subtitle = "Comparing daily volume and velocity between seasons",
    x = "Hour of Day (0–23)", 
    y = "Count"
  )
```

```{r, include=FALSE}
rm(plot_data_full, hours_to_show, full_data, highlight_data)
```

What can we say?

-   Key finding: The "Center of Gravity" for enforcement shifts significantly between seasons. In the Off Season (Nov–Mar), activity peaks at 16:00 and declines rapidly thereafter. In the Hot Season (Apr–Oct), the peak is delayed until 20:00, with high volume sustained well into the evening.
-   Interpretation: Enforcement curve is governed by the "Sun and Temperature" cycle. During the colder months, ridership and commercial delivery activity are compressed into the earlier, slightly warmer/brighter parts of the day. In the Hot Season, the extended daylight and milder evening temperatures allow the "rush" to sustain four hours longer, shifting the peak demand to later in the day.
-   Comment: If we treat `hour` and `hot_season` as separate variables, our models will be "blind" to this shift. So it is best to use interaction terms and bin the hours in the best way possible.
-   Proposed Action: We must check the hours against the targets' ratios to double check our previous binning and then we can consider restricting the bins.

Finally, let us see if the weekend has some kind of effect on the cycle.

TODO: title and subtitle

```{r hour-weekend}
# 1. Prepare data with weekend variable
# Creating readable labels for the plot
plot_data_full <- vehstops %>%
  filter(!is.na(weekend) & !is.na(hour)) %>%
  mutate(
    weekend_label = factor(weekend, levels = c(0, 1), 
                           labels = c("Weekday", "Weekend"))
  ) %>%
  count(weekend_label, hour)

# 2. Define key hours for labels
hours_to_show <- c(0, 4, 8, 12, 16, 20, 23)

# 3. Calculate velocity between the key hours for each group
highlight_data <- plot_data_full %>%
  filter(hour %in% hours_to_show) %>%
  group_by(weekend_label) %>%
  arrange(hour) %>%
  mutate(
    pct_change = (n - lag(n)) / lag(n) * 100,
    label_text = ifelse(is.na(pct_change), 
                        comma(n), 
                        paste0(comma(n), "\n(", ifelse(pct_change > 0, "+", ""), round(pct_change, 1), "%)"))
  )

# 4. Plot faceted by weekend status
ggplot() +
  geom_line(data = plot_data_full, aes(x = hour, y = n), color = "grey85", linewidth = 0.8) +
  geom_point(data = highlight_data, aes(x = hour, y = n), color = "steelblue", size = 2) +
  geom_text(data = highlight_data, aes(x = hour, y = n, label = label_text), 
            vjust = -0.7, size = 2.5, lineheight = 0.8) +
  facet_wrap(~weekend_label, nrow = 1) +
  scale_x_continuous(breaks = c(0, 4, 8, 12, 16, 20, 23)) +
  scale_y_continuous(labels = comma, limits = c(0, NA), expand = expansion(mult = c(0, 0.4))) +
  theme_minimal() +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    plot.title = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(color = "grey50", size = 10),
    strip.text = element_text(size = 12, face = "bold")
  ) +
  labs(
    title = "Hourly Cycle of Enforcement: Weekday vs. Weekend",
    subtitle = "Comparing daily volume and velocity of stops",
    x = "Hour of Day (0–23)", 
    y = "Count"
  )
```

```{r, include=FALSE}
rm(plot_data_full, hours_to_show, highlight_data)
```

What can we say?

-   Key finding: Weekday enforcement volume is approximately 2.5x to 3x higher than weekend enforcement. Furthermore, weekdays exhibit a sharp, "commuter-style" intensity, while weekends follow a significantly flatter, more gradual curve with much lower overall stop counts.
-   Interpretation: The weekday curve mirrors the "rush hour" cycle of a working city. The weekend curve, while still peaking at 20:00, is far less aggressive, suggesting that weekend stops are driven by a different, likely more dispersed, set of riders
-   Comment: The most striking observation is that the peak hour remains 20:00 (8:00 PM) for both weekdays and weekends. This is a critical "Signal vs. Noise" finding. It implies that while the volume changes drastically between Saturday and Tuesday, the policy or behavioral factor driving the evening enforcement is consistent regardless of the day of the week.
-   Proposed Action: Keep the weekend binary flag, but do not stop there. Because the shape of the curve is so different, you should add an Interaction Term (time_bracket \* weekend) to your model. This will allow the model to learn that the "Afternoon/Evening" bracket is a high-volume "danger zone" on a Weekday but much less significant on a Weekend.

Let us check the time brackets previously built.

```{r time-bracket-distribution}
plot_univariate_cat(
  DF = vehstops, 
  FEATURE = time_bracket, 
  X_LABEL = "Time Bracket", 
  TITLE = "Distribution of Time Brackets",
)
```

What can we say?

-   Key finding: Enforcement is heavily skewed toward the latter half of the day. The "Afternoon/evening" (16–19) and "Night" (20–23) brackets collectively account for the majority of stops, with the afternoon/evening block acting as the dominant peak. Consistent time-bracket definitions are essential for maintaining longitudinal integrity. Modifying brackets to "fit" observed peaks introduces selection bias that invalidates historical comparisons across your three-year dataset.
-   Interpretation: This confirms that the "Evening Migration" we observed in the previous charts is a structural trend. This effectively validates that your categorical buckets align with real-world activity peaks, rather than arbitrary time splits.
-   Comment: This is a clean, robust way to feed time data into a machine learning model. By using these brackets, you have significantly reduced the "noise" (like the variance between 16:00 and 17:00) while keeping the "signal" (the fact that 16:00–23:00 is high volume). This reduces the risk of overfitting compared to using the raw hour (0–23) or day_of_year (1–365) variables we discarded earlier.
-   Proposed Action: Maintain your existing time_bracket definitions exactly as they are. Instead of changing them, introduce an interaction term (time_bracket \* year) in your model.

```{r working-hour-distribution}
plot_binary_bar(
  vehstops, working_hour, 
  "No", "Yes",
  "Was it between 9am and 5pm?", "Count", "Distribution of Stops by 9-5 Window"
)
```

What can we say?

-   Key finding: The 9-5 binary feature is an overly reductive "blunt instrument" that collapses high-variance temporal signals into two broad, low-information categories.
-   Interpretation: By lumping the entire work day (9 AM to 5 PM) into a single "Yes" bucket, you are masking the most important insights we’ve already mapped: the morning growth, the "midday bridge" plateau, and the evening peak. The "No" bucket is equally problematic, as it mixes low-activity early morning hours with high-activity evening hours. This creates an "information bottleneck" that prevents the model from seeing the true operational rhythm.
-   Comment: This variable is redundant. You have already engineered time_bracket, which covers the same temporal space but does so with much higher granularity and precision.
-   Proposed Action: Drop the 9-5 feature immediately.

Also, we are going to drop minute as its level of granularity cannot be properly included in the dataset.

```{r drop-working-hour-minute}
vehstops <- vehstops %>%
  select(-working_hour, -minute)
```

## Motorist related predictors

### Reported Age and Age Bracket

We have mainly one issue here. First, we have many instances of drivers with unknown age,

```{r age-bracket-dist}
plot_univariate_cat(
  DF = vehstops, 
  FEATURE = age_bracket, 
  X_LABEL = "Age Bracket", 
  TITLE = "Distribution of Stops by Age Bracket",
)
```

What can we say?

-   Key finding: Enforcement is highly concentrated in the "Early Career" demographic (18–44), with a massive, dominant peak at 25–34. The "Unknown" age category represents roughly 4–5% of the total dataset—a significant enough portion to require a specific strategy, but not enough to undermine the reliability of the overall distribution.
-   Interpretation: The data reflects a clear operational focus on younger, active drivers, which aligns perfectly with a gig-economy or courier demographic. The "Unknown" values are likely "Missing At Random" (MAR)—clerical omissions during the stop process (e.g., driver didn't have ID, officer was in a rush)—rather than a deliberate, systematic suppression of age data.
-   Comment: We do not treat the "Unknown" category as garbage data to be discarded. However, we could find useful to merge some categories such as Under 18 and 18-24, and adults between 55-64 and senior of 65+ years of age. This could be paired with some re-binning to shift the volume of overall stops and reduce the number of bins.
-   Proposed Action: We should check first how this

And now, let us see how this looks when looking at the full distribution

FIXME: Horrible color scale TODO: Title, subtitle, annotations

```{r age-vs-age-bracket}
# 1. Create the brackets and clean the data
age_data <- vehstops %>%
  filter(!is.na(age), age >= 0, age < 100) %>%
  mutate(
    age_group = case_when(
      age < 18 ~ "<18",
      age >= 18 & age <= 24 ~ "18–24",
      age >= 25 & age <= 34 ~ "25–34",
      age >= 35 & age <= 44 ~ "35–44",
      age >= 45 & age <= 54 ~ "45–54",
      age >= 55 & age <= 64 ~ "55–64",
      age >= 65 ~ "65+"
    ),
    # Ensure the order is kept in the plot
    age_group = factor(age_group, levels = c("<18", "18–24", "25–34", "35–44", "45–54", "55–64", "65+"))
  )

# 2. Plotting
ggplot(age_data, aes(x = age, fill = age_group)) +
  # binwidth = 1 keeps the individual resolution
  geom_histogram(binwidth = 1, color = "white", linewidth = 0.1) +
  # Adding clear color differentiation for the brackets
  scale_fill_viridis_d(option = "mako") +
  theme_minimal() +
   theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    plot.title = element_text(size = 14),
    plot.subtitle = element_text(color = "grey50", size = 10),
    strip.text = element_text(size = 11)
  ) +
  labs(
    title = "Age Distribution of Drivers by Target Bracket",
    subtitle = "Individual year resolution with custom enforcement brackets",
    x = "Driver Age", 
    y = "Count",
    fill = "Age Group"
  ) +
  theme(legend.position = "bottom") +
  guides(fill = guide_legend(nrow = 2))
```

```{r, include=FALSE}
rm(age_data)
```

What can we say?

-   Key finding: Enforcement is heavily concentrated within the 20–45 age range, with a distinct operational peak occurring in the mid-20s. After age 50, there is a continuous, rapid decline in stop volume, forming a long, sparse "tail" that extends past 75.
-   Interpretation:
-   Comment:
-   Proposed Action: Filter your dataset to exclude drivers over age 75. This clarifies the "true" shape of your enforcement curve without losing actionable predictive signal. Collapse Sparse Brackets: Since the 55–64 and 65+ buckets have significantly lower counts than the 25–34 bucket, consider collapsing all drivers aged 55+ into a single "55+" category. This increases the sample size for that bin, ensuring the model has enough data to learn a stable weight for the older demographic. After checking ratio.

```{r remove-age-outliers}
vehstops <- vehstops %>%
  filter(age < 76)
```

### Reported Sex

TODO: Make it better

```{r sex-distribution}
plot_univariate_cat(
  DF = vehstops, 
  FEATURE = sex, 
  X_LABEL = "Reported Sex", 
  TITLE = "Distribution of Stops by Reported Sex (Sorted by Count)",
  order_by = "freq"
)
```

What can we say?

-   Key finding: The data exhibits extreme class imbalance. "Male" (M) drivers represent the overwhelming majority of enforcement stops (approximately 96%), while "Female" (F) and "Unknown" (U) represent a negligible fraction of the total volume.
-   Interpretation: This is not a data quality issue—it is an operational reflection of the driver demographic being targeted Because the skew is so profound, "Sex" will inevitably become a high-importance feature in any model you train. The model will essentially learn that "Being Male = High Probability of Stop" simply because that is the status quo in the dataset.
-   Comment: You must be cautious here. If you use this feature as-is, you risk creating a model that is "biased by design." Because the dataset is so skewed, the model could achieve very high mathematical accuracy by simply predicting "Male" for every single case. You are currently looking at exposure (who is on the road), not necessarily propensity (who is most likely to be stopped).
-   Proposed Action: Run a correlation matrix to see if sex is highly correlated with other variables. When splitting your data for training/testing, ensure you use stratified sampling (preserving the M/F/U ratio in both sets). If you don't, your test set might end up with no "Female" or "Unknown" records, making it impossible to evaluate your model's performance on those sub-groups

### Reported Race

TODO: Better title, subtitle, annotation with percentage, less angle

```{r race-distribution}
plot_univariate_cat(
  DF = vehstops, 
  FEATURE = race, 
  X_LABEL = "Reported Race", 
  TITLE = "Distribution of Stops by Reported Race (Sorted by Count)",
  order_by = "freq"
)
```

What can we say?

-   Key finding: The distribution is heavily skewed, with "Hispanic" drivers accounting for the plurality of stops (over 86,000), far outpacing other groups. The "Unknown" category is also statistically significant (\~16,800), mirroring the data quality patterns you observed with age.
-   Interpretation: This data does not reflect a balanced, random sampling of the driving population; it reflects an enforcement density. The high concentration of specific demographics suggests that either the geographic areas targeted for enforcement have high Hispanic populations, or the specific types of vehicles/courier work being monitored are disproportionately staffed by this demographic.
-   Comment: Because the distribution is so imbalanced, a machine learning model will aggressively use "Race" to predict stops. It will effectively "learn" that a stop is more likely for one group over another based on the historical pattern.
-   Proposed Action: Before including race in your regression, perform a correlation check against boro. When training your model, ensure your training and testing splits are stratified by race.

------------------------------------------------------------------------

# Bivariate EDA

## Correlation

FIXME: Do only upper half,

```{r categorical-correlation-heatmap}
# TODO
colonne_da_verificare <- c("arrested", "veh_checkpoint", "sex", "race", "age", "age_bracket",  "area", "boro", "year" , "month", "hot_season", "weekday", "weekend", "holiday", "event_day", "time_bracket", "year")

plot_categorical_correlation(
  DF = vehstops, 
  FEATURES = colonne_da_verificare,
  TITLE = "Correlazione tra le Variabili (Cramer's V)"
)
```

What can we say?

-   Key finding:
-   Interpretation:
-   Comment:
-   Proposed Action:

What we are going to look at now is

## Geospatial Features

### Area

```{r area-vs-arrested-ratio}
ratio_by_cat(
  vehstops, arrested, area,
  "Area", "Arrest Rate (%)", "Arrest Ratio by Area",
  style = "stack", legend_title = "Stop ended in arrest?",
  target_labels = c("No", "Yes"),
  subtitle = "For 2-Wheels Vehicles it shows Increased Arrest Ratio in Low Volume Areas",
  label_vjust = 0.5
)
```

Let us look at the numbers too

```{r confirm-area-vs-arrested-ratio}
vehstops %>%
  group_by(area, arrested) %>%
  summarise(count = n(), .groups = "drop_last") %>%
  mutate(percentage = count / sum(count))
```

What can we say? Remember that Staten Island and Bronx are the areas with the least volume of stops.

-   Key finding:
-   Interpretation:
-   Comment:
-   Proposed Action:

What about checkpoints? For ease of display we are not going to stack the bars, but just displaying the rate percentages of vehicle checkpoints.

```{r area-vs-checkpoint-ratio}
ratio_by_cat(
  vehstops, veh_checkpoint, area,
  "Area", "Vehichle Checkpoint Rate (%)", "Vehichle Checkpoint Rate by Area",
  style = "rate", legend_title = "Stop was done by a checkpoint?",
  target_labels = c("No", "Yes"),
  subtitle = "For 2-Wheels Vehicles it shows Low Checkpoint Ratio in Low Volume Areas"
)
```

Let us look at the numbers too

```{r confirm-area-vs-checkpoint-ratio}
vehstops %>%
  group_by(area, veh_checkpoint) %>%
  summarise(count = n(), .groups = "drop_last") %>%
  mutate(percentage = count / sum(count))
```

What can we say? Low volume areas have less checkpoints, but get proportionally more arrests

-   Key finding:
-   Interpretation:
-   Comment:
-   Proposed Action:

### Borough

```{r boro-vs-arrested-ratio}
ratio_by_cat(
  vehstops, arrested, boro,
  "Borough", "Arrest Rate (%)", "Arrest Ratio by Borough",
  style = "stack", legend_title = "Stop ended in arrest?",
  target_labels = c("No", "Yes"),
  subtitle = "For 2-Wheels Vehicles it confirms the Increased Arrest Ratio in Low Volume Areas",
  label_vjust = 0.5
)
```

What can we say?

-   Key finding:
-   Interpretation:
-   Comment:
-   Proposed Action:

What about the checkpoint?

```{r boro-vs-checkpoint-ratio}
ratio_by_cat(
  vehstops, veh_checkpoint, boro,
  "Borough", "Vehicle Checkpoint Rate (%)", "Vehicle Checkpoint by Borough",
  style = "rate", legend_title = "Stop was done by a checkpoint?",
  target_labels = c("No", "Yes"),
  subtitle = "For 2-Wheels Vehicles it confirms the Low Checkpoint Ratio in Low Volume Areas",
  label_vjust = 0.5
)
```

What can we say?

-   Key finding:
-   Interpretation:
-   Comment:
-   Proposed Action:

We will need to choose between one or the other.

## Time related features

### Year

FIXME: Do by line plot by year

```{r year-vs-arrested-ratio}
# 1. Calculate Arrest Rate per year
arrest_trends <- vehstops %>%
  filter(!is.na(year), !is.na(arrested)) %>%
  group_by(year) %>%
  summarise(
    total_stops = n(),
    arrests = sum(arrested == 1),
    rate = arrests / total_stops, 
    .groups = 'drop'
  ) %>%
  arrange(year) %>%
  # 2. Calculate the year-over-year change in percentage points
  mutate(
    prev_rate = lag(rate),
    delta = (rate - prev_rate) * 100, # Convert to percentage points
    # 3. Create labels: Rate + Delta
    label_text = ifelse(
      is.na(prev_rate), 
      scales::percent(rate, accuracy = 0.1), 
      paste0(scales::percent(rate, accuracy = 0.1), 
             "\n(", ifelse(delta > 0, "+", ""), round(delta, 1), " pts)")
    )
  )

# 4. Plotting
ggplot(arrest_trends, aes(x = factor(year), y = rate, group = 1)) +
  # Draw the trend line
  geom_line(color = "steelblue", linewidth = 1) +
  geom_point(color = "steelblue", size = 3) +
  # Add the labels
  geom_text(aes(label = label_text), vjust = -0.8, size = 3.5, lineheight = 0.8) +
  # Scale y-axis as percentage
  scale_y_continuous(
    labels = scales::percent, 
    limits = c(0, max(arrest_trends$rate) * 1.3), # Extra room for labels
    expand = expansion(mult = c(0, 0.1))
  ) +
  theme_minimal() +
  theme(
    panel.grid.major.x = element_blank(),
    plot.title = element_text(size = 14),
    plot.subtitle = element_text(size = 10, color = "grey50")
  ) +
  labs(
    title = "Annual Arrest Rate has been Slowly Rising in the last 3 years",
    subtitle = "Calculated as the percentage of stops resulting in an arrest",
    x = "Year", y = "Arrest Rate (%)"
  )
```

What can we say?

-   Key finding:
-   Interpretation:
-   Comment:
-   Proposed Action:

```{r year-vs-checkpoint-ratio}
# 1. Calculate Arrest Rate per year
arrest_trends <- vehstops %>%
  filter(!is.na(year), !is.na(veh_checkpoint)) %>%
  group_by(year) %>%
  summarise(
    total_stops = n(),
    arrests = sum(veh_checkpoint == 1),
    rate = arrests / total_stops, 
    .groups = 'drop'
  ) %>%
  arrange(year) %>%
  # 2. Calculate the year-over-year change in percentage points
  mutate(
    prev_rate = lag(rate),
    delta = (rate - prev_rate) * 100, # Convert to percentage points
    # 3. Create labels: Rate + Delta
    label_text = ifelse(
      is.na(prev_rate), 
      scales::percent(rate, accuracy = 0.1), 
      paste0(scales::percent(rate, accuracy = 0.1), 
             "\n(", ifelse(delta > 0, "+", ""), round(delta, 1), " pts)")
    )
  )

# 4. Plotting
ggplot(arrest_trends, aes(x = factor(year), y = rate, group = 1)) +
  # Draw the trend line
  geom_line(color = "steelblue", linewidth = 1) +
  geom_point(color = "steelblue", size = 3) +
  # Add the labels
  geom_text(aes(label = label_text), vjust = -0.8, size = 3.5, lineheight = 0.8) +
  # Scale y-axis as percentage
  scale_y_continuous(
    labels = scales::percent, 
    limits = c(0, max(arrest_trends$rate) * 1.3), # Extra room for labels
    expand = expansion(mult = c(0, 0.1))
  ) +
  theme_minimal() +
  theme(
    panel.grid.major.x = element_blank(),
    plot.title = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 10, color = "grey50")
  ) +
  labs(
    title = "Annual Vehicle Checkpoint Rate has been barely decreasing in the last 3 years",
    subtitle = "Calculated as the percentage of stops from vehicle checkpoints",
    x = "Year", y = "Vehicle Checkpoint Rate (%)"
  )
```

What can we say?

-   Key finding:
-   Interpretation:
-   Comment:
-   Proposed Action:

### Month and Hot Season

```{r month-vs-arrested-ratio}
# 1. Aggregate and calculate monthly arrest rate
monthly_arrest_data <- vehstops %>%
  filter(!is.na(month), !is.na(arrested)) %>%
  group_by(month) %>%
  summarise(
    total_stops = n(),
    arrests = sum(arrested == 1),
    rate = arrests / total_stops,
    .groups = "drop"
  ) %>%
  mutate(month = factor(month, levels = 1:12)) %>%
  arrange(month) %>%
  # 2. Calculate point difference (delta)
  mutate(
    prev_rate = lag(rate),
    delta = (rate - prev_rate) * 100,
    # Create the label: Rate on top, Point change on bottom
    label_text = ifelse(
      is.na(prev_rate), 
      scales::percent(rate, accuracy = 0.1), 
      paste0(scales::percent(rate, accuracy = 0.1), 
             "\n(", ifelse(delta > 0, "+", ""), round(delta, 1), " pts)")
    )
  )

# 3. Plot
monthly_arrest_data %>%
  ggplot(aes(x = month, y = rate, group = 1)) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_point(color = "steelblue", size = 3) +
  # Use nudge_y if labels overlap the line, vjust handles the vertical position
  geom_text(aes(label = label_text), vjust = -1.2, size = 3, lineheight = 0.8) +
  scale_x_discrete(labels = month.abb) +
  scale_y_continuous(
    labels = scales::percent, 
    limits = c(0, max(monthly_arrest_data$rate) * 1.3), 
    expand = expansion(mult = c(0, 0.1))
  ) +
  theme_minimal() +
  theme(
    panel.grid.major.x = element_blank(),
    plot.title = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(color = "grey50", size = 10)
  ) +
  labs(
    title = "Monthly Arrest Rate Trend",
    subtitle = "Aggregated monthly arrest rate; variation shown in percentage points",
    x = "Month",
    y = "Arrest Rate (%)"
  )
```

What can we say?

-   Key finding:
-   Interpretation:
-   Comment:
-   Proposed Action:

```{r month-vs-checkpoint-ratio}
# 1. Aggregate and calculate monthly arrest rate
monthly_arrest_data <- vehstops %>%
  filter(!is.na(month), !is.na(veh_checkpoint)) %>%
  group_by(month) %>%
  summarise(
    total_stops = n(),
    arrests = sum(veh_checkpoint == 1),
    rate = arrests / total_stops,
    .groups = "drop"
  ) %>%
  mutate(month = factor(month, levels = 1:12)) %>%
  arrange(month) %>%
  # 2. Calculate point difference (delta)
  mutate(
    prev_rate = lag(rate),
    delta = (rate - prev_rate) * 100,
    # Create the label: Rate on top, Point change on bottom
    label_text = ifelse(
      is.na(prev_rate), 
      scales::percent(rate, accuracy = 0.1), 
      paste0(scales::percent(rate, accuracy = 0.1), 
             "\n(", ifelse(delta > 0, "+", ""), round(delta, 1), " pts)")
    )
  )

# 3. Plot
monthly_arrest_data %>%
  ggplot(aes(x = month, y = rate, group = 1)) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_point(color = "steelblue", size = 3) +
  # Use nudge_y if labels overlap the line, vjust handles the vertical position
  geom_text(aes(label = label_text), vjust = -1.2, size = 3, lineheight = 0.8) +
  scale_x_discrete(labels = month.abb) +
  scale_y_continuous(
    labels = scales::percent, 
    limits = c(0, max(monthly_arrest_data$rate) * 1.3), 
    expand = expansion(mult = c(0, 0.1))
  ) +
  theme_minimal() +
  theme(
    panel.grid.major.x = element_blank(),
    plot.title = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(color = "grey50", size = 10)
  ) +
  labs(
    title = "Monthly Vehicle Checkpoint Rate Trend",
    subtitle = "Aggregated monthly vehicle checkpoint rate; variation shown in percentage points",
    x = "Month",
    y = "Vehicle Checkpoint Rate (%)"
  )
```

What can we say?

-   Key finding:
-   Interpretation:
-   Comment:
-   Proposed Action:

What about `hot_season`?

FIXME: Correct it

```{r hot-season-vs-arrested-ratio}
ratio_by_cat(
  vehstops, arrested, hot_season,
  "Hot Season", "Arrest Rate (%)", "Arrest Ratio by Season",
  style = "stack", legend_title = "Stop ended in arrest?",
  target_labels = c("No", "Yes"),
  subtitle = "For 2-Wheels Vehicles it shows Increased Arrest Ratio in Low Volume Areas",
  label_vjust = 0.5
)
```

What can we say?

-   Key finding:
-   Interpretation:
-   Comment:
-   Proposed Action:

```{r hot-season-vs-checkpoint-ratio}
ratio_by_cat(
  vehstops, veh_checkpoint, hot_season,
  "Hot Season", "Vehicle Checkpoint Rate (%)", "  Ratio by Season",
  style = "stack", legend_title = " ",
  target_labels = c("No", "Yes"),
  subtitle = "For 2-Wheels Vehicles it shows  ",
  label_vjust = 0.5
)
```

What can we say?

-   Key finding:
-   Interpretation:
-   Comment:
-   Proposed Action:

#### Faceted

FIXME: Fix the annotations

```{r month-faceted-vs-arrested-ratio}
# Data prep: Calculate arrest rates and labeling logic
plot_data <- vehstops %>%
  filter(!is.na(year), !is.na(month), !is.na(arrested)) %>%
  group_by(year, month) %>%
  summarise(
    arrest_rate = mean(arrested == 1), 
    .groups = "drop"
  ) %>%
  mutate(month = factor(month, levels = 1:12)) %>%
  group_by(year) %>%
  mutate(
    rate_jan = arrest_rate[month == 1],
    rate_max = max(arrest_rate),
    is_max = arrest_rate == rate_max,
    
    # Calculate percentage point differences
    # Jan vs current, Max vs current
    diff_jan = (arrest_rate - rate_jan) * 100,
    diff_dec = (arrest_rate - rate_max) * 100, 
    
    # Custom Label Logic
    label_text = case_when(
      month == 1 ~ scales::percent(arrest_rate, accuracy = 0.1), 
      is_max ~ paste0(scales::percent(arrest_rate, accuracy = 0.1), "\n(", ifelse(diff_jan > 0, "+", ""), round(diff_jan, 1), " pts)"),
      month == 12 ~ paste0(scales::percent(arrest_rate, accuracy = 0.1), "\n(", round(diff_dec, 1), " pts)"),
      TRUE ~ ""
    )
  )

# Faceted plot for Arrest Rate
plot_data %>%
  ggplot(aes(x = month, y = arrest_rate, group = 1)) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_point(color = "steelblue", size = 2) +
  geom_text(aes(label = label_text), vjust = -1, hjust = 0.2, size = 2.5, lineheight = 0.8) +
  facet_wrap(~year, nrow = 1) +
  scale_x_discrete(
    labels = c("J", "F", "M", "A", "M", "J", "J", "A", "S", "O", "N", "D"), 
    expand = expansion(mult = c(0.1, 0.1))
  ) +
  # Using percentage scale for Y
  scale_y_continuous(labels = scales::percent, limits = c(0, NA), expand = expansion(mult = c(0, 0.4))) +
  coord_cartesian(clip = "off") +
  theme_minimal() +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    plot.title = element_text(size = 14),
    plot.subtitle = element_text(color = "grey50", size = 10),
    strip.text = element_text(size = 11)
  ) +
  labs(
    title = "Monthly Arrest Rate, by Year, for 2-Wheels Vehicles in NYC",
    subtitle = "Variation shown in percentage points (pts) relative to January and the Annual Peak",
    x = "Month",
    y = "Arrest Rate"
  )
```

What can we say?

-   Key finding:
-   Interpretation:
-   Comment:
-   Proposed Action:

```{r month-faceted-vs-checkpoint-ratio}
# Data prep: Calculate arrest rates and labeling logic
plot_data <- vehstops %>%
  filter(!is.na(year), !is.na(month), !is.na(veh_checkpoint)) %>%
  group_by(year, month) %>%
  summarise(
    arrest_rate = mean(veh_checkpoint == 1), 
    .groups = "drop"
  ) %>%
  mutate(month = factor(month, levels = 1:12)) %>%
  group_by(year) %>%
  mutate(
    rate_jan = arrest_rate[month == 1],
    rate_max = max(arrest_rate),
    is_max = arrest_rate == rate_max,
    
    # Calculate percentage point differences
    # Jan vs current, Max vs current
    diff_jan = (arrest_rate - rate_jan) * 100,
    diff_dec = (arrest_rate - rate_max) * 100, 
    
    # Custom Label Logic
    label_text = case_when(
      month == 1 ~ scales::percent(arrest_rate, accuracy = 0.1), 
      is_max ~ paste0(scales::percent(arrest_rate, accuracy = 0.1), "\n(", ifelse(diff_jan > 0, "+", ""), round(diff_jan, 1), " pts)"),
      month == 12 ~ paste0(scales::percent(arrest_rate, accuracy = 0.1), "\n(", round(diff_dec, 1), " pts)"),
      TRUE ~ ""
    )
  )

# Faceted plot for Arrest Rate
plot_data %>%
  ggplot(aes(x = month, y = arrest_rate, group = 1)) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_point(color = "steelblue", size = 2) +
  geom_text(aes(label = label_text), vjust = -1, hjust = 0.2, size = 2.5, lineheight = 0.8) +
  facet_wrap(~year, nrow = 1) +
  scale_x_discrete(
    labels = c("J", "F", "M", "A", "M", "J", "J", "A", "S", "O", "N", "D"), 
    expand = expansion(mult = c(0.1, 0.1))
  ) +
  # Using percentage scale for Y
  scale_y_continuous(labels = scales::percent, limits = c(0, NA), expand = expansion(mult = c(0, 0.4))) +
  coord_cartesian(clip = "off") +
  theme_minimal() +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    plot.title = element_text(size = 14),
    plot.subtitle = element_text(color = "grey50", size = 10),
    strip.text = element_text(size = 11)
  ) +
  labs(
    title = "Monthly Vehicle Checkpoint Rate, by Year, for 2-Wheels Vehicles in NYC",
    subtitle = "Variation shown in percentage points (pts) relative to January and the Annual Peak",
    x = "Month",
    y = "Vehicle Checkpoint Rate"
  )
```

What can we say?

-   Key finding:
-   Interpretation:
-   Comment:
-   Proposed Action:

```{r faceted-hot-season-vs-arrested-ratio}
# 1. Create the indicator, calculate rate and raw arrests
plot_data <- vehstops %>%
  filter(!is.na(year), !is.na(month), !is.na(arrested)) %>%
  mutate(
    is_hot_season = ifelse(month %in% 4:10, "Hot Season", "Off Season"),
    is_hot_season = factor(is_hot_season, levels = c("Hot Season", "Off Season"))
  ) %>%
  group_by(year, is_hot_season) %>%
  summarise(
    total_stops = n(),
    arrests = sum(arrested == 1),
    arrest_rate = arrests / total_stops,
    .groups = "drop"
  ) %>%
  mutate(
    # Label: Rate (Raw Arrest Count)
    label_text = paste0(scales::percent(arrest_rate, accuracy = 0.1), "\n(", comma(arrests), " arrests)")
  )

# 2. Build the faceted barplot
plot_data %>%
  ggplot(aes(x = is_hot_season, y = arrest_rate, fill = is_hot_season)) +
  geom_col(show.legend = FALSE) +
  # Facet by year
  facet_wrap(~year, nrow = 1) +
  # Add labels
  geom_text(aes(label = label_text), vjust = -0.5, size = 3, lineheight = 0.8) +
  # Use percent scale for the Y axis
  scale_y_continuous(labels = scales::percent, expand = expansion(mult = c(0, 0.2))) +
  scale_fill_brewer(palette = "Paired") +
  theme_minimal() +
  theme(
    panel.grid.major.x = element_blank(),
    plot.title = element_text(size = 14),
    plot.subtitle = element_text(color = "grey50", size = 10)
  ) +
  labs(
    title = "Arrest Rate for 2-Wheels Vehicles: Hot Season vs. Off Season",
    subtitle = "Comparing arrest intensity across seasons by year",
    x = NULL,
    y = "Arrest Rate (%)"
  )
```

What can we say?

-   Key finding:
-   Interpretation:
-   Comment:
-   Proposed Action:

```{r faceted-hot-season-vs-checkpoint-ratio}
# 1. Create the indicator, calculate rate and raw arrests
plot_data <- vehstops %>%
  filter(!is.na(year), !is.na(month), !is.na(veh_checkpoint)) %>%
  mutate(
    is_hot_season = ifelse(month %in% 4:10, "Hot Season", "Off Season"),
    is_hot_season = factor(is_hot_season, levels = c("Hot Season", "Off Season"))
  ) %>%
  group_by(year, is_hot_season) %>%
  summarise(
    total_stops = n(),
    arrests = sum(veh_checkpoint == 1),
    arrest_rate = arrests / total_stops,
    .groups = "drop"
  ) %>%
  mutate(
    # Label: Rate (Raw Arrest Count)
    label_text = paste0(scales::percent(arrest_rate, accuracy = 0.1), "\n(", comma(arrests), " checkpoints)")
  )

# 2. Build the faceted barplot
plot_data %>%
  ggplot(aes(x = is_hot_season, y = arrest_rate, fill = is_hot_season)) +
  geom_col(show.legend = FALSE) +
  # Facet by year
  facet_wrap(~year, nrow = 1) +
  # Add labels
  geom_text(aes(label = label_text), vjust = -0.5, size = 3, lineheight = 0.8) +
  # Use percent scale for the Y axis
  scale_y_continuous(labels = scales::percent, expand = expansion(mult = c(0, 0.2))) +
  scale_fill_brewer(palette = "Paired") +
  theme_minimal() +
  theme(
    panel.grid.major.x = element_blank(),
    plot.title = element_text(size = 14),
    plot.subtitle = element_text(color = "grey50", size = 10)
  ) +
  labs(
    title = "Vehicle Checkpoint Rate for 2-Wheels Vehicles: Hot Season vs. Off Season",
    subtitle = "Comparing arrest intensity across seasons by year",
    x = NULL,
    y = "Arrest Rate (%)"
  )
```

What can we say?

-   Key finding:
-   Interpretation:
-   Comment:
-   Proposed Action:

### Dow and Weekday

```{r dow-vs-arrested-ratio}
```

What can we say?

-   Key finding:
-   Interpretation:
-   Comment:
-   Proposed Action:

```{r dow-vs-checkpoint-ratio}
```

What can we say?

-   Key finding:
-   Interpretation:
-   Comment:
-   Proposed Action:

#### Faceted

```{r dow-faceted-vs-checkpoint-ratio}
```

What can we say?

-   Key finding:
-   Interpretation:
-   Comment:
-   Proposed Action:

### Weekend, Holiday and Event Day

```{r weekend-vs-arrested-ratio}
ratio_by_cat(
  vehstops, arrested, weekend,
  "Area", "Arrest Rate (%)", "Arrest Ratio by Area",
  style = "stack", legend_title = "Stop ended in arrest?",
  target_labels = c("No", "Yes"),
  subtitle = "For 2-Wheels Vehicles it shows Increased Arrest Ratio in Low Volume Areas",
  label_vjust = 0.5
)
```

What can we say?

-   Key finding:
-   Interpretation:
-   Comment:
-   Proposed Action:

```{r weekend-vs-checkpoint-ratio}
ratio_by_cat(
  vehstops, veh_checkpoint, weekend,
  "Area", "Arrest Rate (%)", "Arrest Ratio by Area",
  style = "stack", legend_title = "Stop ended in arrest?",
  target_labels = c("No", "Yes"),
  subtitle = "For 2-Wheels Vehicles it shows Increased Arrest Ratio in Low Volume Areas",
  label_vjust = 0.5
)
```

What can we say?

-   Key finding:
-   Interpretation:
-   Comment:
-   Proposed Action:

```{r holiday-vs-arrested-ratio}
ratio_by_cat(
  vehstops, arrested, holiday,
  "Area", "Arrest Rate (%)", "Arrest Ratio by Area",
  style = "stack", legend_title = "Stop ended in arrest?",
  target_labels = c("No", "Yes"),
  subtitle = "For 2-Wheels Vehicles it shows Increased Arrest Ratio in Low Volume Areas",
  label_vjust = 0.5
)
```

What can we say?

-   Key finding:
-   Interpretation:
-   Comment:
-   Proposed Action:

```{r holiday-vs-checkpoint-ratio}
ratio_by_cat(
  vehstops, veh_checkpoint, holiday,
  "Area", "Arrest Rate (%)", "Arrest Ratio by Area",
  style = "stack", legend_title = "Stop ended in arrest?",
  target_labels = c("No", "Yes"),
  subtitle = "For 2-Wheels Vehicles it shows Increased Arrest Ratio in Low Volume Areas",
  label_vjust = 0.5
)
```

What can we say?

-   Key finding:
-   Interpretation:
-   Comment:
-   Proposed Action:

```{r event-day-vs-arrested-ratio}
ratio_by_cat(
  vehstops, arrested, event_day,
  "Area", "Arrest Rate (%)", "Arrest Ratio by Area",
  style = "stack", legend_title = "Stop ended in arrest?",
  target_labels = c("No", "Yes"),
  subtitle = "For 2-Wheels Vehicles it shows Increased Arrest Ratio in Low Volume Areas",
  label_vjust = 0.5
)
```

What can we say?

-   Key finding:
-   Interpretation:
-   Comment:
-   Proposed Action:

```{r event-day-vs-checkpoint-ratio}
ratio_by_cat(
  vehstops, veh_checkpoint, event_day,
  "Area", "Arrest Rate (%)", "Arrest Ratio by Area",
  style = "stack", legend_title = "Stop ended in arrest?",
  target_labels = c("No", "Yes"),
  subtitle = "For 2-Wheels Vehicles it shows Increased Arrest Ratio in Low Volume Areas",
  label_vjust = 0.5
)
```

What can we say?

-   Key finding:
-   Interpretation:
-   Comment:
-   Proposed Action:

```{r drop-holiday-event-day}
vehstops <- vehstops %>%
  select(-holiday, -event_day)
```

### Hour, Working Hour and Time Bracket

```{r hour-vs-arrested-ratio}
```

What can we say?

-   Key finding:
-   Interpretation:
-   Comment:
-   Proposed Action:

```{r hour-vs-checkpoint-ratio}
```

What can we say?

-   Key finding:
-   Interpretation:
-   Comment:
-   Proposed Action:

```{r hour-vs-target}
vehstops |>
  filter(!is.na(arrested), !is.na(hour), !is.na(year)) |>
  group_by(year, hour) |>
  summarise(rate = mean(arrested), n = n(), .groups = "drop") |>
  ggplot(aes(hour, rate, group = 1)) +
  geom_line(linewidth = 0.6) +
  geom_point(size = 0.8, alpha = 0.7) +
  scale_x_continuous(breaks = seq(0, 23, by = 1)) +
  scale_y_continuous(labels = percent_format()) +
  labs(title = "Escalation rate by Hour", x = "Hour", y = "Rate") +
  facet_wrap(~ year, ncol = 1, scales = "free_x") +
  theme_minimal()
```

Let us look at our grouping now

```{r time-bracket-vs-arrested-ratio}
ratio_by_cat(
  vehstops, arrested, time_bracket,
  "Time bracket", "Rate", "Arrest ratio by time bracket",
  style = "stack"
)
```

What can we say?

-   Key finding:
-   Interpretation:
-   Comment:
-   Proposed Action:

```{r time-bracket-vs-checkpoint-ratio}
ratio_by_cat(
  vehstops, veh_checkpoint, time_bracket,
  "Time bracket", "Rate", "Arrest ratio by time bracket",
  style = "stack"
)
```

What can we say?

-   Key finding:
-   Interpretation:
-   Comment:
-   Proposed Action:

## Motorist related features

### Reported Age

```{r age-by-target}
vehstops %>%
  filter(!is.na(age), between(age, 16, 100)) %>%
  mutate(bin = ntile(age, 20)) %>%
  group_by(bin) %>%
  summarize(
    age_mid = median(age, na.rm = TRUE),
    rate = mean(arrested == 1, na.rm = TRUE),
    n = n(), .groups = "drop"
  ) %>%
  mutate(
    se = sqrt(rate * (1 - rate) / n),
    lo = pmax(rate - 1.96 * se, 0),
    hi = pmin(rate + 1.96 * se, 1)
  ) %>%
  ggplot(aes(age_mid, rate)) +
  geom_point(size = 1.5) +
  geom_line(linewidth = 0.6) +
  geom_ribbon(aes(ymin = lo, ymax = hi), alpha = 0.15) +
  scale_y_continuous(labels = percent) +
  labs(x = "Age (bin median)", y = "Event rate", title = "Escalation rate vs age (deciles)")
```

```{r age-bracket-vs-target}
ratio_by_cat(
  vehstops, arrested, age_bracket,
  "Reported Age Bracket", "Share of stops", "Escalation ratio by reported age bracket",
  style = "stack"
)
```

### Reported Sex

```{r sex-vs-target}
ratio_by_cat(
  vehstops, arrested, sex,
  "Reported sex", "Share of stops", "Escalation ratio by reported sex",
  style = "stack"
)
```

### Reported Race

```{r race-vs-target}
ratio_by_cat(
  vehstops, arrested, race,
  "Reported race", "Share of stops", "Escalation ratio by reported race",
  style = "stack"
)
```

## Vehicle related features vs Target

------------------------------------------------------------------------

# Multivariate EDA

Since multivariate analysis is very complex, and we just want to get rid of some extra features, we are going to look at some of the percentages, with respect to escalation, of multiple variables (ex: sex x race)

## Multicollinearity (binary block)

## Two-way event-rate heatmaps (interactions)

```{r two-way-heatmaps}
boro_order <- vehstops %>%
  group_by(boro) %>%
  summarize(rate = mean(arrested == 1, na.rm = TRUE), .groups = "drop") %>%
  arrange(rate) %>%
  pull(boro)

vehstops %>%
  filter(!is.na(boro)) %>%
  group_by(boro, hour) %>%
  summarize(rate = mean(arrested == 1, na.rm = TRUE), n = n(), .groups = "drop") %>%
  mutate(boro = factor(boro, levels = boro_order)) %>%
  ggplot(aes(hour, boro, fill = rate)) +
  geom_tile(width = 1, height = 1) +
  scale_fill_viridis_c(labels = scales::percent) +
  labs(x = "Hour", y = "Borough", fill = "Rate",
       title = "Escalation rate by borough × hour")

# 2) Sex × Race
vehstops %>%
  filter(!is.na(sex), !is.na(race)) %>%
  group_by(sex, race) %>%
  summarize(rate = mean(arrested == 1, na.rm = TRUE), n = n(), .groups = "drop") %>%
  mutate(race = fct_lump_n(race, n = 8, other_level = "Other")) %>%
  ggplot(aes(race, sex, fill = rate)) +
  geom_tile(width = 1, height = 1) +
  scale_fill_viridis_c(labels = scales::percent) +
  labs(x = "Race (lumped)", y = "Sex", fill = "Rate",
       title = "Escalation rate across sex × race")
```

### Space × Time interaction (gridded)

```{r space-time}
vehstops %>%
  filter(!is.na(latitude), !is.na(longitude),
         between(latitude, 40.48, 40.92), between(longitude, -74.27, -73.68)) %>%
  mutate(hour_bin = cut(hour, breaks = c(-1,5,11,17,23), labels = c("Night","Morning","Afternoon","Evening"))) %>%
  group_by(hour_bin) %>%
  sample_n(size = min(n(), 200000), replace = FALSE) %>%  # cap points per facet for speed
  ggplot(aes(longitude, latitude)) +
  stat_summary_2d(aes(z = arrested), fun = ~ mean(.x == 1, na.rm = TRUE), bins = 50) +
  facet_wrap(~ hour_bin) +
  coord_equal() +
  scale_fill_viridis_c(labels = percent) +
  labs(x = "Longitude", y = "Latitude", fill = "Rate",
       title = "Escalation rate across space by time-of-day")
```

## Enforcement and Escalation features vs Target

## Command Unit, Geospatial and Area related Features vs Target

Per command unit amount of stops during time (per year, per month, per day)

Per command unit escalation rate during time (per year, per month, per day)

### Area

```{r}
ratio_by_cat_split(
  DF = vehstops,
  TARGET = arrested,
  CAT    = area,
  SPLIT  = weekend,
  x_title = "Area",
  y_title = "Share of stops",
  title   = "Escalation ratio by area (split by weekend)",
  target_labels = c("Not escalated","Escalated"),
  split_labels  = c("Weekday","Weekend")
)

```

```{r area-x-hour}
area_order <- vehstops %>%
  group_by(area) %>%
  summarize(rate = mean(arrested == 1, na.rm = TRUE), .groups = "drop") %>%
  arrange(rate) %>%
  pull(area)

vehstops %>%
  filter(!is.na(area)) %>%
  group_by(area, hour) %>%
  summarize(rate = mean(arrested == 1, na.rm = TRUE), n = n(), .groups = "drop") %>%
  mutate(area = factor(area, levels = area_order)) %>%
  ggplot(aes(hour, area, fill = rate)) +
  geom_tile(width = 1, height = 1) +
  scale_fill_viridis_c(labels = scales::percent) +
  labs(x = "Hour", y = "Area", fill = "Rate",
       title = "Escalation rate by area × hour")
```

```{r area-x-sex}
ratio_by_cat_split(
  DF = vehstops,
  TARGET = arrested,
  CAT    = area,
  SPLIT  = sex,
  x_title = "Area",
  y_title = "Share of stops",
  title   = "Escalation ratio by area (split by reported sex)",
  target_labels = c("Not escalated","Escalated"),
  split_labels  = c("No","Yes")
)
```

```{r}
# heat_target_catcat(
#   vehstops, Y = boro, X = time_bracket,
#   X_TITLE = "Time bracket", Y_TITLE = "Borough",
#   TITLE = "Escalation rate by borough × time bracket",
#   min_n = 30, order_y_by = "rate", order_x_by = "freq", alpha_by_n = TRUE
# )
```

```{r}
area_order <- vehstops %>%
  group_by(area) %>%
  summarize(rate = mean(arrested == 1, na.rm = TRUE), .groups = "drop") %>%
  arrange(rate) %>%
  pull(area)

vehstops %>%
  filter(!is.na(area)) %>%
  group_by(area, month) %>%
  summarize(rate = mean(arrested == 1, na.rm = TRUE), n = n(), .groups = "drop") %>%
  mutate(area = factor(area, levels = area_order)) %>%
  ggplot(aes(month, area, fill = rate)) +
  geom_tile(width = 1, height = 1) +
  scale_fill_viridis_c(labels = scales::percent) +
  labs(x = "Month", y = "Area", fill = "Rate",
       title = "Escalation rate by area × month")
```

Per area amount of stops during time (per year, per month, per day)

Per area escalation rate during time (per year, per month, per day)

```{r area-by-weekend-vs-target}
p1 <- ratio_by_cat_split(
  DF = vehstops,
  TARGET = arrested,
  CAT    = area,
  SPLIT  = weekend,
  x_title = "Time bracket",
  y_title = "Share of stops",
  title   = "Escalation ratio by area (split by weekend)",
  target_labels = c("Not escalated","Escalated"),
  split_labels  = c("Weekday","Weekend")
)

p1
```

### Borough

```{r boro-x-year-escalation}
boro_order <- vehstops %>%
  group_by(boro) %>%
  summarize(rate = mean(arrested == 1, na.rm = TRUE), .groups = "drop") %>%
  arrange(rate) %>%
  pull(boro)

vehstops %>%
  filter(!is.na(boro)) %>%
  group_by(boro, year) %>%
  summarize(rate = mean(arrested == 1, na.rm = TRUE), n = n(), .groups = "drop") %>%
  mutate(boro = factor(boro, levels = boro_order)) %>%
  ggplot(aes(year, boro, fill = rate)) +
  geom_tile(width = 1, height = 1) +
  scale_fill_viridis_c(labels = scales::percent) +
  labs(x = "Year", y = "Borough", fill = "Rate",
       title = "Escalation rate by borough × year")


```

```{r boro-x-month-escalation}
boro_order <- vehstops %>%
  group_by(boro) %>%
  summarize(rate = mean(arrested == 1, na.rm = TRUE), .groups = "drop") %>%
  arrange(rate) %>%
  pull(boro)

vehstops %>%
  filter(!is.na(boro)) %>%
  group_by(boro, month) %>%
  summarize(rate = mean(arrested == 1, na.rm = TRUE), n = n(), .groups = "drop") %>%
  mutate(boro = factor(boro, levels = boro_order)) %>%
  ggplot(aes(month, boro, fill = rate)) +
  geom_tile(width = 1, height = 1) +
  scale_fill_viridis_c(labels = scales::percent) +
  labs(x = "Month", y = "Borough", fill = "Rate",
       title = "Escalation rate by borough × month")
```

```{r boro-x-hour-escalation}
boro_order <- vehstops %>%
  group_by(boro) %>%
  summarize(rate = mean(arrested == 1, na.rm = TRUE), .groups = "drop") %>%
  arrange(rate) %>%
  pull(boro)

vehstops %>%
  filter(!is.na(boro)) %>%
  group_by(boro, hour) %>%
  summarize(rate = mean(arrested == 1, na.rm = TRUE), n = n(), .groups = "drop") %>%
  mutate(boro = factor(boro, levels = boro_order)) %>%
  ggplot(aes(hour, boro, fill = rate)) +
  geom_tile(width = 1, height = 1) +
  scale_fill_viridis_c(labels = scales::percent) +
  labs(x = "Hour", y = "Borough", fill = "Rate",
       title = "Escalation rate by borough × hour")
```

### Command Unit

Per command unit amount of stops during time (per year, per month, per day)

Per command unit escalation rate during time (per year, per month, per day)

```{r}
# # Borough × Time bracket (both categorical), hide very sparse cells
# p1 <- heat_target_catcat(
#   vehstops, Y = boro, X = time_bracket,
#   X_TITLE = "Time bracket", Y_TITLE = "Borough",
#   TITLE = "Escalation rate by borough × time bracket",
#   min_n = 30, order_y_by = "rate", order_x_by = "freq", alpha_by_n = TRUE
# )
# 
# # Cleaned vehicle category × Weekend (0/1 already labeled)
# p2 <- heat_target_catcat(
#   vehstops, Y = race, X = sex,
#   "Sex", "Race",
#   "Rate by race × sex",
#   min_n = 50, order_y_by = "freq", order_x_by = "freq"
# )
# 
# 
# p1
# p2

```

```{r}
# heat_target_catcat(
#   vehstops, Y = race, X = sex,
#   "Sex", "Race",
#   "Rate by race × sex",
#   min_n = 50, order_y_by = "freq", order_x_by = "freq"
# )
```

Unknown and Unknown matter somehow!!!

```{r}
# # Borough × hour (hide cells with < 50 stops)
# p_boro_hour <- heat_escalation_yx(
#   DF = vehstops, Y = boro, X = hour,
#   X_TITLE = "Hour of day", Y_TITLE = "Borough",
#   TITLE = "Escalation rate by borough × hour",
#   min_n = 50
# )
# 
# # Area × day-of-week (X is numeric 1–7)
# p_area_dow <- heat_escalation_yx(
#   vehstops, Y = area, X = dow,
#   "Day of week (1–7)", "Area",
#   "Escalation rate by area × DOW", min_n = 25
# )
# 
# # If X is not integer-spaced (e.g., minute) treat as factor to make equal-width tiles
# p_boro_min <- heat_escalation_yx(
#   vehstops, Y = boro, X = minute,
#   "Minute", "Borough", "Rate by borough × minute",
#   x_as_factor = TRUE
# )

```

```{r}
#p_boro_hour
#p_area_dow
#p_boro_min
```

## Time related features

### Hour

#### Grouped by Month

```{r month-lines}
# TODO: use? Move to multivariate
vehstops %>%
  filter(year %in% c(2023, 2024, 2025)) %>%
  group_by(year, month) %>%
  summarize(stops = n(), .groups = "drop") %>%
  ggplot(aes(month, stops, group = factor(year), linetype = factor(year))) +
  geom_line(linewidth = 0.6) + geom_point(size = 1.5) +
  scale_x_continuous(breaks = 1:12) +
  labs(x = "Month", y = "Stops", linetype = "Year",
       title = "Monthly stops: 2023–2025")
```

```{r hour-vs-stops-by-month}
vehstops %>%
  filter(month %in% c(1,2,3,4,5,6,7,8,9,10,11,12)) %>%
  group_by(month, hour) %>%
  summarize(stops = n(), .groups = "drop") %>%
  ggplot(aes(hour, stops, group = factor(month), linetype = factor(month))) +
  geom_line(linewidth = 0.6) + geom_point(size = 1.5) +
  scale_x_continuous(breaks = 0:23) +
  labs(x = "Hour", y = "Stops", linetype = "Month",
       title = "Count of Hourly Stops by Month")
```

```{r hour-vs-dow-by-month}
vehstops %>%
  group_by(dow, hour) %>%
  summarize(stops = n(), .groups = "drop") %>%
  ggplot(aes(hour, stops, group = factor(dow), linetype = factor(dow))) +
  geom_line(linewidth = 0.6) + geom_point(size = 1.5) +
  scale_x_continuous(breaks = 0:23) +
  labs(x = "Hour", y = "Stops", linetype = "Day of the Week",
       title = "Count of Hourly Stops by Day of the Week")
```

```{r hour-vs-target-by-year}
vehstops %>%
  mutate(date = ymd(paste(year, month, day, sep = "-"))) %>%
  group_by(month, hour) %>%
  summarize(rate = mean(arrested == 1, na.rm = TRUE), n = n(), .groups = "drop") %>%
  ggplot(aes(hour, rate, group = factor(month), linetype = factor(month))) +
  geom_line(linewidth = 0.6) + geom_point(aes(size = n)) +
  scale_y_continuous(labels = percent) +
  scale_x_continuous(breaks = 0:23) +
  labs(x = "Hour", y = "Event rate", linetype = "Hour", size = "Stops count",
       title = "Hourly Escalation Rate by Month")
```

```{r dow-hour-heatmap}
vehstops %>%
  filter(!is.na(dow)) %>%
  group_by(dow, hour) %>%
  summarize(rate = mean(arrested == 1, na.rm = TRUE), n = n(), .groups = "drop") %>%
  ggplot(aes(hour, dow, fill = rate)) +
  geom_tile(width = 1, height = 1) +
  scale_fill_viridis_c(labels = percent) +
  scale_x_continuous(breaks = seq(0, 23, 1)) +
  scale_y_continuous(breaks = seq(0, 7, 1)) +
  labs(x = "Hour", y = "Day of week", fill = "Event rate",
       title = "Escalation rate by day and hour")
```

#### Grouped by Year

```{r hour-vs-stops-by-year}
vehstops %>%
  filter(year %in% c(2023, 2024, 2025)) %>%
  group_by(year, hour) %>%
  summarize(stops = n(), .groups = "drop") %>%
  ggplot(aes(hour, stops, group = factor(year), linetype = factor(year))) +
  geom_line(linewidth = 0.6) + geom_point(size = 1.5) +
  scale_x_continuous(breaks = 0:23) +
  labs(x = "Hour", y = "Stops", linetype = "Year",
       title = "Count of Hourly stops by Year")
```

```{r hour-vs-any-escalation-by-year}
vehstops %>%
  mutate(date = ymd(paste(year, month, day, sep = "-"))) %>%
  group_by(year, hour) %>%
  summarize(rate = mean(arrested == 1, na.rm = TRUE), n = n(), .groups = "drop") %>%
  ggplot(aes(hour, rate, group = factor(year), linetype = factor(year))) +
  geom_line(linewidth = 0.6) + geom_point(aes(size = n)) +
  scale_y_continuous(labels = percent) +
  scale_x_continuous(breaks = 0:23) +
  labs(x = "Hour", y = "Event rate", linetype = "Year", size = "Stops count",
       title = "Hourly Escalation Rate by Year")
```

#### Grouped by Day

```{r}
vehstops %>%
  filter(!is.na(dow)) %>%
  group_by(dow, hour) %>%
  summarize(rate = mean(arrested == 1, na.rm = TRUE), n = n(), .groups = "drop") %>%
  ggplot(aes(hour, dow, fill = rate)) +
  geom_tile(width = 1, height = 1) +
  scale_fill_viridis_c(labels = percent) +
  scale_x_continuous(breaks = seq(0, 23, 1)) +
  scale_y_continuous(breaks = seq(0, 7, 1)) +
  labs(x = "Hour", y = "Day of week", fill = "Event rate",
       title = "Escalation rate by day and hour")
```

## Motorist related features

### Reported Age

```{r}
ratio_by_cat_split(
  DF = vehstops,
  TARGET = arrested,
  CAT    = age_bracket,
  SPLIT  = weekend,
  x_title = "Reported Age bracket",
  y_title = "Share of stops",
  title   = "Escalation ratio by reported age (split by weekend)",
  target_labels = c("Not escalated","Escalated"),
  split_labels  = c("Weekday","Weekend")
)
```

### Reported Sex

### Reported Race

------------------------------------------------------------------------
