---
title: "Predicting Excessive Enforcement Patterns in Police Behavior After Vehicles Stops in New York"
subtitle: "EDA and Feature Engineering - 4 Wheels Vehicles"
author: "Gianmaria Pizzo"
output:
  html_document:
    fig_caption: yes
    theme: flatly
    highlight: pygments
    code_folding: show
    keep_md: no
    toc: yes
    toc_depth: 4
    number_sections: no
    toc_float:
      smooth_scroll: no
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

These `.Rmd` files represent the project submission for the course [Statistical Inference and Learning](https://www.unive.it/data/course/521944) by Professor [Cristiano Varin](https://www.unive.it/data/people/5592728) at [Ca' Foscari University of Venice](https://www.unive.it/) for the year 2025/2026.

# Problem Statement

## Goal of this rmarkdown file

This notebook serves as the primary Exploratory Data Analysis (EDA) and (final) Feature Engineering phase for the project. In particular, we focus solely on **4-Wheeled Vehicles**.

The goal is to identify the features that characterize **arrests**.

We approach the dataset through three methodological phases to ensure a robust base for modeling:

1.  **Univariate Analysis**. We begin by analyzing the distributions of individual predictors to understand whether the data shows quality issues, outliers, rare categories, etc...

2.  **Bivariate Analysis**. We evaluate the relationship between potential predictors and our binary target (`arrested`). Given the categorical nature of our data, we will mainly uses **Cramer’s V** as a starting point to measure of effect size. This stage should also point out if predictors have some kind of relation.

Under these premises, we will plot many aspects of the data. Therefore, the way we decided to guide the reader through the Exploratory Data Analysis, is based on four key points that we try to include for any plot:

- **Key finding**: This paragraph describes the main takeaway of the means of visualization employed. This should be a concise description of what we can gather from the plots.
- **Interpretation**: Here we deal with an objective interpretation of our finding, such as understanding how our EDA is affected by it.
- **Comment**: After a formal description, we deal with a more informal discussion that should highlight the qualitative aspects of the analysis. This paragraph is a fusion of previous considerations that we want the reader to be aware of, and a space for us to reason.
- **Proposed Action**: Finally, we switch to a more pragmatic discussion to understand "what to do next", on the basis of what was previously presented. Without claim of completeness, we will name some best-practices, guidelines or ideas about what is best for our goal. We may also limit ourselves to inform the reader that we decided to take no action at all or post-pone it after further investigation.

Note that we do not always include all four points.

## Important Note

We will, sometimes, cut short of the commentary when we make similar discovery with respect to the 2-wheels analysis. This is needed not to repeat ourselves if we go down the same conclusions. In this case, we will omit to elaborate.

------------------------------------------------------------------------

# Environment

The first to do, is to build a reproducible environment for our project.

```{r}
requirements=c( 
  "remotes", "summarytools", "forcats", "scales", "purrr",
  "stringr", "plyr", "dbplyr", "dplyr", "dtplyr", "tidyr", "tidyselect",
  "timeDate", "timechange", "tzdb", "hms", "readr", "tigris", "lsr", "corrplot",
  "lubridate", "prettyunits", "RColorBrewer", "viridis", "rcartocolor", "ggbeeswarm", "emmeans",
  "sf", "plotly", "ggplot2", "ggmap", "ggdensity", "UpSetR", "patchwork"
)
```

```{r install-requirements, include=FALSE}
for (req in requirements){
  if (!require(req, character.only = TRUE)){
      install.packages(req)
  }
}
```

Note that some of the packages may not be used, but to avoid any error related to dependencies, we are going to keep all of these libraries. Now, the loading of all packages should be immediate.

```{r load-requirements, include=FALSE}
for (req in requirements){
  if (!require(req, character.only = TRUE)){
      library(req, character.only = TRUE)
  }
}
```

```{r remove-vars, include=FALSE}
rm(requirements, req)
```

## Plot Functions

Here we add the plot functions to make some kinds of visualizations reproducible.

### Univariate Plot Functions

```{r univariate-plot-helpers}
# ==============================================================================
# 1. Plots for Univariate Analysis
# ==============================================================================

#' @title Plot Binary Bar Chart
#' 
#' @description Creates a simple bar chart counting the occurrences of a binary feature, 
#' with labels showing both the count and the percentage.
#' 
#' @param DF The input dataframe.
#' @param FEATURE The binary column to plot (unquoted).
#' @param OPTION_0 The text label for the "0" or "FALSE" category.
#' @param OPTION_1 The text label for the "1" or "TRUE" category.
#' @param X_LABEL Label for the X axis.
#' @param Y_LABEL Label for the Y axis.
#' @param TITLE Main title of the plot.
#' @param pad Numeric padding for the top of the Y axis to fit text labels.
#' @param text_vjust Vertical adjustment for the text labels on top of bars.
#' 
#' @return A ggplot2 object.
plot_binary_bar <- function(DF, FEATURE, OPTION_0, OPTION_1, X_LABEL, Y_LABEL, TITLE, pad = 0.15, text_vjust = -0.3) {
  # Standardize the binary feature and calculate percentages
  df_sum <- DF %>%
    mutate(
      .raw  = {{ FEATURE }},
      .feat = dplyr::case_when(
        as.character(.raw) %in% c("1","TRUE","true")  ~ 1L,
        as.character(.raw) %in% c("0","FALSE","false") ~ 0L,
        TRUE ~ NA_integer_
      )
    ) %>%
    filter(!is.na(.feat)) %>%
    count(.feat, name = "n") %>%
    mutate(
      .lab = factor(.feat, levels = c(0, 1), labels = c(OPTION_0, OPTION_1)),
      # Calculate percentage and create combined label
      pct = n / sum(n),
      label_text = paste0(scales::comma(n), "\n(", scales::percent(pct, accuracy = 0.1), ")")
    )
  
  # Get the maximum bar height
  ymax <- max(df_sum$n, na.rm = TRUE)
  
  # Build the ggplot
  ggplot(df_sum, aes(x = .lab, y = n, fill = .lab)) +
    geom_col(width = 0.8, show.legend = FALSE) +
    # Updated to use label_text
    geom_text(aes(label = label_text), vjust = text_vjust, lineheight = 0.8) +
    # Expand the Y axis slightly for the taller multi-line labels
    scale_y_continuous(limits = c(0, ymax * (1 + pad)), expand = expansion(mult = c(0, 0.02))) +
    coord_cartesian(clip = "off") +
    theme_minimal() +
    theme(panel.grid.major.x = element_blank(), plot.margin = margin(t = 20, r = 5.5, b = 5.5, l = 5.5)) +
    scale_fill_viridis_d(option = "mako", begin = 0.3, end = 0.8) +
    labs(x = X_LABEL, y = Y_LABEL, title = TITLE)
}


#' @title Plot Univariate Bar Chart (Categorical)
#' 
#' @description Creates a bar chart counting the occurrences of a categorical feature with 2+ classes.
#' Allows dynamic ordering (by frequency or original/alphabetical order).
#' 
#' @param DF The input dataframe.
#' @param FEATURE The categorical column to analyze (unquoted).
#' @param X_LABEL Label for the X axis.
#' @param Y_LABEL Label for the Y axis (default is "Count of Stops").
#' @param TITLE Main title of the plot.
#' @param SUBTITLE Subtitle of the plot.
#' @param top_n Optional integer; if set, keeps only the top N most frequent categories.
#' @param order_by Method to order the X axis: "freq" (descending count) or "none" (alphabetical/original).
#' @param pad Numeric padding for the top of the Y axis to fit text labels.
#' @param text_vjust Vertical adjustment for the text labels on top of bars.
#' 
#' @return A ggplot2 object.
plot_univariate_cat <- function(DF, FEATURE, X_LABEL, Y_LABEL = "Count of Stops", TITLE, SUBTITLE, top_n = NULL, order_by = c("freq", "none"), pad = 0.12, text_vjust = -0.35) {
  
  order_by <- match.arg(order_by)
  
  # Compute counts
  df_sum <- DF %>%
    filter(!is.na({{ FEATURE }})) %>%
    count({{ FEATURE }}, name = "n")
  
  # Filter top n results
  if (!is.null(top_n)) {
    df_sum <- df_sum %>% slice_max(order_by = n, n = top_n)
  }
  
  # X axis order (either frequency or std)
  cat_levels <- switch(
    order_by,
    freq = df_sum %>% arrange(desc(n)) %>% pull({{ FEATURE }}),
    none = sort(unique(df_sum %>% pull({{ FEATURE }})))
  )
  
  # Apply sorting
  df_sum <- df_sum %>%
    mutate(.feat_f = factor(as.character({{ FEATURE }}), levels = as.character(cat_levels)))
  
  # Max height for dynamic padding
  ymax <- max(df_sum$n, na.rm = TRUE)

  # Construct plot
  ggplot(df_sum, aes(x = .feat_f, y = n, fill = .feat_f)) +
    geom_col(width = 0.8, show.legend = FALSE) +
    geom_text(aes(label = scales::comma(n)), vjust = text_vjust, size = 3) +
    scale_y_continuous(limits = c(0, ymax * (1 + pad)),
                       expand = expansion(mult = c(0, 0.02))) +
    coord_cartesian(clip = "off") +
    theme_minimal() +
    theme(
      plot.title = element_text(size = 14),
      plot.subtitle = element_text(size = 10, color = "grey50"),
      axis.text.x = element_text(angle = 45, hjust = 1),
      panel.grid.major.x = element_blank(),
      plot.margin = margin(t = 10, r = 5.5, b = 5.5, l = 5.5)) +
    scale_fill_viridis_d(option = "mako", begin = 0.3, end = 0.8) +
    labs(x = X_LABEL, y = Y_LABEL, title = TITLE, subtitle = SUBTITLE)
}
```

### Bivariate Plot Functions

```{r biviariate-plot-helpers}
# ==============================================================================
# 2. Plots for Bivariate Analysis
# ==============================================================================

#' @title Calculate Cramer's V
#' 
#' @description Calculates the Cramer's V measure of association between two 
#' categorical vectors. Returns a value between 0 (no association) and 1 (perfect association).
#' 
#' @param x First categorical vector.
#' @param y Second categorical vector.
#' 
#' @return A numeric value representing Cramer's V.
cramer_v <- function(x, y) {
  # Contingency Table
  tbl <- table(x, y)
  
  # Execute Chi2 Test
  chi2 <- suppressWarnings(chisq.test(tbl, correct = FALSE)$statistic)
  
  # Compute dim and min value
  n <- sum(tbl)
  min_dim <- min(nrow(tbl) - 1, ncol(tbl) - 1)
  
  # Avoid math error for 1 categ
  if(min_dim == 0) return(0) 
  
  # Cramer's formula
  v <- sqrt(chi2 / (n * min_dim))
  return(as.numeric(v))
}

# ==============================================================================
# Categorical "Correlation" Heatmap - Association Matrix based on Cramer's V
# ==============================================================================

#' @title Plot Categorical Correlation Matrix (Cramer's V)
#' 
#' @description Creates a correlation-like heatmap for multiple categorical variables
#' using Cramer's V. Answers "How much do these categorical features overlap with each other?"
#' 
#' @param DF The input dataframe.
#' @param FEATURES A character vector containing the names of the categorical columns to compare.
#' @param TITLE Main title of the plot.
#' 
#' @return A ggplot2 object.
plot_categorical_correlation <- function(DF, FEATURES, TITLE = "Association Matrix for Categorical Predictors (Cramer's V)", text_size = 3) {
  
  # Select columns and drop NAs
  df_sub <- DF %>% 
    select(all_of(FEATURES)) %>%
    drop_na()
  
  n_vars <- ncol(df_sub)
  
  # Initialize matrix and set main diag to 1s
  mat <- matrix(1, nrow = n_vars, ncol = n_vars)
  rownames(mat) <- colnames(df_sub)
  colnames(mat) <- colnames(df_sub)

  # Compute Cramer's V for each possible pair
  for(i in 1:(n_vars - 1)) {
    for(j in (i + 1):n_vars) {
      v <- cramer_v(df_sub[[i]], df_sub[[j]])
      # Symmetric 
      mat[i, j] <- v
      mat[j, i] <- v
    }
  }
  
  # Make matrix compatible with ggplot2
  mat_long <- as.data.frame(as.table(mat))
  colnames(mat_long) <- c("Var1", "Var2", "CramerV")
  
  # Create heatmap
  ggplot(mat_long, aes(x = Var1, y = Var2, fill = CramerV)) +
    geom_tile(color = "white", width = 1, height = 1) +
    # Rounding
    geom_text(aes(label = round(CramerV, 2)), size = text_size) +
    # Setting colors
    scale_fill_gradient(low = "whitesmoke", high = "steelblue", limits = c(0, 1), name = "Cramer's V") +
    labs(title = TITLE, x = "", y = "") +
    theme_minimal() +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1, size = 10),
      axis.text.y = element_text(size = 10),
      panel.grid = element_blank()
    )
}

# ==============================================================================
# Plotting of Targets' Rate Against Categorical Predictors
# ==============================================================================

#' @title Plot Target Ratio by Category
#' 
#' @description Creates a bivariate bar chart showing the relationship between a binary 
#' TARGET and a single categorical variable (CAT). It supports two visual styles: 
#' a 100% stacked bar chart ("stack") or a simple bar chart showing the rate of the target ("rate").
#' 
#' @param DF The input dataframe.
#' @param TARGET The binary column to analyze (unquoted). It will be standardized to 0/1.
#' @param CAT The categorical column for the X axis (unquoted).
#' @param x_title Label for the X axis.
#' @param y_title Label for the Y axis.
#' @param title Main title of the plot.
#' @param target_labels A character vector of length 2 defining the labels for 0 and 1 (e.g., c("Not escalated", "Escalated")).
#' @param style The visual style of the plot: "stack" (100% stacked bars) or "rate" (single bar for target = 1).
#' @param order_by Method to order the X axis categories: by "rate" (ascending target rate), "freq" (ascending frequency), or "none" (alphabetical/original).
#' @param percent_accuracy Accuracy for the percent labels (e.g., 0.1 for one decimal place).
#' @param colors A character vector of length 2 defining the hex color codes for the 0 and 1 categories respectively.
#' 
#' @return A ggplot2 object.
ratio_by_cat <- function(
    DF, TARGET, CAT, x_title, y_title, title, 
    target_labels = c("Not escalated", "Escalated"), 
    style = c("stack", "rate"), order_by = c("rate", "freq", "none"), 
    percent_accuracy = 0.1, palette = "Paired",
    colors = c("grey80", "steelblue"), legend_title = "Legend",
    subtitle = NULL, bar_width = 0.7, aspect_ratio = 1.0, label_vjust = 0.5,
    text_size = 2
) {
  
  # Validate arguments
  style    <- match.arg(style)
  order_by <- match.arg(order_by)
  
  # Standardize target and extract the category
  dat <- DF %>%
    mutate(
      .t_raw = {{ TARGET }},
      .t = case_when(
        as.character(.t_raw) %in% c("1","TRUE","true") ~ 1L,
        as.character(.t_raw) %in% c("0","FALSE","false") ~ 0L,
        TRUE ~ NA_integer_
      ),
      .cat = {{ CAT }}
    ) %>%
    filter(!is.na(.t), !is.na(.cat))
  
  # Compute rates and frequencies
  rates <- dat %>% 
    group_by(.cat) %>%
    summarise(total = n(), rate = mean(.t == 1), .groups = "drop")
  
  # Determine ordering
  cat_levels <- switch(
    order_by,
    rate = rates %>% arrange(rate) %>% pull(.cat),
    freq = rates %>% arrange(total) %>% pull(.cat),
    none = unique(dat$.cat)
  )
  
  # Build the plot based on style
  if (style == "stack") {
    stack <- dat %>%
      count(.cat, .t, name = "n") %>%
      group_by(.cat) %>%
      mutate(ratio = n/sum(n)) %>%
      ungroup() %>%
      mutate(
        .cat_f = factor(as.character(.cat), levels = as.character(cat_levels)),
        .t_f = factor(.t, levels = c(0,1), labels = target_labels)
      )
    
    ggplot(stack, aes(x = .cat_f, y = ratio, fill = .t_f)) +
      geom_col(position = "fill", width = bar_width) +
      geom_text(
        aes(label = scales::percent(ratio, accuracy = percent_accuracy)),
        position = position_fill(vjust = label_vjust), 
        size = text_size
      ) +
      scale_y_continuous(labels = scales::percent) +
      scale_fill_manual(values = setNames(colors, target_labels)) +
      labs(x = x_title, y = y_title, fill = legend_title, title = title, subtitle = subtitle) +
      theme_minimal() +
      theme(
        axis.text.x = element_text(angle = 45, hjust = 1),
        panel.grid = element_blank(),
        plot.subtitle = element_text(color = "grey50"),
        aspect.ratio = aspect_ratio
      )
      
  } else {
    # Rate style
    rates_data <- rates %>%
      mutate(.cat_f = factor(as.character(.cat), levels = as.character(cat_levels)))
    
    ggplot(rates_data, aes(x = .cat_f, y = rate)) +
      geom_col(width = 0.8, fill = colors[2]) +
      geom_text(
        aes(label = scales::percent(rate, accuracy = percent_accuracy)),
        vjust = -0.35, size = text_size
      ) +
      scale_y_continuous(labels = scales::percent, expand = expansion(mult = c(0, 0.12))) +
      coord_cartesian(clip = "off") +
      labs(x = x_title, y = y_title, title = title, subtitle = subtitle) +
      theme_minimal() +
      theme(
        axis.text.x = element_text(angle = 45, hjust = 1),
        panel.grid = element_blank(),
        plot.subtitle = element_text(color = "grey50")
      )
  }
}
```

------------------------------------------------------------------------

Let us load the data now, and remove the extra feature `veh_category` because we are **exclusively analyzing 4 wheels vehicles now**.

```{r load-data-drop-veh-categ}
vehstops = read.csv("data/vehiclestops_4_wheels.csv")

vehstops <- vehstops %>%
  select(-veh_categ)
```

So now we have the following dataset, that we present with the usual summary

```{r summary-data}
print(dfSummary(vehstops), method="render")
```

This preview is already interesting.

------------------------------------------------------------------------

# Univariate EDA

## Target

```{r enforcement-dist}
p1 <- plot_binary_bar(
  vehstops, arrested, 
  "No", "Yes",
  "Was the driver arrested?", "Count of Stops (Percentage %)", ""
)

row1 <- p1

row1 + 
plot_annotation(title = "Distribution of Arrests for Cars Stopped in NYC") &
theme(legend.position = "bottom") 
```

```{r, include=FALSE}
rm(p1, row1)
```

What can we say?

What can we say?

- **Key finding**: The dataset for 4-wheels vehicles **exhibits extreme class imbalance** that is even more pronounced than what we observed in the 2-wheels analysis. Arrests are highly sparse, representing only approximately $3.3\%$ of the outcomes ($70,600$ "Yes" out of $2,157,445$ total records) compared to the $5.6\%$ rate found in 2-wheeled vehicle stops.
- **Interpretation**: The target variable constitutes a rare event when we contemplate the stops of 4-wheels vehicles. While the overall volume of stops is drastically higher for 4-wheels vehicles, an arrest remains an infrequent occurrence with respect to the total volume of police stops.
- **Comment**: The high class disparity *makes accuracy a limiting metric* for our models' evaluation, here too.
- **Proposed Action**: Same as the 2-wheels analysis.

## Geospatial predictors

### Area

```{r count-stops-by-area}
plot_univariate_cat(
  DF = vehstops, 
  FEATURE = area, 
  X_LABEL = "Area", 
  TITLE = "Distribution of Car Stops by Area in NYC",
  SUBTITLE = "Brooklyn is where the most amount of vehicles are stopped",
  order_by = "freq"
)
```

What can we say?

- **Key finding**: Since the bars are distributed by volume, in descending order, we can clearly observe a "ladder" trend in volume. We still have variance, but thanks to the volume we now have a more balanced case. We observe two clear "high-volume" areas (Brooklyn with over $640$k stops and Queens with over $550$k stops) followed by two "medium-volume" areas (Manhattan with $418$k and the Bronx with $381$k) and a distinct "low-volume" area (Staten Island, $166$k).
- **Interpretation**: Compared to the 2-wheels analysis where Staten Island was a tiny outlier ($1.5$k), here it holds a substantial sample size ($166$k). The shifting proportions, particularly the rise of the Bronx into a solid medium-volume category, suggests that geographic location could be a strong predictor of the probability of being stopped (in case the targets' ratio distributed by area follows a specific pattern).
- **Comment**: Unlike the 2-wheels analysis, Staten Island is not anymore a very rare label, even if the distribution is the lowest among the 5 areas. We have a slightly lower risk of overfitting here, but it is still to early to explore this aspect.
- **Proposed Action**: Same as the 2-wheels analysis.

### Borough

```{r count-stops-by-boro}
plot_univariate_cat(
  DF = vehstops, 
  FEATURE = boro, 
  X_LABEL = "Borough", 
  TITLE = "Distribution of Car Stops by Borough in NYC",
  SUBTITLE = "A more fine-grained bracketing of areas shows that Brookly (South) and Bronx dominate for amount of stops"
)
```

What can we say?

- **Key finding**: The granular bracketing shows that "Brooklyn (South)" dominates with $398,062$ stops, closely followed by the "Bronx" at $381,239$ stops, whereas in the 2-wheels analysis "North Queens" was the dominant category. Except the top-two volume boroughs, the remaining boroughs seem to have an almost uniform (decreasing) distribution ranging between $166$k - $276$k.
- **Interpretation**: For what concerns granularity, we can draw the same conclsions as the 2-wheels case. On the other hand, since every category contains a massive amount of observations, we probably do not run the risk encountered in the 2-wheels context where low-frequency categories caused high variance and wide confidence intervals in our parameter estimates.
- **Comment**: Same as the 2-wheels analysis.
- **Proposed Action**: Same as the 2-wheels analysis.

### Latitude and Longitude for Stop Density

For what concerns these two predictors, we will use the to plot a spatial heat map of the stops with these values available.

```{r stop-density}
nyc_map <- counties(state = "NY", cb = TRUE) %>%
  filter(NAME %in% c("Bronx", "Kings", "New York", "Queens", "Richmond"))

bins <- 100
lon_breaks <- seq(-74.26, -73.71, length.out = bins + 1)
lat_breaks <- seq(40.48, 40.92, length.out = bins + 1)

grid_sum <- vehstops %>%
  filter(!is.na(longitude) & !is.na(latitude)) %>%
  mutate(
    xbin = cut(longitude, breaks = lon_breaks, include.lowest = TRUE, right = FALSE),
    ybin = cut(latitude, breaks = lat_breaks, include.lowest = TRUE, right = FALSE)
  ) %>%
  mutate(
    x = as.numeric(gsub(".*\\[(.*),.*", "\\1", xbin)) + (lon_breaks[2] - lon_breaks[1])/2,
    y = as.numeric(gsub(".*\\[(.*),.*", "\\1", ybin)) + (lat_breaks[2] - lat_breaks[1])/2
  ) %>%
  group_by(x, y) %>%
  summarise(n = n(), .groups = "drop")

ggplot() +
  geom_sf(data = nyc_map, fill = "white", color = "grey80", size = 0.2) +
  geom_tile(data = grid_sum, aes(x, y, fill = n), alpha = 0.8) +
  scale_fill_viridis_c(trans = "log10", name = "Stops (log10)", option = "mako", direction = -1) +
  coord_sf(xlim = c(-74.26, -73.71), ylim = c(40.48, 40.92)) +
  theme_minimal() +
  theme(panel.grid = element_blank()) +
  labs(title = "Heatmap of Cars Stops Density in NYC - 100 Bins", x = "Longitude", y = "Latitude")
```

```{r, include=FALSE}
rm(grid_sum, nyc_map, bins, lat_breaks, lon_breaks)
```

What can we say?

- **Key finding**: The heatmap reveals that, with respect to the 2-wheels case, the hotspots are the same, although the increase in volume is notable.
- **Interpretation**: Same as the 2-wheels analysis.
- **Proposed Action**: Same as the 2-wheels analysis.

## Time-related predictors

### Year

```{r count-stops-by-year}
vehstops %>%
  filter(!is.na(year)) %>%
  count(year) %>%
  mutate(year = as.numeric(as.character(year))) %>%
  arrange(year) %>%
  mutate(
    prev_n = lag(n), pct_change = (n - prev_n) / prev_n * 100,
    label_text = ifelse(
      is.na(pct_change), 
      comma(n), 
      paste0(comma(n), "\n(", ifelse(pct_change > 0, "+", ""), round(pct_change, 1), "%)")
    )) %>%
  ggplot(aes(x = factor(year), y = n, group = 1)) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_point(color = "steelblue", size = 3) +
  geom_text(aes(label = label_text), vjust = -0.8, nudge_x = -0.1, size = 3.5, lineheight = 0.8) +
  scale_y_continuous(labels = comma, limits = c(0, NA), expand = expansion(mult = c(0, 0.2))) +
  theme_minimal() +
  theme(
    panel.grid.major.x = element_blank(), 
    plot.title = element_text(size = 14),
    plot.subtitle = element_text(size = 10, color = "grey50")
    ) +
  labs(
    title = "Annual Volume of Stops for Cars in NYC",
    subtitle = "In 3 Years the Total Volume Increased by almost 200k Stops",
    x = "Year", y = "Count of Stops"
    )
```

What can we say?

- **Key finding**: There has been a steady, multi-year increase in the volume of police stops involving 4-wheeled vehicles, with the absolute number of encounters expanding from $611,734$ in $2023$ to $799,853$ in $2025$. Overall, the volume of stops does not show a very unbalanced distribution among the three years.
- **Interpretation**: The growth trajectory is non-linear but far more stable than the explosive surge seen in the 2-wheels analysis. The $21.9\%$ jump between $2023$ and $2024$ represents a significant upward trend, which is suggesting a broader intensification of traffic enforcement across the city. On the other hand, the $7.2\%$ increase in 2025 indicates that while the absolute volume of activity is still rising, the rate of acceleration is starting to stabilize.
- **Comment**: Same as the 2-wheels analysis.
- **Proposed Action**: Same as the 2-wheels analysis.

### Month

```{r count-stops-by-month}
month_data <- vehstops %>%
  filter(!is.na(month)) %>%
  count(month) %>%
  mutate(month = factor(month, levels = 1:12)) %>%
  arrange(month) %>%
  mutate(
    prev_n = lag(n), pct_change = (n - prev_n) / prev_n * 100,
    label_text = ifelse(
      is.na(pct_change), comma(n),
      paste0(comma(n), "\n(", ifelse(pct_change > 0, "+", ""), round(pct_change, 1), "%)"))
  )

month_data %>%
  ggplot(aes(x = month, y = n, group = 1)) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_point(color = "steelblue", size = 3) +
  geom_text(aes(label = label_text), vjust = -1, nudge_x = -0.1, size = 3, lineheight = 0.8) +
  scale_x_discrete(labels = month.abb) +
  scale_y_continuous(labels = comma, limits = c(0, NA), expand = expansion(mult = c(0, 0.3))) +
  theme_minimal() +
  theme(
    panel.grid.major.x = element_blank(),
    plot.title = element_text(size = 14),
    plot.subtitle = element_text(color = "grey50", size = 10)
  ) +
  labs(
    title = "Monthly Volume of Stops for Cars in NYC",
    subtitle = "Over 3 Years, Summer is when the Least Amount of Stops take place",
    x = "Month",
    y = "Count of Stops"
  )
```

```{r, include=FALSE}
rm(month_data)
```

What can we say?

- **Key finding**: The data exhibits a highly distinct monthly pattern for 4-wheels vehicles, contrasting sharply with the June to October summer plateau observed for bikers. Here, stop volumes peak in the early months of the year, reaching an absolute maximum in March at $208,661$ stops, before experiencing a plateau, starting in June, that bottoms out in September at $154,192$ stops (a $-7.4\%$ drop from August).
- **Interpretation**: Unlike 2-wheeled vehicles, **4-wheeled vehicles are stopped regardless of the season and the related weather conditions**. This explains the high volume baseline maintained in January ($204,603$) and February ($200,831$). The steady downward trajectory into June ($-16.4\%$) and July may suggest a shift in enforcement.
- **Comment**: Most of the conclusions we can make here come from the 2-wheels analysis. The interesting aspect is that the peaks are almost completely counter-cyclical to the 2-wheels plot. This may suggest that, for standard automobiles, the police behavior shows a mechanic that is entirely independent from seasonal weather. If this was true, we could not compare fairly the two datasets.
- **Proposed Action**: Same as the 2-wheels analysis.

Let us see this with respect to the year

```{r year-monthly-distribution}
plot_data <- vehstops %>%
  filter(!is.na(year) & !is.na(month)) %>%
  count(year, month) %>%
  mutate(month = factor(month, levels = 1:12)) %>%
  group_by(year) %>%
  mutate(
    n_jan = n[month == 1],
    n_max = max(n),
    is_max = n == n_max,
    pct_max = round((n - n_jan) / n_jan * 100, 0),
    pct_dec = round((n - n_max) / n_max * 100, 0),
    
    label_text = case_when(
      month == 1 ~ comma(n), 
      is_max ~ paste0(comma(n), "\n(", ifelse(pct_max > 0, "+", ""), pct_max, "%)"),
      month == 12 ~ paste0(comma(n), "\n(", pct_dec, "%)"),
      TRUE ~ ""
    )
  )

# Faceted plot
plot_data %>%
  ggplot(aes(x = month, y = n, group = 1)) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_point(color = "steelblue", size = 2) +
  geom_text(aes(label = label_text), vjust = -1, hjust = 0.2, size = 2.5, lineheight = 0.8) +
  facet_wrap(~year, nrow = 1) +
  scale_x_discrete(
    labels = c("J", "F", "M", "A", "M", "J", "J", "A", "S", "O", "N", "D"), 
    expand = expansion(mult = c(0.1, 0.1))
    ) +
  scale_y_continuous(labels = comma, limits = c(0, NA), expand = expansion(mult = c(0, 0.4))) +
  coord_cartesian(clip = "off") +
  theme_minimal() +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    plot.title = element_text(size = 14),
    plot.subtitle = element_text(color = "grey50", size = 10),
    strip.text = element_text(size = 11)
  ) +
  labs(
    title = "Monthly Volume of Stops, by Year, for Cars in NYC",
    subtitle = "Custom Growth/Decline analysis: January vs Max, and Max vs December",
    x = "Month",
    y = "Count of Stops"
  )
```

```{r, include=FALSE}
rm(plot_data)
```

What can we say?

- **Key finding**: There is a pronounced shifting trend in the intra-year volume distribution over the three-year period. In $2023$, the annual maximum occurred early in March ($64,991$, an $11\%$ increase from January) followed by a sharp drop to December ($38,193$, a $-41\%$ decline). In $2024$ the monthly amount of stops follows a slightly decreasing trend. By $2025$, the behavior changes, as the year begins at a much higher baseline peak in January ($81,793$). Then it shows a steep mid-year contraction through summer, and it ends with a surge (in October) that leads to a much softer winter decline by December ($62,748$, a $-23\%$ drop from the annual maximum).
- **Interpretation**: The data indicates that enforcement patterns for 4-wheels vehicles are having structural adjustments over the years, although the overall volume keeps increasing. The widening gap between the high-volume peaks and the low-volume troughs seen in the 2-wheels context does not apply here. Instead, the "off-season" drop in December is becoming significantly less severe, moving from $-41\%$ in $2023$ to $-23\%$ in $2025$, which implies a stabilization of late-year policing activities.
- **Comment**: The plot clearly shows that while the baseline (January) is rising substantially over the years. Furthermore, we notice that the *shape* of the annual trend is shifting from looking like a spring, to a 2-phase distribution with a notable winter increase. This shift confirms our hypothesis that seasonal **vehicle usage is not the primary factor** for stops. Instead, this is probably due to the policy shifts within the police department.
- **Proposed Action**: Here is impossible to include the same `hot_season` factor, as we would like to restrict the focus months over the June-September period. Therefore, we skip it.

### Dow/Weekday, Day and Doy

We have decided to exclude 'Day of Month' (`day`) and 'Day of Year' (`doy`) from our feature set to mitigate 'calendar artifact' bias.

As we can see below, the variation in month length creates false activity dips for end-of-month dates

```{r day-bias}
vehstops %>%
  filter(!is.na(day)) %>%
  count(day) %>%
  ggplot(aes(x = day, y = n)) +
  geom_line(color = "grey60", alpha = 0.5) +
  geom_smooth(method = "loess", color = "steelblue", se = FALSE) +
  theme_minimal() +
   theme(
    panel.grid.major.x = element_blank(),
    plot.title = element_text(size = 14),
    plot.subtitle = element_text(color = "grey50", size = 10)
  ) +
  labs(title = "Volume of Stops for Cars, by Day of the Month (1-31), in NYC",
       subtitle = "The drop towards 30-31 indicates it is not a behavior, but depends on the month!",
       x = "Day of the Month", y = "Count of Stops")
```

Furthermore, `doy` variable suffers from inconsistent data density.

```{r doy-incosistent-3-years-span}
plot_data <- vehstops %>%
  mutate(date = as.Date(paste(year, month, day, sep = "-"))) %>%
  count(date)

plot_data %>%
  ggplot(aes(x = date, y = n)) +
  geom_line(color = "grey70", alpha = 0.6, linewidth = 0.5) +
  geom_smooth(method = "loess", span = 0.2, color = "steelblue", se = FALSE, linewidth = 1.2) +
  scale_x_date(date_breaks = "3 months", date_labels = "%b %Y") +
  theme_minimal() +
  theme(
    panel.grid.minor = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(size = 14),
    plot.subtitle = element_text(color = "grey50", size = 10)
  ) +
  labs(
    title = "Volume of Stops for Cars, by Day of the Year (1-365), in NYC",
    subtitle = "Continuous time series shows consistent peak in February",
    x = "Period", 
    y = "Count of Stops"
  )
```

```{r, include=FALSE}
rm(plot_data)
```

While we observe a consistent seasonal clustering peaking in late winter and early spring, our current 3-year dataset captures three seasonal cycles. This is sufficient to identify a recurring pattern, but insufficient to establish it as a long-term trend.

The next plot, demonstrates high daily volatility (indicated by the grey jitter) that obscures the underlying seasonal trend (the blue trend line). Because the daily fluctuation is relative to the overall pattern, including 'Day of Year' as a feature risks overfitting.

```{r doy-inconsistent-density}
vehstops %>%
  filter(!is.na(doy)) %>%
  count(doy) %>%
  ggplot(aes(x = doy, y = n)) +
  geom_line(color = "grey60", alpha = 0.5) +
  geom_smooth(method = "loess", color = "steelblue", se = FALSE) +
  theme_minimal() +
   theme(
    panel.grid.major.x = element_blank(),
    plot.title = element_text(size = 14),
    plot.subtitle = element_text(color = "grey50", size = 10)
  ) +
  labs(title = "Volume of Stops for Cars, by Day of the Year (1-365), in NYC",
       subtitle = "The extreme jitter confirms this feature is too noisy to predict, but the trend favors colder months",
       x = "Day of Year", y = "Count of Stops")
```

The extreme jitter confirms this feature is too noisy to predict, but unlike the 2-wheels analysis where the trend favors hot months, the 4-wheels trend remains relatively elevated in the first half of the year before dipping during the summer months.

On the other hand, we can study the "Day of the Week" which usually abstract better.

```{r dow-dis}
vehstops %>%
  filter(!is.na(dow)) %>%
  count(dow) %>%
  mutate(
    dow = factor(dow, levels = 1:7, 
                 labels = c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun")),
    pct_change = (n - lag(n)) / lag(n) * 100,
    label_text = ifelse(is.na(pct_change), 
                        comma(n), 
                        paste0(comma(n), "\n(", ifelse(pct_change > 0, "+", ""), round(pct_change, 1), "%)"))
  ) %>%
  ggplot(aes(x = dow, y = n, group = 1)) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_point(color = "steelblue", size = 3) +
  geom_text(aes(label = label_text), vjust = -1, size = 3, lineheight = 0.8) +
  scale_y_continuous(labels = comma, 
                     limits = c(0, NA), 
                     expand = expansion(mult = c(0, 0.4))) +
  theme_minimal() +
  theme(
    panel.grid.major.x = element_blank(),
    plot.title = element_text(size = 14),
    plot.subtitle = element_text(color = "grey50", size = 10)
  ) +
  labs(
    title = "Daily Volume of Stops for Cars in NYC",
    subtitle = "Daily volume and day-over-day growth/decline show high stop counts during mid-week",
    x = "Day of the Week", 
    y = "Count of Stops"
  )
```

What can we say?

- **Key finding**: Enforcement activity follows a distinct "Mid-Week Peak" cycle that strongly mirrors the weekly rhythm seen in the 2-wheels analysis. The volume starts at its absolute lowest on Monday ($230,725$), raises aggressively ($+52\%$) into Tuesday ($350,695$), peaks on Wednesday at $380,203$ stops ($+8.4\%$), and then undergoes a multi-day decline through the end of the week, bottoming out on Sunday ($236,756$).
- **Comment**: Same as the 2-wheels analysis.
- **Proposed Action**: Same as the 2-wheels analysis.

### Hour and Time Bracket

First of all we look at the overall cycle of stops volume

```{r hour-dist}
hours_to_show <- c(0, 4, 8, 12, 16, 19, 20, 23)

full_data <- vehstops %>%
  filter(!is.na(hour)) %>%
  count(hour)

highlight_data <- full_data %>%
  filter(hour %in% hours_to_show) %>%
  arrange(hour) %>%
  mutate(
    pct_change = (n - lag(n)) / lag(n) * 100,
    label_text = ifelse(
      is.na(pct_change), 
      comma(n),
      paste0(comma(n), "\n(", ifelse(pct_change > 0, "+", ""), round(pct_change, 1), "%)"))
  )

ggplot() +
  geom_line(data = full_data, aes(x = hour, y = n), color = "grey80", linewidth = 1) +
  geom_point(data = highlight_data, aes(x = hour, y = n), color = "steelblue", size = 3) +
  geom_text(data = highlight_data, aes(x = hour, y = n, label = label_text), 
            vjust = -0.6, size = 3, lineheight = 0.8) +
  scale_x_continuous(breaks = 0:23) +
  scale_y_continuous(labels = comma, limits = c(0, NA), expand = expansion(mult = c(0, 0.4))) +
  theme_minimal() +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    plot.title = element_text(size = 14),
    plot.subtitle = element_text(color = "grey50", size = 10)
  ) +
  labs(
    title = "Hourly Volume of Stops for Cars in NYC",
    subtitle = "The stops increase during main commute hours, with a peak at 4:00 PM ",
    x = "Hour of Day (0-23)", 
    y = "Count of Stops"
  )
```

```{r, include=FALSE}
rm(hours_to_show, full_data, highlight_data)
```

What can we say?

- **Key finding**: Volume of stops is characterized by a 3-stages daily cycle. First, it follows a night time decline towards an absolute minimum at $4:00$ ($23,374$, a $-79.7\%$ drop with respect to $00:00$). Then, we have a rapid morning ascent to before localized peak at $9:00$ (at $8:00$ we have $105,378$, which is an increase of $+350.8\%$ w.r.t. 4 AM), followed by a midday decline before a rapid afternoon surge to an absolute peak at $16:00$ ($144,954$, $+80.6\%$ w.r.t. 12 PM), and a sustained high-volume evening plateau before tailing off late at night.
- **Interpretation**: This distribution indicates that the enforcement cycle is driven by three phases: the **Morning Commuter Phase** (centered around $8:00$), the **Midday Transition Phase** (which causes the temporary dip down to $80,271$ at $12:00$), and the **Late Afternoon Rush Phase** (surging aggressively to the absolute peak at $16:00$ and tapering slowly through the evening hours). Unlike the 2-wheels analysis, the 4-wheels data shows shifts that perfectly align with major automotive traffic shifts throughout the city.
- **Proposed Action**: We will maintain `hour` as a temporal feature in our predictive framework. As an alternative, we *could* try to properly capture this pattern. For instance, we could **transform the continuous hour variable using a periodic basis function**, so that it becomes a continuous feature.

Before anything else, let us see whether this cycle changed throughout the years.

```{r hour-year-dist}
hours_to_show <- c(0, 4, 8, 12, 16, 20, 23)

full_data <- vehstops %>%
  filter(!is.na(year) & !is.na(hour)) %>%
  count(year, hour)

highlight_data <- full_data %>%
  filter(hour %in% hours_to_show) %>%
  group_by(year) %>%
  arrange(hour) %>%
  mutate(
    pct_change = (n - lag(n)) / lag(n) * 100,
    label_text = ifelse(
      is.na(pct_change), 
      comma(n), 
      paste0(comma(n), "\n(", ifelse(pct_change > 0, "+", ""), round(pct_change, 1), "%)"))
  )

ggplot() +
  geom_line(data = full_data, aes(x = hour, y = n), color = "grey85", linewidth = 0.8) +
  geom_point(data = highlight_data, aes(x = hour, y = n), color = "steelblue", size = 2) +
  geom_text(data = highlight_data, aes(x = hour, y = n, label = label_text), 
            vjust = -0.7, size = 2.5, lineheight = 0.8) +
  facet_wrap(~year, nrow = 1) +
  scale_x_continuous(breaks = c(0, 4, 8, 12, 16, 20, 23)) +
  scale_y_continuous(labels = comma, limits = c(0, NA), expand = expansion(mult = c(0, 0.4))) +
  theme_minimal() +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    plot.title = element_text(size = 14),
    plot.subtitle = element_text(color = "grey50", size = 10),
    strip.text = element_text(size = 11)
  ) +
  labs(
    title = "Hourly Volume of Stops, by Year, for Cars in NYC",
    subtitle = "The same enforcement pattern emerges, throughout the 3 years period",
    x = "Hour of Day (0-23)", 
    y = "Count of Stops"
  )
```

```{r, include=FALSE}
rm(hours_to_show, full_data, highlight_data)
```

What can we say?

- **Key finding**: Across all the three years, the enforcement volume shows the same pattern as stated for the previous plot, except for the period that goes from $16:00$ to $23:00$ which seems to be shifting. Two aspects remain evident: there are a minor peak during the morning rush at $8:00$ and a dominant peak during the late afternoon and evening. In $2023$ and $2024$, the absolute highest volume of stops occurred at $16:00$ ($4:00$ PM). By $2025$, while the $16:00$ peak remains the highest absolute value ($51011$ stops), the volume remains high and remains nearly unchanged until $20:00$ with $49857$ stops.
- **Interpretation**: Unlike what we observed for two-wheeled vehicles, car stops maintain a consistent primary afternoon peak at $16:00$ across the entire $2023$ to $2025$ period. The increase in volume at $20:00$ in $2025$ suggests that evening traffic enforcement has been growing. Instead, the morning rush hour peak at $8:00$ remains stable in structure across all years.

Let us check the time brackets previously built.

```{r time-bracket-distribution}
plot_univariate_cat(
  DF = vehstops, 
  FEATURE = time_bracket, 
  X_LABEL = "Time Bracket", 
  TITLE = "Distribution of Time Brackets for Cars in NYC",
  SUBTITLE = "Enforcement is heavily skewed toward the latter half of the day (16-23)"
)
```

What can we say?

- **Key finding**: The "Afternoon/evening ($16-19$)" bracket stands as the absolute dominant peak with $546,748$ stops, closely followed by the "Night ($20-23$)" and "Morning commute ($8-11$)" brackets with $449,686$ and $410,555$ stops respectively. The lowest volume is in the "Early morning ($4-7$)" window that registers just $138,852$ stops.
- **Interpretation**: Just like for 2-wheels vehicles, the volume of enforcement is tied around the latter half of the day ($16-23$), which confirm the pattern shown by the previous bar plots.
- **Comment**: In contrast to the two-wheeled vehicle analysis, the car data shows a smoother distribution.
- **Proposed Action**: We can safely maintain our existing `time_bracket` feature engineering exactly as it is for the upcoming modeling phase.

Also, we are going to drop minute as its level of granularity cannot be properly included in the dataset.

## Driver related predictors

### Reported Age

We have mainly one issue here. First, we have many instances of drivers with unknown age,

```{r age-bracket-dist}
plot_univariate_cat(
  DF = vehstops, 
  FEATURE = age_bracket, 
  X_LABEL = "Age Bracket", 
  TITLE = "Distribution of Car Stops in NYC by Age Bracket",
  SUBTITLE = "Enforcement volume is highly concentrated in the demographic (18-44)"
)
```

What can we say?

- **Key finding**: The demographic distribution of stops is highly concentrated in the $25-34$ age category with $631,234$ observations, followed by the $35-44$ and $18-24$ brackets. Furthermore, the "Unknown" age category represents a significant portion with $118,990$ records, while the "Under 18" category is the smallest group with just $9298$ stops.
- **Interpretation**: Although, older age groups show lower frequency, they are much more well-represented here compared to the two-wheeled vehicle analysis. In fact, most categories are well represented here too.
- **Comment**: Same as 2-wheels analysis.
- **Proposed Action**: We will retain the "Unknown" category as a level within our `age_bracket` feature rather than using imputation.

And now, let us see how this looks when looking at the full distribution

```{r age-vs-age-bracket}
age_data <- vehstops %>%
  filter(!is.na(age), age >= 0, age < 100) %>%
  mutate(
    age_group = case_when(
      age < 18 ~ "<18",
      age >= 18 & age <= 24 ~ "18–24",
      age >= 25 & age <= 34 ~ "25–34",
      age >= 35 & age <= 44 ~ "35–44",
      age >= 45 & age <= 54 ~ "45–54",
      age >= 55 & age <= 64 ~ "55–64",
      age >= 65 ~ "65+"
    ),
    age_group = factor(age_group, levels = c("<18", "18–24", "25–34", "35–44", "45–54", "55–64", "65+"))
  )

ggplot(age_data, aes(x = age, fill = age_group)) +
  geom_histogram(binwidth = 1, color = "white", linewidth = 0.1) +
  scale_fill_viridis_d() +
  theme_minimal() +
   theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    plot.title = element_text(size = 14),
    plot.subtitle = element_text(color = "grey50", size = 10),
    strip.text = element_text(size = 11)
  ) +
  labs(
    title = "Age Distribution of Stopped Car Drivers in NYC by Target Bracket",
    subtitle = "Enforcement volume is confirmed to be concentrated in 20-45 age range",
    x = "Driver Age", 
    y = "Count of Stops",
    fill = "Age Group"
  ) +
  theme(legend.position = "bottom") +
  guides(fill = guide_legend(nrow = 2))
```

```{r, include=FALSE}
rm(age_data)
```

What can we say?

- **Key finding**: The enforcement volume shows a rapid increase, from the underage year, that becomes a sustained plateau that spans from around the age $23$ through age $35$. Here, most age labels can count on more than $60000$ stops. After the age of $35$, the volume starts to steadily decline through middle age. After that, it shifts into a long tail that, past age $75$, becomes extremely flat.
- **Comment**: The four-wheel stops distribution show a **much broader plateau**, with respect to the 2-wheels analysis, which is probably due to the higher concentration of older drivers (hinted by the slow decreasing pattern).
- **Proposed Action**: Same as 2-wheels analysis, even if here the best thing to do is to filter over $85$, but we want it to be comparable with the other analysis.

Let us remove the outliers

```{r remove-age-outliers}
# Keep the NAs
vehstops <- vehstops %>%
  filter(age < 76 | is.na(age))
```

From now on we will consider drivers of age in the range 16-75.

```{r age-outlier-boxplot-after}
vehstops %>%
  filter(!is.na(age)) %>%
  ggplot(aes(y = age)) +
  geom_boxplot(fill = "steelblue", alpha = 0.6, width = 0.4,
               outlier.size = 1.5, outlier.alpha = 0.4) +
  coord_flip() +
  theme_minimal() +
  theme(
    panel.grid.major.y = element_blank(),
    plot.title = element_text(size = 14),
    plot.subtitle = element_text(color = "grey50", size = 10)
  ) +
  labs(
    title = "Age Distribution after Outlier Removal (76+ excluded)",
    subtitle = "Remaining values span the plausible 16-75 enforcement range",
    y = "Reported Age", x = NULL
  )
```

What can we say?

- **Key finding**: The median is centered around $35$, with the interquartile range ($IQR$) stretching roughly from $27$ to $47$. The lower whisker reaches near the minimum legal driving age, and the upper whisker gets up to $75$ without showing any point as outlier.
- **Interpretation**: The fact that the $IQR$ spans $27-47$ reflects the increased presence of older drivers in the car dataset.
- **Comment**: While the boxplot for the two-wheeled vehicle analysis shows sparsity for older drivers ($65-75$), here the car data does not label them as possible outliers for the current distribution.
- **Proposed Action**: Same as 2-wheels analysis.

### Reported Sex

```{r sex-distribution}
plot_univariate_cat(
  DF = vehstops, 
  FEATURE = sex, 
  X_LABEL = "Reported Sex", 
  TITLE = "Distribution of Car Stops in NYC by Reported Sex",
  SUBTITLE = "The vast majority of stopped drivers are males",
  order_by = "freq"
)
```

What can we say?

- **Key finding**: The "Male" (M) drivers represent th majority of stops with $1,736,848$ observations. Unlike the 2-wheels case, "Female" (F) drivers are not under-represented and they account for $395,347$ stops, while the "Unknown" (U) category remains a small minority with $13,392$ records.
- **Interpretation**: Same as 2-wheels analysis.
- **Comment**: In comparison to the two-wheeled vehicle analysis, the female proportion is significantly larger here.
- **Proposed Action**: Same as 2-wheels analysis.

### Reported Ethnicity

```{r ethnicity-distribution}
plot_univariate_cat(
  DF = vehstops, 
  FEATURE = ethnicity, 
  X_LABEL = "Reported Ethnicity", 
  TITLE = "Distribution of Car Stops in NYC by Reported Ethnicity",
  SUBTITLE = "Hispanic drivers seem to account for the plurality of stops",
  order_by = "freq"
)
```

What can we say?

- **Key finding**: "Black" drivers represent the largest single group with $656,109$ stops, followed by "Hispanic" drivers with $565,531$ observations. Together, they represent the majority of the data. Furthermore, the amount of stops for "White" car drivers is $426,594$, which is incredibly high with respect to the 4th category (by amount of stops), "Asian / Pacific Islander" ($270,311$). In fact, the 2-wheels analysis found that these two labels had almost the same amount of observations. For what concerns the remaining categories, we have "Unknown" ($193,050$), and "Other" ($33,992$) for last.
- **Interpretation**: This profile indicates that four-wheeled vehicle enforcement is heavily concentrated within Black and Hispanic demographics across the city. The high absolute count for the "Unknown" category matches the pattern we observed with missing demographic metrics elsewhere in our analysis. Given the large sample size across all 4 main groups (White, Black, Hispanic, and Asian) we can be sure to have plenty of variability to learn from, unlike the other dataset.
- **Comment**: In comparison to the previous (2-wheels) analysis, where the distribution was dominated by Hispanic riders, the four-wheeled data shows a more distributed but still unequal breakdown. Here, Black drivers constitute the absolute peak subset of stops, even if the majority of the NYC population is White.
- **Proposed Action**: Same as 2-wheels analysis.

------------------------------------------------------------------------

# Bivariate EDA

## Categorical Association of Binary Response vs. Categorical Predictors

When dealing with a dataset of this size, traditional p-values from chi-squared tests become meaningless. To get around this, we use Cramer’s V to measure effect size (association).

Because of how the `table()` function works in R, we can also include integer predictors like `hour` and `age`: they do not take truly continuous values in this dataset, so treating each integer as its own category is valid.

```{r categorical-correlation-heatmap}
# Excluding doy (redundant) and dow (we use weekday instead)
columns <- c(
  "year", "month", "weekday", 
  "hour", "time_bracket",
  "area", "boro", 
  "sex", "ethnicity", "age", "age_bracket",
  "arrested"
  )

plot_categorical_correlation(
  DF = vehstops, 
  FEATURES = columns,
  TITLE = "Correlation between Categorical Variables (Cramer's V)"
)
```

What can we say?

- **Key finding**: The association matrix reveals that the only predictors showing a noticeable signal with `arrested` are, `ethnicity` ($0.1$), `hour`/`time_bracket` (both at $0.09$). Here, other demographic features such as `age`/`age_bracket` ($0.07$) and `sex` ($0.05$) have a more modest association. Spatial features like `boro` ($0.06$)/`area` ($0.05$) show very low, but still relevant signals. The matrix highlights a moderate association between these top predictors, specifically:
  - between `age`/`age_bracket` $\leftrightarrow$ `hour` / `time_bracket` ($\approx 0.08-0.09$)
  - between `sex` $\leftrightarrow$ `hour` / `time_bracket` ($\approx 0.07-0.08$)
  - between `ethnicity` $\leftrightarrow$ `hour` / `time_bracket` ($\approx 0.07-0.08$)
  - between `ethnicity` $\leftrightarrow$ `area`/`boro` ($0.17 - 0.19$)
  - between `ethnicity` $\leftrightarrow$ `sex` ($0.15$).
- **Interpretation**: Temporal and ethnic features carry the strongest individual baseline impacts when predicting an arrest, while standalone spatial attributes are less informative on their own for four-wheeled vehicles. The persistent $0.17$ to $0.19$ association between ethnicity and geography suggests that the demographic composition of stops remains non-uniformly distributed across different parts of the city.
- **Comment**: While the previous analysis shows a stronger connection to geographical signals (where `area` and `boro` scored $0.13$), here we probably encounter a more distributed arrest pattern. However, because no single predictor strongly correlates with `arrested`, models relying purely on main effects will likely underfit the data. Therefore, we must be careful when we consider the interactions.
- **Proposed Action**: Same as 2-wheels analysis, with the sole difference that here we focus a bit more on demographics then spatial features.

## Correlation of Binary Response vs. Numerical Predictors

```{r numerical-correlation-heatmap}
# Ignoring Latitude and Longitude
numerical_cols <- c(
  "arrested",
  "year", "month", "day", "dow", "doy", "hour", 
  "age"
  )

cor_matrix <- cor(vehstops[, numerical_cols], use = "complete.obs")

corrplot(cor_matrix, method = "color", type = "upper",
         order = "hclust", addCoef.col = "black", number.cex = 0.7,
         tl.col = "black", tl.srt = 45,
         title = "Correlation Matrix of Numerical Features",
         mar = c(0, 0, 2, 0))
```

What can we say?

- **Key finding**: The Pearson correlation matrix shows that our binary target `arrested` has a negligible linear relationship with all continuous numerical features. The strongest standalone linear relationship with the target appears in `age` ($-0.07$), which indicates a very slight negative trend, followed by `hour` ($0.03$). Calendar features like `month`, `doy` (day of year), `year`, `dow` (day of week), and `day` all show absolute correlation coefficients of $0.01$ or lower, indicating a complete absence of direct global linear relationships.
- **Interpretation**: The near-zero linear correlations across the board demonstrate that the probability of an arrest cannot be effectively modeled or predicted using standard linear relationships on raw numerical predictors. The slight negative value for `age` implies that older stopped drivers have a marginally lower likelihood of being arrested.
- **Comment**: Linear correlation metrics are not suited for capturing the dynamics of this dataset. This is probably due to the fact, both the features (like cyclical hours or seasonal months) and the target variable (`arrested`) act non-linearly.
- **Proposed Action**: We could explicitly map the continuous predictors using non-linear modeling frameworks, such as incorporating polynomial expansions or using Generalized Additive Models (GAMs) with smoothing splines for `age` and `hour`.

## Geospatial Features

### Area

```{r area-vs-arrested-ratio}
ratio_by_cat(
  vehstops, arrested, area,
  "Area", "Arrest Rate (%)", "Arrest Ratio by Area for Cars Stopped in NYC",
  style = "stack", legend_title = "Stop ended in arrest?",
  target_labels = c("No", "Yes"),
  subtitle = "Although the ratios are low, the amount of arrests, in proportion, are incredibly high",
  label_vjust = 0.5,
  bar_width = 0.6,  
  text_size = 2.5,
  aspect_ratio = 0.9
)
```

Let us look at the numbers too

```{r confirm-area-vs-arrested-ratio}
vehstops %>%
  group_by(area, arrested) %>%
  summarise(count = n(), .groups = "drop_last") %>%
  mutate(percentage = count / sum(count))
```

What can we say?

- **Key finding**: The arrest rate ranges from a minimum of $2.5\%$ in Queens to a maximum of $4.8\%$ in the Bronx. This represents nearly a 2x increase in the conditional probability of an arrest depending purely on **where the stop occurs**. While the absolute majority of stops do not result in arrests across all territories, the Bronx consistently maintains the highest conversion rate, followed by Brooklyn at $3.5\%$, Staten Island at $3.0\%$, and Manhattan at $2.8\%$.
- **Interpretation**: The variance here is more stable, thanks to the size of the sample. As a consequence, even small fluctuations like the difference between $2.5\%$ in Queens and $4.8\%$ in the Bronx can be considered as a robust statistical signal. This confirms that geographic features are very important for our prediction.
- **Comment**: When compared against the previous analysis, car stops have a tighter distribution. However, the curious aspect we found concerns Bronx and Brooklyn areas. While they show a great difference in the total volume of stops, the amount of arrest (following a stop) is still comparable for those area. Furthermore, the amount of stops that **did not end up in an arrest**, in Brooklyn, is almost 2x with respect to the Bronx. In general we could say that the spatial distribution is smoother, but the Bronx still remains a problematic area. For this reason we wonder whether we could say that the Bronx is just a big "outlier" with respect to the remaining areas.
- **Proposed Action**: Same as 2-wheels analysis.

### Borough

```{r boro-vs-arrested-ratio}
ratio_by_cat(
  vehstops, arrested, boro,
  "Borough", "Arrest Rate (%)", "Arrest Ratio by Borough for Cars Stopped in NYC",
  style = "stack", legend_title = "Stop ended in arrest?",
  target_labels = c("No", "Yes"),
  subtitle = "The arrest rate seems to be well distributed among the boroughs",
  label_vjust = 0.5,
  bar_width = 0.8,
  text_size = 2.5
)
```

```{r confirm-borough-vs-arrested-ratio}
vehstops %>%
  group_by(boro, arrested) %>%
  summarise(count = n(), .groups = "drop_last") %>%
  mutate(percentage = count / sum(count))
```

What can we say?

- **Key finding**: Disaggregating (some of) the geographic data into finer regions does not preserve the overall ranking and reveals a minor intra-borough variance. While Manhattan's arrest rate as a whole appeared a bit low in our previous chart, here we observe steep internal change in rates: from a minimum of $1.6\%$ in the south of Manhattan to a substantial $3.9\%$ in the northern part of the same area. The remaining separated regions display an upward slope, with Queens ranging from $2.1\%$ (North) to $2.8\%$ (South), and Brooklyn climbing from $3.3\%$ (South) to $3.7\%$ (North).
- **Interpretation**: This fine-grained visualization confirms that the macro patterns are influenced by specific zones. For instance, within the same area of Manhattan, a stop in the northen side is more than twice as likely to result in an arrest compared to one in the southern part. This suggests that a finer geographic division will improve the explanatory power of our models.
- **Comment**: In the two-wheel case, the Bronx looks like an isolated outlier. Here, arrest conversion rates seem to be highly dependent on specific command zones within the same borough or neighboring ones. In fact, Manhattan (North) rivals the Bronx in arrest rate, while Manhattan (South) remains the absolute lowest (by both rate and number of arrests). This remarks the sharp division in enforcement behavior.
- **Proposed Action**:  Same as 2-wheels analysis.

We will need to choose between one or the other.

### Latitude and Longitude for Stop Rate

```{r stop-ratio-density}
nyc_map <- counties(state = "NY", cb = TRUE) %>%
  filter(NAME %in% c("Bronx", "Kings", "New York", "Queens", "Richmond"))

bins <- 100
lon_breaks <- seq(-74.26, -73.71, length.out = bins + 1)
lat_breaks <- seq(40.48, 40.92, length.out = bins + 1)

grid_sum_rate <- vehstops %>%
  filter(!is.na(longitude), !is.na(latitude), !is.na(arrested)) %>%
  mutate(
    xbin = cut(longitude, breaks = lon_breaks, include.lowest = TRUE, right = FALSE),
    ybin = cut(latitude, breaks = lat_breaks, include.lowest = TRUE, right = FALSE)
  ) %>%
  mutate(
    x = as.numeric(gsub(".*\\[(.*),.*", "\\1", xbin)) + (lon_breaks[2] - lon_breaks[1])/2,
    y = as.numeric(gsub(".*\\[(.*),.*", "\\1", ybin)) + (lat_breaks[2] - lat_breaks[1])/2
  ) %>%
  group_by(x, y) %>%
  summarise(
    total_stops = n(),
    arrest_rate = mean(arrested == 1),
    .groups = "drop"
  ) %>%
  filter(total_stops >= 5)

ggplot() +
  geom_sf(data = nyc_map, fill = "white", color = "grey80", linewidth = 0.2) +
  geom_tile(data = grid_sum_rate, aes(x, y, fill = arrest_rate), alpha = 0.8) +
  scale_fill_viridis_c(
    labels = scales::percent, 
    name = "Arrest Rate (%)", 
    option = "inferno",
    direction = -1
  ) +
  coord_sf(xlim = c(-74.26, -73.71), ylim = c(40.48, 40.92)) +
  theme_minimal() +
  theme(
    panel.grid = element_blank(),
    plot.title = element_text(size = 14),
    plot.subtitle = element_text(color = "grey50", size = 10)
  ) +
  labs(
    title = "Spatial Distribution of Arrest Rates for Cars Stopped in NYC",
    subtitle = "Percentage of stops resulting in arrest (Bins with n < 5 removed)",
    x = "Longitude", 
    y = "Latitude"
  )
```

What can we say?

- **Key finding**: The plot shows a more decentralized distribution of arrest rates for cars, characterized mainly by a uniform baseline with light yellow to orange shades ($0\%$ to $15\%$). On the other hand, isolated hot spots are visible (dark purple and black pixels) across northern Manhattan, parts of the Bronx, eastern Queens, and central Staten Island. In some cases, these hotspots show arrest rates spiking above $40\%$, but this is due to the visualization threshold and scale.
- **Interpretation**: Same as 2-wheels analysis.
- **Comment**: Car enforcement presents a much more fragmented scene. The baseline risk is quite homogenous across the five boroughs, meaning that the spikes are highly localized exceptions rather than trends. It could be interesting to build a map for different boroughs and analyze how this visualization changes for different hours of the day, or periods of the year.

Note: This was made to be comparable with the EDA for 2-wheels case but the visualization itself would probably benefit from better binning and a higher threshold. In fact, it becomes a harder to demonstrate the effects this way.

## Time related features

### Year

```{r year-vs-arrested-ratio}
arrest_trends <- vehstops %>%
  filter(!is.na(year), !is.na(arrested)) %>%
  group_by(year) %>%
  summarise(
    total_stops = n(),
    arrests = sum(arrested == 1),
    rate = arrests / total_stops, 
    .groups = 'drop'
  ) %>%
  arrange(year) %>%
  mutate(
    prev_rate = lag(rate),
    delta = (rate - prev_rate) * 100,
    label_text = ifelse(
      is.na(prev_rate), 
      scales::percent(rate, accuracy = 0.1), 
      paste0(scales::percent(rate, accuracy = 0.1), 
             "\n(", ifelse(delta > 0, "+", ""), round(delta, 1), " pts)")
    )
  )

ggplot(arrest_trends, aes(x = factor(year), y = rate, group = 1)) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_point(color = "steelblue", size = 3) +
  geom_text(aes(label = label_text), vjust = -0.8, size = 3.5, lineheight = 0.8) +
  scale_y_continuous(
    labels = scales::percent, 
    limits = c(0, max(arrest_trends$rate) * 1.3),
    expand = expansion(mult = c(0, 0.1))
  ) +
  theme_minimal() +
  theme(
    panel.grid.major.x = element_blank(),
    plot.title = element_text(size = 14),
    plot.subtitle = element_text(size = 10, color = "grey50")
  ) +
  labs(
    title = "Annual Arrest Rate has Stabilized in the last 3 years for Cars Stopped in NYC",
    subtitle = "The percentage of arrests seems to have reached a stable point for the first time",
    x = "Year", y = "Arrest Rate (%)"
  )
```

What can we say?

- **Key finding**: The line chart displays the trajectory of the annual arrest rate for four-wheeled vehicles over a three-year period, revealing an initial rise followed by stabilization. The proportion of stops resulting in an arrest rose from $2.8\%$ in $2023$ to $3.5\%$ in $2024$ (a jump of $+0.7$ percentage points), followed by a marginal reduction to $3.4\%$ in $2025$.
- **Interpretation**: From a time-related perspective, this shows that the probability (of an arrest given a stop) shifted between $2023$ and $2024$ but has plateaued.
- **Proposed Action**: Same as 2-wheels analysis.

### Month

```{r month-vs-arrested-ratio}
monthly_arrest_data <- vehstops %>%
  filter(!is.na(month), !is.na(arrested)) %>%
  group_by(month) %>%
  summarise(
    total_stops = n(),
    arrests = sum(arrested == 1),
    rate = arrests / total_stops,
    .groups = "drop"
  ) %>%
  mutate(month = factor(month, levels = 1:12)) %>%
  arrange(month) %>%
  mutate(
    prev_rate = lag(rate),
    delta = (rate - prev_rate) * 100,
    label_text = ifelse(
      is.na(prev_rate), 
      scales::percent(rate, accuracy = 0.1), 
      paste0(scales::percent(rate, accuracy = 0.1), 
             "\n(", ifelse(delta > 0, "+", ""), round(delta, 1), " pts)")
    )
  )

monthly_arrest_data %>%
  ggplot(aes(x = month, y = rate, group = 1)) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_point(color = "steelblue", size = 3) +
  geom_text(aes(label = label_text), vjust = -1.2, size = 3, lineheight = 0.8) +
  scale_x_discrete(labels = month.abb) +
  scale_y_continuous(
    labels = scales::percent, 
    limits = c(0, max(monthly_arrest_data$rate) * 1.3), 
    expand = expansion(mult = c(0, 0.1))
  ) +
  theme_minimal() +
  theme(
    panel.grid.major.x = element_blank(),
    plot.title = element_text(size = 14),
    plot.subtitle = element_text(color = "grey50", size = 10)
  ) +
  labs(
    title = "Monthly Arrest Rate Trend for Cars Stopped in NYC",
    subtitle = "Aggregated monthly arrest rate, the peak is found between January and December",
    x = "Month",
    y = "Arrest Rate (%)"
  )
```

What can we say?

- **Key finding**: The monthly arrest rate across the years, shows a highly stable pattern with some slight peaks concentrated at the absolute ends of the calendar year. The arrest rate starts at $3.6\%$ in January, then drops and plateaus between $3.1\%$ and $3.3\%$ from February through November. Finally, it shows a small surge to its annual peak of $3.7\%$ in December ($+0.4$ percentage points).
- **Interpretation**: From a temporal perspective, it seems that the probability of an arrest given a stop is essentially invariant.

#### Faceted

```{r month-faceted-vs-arrested-ratio}
plot_data <- vehstops %>%
  filter(!is.na(year), !is.na(month), !is.na(arrested)) %>%
  group_by(year, month) %>%
  summarise(
    arrest_rate = mean(arrested == 1), 
    .groups = "drop"
  ) %>%
  mutate(month = factor(month, levels = 1:12)) %>%
  group_by(year) %>%
  mutate(
    rate_jan = arrest_rate[month == 1],
    rate_max = max(arrest_rate),
    is_max = arrest_rate == rate_max,
  
    diff_jan = (arrest_rate - rate_jan) * 100,
    diff_dec = (arrest_rate - rate_max) * 100, 
    
    label_text = case_when(
      month == 1 ~ scales::percent(arrest_rate, accuracy = 0.1), 
      is_max ~ paste0(scales::percent(arrest_rate, accuracy = 0.1), "\n(", ifelse(diff_jan > 0, "+", ""), round(diff_jan, 1), " pts)"),
      month == 12 ~ paste0(scales::percent(arrest_rate, accuracy = 0.1), "\n(", round(diff_dec, 1), " pts)"),
      TRUE ~ ""
    )
  )


plot_data %>%
  ggplot(aes(x = month, y = arrest_rate, group = 1)) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_point(color = "steelblue", size = 2) +
  geom_text(aes(label = label_text), vjust = -1, hjust = 0.2, size = 2.5, lineheight = 0.8) +
  facet_wrap(~year, nrow = 1) +
  scale_x_discrete(
    labels = c("J", "F", "M", "A", "M", "J", "J", "A", "S", "O", "N", "D"), 
    expand = expansion(mult = c(0.1, 0.1))
  ) +
  scale_y_continuous(labels = scales::percent, limits = c(0, NA), expand = expansion(mult = c(0, 0.4))) +
  coord_cartesian(clip = "off") +
  theme_minimal() +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    plot.title = element_text(size = 14),
    plot.subtitle = element_text(color = "grey50", size = 10),
    strip.text = element_text(size = 11)
  ) +
  labs(
    title = "Monthly Arrest Rate, by Year, for Cars in NYC",
    subtitle = "The arrest rate pattern is still drifting",
    x = "Month",
    y = "Arrest Rate (%)"
  )
```

What can we say?

- **Key finding**: This faceted line chart breaks down the monthly arrest rates for cars across three consecutive years. This reveals that the internal pattern keeps changing over time. For instance, $2023$ shows an upward climb from a minimum of $2.4\%$ in February to a peak of $3.7\%$ in December ($+1.1$ percentage points), while $2024$ and $2025$ invert this pattern. In $2024$, the peak is $3.9\%$ in January, which then drops through summer, and reaches again $3.9\%$ in December. By $2025$, the absolute peak occurs in January at $4.1\%$ before showing a downward trend.
- **Comment**: Car enforcement does not seem to be bound by a meteorological cycle like delivery riders. On the other hand, the rapid shift of the peak from December ($2023$) to January ($2024$ and $2025$) suggests that this is more related to administrative changes (which we cannot account for).
- **Proposed Action**: It does not seem that we can effectively rely on the month (or even year) to capture how the probability of `arrested` scales across this timeline. We could try to include an interaction term `month * year` within our model.

### Day, Dow and Weekday

```{r day-vs-arrested-ratio}
daily_arrest_rate <- vehstops %>%
  filter(!is.na(day), !is.na(arrested)) %>%
  group_by(day) %>%
  summarise(
    total_stops = n(),
    arrests = sum(arrested == 1),
    rate = arrests / total_stops,
    .groups = "drop"
  )

daily_arrest_rate %>%
  ggplot(aes(x = day, y = rate)) +
  geom_line(color = "grey60", alpha = 0.5) +
  geom_smooth(method = "loess", color = "steelblue", se = FALSE) +
  scale_y_continuous(labels = scales::percent, limits = c(0, NA)) +
  theme_minimal() + 
  theme(
    panel.grid.major.x = element_blank(),
    plot.title = element_text(size = 14),
    plot.subtitle = element_text(color = "grey50", size = 10)
  ) +
  labs(
    title = "Arrest Rate by Day of the Month (1-31) for Cars Stopped in NYC",
    subtitle = "The trend seems to be stable across every day",
    x = "Day of the Month",
    y = "Arrest Rate (%)"
  )
```

What can we say?

- **Key finding**: Arrest rates are broadly flat across days 1 through 31, showing no meaningful trend.
- **Interpretation**: The flat pattern across days 1-28 indicates that the day-of-month variable carries no genuine enforcement signal.
- **Proposed Action**: Exclude `day` from our models.

```{r dow-vs-arrested-ratio}
daily_arrest_rate <- vehstops %>%
  filter(!is.na(dow), !is.na(arrested)) %>%
  group_by(dow) %>%
  summarise(
    total_stops = n(),
    arrests = sum(arrested == 1),
    rate = arrests / total_stops,
    .groups = "drop"
  ) %>%
  mutate(
    dow_label = factor(dow, levels = 1:7,
                       labels = c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun")),
    prev_rate = lag(rate),
    delta = (rate - prev_rate) * 100,
    label_text = ifelse(
      is.na(prev_rate),
      scales::percent(rate, accuracy = 0.1),
      paste0(scales::percent(rate, accuracy = 0.1),
             "\n(", ifelse(delta > 0, "+", ""), round(delta, 1), " pts)")
    )
  )

daily_arrest_rate %>%
  ggplot(aes(x = dow_label, y = rate, group = 1)) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_point(color = "steelblue", size = 3) +
  geom_text(aes(label = label_text), vjust = -1.0, size = 3, lineheight = 0.8) +
  scale_y_continuous(
    labels = scales::percent,
    limits = c(0, max(daily_arrest_rate$rate) * 1.3),
    expand = expansion(mult = c(0, 0.1))
  ) +
  theme_minimal() +
  theme(
    panel.grid.major.x = element_blank(),
    plot.title = element_text(size = 14),
    plot.subtitle = element_text(color = "grey50", size = 10)
  ) +
  labs(
    title = "Arrest Rate by Day of the Week for Cars Stopped in NYC",
    subtitle = "Arrest rate peaks on weekends",
    x = "Day of the Week",
    y = "Arrest Rate (%)"
  )
```

What can we say?

- **Key finding**: The plot displays a monotonic rise in the weekly arrest rate as we approach the weekend. The probability of an arrest given a stop begins at its minimum on Monday at $3.0\%$, hovers around $3.1\%$ to $3.2\%$ through Thursday, and then experiences a clear upward lift starting on Friday ($3.5\%$) before reaching its absolute maximum on Saturday at $3.8\%$ ($+0.3$ percentage points).
- **Interpretation**: The variation across the week is exceptionally tight, as the fluctuations span around $0.8$ percentage points between the Monday low and the Saturday peak. This tells us that the day of the week has a relatively weak impact compared to spatial or temporal features.
- **Comment**: We notice a shift toward the weekend with respect to the previous analysis, even if at lower ratios. Both vehicle types show an absolute maximum arrest rate on Saturday. Interestingly, car stops follow a more predictable increase, which highlights the transition from weekday traffic to weekend police behavior.
- **Proposed Action**: Same as 2-wheels analysis.

### Hour and Time Bracket

```{r hour-vs-arrested-ratio}
hourly_arrest_rate <- vehstops %>%
  filter(!is.na(hour), !is.na(arrested)) %>%
  group_by(hour) %>%
  summarise(
    total_stops = n(),
    arrests = sum(arrested == 1),
    rate = arrests / total_stops,
    .groups = "drop"
  )

hourly_arrest_rate %>%
  ggplot(aes(x = hour, y = rate)) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_point(color = "steelblue", size = 2) +
  scale_x_continuous(breaks = 0:23) +
  scale_y_continuous(labels = scales::percent, limits = c(0, NA),
                     expand = expansion(mult = c(0, 0.15))) +
  theme_minimal() +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    plot.title = element_text(size = 14),
    plot.subtitle = element_text(color = "grey50", size = 10)
  ) +
  labs(
    title = "Arrest Rate by Hour of Day for Cars Stopped in NYC",
    subtitle = "The hourly arrest rate follows a U-shaped diurnal pattern",
    x = "Hour of Day (0-23)",
    y = "Arrest Rate (%)"
  )
```

What can we say?

- **Key finding**: The line plot underlines a distinct time-dependent pattern in the arrest rate resulting from stops across a $24$-hour cycle. We observe a global minimum at $7:00$ ($07:00$ AM) where the arrest rate drops to around $0.6\%$, followed by a steady climb to a minor afternoon peak around $14:00$ ($02:00$ PM) near $2.9\%$. The absolute maximum occurs during the late-night and early-morning hours, specifically peaking at $3:00$ ($03:00$ AM) with an arrest rate reaching approximately $5.3\%$.
- **Interpretation**: This strong fluctuation suggests that, depending on the hour of the day, a stop is more likely to turn into an arrest. However, we are not able to tell whether this is the result of a change in police behavior or the driver's.
- **Comment**: When comparing this curve to the 2-wheeled analysis, we notice a remarkable similarity in their shape, even if here we start at a lower absolute maximum ($5.3\%$ versus nearly $11\%$). The remaining observations are the same of the 2-wheels analysis.
- **Proposed Action**: We may opt for the engineered `time_bracket` predictor. Also, we can test whether a continuous spline function using `hour` inside a Generalized Additive Model performs better than bucketing.

Let us look at our grouping now

```{r time-bracket-vs-arrested-ratio}
ratio_by_cat(
  vehstops, arrested, time_bracket,
  "Time bracket", "Arrest Rate (%)", "Arrest ratio by Custom Time Bracket for Cars Stopped in NYC",
  style = "stack", legend_title = "Stop ended in arrest?", target_labels = c("No", "Yes"),
  subtitle = "The night brackets show a slightly higher arrest rates among all"
)
```

```{r confirm-time-bracket-vs-target-ratio}
vehstops %>%
  group_by(time_bracket, arrested) %>%
  summarise(count = n(), .groups = "drop_last") %>%
  mutate(percentage = count / sum(count))
```

What can we say?

- **Key finding**: The ordering of brackets by arrest rate follows (almost) the expected pattern from the diurnal curve: Morning commute ($1.0\%$), Early morning ($1.4\%$), Midday ($2.6\%$), Afternoon/evening ($3.6\%$), Late night ($4.9\%$), and Night ($5.1\%$). The nighttime brackets show an arrest probability that stands above the global mean, with $19,449$ and $22,763$ arrests.
- **Comment**: Car stops follow the exact same monotonic "risk" curve as 2-wheels stops, even if the values are found in a lower range. As before, the absolute majority of interactions ($94.9\%$) still do not result in an arrest. We can try to argue that, even though the relative risk increases significantly during late hours, a stop ending in an arrest remains a rare event.
- **Proposed Action**: Same as 2-wheels analysis.

## Driver related features

### Reported Age

```{r age-bracket-vs-target}
ratio_by_cat(
  vehstops, arrested, age_bracket,
  "Reported Age Bracket", "Arrest Rate (%)", "Arrest Ratio by Reported Age Bracket for Cars Stopped in NYC",
  style = "stack", legend_title = "Stop ended in arrest?", target_labels = c("No", "Yes"),
  subtitle = "The 18-34 age bracket shows the highest arrest rates",
  label_vjust = 0.5,
)
```

```{r confirm-age-bracket-vs-target-ratio}
vehstops %>%
  group_by(age_bracket, arrested) %>%
  summarise(count = n(), .groups = "drop_last") %>%
  mutate(percentage = count / sum(count))
```

What can we say?

- **Key finding**: This plot shows a clear demographic trend where the arrest rate generally increases as the reported age bracket gets younger, reaching its absolute peak in the $25–34$ label. The lowest arrest rate is found within the $65+$ demographic at $0.66\%$ (specifically $520$ arrests). The highest ratio belongs to the $25–34$ bracket, where the arrest rate reaches $4.57\%$ aka $28,821$ arrests. The $18–24$ group closely follows with a $4.38\%$ arrest rate ($14,380$ arrests). Unlike the 2-wheeled analysis, the under $18$ bracket sits at a lower rate of $2.99\%$ ($278$ arrests), and the Unknown category shows a $1.93\%$ arrest rate ($2291$ arrests).
- **Interpretation**: The data shows a pronounced effect where young adults (ages $18$ to $34$) face the highest probability of arrest upon being stopped in a car. This differs from the previous analysis where minors (Under $18$) showed the highest arrest rate.
- **Comment**: For 2-wheeled vehicles, the younger brackets (especially minors) were heavily penalized, whereas for cars, the risk concentrates heavily on the $25–34$ bracket.
- **Proposed Action**: Same as 2-wheels analysis.

### Reported Sex

```{r sex-vs-target, fig.width=6}
ratio_by_cat(
  vehstops, arrested, sex,
  "Reported sex", "Arrest Rate (%)", "Arrest Ratio by Reported sex for Cars Stopped in NYC",
  style = "stack", legend_title = "Stop ended in arrest?", target_labels = c("No", "Yes"),
  subtitle = "Male drivers are subject to higher arrest rates, followed by Unknown category",
)
```

```{r confirm-sex-vs-target-ratio}
vehstops %>%
  group_by(sex, arrested) %>%
  summarise(count = n(), .groups = "drop_last") %>%
  mutate(percentage = count / sum(count))
```

What can we say?

- **Key finding**: This highlights a clear disparity in arrest rates across reported sex categories for 4-wheeled vehicle stops. Female drivers (`F`) experience the lowest arrest rate at $1.51\%$ ($5966$ arrests), followed by the unknown or unrecorded category (`U`) at $2.06\%$ (with $276$ arrests). Male drivers (`M`) are subject to the highest rate, with $3.70\%$ of their total stops resulting in an arrest ($64,333$).
- **Interpretation**: The ratio of arrest, given a stop, **is more than twice as high for male drivers compared to female** drivers. This variation affects our analysis by showing that there is a consistent and proportional gap between the 2 analyses. This confirms that reported sex remains a key predictor that needs to be included in our statistical models.
- **Comment**: The relative arrest rates are generally lower across all categories for cars than for 2-wheeled vehicles (e.g., $3.70\%$ vs $5.7\%$ for males).
- **Proposed Action**: Same as 2-wheels analysis.

### Reported Ethnicity

```{r ethnicity-vs-target, fig.width=6}
ratio_by_cat(
  vehstops, arrested, ethnicity,
  "Reported ethnicity", "Arrest Rate (%)", "Escalation ratio by Reported Ethnicity for Cars Stopped in NYC",
  style = "stack", legend_title = "Stop ended in arrest?", target_labels = c("No", "Yes"),
  subtitle = "Hispanic and Black Motorists have the highest arrest rates and account for most arrests",
)
```

```{r confirm-ethnicity-vs-target-ratio}
vehstops %>%
  group_by(ethnicity, arrested) %>%
  summarise(count = n(), .groups = "drop_last") %>%
  mutate(percentage = count / sum(count))
```

What can we say?

- **Key finding**: Black drivers represent the largest slice with $656,109$ total stops and the highest absolute number of arrests at $32,954$ ($5.02\%$). In second place, we find Hispanic drivers with $26,330$ arrests out of $565,531$ stops ($4.66\%$). Then, White drivers account for $426,594$ stops and $4912$ arrests ($1.15\%$). Finally, Asian / Pacific Islander, Other, and Unknown categories maintain the lowest arrest rates (all remaining near or below $1.51\%$).
- **Interpretation**: The probability of a stop resulting in an arrest is more than four times higher for Black and Hispanic drivers compared to White drivers. These rates are calculated from a huge sample, meaning that they are stable patterns.
- **Comment**: For 2-wheeled vehicles, Hispanic motorists faced the absolute highest arrest rate ($7.24\%$), instead in our case, Black drivers experience both the highest total stop volume and the highest escalation rate ($5.02\%$).
- **Proposed Action**: Same as 2-wheels analysis.

------------------------------------------------------------------------

# Final touches

## Re-Ordering Features

We re-order the features for better display

```{r re-order-dataset}
vehstops_ord <- vehstops %>%
  select(
    year, month, weekday,
    hour, time_bracket,
    area, boro,
    sex, ethnicity, 
    age, age_bracket,
    arrested
  )
```

Here is the final version.

```{r summary-final-dataset}
print(dfSummary(vehstops_ord), method="render")
```

## Saving the dataset

```{r save-dataset}
write.csv(vehstops_ord, "data/vehiclestops_4_wheels_final.csv", row.names = FALSE)
```
