---
title: "Predicting Excessive Enforcement Patterns in Police Behavior After Vehicles Stops in New York"
subtitle: "EDA and Feature Engineering - 2 Wheels Vehicles"
author: "Gianmaria Pizzo"
output:
  html_document:
    fig_caption: yes
    theme: flatly
    highlight: pygments
    code_folding: show
    kee_md: yes
    toc: yes
    toc_depth: 4
    number_sections: no
    toc_float:
      smooth_scroll: no
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

------------------------------------------------------------------------

# Environment

The first to do, is to build a reproducible environment for our project.

```{r}
requirements=c( 
  "remotes", "summarytools", "forcats", "scales", "purrr",
  "stringr", "plyr", "dbplyr", "dplyr", "dtplyr", "tidyr", "tidyselect",
  "timeDate", "timechange", "tzdb", "hms", "readr",
  "lubridate", "prettyunits", "RColorBrewer", "viridis", "rcartocolor", "ggbeeswarm", "emmeans",
  "sf", "plotly", "ggplot2", "ggmap", "ggdensity", "UpSetR", "patchwork"
)
```


```{r install-requirements, include=FALSE}
for (req in requirements){
  if (!require(req, character.only = TRUE)){
      install.packages(req)
  }
}
```

Note that some of the packages may not be used, but to avoid any error related to dependencies, we are going to keep all of these libraries. Now, the loading of all packages should be immediate.

```{r load-requirements, include=FALSE}
for (req in requirements){
  if (!require(req, character.only = TRUE)){
      library(req)
  }
}
```

```{r remove-vars, include=FALSE}
rm(requirements, req)
```

## Plot Functions 

Here we add the plot functions to make some kinds of visualizations reproducible.

```{r univariate-plot-helpers}
# ==============================================================================
# 1. Plots for Univariate Analysis
# ==============================================================================

#' @title Plot Binary Bar Chart
#' 
#' @description Creates a simple bar chart counting the occurrences of a binary feature.
#' It automatically handles different representations of binary values (1/0, TRUE/FALSE).
#' 
#' @param DF The input dataframe.
#' @param FEATURE The binary column to plot (unquoted).
#' @param OPTION_0 The text label for the "0" or "FALSE" category.
#' @param OPTION_1 The text label for the "1" or "TRUE" category.
#' @param X_LABEL Label for the X axis.
#' @param Y_LABEL Label for the Y axis.
#' @param TITLE Main title of the plot.
#' @param pad Numeric padding for the top of the Y axis to fit text labels.
#' @param text_vjust Vertical adjustment for the text labels on top of bars.
#' 
#' @return A ggplot2 object.
plot_binary_bar <- function(DF, FEATURE, OPTION_0, OPTION_1, X_LABEL, Y_LABEL, TITLE, pad = 0.12, text_vjust = -0.35) {
  # Standardize the binary feature
  df_sum <- DF %>%
    mutate(
      .raw  = {{ FEATURE }},
      # Convert into standard 1L or 0L integers
      .feat = dplyr::case_when(
        as.character(.raw) %in% c("1","TRUE","true")  ~ 1L,
        as.character(.raw) %in% c("0","FALSE","false") ~ 0L,
        TRUE ~ NA_integer_
      )
    ) %>%
    # Remove missing values
    filter(!is.na(.feat)) %>%
    # Count of 0s and 1s
    count(.feat, name = "n") %>%
    # Create factor variable with custom labels for plotting 
    mutate(.lab = factor(.feat, levels = c(0, 1), labels = c(OPTION_0, OPTION_1)))
  
  # Get the maximum bar height
  ymax <- max(df_sum$n, na.rm = TRUE)
  
  # Build the ggplot
  ggplot(df_sum, aes(x = .lab, y = n, fill = .lab)) +
    geom_col(width = 0.8, show.legend = FALSE) +
    # Add count numbers on top of the bars
    geom_text(aes(label = scales::comma(n)), vjust = text_vjust) +
    # Expand the Y axis slightly so the text labels are not cut off
    scale_y_continuous(limits = c(0, ymax * (1 + pad)),
                       expand = expansion(mult = c(0, 0.02))) +
    coord_cartesian(clip = "off") +
    theme(plot.margin = margin(t = 10, r = 5.5, b = 5.5, l = 5.5)) +
    labs(x = X_LABEL, y = Y_LABEL, title = TITLE)
}

# FIXME: Document and comment better
#' @title Plot Univariate Bar Chart (Categorical)
#' 
#' @description Creates a bar chart counting the occurrences of a categorical feature.
#' Allows dynamic ordering (by frequency or original/alphabetical order).
#' 
#' @param DF The input dataframe.
#' @param FEATURE The categorical column to analyze (unquoted).
#' @param X_LABEL Label for the X axis.
#' @param Y_LABEL Label for the Y axis (default is "Count").
#' @param TITLE Main title of the plot.
#' @param top_n Optional integer; if set, keeps only the top N most frequent categories.
#' @param order_by Method to order the X axis: "freq" (descending count) or "none" (alphabetical/original).
#' @param pad Numeric padding for the top of the Y axis to fit text labels.
#' @param text_vjust Vertical adjustment for the text labels on top of bars.
#' 
#' @return A ggplot2 object.
plot_univariate_cat <- function(DF, FEATURE, X_LABEL, Y_LABEL = "Count", TITLE, 
                                top_n = NULL, order_by = c("freq", "none"), 
                                pad = 0.12, text_vjust = -0.35) {
  
  order_by <- match.arg(order_by)
  
  # Step 1: Calcola i conteggi
  df_sum <- DF %>%
    filter(!is.na({{ FEATURE }})) %>%
    count({{ FEATURE }}, name = "n")
  
  # Step 2: Se richiesto 'top_n', mantieni solo i più frequenti
  if (!is.null(top_n)) {
    df_sum <- df_sum %>% slice_max(order_by = n, n = top_n)
  }
  
  # Step 3: Determina l'ordine per l'asse X
  cat_levels <- switch(
    order_by,
    freq = df_sum %>% arrange(desc(n)) %>% pull({{ FEATURE }}),
    none = sort(unique(df_sum %>% pull({{ FEATURE }}))) # Ordine alfabetico / originale
  )
  
  # Applica l'ordine scelto
  df_sum <- df_sum %>%
    mutate(.feat_f = factor(as.character({{ FEATURE }}), levels = as.character(cat_levels)))
  
  # Step 4: Calcola l'altezza massima per il padding dinamico
  ymax <- max(df_sum$n, na.rm = TRUE)

  # Step 5: Costruisci il grafico
  ggplot(df_sum, aes(x = .feat_f, y = n, fill = .feat_f)) +
    geom_col(width = 0.8, show.legend = FALSE) +
    geom_text(aes(label = scales::comma(n)), vjust = text_vjust, size = 3) +
    scale_y_continuous(limits = c(0, ymax * (1 + pad)),
                       expand = expansion(mult = c(0, 0.02))) +
    coord_cartesian(clip = "off") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
          panel.grid.major.x = element_blank(),
          plot.margin = margin(t = 10, r = 5.5, b = 5.5, l = 5.5)) +
    scale_fill_viridis_d(option = "mako", begin = 0.3, end = 0.8) +
    labs(x = X_LABEL, y = Y_LABEL, title = TITLE)
}


#' @title Plot Univariate Histogram
#' 
#' @description Creates a histogram to show the distribution of a continuous 
#' numerical variable.
#' 
#' @param DF The input dataframe.
#' @param FEATURE The numerical column to analyze (unquoted).
#' @param X_LABEL Label for the X axis.
#' @param TITLE Main title of the plot.
#' @param bin_width The width of each bin in the histogram (e.g., 5 for 5-year age groups).
#' 
#' @return A ggplot2 object.
plot_univariate_hist <- function(DF, FEATURE, X_LABEL, TITLE, bin_width = 5) {
  
  ggplot(DF, aes(x = {{ FEATURE }})) +
    # Create the histogram with a specific bin width
    geom_histogram(binwidth = bin_width, fill = "seagreen", color = "white", alpha = 0.8) +
    scale_y_continuous(labels = scales::comma) +
    labs(x = X_LABEL, y = "Frequenza", title = TITLE) +
    theme_minimal() +
    theme(panel.grid.minor = element_blank())
}
```

```{r biviariate-plot-helpers}
# ==============================================================================
# 2. Plots for Bivariate Analysis
# ==============================================================================

# FIXME: Percentage shown in the wrong place.

#' @title Plot Target Ratio by Category
#' 
#' @description Creates a bivariate bar chart showing the relationship between a binary 
#' TARGET and a single categorical variable (CAT). It supports two visual styles: 
#' a 100% stacked bar chart ("stack") or a simple bar chart showing the rate of the target ("rate").
#' 
#' @param DF The input dataframe.
#' @param TARGET The binary column to analyze (unquoted). It will be standardized to 0/1.
#' @param CAT The categorical column for the X axis (unquoted).
#' @param x_title Label for the X axis.
#' @param y_title Label for the Y axis.
#' @param title Main title of the plot.
#' @param target_labels A character vector of length 2 defining the labels for 0 and 1 (e.g., c("Not escalated", "Escalated")).
#' @param style The visual style of the plot: "stack" (100% stacked bars) or "rate" (single bar for target = 1).
#' @param order_by Method to order the X axis categories: by "rate" (ascending target rate), "freq" (ascending frequency), or "none" (alphabetical/original).
#' @param percent_accuracy Accuracy for the percent labels (e.g., 0.1 for one decimal place).
#' @param colors A character vector of length 2 defining the hex color codes for the 0 and 1 categories respectively.
#' 
#' @return A ggplot2 object.
ratio_by_cat <- function(DF, TARGET, CAT, x_title, y_title, title, target_labels = c("Not escalated","Escalated"), style = c("stack","rate"), order_by = c("rate","freq","none"), percent_accuracy = 0.1, colors = c("#00bfc4", "#f8766d")){
  # Validate arguments
  style    <- match.arg(style)
  order_by <- match.arg(order_by)

  # Standardize target and extract the category
  dat <- DF %>%
    mutate(
      .t_raw = {{ TARGET }},
      # Convert into standard 1L or 0L integers
      .t = case_when(
        as.character(.t_raw) %in% c("1","TRUE","true") ~ 1L,
        as.character(.t_raw) %in% c("0","FALSE","false") ~ 0L,
        TRUE ~ NA_integer_
      ),
      .cat = {{ CAT }}
    ) %>%
    # Filter NAs
    filter(!is.na(.t), !is.na(.cat))

  # Compute rates and frequencies
  rates <- dat %>% 
    group_by(.cat) %>%
    summarise(total = n(), rate = mean(.t == 1), .groups = "drop")

  # Determine the ordering for X axis
  cat_levels <- switch(
    order_by,
    rate = rates %>% arrange(rate) %>% pull(.cat),
    freq = rates %>% arrange(total) %>% pull(.cat),
    none = unique(dat$.cat)
  )

  # Build the plot based on the chosen 'style'
  if (style == "stack") {
    # Prepare data for stacked bar chart
    stack <- dat %>%
      count(.cat, .t, name = "n") %>%
      group_by(.cat) %>%
      mutate(ratio = n/sum(n)) %>%
      ungroup() %>%
      mutate(
        # Apply the ordering
        .cat_f = factor(as.character(.cat), levels = as.character(cat_levels)),
        # Apply custom labels
        .t_f = factor(.t, levels = c(0,1), labels = target_labels)
      ) %>%
      # Join with the rates
      left_join(rates, by = ".cat")
    
    # Build the stacked plot
    ggplot(stack, aes(x = .cat_f, y = ratio, fill = .t_f)) +
      # Forces the bars to stack to 100%
      geom_col(position = "fill") +
      # Add exact percentage labels (only for the TARGET == 1 category)
      geom_text(
        data = subset(stack, .t_f == target_labels[2]),
        aes(label = scales::percent(rate, accuracy = percent_accuracy)),
        position = position_fill(vjust = 0.5), 
        size = 3
      ) +
      # Make continuous
      scale_y_continuous(labels = scales::percent) +
      # Map colors
      scale_fill_manual(values = setNames(colors, target_labels)) +
      labs(x = x_title, y = y_title, fill = NULL, title = title) +
      theme_minimal() +
      theme(axis.text.x = element_text(angle = 45, hjust = 1),
            panel.grid = element_blank())
  } else {
    # Prepare for rate bar chart (bars represent % of TARGET == 1)
    rates <- rates %>%
      mutate(
        .cat_f = factor(as.character(.cat),levels = as.character(cat_levels))
      )

    # Build the rate plot
    ggplot(rates, aes(x = .cat_f, y = rate)) +
      # Use second color to represent TARGET == 1
      geom_col(width = 0.8, fill = colors[2]) +
      # Add percentage labels
      geom_text(
        aes(label = scales::percent(rate, accuracy = percent_accuracy)),
        vjust = -0.35, size = 3
      ) +
      # Make continuous and expand
      scale_y_continuous(labels = scales::percent, expand = expansion(mult = c(0, 0.12))) +
      coord_cartesian(clip = "off") +
      labs(x = x_title, y = y_title, title = title,
           subtitle = "Bars show % of target = 1") +
      theme_minimal() +
      theme(axis.text.x = element_text(angle = 45, hjust = 1),
            panel.grid = element_blank())
  }
}

```

```{r multivariate-plot-helpers}
# ==============================================================================
# 3. Plots for Multivariate Analysis
# ==============================================================================

#' @title Heatmap of Target Rate (Categorical Y vs Numeric X)
#' 
#' @description Plots a heatmap showing the rate of a binary TARGET across a categorical
#' Y variable and a numeric (or factor) X variable.
#' 
#' @param DF The input dataframe.
#' @param TARGET The binary column to calculate the rate for (unquoted).
#' @param Y Categorical variable for the Y axis.
#' @param X Numeric or categorical variable for the X axis.
#' @param X_TITLE Label for the X axis.
#' @param Y_TITLE Label for the Y axis.
#' @param TITLE Main title of the plot.
#' @param min_n Minimum number of observations required to show a tile.
#' @param drop_na Boolean; if TRUE, rows with NA in X or Y are removed.
#' @param x_as_factor Boolean; if TRUE, forces the X variable to be treated as a categorical factor.
#' 
#' @return A ggplot2 object.
heat_target_yx <- function(DF, TARGET, Y, X, X_TITLE, Y_TITLE, TITLE, min_n = 0, drop_na = TRUE, x_as_factor = FALSE) {
  # Standardize the binary target
  dat <- DF %>%
    mutate(
      .target_raw = {{ TARGET }},
      .target_bin = dplyr::case_when(
        as.character(.target_raw) %in% c("1", "TRUE", "true")  ~ 1L,
        as.character(.target_raw) %in% c("0", "FALSE", "false") ~ 0L,
        TRUE ~ NA_integer_
      )
    )

  # Filter NAs
  if (drop_na) {
    dat <- dat %>% filter(!is.na({{Y}}), !is.na({{X}}), !is.na(.target_bin))
  } else {
    dat <- dat %>% filter(!is.na(.target_bin))
  }

  # Incrementally sort Y axis based on the mean ratio of the target
  y_order <- dat %>%
    group_by({{Y}}) %>%
    summarise(rate = mean(.target_bin == 1, na.rm = TRUE), .groups = "drop") %>%
    arrange(rate) %>%
    pull({{Y}})

  # Compute the target rate for each combination of Y x X
  grid <- dat %>%
    group_by({{Y}}, {{X}}) %>%
    summarise(
      rate = mean(.target_bin == 1, na.rm = TRUE),
      n = n(), .groups = "drop"
    ) %>%
    mutate(
      .y = factor(as.character({{Y}}), levels = as.character(y_order)),
      .x = if (x_as_factor) factor({{X}}) else {{X}},
      # Hide cells that do not reach the minimum amount of observations
      rate_plot = ifelse(n >= min_n, rate, NA_real_)
    )

  # Build ggplot heatmap
  ggplot(grid, aes(.x, .y, fill = rate_plot)) +
    geom_tile(width = 1, height = 0.9) +
    scale_fill_viridis_c(labels = scales::percent, na.value = "grey90", name = "Rate") +
    labs(x = X_TITLE, y = Y_TITLE, title = TITLE) +
    theme_minimal() +
    theme(panel.grid = element_blank())
}


#' @title Heatmap of Target Rate (Categorical Y vs Categorical X)
#' 
#' @description Plots a heatmap showing the rate of a binary TARGET across two categorical
#' variables (Y and X). Allows dynamic ordering and shading based on sample size.
#' 
#' @param DF The input dataframe.
#' @param TARGET The binary column to calculate the rate for (unquoted).
#' @param Y Categorical variable for the Y axis.
#' @param X Categorical variable for the X axis.
#' @param X_TITLE Label for the X axis.
#' @param Y_TITLE Label for the Y axis.
#' @param TITLE Main title of the plot.
#' @param min_n Minimum number of observations required to show a tile (filters noise).
#' @param order_y_by Method to order the Y axis: by 'rate', 'freq' (frequency), or 'none'.
#' @param order_x_by Method to order the X axis: by 'rate', 'freq' (frequency), or 'none'.
#' @param alpha_by_n Boolean; if TRUE, shades tiles by count using alpha (transparency).
#' 
#' @return A ggplot2 object.
heat_target_catcat <- function(DF, TARGET, Y, X, X_TITLE, Y_TITLE, TITLE, min_n = 0, order_y_by = c("rate","freq","none"), order_x_by = c("rate","freq","none"), alpha_by_n = FALSE ) {
  order_y_by <- match.arg(order_y_by)
  order_x_by <- match.arg(order_x_by)

  # Standardize target
  dat <- DF %>%
    mutate(
      .target_raw = {{ TARGET }},
      .target_bin = dplyr::case_when(
        as.character(.target_raw) %in% c("1", "TRUE", "true")  ~ 1L,
        as.character(.target_raw) %in% c("0", "FALSE", "false") ~ 0L,
        TRUE ~ NA_integer_
      )
    ) %>%
    # Filter NAs
    filter(!is.na({{Y}}), !is.na({{X}}), !is.na(.target_bin))

  # Compute stats to determine the Y axis order
  y_stats <- dat %>%
    group_by({{Y}}) %>%
    summarise(
      rate = mean(.target_bin == 1, na.rm = TRUE),
      n = n(), .groups = "drop"
    )
  
  # Compute stats to determine the X axis order
  x_stats <- dat %>%
    group_by({{X}}) %>%
    summarise(
      rate = mean(.target_bin == 1, na.rm = TRUE), 
      n = n(), .groups = "drop"
    )

  # Apply chosen "order by" to Y
  y_levels <- switch(
    order_y_by,
    rate = y_stats %>% arrange(rate) %>% pull({{Y}}),
    freq = y_stats %>% arrange(n) %>% pull({{Y}}),
    none = unique(dat %>% pull({{Y}}))
  )
  
  # Apply chosen "order by" to X
  x_levels <- switch(
    order_x_by,
    rate = x_stats %>% arrange(rate) %>% pull({{X}}),
    freq = x_stats %>% arrange(n) %>% pull({{X}}),
    none = unique(dat %>% pull({{X}}))
  )

  # Compute the cell's target rate and counts for each combination of Y x X
  grid <- dat %>%
    group_by({{Y}}, {{X}}) %>%
    summarise(
      rate = mean(.target_bin == 1, na.rm = TRUE),
      n = n(), .groups = "drop"
    ) %>%
    mutate(
      .y = factor(as.character({{Y}}), levels = as.character(y_levels)),
      .x = factor(as.character({{X}}), levels = as.character(x_levels)),
      rate_plot = ifelse(n >= min_n, rate, NA_real_)
    )

  # Build the plot
  p <- ggplot(grid, aes(.x, .y)) +
    labs(x = X_TITLE, y = Y_TITLE, title = TITLE) +
    theme_minimal() +
    theme(panel.grid = element_blank()) +
    # Use scales for displaying rates
    scale_fill_viridis_c(labels = scales::percent, na.value = "grey90", name = "Rate")

  if (alpha_by_n) {
    p <- p +
      geom_tile(aes(fill = rate_plot, alpha = n), width = 0.95, height = 0.95) +
      scale_alpha_continuous(range = c(0.25, 1), name = "Stops")
  } else {
    p <- p +
      geom_tile(aes(fill = rate_plot), width = 0.95, height = 0.95)
  }

  p
}


#' @title Stacked Ratio Bars Split by a Third Variable
#' 
#' @description Creates a faceted 100% stacked bar chart. This is highly effective 
#' for visualizing how the rate of a binary TARGET changes across a primary category (CAT), 
#' while simultaneously splitting the data by a third variable (SPLIT) to create side-by-side plots.
#' 
#' @param DF The input dataframe.
#' @param TARGET The binary column to analyze (unquoted). It will be standardized to 0/1.
#' @param CAT The main categorical variable for the X axis (unquoted).
#' @param SPLIT The categorical variable used to split/facet the plots (unquoted).
#' @param x_title Label for the X axis.
#' @param y_title Label for the Y axis.
#' @param title Main title of the entire plot.
#' @param target_labels Character vector of length 2 defining labels for the binary outcomes (e.g., c("Not escalated", "Escalated")).
#' @param split_labels Optional character vector to provide custom titles for the split panels (facets).
#' @param palette A character string specifying an RColorBrewer palette name (e.g., "Paired", "Set1").
#' @param percent_accuracy Accuracy for the percentage labels (e.g., 0.1 for one decimal place).
#' 
#' @return A ggplot2 object.
ratio_by_cat_split <- function(DF, TARGET, CAT, SPLIT, x_title, y_title, title, target_labels = c("Not escalated", "Escalated"), split_labels = NULL, palette = "Paired", percent_accuracy = 0.1){
  # Standardize target
  dat <- DF %>%
    mutate(
      .target_raw = {{ TARGET }},
      # Conversion to 0/1
      .target_bin = case_when(
        as.character(.target_raw) %in% c("1", "TRUE", "true") ~ 1L,
        as.character(.target_raw) %in% c("0", "FALSE", "false") ~ 0L,
        TRUE ~ NA_integer_
      ),
      .cat = {{ CAT }},
      .split = {{ SPLIT }}
    ) %>%
    # Filter NAs
    filter(!is.na(.cat), !is.na(.split), !is.na(.target_bin))

  # Order the main categories based on the overall target rate (from lowest to highest)
  cat_order <- dat %>%
    group_by(.cat) %>%
    summarise(
      rate = mean(.target_bin == 1),
      .groups = "drop"
    ) %>%
    arrange(rate) %>% 
    pull(.cat)

  # Tidy labels by converting character to factors
  dat <- dat %>%
    mutate(
      # Apply the order to X axis
      .cat = factor(as.character(.cat), levels = as.character(cat_order)),
      # Apply custom labels to the target
      .target_f = factor(.target_bin, levels = c(0, 1), labels = target_labels),
      # Format the split variable
      .split_f = {
        lv <- unique(as.character(.split))
        # If custom split labels are provided and the length matches, use them
        if (!is.null(split_labels) && length(split_labels) == length(lv))
          factor(as.character(.split), levels = lv, labels = split_labels)
        else factor(as.character(.split), levels = lv)
      }
    )

  # Compute rates
  plot_df <- dat %>%
    count(.split_f, .cat, .target_f, name = "n") %>%
    group_by(.split_f, .cat) %>%
    mutate(ratio = n/sum(n)) %>%
    ungroup()

  # Build plot
  ggplot(plot_df, aes(x = .cat, y = ratio, fill = .target_f)) +
    # Stacks the bars to exactly 100%
    geom_col(position = "fill") +
    # Add ratio text labels
    geom_text(
      aes(label = percent(ratio, accuracy = percent_accuracy)),
      position = position_fill(vjust = 0.5), 
      size = 3
    ) +
    # Format Y axis as percentages and make continuous
    scale_y_continuous(labels = percent) +
    # Use a clean palette for the target categories
    scale_fill_brewer(palette = palette, name = rlang::as_name(rlang::enquo(TARGET))) +
    labs(x = x_title, y = y_title, title = title) +
    # Split the plot into separate panels based on the .split_f variable
    facet_wrap(~ .split_f, nrow = 1) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
          panel.grid = element_blank())
}
```

---

Let us load the data now, and remove the extra feature `veh_category` because we are **exclusively analyzing 2 wheels vehicles now**.

```{r load-data-drop-veh-categ}
vehstops = read.csv("data/vehiclestops_2_wheels.csv")

vehstops <- vehstops %>%
  select(-veh_categ)
```

So now we have the following dataset, that we present with the usual summary

```{r summary-data}
print(dfSummary(vehstops), method="render")
```

This preview is already interesting.

---

# Univariate EDA: Distribution of Targets and Predictors

## Targets


## Command Unit, Geospatial and Area-related predictors

### Area

### Borough

### Command Unit

### Latitude and Longitude


## Time-related predictors

### Year

### Month

### Day, Doy and Dow

### Hour, Working Hour and Time Bracket

### Weekend, Holiday and Event Day


## Motorist related predictors

### Reported Age and Age Bracket

### Reported Sex

### Reported Race

---

# Bivariate EDA: Targets vs Predictors

## Correlation

TODO: If there i

## Collinearity

## Arrested vs Vehicle Checkpoint

Sarebbe utilissimo incrociarle! La polizia effettua più arresti durante i normali stop di pattuglia o durante i posti di blocco?
Puoi usare la funzione ratio_by_cat che abbiamo creato prima impostando TARGET = arrested e CAT = veh_checkpoint.

## Command Unit, Geospatial and Area-related predictors


## Time-related predictors


## Motorist related predictors


---

# Multivariate EDA

## Multicollinearity




