---
title: "Predicting Excessive Enforcement Patterns in Police Behavior After Vehicles Stops in New York"
subtitle: "EDA and Feature Engineering - 4 Wheels Vehicles"
author: "Gianmaria Pizzo"
output:
  html_document:
    fig_caption: yes
    theme: flatly
    highlight: pygments
    code_folding: show
    kee_md: yes
    toc: yes
    toc_depth: 4
    number_sections: no
    toc_float:
      smooth_scroll: no
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

These `.Rmd` files represent the project submission for the course [Statistical Inference and Learning](https://www.unive.it/data/course/521944) by Professor [Cristiano Varin](https://www.unive.it/data/people/5592728) at [Ca' Foscari University of Venice](https://www.unive.it/) for the year 2025/2026.

# Problem Statement

## Goal of this rmarkdown file

This notebook serves as the primary Exploratory Data Analysis (EDA) and (final) Feature Engineering phase for the project. In particular, we focus solely on **4-Wheeled Vehicles**.

The goal is to identify the features that characterize **arrests**.

We approach the dataset through three methodological phases to ensure a robust base for modeling:

1.  **Univariate Analysis**. We begin by analyzing the distributions of individual predictors to understand whether the data shows quality issues, outliers, rare categories, etc...

2.  **Bivariate Analysis**. We evaluate the relationship between potential predictors and our binary target (`arrested`). Given the categorical nature of our data, we will mainly uses **Cramer’s V** as a starting point to measure of effect size. This stage should also point out if predictors have some kind of relation.

Under these premises, we will plot many aspects of the data. Therefore, the way we decided to guide the reader through the Exploratory Data Analysis, is based on four key points that we try to include for any plot:

- **Key finding**: This paragraph describes the main takeaway of the means of visualization employed. This should be a concise description of what we can gather from the plots.
- **Interpretation**: Here we deal with an objective interpretation of our finding, such as understanding how our EDA is affected by it.
- **Comment**: After a formal description, we deal with a more informal discussion that should highlight the qualitative aspects of the analysis. This paragraph is a fusion of previous considerations that we want the reader to be aware of, and a space for us to reason.
- **Proposed Action**: Finally, we switch to a more pragmatic discussion to understand "what to do next", on the basis of what was previously presented. Without claim of completeness, we will name some best-practices, guidelines or ideas about what is best for our goal. We may also limit ourselves to inform the reader that we decided to take no action at all or post-pone it after further investigation.

Note that we do not always include all four points.

------------------------------------------------------------------------

# Environment

The first to do, is to build a reproducible environment for our project.

```{r}
requirements=c( 
  "remotes", "summarytools", "forcats", "scales", "purrr",
  "stringr", "plyr", "dbplyr", "dplyr", "dtplyr", "tidyr", "tidyselect",
  "timeDate", "timechange", "tzdb", "hms", "readr", "tigris", "lsr", "corrplot",
  "lubridate", "prettyunits", "RColorBrewer", "viridis", "rcartocolor", "ggbeeswarm", "emmeans",
  "sf", "plotly", "ggplot2", "ggmap", "ggdensity", "UpSetR", "patchwork"
)
```

```{r install-requirements, include=FALSE}
for (req in requirements){
  if (!require(req, character.only = TRUE)){
      install.packages(req)
  }
}
```

Note that some of the packages may not be used, but to avoid any error related to dependencies, we are going to keep all of these libraries. Now, the loading of all packages should be immediate.

```{r load-requirements, include=FALSE}
for (req in requirements){
  if (!require(req, character.only = TRUE)){
      library(req, character.only = TRUE)
  }
}
```

```{r remove-vars, include=FALSE}
rm(requirements, req)
```

## Plot Functions

Here we add the plot functions to make some kinds of visualizations reproducible.

### Univariate Plot Functions

```{r univariate-plot-helpers}
# ==============================================================================
# 1. Plots for Univariate Analysis
# ==============================================================================

#' @title Plot Binary Bar Chart
#' 
#' @description Creates a simple bar chart counting the occurrences of a binary feature, 
#' with labels showing both the count and the percentage.
#' 
#' @param DF The input dataframe.
#' @param FEATURE The binary column to plot (unquoted).
#' @param OPTION_0 The text label for the "0" or "FALSE" category.
#' @param OPTION_1 The text label for the "1" or "TRUE" category.
#' @param X_LABEL Label for the X axis.
#' @param Y_LABEL Label for the Y axis.
#' @param TITLE Main title of the plot.
#' @param pad Numeric padding for the top of the Y axis to fit text labels.
#' @param text_vjust Vertical adjustment for the text labels on top of bars.
#' 
#' @return A ggplot2 object.
plot_binary_bar <- function(DF, FEATURE, OPTION_0, OPTION_1, X_LABEL, Y_LABEL, TITLE, pad = 0.15, text_vjust = -0.3) {
  # Standardize the binary feature and calculate percentages
  df_sum <- DF %>%
    mutate(
      .raw  = {{ FEATURE }},
      .feat = dplyr::case_when(
        as.character(.raw) %in% c("1","TRUE","true")  ~ 1L,
        as.character(.raw) %in% c("0","FALSE","false") ~ 0L,
        TRUE ~ NA_integer_
      )
    ) %>%
    filter(!is.na(.feat)) %>%
    count(.feat, name = "n") %>%
    mutate(
      .lab = factor(.feat, levels = c(0, 1), labels = c(OPTION_0, OPTION_1)),
      # Calculate percentage and create combined label
      pct = n / sum(n),
      label_text = paste0(scales::comma(n), "\n(", scales::percent(pct, accuracy = 0.1), ")")
    )
  
  # Get the maximum bar height
  ymax <- max(df_sum$n, na.rm = TRUE)
  
  # Build the ggplot
  ggplot(df_sum, aes(x = .lab, y = n, fill = .lab)) +
    geom_col(width = 0.8, show.legend = FALSE) +
    # Updated to use label_text
    geom_text(aes(label = label_text), vjust = text_vjust, lineheight = 0.8) +
    # Expand the Y axis slightly for the taller multi-line labels
    scale_y_continuous(limits = c(0, ymax * (1 + pad)), expand = expansion(mult = c(0, 0.02))) +
    coord_cartesian(clip = "off") +
    theme_minimal() +
    theme(panel.grid.major.x = element_blank(), plot.margin = margin(t = 20, r = 5.5, b = 5.5, l = 5.5)) +
    scale_fill_viridis_d(option = "mako", begin = 0.3, end = 0.8) +
    labs(x = X_LABEL, y = Y_LABEL, title = TITLE)
}


#' @title Plot Univariate Bar Chart (Categorical)
#' 
#' @description Creates a bar chart counting the occurrences of a categorical feature with 2+ classes.
#' Allows dynamic ordering (by frequency or original/alphabetical order).
#' 
#' @param DF The input dataframe.
#' @param FEATURE The categorical column to analyze (unquoted).
#' @param X_LABEL Label for the X axis.
#' @param Y_LABEL Label for the Y axis (default is "Count of Stops").
#' @param TITLE Main title of the plot.
#' @param SUBTITLE Subtitle of the plot.
#' @param top_n Optional integer; if set, keeps only the top N most frequent categories.
#' @param order_by Method to order the X axis: "freq" (descending count) or "none" (alphabetical/original).
#' @param pad Numeric padding for the top of the Y axis to fit text labels.
#' @param text_vjust Vertical adjustment for the text labels on top of bars.
#' 
#' @return A ggplot2 object.
plot_univariate_cat <- function(DF, FEATURE, X_LABEL, Y_LABEL = "Count of Stops", TITLE, SUBTITLE, top_n = NULL, order_by = c("freq", "none"), pad = 0.12, text_vjust = -0.35) {
  
  order_by <- match.arg(order_by)
  
  # Compute counts
  df_sum <- DF %>%
    filter(!is.na({{ FEATURE }})) %>%
    count({{ FEATURE }}, name = "n")
  
  # Filter top n results
  if (!is.null(top_n)) {
    df_sum <- df_sum %>% slice_max(order_by = n, n = top_n)
  }
  
  # X axis order (either frequency or std)
  cat_levels <- switch(
    order_by,
    freq = df_sum %>% arrange(desc(n)) %>% pull({{ FEATURE }}),
    none = sort(unique(df_sum %>% pull({{ FEATURE }})))
  )
  
  # Apply sorting
  df_sum <- df_sum %>%
    mutate(.feat_f = factor(as.character({{ FEATURE }}), levels = as.character(cat_levels)))
  
  # Max height for dynamic padding
  ymax <- max(df_sum$n, na.rm = TRUE)

  # Construct plot
  ggplot(df_sum, aes(x = .feat_f, y = n, fill = .feat_f)) +
    geom_col(width = 0.8, show.legend = FALSE) +
    geom_text(aes(label = scales::comma(n)), vjust = text_vjust, size = 3) +
    scale_y_continuous(limits = c(0, ymax * (1 + pad)),
                       expand = expansion(mult = c(0, 0.02))) +
    coord_cartesian(clip = "off") +
    theme_minimal() +
    theme(
      plot.title = element_text(size = 14),
      plot.subtitle = element_text(size = 10, color = "grey50"),
      axis.text.x = element_text(angle = 45, hjust = 1),
      panel.grid.major.x = element_blank(),
      plot.margin = margin(t = 10, r = 5.5, b = 5.5, l = 5.5)) +
    scale_fill_viridis_d(option = "mako", begin = 0.3, end = 0.8) +
    labs(x = X_LABEL, y = Y_LABEL, title = TITLE, subtitle = SUBTITLE)
}
```

### Bivariate Plot Functions

```{r biviariate-plot-helpers}
# ==============================================================================
# 2. Plots for Bivariate Analysis
# ==============================================================================

#' @title Calculate Cramer's V
#' 
#' @description Calculates the Cramer's V measure of association between two 
#' categorical vectors. Returns a value between 0 (no association) and 1 (perfect association).
#' 
#' @param x First categorical vector.
#' @param y Second categorical vector.
#' 
#' @return A numeric value representing Cramer's V.
cramer_v <- function(x, y) {
  # Contingency Table
  tbl <- table(x, y)
  
  # Execute Chi2 Test
  chi2 <- suppressWarnings(chisq.test(tbl, correct = FALSE)$statistic)
  
  # Compute dim and min value
  n <- sum(tbl)
  min_dim <- min(nrow(tbl) - 1, ncol(tbl) - 1)
  
  # Avoid math error for 1 categ
  if(min_dim == 0) return(0) 
  
  # Cramer's formula
  v <- sqrt(chi2 / (n * min_dim))
  return(as.numeric(v))
}

# ==============================================================================
# Categorical "Correlation" Heatmap - Association Matrix based on Cramer's V
# ==============================================================================

#' @title Plot Categorical Correlation Matrix (Cramer's V)
#' 
#' @description Creates a correlation-like heatmap for multiple categorical variables
#' using Cramer's V. Answers "How much do these categorical features overlap with each other?"
#' 
#' @param DF The input dataframe.
#' @param FEATURES A character vector containing the names of the categorical columns to compare.
#' @param TITLE Main title of the plot.
#' 
#' @return A ggplot2 object.
plot_categorical_correlation <- function(DF, FEATURES, TITLE = "Association Matrix for Categorical Predictors (Cramer's V)", text_size = 3) {
  
  # Select columns and drop NAs
  df_sub <- DF %>% 
    select(all_of(FEATURES)) %>%
    drop_na()
  
  n_vars <- ncol(df_sub)
  
  # Initialize matrix and set main diag to 1s
  mat <- matrix(1, nrow = n_vars, ncol = n_vars)
  rownames(mat) <- colnames(df_sub)
  colnames(mat) <- colnames(df_sub)

  # Compute Cramer's V for each possible pair
  for(i in 1:(n_vars - 1)) {
    for(j in (i + 1):n_vars) {
      v <- cramer_v(df_sub[[i]], df_sub[[j]])
      # Symmetric 
      mat[i, j] <- v
      mat[j, i] <- v
    }
  }
  
  # Make matrix compatible with ggplot2
  mat_long <- as.data.frame(as.table(mat))
  colnames(mat_long) <- c("Var1", "Var2", "CramerV")
  
  # Create heatmap
  ggplot(mat_long, aes(x = Var1, y = Var2, fill = CramerV)) +
    geom_tile(color = "white", width = 1, height = 1) +
    # Rounding
    geom_text(aes(label = round(CramerV, 2)), size = text_size) +
    # Setting colors
    scale_fill_gradient(low = "whitesmoke", high = "steelblue", limits = c(0, 1), name = "Cramer's V") +
    labs(title = TITLE, x = "", y = "") +
    theme_minimal() +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1, size = 10),
      axis.text.y = element_text(size = 10),
      panel.grid = element_blank()
    )
}

# ==============================================================================
# Plotting of Targets' Rate Against Categorical Predictors
# ==============================================================================

#' @title Plot Target Ratio by Category
#' 
#' @description Creates a bivariate bar chart showing the relationship between a binary 
#' TARGET and a single categorical variable (CAT). It supports two visual styles: 
#' a 100% stacked bar chart ("stack") or a simple bar chart showing the rate of the target ("rate").
#' 
#' @param DF The input dataframe.
#' @param TARGET The binary column to analyze (unquoted). It will be standardized to 0/1.
#' @param CAT The categorical column for the X axis (unquoted).
#' @param x_title Label for the X axis.
#' @param y_title Label for the Y axis.
#' @param title Main title of the plot.
#' @param target_labels A character vector of length 2 defining the labels for 0 and 1 (e.g., c("Not escalated", "Escalated")).
#' @param style The visual style of the plot: "stack" (100% stacked bars) or "rate" (single bar for target = 1).
#' @param order_by Method to order the X axis categories: by "rate" (ascending target rate), "freq" (ascending frequency), or "none" (alphabetical/original).
#' @param percent_accuracy Accuracy for the percent labels (e.g., 0.1 for one decimal place).
#' @param colors A character vector of length 2 defining the hex color codes for the 0 and 1 categories respectively.
#' 
#' @return A ggplot2 object.
ratio_by_cat <- function(
    DF, TARGET, CAT, x_title, y_title, title, 
    target_labels = c("Not escalated", "Escalated"), 
    style = c("stack", "rate"), order_by = c("rate", "freq", "none"), 
    percent_accuracy = 0.1, palette = "Paired",
    colors = c("grey80", "steelblue"), legend_title = "Legend",
    subtitle = NULL, bar_width = 0.7, aspect_ratio = 1.0, label_vjust = 0.5,
    text_size = 2
) {
  
  # Validate arguments
  style    <- match.arg(style)
  order_by <- match.arg(order_by)
  
  # Standardize target and extract the category
  dat <- DF %>%
    mutate(
      .t_raw = {{ TARGET }},
      .t = case_when(
        as.character(.t_raw) %in% c("1","TRUE","true") ~ 1L,
        as.character(.t_raw) %in% c("0","FALSE","false") ~ 0L,
        TRUE ~ NA_integer_
      ),
      .cat = {{ CAT }}
    ) %>%
    filter(!is.na(.t), !is.na(.cat))
  
  # Compute rates and frequencies
  rates <- dat %>% 
    group_by(.cat) %>%
    summarise(total = n(), rate = mean(.t == 1), .groups = "drop")
  
  # Determine ordering
  cat_levels <- switch(
    order_by,
    rate = rates %>% arrange(rate) %>% pull(.cat),
    freq = rates %>% arrange(total) %>% pull(.cat),
    none = unique(dat$.cat)
  )
  
  # Build the plot based on style
  if (style == "stack") {
    stack <- dat %>%
      count(.cat, .t, name = "n") %>%
      group_by(.cat) %>%
      mutate(ratio = n/sum(n)) %>%
      ungroup() %>%
      mutate(
        .cat_f = factor(as.character(.cat), levels = as.character(cat_levels)),
        .t_f = factor(.t, levels = c(0,1), labels = target_labels)
      )
    
    ggplot(stack, aes(x = .cat_f, y = ratio, fill = .t_f)) +
      geom_col(position = "fill", width = bar_width) +
      geom_text(
        aes(label = scales::percent(ratio, accuracy = percent_accuracy)),
        position = position_fill(vjust = label_vjust), 
        size = text_size
      ) +
      scale_y_continuous(labels = scales::percent) +
      scale_fill_manual(values = setNames(colors, target_labels)) +
      labs(x = x_title, y = y_title, fill = legend_title, title = title, subtitle = subtitle) +
      theme_minimal() +
      theme(
        axis.text.x = element_text(angle = 45, hjust = 1),
        panel.grid = element_blank(),
        plot.subtitle = element_text(color = "grey50"),
        aspect.ratio = aspect_ratio
      )
      
  } else {
    # Rate style
    rates_data <- rates %>%
      mutate(.cat_f = factor(as.character(.cat), levels = as.character(cat_levels)))
    
    ggplot(rates_data, aes(x = .cat_f, y = rate)) +
      geom_col(width = 0.8, fill = colors[2]) +
      geom_text(
        aes(label = scales::percent(rate, accuracy = percent_accuracy)),
        vjust = -0.35, size = text_size
      ) +
      scale_y_continuous(labels = scales::percent, expand = expansion(mult = c(0, 0.12))) +
      coord_cartesian(clip = "off") +
      labs(x = x_title, y = y_title, title = title, subtitle = subtitle) +
      theme_minimal() +
      theme(
        axis.text.x = element_text(angle = 45, hjust = 1),
        panel.grid = element_blank(),
        plot.subtitle = element_text(color = "grey50")
      )
  }
}
```

------------------------------------------------------------------------

Let us load the data now, and remove the extra feature `veh_category` because we are **exclusively analyzing 4 wheels vehicles now**.

```{r load-data-drop-veh-categ}
vehstops = read.csv("data/vehiclestops_4_wheels.csv")

vehstops <- vehstops %>%
  select(-veh_categ)
```

So now we have the following dataset, that we present with the usual summary

```{r summary-data}
print(dfSummary(vehstops), method="render")
```

This preview is already interesting.

------------------------------------------------------------------------

# Univariate EDA

## Target

```{r enforcement-dist}
p1 <- plot_binary_bar(
  vehstops, arrested, 
  "No", "Yes",
  "Was the driver arrested?", "Count of Stops (Percentage %)", ""
)

row1 <- p1

row1 + 
plot_annotation(title = "Distribution of Car Stops by Arrested") &
theme(legend.position = "bottom") 
```

```{r, include=FALSE}
rm(p1, row1)
```

What can we say?

What can we say?

- **Key finding**: The dataset for 4-wheels vehicles **exhibits extreme class imbalance** that is even more pronounced than what we observed in the 2-wheels analysis. Arrests are highly sparse, representing only approximately $3.3\%$ of the outcomes ($70,600$ "Yes" out of $2,157,445$ total records) compared to the $5.6\%$ rate found in 2-wheeled vehicle stops.
- **Interpretation**: The target variable constitutes a rare event when we contemplate the stops of 4-wheels vehicles. While the overall volume of stops is drastically higher for 4-wheels vehicles, an arrest remains an infrequent occurrence with respect to the total volume of police stops.
- **Comment**: The high class disparity *makes accuracy a limiting metric* for our models' evaluation. This is due to the fact a predictive model aiming for accuracy could achieve a high score of $96.7\%$ by simply classifying every stop as "No". Therefore, it completely fails to capture what characterizes an arrest after a stop. This is also sub-optimal for our research, as capturing the true patterns is what drives our investigation towards enforcement bias. The fact that the positive class is even rarer here means our models will face a higher risk of failing to learn the underlying patterns of the minority class.
- **Proposed Action**: Although we could employ cost-sensitive learning models, this is outside the scope of the course. As an alternative, we will **employ stratified down-sampling of the majority class** ("No") to create a balanced training set, ensuring compatibility across all models (LDA, QDA, KNN, etc...). For the **GLM-based models** (Logistic, Lasso, Ridge), we will specifically **utilize observation weighting within the likelihood function** to account for the class disparity. Finally, we will move away from the default $0.5$ classification threshold, **optimizing the decision boundary based on the Precision-Recall curve** to ensure our model maintains sensitivity to the minority class ("Yes").

## Geospatial predictors

### Area

```{r count-stops-by-area}
plot_univariate_cat(
  DF = vehstops, 
  FEATURE = area, 
  X_LABEL = "Area", 
  TITLE = "Distribution of Car Stops by Area (Sorted by Count Desc)",
  SUBTITLE = "Brooklyn is where the most amount of vehicles are stopped",
  order_by = "freq"
)
```

What can we say?

- **Key finding**: There is a high variance in the volume of stops across areas, though the overall scale is massively larger for 4-wheels vehicles. We observe two clear "high-volume" areas (Brooklyn with over $640$k stops and Queens with over $550$k stops) followed by two "medium-volume" areas (Manhattan with $418$k and the Bronx with $381$k) and a distinct "low-volume" area (Staten Island, $166$k).
- **Interpretation**: The distribution is highly non-uniform, or better, uneven. However, compared to the 2-wheels analysis where Staten Island was a tiny outlier ($1.5$k), here it holds a substantial sample size ($166$k). The shifting proportions, particularly the rise of the Bronx into a solid medium-volume category, suggests that geographic location could be a strong predictor of the probability of being stopped (in case the targets' ratio distributed by area follows a specific pattern).
- **Comment**: From a modeling perspective, this introduces a risk of **overfitting to the high-volume areas**. Since some of the available models (especially KNN or QDA) rely on local neighborhood data or global likelihoods, they will naturally prioritize minimizing errors in the areas with the most observations. Unlike the 2-wheels context where the extreme sparsity of Staten Island risked high variance in parameter estimates, the 4-wheels data provides ample observations across all levels, which will stabilize our coefficients. The visualization succeeds in highlighting a pattern, but it leaves the causal mechanisms open to speculation.
- **Proposed Action**: We still need to check its association with subject-related predictors. Also, we still want to explore the finer-grained version of this geographical predictor (`boro`), which may achieve better results in the models. If using **Lasso or Ridge regression**, we will carefully tune the penalty parameter ($\lambda$) to ensure that the "Area" coefficients are not inflated/suppressed due to the sample size difference.

### Borough

```{r count-stops-by-boro}
plot_univariate_cat(
  DF = vehstops, 
  FEATURE = boro, 
  X_LABEL = "Borough", 
  TITLE = "Distribution of Car Stops by Borough (Sorted by Count Desc)",
  SUBTITLE = "A more fine-grained bracketing of areas shows that Brookly (South) and Bronx dominate for amount of stops"
)
```

What can we say?

- **Key finding**: The comparison reveals that while "Staten Island" and "Bronx" remain single units, boroughs like "Brooklyn", "Queens", and "Manhattan" are split into internally heterogeneous sub-regions. The granular bracketing shows that "Brooklyn (South)" dominates with $398,062$ stops, closely followed by the "Bronx" at $381,239$ stops, whereas in the 2-wheels analysis "North Queens" was the dominant category.
- **Interpretation**: If the arrest rate in "Brooklyn (South)" (the highest-volume area) is statistically distinct from "Brooklyn (North)", the `area` variable acts as a source of **aggregation bias**. By averaging these rates into a single parent borough, we lose the model’s ability to differentiate between high-risk and low-risk zones within the same borough (potentially obscuring patterns of excessive enforcement). On the other hand, because every single granular category in this 4-wheels dataset contains a massive amount of observations (ranging from $166,627$ in Staten Island to nearly $400$k in South Brooklyn), we do not run the risk encountered in the 2-wheels context where low-frequency categories caused high variance and wide confidence intervals in our parameter estimates.
- **Comment**: Discarding granularity prematurely is a risk. If a specific sub-borough has a disproportionately higher event rate than its parent borough, the model will fail to capture that specific hotspot. We should not treat granularity as "noise" until we have empirical evidence that the target ratios are homogeneous within the boroughs. In fact, if the target ratio differs significantly between "Brooklyn (South)" and "Brooklyn (North)", aggregating them would indeed "wash out" a genuine, predictive signal, resulting in an under-fitted model. Given the substantial sample sizes across all subdivisions here, the statistical power to detect localized differences is significantly elevated.
- **Proposed Action**: As previously stated, we wish to plot the arrest rate for each borough level, as well for the areas. Then, based on our observations, we could train two alternative versions of the same model using `area` and `boro`. In fact, we could employ means of model comparison (for instance, Stratified K-fold Cross Validation) and, in case the granular version yields lower CV error, we may be able to discard the hypothesis of introducing noise/variability.

### Latitude and Longitude for Stop Density

For what concerns these two predictors, we will use the to plot a spatial heat map of the stops with these values available.

```{r stop-density}
nyc_map <- counties(state = "NY", cb = TRUE) %>%
  filter(NAME %in% c("Bronx", "Kings", "New York", "Queens", "Richmond"))

bins <- 100
lon_breaks <- seq(-74.26, -73.71, length.out = bins + 1)
lat_breaks <- seq(40.48, 40.92, length.out = bins + 1)

grid_sum <- vehstops %>%
  filter(!is.na(longitude) & !is.na(latitude)) %>%
  mutate(
    xbin = cut(longitude, breaks = lon_breaks, include.lowest = TRUE, right = FALSE),
    ybin = cut(latitude, breaks = lat_breaks, include.lowest = TRUE, right = FALSE)
  ) %>%
  mutate(
    x = as.numeric(gsub(".*\\[(.*),.*", "\\1", xbin)) + (lon_breaks[2] - lon_breaks[1])/2,
    y = as.numeric(gsub(".*\\[(.*),.*", "\\1", ybin)) + (lat_breaks[2] - lat_breaks[1])/2
  ) %>%
  group_by(x, y) %>%
  summarise(n = n(), .groups = "drop")

ggplot() +
  geom_sf(data = nyc_map, fill = "white", color = "grey80", size = 0.2) +
  geom_tile(data = grid_sum, aes(x, y, fill = n), alpha = 0.8) +
  scale_fill_viridis_c(trans = "log10", name = "Stops (log10)", option = "mako", direction = -1) +
  coord_sf(xlim = c(-74.26, -73.71), ylim = c(40.48, 40.92)) +
  theme_minimal() +
  theme(panel.grid = element_blank()) +
  labs(title = "Heatmap of Cars Stops Density - 100 Bins", x = "Longitude", y = "Latitude")
```

```{r, include=FALSE}
rm(grid_sum, nyc_map, bins, lat_breaks, lon_breaks)
```

What can we say?

- **Key finding**: The heatmap reveals a highly heterogeneous spatial distribution, characterized by distinct high-density hotspots that cover a much wider and denser geographic layout across the city compared to the 2-wheels analysis. We observe large clusters of dense activity (reaching values up to $1\text{e}+04$ on the $\log_{10}$ scale) spreading across major sections of Manhattan, Brooklyn, Queens, and the Bronx, while Staten Island shows more dispersed, lower-density tiles.
- **Interpretation**: This spatial clustering indicates that police enforcement is not randomly distributed but it is likely to be following the major traffic arteries and commercial hubs. The extensive coverage of high-density tiles reflects the ubiquitous nature of 4-wheeled transportation in the urban grid, meaning that spatial coordinates share a high amount of mutual information with the overall distribution of traffic volume.
- **Comment**: When we look at this from a qualitative perspective, mapping continuous geographic coordinates directly into predictive models introduces a serious risk of memorizing spatial noise. Since our goal is to isolate systematic enforcement bias rather than building a geographical positioning system for police presence, treating latitude and longitude as raw predictors could cause the models to proxy localized demographic variations implicitly. This is problematic because it can create a false sense of predictive accuracy while failing to capture the structural variables that actually drive the `arrest_made` target.
- **Proposed Action**: To prevent data leakage we cannot just include them in the models. These two predictors will be later removed. Instead of relying on raw coordinates, we will rely on the administrative categorical boundaries (such as `boro` or `area`) which aggregate these spatial patterns into structurally meaningful units for the classification models.

## Time-related predictors

### Year

```{r count-stops-by-year}
vehstops %>%
  filter(!is.na(year)) %>%
  count(year) %>%
  mutate(year = as.numeric(as.character(year))) %>%
  arrange(year) %>%
  mutate(
    prev_n = lag(n), pct_change = (n - prev_n) / prev_n * 100,
    label_text = ifelse(
      is.na(pct_change), 
      comma(n), 
      paste0(comma(n), "\n(", ifelse(pct_change > 0, "+", ""), round(pct_change, 1), "%)")
    )) %>%
  ggplot(aes(x = factor(year), y = n, group = 1)) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_point(color = "steelblue", size = 3) +
  geom_text(aes(label = label_text), vjust = -0.8, nudge_x = -0.1, size = 3.5, lineheight = 0.8) +
  scale_y_continuous(labels = comma, limits = c(0, NA), expand = expansion(mult = c(0, 0.2))) +
  theme_minimal() +
  theme(
    panel.grid.major.x = element_blank(), 
    plot.title = element_text(size = 14),
    plot.subtitle = element_text(size = 10, color = "grey50")
    ) +
  labs(
    title = "Annual Volume of Stops for Cars in NYC",
    subtitle = "In 3 Years the Total Volume Increased by ... Times",
    x = "Year", y = "Count of Stops"
    )
```

What can we say?

- **Key finding**: There has been a steady, multi-year increase in the volume of police stops involving 4-wheeled vehicles, with the absolute number of encounters expanding from $611,734$ in $2023$ to $799,853$ in $2025$.
- **Interpretation**: The growth trajectory is non-linear but far more stable than the explosive surge seen in the 2-wheels analysis. The $21.9\%$ jump between $2023$ and $2024$ represents a significant upward trend, which is suggesting a broader intensification of traffic enforcement across the city. On the other hand, the $7.2\%$ increase in 2025 indicates that while the absolute volume of activity is still rising, the rate of acceleration is starting to stabilize.
- **Comment**: The visualization clearly demonstrates that the absolute "public exposure" to these stops is vast, operating on a scale nearly ten times larger than that of 2-wheeled vehicles. However, since the dataset analyzed represents the foundational baseline of standard automotive traffic, this moderate growth likely reflects institutional shifts in overall city-wide policing strategy rather than highly specific, hyper-targeted campaigns. Plus, 3 years is not a sufficient time span for us to deliver some substantial proof regarding long-term structural trends.
- **Proposed Action**: From a modeling perspective, we need to ensure that the temporal variable `year` does not introduce confounding effects when evaluating our target variable. We will include `year` as a categorical factor or dummy variable in our classification models to control for these shifting annual baselines. Furthermore, we must check if the arrest rate remained stable during this expansion or if the influx of new stops diluted the proportion of `arrest_made` outcomes, which we can verify by plotting the conditional probability of arrests over time.

### Month

```{r count-stops-by-month}
month_data <- vehstops %>%
  filter(!is.na(month)) %>%
  count(month) %>%
  mutate(month = factor(month, levels = 1:12)) %>%
  arrange(month) %>%
  mutate(
    prev_n = lag(n), pct_change = (n - prev_n) / prev_n * 100,
    label_text = ifelse(
      is.na(pct_change), comma(n),
      paste0(comma(n), "\n(", ifelse(pct_change > 0, "+", ""), round(pct_change, 1), "%)"))
  )

month_data %>%
  ggplot(aes(x = month, y = n, group = 1)) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_point(color = "steelblue", size = 3) +
  geom_text(aes(label = label_text), vjust = -1, nudge_x = -0.1, size = 3, lineheight = 0.8) +
  scale_x_discrete(labels = month.abb) +
  scale_y_continuous(labels = comma, limits = c(0, NA), expand = expansion(mult = c(0, 0.3))) +
  theme_minimal() +
  theme(
    panel.grid.major.x = element_blank(),
    plot.title = element_text(size = 14),
    plot.subtitle = element_text(color = "grey50", size = 10) # Styled as requested
  ) +
  labs(
    title = "Monthly Volume of Stops for Cars in NYC",
    subtitle = "Over 3 Years, Most Vehicles are Stopped between June and October",
    x = "Month",
    y = "Count of Stops"
  )
```

```{r, include=FALSE}
rm(month_data)
```

What can we say?

- **Key finding**: The data exhibits a highly distinct monthly pattern for 4-wheels vehicles, contrasting sharply with the June to October summer plateau observed for bikers. Here, stop volumes peak in the early months of the year, reaching an absolute maximum in March at $208,661$ stops, before experiencing a notable multi-month contraction that bottoms out in September at $154,192$ stops (a $-7.4\%$ drop from August).
- **Interpretation**: Unlike 2-wheeled vehicles, 4-wheeled vehicles operate year-round regardless of harsh winter conditions, explaining the high volume baseline maintained in January ($204,603$) and February ($200,831$). The steady downward trajectory into June ($-16.4\%$) and July indicates that warm weather does not drive an increase in 4-wheels encounters, suggesting instead that shifting law enforcement administrative cycles, fiscal year targets, or changing city-wide patrol strategies dictate the distribution of these stops.
- **Comment**: It would be easy to state that the decline in stops during late summer and early autumn is likely due to structural shifts in commuter behavior rather than a change in police behavior. However, both mechanisms are not distinguishable from the data alone. What is qualitatively vital here is that the peaks and troughs run almost completely counter-cyclical to the 2-wheels distribution, proving that the underlying operational mechanics driving police encounters for standard automobiles are entirely decoupled from seasonal weather dependencies.
- **Proposed Action**: Given this clear monthly pattern, we ensure our predictive models include this time-related feature. We will incorporate `month` as a categorical factor to capture these cyclical baseline variations across the calendar year. Furthermore, to verify if these volume shifts alter the underlying target probability, we will test for interaction terms between `month` and the vehicle type during downstream classification tasks to see if seasonal variation modulates the likelihood of the `arrest_made` outcome.

Let us see this with respect to the year

```{r year-monthly-distribution}
plot_data <- vehstops %>%
  filter(!is.na(year) & !is.na(month)) %>%
  count(year, month) %>%
  mutate(month = factor(month, levels = 1:12)) %>%
  group_by(year) %>%
  mutate(
    n_jan = n[month == 1],
    n_max = max(n),
    is_max = n == n_max,
    pct_max = round((n - n_jan) / n_jan * 100, 0),
    pct_dec = round((n - n_max) / n_max * 100, 0),
    
    label_text = case_when(
      month == 1 ~ comma(n), 
      is_max ~ paste0(comma(n), "\n(", ifelse(pct_max > 0, "+", ""), pct_max, "%)"),
      month == 12 ~ paste0(comma(n), "\n(", pct_dec, "%)"), # Dec is usually negative, so '-' is automatic
      TRUE ~ ""
    )
  )

# Faceted plot
plot_data %>%
  ggplot(aes(x = month, y = n, group = 1)) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_point(color = "steelblue", size = 2) +
  geom_text(aes(label = label_text), vjust = -1, hjust = 0.2, size = 2.5, lineheight = 0.8) +
  facet_wrap(~year, nrow = 1) +
  scale_x_discrete(
    labels = c("J", "F", "M", "A", "M", "J", "J", "A", "S", "O", "N", "D"), 
    expand = expansion(mult = c(0.1, 0.1))
    ) +
  scale_y_continuous(labels = comma, limits = c(0, NA), expand = expansion(mult = c(0, 0.4))) +
  coord_cartesian(clip = "off") +
  theme_minimal() +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    plot.title = element_text(size = 14),
    plot.subtitle = element_text(color = "grey50", size = 10),
    strip.text = element_text(size = 11)
  ) +
  labs(
    title = "Monthly Volume of Stops, by Year, for Cars in NYC",
    subtitle = "Custom Growth/Decline analysis: January vs Max, and Max vs December",
    x = "Month",
    y = "Count of Stops"
  )
```

```{r, include=FALSE}
rm(plot_data)
```

What can we say?

- **Key finding**: There is a pronounced shifting trend in the intra-year volume distribution over the three-year period. In $2023$, the annual maximum occurred early in March ($64,991$, an $11\%$ increase from January) followed by a sharp drop to December ($38,193$, a $-41\%$ decline). By $2025$, the behavior changes markedly, as the year begins at a much higher baseline peak in January ($81,793$), experiences a steep mid-year contraction through summer, and finishes with a late-year resurgence that leads to a much softer winter decline by December ($62,748$, a $-23\%$ drop from the annual maximum).
- **Interpretation**: The data indicates that enforcement patterns for 4-wheels vehicles are undergoing structural adjustments over time. The widening gap between the high-volume peaks and the low-volume troughs seen in the 2-wheels context does not apply here. Instead, the "off-season" drop in December is becoming significantly less severe, moving from $-41\%$ in $2023$ to $-23\%$ in $2025$, which implies a stabilization of late-year policing activities.
- **Comment**: It clearly shows that while the baseline (January) is rising substantially over the years, the *shape* of the annual trend is shifting from a spring-centric peak to a bimodal distribution with a prominent winter surge. This qualitative shift confirms that seasonal vehicle usage is not the primary mechanism driving these 4-wheels encounters, pointing instead toward evolving, multi-year administrative directives or policy shifts within the police department.
- **Proposed Action**: Rather than creating a static binary indicator for a warm season as we did for 2-wheels vehicles, we will **include interaction terms between `year` and `month**` in our classification framework. This will allow our GLM and distance-based models to capture the changing seasonal dynamics and prevent the model from assuming a fixed monthly effect across the entire $2023$ to $2025$ timeline.

### Checking out `hot_season`

Let us proceed with this intuition

```{r add-hot-season}
vehstops <- vehstops %>%
  mutate(hot_season = ifelse(month %in% 4:10, 1, 0))
```

Now, for plotting

```{r year-hot-season-distribution}
plot_data <- vehstops %>%
  filter(!is.na(year) & !is.na(month)) %>%
  mutate(
    is_hot_season = ifelse(month %in% 4:10, "Hot Season", "Off Season"),
    is_hot_season = factor(is_hot_season, levels = c("Hot Season", "Off Season"))
  ) %>%
  count(year, is_hot_season) %>%
  group_by(year) %>%
  mutate(
    year_total = sum(n),
    pct = n / year_total,
    label_text = paste0(comma(n), "\n(", round(pct * 100, 1), "%)")
  )

plot_data %>%
  ggplot(aes(x = is_hot_season, y = n, fill = is_hot_season)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~year, nrow = 1) +
  geom_text(aes(label = label_text), vjust = -0.5, size = 3.5, lineheight = 0.8) +
  scale_y_continuous(labels = comma, expand = expansion(mult = c(0, 0.2))) +
  scale_fill_brewer(palette = "Paired") +
  theme_minimal() +
  theme(
    panel.grid.major.x = element_blank(),
    plot.title = element_text(size = 14),
    plot.subtitle = element_text(color = "grey50", size = 10)
  ) +
  labs(
    title = "Volume of Stops for Cars: Hot Season vs. Off Season",
    subtitle = "While Overall Volume Increase, Percentages slightly Drift Towards Hot Season",
    x = NULL,
    y = "Count of Stops"
  )
```

```{r, include=FALSE}
rm(plot_data)
```

What can we say?

- **Key finding**: Total enforcement volume is surging but, more importantly, the enforcement pattern is slowly becoming **increasingly concentrated** in the "Off Season" rather than the "Hot Season." In fact, while the hot season accounted for $55.8\%$ of stops in $2023$ ($341,205$ stops), its share slightly drifted downward to $54.5\%$ in $2025$ ($436,037$ stops) as off-season stops climbed from $44.2\%$ to $45.5\%$ ($363,816$ stops).
- **Interpretation**: The growth in stops is not uniform throughout the year. The declining percentage of hot season stops indicates that the "Off Season" (Nov-March) is becoming a slightly faster driver of the expanding annual stop totals year-over-year, which runs in complete opposition to the 2-wheels concentration behavior. However, the $3$ year span of the data does not make this minor drift definite.
- **Comment**: This plot confirms that while a "Hot Season" vs "Off Season" division captures a clear baseline volume gap, it behaves fundamentally differently for 4-wheeled vehicles. The structural shift toward a higher off-season proportion highlights that winter months are not a deterrent for automotive policing, indicating that standard vehicle enforcement is sustained at massive scales regardless of weather shifts. On the other hand, this marginal percentage drift cannot be just inferred by looking at the year.
- **Proposed Action**: Because the seasonal proportions are shifting across the years in a distinct pattern, we will **include the interaction between** `year` **and our seasonal features** during model training. This will allow our classification algorithms to properly evaluate if the probability of the `arrest_made` target is being modulated by these changing volume dynamics rather than assuming a fixed seasonal weight over time.

### Dow/Weekday, Day and Doy

We have decided to exclude 'Day of Month' (`day`) and 'Day of Year' (`doy`) from our feature set to mitigate 'calendar artifact' bias.

As we can see below, the variation in month length creates false activity dips for end-of-month dates

```{r day-bias}
vehstops %>%
  filter(!is.na(day)) %>%
  count(day) %>%
  ggplot(aes(x = day, y = n)) +
  geom_line(color = "grey60", alpha = 0.5) +
  geom_smooth(method = "loess", color = "steelblue", se = FALSE) +
  theme_minimal() +
   theme(
    panel.grid.major.x = element_blank(),
    plot.title = element_text(size = 14),
    plot.subtitle = element_text(color = "grey50", size = 10)
  ) +
  labs(title = "Volume of Stops by Day of the Month (1-31) for Cars",
       subtitle = "The drop towards 30-31 indicates it is not a behavior, but depends on the month!",
       x = "Day of the Month", y = "Count of Stops")
```

Furthermore, `doy` variable suffers from inconsistent data density.

```{r doy-incosistent-3-years-span}
plot_data <- vehstops %>%
  mutate(date = as.Date(paste(year, month, day, sep = "-"))) %>%
  count(date)

plot_data %>%
  ggplot(aes(x = date, y = n)) +
  geom_line(color = "grey70", alpha = 0.6, linewidth = 0.5) +
  geom_smooth(method = "loess", span = 0.2, color = "steelblue", se = FALSE, linewidth = 1.2) +
  scale_x_date(date_breaks = "3 months", date_labels = "%b %Y") +
  theme_minimal() +
  theme(
    panel.grid.minor = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(size = 14),
    plot.subtitle = element_text(color = "grey50", size = 10)
  ) +
  labs(
    title = "Volume of Stops by Day of the Year (1-365) for Cars",
    subtitle = "TODO",
    x = "Period", 
    y = "Count of Stops"
  )
```

```{r, include=FALSE}
rm(plot_data)
```

While we observe a consistent seasonal clustering peaking in late winter and early spring, our current 3-year dataset captures three seasonal cycles. This is sufficient to identify a recurring pattern, but insufficient to establish it as a long-term trend.

The next plot, demonstrates high daily volatility (indicated by the grey jitter) that obscures the underlying seasonal trend (the blue trend line). Because the daily fluctuation is relative to the overall pattern, including 'Day of Year' as a feature risks overfitting.

```{r doy-inconsistent-density}
vehstops %>%
  filter(!is.na(doy)) %>%
  count(doy) %>%
  ggplot(aes(x = doy, y = n)) +
  geom_line(color = "grey60", alpha = 0.5) +
  geom_smooth(method = "loess", color = "steelblue", se = FALSE) +
  theme_minimal() +
   theme(
    panel.grid.major.x = element_blank(),
    plot.title = element_text(size = 14),
    plot.subtitle = element_text(color = "grey50", size = 10)
  ) +
  labs(title = "Volume of Stops by Day of the Year (1-365) for Cars",
       subtitle = "TODO",
       x = "Day of Year", y = "Count of Stops")
```

The extreme jitter confirms this feature is too noisy to predict, but unlike the 2-wheels analysis where the trend favors hot months, the 4-wheels trend remains relatively elevated in the first half of the year before dipping during the summer months.

On the other hand, we can study the "Day of the Week" which usually abstract better.

```{r dow-dis}
vehstops %>%
  filter(!is.na(dow)) %>%
  count(dow) %>%
  mutate(
    dow = factor(dow, levels = 1:7, 
                 labels = c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun")),
    pct_change = (n - lag(n)) / lag(n) * 100,
    label_text = ifelse(is.na(pct_change), 
                        comma(n), 
                        paste0(comma(n), "\n(", ifelse(pct_change > 0, "+", ""), round(pct_change, 1), "%)"))
  ) %>%
  ggplot(aes(x = dow, y = n, group = 1)) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_point(color = "steelblue", size = 3) +
  geom_text(aes(label = label_text), vjust = -1, size = 3, lineheight = 0.8) +
  scale_y_continuous(labels = comma, 
                     limits = c(0, NA), 
                     expand = expansion(mult = c(0, 0.4))) + # Extra room for labels
  theme_minimal() +
  theme(
    panel.grid.major.x = element_blank(),
    plot.title = element_text(size = 14),
    plot.subtitle = element_text(color = "grey50", size = 10)
  ) +
  labs(
    title = "Daily Volume of Stops for Cars",
    subtitle = "TODO",
    x = "Day of the Week", 
    y = "Count of Stops"
  )
```

What can we say?

- **Key finding**: Enforcement activity follows a distinct "Mid-Week Peak" cycle that strongly mirrors the weekly rhythm seen in the 2-wheels analysis. The volume starts at its absolute lowest on Monday ($230,725$), raises aggressively ($+52\%$) into Tuesday ($350,695$), peaks on Wednesday at $380,203$ stops ($+8.4\%$), and then undergoes a multi-day decline through the end of the week, bottoming out on Sunday ($236,756$).
- **Interpretation**: The distribution confirms that weekly enforcement patterns are highly structured. The dramatic rise from Monday to Wednesday followed by a steady drop through Friday ($-14.1\%$), Saturday ($-14.5\%$), and Sunday ($-12.9\%$) indicates that the volume of stops is tightly bound to general weekday activity cycles rather than weekend leisure or recreational travel.
- **Comment**: This pattern suggests that these stops are primarily linked to standard **commercial and commuter usage** (such as regular weekday work traffic and business-hour operations) rather than weekend travel. While the 2-wheeled analysis pointed to specific commercial sub-sectors like courier services, the massive scale of the 4-wheels data indicates a broader institutional alignment with the typical NYC business week. The interpretation is plausible but not derivable from these counts alone.
- **Proposed Action**: We may keep `weekday` as the new `day` variable in the model to serve as a robust categorical factor. Also we need first to check the signal from other indicators previously built, verifying whether the probability of an `arrest_made` outcome shifts during these peak volume days or remains uniform across the weekly cycle.

### Hour and Time Bracket

First of all we look at the overall cycle of stops volume

```{r hour-dist}
hours_to_show <- c(0, 4, 8, 12, 16, 19, 20, 23)

full_data <- vehstops %>%
  filter(!is.na(hour)) %>%
  count(hour)

highlight_data <- full_data %>%
  filter(hour %in% hours_to_show) %>%
  arrange(hour) %>%
  mutate(
    pct_change = (n - lag(n)) / lag(n) * 100,
    label_text = ifelse(
      is.na(pct_change), 
      comma(n),
      paste0(comma(n), "\n(", ifelse(pct_change > 0, "+", ""), round(pct_change, 1), "%)"))
  )

ggplot() +
  geom_line(data = full_data, aes(x = hour, y = n), color = "grey80", linewidth = 1) +
  geom_point(data = highlight_data, aes(x = hour, y = n), color = "steelblue", size = 3) +
  geom_text(data = highlight_data, aes(x = hour, y = n, label = label_text), 
            vjust = -0.6, size = 3, lineheight = 0.8) +
  scale_x_continuous(breaks = 0:23) +
  scale_y_continuous(labels = comma, limits = c(0, NA), expand = expansion(mult = c(0, 0.4))) +
  theme_minimal() +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    plot.title = element_text(size = 14),
    plot.subtitle = element_text(color = "grey50", size = 10)
  ) +
  labs(
    title = "Hourly Volume of Stops for Cars",
    subtitle = "TODO",
    x = "Hour of Day (0-23)", 
    y = "Count of Stops"
  )
```

```{r, include=FALSE}
rm(hours_to_show, full_data, highlight_data)
```

What can we say?

- **Key finding**: Enforcement volume is characterized by a "fractured" daily cycle that is clearly divided into three distinct operational phases. It follows a predictable overnight decline to an absolute minimum at $4:00$ ($23,374$, a $-79.7\%$ drop), a rapid morning ascent to a localized peak at $8:00$ ($105,378$, $+350.8\%$), a distinct midday dip bottoming out before a massive afternoon surge to an absolute peak at $16:00$ ($144,954$, $+80.6\%$), and a sustained high-volume evening plateau before tailing off late at night.
- **Interpretation**: This distribution indicates that the enforcement cycle is driven by three distinct phases: the **Morning Commuter Phase** (centered around $8:00$), the **Midday Transition Phase** (which causes the temporary dip down to $80,271$ at $12:00$), and the **Late Afternoon Rush Phase** (surging aggressively to the absolute peak at $16:00$ and tapering slowly through the evening hours). Unlike the 2-wheels analysis which lacked a clear midday drop and peaked much later at $19:00$, the 4-wheels data captures structural shifts that perfectly align with major automotive traffic shifts throughout the city.
- **Comment**: From a qualitative perspective, mapping a cyclical $24$-hour clock directly into a classification model as a simple linear feature would introduce massive model misspecification. The distinct presence of three separate volume phases demonstrates that time of day has a highly non-linear relationship with vehicle stops. If we ignore this tri-modal structure, our models will completely fail to account for the operational shifts of police patrols and daily traffic density, which might heavily confound the probability of the `arrest_made` target.
- **Proposed Action**: We will maintain `hour` as a vital temporal factor in our predictive framework. To properly capture this complex three-phase behavior without creating separate models for different times of day, we will **transform the continuous hour variable using a periodic basis function**, specifically by **employing a combination of sine and cosine transformations** with a $24$-hour period or utilizing a flexible natural cubic spline with higher degrees of freedom ($df = 4$ or $5$). This feature engineering step will allow our linear and distance-based classifiers to treat $23:00$ and $0:00$ as continuous points and adapt smoothly to the shifting daily cycles.

Before anything else, let us see whether this cycle changed throughout the years.

```{r hour-year-dist}
hours_to_show <- c(0, 4, 8, 12, 16, 20, 23)

full_data <- vehstops %>%
  filter(!is.na(year) & !is.na(hour)) %>%
  count(year, hour)

highlight_data <- full_data %>%
  filter(hour %in% hours_to_show) %>%
  group_by(year) %>%
  arrange(hour) %>%
  mutate(
    pct_change = (n - lag(n)) / lag(n) * 100,
    label_text = ifelse(
      is.na(pct_change), 
      comma(n), 
      paste0(comma(n), "\n(", ifelse(pct_change > 0, "+", ""), round(pct_change, 1), "%)"))
  )

ggplot() +
  geom_line(data = full_data, aes(x = hour, y = n), color = "grey85", linewidth = 0.8) +
  geom_point(data = highlight_data, aes(x = hour, y = n), color = "steelblue", size = 2) +
  geom_text(data = highlight_data, aes(x = hour, y = n, label = label_text), 
            vjust = -0.7, size = 2.5, lineheight = 0.8) +
  facet_wrap(~year, nrow = 1) +
  scale_x_continuous(breaks = c(0, 4, 8, 12, 16, 20, 23)) +
  scale_y_continuous(labels = comma, limits = c(0, NA), expand = expansion(mult = c(0, 0.4))) +
  theme_minimal() +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    plot.title = element_text(size = 14),
    plot.subtitle = element_text(color = "grey50", size = 10),
    strip.text = element_text(size = 11)
  ) +
  labs(
    title = "Hourly Volume of Stops, by Year, for Cars",
    subtitle = "TODO",
    x = "Hour of Day (0-23)", 
    y = "Count of Stops"
  )
```

```{r, include=FALSE}
rm(hours_to_show, full_data, highlight_data)
```

What can we say?

- **Key finding**: Across all three years, the enforcement volume shows a distinct bimodal pattern with a minor peak during the morning rush at $8:00$ and a dominant peak during the late afternoon and evening. In $2023$ and $2024$, the absolute highest volume of stops occurred at $16:00$ ($4:00$ PM). By $2025$, while the $16:00$ peak remains the highest absolute value ($51011$ stops), the volume sustains a remarkably high plateau into the evening, remaining nearly unchanged at $20:00$ with $49857$ stops.
- **Interpretation**: Unlike the pattern observed for two-wheeled vehicles where the peak completely migrated to later hours, four-wheeled vehicle stops maintain a consistent primary afternoon peak at $16:00$ across the entire $2023$ to $2025$ window. The expanding volume at $20:00$ in $2025$ suggests that evening traffic enforcement is intensifying overall rather than shifting away from the afternoon rush hour. The morning rush hour peak at $8:00$ remains stable in structure across all years, which shows that commuting patterns provide a highly predictable baseline for traffic stop volume.
- **Comment**: When comparing these distributions to our previous two-wheeled vehicle analysis, we notice a much more rigid and traditional bimodal rush hour structure here. While two-wheelers saw a drastic tripling of evening volume due to changing delivery dynamics or specific late evening crackdowns, four-wheeled vehicle enforcement aligns more closely with standard standard NYC commuting hours. The sustained plateau between $16:00$ and $20:00$ in $2025$ indicates an overall expansion of police activity during the entire evening transition period rather than a targeted tactical shift.
- **Proposed Action**: For our downstream predictive modeling of `arrest_made`, we should engineer a categorical feature capturing these distinct diurnal periods, such as defining an explicit evening rush window from $16:00$ to $20:00$. Because the hourly volume distribution shapes remain fairly consistent across years while absolute counts grow, we should include `hour` as a non-linear effect using a generalized additive framework or a tree-based classifier to properly map how the probability of an arrest scales during peak traffic hours.

Now let us see if the hot season might shift this cycle

```{r hour-hot-season}
plot_data_full <- vehstops %>%
  mutate(
    hot_season = ifelse(month %in% 4:10, 1, 0),
    hot_season_label = factor(hot_season, levels = c(0, 1), 
                              labels = c("Off Season (Nov-Mar)", "Hot Season (Apr-Oct)"))
  ) %>%
  filter(!is.na(hour))

full_data <- plot_data_full %>%
  count(hot_season_label, hour)

hours_to_show <- c(0, 4, 8, 12, 16, 20, 23)

highlight_data <- full_data %>%
  filter(hour %in% hours_to_show) %>%
  group_by(hot_season_label) %>%
  arrange(hour) %>%
  mutate(
    pct_change = (n - lag(n)) / lag(n) * 100,
    label_text = ifelse(is.na(pct_change), 
                        comma(n), 
                        paste0(comma(n), "\n(", ifelse(pct_change > 0, "+", ""), round(pct_change, 1), "%)"))
  )

ggplot() +
  geom_line(data = full_data, aes(x = hour, y = n), color = "grey85", linewidth = 0.8) +
  geom_point(data = highlight_data, aes(x = hour, y = n), color = "steelblue", size = 2) +
  geom_text(data = highlight_data, aes(x = hour, y = n, label = label_text), 
            vjust = -0.7, size = 2.5, lineheight = 0.8) +
  facet_wrap(~hot_season_label, nrow = 1) +
  scale_x_continuous(breaks = c(0, 4, 8, 12, 16, 20, 23)) +
  scale_y_continuous(labels = comma, limits = c(0, NA), expand = expansion(mult = c(0, 0.4))) +
  theme_minimal() +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    plot.title = element_text(size = 14),
    plot.subtitle = element_text(color = "grey50", size = 10),
    strip.text = element_text(size = 12)
  ) +
  labs(
    title = "Hourly Cycle of Enforcement: Hot Season vs. Off Season",
    subtitle = "TODO",
    x = "Hour of Day (0-23)", 
    y = "Count of Stops"
  )
```

```{r, include=FALSE}
rm(plot_data_full, hours_to_show, full_data, highlight_data)
```

What can we say?

- **Key finding**: The fundamental bimodal shape of traffic enforcement remains remarkably stable across both periods, peaking consistently at $16:00$ in both the Off Season ($70134$ stops) and the Hot Season ($74820$ stops). However, the Hot Season exhibits a significantly higher baseline volume throughout the entire day, particularly at midnight ($66252$ stops compared to $49057$ in the Off Season) and during the evening hours, where volume remains elevated at $20:00$ ($72847$ stops) before tapering off.
- **Interpretation**: Unlike two-wheeled vehicles where seasonal shifts completely migrate the peak hour, four-wheeled vehicle enforcement seems governed by stable, year-round commuting infrastructures that anchor the primary peak at $16:00$. The primary seasonal effect is volume-based rather than structural, meaning that warmer weather expands the absolute quantity of nighttime and evening stops without disrupting the daytime baseline trends. The sharp drop to the daily minimum at $4:00$ is universal, which reflects a systemic low in overall vehicular activity regardless of the season.
- **Comment**: This presents a contrasting narrative to the two-wheel case. While two-wheelers are heavily dependent on weather conditions for their operational viability (which explains why their enforcement curve shifts dramatically with the "Sun and Temperature" cycle), four-wheeled vehicles operate constantly across all seasons. The data suggests that police deployment strategies for standard vehicles track general traffic density trends, though they take advantage of increased nighttime activity during the warmer months to sustain higher enforcement levels late into the evening.
- **Proposed Action**: For our classification framework targeting `arrest_made`, we should model season as an interaction term with hour rather than just a standalone main effect. Since the peak hours do not shift but the late-night volumes change significantly, an interaction term like `hour * season` will allow the model to capture whether a stop made at $20:00$ or $0:00$ carries a different conditional probability of arrest depending on the time of year.

Let us check the time brackets previously built.

```{r time-bracket-distribution}
plot_univariate_cat(
  DF = vehstops, 
  FEATURE = time_bracket, 
  X_LABEL = "Time Bracket", 
  TITLE = "Distribution of Time Brackets for Cars",
  SUBTITLE = "TODO"
)
```

What can we say?

- **Key finding**: The "Afternoon/evening ($16-19$)" bracket stands as the absolute dominant peak with $546748$ stops, closely followed by the "Night ($20-23$)" and "Morning commute ($8-11$)" brackets with $449686$ and $410555$ stops respectively. The lowest volume of enforcement is concentrated in the "Early morning ($4-7$)" window, which registers only $138852$ stops.
- **Interpretation**: This distribution confirms that the temporal structure of four-wheeled vehicle enforcement is heavily anchored around the latter half of the day ($16-23$), mirroring the core pattern observed in our hourly time series. The relatively high volume in the "Morning commute ($8-11$)" bucket relative to "Midday ($12-15$)" validates that our feature engineering captures the traditional bimodal traffic peaks effectively, aligning well with systemic commuter workflows in NYC.
- **Comment**: In contrast to the two-wheeled vehicle analysis where the late night blocks saw massive, policy-driven exponential spikes that skewed the buckets, the four-wheeled data shows a smoother, more classic step-down distribution. The progressive decrease from afternoon to late night and into early morning fits standard urban traffic density expectations. This confirms that our current `time_bracket` categorization acts as an excellent, representative aggregator of the raw hourly data without washing away the underlying cyclical patterns.
- **Proposed Action**: We can safely maintain our existing `time_bracket` feature engineering exactly as it is for the upcoming modeling phase. Since these discrete intervals successfully segregate the high-volume rush periods from the low-volume operational troughs, they will provide a clean categorical input to evaluate how the conditional probability of our binary target `arrest_made` fluctuates across distinct parts of the day.

Also, we are going to drop minute as its level of granularity cannot be properly included in the dataset.

## Driver related predictors

### Reported Age

We have mainly one issue here. First, we have many instances of drivers with unknown age,

```{r age-bracket-dist}
plot_univariate_cat(
  DF = vehstops, 
  FEATURE = age_bracket, 
  X_LABEL = "Age Bracket", 
  TITLE = "Distribution of Car Stops by Age Bracket (Sorted by Count Desc)",
  SUBTITLE = "TODO"
)
```

What can we say?

- **Key finding**: The demographic distribution of stops is highly concentrated in the $25-34$ age category with $631234$ observations, followed sequentially by the $35-44$ and $18-24$ brackets. Furthermore, the "Unknown" age category represents a significant portion of the distribution with $118990$ records, while the "Under 18" bracket forms the smallest group with just $9298$ stops.
- **Interpretation**: The data reflects a clear operational density focusing on young to middle-aged adults, which together account for the vast majority of four-wheeled vehicle stops. While older age groups like the $65+$ bracket ($90845$ stops) show lower frequency, they are much more well-represented here compared to the two-wheeled vehicle analysis. This confirms that four-wheeled drivers span a broader generational spectrum, though the primary concentration remains securely centered around individuals between $18$ and $44$ years old.
- **Comment**: We do not treat the "Unknown" category as garbage data to be discarded because its substantial size means it could hold structured missingness related to specific operational scenarios, such as instances where an officer is unable to verify a driver's ID or when a stop escalates too rapidly to document demographic details. Interestingly, the presence of a strong $35-44$ cohort here showcases the standard distribution of licensed drivers in the city, differentiating this group from the heavily youth-skewed delivery or micromobility patterns seen in two-wheelers.
- **Proposed Action**: We will retain the "Unknown" category as a distinct nominal level within our `age_bracket` feature rather than using imputation, which allows the predictive model to capture any potential correlation between undocumented demographics and the `arrest_made` outcome. Due to their limited sample sizes, we will monitor the "Under 18" and $65+$ categories during the modeling stage to verify if they provide enough statistical power, or if they should eventually be pooled into adjacent brackets to stabilize the variance of our parameter estimates.

And now, let us see how this looks when looking at the full distribution

```{r age-vs-age-bracket}
age_data <- vehstops %>%
  filter(!is.na(age), age >= 0, age < 100) %>%
  mutate(
    age_group = case_when(
      age < 18 ~ "<18",
      age >= 18 & age <= 24 ~ "18–24",
      age >= 25 & age <= 34 ~ "25–34",
      age >= 35 & age <= 44 ~ "35–44",
      age >= 45 & age <= 54 ~ "45–54",
      age >= 55 & age <= 64 ~ "55–64",
      age >= 65 ~ "65+"
    ),
    age_group = factor(age_group, levels = c("<18", "18–24", "25–34", "35–44", "45–54", "55–64", "65+"))
  )

ggplot(age_data, aes(x = age, fill = age_group)) +
  geom_histogram(binwidth = 1, color = "white", linewidth = 0.1) +
  scale_fill_viridis_d() +
  theme_minimal() +
   theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    plot.title = element_text(size = 14),
    plot.subtitle = element_text(color = "grey50", size = 10),
    strip.text = element_text(size = 11)
  ) +
  labs(
    title = "Age Distribution of Car Drivers by Target Bracket",
    subtitle = "TODO",
    x = "Driver Age", 
    y = "Count of Stops",
    fill = "Age Group"
  ) +
  theme(legend.position = "bottom") +
  guides(fill = guide_legend(nrow = 2))
```

```{r, include=FALSE}
rm(age_data)
```

What can we say?

- **Key finding**: Enforcement volume scales rapidly from the late teens, climbing into a broad, highly sustained plateau that spans from roughly age $23$ through age $35$ with individual yearly counts consistently exceeding $60000$ stops. Beyond age $35$, the volume begins a steady, continuous decline through middle age, which transitions into a long, dwindling tail that tapers off gradually past age $75$ and approaches zero near age $90$.
- **Interpretation**: The continuous age distribution reveals that four-wheeled vehicle stops do not possess a single sharp operational peak like two-wheelers, but are instead characterized by a wide window of maximum exposure between $23$ and $35$ years old. The decline after age $35$ is relatively gradual rather than steep, reflecting that a substantial volume of standard vehicle drivers remain active well into their $45-54$ and $55-64$ cohorts. The sparse tail at the far right indicates that observations past age $80$ represent extreme values in our exposure domain.
- **Comment**: When juxtaposed against the two-wheeled vehicle profile, this visualization highlights the universal nature of four-wheeled transport in NYC. While micromobility and delivery trends compress the two-wheel enforcement distribution tightly into the mid-20s, the four-wheel curve shows a much broader plateau that matches general adult demographic trends. The slower decay through middle age confirms that older drivers are much more prevalent in this subset, providing a much higher baseline of data points across all intermediate age classes.
- **Proposed Action**: We do not need to collapse the $55-64$ and $65+$ categories into a single group because the absolute count of stops for these senior demographics remains high enough to provide sufficient statistical power. However, to enhance model stability and protect against leverage points in our statistical framework, we will filter our dataset to exclude observations where `Driver Age` exceeds $85$ years old. This minor truncation will eliminate the sparsest portion of the tail without losing meaningful predictive signal for our target variable `arrest_made`.

Let us remove the outliers

```{r remove-age-outliers}
# Keep the NAs
vehstops <- vehstops %>%
  filter(age < 76 | is.na(age))
```

From now on we will consider drivers of age in the range 16-75.

```{r age-outlier-boxplot-after}
vehstops %>%
  filter(!is.na(age)) %>%
  ggplot(aes(y = age)) +
  geom_boxplot(fill = "steelblue", alpha = 0.6, width = 0.4,
               outlier.size = 1.5, outlier.alpha = 0.4) +
  coord_flip() +
  theme_minimal() +
  theme(
    panel.grid.major.y = element_blank(),
    plot.title = element_text(size = 14),
    plot.subtitle = element_text(color = "grey50", size = 10)
  ) +
  labs(
    title = "Age Distribution after Outlier Removal (76+ excluded)",
    subtitle = "Remaining values span the plausible 16-75 enforcement range",
    y = "Reported Age", x = NULL
  )
```

What can we say?

- **Key finding**: Following the exclusion of the extreme upper tail ($76+$), the remaining age distribution for four-wheeled vehicle drivers displays a moderately right-skewed profile. The median is centered around $35$, with the interquartile range ($IQR$) stretching roughly from $27$ to $47$. The lower whisker reaches down near the minimum legal driving age around $16-17$, while the upper whisker extends smoothly up to $75$ without showing any isolated points flagged as statistical outliers.
- **Interpretation**: The fact that the $IQR$ spans a wide $20$-year window ($27-47$) reflects the substantial presence of mature drivers in the four-wheel dataset. This contrasts with narrower age profiles, suggesting less vulnerability to high statistical leverage at the upper bounds since the data density remains relatively thick up through age $55$. While the distribution decays after the late $30$s, the absence of detached outlier points below age $76$ suggests that the remaining continuous profile is stable and structurally coherent.
- **Comment**: In comparison to the two-wheeled vehicle analysis where older drivers were so sparse that the $65-75$ range was flagged as an outlier group, the four-wheeled cohort accommodates these ages naturally within the main body of the distribution. This highlights a fundamental sociological shift between the two vehicle types, as driving a standard four-wheeled automobile remains a core, persistent lifestyle feature for older New Yorkers. Consequently, the statistical distribution is much more robust across the entire adult lifespan.
- **Proposed Action**: We will proceed by prioritizing the continuous `Reported Age` variable in our exploratory model formulations, as the high data density across middle and older age classes reduces the risk of parameter instability. However, to evaluate potential non-linear impacts on `arrest_made` and ensure a balanced comparison, we will build a competitive model configuration using the discretized `age_bracket` feature. This dual approach will allow us to formally test whether a non-linear continuous formulation or a categorical pooling structure maximizes the predictive accuracy of our framework.

### Reported Sex

```{r sex-distribution}
plot_univariate_cat(
  DF = vehstops, 
  FEATURE = sex, 
  X_LABEL = "Reported Sex", 
  TITLE = "Distribution of Car Stops by Reported Sex (Sorted by Count Desc)",
  SUBTITLE = "TODO",
  order_by = "freq"
)
```

What can we say?

- **Key finding**: The data exhibits a pronounced class imbalance where "Male" (M) drivers represent the clear majority of enforcement stops with $1736848$ observations. However, "Female" (F) drivers constitute a highly substantive subset of the distribution with $395347$ stops, while the "Unknown" (U) category remains a small minority with $13392$ records.
- **Interpretation**: This distribution reflects the baseline demographic profile of drivers subjected to traffic stops across the city. Although males are stopped at a much higher frequency, the presence of nearly $400000$ female driver records provides an extensive sample size for this subgroup. This indicates that while gender exposure is highly unequal, we possess sufficient data density in both the male and female strata to reliably evaluate their respective effects on the target variable.
- **Comment**: In comparison to the two-wheeled vehicle analysis where the male category was an overwhelming $96\%$ of the data and left the female cohort statistically negligible, the four-wheeled data shows a noticeably different demographic footprint. The female proportion is significantly larger here, climbing to roughly $18\%$ of the total volume. This highlights a clear sociological shift, as four-wheeled vehicle ownership and operation in the city are far more gender-balanced than the youth-centric, male-dominated delivery and micromobility patterns that characterize the two-wheel market.
- **Proposed Action**: We will maintain `Reported Sex` as a major nominal feature in our modeling pipeline. Given the massive absolute sizes of both the M and F cohorts, our models will easily learn stable parameters for these groups without risking variance instability. However, because the "Unknown" category remains quite small and could lead to high variance or separation issues in minor sub-samples, we will implement stratified splitting techniques to guarantee that the M/F/U ratio is identically preserved across our training and validation datasets.

### Reported Ethnicity

```{r ethnicity-distribution}
plot_univariate_cat(
  DF = vehstops, 
  FEATURE = ethnicity, 
  X_LABEL = "Reported Ethnicity", 
  TITLE = "Distribution of 2-Wheels Vehicles Stops by Reported Ethnicity (Sorted by Count Desc)",
  SUBTITLE = "Hispanic drivers seem to account for the plurality of stops",
  order_by = "freq"
)
```

What can we say?

- **Key finding**: The distribution shows that "Black" drivers represent the largest single group subjected to stops with $656109$ observations, followed by "Hispanic" drivers with $565531$ observations. Together, they comprise the substantial majority of the data, while the "White" ($426594$), "Asian / Pacific Islander" ($270311$), "Unknown" ($193050$), and "Other" ($33992$) categories form progressively smaller steps in the distribution.
- **Interpretation**: This profile indicates that four-wheeled vehicle enforcement is heavily concentrated within Black and Hispanic demographics across the city. The high absolute count for the "Unknown" category matches the pattern we observed with missing demographic metrics elsewhere in our analysis. The substantial sample size across all major groups (White, Black, Hispanic, and Asian) ensures that any downstream statistical evaluation will have ample data density to measure group-specific variances accurately.
- **Comment**: In comparison to the two-wheeled vehicle analysis where the distribution was dominated by a single massive plurality of Hispanic riders (highly characteristic of NYC's delivery workforce), the four-wheeled data shows a more distributed but still highly unequal demographic breakdown. Here, Black drivers constitute the absolute peak volume of enforcement stops. This variation highlights that the underlying operational or geographic mechanisms influencing four-wheeled vehicle stops differ fundamentally from the delivery-centric regulations that governed the two-wheel market.
- **Proposed Action**: Before incorporating `Reported Ethnicity` into our predictive modeling framework for `arrest_made`, we will perform a thorough mutual information and chi-squared check against spatial and temporal predictors to identify potential confounding geographic variables. Additionally, because certain minor classes like "Other" have limited volume compared to the dominant strata, we will utilize stratified sampling across ethnicity and sex to ensure that all demographic subgroups are balanced identically between our training and testing partitions.

------------------------------------------------------------------------

# Bivariate EDA

## Association of Binary Response vs. Categorical Predictors

When dealing with a dataset of this size, traditional p-values from chi-squared tests become meaningless. To get around this, we use Cramer’s V to measure effect size (association).

Because of how the `table()` function works in R, we can also include integer predictors like `hour` and `age`: they do not take truly continuous values in this dataset, so treating each integer as its own category is valid.

```{r categorical-correlation-heatmap}
# Excluding doy (redundant) and dow (we use weekday instead)
columns <- c(
  "year", "month", "hot_season", "weekday", 
  "hour", "time_bracket",
  "area", "boro", 
  "sex", "ethnicity", "age", "age_bracket",
  "arrested"
  )

plot_categorical_correlation(
  DF = vehstops, 
  FEATURES = columns,
  TITLE = "Correlation between Categorical Variables (Cramer's V)"
)
```

What can we say?

What can we say?

- **Key finding**: The association matrix reveals that the predictors showing the most noticeable standalone signal with our target variable `arrested` are `ethnicity` ($0.1$) followed closely by the temporal drivers `hour` and `time_bracket` (both at $0.09$). Demographic features such as `age` and `age_bracket` share a modest pairwise association ($0.07$), while spatial features like `boro` ($0.06$) and `area` ($0.05$) along with `sex` ($0.05$) exhibit very low standalone signals. The matrix highlights critical structural associations between predictors, specifically a perfect relationship between the geographic blocks `area` $\leftrightarrow$ `boro` (Cramer's V of $1.00$), a strong redundancy between `hour` $\leftrightarrow$ `time_bracket` ($0.99$), and notable cross-associations between `ethnicity` $\leftrightarrow$ `boro` ($0.19$) and `ethnicity` $\leftrightarrow$ `sex` ($0.15$).
- **Interpretation**: Temporal and ethnic features carry the strongest individual baseline impacts when predicting an arrest, while standalone spatial attributes are less informative on their own for four-wheeled vehicles. The perfect $1.00$ association between `area` and `boro` indicates absolute structural redundancy, which means including both features simultaneously in their raw forms will introduce perfect multi-collinearity into our framework. Furthermore, the persistent $0.17$ to $0.19$ association between ethnicity and geography suggests that the demographic composition of stops remains non-uniformly distributed across different parts of the city.
- **Comment**: We observe a fascinating shift when comparing these dynamics to the two-wheeled vehicle analysis. While two-wheelers showed a stronger direct connection to specific geographic enforcement zones (where `area` and `boro` scored $0.13$), four-wheeled vehicle stops display a more geographically dispersed arrest pattern. However, because no single predictor strongly correlates with `arrested`, models relying purely on main effects will likely underfit the data. The true operational narrative resides within conditional interactions, since the relationships between driver demographics, time of day, and spatial deployment change depending on the combination of features.
- **Proposed Action**: We will structure our predictive framework to mitigate multi-collinearity by dropping `area` and retaining `boro` as our primary spatial representative, while similarly prioritizing `time_bracket` over raw `hour` to conserve degrees of freedom. To capture the underlying dependencies without exploding the parameter space, we will explicitly evaluate interaction terms such as (`ethnicity` $\times$ `boro`) within a regularized logistic regression framework. This will allow us to test whether the probability of an arrest spikes conditionally across specific demographic groups in specific regions of the city.

## Correlation of Binary Response vs. Numerical Predictors

```{r numerical-correlation-heatmap}
# Ignoring Latitude and Longitude
numerical_cols <- c(
  "arrested",
  "year", "month", "day", "dow", "doy", "hour", 
  "age"
  )

cor_matrix <- cor(vehstops[, numerical_cols], use = "complete.obs")

corrplot(cor_matrix, method = "color", type = "upper",
         order = "hclust", addCoef.col = "black", number.cex = 0.7,
         tl.col = "black", tl.srt = 45,
         title = "Correlation Matrix of Numerical Features",
         mar = c(0, 0, 2, 0))
```

What can we say?

- **Key finding**: The Pearson correlation matrix shows that our binary target `arrested` has a negligible linear relationship with all continuous numerical features. The strongest standalone linear relationship with the target appears in `age` ($-0.07$), which indicates a very slight negative trend, followed by `hour` ($0.03$). Calendar features like `month`, `doy` (day of year), `year`, `dow` (day of week), and `day` all show absolute correlation coefficients of $0.01$ or lower, pointing to a complete absence of direct global linear relationships. Additionally, a perfect linear correlation of $1.00$ exists between `month` and `doy`.
- **Interpretation**: The near-zero linear correlations across the board demonstrate that the probability of an arrest cannot be effectively modeled or predicted using standard linear relationships on raw numerical predictors. The perfect $1.00$ relationship between `month` and `doy` is expected, since the day of the year is mathematically deterministic based on the calendar month, creating total structural redundancy. The slight negative value for `age` implies that older stopped individuals have a marginally lower baseline likelihood of being arrested, but this tells us nothing about potential non-linear shifts within intermediate age ranges.
- **Comment**: When comparing these outcomes to our previous two-wheeled vehicle analysis, we observe a shared statistical reality. Linear correlation metrics are fundamentally ill-suited for capturing the dynamics of this dataset because both the features (like cyclical hours or seasonal months) and the target variable (`arrested`) operate non-linearly. A value of $0.03$ for `hour` does not mean time of day is irrelevant. Instead, it simply means that the bimodal rush-hour distributions we discovered earlier cancel each other out when compressed into a single global linear trend line.
- **Proposed Action**: We must avoid using raw linear models like standard un-penalized linear regression, as they will completely fail to extract predictive signals from these continuous variables. To handle the perfect collinearity between calendar features, we will drop `month` and retain `doy` or convert them into seasonal categories. Furthermore, we must explicitly map the continuous predictors using non-linear modeling frameworks, such as incorporating polynomial expansions, utilizing Generalized Additive Models (GAMs) with smoothing splines for `age` and `hour`, or relying on tree-based architectures like Random Forests that naturally capture non-monotonic step functions.

## Geospatial Features

### Area

```{r area-vs-arrested-ratio}
ratio_by_cat(
  vehstops, arrested, area,
  "Area", "Arrest Rate (%)", "Arrest Ratio by Area for Cars Stopped in NYC",
  style = "stack", legend_title = "Stop ended in arrest?",
  target_labels = c("No", "Yes"),
  subtitle = "TODO",
  label_vjust = 0.5,
  bar_width = 0.6,  
  text_size = 2.5,
  aspect_ratio = 0.9
)
```

Let us look at the numbers too

```{r confirm-area-vs-arrested-ratio}
vehstops %>%
  group_by(area, arrested) %>%
  summarise(count = n(), .groups = "drop_last") %>%
  mutate(percentage = count / sum(count))
```

What can we say?

- **Key finding**: The plot reveals a comparatively narrower yet statistically significant disparity in the proportions of four-wheeled vehicle stops that culminate in an arrest across New York City areas. The arrest rate ranges from a minimum of $2.5\%$ in Queens to a maximum of $4.8\%$ in the Bronx. This represents nearly a 2x increase in the conditional probability of an arrest depending purely on where the stop occurs. While the absolute majority of stops do not result in arrests across all territories, the Bronx consistently maintains the highest conversion rate, followed by Brooklyn at $3.5\%$, Staten Island at $3.0\%$, and Manhattan at $2.8\%$.
- **Interpretation**: The variance in four-wheeled vehicle arrest rates across geography is structurally less volatile than what was observed in the two-wheel dataset. Because four-wheeled vehicles represent a massive overall sample size across all NYC boroughs, even small percentage fluctuations like the difference between $2.5\%$ in Queens and $4.8\%$ in the Bronx represent highly robust statistical signals with minimal exposure to sampling noise or wide confidence intervals. This confirms that geographic territory exerts a stable, persistent main effect on the target variable.
- **Comment**: When contrasted against the two-wheeled vehicle analysis, where arrest rates skyrocketed up to $12.1\%$ in the Bronx, four-wheeled vehicle stops demonstrate a much tighter distribution window. The multi-fold spikes seen in two-wheelers were highly localized anomalies, likely tied to aggressive enforcement of specific delivery or unregistered vehicle regulations. For standard cars, the spatial distribution is smoother, yet it confirms that police behavior and the baseline threshold for making an arrest continue to vary systematically by borough. The Bronx remains the clear operational peak for arrest probability across both analyses.
- **Proposed Action**: We will include `boro` as a core categorical predictor in our modeling pipeline to capture these geographic baseline shifts. Given that the absolute volumes of four-wheeled stops are massive across all five boroughs, we will not need to group or collapse any regional levels. However, we will ensure that our train-test splits are fully stratified across the geographic blocks to prevent any spatial imbalance from biasing our final parameter estimates for `arrested`.

### Borough

```{r boro-vs-arrested-ratio}
ratio_by_cat(
  vehstops, arrested, boro,
  "Borough", "Arrest Rate (%)", "Arrest Ratio by Borough for Cars Stopped in NYC",
  style = "stack", legend_title = "Stop ended in arrest?",
  target_labels = c("No", "Yes"),
  subtitle = "TODO",
  label_vjust = 0.5,
  bar_width = 0.8,
  text_size = 2.5
)
```

What can we say?

- **Key finding**: Disaggregating the geographic data into finer operational regions preserves the macro position of the Bronx ($4.8\%$) as the absolute peak but exposes massive intra-borough variance that was previously hidden. While Manhattan as a whole appeared moderately low in our previous chart, this breakdown reveals a steep internal gradient ranging from a citywide minimum of $1.6\%$ in Manhattan (South) to a substantial $3.9\%$ in Manhattan (North). The remaining separated regions display a steady upward slope, with Queens ranging from $2.1\%$ (North) to $2.8\%$ (South), and Brooklyn climbing from $3.3\%$ (South) to $3.7\%$ (North).
- **Interpretation**: This fine-grained visualization confirms that the macro patterns are heavily influenced by specific administrative zones. The stark divergence within Manhattan is particularly illuminating, as a stop in Manhattan (North) is more than twice as likely to result in an arrest compared to one in Manhattan (South). This strong localized variance suggests that utilizing a finer geographic division will significantly improve the explanatory power of our models, as macro borough-level metrics average out these critical operational differences.
- **Comment**: When comparing this to the two-wheeled vehicle analysis, the four-wheel data tells a much more localized story. In the two-wheel case, the entire borough of the Bronx was an isolated outlier due to targeted citywide regulations. Here, standard car stops show that police behavior and arrest conversion thresholds are highly dependent on specific command zones within the same borough. The fact that Manhattan (North) rivals the Bronx in arrest probability, while Manhattan (South) remains the absolute lowest, highlights a sharp geographical divide in enforcement outcomes that a coarse analysis would completely obscure.
- **Proposed Action**: We will perform a likelihood ratio test on nested logistic regression models to formally evaluate whether this high-resolution `Borough` feature achieves a statistically significant improvement in deviance over the simpler macro-area classifier. If the test yields a significant result, we will incorporate the finer regional categories into our final model to capture this localized variance. Conversely, if the additional complexity does not provide sufficient statistical justification, we will favor the simpler, more parsimonious macro feature to prevent overfitting.

We will need to choose between one or the other.

### Latitude and Longitude for Stop Rate

```{r}
nyc_map <- counties(state = "NY", cb = TRUE) %>%
  filter(NAME %in% c("Bronx", "Kings", "New York", "Queens", "Richmond"))

bins <- 100
lon_breaks <- seq(-74.26, -73.71, length.out = bins + 1)
lat_breaks <- seq(40.48, 40.92, length.out = bins + 1)

grid_sum_rate <- vehstops %>%
  filter(!is.na(longitude), !is.na(latitude), !is.na(arrested)) %>%
  mutate(
    xbin = cut(longitude, breaks = lon_breaks, include.lowest = TRUE, right = FALSE),
    ybin = cut(latitude, breaks = lat_breaks, include.lowest = TRUE, right = FALSE)
  ) %>%
  mutate(
    x = as.numeric(gsub(".*\\[(.*),.*", "\\1", xbin)) + (lon_breaks[2] - lon_breaks[1])/2,
    y = as.numeric(gsub(".*\\[(.*),.*", "\\1", ybin)) + (lat_breaks[2] - lat_breaks[1])/2
  ) %>%
  group_by(x, y) %>%
  summarise(
    total_stops = n(),
    arrest_rate = mean(arrested == 1),
    .groups = "drop"
  ) %>%
  filter(total_stops >= 5)

ggplot() +
  geom_sf(data = nyc_map, fill = "white", color = "grey80", linewidth = 0.2) +
  geom_tile(data = grid_sum_rate, aes(x, y, fill = arrest_rate), alpha = 0.8) +
  scale_fill_viridis_c(
    labels = scales::percent, 
    name = "Arrest Rate (%)", 
    option = "inferno",
    direction = -1
  ) +
  coord_sf(xlim = c(-74.26, -73.71), ylim = c(40.48, 40.92)) +
  theme_minimal() +
  theme(
    panel.grid = element_blank(),
    plot.title = element_text(size = 14),
    plot.subtitle = element_text(color = "grey50", size = 10)
  ) +
  labs(
    title = "Spatial Distribution of Arrest Rates for Cars Stopped in NYC",
    subtitle = "Percentage of stops resulting in arrest (Bins with n < 5 removed)",
    x = "Longitude", 
    y = "Latitude"
  )
```

What can we say?

- **Key finding**: The spatial visualization exhibits a highly decentralized distribution of arrest rates for standard vehicles, characterized by a predominantly uniform baseline across the city with light yellow to orange shades ($0\%$ to $15\%$). However, isolated high-intensity hot spots are visible as dark purple and black pixels scattered across northern Manhattan, parts of the Bronx, eastern Queens, and central Staten Island, where specific geographic bins show arrest rates spiking above $40\%$.
- **Interpretation**: While this highlights a localized geographic heterogeneity, it exposes a critical limitation regarding direct model integration because spatial coordinates violate the standard assumption of independent and identically distributed ($i.i.d.$) errors. The presence of isolated high-arrest bins amidst broad fields of low arrest rates suggests these spikes could be driven by micro-level spatial anomalies, localized checkpoints, or sparse sample sizes that survived the $n < 5$ filter but remain highly sensitive to single-event fluctuations.
- **Comment**: In comparison to the two-wheeled vehicle analysis, which showcased broad and contiguous regional hot spots reflecting targeted borough-wide crackdowns, four-wheeled vehicle enforcement presents a much more fragmented spatial narrative. Standard vehicle stops do not yield massive, sweeping clusters of high arrest rates across entire neighborhoods. Instead, the baseline risk is fairly homogenous across the five boroughs, meaning that the extreme $40\%$ plus spikes are highly localized exceptions rather than representative regional trends.
- **Proposed Action**: We should avoid feeding raw, high-resolution continuous coordinates directly into a standard generalized linear model due to the risk of spatial autocorrelation distorting our standard errors. Instead, we can extract value from this micro-spatial structure by engineering a regional density feature or by using spatial aggregation methods like K-means clustering to group adjacent bins into cohesive categorical inputs. Alternatively, we can utilize mixed-effects modeling frameworks where geographic zones are treated as random intercepts to isolate the true local variance without breaking our statistical assumptions.

## Time related features

### Year

```{r year-vs-arrested-ratio}
arrest_trends <- vehstops %>%
  filter(!is.na(year), !is.na(arrested)) %>%
  group_by(year) %>%
  summarise(
    total_stops = n(),
    arrests = sum(arrested == 1),
    rate = arrests / total_stops, 
    .groups = 'drop'
  ) %>%
  arrange(year) %>%
  mutate(
    prev_rate = lag(rate),
    delta = (rate - prev_rate) * 100,
    label_text = ifelse(
      is.na(prev_rate), 
      scales::percent(rate, accuracy = 0.1), 
      paste0(scales::percent(rate, accuracy = 0.1), 
             "\n(", ifelse(delta > 0, "+", ""), round(delta, 1), " pts)")
    )
  )

ggplot(arrest_trends, aes(x = factor(year), y = rate, group = 1)) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_point(color = "steelblue", size = 3) +
  geom_text(aes(label = label_text), vjust = -0.8, size = 3.5, lineheight = 0.8) +
  scale_y_continuous(
    labels = scales::percent, 
    limits = c(0, max(arrest_trends$rate) * 1.3), # Extra room for labels
    expand = expansion(mult = c(0, 0.1))
  ) +
  theme_minimal() +
  theme(
    panel.grid.major.x = element_blank(),
    plot.title = element_text(size = 14),
    plot.subtitle = element_text(size = 10, color = "grey50")
  ) +
  labs(
    title = "Annual Arrest Rate has Stabilized in the last 3 years for Cars Stopped in NYC",
    subtitle = "TODO",
    x = "Year", y = "Arrest Rate (%)"
  )
```

What can we say?

- **Key finding**: The line chart displays the trajectory of the annual arrest rate for four-wheeled vehicles over a three-year period, revealing an initial rise followed by stabilization. The proportion of stops resulting in an arrest rose from $2.8\%$ in $2023$ to $3.5\%$ in $2024$ (a jump of $+0.7$ percentage points), followed by a marginal contraction to $3.4\%$ in $2025$ (a slight decrease of $-0.1$ percentage points).
- **Interpretation**: From a time-related perspective, this demonstrates that the probability of an arrest given a stop shifted between $2023$ and $2024$ but has since plateaued. In a predictive framework, this behavior implies that while data from $2023$ might underestimate subsequent enforcement patterns, data points from $2024$ and $2025$ are structurally homogenous. The relative stabilization indicates that the macro baseline threshold for executing an arrest has reached an equilibrium point in the final two years of the window.
- **Comment**: In comparison to the two-wheeled vehicle analysis, which exhibited a continuous upward drift up to $6.0\%$, four-wheeled vehicle stops show a significantly lower overall arrest conversion rate and a distinct plateauing pattern. The continuous surge in two-wheel arrests was highly representative of evolving micromobility crackdowns and successive regulatory iterations stretching into $2025$. Conversely, standard vehicle enforcement represents a more mature operational environment, where policy shifts yielded a one-time adjustment in $2024$ before returning to a stable baseline.
- **Proposed Action**: We should check for an association effect between time and geography by plotting these three-year trends faceted by borough to see if the $2024$ peak and subsequent $2025$ stabilization are uniform across the city. Furthermore, we will include `year` as a categorical predictor in our experimental model pipeline to absorb the initial yearly macro trend shift, though we may consider pooling $2024$ and $2025$ if a parsimonious model formulation is desired.

### Month and Hot Season

```{r month-vs-arrested-ratio}
monthly_arrest_data <- vehstops %>%
  filter(!is.na(month), !is.na(arrested)) %>%
  group_by(month) %>%
  summarise(
    total_stops = n(),
    arrests = sum(arrested == 1),
    rate = arrests / total_stops,
    .groups = "drop"
  ) %>%
  mutate(month = factor(month, levels = 1:12)) %>%
  arrange(month) %>%
  mutate(
    prev_rate = lag(rate),
    delta = (rate - prev_rate) * 100,
    label_text = ifelse(
      is.na(prev_rate), 
      scales::percent(rate, accuracy = 0.1), 
      paste0(scales::percent(rate, accuracy = 0.1), 
             "\n(", ifelse(delta > 0, "+", ""), round(delta, 1), " pts)")
    )
  )

monthly_arrest_data %>%
  ggplot(aes(x = month, y = rate, group = 1)) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_point(color = "steelblue", size = 3) +
  geom_text(aes(label = label_text), vjust = -1.2, size = 3, lineheight = 0.8) +
  scale_x_discrete(labels = month.abb) +
  scale_y_continuous(
    labels = scales::percent, 
    limits = c(0, max(monthly_arrest_data$rate) * 1.3), 
    expand = expansion(mult = c(0, 0.1))
  ) +
  theme_minimal() +
  theme(
    panel.grid.major.x = element_blank(),
    plot.title = element_text(size = 14),
    plot.subtitle = element_text(color = "grey50", size = 10)
  ) +
  labs(
    title = "Monthly Arrest Rate Trend for Cars Stopped in NYC",
    subtitle = "Aggregated monthly arrest rate, the peak is found in TODO",
    x = "Month",
    y = "Arrest Rate (%)"
  )
```

What can we say?

- **Key finding**: The monthly arrest rate line chart describes a highly stable, flat intra-year pattern with slight structural peaks concentrated at the absolute ends of the calendar year. The arrest rate begins at $3.6\%$ in January, drops and plateaus between $3.1\%$ and $3.3\%$ from February through November, and then experiences a slight late-year surge to its annual peak of $3.7\%$ in December ($+0.4$ percentage points).
- **Interpretation**: From a temporal perspective, the conditional probability of an arrest given a stop is essentially invariant across the spring, summer, and autumn months for four-wheeled vehicles. The slight elevation observed in December and January suggests that seasonal factors, such as holiday-season traffic initiatives, increased checkpoint density, or changes in driving behavior during mid-winter, introduce a minor but distinct uptick in arrest efficiency at the turn of the year.
- **Comment**: This visualization presents a stark contrast to the two-wheeled vehicle analysis, where arrest rates exhibited a massive sinusoidal curve tied directly to the summer season, peaking at $6.5\%$ in July. For standard automobiles, there is no summer surge, which further underscores that standard vehicle operations maintain a highly predictable, institutional baseline year-round. The minor spikes in December and January are likely driven by citywide holiday enforcement campaigns rather than environmental weather changes, revealing a completely different operational timeline than the two-wheel market.
- **Proposed Action**: We should avoid relying on the binary `hot_season` feature for the four-wheeled vehicle framework, as the flat summer profile means this indicator will fail to capture any meaningful predictive signal. Instead, we can extract the true seasonal variance by engineering a specialized categorical feature that isolates the holiday transition window, such as a binary indicator for `holiday_season` spanning December and January, or by utilizing `month` as a nominal variable with January or December established as the reference levels in our regression pipeline.

What about `hot_season`?

```{r hot-season-vs-arrested-ratio}
ratio_by_cat(
  vehstops, arrested, hot_season,
  "Hot Season", "Arrest Rate (%)", "Arrest Ratio by Season for Cars Stopped in NYC",
  style = "stack", legend_title = "Stop ended in arrest?",
  target_labels = c("No", "Yes"),
  subtitle = "For Cars it shows Increased Arrest Ratio during TODO",
  label_vjust = 0.5,
  text_size = 2.5,
  bar_width = 0.7
)
```

What can we say?

- **Key finding**: The visualization shows an extremely narrow variation in the arrest rate between the two seasonal blocks. When the `Hot Season` flag is active ($1$), the proportion of stops culminating in an arrest stands at $3.2\%$, whereas during the off-season ($0$), the conversion rate is marginally higher at $3.4\%$. This represents a negligible difference of only $0.2$ percentage points between the two operational periods.
- **Interpretation**: From a feature engineering perspective, compressing the monthly calendar timeline into a binary macro-season variable yields almost zero explanatory return for four-wheeled vehicles. The negligible $0.2$ percentage point shift indicates that the global probability of an arrest given a stop is practically orthogonal to this specific definition of seasonality. This statistical invariance suggests that any predictive signal present in individual months is effectively washed out when aggregated into these broad binary blocks.
- **Comment**: This outcome stands in sharp contrast to the two-wheeled vehicle analysis, where the hot season triggered a distinct $1.1$ percentage point surge that validated the binary variable design. For two-wheelers, the weather directly governs ridership demographics and specific tactical initiatives. For standard cars, the year-round consistency of traffic composition ensures a stable background rate. Interestingly, the off-season ($0$) actually shows the minor upper hand here, which aligns with our previous monthly finding where the localized holiday spikes in December and January elevated the winter baseline.
- **Proposed Action**: We will exclude the global binary `hot_season` feature from our final predictive models for four-wheeled vehicles, as it lacks sufficient variance to provide a meaningful coefficient. Instead, we will favor more targeted indicator variables that isolate specific anomalies, such as a dedicated binary flag for the winter holiday months, or utilize a multi-level categorical variable for quarters to capture temporal transitions without oversimplifying the operational reality.

#### Faceted

```{r month-faceted-vs-arrested-ratio}
plot_data <- vehstops %>%
  filter(!is.na(year), !is.na(month), !is.na(arrested)) %>%
  group_by(year, month) %>%
  summarise(
    arrest_rate = mean(arrested == 1), 
    .groups = "drop"
  ) %>%
  mutate(month = factor(month, levels = 1:12)) %>%
  group_by(year) %>%
  mutate(
    rate_jan = arrest_rate[month == 1],
    rate_max = max(arrest_rate),
    is_max = arrest_rate == rate_max,
  
    diff_jan = (arrest_rate - rate_jan) * 100,
    diff_dec = (arrest_rate - rate_max) * 100, 
    
    label_text = case_when(
      month == 1 ~ scales::percent(arrest_rate, accuracy = 0.1), 
      is_max ~ paste0(scales::percent(arrest_rate, accuracy = 0.1), "\n(", ifelse(diff_jan > 0, "+", ""), round(diff_jan, 1), " pts)"),
      month == 12 ~ paste0(scales::percent(arrest_rate, accuracy = 0.1), "\n(", round(diff_dec, 1), " pts)"),
      TRUE ~ ""
    )
  )


plot_data %>%
  ggplot(aes(x = month, y = arrest_rate, group = 1)) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_point(color = "steelblue", size = 2) +
  geom_text(aes(label = label_text), vjust = -1, hjust = 0.2, size = 2.5, lineheight = 0.8) +
  facet_wrap(~year, nrow = 1) +
  scale_x_discrete(
    labels = c("J", "F", "M", "A", "M", "J", "J", "A", "S", "O", "N", "D"), 
    expand = expansion(mult = c(0.1, 0.1))
  ) +
  scale_y_continuous(labels = scales::percent, limits = c(0, NA), expand = expansion(mult = c(0, 0.4))) +
  coord_cartesian(clip = "off") +
  theme_minimal() +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    plot.title = element_text(size = 14),
    plot.subtitle = element_text(color = "grey50", size = 10),
    strip.text = element_text(size = 11)
  ) +
  labs(
    title = "Monthly Arrest Rate, by Year, for Cars in NYC",
    subtitle = "TODO",
    x = "Month",
    y = "Arrest Rate (%)"
  )
```

What can we say?

- **Key finding**: The faceted line chart breaks down the monthly arrest rates for four-wheeled vehicles across three consecutive years, revealing that the internal shape of the seasonal trajectory is fundamentally changing shape over time alongside its vertical shifts. While $2023$ shows a steady upward climb from a minimum of $2.4\%$ in February to a peak of $3.7\%$ in December ($+1.1$ percentage points), $2024$ and $2025$ invert this structure. In $2024$, the year peaks early at $3.9\%$ in January, drops through summer, and builds back to $3.9\%$ in December. By $2025$, the absolute peak occurs in January at $4.1\%$ before experiencing a gradual downward trend that bottoms out in autumn and rises marginally to finish at $3.5\%$ in December.
- **Interpretation**: From a longitudinal perspective, this evolution demonstrates that the seasonal distribution of the arrest probability is non-stationary. The sharp contrast between the upward trend of $2023$ and the front-loaded, downward-sloping profiles of $2024$ and $2025$ indicates that a fixed, static monthly effect will fail to capture the baseline transition. The data suggests that tactical changes in traffic enforcement strategies or policy mandates implemented between late $2023$ and early $2024$ permanently re-engineered the annual distribution, shifting the highest conditional probability of an arrest to the beginning of the calendar year.
- **Comment**: When contrasted against the two-wheeled vehicle analysis, which exhibited a highly stable, recurring summer peak across all panels, the four-wheeled data exposes a complete structural transformation. Standard vehicle enforcement is not bound by a predictable meteorological cycle like delivery riders or micromobility users. Instead, the rapid migration of the peak from December ($2023$) to January ($2024$ and $2025$) underscores that administrative, policy-driven behavior dominates this timeline. Rather than a recurring mid-summer dip, standard cars see a sustained, institutional baseline shift where early-winter enforcement has become progressively more efficient at securing arrests over the three-year window.
- **Proposed Action**: We cannot rely on a simple main effect for month or a static seasonal category in our final predictive architecture, as the shifting intra-year structure violates the assumption of temporal stationarity. To accurately capture how the probability of `arrested` scales across this timeline, we must explicitly include a macro interaction term structured as `month * year` within our model. This formulation will allow the classification framework to adjust the monthly weights dynamically across the three-year horizon, preventing the inverted trends of $2024$ and $2025$ from canceling out the upward trajectory observed in $2023$.

```{r faceted-hot-season-vs-arrested-ratio}
plot_data <- vehstops %>%
  filter(!is.na(year), !is.na(month), !is.na(arrested)) %>%
  mutate(
    is_hot_season = ifelse(month %in% 4:10, "Hot Season", "Off Season"),
    is_hot_season = factor(is_hot_season, levels = c("Hot Season", "Off Season"))
  ) %>%
  group_by(year, is_hot_season) %>%
  summarise(
    total_stops = n(),
    arrests = sum(arrested == 1),
    arrest_rate = arrests / total_stops,
    .groups = "drop"
  ) %>%
  mutate(
    label_text = paste0(scales::percent(arrest_rate, accuracy = 0.1), "\n(", comma(arrests), " arrests)")
  )

plot_data %>%
  ggplot(aes(x = is_hot_season, y = arrest_rate, fill = is_hot_season)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~year, nrow = 1) +
  geom_text(aes(label = label_text), vjust = -0.5, size = 3, lineheight = 0.8) +
  scale_y_continuous(labels = scales::percent, expand = expansion(mult = c(0, 0.2))) +
  scale_fill_brewer(palette = "Paired") +
  theme_minimal() +
  theme(
    panel.grid.major.x = element_blank(),
    plot.title = element_text(size = 14),
    plot.subtitle = element_text(color = "grey50", size = 10)
  ) +
  labs(
    title = "Arrest Rate for Cars in NYC: Hot Season vs. Off Season",
    subtitle = "TODO",
    x = NULL,
    y = "Arrest Rate (%)"
  )
```

What can we say?

- **Key finding**: The faceted bar chart reveals that the seasonal relationship with the arrest rate for standard vehicles is weak and varies over time. In $2023$, the Hot Season arrest rate was slightly higher than the Off Season rate ($2.9\%$ vs $2.8\%$), whereas in $2024$ and $2025$, this pattern inverted with the Off Season displaying a higher conditional probability of arrest ($3.7\%$ vs $3.3\%$ in $2024$, and $3.6\%$ vs $3.3\%$ in $2025$).
- **Interpretation**: The visualization shows the dark blue bars (Off Season) overtaking the light blue bars (Hot Season) in the latter two years. Furthermore, looking at the absolute volumes shows that the Off Season accounts for a massive and expanding portion of the overall arrest count, yielding $11930$ arrests in $2024$ and growing to $13078$ arrests by $2025$. This confirms that our global binary seasonal categorization fails to capture a stable, consistent operational effect for standard vehicle stops.
- **Comment**: This presents a complete contradiction to the two-wheeled vehicle analysis, where the Hot Season arrest rate was uniformly dominant across all years. For two-wheelers, warmer weather drives the operational volume and targeted enforcement. For four-wheeled vehicles, the data suggests that winter and late-autumn months actually feature more efficient or higher-probability arrest encounters in recent years. This inversion reinforces our previous finding regarding the monthly non-stationarity of standard car stops, proving that seasonal patterns are heavily dependent on yearly administrative or tactical policy changes.
- **Proposed Action**: We must avoid using the static `hot_season` binary flag as a standalone macro predictor in our final classification models, since its fluctuating direction across the years will dilute the predictive signal. Instead, we should model seasonal trends dynamically by including the interaction term (`hot_season` $\times$ `year`). This approach will allow the model to adjust for the structural shift where the conditional probability of an arrest moved from the warmer months to the colder months after $2023$.

### Day, Dow and Weekday

```{r day-vs-arrested-ratio}
daily_arrest_rate <- vehstops %>%
  filter(!is.na(day), !is.na(arrested)) %>%
  group_by(day) %>%
  summarise(
    total_stops = n(),
    arrests = sum(arrested == 1),
    rate = arrests / total_stops,
    .groups = "drop"
  )

daily_arrest_rate %>%
  ggplot(aes(x = day, y = rate)) +
  geom_line(color = "grey60", alpha = 0.5) +
  geom_smooth(method = "loess", color = "steelblue", se = FALSE) +
  scale_y_continuous(labels = scales::percent, limits = c(0, NA)) +
  theme_minimal() + 
  theme(
    panel.grid.major.x = element_blank(),
    plot.title = element_text(size = 14),
    plot.subtitle = element_text(color = "grey50", size = 10)
  ) +
  labs(
    title = "Arrest Rate by Day of the Month (1-31) for Cars Stopped in NYC",
    subtitle = "The trend seems to be stable across every day",
    x = "Day of the Month",
    y = "Arrest Rate (%)"
  )
```

What can we say?

- **Key finding**: Arrest rates are broadly flat across days 1 through 31, showing no meaningful trend.
- **Interpretation**: The flat pattern across days 1-28 indicates that the day-of-month variable carries no genuine enforcement signal.
- **Proposed Action**: Exclude `day` from our models.

```{r dow-vs-arrested-ratio}
daily_arrest_rate <- vehstops %>%
  filter(!is.na(dow), !is.na(arrested)) %>%
  group_by(dow) %>%
  summarise(
    total_stops = n(),
    arrests = sum(arrested == 1),
    rate = arrests / total_stops,
    .groups = "drop"
  ) %>%
  mutate(
    dow_label = factor(dow, levels = 1:7,
                       labels = c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun")),
    prev_rate = lag(rate),
    delta = (rate - prev_rate) * 100,
    label_text = ifelse(
      is.na(prev_rate),
      scales::percent(rate, accuracy = 0.1),
      paste0(scales::percent(rate, accuracy = 0.1),
             "\n(", ifelse(delta > 0, "+", ""), round(delta, 1), " pts)")
    )
  )

daily_arrest_rate %>%
  ggplot(aes(x = dow_label, y = rate, group = 1)) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_point(color = "steelblue", size = 3) +
  geom_text(aes(label = label_text), vjust = -1.0, size = 3, lineheight = 0.8) +
  scale_y_continuous(
    labels = scales::percent,
    limits = c(0, max(daily_arrest_rate$rate) * 1.3),
    expand = expansion(mult = c(0, 0.1))
  ) +
  theme_minimal() +
  theme(
    panel.grid.major.x = element_blank(),
    plot.title = element_text(size = 14),
    plot.subtitle = element_text(color = "grey50", size = 10)
  ) +
  labs(
    title = "Arrest Rate by Day of the Week for Cars Stopped in NYC",
    subtitle = "TODO",
    x = "Day of the Week",
    y = "Arrest Rate (%)"
  )
```

What can we say?

- **Key finding**: The line chart displays a gradual, monotonic rise in the weekly arrest rate as we approach the weekend. The probability of an arrest given a stop begins at its minimum on Monday at $3.0\%$, hovers around $3.1\%$ to $3.2\%$ through Thursday, and then experiences a clear upward lift starting on Friday ($3.5\%$) before reaching its absolute maximum on Saturday at $3.8\%$ ($+0.3$ percentage points).
- **Interpretation**: The overall variation across the week is exceptionally tight, spanning a narrow range of only $0.8$ percentage points between the Monday low and the Saturday peak. From a predictive perspective, this tells us that the day of the week carries a relatively weak standalone signal compared to spatial or hourly features. The stable baseline during the working week suggests that routine commuting days present a highly uniform threshold for arrest conversions.
- **Comment**: When comparing this trajectory to the two-wheeled vehicle analysis, we notice a shared behavioral shift toward the weekend. Both vehicle types experience an absolute maximum arrest rate on Saturday, which suggests a universal operational pattern where nighttime driving, leisure travel, and weekend enforcement strategies alter the composition of stops. Interestingly, while two-wheelers saw a fluctuating mid-week peak on Tuesday, standard car stops follow a smoother, more predictable ramp-up, highlighting the steady transition from structured weekday traffic to weekend police behaviors.
- **Proposed Action**: We should treat the day of the week as a categorical factor variable using dummy encoding within our predictive pipeline. Given the narrow global variation, we will evaluate whether collapsing the feature into a simplified binary indicator, such as `is_weekend` (isolating Friday through Sunday), preserves the predictive signal while creating a more parsimonious model structure.

### Hour and Time Bracket

```{r hour-vs-arrested-ratio}
hourly_arrest_rate <- vehstops %>%
  filter(!is.na(hour), !is.na(arrested)) %>%
  group_by(hour) %>%
  summarise(
    total_stops = n(),
    arrests = sum(arrested == 1),
    rate = arrests / total_stops,
    .groups = "drop"
  )

hourly_arrest_rate %>%
  ggplot(aes(x = hour, y = rate)) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_point(color = "steelblue", size = 2) +
  scale_x_continuous(breaks = 0:23) +
  scale_y_continuous(labels = scales::percent, limits = c(0, NA),
                     expand = expansion(mult = c(0, 0.15))) +
  theme_minimal() +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    plot.title = element_text(size = 14),
    plot.subtitle = element_text(color = "grey50", size = 10)
  ) +
  labs(
    title = "Arrest Rate by Hour of Day for Cars Stopped in NYC",
    subtitle = "TODO",
    x = "Hour of Day (0-23)",
    y = "Arrest Rate (%)"
  )
```

What can we say?

- **Key finding**: The line plot underlines a distinct time-dependent pattern in the arrest rate resulting from stops across a $24$-hour cycle. We observe a global minimum at $7:00$ ($07:00$ AM) where the arrest rate drops to around $0.6\%$, followed by a steady climb to a minor afternoon peak around $14:00$ ($02:00$ PM) near $2.9\%$. The absolute maximum occurs during the late-night and early-morning hours, specifically peaking at $3:00$ ($03:00$ AM) with an arrest rate reaching approximately $5.3\%$.
- **Interpretation**: This strong diurnal fluctuation suggests that our exploratory data analysis is heavily influenced by variations in driver behavior, law enforcement patterns, or both depending on the hour of the day. The steep descent between $3:00$ and $7:00$ and the rapid ascent between $16:00$ and $19:00$ reveal that the conditional probability of an arrest given a stop is highly dynamic, making the exact time of the encounter a highly informative feature.
- **Comment**: When comparing this curve to the two-wheeled vehicle analysis, we observe a remarkably similar U-shaped temporal structure, although the four-wheel peak is compressed at a lower absolute maximum ($5.3\%$ versus nearly $11\%$). The low arrest rates during the morning commute hours (from $6:00$ to $10:00$) suggest that law enforcement actions during this window might focus on high-volume traffic management and routing. Conversely, the high plateau maintained from $19:00$ to $3:00$ likely reflects stops related to more severe nighttime offenses, such as operating a vehicle under the influence or criminal contraband stops.
- **Proposed Action**: We may opt for the engineered `time_bracket` predictor as our primary spatial-temporal input to capture this prominent U-shape in a parsimonious way. However, given the sharp changes visible during the morning drop and evening rise, we will test whether a continuous spline function using `hour` inside a Generalized Additive Model framework fits the non-linear transition thresholds significantly better than discrete categorical buckets.

Let us look at our grouping now

```{r time-bracket-vs-arrested-ratio}
ratio_by_cat(
  vehstops, arrested, time_bracket,
  "Time bracket", "Arrest Rate (%)", "Arrest ratio by Custom Time Bracket for Cars Stopped in NYC",
  style = "stack", legend_title = "Stop ended in arrest?", target_labels = c("No", "Yes"),
  subtitle = "TODO"
)
```

What can we say?

- **Key finding**: The ordering of custom brackets by arrest rate precisely mirrors the expected patterns extracted from the continuous diurnal curve, sorting from lowest to highest conversion probability: Morning commute ($1.0\%$), Early morning ($1.4\%$), Midday ($2.6\%$), Afternoon/evening ($3.6\%$), Late night ($4.9\%$), and Night ($5.1\%$). The nighttime blocks show an arrest probability that stands substantially above the global mean baseline.
- **Interpretation**: This step-up distribution confirms that our discrete custom buckets act as highly stable aggregators of the underlying hourly timeline. The progression validates that the relative risk of an arrest given a stop increases fivefold from the morning commute minimum to the night maximum, proving that the temporal circumstance of a vehicle stop exerts a powerful structural influence on the target outcome.
- **Comment**: When comparing these ratios to the two-wheeled vehicle analysis, we note that standard car stops follow the exact same monotonic risk curve as the day progresses, though the values are compressed into a lower range. Even during the peak window in the Night bracket, the absolute majority of interactions ($94.9\%$) still do not result in an arrest. We can argue that, while the relative risk surges significantly during late hours, an encounter ending in an arrest remains a clear minority event across all examined operational blocks.
- **Proposed Action**: We choose `time_bracket` as the primary hour-related categorical predictor in our final arrest classification models. Because the arrest rate scales linearly across these sequential blocks, this custom categorization will allow our framework to absorb the non-monotonic temporal pattern efficiently while avoiding the excessive degrees of freedom associated with the raw $24$-hour variable.

## Driver related features

### Reported Age

```{r age-bracket-vs-target}
ratio_by_cat(
  vehstops, arrested, age_bracket,
  "Reported Age Bracket", "Arrest Rate (%)", "Arrest Ratio by Reported Age Bracket for Cars Stopped in NYC",
  style = "stack", legend_title = "Stop ended in arrest?", target_labels = c("No", "Yes"),
  subtitle = "TODO",
  label_vjust = 0.5,
)
```

```{r confirm-age-bracket-vs-target-ratio}
vehstops %>%
  group_by(age_bracket, arrested) %>%
  summarise(count = n(), .groups = "drop_last") %>%
  mutate(percentage = count / sum(count))
```

What can we say?

- **Key finding**: This plot shows a clear demographic trend where the arrest rate generally increases as the reported age bracket gets younger, reaching its absolute peak in the $25–34$ cohort. The lowest arrest rate is found within the $65+$ demographic at approximately $0.66\%$ (specifically $520$ arrests). The highest ratio belongs to the $25–34$ bracket, where the arrest rate reaches $4.57\%$, which also accounts for the absolute highest volume of individuals arrested with $28821$ instances. The $18–24$ group closely follows with a $4.38\%$ arrest rate ($14380$ arrests). Unlike the 2-wheeled vehicle case where the Under $18$ bracket was the distinct peak, for 4-wheeled vehicles, the under $18$ cohort sits at a lower rate of $2.99\%$ ($278$ arrests), while the Unknown category shows a $1.93\%$ arrest rate ($2291$ arrests).
- **Interpretation**: The data demonstrates a non-linear but highly pronounced age effect where young adults (ages $18$ to $34$) face the highest conditional probability of arrest upon being stopped in a 4-wheeled vehicle. This differs notably from the 2-wheeled analysis where minors (Under $18$) exhibited the highest vulnerability to arrest, suggesting that the dynamics of vehicle ownership, licensing compliance, or enforcement behavior shift considerably between vehicle classes.
- **Comment**: When we look qualitatively at these findings, the contrast between the two analyses is striking. For 2-wheeled vehicles, the younger brackets (especially minors) were heavily penalized, whereas for cars, the risk concentrates heavily on young adults in their prime driving years. The sheer volume of stops and arrests in the $25–34$ bracket highlights that this group dominates the absolute impact on the judicial system. The lower relative arrest rate for minors in 4-wheeled vehicles might reflect stricter initial barriers to entry for driving cars compared to scooters or motorcycles, which yields a different subset of interactions with law enforcement.
- **Proposed Action**: We want to include `age_bracket` as a predictor in our classification models instead of just using the raw continuous `age` feature. Given the structural differences in how age interacts with the target variable across vehicle types, we should also explore an interaction term between `age_bracket` and vehicle type or build separate stratified models to preserve these distinct behavioral patterns.

### Reported Sex

```{r sex-vs-target, fig.width=6}
ratio_by_cat(
  vehstops, arrested, sex,
  "Reported sex", "Arrest Rate (%)", "Arrest Ratio by Reported sex for Cars Stopped in NYC",
  style = "stack", legend_title = "Stop ended in arrest?", target_labels = c("No", "Yes"),
  subtitle = "TODO",
)
```

```{r confirm-sex-vs-target-ratio}
vehstops %>%
  group_by(sex, arrested) %>%
  summarise(count = n(), .groups = "drop_last") %>%
  mutate(percentage = count / sum(count))
```

What can we say?

- **Key finding**: This highlights a clear disparity in arrest rates across reported sex categories for 4-wheeled vehicle stops. Female drivers (`F`) experience the lowest arrest rate at $1.51\%$ ($5966$ arrests), followed by the unknown or unrecorded category (`U`) at $2.06\%$ (with $276$ arrests). Male drivers (`M`) are subject to the highest rate, with $3.70\%$ of their total stops resulting in an arrest. In fact, the total Male driver arrests stand heavily at $64333$, which represents the vast majority of the overall arrest volume.
- **Interpretation**: The ratio of arrest, given a stop, **is more than twice as high for male drivers compared to female** drivers. This variation affects our current analysis by demonstrating that the proportional gap observed in the 2-wheeled analysis persists strongly in the 4-wheeled domain, confirming that reported sex remains a key predictor that needs to be included in our statistical models.
- **Comment**: The higher arrest rate among male drivers across both vehicle types aligns with broader demographic trends in traffic violations, where they are statistically associated with riskier driving behaviors or more severe infractions. Interestingly, the relative arrest rates are generally lower across all categories for 4-wheeled vehicles than for 2-wheeled vehicles (e.g., $3.70\%$ vs $5.7\%$ for males), but the sheer volume of 4-wheeled stops means that the absolute impact on law enforcement resources is driven entirely by cars. The Unknown category continues to show an intermediate rate, which suggests that incomplete data entry might behave differently depending on the specific circumstances of the encounter.
- **Proposed Action**: We will try to understand the importance of this component, because the low overall baseline association with the target and the heavily imbalanced nature of this feature (especially with the dominant volume of male drivers) may become tricky in the modeling phase. We should consider evaluating feature importance metrics or using techniques like downsampling to ensure our models do not over-rely on the dominant demographic class.

### Reported Ethnicity

```{r ethnicity-vs-target, fig.width=6}
ratio_by_cat(
  vehstops, arrested, ethnicity,
  "Reported ethnicity", "Arrest Rate (%)", "Escalation ratio by Reported Ethnicity for Cars Stopped in NYC",
  style = "stack", legend_title = "Stop ended in arrest?", target_labels = c("No", "Yes"),
  subtitle = "TODO",
)
```

```{r confirm-ethnicity-vs-target-ratio}
vehstops %>%
  group_by(ethnicity, arrested) %>%
  summarise(count = n(), .groups = "drop_last") %>%
  mutate(percentage = count / sum(count))
```

What can we say?

- **Key finding**: Black drivers represent the largest slice with $656109$ total stops and the highest absolute number of arrests at $32954$ ($5.02\%$). Hispanic drivers experience the second highest escalation rate with $26330$ arrests out of $565531$ stops ($4.66\%$). In contrast, White drivers account for $426594$ stops and $4912$ arrests ($1.15\%$). Finally, Asian / Pacific Islander, Other, and Unknown categories maintain the lowest arrest rates, all remaining near or below $1.51\%$.
- **Interpretation**: The conditional probability of a stop resulting in an arrest is more than four times higher for Black and Hispanic drivers compared to White drivers. These rates are calculated from a massive sample size exceeding two million total records, indicating they are highly stable statistical patterns and completely isolated from small sample variance.
- **Comment**: When comparing these results to the 2-wheeled vehicle case, we observe a noticeable structural shift in volume and magnitude, though the relative disparity remains stark. For 2-wheeled vehicles, Hispanic operators faced the absolute highest arrest rate ($7.24\%$), whereas in the 4-wheeled domain, Black drivers experience both the highest total stop volume and the highest escalation rate ($5.02\%$). While the overall baseline arrest rates are lower for cars than for 2-wheelers across all demographics, the proportional gap between minority groups and White drivers has actually widened significantly from roughly a two fold difference to a four fold difference, signaling a distinct operational dynamic in how car stops escalate.
- **Proposed Action**: This severe imbalance indicates that reported ethnicity is a critical factor that must be included in our models. Due to the high disparity and potential proxy correlations with geographic factors, we should assess downstream model fairness using demographic parity metrics to understand how including `ethnicity` impacts predictive equity across different vehicle classes.

------------------------------------------------------------------------

# Final touches

## Re-Ordering Features

We re-order the features for better display

```{r re-order-dataset}
vehstops_ord <- vehstops %>%
  select(
    year, month, hot_season, weekday,
    hour, time_bracket,
    area, boro,
    sex, ethnicity, 
    age, age_bracket,
    arrested
  )
```

Here is the final version.

```{r summary-final-dataset}
print(dfSummary(vehstops_ord), method="render")
```

## Saving the dataset

```{r save-dataset}
write.csv(vehstops_ord, "data/vehiclestops_4_wheels_final.csv", row.names = FALSE)
```
