---
title: "Predicting Excessive Enforcement Patterns in Police Behavior After Vehicles Stops in New York"
subtitle: "EDA and Feature Engineering - 4 Wheels Vehicles"
author: "Gianmaria Pizzo"
output:
  html_document:
    fig_caption: yes
    theme: flatly
    highlight: pygments
    code_folding: show
    kee_md: yes
    toc: yes
    toc_depth: 4
    number_sections: no
    toc_float:
      smooth_scroll: no
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

These `.Rmd` files represent the project submission for the course [Statistical Inference and Learning](https://www.unive.it/data/course/521944) by Professor [Cristiano Varin](https://www.unive.it/data/people/5592728) at [Ca' Foscari University of Venice](https://www.unive.it/) for the year 2025/2026.

# Problem Statement

## Goal of this rmarkdown file

This notebook serves as the primary Exploratory Data Analysis (EDA) and (final) Feature Engineering phase for the project. In particular, we focus solely on **4-Wheeled Vehicles**.

The goal is to identify the features that characterize **arrests**.

We approach the dataset through three methodological phases to ensure a robust base for modeling:

1.  **Univariate Analysis**. We begin by analyzing the distributions of individual predictors to understand whether the data shows quality issues, outliers, rare categories, etc...

2.  **Bivariate Analysis**. We evaluate the relationship between potential predictors and our binary target (`arrested`). Given the categorical nature of our data, we will mainly uses **Cramer’s V** as a starting point to measure of effect size. This stage should also point out if predictors have some kind of relation.

Under these premises, we will plot many aspects of the data. Therefore, the way we decided to guide the reader through the Exploratory Data Analysis, is based on four key points that we try to include for any plot:

- **Key finding**: This paragraph describes the main takeaway of the means of visualization employed. This should be a concise description of what we can gather from the plots.
- **Interpretation**: Here we deal with an objective interpretation of our finding, such as understanding how our EDA is affected by it.
- **Comment**: After a formal description, we deal with a more informal discussion that should highlight the qualitative aspects of the analysis. This paragraph is a fusion of previous considerations that we want the reader to be aware of, and a space for us to reason.
- **Proposed Action**: Finally, we switch to a more pragmatic discussion to understand "what to do next", on the basis of what was previously presented. Without claim of completeness, we will name some best-practices, guidelines or ideas about what is best for our goal. We may also limit ourselves to inform the reader that we decided to take no action at all or post-pone it after further investigation.

Note that we do not always include all four points.

## Important Note

We will, sometimes, cut short of the commentary when we make similar discovery with respect to the 2-wheels analysis. This is needed not to repeat ourselves if we go down the same conclusions. In this case, we will omit to elaborate.

------------------------------------------------------------------------

# Environment

The first to do, is to build a reproducible environment for our project.

```{r}
requirements=c( 
  "remotes", "summarytools", "forcats", "scales", "purrr",
  "stringr", "plyr", "dbplyr", "dplyr", "dtplyr", "tidyr", "tidyselect",
  "timeDate", "timechange", "tzdb", "hms", "readr", "tigris", "lsr", "corrplot",
  "lubridate", "prettyunits", "RColorBrewer", "viridis", "rcartocolor", "ggbeeswarm", "emmeans",
  "sf", "plotly", "ggplot2", "ggmap", "ggdensity", "UpSetR", "patchwork"
)
```

```{r install-requirements, include=FALSE}
for (req in requirements){
  if (!require(req, character.only = TRUE)){
      install.packages(req)
  }
}
```

Note that some of the packages may not be used, but to avoid any error related to dependencies, we are going to keep all of these libraries. Now, the loading of all packages should be immediate.

```{r load-requirements, include=FALSE}
for (req in requirements){
  if (!require(req, character.only = TRUE)){
      library(req, character.only = TRUE)
  }
}
```

```{r remove-vars, include=FALSE}
rm(requirements, req)
```

## Plot Functions

Here we add the plot functions to make some kinds of visualizations reproducible.

### Univariate Plot Functions

```{r univariate-plot-helpers}
# ==============================================================================
# 1. Plots for Univariate Analysis
# ==============================================================================

#' @title Plot Binary Bar Chart
#' 
#' @description Creates a simple bar chart counting the occurrences of a binary feature, 
#' with labels showing both the count and the percentage.
#' 
#' @param DF The input dataframe.
#' @param FEATURE The binary column to plot (unquoted).
#' @param OPTION_0 The text label for the "0" or "FALSE" category.
#' @param OPTION_1 The text label for the "1" or "TRUE" category.
#' @param X_LABEL Label for the X axis.
#' @param Y_LABEL Label for the Y axis.
#' @param TITLE Main title of the plot.
#' @param pad Numeric padding for the top of the Y axis to fit text labels.
#' @param text_vjust Vertical adjustment for the text labels on top of bars.
#' 
#' @return A ggplot2 object.
plot_binary_bar <- function(DF, FEATURE, OPTION_0, OPTION_1, X_LABEL, Y_LABEL, TITLE, pad = 0.15, text_vjust = -0.3) {
  # Standardize the binary feature and calculate percentages
  df_sum <- DF %>%
    mutate(
      .raw  = {{ FEATURE }},
      .feat = dplyr::case_when(
        as.character(.raw) %in% c("1","TRUE","true")  ~ 1L,
        as.character(.raw) %in% c("0","FALSE","false") ~ 0L,
        TRUE ~ NA_integer_
      )
    ) %>%
    filter(!is.na(.feat)) %>%
    count(.feat, name = "n") %>%
    mutate(
      .lab = factor(.feat, levels = c(0, 1), labels = c(OPTION_0, OPTION_1)),
      # Calculate percentage and create combined label
      pct = n / sum(n),
      label_text = paste0(scales::comma(n), "\n(", scales::percent(pct, accuracy = 0.1), ")")
    )
  
  # Get the maximum bar height
  ymax <- max(df_sum$n, na.rm = TRUE)
  
  # Build the ggplot
  ggplot(df_sum, aes(x = .lab, y = n, fill = .lab)) +
    geom_col(width = 0.8, show.legend = FALSE) +
    # Updated to use label_text
    geom_text(aes(label = label_text), vjust = text_vjust, lineheight = 0.8) +
    # Expand the Y axis slightly for the taller multi-line labels
    scale_y_continuous(limits = c(0, ymax * (1 + pad)), expand = expansion(mult = c(0, 0.02))) +
    coord_cartesian(clip = "off") +
    theme_minimal() +
    theme(panel.grid.major.x = element_blank(), plot.margin = margin(t = 20, r = 5.5, b = 5.5, l = 5.5)) +
    scale_fill_viridis_d(option = "mako", begin = 0.3, end = 0.8) +
    labs(x = X_LABEL, y = Y_LABEL, title = TITLE)
}


#' @title Plot Univariate Bar Chart (Categorical)
#' 
#' @description Creates a bar chart counting the occurrences of a categorical feature with 2+ classes.
#' Allows dynamic ordering (by frequency or original/alphabetical order).
#' 
#' @param DF The input dataframe.
#' @param FEATURE The categorical column to analyze (unquoted).
#' @param X_LABEL Label for the X axis.
#' @param Y_LABEL Label for the Y axis (default is "Count of Stops").
#' @param TITLE Main title of the plot.
#' @param SUBTITLE Subtitle of the plot.
#' @param top_n Optional integer; if set, keeps only the top N most frequent categories.
#' @param order_by Method to order the X axis: "freq" (descending count) or "none" (alphabetical/original).
#' @param pad Numeric padding for the top of the Y axis to fit text labels.
#' @param text_vjust Vertical adjustment for the text labels on top of bars.
#' 
#' @return A ggplot2 object.
plot_univariate_cat <- function(DF, FEATURE, X_LABEL, Y_LABEL = "Count of Stops", TITLE, SUBTITLE, top_n = NULL, order_by = c("freq", "none"), pad = 0.12, text_vjust = -0.35) {
  
  order_by <- match.arg(order_by)
  
  # Compute counts
  df_sum <- DF %>%
    filter(!is.na({{ FEATURE }})) %>%
    count({{ FEATURE }}, name = "n")
  
  # Filter top n results
  if (!is.null(top_n)) {
    df_sum <- df_sum %>% slice_max(order_by = n, n = top_n)
  }
  
  # X axis order (either frequency or std)
  cat_levels <- switch(
    order_by,
    freq = df_sum %>% arrange(desc(n)) %>% pull({{ FEATURE }}),
    none = sort(unique(df_sum %>% pull({{ FEATURE }})))
  )
  
  # Apply sorting
  df_sum <- df_sum %>%
    mutate(.feat_f = factor(as.character({{ FEATURE }}), levels = as.character(cat_levels)))
  
  # Max height for dynamic padding
  ymax <- max(df_sum$n, na.rm = TRUE)

  # Construct plot
  ggplot(df_sum, aes(x = .feat_f, y = n, fill = .feat_f)) +
    geom_col(width = 0.8, show.legend = FALSE) +
    geom_text(aes(label = scales::comma(n)), vjust = text_vjust, size = 3) +
    scale_y_continuous(limits = c(0, ymax * (1 + pad)),
                       expand = expansion(mult = c(0, 0.02))) +
    coord_cartesian(clip = "off") +
    theme_minimal() +
    theme(
      plot.title = element_text(size = 14),
      plot.subtitle = element_text(size = 10, color = "grey50"),
      axis.text.x = element_text(angle = 45, hjust = 1),
      panel.grid.major.x = element_blank(),
      plot.margin = margin(t = 10, r = 5.5, b = 5.5, l = 5.5)) +
    scale_fill_viridis_d(option = "mako", begin = 0.3, end = 0.8) +
    labs(x = X_LABEL, y = Y_LABEL, title = TITLE, subtitle = SUBTITLE)
}
```

### Bivariate Plot Functions

```{r biviariate-plot-helpers}
# ==============================================================================
# 2. Plots for Bivariate Analysis
# ==============================================================================

#' @title Calculate Cramer's V
#' 
#' @description Calculates the Cramer's V measure of association between two 
#' categorical vectors. Returns a value between 0 (no association) and 1 (perfect association).
#' 
#' @param x First categorical vector.
#' @param y Second categorical vector.
#' 
#' @return A numeric value representing Cramer's V.
cramer_v <- function(x, y) {
  # Contingency Table
  tbl <- table(x, y)
  
  # Execute Chi2 Test
  chi2 <- suppressWarnings(chisq.test(tbl, correct = FALSE)$statistic)
  
  # Compute dim and min value
  n <- sum(tbl)
  min_dim <- min(nrow(tbl) - 1, ncol(tbl) - 1)
  
  # Avoid math error for 1 categ
  if(min_dim == 0) return(0) 
  
  # Cramer's formula
  v <- sqrt(chi2 / (n * min_dim))
  return(as.numeric(v))
}

# ==============================================================================
# Categorical "Correlation" Heatmap - Association Matrix based on Cramer's V
# ==============================================================================

#' @title Plot Categorical Correlation Matrix (Cramer's V)
#' 
#' @description Creates a correlation-like heatmap for multiple categorical variables
#' using Cramer's V. Answers "How much do these categorical features overlap with each other?"
#' 
#' @param DF The input dataframe.
#' @param FEATURES A character vector containing the names of the categorical columns to compare.
#' @param TITLE Main title of the plot.
#' 
#' @return A ggplot2 object.
plot_categorical_correlation <- function(DF, FEATURES, TITLE = "Association Matrix for Categorical Predictors (Cramer's V)", text_size = 3) {
  
  # Select columns and drop NAs
  df_sub <- DF %>% 
    select(all_of(FEATURES)) %>%
    drop_na()
  
  n_vars <- ncol(df_sub)
  
  # Initialize matrix and set main diag to 1s
  mat <- matrix(1, nrow = n_vars, ncol = n_vars)
  rownames(mat) <- colnames(df_sub)
  colnames(mat) <- colnames(df_sub)

  # Compute Cramer's V for each possible pair
  for(i in 1:(n_vars - 1)) {
    for(j in (i + 1):n_vars) {
      v <- cramer_v(df_sub[[i]], df_sub[[j]])
      # Symmetric 
      mat[i, j] <- v
      mat[j, i] <- v
    }
  }
  
  # Make matrix compatible with ggplot2
  mat_long <- as.data.frame(as.table(mat))
  colnames(mat_long) <- c("Var1", "Var2", "CramerV")
  
  # Create heatmap
  ggplot(mat_long, aes(x = Var1, y = Var2, fill = CramerV)) +
    geom_tile(color = "white", width = 1, height = 1) +
    # Rounding
    geom_text(aes(label = round(CramerV, 2)), size = text_size) +
    # Setting colors
    scale_fill_gradient(low = "whitesmoke", high = "steelblue", limits = c(0, 1), name = "Cramer's V") +
    labs(title = TITLE, x = "", y = "") +
    theme_minimal() +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1, size = 10),
      axis.text.y = element_text(size = 10),
      panel.grid = element_blank()
    )
}

# ==============================================================================
# Plotting of Targets' Rate Against Categorical Predictors
# ==============================================================================

#' @title Plot Target Ratio by Category
#' 
#' @description Creates a bivariate bar chart showing the relationship between a binary 
#' TARGET and a single categorical variable (CAT). It supports two visual styles: 
#' a 100% stacked bar chart ("stack") or a simple bar chart showing the rate of the target ("rate").
#' 
#' @param DF The input dataframe.
#' @param TARGET The binary column to analyze (unquoted). It will be standardized to 0/1.
#' @param CAT The categorical column for the X axis (unquoted).
#' @param x_title Label for the X axis.
#' @param y_title Label for the Y axis.
#' @param title Main title of the plot.
#' @param target_labels A character vector of length 2 defining the labels for 0 and 1 (e.g., c("Not escalated", "Escalated")).
#' @param style The visual style of the plot: "stack" (100% stacked bars) or "rate" (single bar for target = 1).
#' @param order_by Method to order the X axis categories: by "rate" (ascending target rate), "freq" (ascending frequency), or "none" (alphabetical/original).
#' @param percent_accuracy Accuracy for the percent labels (e.g., 0.1 for one decimal place).
#' @param colors A character vector of length 2 defining the hex color codes for the 0 and 1 categories respectively.
#' 
#' @return A ggplot2 object.
ratio_by_cat <- function(
    DF, TARGET, CAT, x_title, y_title, title, 
    target_labels = c("Not escalated", "Escalated"), 
    style = c("stack", "rate"), order_by = c("rate", "freq", "none"), 
    percent_accuracy = 0.1, palette = "Paired",
    colors = c("grey80", "steelblue"), legend_title = "Legend",
    subtitle = NULL, bar_width = 0.7, aspect_ratio = 1.0, label_vjust = 0.5,
    text_size = 2
) {
  
  # Validate arguments
  style    <- match.arg(style)
  order_by <- match.arg(order_by)
  
  # Standardize target and extract the category
  dat <- DF %>%
    mutate(
      .t_raw = {{ TARGET }},
      .t = case_when(
        as.character(.t_raw) %in% c("1","TRUE","true") ~ 1L,
        as.character(.t_raw) %in% c("0","FALSE","false") ~ 0L,
        TRUE ~ NA_integer_
      ),
      .cat = {{ CAT }}
    ) %>%
    filter(!is.na(.t), !is.na(.cat))
  
  # Compute rates and frequencies
  rates <- dat %>% 
    group_by(.cat) %>%
    summarise(total = n(), rate = mean(.t == 1), .groups = "drop")
  
  # Determine ordering
  cat_levels <- switch(
    order_by,
    rate = rates %>% arrange(rate) %>% pull(.cat),
    freq = rates %>% arrange(total) %>% pull(.cat),
    none = unique(dat$.cat)
  )
  
  # Build the plot based on style
  if (style == "stack") {
    stack <- dat %>%
      count(.cat, .t, name = "n") %>%
      group_by(.cat) %>%
      mutate(ratio = n/sum(n)) %>%
      ungroup() %>%
      mutate(
        .cat_f = factor(as.character(.cat), levels = as.character(cat_levels)),
        .t_f = factor(.t, levels = c(0,1), labels = target_labels)
      )
    
    ggplot(stack, aes(x = .cat_f, y = ratio, fill = .t_f)) +
      geom_col(position = "fill", width = bar_width) +
      geom_text(
        aes(label = scales::percent(ratio, accuracy = percent_accuracy)),
        position = position_fill(vjust = label_vjust), 
        size = text_size
      ) +
      scale_y_continuous(labels = scales::percent) +
      scale_fill_manual(values = setNames(colors, target_labels)) +
      labs(x = x_title, y = y_title, fill = legend_title, title = title, subtitle = subtitle) +
      theme_minimal() +
      theme(
        axis.text.x = element_text(angle = 45, hjust = 1),
        panel.grid = element_blank(),
        plot.subtitle = element_text(color = "grey50"),
        aspect.ratio = aspect_ratio
      )
      
  } else {
    # Rate style
    rates_data <- rates %>%
      mutate(.cat_f = factor(as.character(.cat), levels = as.character(cat_levels)))
    
    ggplot(rates_data, aes(x = .cat_f, y = rate)) +
      geom_col(width = 0.8, fill = colors[2]) +
      geom_text(
        aes(label = scales::percent(rate, accuracy = percent_accuracy)),
        vjust = -0.35, size = text_size
      ) +
      scale_y_continuous(labels = scales::percent, expand = expansion(mult = c(0, 0.12))) +
      coord_cartesian(clip = "off") +
      labs(x = x_title, y = y_title, title = title, subtitle = subtitle) +
      theme_minimal() +
      theme(
        axis.text.x = element_text(angle = 45, hjust = 1),
        panel.grid = element_blank(),
        plot.subtitle = element_text(color = "grey50")
      )
  }
}
```

------------------------------------------------------------------------

Let us load the data now, and remove the extra feature `veh_category` because we are **exclusively analyzing 4 wheels vehicles now**.

```{r load-data-drop-veh-categ}
vehstops = read.csv("data/vehiclestops_4_wheels.csv")

vehstops <- vehstops %>%
  select(-veh_categ)
```

So now we have the following dataset, that we present with the usual summary

```{r summary-data}
print(dfSummary(vehstops), method="render")
```

This preview is already interesting.

------------------------------------------------------------------------

# Univariate EDA

## Target

```{r enforcement-dist}
p1 <- plot_binary_bar(
  vehstops, arrested, 
  "No", "Yes",
  "Was the driver arrested?", "Count of Stops (Percentage %)", ""
)

row1 <- p1

row1 + 
plot_annotation(title = "Distribution of Arrests for Cars Stopped in NYC") &
theme(legend.position = "bottom") 
```

```{r, include=FALSE}
rm(p1, row1)
```

What can we say?

What can we say?

- **Key finding**: The dataset for 4-wheels vehicles **exhibits extreme class imbalance** that is even more pronounced than what we observed in the 2-wheels analysis. Arrests are highly sparse, representing only approximately $3.3\%$ of the outcomes ($70,600$ "Yes" out of $2,157,445$ total records) compared to the $5.6\%$ rate found in 2-wheeled vehicle stops.
- **Interpretation**: The target variable constitutes a rare event when we contemplate the stops of 4-wheels vehicles. While the overall volume of stops is drastically higher for 4-wheels vehicles, an arrest remains an infrequent occurrence with respect to the total volume of police stops.
- **Comment**: The high class disparity *makes accuracy a limiting metric* for our models' evaluation, here too.
- **Proposed Action**: Same as the 2-wheels analysis.

## Geospatial predictors

### Area

```{r count-stops-by-area}
plot_univariate_cat(
  DF = vehstops, 
  FEATURE = area, 
  X_LABEL = "Area", 
  TITLE = "Distribution of Car Stops by Area in NYC",
  SUBTITLE = "Brooklyn is where the most amount of vehicles are stopped",
  order_by = "freq"
)
```

What can we say?

- **Key finding**: Since the bars are distributed by volume, in descending order, we can clearly observe a "ladder" trend in volume. We still have variance, but thanks to the volume we now have a more balanced case. We observe two clear "high-volume" areas (Brooklyn with over $640$k stops and Queens with over $550$k stops) followed by two "medium-volume" areas (Manhattan with $418$k and the Bronx with $381$k) and a distinct "low-volume" area (Staten Island, $166$k).
- **Interpretation**: Compared to the 2-wheels analysis where Staten Island was a tiny outlier ($1.5$k), here it holds a substantial sample size ($166$k). The shifting proportions, particularly the rise of the Bronx into a solid medium-volume category, suggests that geographic location could be a strong predictor of the probability of being stopped (in case the targets' ratio distributed by area follows a specific pattern).
- **Comment**: Unlike the 2-wheels analysis, Staten Island is not anymore a very rare label, even if the distribution is the lowest among the 5 areas. We have a slightly lower risk of overfitting here, but it is still to early to explore this aspect.
- **Proposed Action**: Same as the 2-wheels analysis.

### Borough

```{r count-stops-by-boro}
plot_univariate_cat(
  DF = vehstops, 
  FEATURE = boro, 
  X_LABEL = "Borough", 
  TITLE = "Distribution of Car Stops by Borough in NYC",
  SUBTITLE = "A more fine-grained bracketing of areas shows that Brookly (South) and Bronx dominate for amount of stops"
)
```

What can we say?

- **Key finding**: The granular bracketing shows that "Brooklyn (South)" dominates with $398,062$ stops, closely followed by the "Bronx" at $381,239$ stops, whereas in the 2-wheels analysis "North Queens" was the dominant category. Except the top-two volume boroughs, the remaining boroughs seem to have an almost uniform (decreasing) distribution ranging between $166$k - $276$k.
- **Interpretation**: For what concerns granularity, we can draw the same conclsions as the 2-wheels case. On the other hand, since every category contains a massive amount of observations, we probably do not run the risk encountered in the 2-wheels context where low-frequency categories caused high variance and wide confidence intervals in our parameter estimates.
- **Comment**: Same as the 2-wheels analysis.
- **Proposed Action**: Same as the 2-wheels analysis.

### Latitude and Longitude for Stop Density

For what concerns these two predictors, we will use the to plot a spatial heat map of the stops with these values available.

```{r stop-density}
nyc_map <- counties(state = "NY", cb = TRUE) %>%
  filter(NAME %in% c("Bronx", "Kings", "New York", "Queens", "Richmond"))

bins <- 100
lon_breaks <- seq(-74.26, -73.71, length.out = bins + 1)
lat_breaks <- seq(40.48, 40.92, length.out = bins + 1)

grid_sum <- vehstops %>%
  filter(!is.na(longitude) & !is.na(latitude)) %>%
  mutate(
    xbin = cut(longitude, breaks = lon_breaks, include.lowest = TRUE, right = FALSE),
    ybin = cut(latitude, breaks = lat_breaks, include.lowest = TRUE, right = FALSE)
  ) %>%
  mutate(
    x = as.numeric(gsub(".*\\[(.*),.*", "\\1", xbin)) + (lon_breaks[2] - lon_breaks[1])/2,
    y = as.numeric(gsub(".*\\[(.*),.*", "\\1", ybin)) + (lat_breaks[2] - lat_breaks[1])/2
  ) %>%
  group_by(x, y) %>%
  summarise(n = n(), .groups = "drop")

ggplot() +
  geom_sf(data = nyc_map, fill = "white", color = "grey80", size = 0.2) +
  geom_tile(data = grid_sum, aes(x, y, fill = n), alpha = 0.8) +
  scale_fill_viridis_c(trans = "log10", name = "Stops (log10)", option = "mako", direction = -1) +
  coord_sf(xlim = c(-74.26, -73.71), ylim = c(40.48, 40.92)) +
  theme_minimal() +
  theme(panel.grid = element_blank()) +
  labs(title = "Heatmap of Cars Stops Density in NYC - 100 Bins", x = "Longitude", y = "Latitude")
```

```{r, include=FALSE}
rm(grid_sum, nyc_map, bins, lat_breaks, lon_breaks)
```

What can we say?

- **Key finding**: The heatmap reveals that, with respect to the 2-wheels case, the hotspots are the same, although the increase in volume is notable.
- **Interpretation**: Same as the 2-wheels analysis.
- **Proposed Action**: Same as the 2-wheels analysis.

## Time-related predictors

### Year

```{r count-stops-by-year}
vehstops %>%
  filter(!is.na(year)) %>%
  count(year) %>%
  mutate(year = as.numeric(as.character(year))) %>%
  arrange(year) %>%
  mutate(
    prev_n = lag(n), pct_change = (n - prev_n) / prev_n * 100,
    label_text = ifelse(
      is.na(pct_change), 
      comma(n), 
      paste0(comma(n), "\n(", ifelse(pct_change > 0, "+", ""), round(pct_change, 1), "%)")
    )) %>%
  ggplot(aes(x = factor(year), y = n, group = 1)) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_point(color = "steelblue", size = 3) +
  geom_text(aes(label = label_text), vjust = -0.8, nudge_x = -0.1, size = 3.5, lineheight = 0.8) +
  scale_y_continuous(labels = comma, limits = c(0, NA), expand = expansion(mult = c(0, 0.2))) +
  theme_minimal() +
  theme(
    panel.grid.major.x = element_blank(), 
    plot.title = element_text(size = 14),
    plot.subtitle = element_text(size = 10, color = "grey50")
    ) +
  labs(
    title = "Annual Volume of Stops for Cars in NYC",
    subtitle = "In 3 Years the Total Volume Increased by almost 200k Stops",
    x = "Year", y = "Count of Stops"
    )
```

What can we say?

- **Key finding**: There has been a steady, multi-year increase in the volume of police stops involving 4-wheeled vehicles, with the absolute number of encounters expanding from $611,734$ in $2023$ to $799,853$ in $2025$. Overall, the volume of stops does not show a very unbalanced distribution among the three years.
- **Interpretation**: The growth trajectory is non-linear but far more stable than the explosive surge seen in the 2-wheels analysis. The $21.9\%$ jump between $2023$ and $2024$ represents a significant upward trend, which is suggesting a broader intensification of traffic enforcement across the city. On the other hand, the $7.2\%$ increase in 2025 indicates that while the absolute volume of activity is still rising, the rate of acceleration is starting to stabilize.
- **Comment**: Same as the 2-wheels analysis.
- **Proposed Action**: Same as the 2-wheels analysis.

### Month

```{r count-stops-by-month}
month_data <- vehstops %>%
  filter(!is.na(month)) %>%
  count(month) %>%
  mutate(month = factor(month, levels = 1:12)) %>%
  arrange(month) %>%
  mutate(
    prev_n = lag(n), pct_change = (n - prev_n) / prev_n * 100,
    label_text = ifelse(
      is.na(pct_change), comma(n),
      paste0(comma(n), "\n(", ifelse(pct_change > 0, "+", ""), round(pct_change, 1), "%)"))
  )

month_data %>%
  ggplot(aes(x = month, y = n, group = 1)) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_point(color = "steelblue", size = 3) +
  geom_text(aes(label = label_text), vjust = -1, nudge_x = -0.1, size = 3, lineheight = 0.8) +
  scale_x_discrete(labels = month.abb) +
  scale_y_continuous(labels = comma, limits = c(0, NA), expand = expansion(mult = c(0, 0.3))) +
  theme_minimal() +
  theme(
    panel.grid.major.x = element_blank(),
    plot.title = element_text(size = 14),
    plot.subtitle = element_text(color = "grey50", size = 10)
  ) +
  labs(
    title = "Monthly Volume of Stops for Cars in NYC",
    subtitle = "Over 3 Years, Summer is when the Least Amount of Stops take place",
    x = "Month",
    y = "Count of Stops"
  )
```

```{r, include=FALSE}
rm(month_data)
```

What can we say?

- **Key finding**: The data exhibits a highly distinct monthly pattern for 4-wheels vehicles, contrasting sharply with the June to October summer plateau observed for bikers. Here, stop volumes peak in the early months of the year, reaching an absolute maximum in March at $208,661$ stops, before experiencing a plateau, starting in June, that bottoms out in September at $154,192$ stops (a $-7.4\%$ drop from August).
- **Interpretation**: Unlike 2-wheeled vehicles, **4-wheeled vehicles are stopped regardless of the season and the related weather conditions**. This explains the high volume baseline maintained in January ($204,603$) and February ($200,831$). The steady downward trajectory into June ($-16.4\%$) and July may suggest a shift in enforcement.
- **Comment**: Most of the conclusions we can make here come from the 2-wheels analysis. The interesting aspect is that the peaks are almost completely counter-cyclical to the 2-wheels plot. This may suggest that, for standard automobiles, the police behavior shows a mechanic that is entirely independent from seasonal weather. If this was true, we could not compare fairly the two datasets.
- **Proposed Action**: Same as the 2-wheels analysis.

Let us see this with respect to the year

```{r year-monthly-distribution}
plot_data <- vehstops %>%
  filter(!is.na(year) & !is.na(month)) %>%
  count(year, month) %>%
  mutate(month = factor(month, levels = 1:12)) %>%
  group_by(year) %>%
  mutate(
    n_jan = n[month == 1],
    n_max = max(n),
    is_max = n == n_max,
    pct_max = round((n - n_jan) / n_jan * 100, 0),
    pct_dec = round((n - n_max) / n_max * 100, 0),
    
    label_text = case_when(
      month == 1 ~ comma(n), 
      is_max ~ paste0(comma(n), "\n(", ifelse(pct_max > 0, "+", ""), pct_max, "%)"),
      month == 12 ~ paste0(comma(n), "\n(", pct_dec, "%)"),
      TRUE ~ ""
    )
  )

# Faceted plot
plot_data %>%
  ggplot(aes(x = month, y = n, group = 1)) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_point(color = "steelblue", size = 2) +
  geom_text(aes(label = label_text), vjust = -1, hjust = 0.2, size = 2.5, lineheight = 0.8) +
  facet_wrap(~year, nrow = 1) +
  scale_x_discrete(
    labels = c("J", "F", "M", "A", "M", "J", "J", "A", "S", "O", "N", "D"), 
    expand = expansion(mult = c(0.1, 0.1))
    ) +
  scale_y_continuous(labels = comma, limits = c(0, NA), expand = expansion(mult = c(0, 0.4))) +
  coord_cartesian(clip = "off") +
  theme_minimal() +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    plot.title = element_text(size = 14),
    plot.subtitle = element_text(color = "grey50", size = 10),
    strip.text = element_text(size = 11)
  ) +
  labs(
    title = "Monthly Volume of Stops, by Year, for Cars in NYC",
    subtitle = "Custom Growth/Decline analysis: January vs Max, and Max vs December",
    x = "Month",
    y = "Count of Stops"
  )
```

```{r, include=FALSE}
rm(plot_data)
```

What can we say?

- **Key finding**: There is a pronounced shifting trend in the intra-year volume distribution over the three-year period. In $2023$, the annual maximum occurred early in March ($64,991$, an $11\%$ increase from January) followed by a sharp drop to December ($38,193$, a $-41\%$ decline). In $2024$ the monthly amount of stops follows a slightly decreasing trend. By $2025$, the behavior changes, as the year begins at a much higher baseline peak in January ($81,793$). Then it shows a steep mid-year contraction through summer, and it ends with a surge (in October) that leads to a much softer winter decline by December ($62,748$, a $-23\%$ drop from the annual maximum).
- **Interpretation**: The data indicates that enforcement patterns for 4-wheels vehicles are having structural adjustments over the years, although the overall volume keeps increasing. The widening gap between the high-volume peaks and the low-volume troughs seen in the 2-wheels context does not apply here. Instead, the "off-season" drop in December is becoming significantly less severe, moving from $-41\%$ in $2023$ to $-23\%$ in $2025$, which implies a stabilization of late-year policing activities.
- **Comment**: The plot clearly shows that while the baseline (January) is rising substantially over the years. Furthermore, we notice that the *shape* of the annual trend is shifting from looking like a spring, to a 2-phase distribution with a notable winter increase. This shift confirms our hypothesis that seasonal **vehicle usage is not the primary factor** for stops. Instead, this is probably due to the policy shifts within the police department.
- **Proposed Action**: Here is impossible to include the same `hot_season` factor, as we would like to restrict the focus months over the June-September period. Therefore, we skip it.

### Dow/Weekday, Day and Doy

We have decided to exclude 'Day of Month' (`day`) and 'Day of Year' (`doy`) from our feature set to mitigate 'calendar artifact' bias.

As we can see below, the variation in month length creates false activity dips for end-of-month dates

```{r day-bias}
vehstops %>%
  filter(!is.na(day)) %>%
  count(day) %>%
  ggplot(aes(x = day, y = n)) +
  geom_line(color = "grey60", alpha = 0.5) +
  geom_smooth(method = "loess", color = "steelblue", se = FALSE) +
  theme_minimal() +
   theme(
    panel.grid.major.x = element_blank(),
    plot.title = element_text(size = 14),
    plot.subtitle = element_text(color = "grey50", size = 10)
  ) +
  labs(title = "Volume of Stops for Cars, by Day of the Month (1-31), in NYC",
       subtitle = "The drop towards 30-31 indicates it is not a behavior, but depends on the month!",
       x = "Day of the Month", y = "Count of Stops")
```

Furthermore, `doy` variable suffers from inconsistent data density.

```{r doy-incosistent-3-years-span}
plot_data <- vehstops %>%
  mutate(date = as.Date(paste(year, month, day, sep = "-"))) %>%
  count(date)

plot_data %>%
  ggplot(aes(x = date, y = n)) +
  geom_line(color = "grey70", alpha = 0.6, linewidth = 0.5) +
  geom_smooth(method = "loess", span = 0.2, color = "steelblue", se = FALSE, linewidth = 1.2) +
  scale_x_date(date_breaks = "3 months", date_labels = "%b %Y") +
  theme_minimal() +
  theme(
    panel.grid.minor = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(size = 14),
    plot.subtitle = element_text(color = "grey50", size = 10)
  ) +
  labs(
    title = "Volume of Stops for Cars, by Day of the Year (1-365), in NYC",
    subtitle = "Continuous time series shows consistent peak in February",
    x = "Period", 
    y = "Count of Stops"
  )
```

```{r, include=FALSE}
rm(plot_data)
```

While we observe a consistent seasonal clustering peaking in late winter and early spring, our current 3-year dataset captures three seasonal cycles. This is sufficient to identify a recurring pattern, but insufficient to establish it as a long-term trend.

The next plot, demonstrates high daily volatility (indicated by the grey jitter) that obscures the underlying seasonal trend (the blue trend line). Because the daily fluctuation is relative to the overall pattern, including 'Day of Year' as a feature risks overfitting.

```{r doy-inconsistent-density}
vehstops %>%
  filter(!is.na(doy)) %>%
  count(doy) %>%
  ggplot(aes(x = doy, y = n)) +
  geom_line(color = "grey60", alpha = 0.5) +
  geom_smooth(method = "loess", color = "steelblue", se = FALSE) +
  theme_minimal() +
   theme(
    panel.grid.major.x = element_blank(),
    plot.title = element_text(size = 14),
    plot.subtitle = element_text(color = "grey50", size = 10)
  ) +
  labs(title = "Volume of Stops for Cars, by Day of the Year (1-365), in NYC",
       subtitle = "The extreme jitter confirms this feature is too noisy to predict, but the trend favors colder months",
       x = "Day of Year", y = "Count of Stops")
```

The extreme jitter confirms this feature is too noisy to predict, but unlike the 2-wheels analysis where the trend favors hot months, the 4-wheels trend remains relatively elevated in the first half of the year before dipping during the summer months.

On the other hand, we can study the "Day of the Week" which usually abstract better.

```{r dow-dis}
vehstops %>%
  filter(!is.na(dow)) %>%
  count(dow) %>%
  mutate(
    dow = factor(dow, levels = 1:7, 
                 labels = c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun")),
    pct_change = (n - lag(n)) / lag(n) * 100,
    label_text = ifelse(is.na(pct_change), 
                        comma(n), 
                        paste0(comma(n), "\n(", ifelse(pct_change > 0, "+", ""), round(pct_change, 1), "%)"))
  ) %>%
  ggplot(aes(x = dow, y = n, group = 1)) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_point(color = "steelblue", size = 3) +
  geom_text(aes(label = label_text), vjust = -1, size = 3, lineheight = 0.8) +
  scale_y_continuous(labels = comma, 
                     limits = c(0, NA), 
                     expand = expansion(mult = c(0, 0.4))) +
  theme_minimal() +
  theme(
    panel.grid.major.x = element_blank(),
    plot.title = element_text(size = 14),
    plot.subtitle = element_text(color = "grey50", size = 10)
  ) +
  labs(
    title = "Daily Volume of Stops for Cars in NYC",
    subtitle = "Daily volume and day-over-day growth/decline show high stop counts during mid-week",
    x = "Day of the Week", 
    y = "Count of Stops"
  )
```

What can we say?

- **Key finding**: Enforcement activity follows a distinct "Mid-Week Peak" cycle that strongly mirrors the weekly rhythm seen in the 2-wheels analysis. The volume starts at its absolute lowest on Monday ($230,725$), raises aggressively ($+52\%$) into Tuesday ($350,695$), peaks on Wednesday at $380,203$ stops ($+8.4\%$), and then undergoes a multi-day decline through the end of the week, bottoming out on Sunday ($236,756$).
- **Comment**: Same as the 2-wheels analysis.
- **Proposed Action**: Same as the 2-wheels analysis.

### Hour and Time Bracket

First of all we look at the overall cycle of stops volume

```{r hour-dist}
hours_to_show <- c(0, 4, 8, 12, 16, 19, 20, 23)

full_data <- vehstops %>%
  filter(!is.na(hour)) %>%
  count(hour)

highlight_data <- full_data %>%
  filter(hour %in% hours_to_show) %>%
  arrange(hour) %>%
  mutate(
    pct_change = (n - lag(n)) / lag(n) * 100,
    label_text = ifelse(
      is.na(pct_change), 
      comma(n),
      paste0(comma(n), "\n(", ifelse(pct_change > 0, "+", ""), round(pct_change, 1), "%)"))
  )

ggplot() +
  geom_line(data = full_data, aes(x = hour, y = n), color = "grey80", linewidth = 1) +
  geom_point(data = highlight_data, aes(x = hour, y = n), color = "steelblue", size = 3) +
  geom_text(data = highlight_data, aes(x = hour, y = n, label = label_text), 
            vjust = -0.6, size = 3, lineheight = 0.8) +
  scale_x_continuous(breaks = 0:23) +
  scale_y_continuous(labels = comma, limits = c(0, NA), expand = expansion(mult = c(0, 0.4))) +
  theme_minimal() +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    plot.title = element_text(size = 14),
    plot.subtitle = element_text(color = "grey50", size = 10)
  ) +
  labs(
    title = "Hourly Volume of Stops for Cars in NYC",
    subtitle = "The stops increase during main commute hours, with a peak at 4:00 PM ",
    x = "Hour of Day (0-23)", 
    y = "Count of Stops"
  )
```

```{r, include=FALSE}
rm(hours_to_show, full_data, highlight_data)
```

What can we say?

- **Key finding**: Volume of stops is characterized by a 3-stages daily cycle. First, it follows a night time decline towards an absolute minimum at $4:00$ ($23,374$, a $-79.7\%$ drop with respect to $00:00$). Then, we have a rapid morning ascent to before localized peak at $9:00$ (at $8:00$ we have $105,378$, which is an increase of $+350.8\%$ w.r.t. 4 AM), followed by a midday decline before a rapid afternoon surge to an absolute peak at $16:00$ ($144,954$, $+80.6\%$ w.r.t. 12 PM), and a sustained high-volume evening plateau before tailing off late at night.
- **Interpretation**: This distribution indicates that the enforcement cycle is driven by three phases: the **Morning Commuter Phase** (centered around $8:00$), the **Midday Transition Phase** (which causes the temporary dip down to $80,271$ at $12:00$), and the **Late Afternoon Rush Phase** (surging aggressively to the absolute peak at $16:00$ and tapering slowly through the evening hours). Unlike the 2-wheels analysis, the 4-wheels data shows shifts that perfectly align with major automotive traffic shifts throughout the city.
- **Proposed Action**: We will maintain `hour` as a temporal feature in our predictive framework. As an alternative, we *could* try to properly capture this pattern. For instance, we could **transform the continuous hour variable using a periodic basis function**, so that it becomes a continuous feature.

Before anything else, let us see whether this cycle changed throughout the years.

```{r hour-year-dist}
hours_to_show <- c(0, 4, 8, 12, 16, 20, 23)

full_data <- vehstops %>%
  filter(!is.na(year) & !is.na(hour)) %>%
  count(year, hour)

highlight_data <- full_data %>%
  filter(hour %in% hours_to_show) %>%
  group_by(year) %>%
  arrange(hour) %>%
  mutate(
    pct_change = (n - lag(n)) / lag(n) * 100,
    label_text = ifelse(
      is.na(pct_change), 
      comma(n), 
      paste0(comma(n), "\n(", ifelse(pct_change > 0, "+", ""), round(pct_change, 1), "%)"))
  )

ggplot() +
  geom_line(data = full_data, aes(x = hour, y = n), color = "grey85", linewidth = 0.8) +
  geom_point(data = highlight_data, aes(x = hour, y = n), color = "steelblue", size = 2) +
  geom_text(data = highlight_data, aes(x = hour, y = n, label = label_text), 
            vjust = -0.7, size = 2.5, lineheight = 0.8) +
  facet_wrap(~year, nrow = 1) +
  scale_x_continuous(breaks = c(0, 4, 8, 12, 16, 20, 23)) +
  scale_y_continuous(labels = comma, limits = c(0, NA), expand = expansion(mult = c(0, 0.4))) +
  theme_minimal() +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    plot.title = element_text(size = 14),
    plot.subtitle = element_text(color = "grey50", size = 10),
    strip.text = element_text(size = 11)
  ) +
  labs(
    title = "Hourly Volume of Stops, by Year, for Cars in NYC",
    subtitle = "The same enforcement pattern emerges, throughout the 3 years period",
    x = "Hour of Day (0-23)", 
    y = "Count of Stops"
  )
```

```{r, include=FALSE}
rm(hours_to_show, full_data, highlight_data)
```

What can we say?

- **Key finding**: Across all the three years, the enforcement volume shows the same pattern as stated for the previous plot, except for the period that goes from $16:00$ to $23:00$ which seems to be shifting. Two aspects remain evident: there are a minor peak during the morning rush at $8:00$ and a dominant peak during the late afternoon and evening. In $2023$ and $2024$, the absolute highest volume of stops occurred at $16:00$ ($4:00$ PM). By $2025$, while the $16:00$ peak remains the highest absolute value ($51011$ stops), the volume remains high and remains nearly unchanged until $20:00$ with $49857$ stops.
- **Interpretation**: Unlike what we observed for two-wheeled vehicles, car stops maintain a consistent primary afternoon peak at $16:00$ across the entire $2023$ to $2025$ period. The increase in volume at $20:00$ in $2025$ suggests that evening traffic enforcement has been growing. Instead, the morning rush hour peak at $8:00$ remains stable in structure across all years.

Let us check the time brackets previously built.

```{r time-bracket-distribution}
plot_univariate_cat(
  DF = vehstops, 
  FEATURE = time_bracket, 
  X_LABEL = "Time Bracket", 
  TITLE = "Distribution of Time Brackets for Cars in NYC",
  SUBTITLE = "Enforcement is heavily skewed toward the latter half of the day (16-23)"
)
```

What can we say?

- **Key finding**: The "Afternoon/evening ($16-19$)" bracket stands as the absolute dominant peak with $546,748$ stops, closely followed by the "Night ($20-23$)" and "Morning commute ($8-11$)" brackets with $449,686$ and $410,555$ stops respectively. The lowest volume is in the "Early morning ($4-7$)" window that registers just $138,852$ stops.
- **Interpretation**: Just like for 2-wheels vehicles, the volume of enforcement is tied around the latter half of the day ($16-23$), which confirm the pattern shown by the previous bar plots.
- **Comment**: In contrast to the two-wheeled vehicle analysis, the car data shows a smoother distribution.
- **Proposed Action**: We can safely maintain our existing `time_bracket` feature engineering exactly as it is for the upcoming modeling phase.

Also, we are going to drop minute as its level of granularity cannot be properly included in the dataset.

## Driver related predictors

### Reported Age

We have mainly one issue here. First, we have many instances of drivers with unknown age,

```{r age-bracket-dist}
plot_univariate_cat(
  DF = vehstops, 
  FEATURE = age_bracket, 
  X_LABEL = "Age Bracket", 
  TITLE = "Distribution of Car Stops in NYC by Age Bracket",
  SUBTITLE = "Enforcement volume is highly concentrated in the demographic (18-44)"
)
```

What can we say?

- **Key finding**: The demographic distribution of stops is highly concentrated in the $25-34$ age category with $631,234$ observations, followed by the $35-44$ and $18-24$ brackets. Furthermore, the "Unknown" age category represents a significant portion with $118,990$ records, while the "Under 18" category is the smallest group with just $9298$ stops.
- **Interpretation**: Although, older age groups show lower frequency, they are much more well-represented here compared to the two-wheeled vehicle analysis. In fact, most categories are well represented here too.
- **Comment**: Same as 2-wheels analysis.
- **Proposed Action**: We will retain the "Unknown" category as a level within our `age_bracket` feature rather than using imputation.

And now, let us see how this looks when looking at the full distribution

```{r age-vs-age-bracket}
age_data <- vehstops %>%
  filter(!is.na(age), age >= 0, age < 100) %>%
  mutate(
    age_group = case_when(
      age < 18 ~ "<18",
      age >= 18 & age <= 24 ~ "18–24",
      age >= 25 & age <= 34 ~ "25–34",
      age >= 35 & age <= 44 ~ "35–44",
      age >= 45 & age <= 54 ~ "45–54",
      age >= 55 & age <= 64 ~ "55–64",
      age >= 65 ~ "65+"
    ),
    age_group = factor(age_group, levels = c("<18", "18–24", "25–34", "35–44", "45–54", "55–64", "65+"))
  )

ggplot(age_data, aes(x = age, fill = age_group)) +
  geom_histogram(binwidth = 1, color = "white", linewidth = 0.1) +
  scale_fill_viridis_d() +
  theme_minimal() +
   theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    plot.title = element_text(size = 14),
    plot.subtitle = element_text(color = "grey50", size = 10),
    strip.text = element_text(size = 11)
  ) +
  labs(
    title = "Age Distribution of Stopped Car Drivers in NYC by Target Bracket",
    subtitle = "Enforcement volume is confirmed to be concentrated in 20-45 age range",
    x = "Driver Age", 
    y = "Count of Stops",
    fill = "Age Group"
  ) +
  theme(legend.position = "bottom") +
  guides(fill = guide_legend(nrow = 2))
```

```{r, include=FALSE}
rm(age_data)
```

What can we say?

- **Key finding**: The enforcement volume shows a rapid increase, from the underage year, that becomes a sustained plateau that spans from around the age $23$ through age $35$. Here, most age labels can count on more than $60000$ stops. After the age of $35$, the volume starts to steadily decline through middle age. After that, it shifts into a long tail that, past age $75$, becomes extremely flat.
- **Comment**: The four-wheel stops distribution show a **much broader plateau**, with respect to the 2-wheels analysis, which is probably due to the higher concentration of older drivers (hinted by the slow decreasing pattern).
- **Proposed Action**: Same as 2-wheels analysis, even if here the best thing to do is to filter over $85$, but we want it to be comparable with the other analysis.

Let us remove the outliers

```{r remove-age-outliers}
# Keep the NAs
vehstops <- vehstops %>%
  filter(age < 76 | is.na(age))
```

From now on we will consider drivers of age in the range 16-75.

```{r age-outlier-boxplot-after}
vehstops %>%
  filter(!is.na(age)) %>%
  ggplot(aes(y = age)) +
  geom_boxplot(fill = "steelblue", alpha = 0.6, width = 0.4,
               outlier.size = 1.5, outlier.alpha = 0.4) +
  coord_flip() +
  theme_minimal() +
  theme(
    panel.grid.major.y = element_blank(),
    plot.title = element_text(size = 14),
    plot.subtitle = element_text(color = "grey50", size = 10)
  ) +
  labs(
    title = "Age Distribution after Outlier Removal (76+ excluded)",
    subtitle = "Remaining values span the plausible 16-75 enforcement range",
    y = "Reported Age", x = NULL
  )
```

What can we say?

- **Key finding**: The median is centered around $35$, with the interquartile range ($IQR$) stretching roughly from $27$ to $47$. The lower whisker reaches near the minimum legal driving age, and the upper whisker gets up to $75$ without showing any point as outlier.
- **Interpretation**: The fact that the $IQR$ spans $27-47$ reflects the increased presence of older drivers in the car dataset.
- **Comment**: While the boxplot for the two-wheeled vehicle analysis shows sparsity for older drivers ($65-75$), here the car data does not label them as possible outliers for the current distribution.
- **Proposed Action**: Same as 2-wheels analysis.

### Reported Sex

```{r sex-distribution}
plot_univariate_cat(
  DF = vehstops, 
  FEATURE = sex, 
  X_LABEL = "Reported Sex", 
  TITLE = "Distribution of Car Stops in NYC by Reported Sex",
  SUBTITLE = "The vast majority of stopped drivers are males",
  order_by = "freq"
)
```

What can we say?

- **Key finding**: The "Male" (M) drivers represent th majority of stops with $1,736,848$ observations. Unlike the 2-wheels case, "Female" (F) drivers are not under-represented and they account for $395,347$ stops, while the "Unknown" (U) category remains a small minority with $13,392$ records.
- **Interpretation**: Same as 2-wheels analysis.
- **Comment**: In comparison to the two-wheeled vehicle analysis, the female proportion is significantly larger here.
- **Proposed Action**: Same as 2-wheels analysis.

### Reported Ethnicity

```{r ethnicity-distribution}
plot_univariate_cat(
  DF = vehstops, 
  FEATURE = ethnicity, 
  X_LABEL = "Reported Ethnicity", 
  TITLE = "Distribution of Car Stops in NYC by Reported Ethnicity",
  SUBTITLE = "Hispanic drivers seem to account for the plurality of stops",
  order_by = "freq"
)
```

What can we say?

- **Key finding**: "Black" drivers represent the largest single group with $656,109$ stops, followed by "Hispanic" drivers with $565,531$ observations. Together, they represent the majority of the data. Furthermore, the amount of stops for "White" car drivers is $426,594$, which is incredibly high with respect to the 4th category (by amount of stops), "Asian / Pacific Islander" ($270,311$). In fact, the 2-wheels analysis found that these two labels had almost the same amount of observations. For what concerns the remaining categories, we have "Unknown" ($193,050$), and "Other" ($33,992$) for last.
- **Interpretation**: This profile indicates that four-wheeled vehicle enforcement is heavily concentrated within Black and Hispanic demographics across the city. The high absolute count for the "Unknown" category matches the pattern we observed with missing demographic metrics elsewhere in our analysis. Given the large sample size across all 4 main groups (White, Black, Hispanic, and Asian) we can be sure to have plenty of variability to learn from, unlike the other dataset.
- **Comment**: In comparison to the previous (2-wheels) analysis, where the distribution was dominated by Hispanic riders, the four-wheeled data shows a more distributed but still unequal breakdown. Here, Black drivers constitute the absolute peak subset of stops, even if the majority of the NYC population is White.
- **Proposed Action**: Same as 2-wheels analysis.

------------------------------------------------------------------------

# Bivariate EDA

## Association of Binary Response vs. Categorical Predictors

When dealing with a dataset of this size, traditional p-values from chi-squared tests become meaningless. To get around this, we use Cramer’s V to measure effect size (association).

Because of how the `table()` function works in R, we can also include integer predictors like `hour` and `age`: they do not take truly continuous values in this dataset, so treating each integer as its own category is valid.

```{r categorical-correlation-heatmap}
# Excluding doy (redundant) and dow (we use weekday instead)
columns <- c(
  "year", "month", "weekday", 
  "hour", "time_bracket",
  "area", "boro", 
  "sex", "ethnicity", "age", "age_bracket",
  "arrested"
  )

plot_categorical_correlation(
  DF = vehstops, 
  FEATURES = columns,
  TITLE = "Correlation between Categorical Variables (Cramer's V)"
)
```

What can we say?

- **Key finding**: The association matrix reveals that the only predictors showing a noticeable signal with `arrested` are, `ethnicity` ($0.1$), `hour`/`time_bracket` (both at $0.09$). Here, other demographic features such as `age`/`age_bracket` ($0.07$) and `sex` ($0.05$) have a more modest association. Spatial features like `boro` ($0.06$)/`area` ($0.05$) show very low, but still relevant signals. The matrix highlights a moderate association between these top predictors, specifically:
  - between `age`/`age_bracket` $\leftrightarrow$ `hour` / `time_bracket` ($\approx 0.08-0.09$)
  - between `sex` $\leftrightarrow$ `hour` / `time_bracket` ($\approx 0.07-0.08$)
  - between `ethnicity` $\leftrightarrow$ `hour` / `time_bracket` ($\approx 0.07-0.08$)
  - between `ethnicity` $\leftrightarrow$ `area`/`boro` ($0.17 - 0.19$)
  - between `ethnicity` $\leftrightarrow$ `sex` ($0.15$).
- **Interpretation**: Temporal and ethnic features carry the strongest individual baseline impacts when predicting an arrest, while standalone spatial attributes are less informative on their own for four-wheeled vehicles. The persistent $0.17$ to $0.19$ association between ethnicity and geography suggests that the demographic composition of stops remains non-uniformly distributed across different parts of the city.
- **Comment**: While the previous analysis shows a stronger connection to geographical signals (where `area` and `boro` scored $0.13$), here we probably encounter a more distributed arrest pattern. However, because no single predictor strongly correlates with `arrested`, models relying purely on main effects will likely underfit the data. Therefore, we must be careful when we consider the interactions.
- **Proposed Action**: Same as 2-wheels analysis, with the sole difference that here we focus a bit more on demographics then spatial features.

## Correlation of Binary Response vs. Numerical Predictors

```{r numerical-correlation-heatmap}
# Ignoring Latitude and Longitude
numerical_cols <- c(
  "arrested",
  "year", "month", "day", "dow", "doy", "hour", 
  "age"
  )

cor_matrix <- cor(vehstops[, numerical_cols], use = "complete.obs")

corrplot(cor_matrix, method = "color", type = "upper",
         order = "hclust", addCoef.col = "black", number.cex = 0.7,
         tl.col = "black", tl.srt = 45,
         title = "Correlation Matrix of Numerical Features",
         mar = c(0, 0, 2, 0))
```

What can we say?

- **Key finding**: The Pearson correlation matrix shows that our binary target `arrested` has a negligible linear relationship with all continuous numerical features. The strongest standalone linear relationship with the target appears in `age` ($-0.07$), which indicates a very slight negative trend, followed by `hour` ($0.03$). Calendar features like `month`, `doy` (day of year), `year`, `dow` (day of week), and `day` all show absolute correlation coefficients of $0.01$ or lower, indicating a complete absence of direct global linear relationships.
- **Interpretation**: The near-zero linear correlations across the board demonstrate that the probability of an arrest cannot be effectively modeled or predicted using standard linear relationships on raw numerical predictors. The slight negative value for `age` implies that older stopped drivers have a marginally lower likelihood of being arrested.
- **Comment**: Linear correlation metrics are not suited for capturing the dynamics of this dataset. This is probably due to the fact, both the features (like cyclical hours or seasonal months) and the target variable (`arrested`) act non-linearly.
- **Proposed Action**: We could explicitly map the continuous predictors using non-linear modeling frameworks, such as incorporating polynomial expansions or using Generalized Additive Models (GAMs) with smoothing splines for `age` and `hour`.

## Geospatial Features

### Area

```{r area-vs-arrested-ratio}
ratio_by_cat(
  vehstops, arrested, area,
  "Area", "Arrest Rate (%)", "Arrest Ratio by Area for Cars Stopped in NYC",
  style = "stack", legend_title = "Stop ended in arrest?",
  target_labels = c("No", "Yes"),
  subtitle = "Although the ratios are low, the amount of arrests, in proportion, are incredibly high",
  label_vjust = 0.5,
  bar_width = 0.6,  
  text_size = 2.5,
  aspect_ratio = 0.9
)
```

Let us look at the numbers too

```{r confirm-area-vs-arrested-ratio}
vehstops %>%
  group_by(area, arrested) %>%
  summarise(count = n(), .groups = "drop_last") %>%
  mutate(percentage = count / sum(count))
```

What can we say?

- **Key finding**: The plot reveals a comparatively narrower yet statistically significant disparity in the proportions of four-wheeled vehicle stops that culminate in an arrest across New York City areas. The arrest rate ranges from a minimum of $2.5\%$ in Queens to a maximum of $4.8\%$ in the Bronx. This represents nearly a 2x increase in the conditional probability of an arrest depending purely on where the stop occurs. While the absolute majority of stops do not result in arrests across all territories, the Bronx consistently maintains the highest conversion rate, followed by Brooklyn at $3.5\%$, Staten Island at $3.0\%$, and Manhattan at $2.8\%$.
- **Interpretation**: The variance in four-wheeled vehicle arrest rates across geography is structurally less volatile than what was observed in the two-wheel dataset. Because four-wheeled vehicles represent a massive overall sample size across all NYC boroughs, even small percentage fluctuations like the difference between $2.5\%$ in Queens and $4.8\%$ in the Bronx represent highly robust statistical signals with minimal exposure to sampling noise or wide confidence intervals. This confirms that geographic territory exerts a stable, persistent main effect on the target variable.
- **Comment**: When contrasted against the two-wheeled vehicle analysis, where arrest rates skyrocketed up to $12.1\%$ in the Bronx, four-wheeled vehicle stops demonstrate a much tighter distribution window. The multi-fold spikes seen in two-wheelers were highly localized anomalies, likely tied to aggressive enforcement of specific delivery or unregistered vehicle regulations. For standard cars, the spatial distribution is smoother, yet it confirms that police behavior and the baseline threshold for making an arrest continue to vary systematically by borough. The Bronx remains the clear operational peak for arrest probability across both analyses.
- **Proposed Action**: We will include `boro` as a core categorical predictor in our modeling pipeline to capture these geographic baseline shifts. Given that the absolute volumes of four-wheeled stops are massive across all five boroughs, we will not need to group or collapse any regional levels. However, we will ensure that our train-test splits are fully stratified across the geographic blocks to prevent any spatial imbalance from biasing our final parameter estimates for `arrested`.

### Borough

```{r boro-vs-arrested-ratio}
ratio_by_cat(
  vehstops, arrested, boro,
  "Borough", "Arrest Rate (%)", "Arrest Ratio by Borough for Cars Stopped in NYC",
  style = "stack", legend_title = "Stop ended in arrest?",
  target_labels = c("No", "Yes"),
  subtitle = "The arrest rate seems to be well distributed among the boroughs",
  label_vjust = 0.5,
  bar_width = 0.8,
  text_size = 2.5
)
```

What can we say?

- **Key finding**: Disaggregating the geographic data into finer operational regions preserves the macro position of the Bronx ($4.8\%$) as the absolute peak but exposes massive intra-borough variance that was previously hidden. While Manhattan as a whole appeared moderately low in our previous chart, this breakdown reveals a steep internal gradient ranging from a citywide minimum of $1.6\%$ in Manhattan (South) to a substantial $3.9\%$ in Manhattan (North). The remaining separated regions display a steady upward slope, with Queens ranging from $2.1\%$ (North) to $2.8\%$ (South), and Brooklyn climbing from $3.3\%$ (South) to $3.7\%$ (North).
- **Interpretation**: This fine-grained visualization confirms that the macro patterns are heavily influenced by specific administrative zones. The stark divergence within Manhattan is particularly illuminating, as a stop in Manhattan (North) is more than twice as likely to result in an arrest compared to one in Manhattan (South). This strong localized variance suggests that utilizing a finer geographic division will significantly improve the explanatory power of our models, as macro borough-level metrics average out these critical operational differences.
- **Comment**: When comparing this to the two-wheeled vehicle analysis, the four-wheel data tells a much more localized story. In the two-wheel case, the entire borough of the Bronx was an isolated outlier due to targeted citywide regulations. Here, standard car stops show that police behavior and arrest conversion thresholds are highly dependent on specific command zones within the same borough. The fact that Manhattan (North) rivals the Bronx in arrest probability, while Manhattan (South) remains the absolute lowest, highlights a sharp geographical divide in enforcement outcomes that a coarse analysis would completely obscure.
- **Proposed Action**: We will perform a likelihood ratio test on nested logistic regression models to formally evaluate whether this high-resolution `Borough` feature achieves a statistically significant improvement in deviance over the simpler macro-area classifier. If the test yields a significant result, we will incorporate the finer regional categories into our final model to capture this localized variance. Conversely, if the additional complexity does not provide sufficient statistical justification, we will favor the simpler, more parsimonious macro feature to prevent overfitting.

We will need to choose between one or the other.

### Latitude and Longitude for Stop Rate

```{r stop-ratio-density}
nyc_map <- counties(state = "NY", cb = TRUE) %>%
  filter(NAME %in% c("Bronx", "Kings", "New York", "Queens", "Richmond"))

bins <- 100
lon_breaks <- seq(-74.26, -73.71, length.out = bins + 1)
lat_breaks <- seq(40.48, 40.92, length.out = bins + 1)

grid_sum_rate <- vehstops %>%
  filter(!is.na(longitude), !is.na(latitude), !is.na(arrested)) %>%
  mutate(
    xbin = cut(longitude, breaks = lon_breaks, include.lowest = TRUE, right = FALSE),
    ybin = cut(latitude, breaks = lat_breaks, include.lowest = TRUE, right = FALSE)
  ) %>%
  mutate(
    x = as.numeric(gsub(".*\\[(.*),.*", "\\1", xbin)) + (lon_breaks[2] - lon_breaks[1])/2,
    y = as.numeric(gsub(".*\\[(.*),.*", "\\1", ybin)) + (lat_breaks[2] - lat_breaks[1])/2
  ) %>%
  group_by(x, y) %>%
  summarise(
    total_stops = n(),
    arrest_rate = mean(arrested == 1),
    .groups = "drop"
  ) %>%
  filter(total_stops >= 5)

ggplot() +
  geom_sf(data = nyc_map, fill = "white", color = "grey80", linewidth = 0.2) +
  geom_tile(data = grid_sum_rate, aes(x, y, fill = arrest_rate), alpha = 0.8) +
  scale_fill_viridis_c(
    labels = scales::percent, 
    name = "Arrest Rate (%)", 
    option = "inferno",
    direction = -1
  ) +
  coord_sf(xlim = c(-74.26, -73.71), ylim = c(40.48, 40.92)) +
  theme_minimal() +
  theme(
    panel.grid = element_blank(),
    plot.title = element_text(size = 14),
    plot.subtitle = element_text(color = "grey50", size = 10)
  ) +
  labs(
    title = "Spatial Distribution of Arrest Rates for Cars Stopped in NYC",
    subtitle = "Percentage of stops resulting in arrest (Bins with n < 5 removed)",
    x = "Longitude", 
    y = "Latitude"
  )
```

What can we say?

- **Key finding**: The spatial visualization exhibits a highly decentralized distribution of arrest rates for standard vehicles, characterized by a predominantly uniform baseline across the city with light yellow to orange shades ($0\%$ to $15\%$). However, isolated high-intensity hot spots are visible as dark purple and black pixels scattered across northern Manhattan, parts of the Bronx, eastern Queens, and central Staten Island, where specific geographic bins show arrest rates spiking above $40\%$.
- **Interpretation**: While this highlights a localized geographic heterogeneity, it exposes a critical limitation regarding direct model integration because spatial coordinates violate the standard assumption of independent and identically distributed ($i.i.d.$) errors. The presence of isolated high-arrest bins amidst broad fields of low arrest rates suggests these spikes could be driven by micro-level spatial anomalies, localized checkpoints, or sparse sample sizes that survived the $n < 5$ filter but remain highly sensitive to single-event fluctuations.
- **Comment**: In comparison to the two-wheeled vehicle analysis, which showcased broad and contiguous regional hot spots reflecting targeted borough-wide crackdowns, four-wheeled vehicle enforcement presents a much more fragmented spatial narrative. Standard vehicle stops do not yield massive, sweeping clusters of high arrest rates across entire neighborhoods. Instead, the baseline risk is fairly homogenous across the five boroughs, meaning that the extreme $40\%$ plus spikes are highly localized exceptions rather than representative regional trends.
- **Proposed Action**: We should avoid feeding raw, high-resolution continuous coordinates directly into a standard generalized linear model due to the risk of spatial autocorrelation distorting our standard errors. Instead, we can extract value from this micro-spatial structure by engineering a regional density feature or by using spatial aggregation methods like K-means clustering to group adjacent bins into cohesive categorical inputs. Alternatively, we can utilize mixed-effects modeling frameworks where geographic zones are treated as random intercepts to isolate the true local variance without breaking our statistical assumptions.

## Time related features

### Year

```{r year-vs-arrested-ratio}
arrest_trends <- vehstops %>%
  filter(!is.na(year), !is.na(arrested)) %>%
  group_by(year) %>%
  summarise(
    total_stops = n(),
    arrests = sum(arrested == 1),
    rate = arrests / total_stops, 
    .groups = 'drop'
  ) %>%
  arrange(year) %>%
  mutate(
    prev_rate = lag(rate),
    delta = (rate - prev_rate) * 100,
    label_text = ifelse(
      is.na(prev_rate), 
      scales::percent(rate, accuracy = 0.1), 
      paste0(scales::percent(rate, accuracy = 0.1), 
             "\n(", ifelse(delta > 0, "+", ""), round(delta, 1), " pts)")
    )
  )

ggplot(arrest_trends, aes(x = factor(year), y = rate, group = 1)) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_point(color = "steelblue", size = 3) +
  geom_text(aes(label = label_text), vjust = -0.8, size = 3.5, lineheight = 0.8) +
  scale_y_continuous(
    labels = scales::percent, 
    limits = c(0, max(arrest_trends$rate) * 1.3),
    expand = expansion(mult = c(0, 0.1))
  ) +
  theme_minimal() +
  theme(
    panel.grid.major.x = element_blank(),
    plot.title = element_text(size = 14),
    plot.subtitle = element_text(size = 10, color = "grey50")
  ) +
  labs(
    title = "Annual Arrest Rate has Stabilized in the last 3 years for Cars Stopped in NYC",
    subtitle = "The percentage of arrests seems to have reached a stable point for the first time",
    x = "Year", y = "Arrest Rate (%)"
  )
```

What can we say?

- **Key finding**: The line chart displays the trajectory of the annual arrest rate for four-wheeled vehicles over a three-year period, revealing an initial rise followed by stabilization. The proportion of stops resulting in an arrest rose from $2.8\%$ in $2023$ to $3.5\%$ in $2024$ (a jump of $+0.7$ percentage points), followed by a marginal contraction to $3.4\%$ in $2025$ (a slight decrease of $-0.1$ percentage points).
- **Interpretation**: From a time-related perspective, this demonstrates that the probability of an arrest given a stop shifted between $2023$ and $2024$ but has since plateaued. In a predictive framework, this behavior implies that while data from $2023$ might underestimate subsequent enforcement patterns, data points from $2024$ and $2025$ are structurally homogenous. The relative stabilization indicates that the macro baseline threshold for executing an arrest has reached an equilibrium point in the final two years of the window.
- **Comment**: In comparison to the two-wheeled vehicle analysis, which exhibited a continuous upward drift up to $6.0\%$, four-wheeled vehicle stops show a significantly lower overall arrest conversion rate and a distinct plateauing pattern. The continuous surge in two-wheel arrests was highly representative of evolving micromobility crackdowns and successive regulatory iterations stretching into $2025$. Conversely, standard vehicle enforcement represents a more mature operational environment, where policy shifts yielded a one-time adjustment in $2024$ before returning to a stable baseline.
- **Proposed Action**: We should check for an association effect between time and geography by plotting these three-year trends faceted by borough to see if the $2024$ peak and subsequent $2025$ stabilization are uniform across the city. Furthermore, we will include `year` as a categorical predictor in our experimental model pipeline to absorb the initial yearly macro trend shift, though we may consider pooling $2024$ and $2025$ if a parsimonious model formulation is desired.

### Month

```{r month-vs-arrested-ratio}
monthly_arrest_data <- vehstops %>%
  filter(!is.na(month), !is.na(arrested)) %>%
  group_by(month) %>%
  summarise(
    total_stops = n(),
    arrests = sum(arrested == 1),
    rate = arrests / total_stops,
    .groups = "drop"
  ) %>%
  mutate(month = factor(month, levels = 1:12)) %>%
  arrange(month) %>%
  mutate(
    prev_rate = lag(rate),
    delta = (rate - prev_rate) * 100,
    label_text = ifelse(
      is.na(prev_rate), 
      scales::percent(rate, accuracy = 0.1), 
      paste0(scales::percent(rate, accuracy = 0.1), 
             "\n(", ifelse(delta > 0, "+", ""), round(delta, 1), " pts)")
    )
  )

monthly_arrest_data %>%
  ggplot(aes(x = month, y = rate, group = 1)) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_point(color = "steelblue", size = 3) +
  geom_text(aes(label = label_text), vjust = -1.2, size = 3, lineheight = 0.8) +
  scale_x_discrete(labels = month.abb) +
  scale_y_continuous(
    labels = scales::percent, 
    limits = c(0, max(monthly_arrest_data$rate) * 1.3), 
    expand = expansion(mult = c(0, 0.1))
  ) +
  theme_minimal() +
  theme(
    panel.grid.major.x = element_blank(),
    plot.title = element_text(size = 14),
    plot.subtitle = element_text(color = "grey50", size = 10)
  ) +
  labs(
    title = "Monthly Arrest Rate Trend for Cars Stopped in NYC",
    subtitle = "Aggregated monthly arrest rate, the peak is found between January and December",
    x = "Month",
    y = "Arrest Rate (%)"
  )
```

What can we say?

- **Key finding**: The monthly arrest rate line chart describes a highly stable, flat intra-year pattern with slight structural peaks concentrated at the absolute ends of the calendar year. The arrest rate begins at $3.6\%$ in January, drops and plateaus between $3.1\%$ and $3.3\%$ from February through November, and then experiences a slight late-year surge to its annual peak of $3.7\%$ in December ($+0.4$ percentage points).
- **Interpretation**: From a temporal perspective, the conditional probability of an arrest given a stop is essentially invariant across the spring, summer, and autumn months for four-wheeled vehicles. The slight elevation observed in December and January suggests that seasonal factors, such as holiday-season traffic initiatives, increased checkpoint density, or changes in driving behavior during mid-winter, introduce a minor but distinct uptick in arrest efficiency at the turn of the year.
- **Comment**: This visualization presents a stark contrast to the two-wheeled vehicle analysis, where arrest rates exhibited a massive sinusoidal curve tied directly to the summer season, peaking at $6.5\%$ in July. For standard automobiles, there is no summer surge, which further underscores that standard vehicle operations maintain a highly predictable, institutional baseline year-round. The minor spikes in December and January are likely driven by citywide holiday enforcement campaigns rather than environmental weather changes, revealing a completely different operational timeline than the two-wheel market.
- **Proposed Action**: We should avoid relying on the binary `hot_season` feature for the four-wheeled vehicle framework, as the flat summer profile means this indicator will fail to capture any meaningful predictive signal. Instead, we can extract the true seasonal variance by engineering a specialized categorical feature that isolates the holiday transition window, such as a binary indicator for `holiday_season` spanning December and January, or by utilizing `month` as a nominal variable with January or December established as the reference levels in our regression pipeline.

#### Faceted

```{r month-faceted-vs-arrested-ratio}
plot_data <- vehstops %>%
  filter(!is.na(year), !is.na(month), !is.na(arrested)) %>%
  group_by(year, month) %>%
  summarise(
    arrest_rate = mean(arrested == 1), 
    .groups = "drop"
  ) %>%
  mutate(month = factor(month, levels = 1:12)) %>%
  group_by(year) %>%
  mutate(
    rate_jan = arrest_rate[month == 1],
    rate_max = max(arrest_rate),
    is_max = arrest_rate == rate_max,
  
    diff_jan = (arrest_rate - rate_jan) * 100,
    diff_dec = (arrest_rate - rate_max) * 100, 
    
    label_text = case_when(
      month == 1 ~ scales::percent(arrest_rate, accuracy = 0.1), 
      is_max ~ paste0(scales::percent(arrest_rate, accuracy = 0.1), "\n(", ifelse(diff_jan > 0, "+", ""), round(diff_jan, 1), " pts)"),
      month == 12 ~ paste0(scales::percent(arrest_rate, accuracy = 0.1), "\n(", round(diff_dec, 1), " pts)"),
      TRUE ~ ""
    )
  )


plot_data %>%
  ggplot(aes(x = month, y = arrest_rate, group = 1)) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_point(color = "steelblue", size = 2) +
  geom_text(aes(label = label_text), vjust = -1, hjust = 0.2, size = 2.5, lineheight = 0.8) +
  facet_wrap(~year, nrow = 1) +
  scale_x_discrete(
    labels = c("J", "F", "M", "A", "M", "J", "J", "A", "S", "O", "N", "D"), 
    expand = expansion(mult = c(0.1, 0.1))
  ) +
  scale_y_continuous(labels = scales::percent, limits = c(0, NA), expand = expansion(mult = c(0, 0.4))) +
  coord_cartesian(clip = "off") +
  theme_minimal() +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    plot.title = element_text(size = 14),
    plot.subtitle = element_text(color = "grey50", size = 10),
    strip.text = element_text(size = 11)
  ) +
  labs(
    title = "Monthly Arrest Rate, by Year, for Cars in NYC",
    subtitle = "The arrest rate pattern is still drifting",
    x = "Month",
    y = "Arrest Rate (%)"
  )
```

What can we say?

- **Key finding**: The faceted line chart breaks down the monthly arrest rates for four-wheeled vehicles across three consecutive years, revealing that the internal shape of the seasonal trajectory is fundamentally changing shape over time alongside its vertical shifts. While $2023$ shows a steady upward climb from a minimum of $2.4\%$ in February to a peak of $3.7\%$ in December ($+1.1$ percentage points), $2024$ and $2025$ invert this structure. In $2024$, the year peaks early at $3.9\%$ in January, drops through summer, and builds back to $3.9\%$ in December. By $2025$, the absolute peak occurs in January at $4.1\%$ before experiencing a gradual downward trend that bottoms out in autumn and rises marginally to finish at $3.5\%$ in December.
- **Interpretation**: From a longitudinal perspective, this evolution demonstrates that the seasonal distribution of the arrest probability is non-stationary. The sharp contrast between the upward trend of $2023$ and the front-loaded, downward-sloping profiles of $2024$ and $2025$ indicates that a fixed, static monthly effect will fail to capture the baseline transition. The data suggests that tactical changes in traffic enforcement strategies or policy mandates implemented between late $2023$ and early $2024$ permanently re-engineered the annual distribution, shifting the highest conditional probability of an arrest to the beginning of the calendar year.
- **Comment**: When contrasted against the two-wheeled vehicle analysis, which exhibited a highly stable, recurring summer peak across all panels, the four-wheeled data exposes a complete structural transformation. Standard vehicle enforcement is not bound by a predictable meteorological cycle like delivery riders or micromobility users. Instead, the rapid migration of the peak from December ($2023$) to January ($2024$ and $2025$) underscores that administrative, policy-driven behavior dominates this timeline. Rather than a recurring mid-summer dip, standard cars see a sustained, institutional baseline shift where early-winter enforcement has become progressively more efficient at securing arrests over the three-year window.
- **Proposed Action**: We cannot rely on a simple main effect for month or a static seasonal category in our final predictive architecture, as the shifting intra-year structure violates the assumption of temporal stationarity. To accurately capture how the probability of `arrested` scales across this timeline, we must explicitly include a macro interaction term structured as `month * year` within our model. This formulation will allow the classification framework to adjust the monthly weights dynamically across the three-year horizon, preventing the inverted trends of $2024$ and $2025$ from canceling out the upward trajectory observed in $2023$.

### Day, Dow and Weekday

```{r day-vs-arrested-ratio}
daily_arrest_rate <- vehstops %>%
  filter(!is.na(day), !is.na(arrested)) %>%
  group_by(day) %>%
  summarise(
    total_stops = n(),
    arrests = sum(arrested == 1),
    rate = arrests / total_stops,
    .groups = "drop"
  )

daily_arrest_rate %>%
  ggplot(aes(x = day, y = rate)) +
  geom_line(color = "grey60", alpha = 0.5) +
  geom_smooth(method = "loess", color = "steelblue", se = FALSE) +
  scale_y_continuous(labels = scales::percent, limits = c(0, NA)) +
  theme_minimal() + 
  theme(
    panel.grid.major.x = element_blank(),
    plot.title = element_text(size = 14),
    plot.subtitle = element_text(color = "grey50", size = 10)
  ) +
  labs(
    title = "Arrest Rate by Day of the Month (1-31) for Cars Stopped in NYC",
    subtitle = "The trend seems to be stable across every day",
    x = "Day of the Month",
    y = "Arrest Rate (%)"
  )
```

What can we say?

- **Key finding**: Arrest rates are broadly flat across days 1 through 31, showing no meaningful trend.
- **Interpretation**: The flat pattern across days 1-28 indicates that the day-of-month variable carries no genuine enforcement signal.
- **Proposed Action**: Exclude `day` from our models.

```{r dow-vs-arrested-ratio}
daily_arrest_rate <- vehstops %>%
  filter(!is.na(dow), !is.na(arrested)) %>%
  group_by(dow) %>%
  summarise(
    total_stops = n(),
    arrests = sum(arrested == 1),
    rate = arrests / total_stops,
    .groups = "drop"
  ) %>%
  mutate(
    dow_label = factor(dow, levels = 1:7,
                       labels = c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun")),
    prev_rate = lag(rate),
    delta = (rate - prev_rate) * 100,
    label_text = ifelse(
      is.na(prev_rate),
      scales::percent(rate, accuracy = 0.1),
      paste0(scales::percent(rate, accuracy = 0.1),
             "\n(", ifelse(delta > 0, "+", ""), round(delta, 1), " pts)")
    )
  )

daily_arrest_rate %>%
  ggplot(aes(x = dow_label, y = rate, group = 1)) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_point(color = "steelblue", size = 3) +
  geom_text(aes(label = label_text), vjust = -1.0, size = 3, lineheight = 0.8) +
  scale_y_continuous(
    labels = scales::percent,
    limits = c(0, max(daily_arrest_rate$rate) * 1.3),
    expand = expansion(mult = c(0, 0.1))
  ) +
  theme_minimal() +
  theme(
    panel.grid.major.x = element_blank(),
    plot.title = element_text(size = 14),
    plot.subtitle = element_text(color = "grey50", size = 10)
  ) +
  labs(
    title = "Arrest Rate by Day of the Week for Cars Stopped in NYC",
    subtitle = "Arrest rate peaks on weekends",
    x = "Day of the Week",
    y = "Arrest Rate (%)"
  )
```

What can we say?

- **Key finding**: The line chart displays a gradual, monotonic rise in the weekly arrest rate as we approach the weekend. The probability of an arrest given a stop begins at its minimum on Monday at $3.0\%$, hovers around $3.1\%$ to $3.2\%$ through Thursday, and then experiences a clear upward lift starting on Friday ($3.5\%$) before reaching its absolute maximum on Saturday at $3.8\%$ ($+0.3$ percentage points).
- **Interpretation**: The overall variation across the week is exceptionally tight, spanning a narrow range of only $0.8$ percentage points between the Monday low and the Saturday peak. From a predictive perspective, this tells us that the day of the week carries a relatively weak standalone signal compared to spatial or hourly features. The stable baseline during the working week suggests that routine commuting days present a highly uniform threshold for arrest conversions.
- **Comment**: When comparing this trajectory to the two-wheeled vehicle analysis, we notice a shared behavioral shift toward the weekend. Both vehicle types experience an absolute maximum arrest rate on Saturday, which suggests a universal operational pattern where nighttime driving, leisure travel, and weekend enforcement strategies alter the composition of stops. Interestingly, while two-wheelers saw a fluctuating mid-week peak on Tuesday, standard car stops follow a smoother, more predictable ramp-up, highlighting the steady transition from structured weekday traffic to weekend police behaviors.
- **Proposed Action**: We should treat the day of the week as a categorical factor variable using dummy encoding within our predictive pipeline. Given the narrow global variation, we will evaluate whether collapsing the feature into a simplified binary indicator, such as `is_weekend` (isolating Friday through Sunday), preserves the predictive signal while creating a more parsimonious model structure.

### Hour and Time Bracket

```{r hour-vs-arrested-ratio}
hourly_arrest_rate <- vehstops %>%
  filter(!is.na(hour), !is.na(arrested)) %>%
  group_by(hour) %>%
  summarise(
    total_stops = n(),
    arrests = sum(arrested == 1),
    rate = arrests / total_stops,
    .groups = "drop"
  )

hourly_arrest_rate %>%
  ggplot(aes(x = hour, y = rate)) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_point(color = "steelblue", size = 2) +
  scale_x_continuous(breaks = 0:23) +
  scale_y_continuous(labels = scales::percent, limits = c(0, NA),
                     expand = expansion(mult = c(0, 0.15))) +
  theme_minimal() +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    plot.title = element_text(size = 14),
    plot.subtitle = element_text(color = "grey50", size = 10)
  ) +
  labs(
    title = "Arrest Rate by Hour of Day for Cars Stopped in NYC",
    subtitle = "The hourly arrest rate follows a U-shaped diurnal pattern",
    x = "Hour of Day (0-23)",
    y = "Arrest Rate (%)"
  )
```

What can we say?

- **Key finding**: The line plot underlines a distinct time-dependent pattern in the arrest rate resulting from stops across a $24$-hour cycle. We observe a global minimum at $7:00$ ($07:00$ AM) where the arrest rate drops to around $0.6\%$, followed by a steady climb to a minor afternoon peak around $14:00$ ($02:00$ PM) near $2.9\%$. The absolute maximum occurs during the late-night and early-morning hours, specifically peaking at $3:00$ ($03:00$ AM) with an arrest rate reaching approximately $5.3\%$.
- **Interpretation**: This strong diurnal fluctuation suggests that our exploratory data analysis is heavily influenced by variations in driver behavior, law enforcement patterns, or both depending on the hour of the day. The steep descent between $3:00$ and $7:00$ and the rapid ascent between $16:00$ and $19:00$ reveal that the conditional probability of an arrest given a stop is highly dynamic, making the exact time of the encounter a highly informative feature.
- **Comment**: When comparing this curve to the two-wheeled vehicle analysis, we observe a remarkably similar U-shaped temporal structure, although the four-wheel peak is compressed at a lower absolute maximum ($5.3\%$ versus nearly $11\%$). The low arrest rates during the morning commute hours (from $6:00$ to $10:00$) suggest that law enforcement actions during this window might focus on high-volume traffic management and routing. Conversely, the high plateau maintained from $19:00$ to $3:00$ likely reflects stops related to more severe nighttime offenses, such as operating a vehicle under the influence or criminal contraband stops.
- **Proposed Action**: We may opt for the engineered `time_bracket` predictor as our primary spatial-temporal input to capture this prominent U-shape in a parsimonious way. However, given the sharp changes visible during the morning drop and evening rise, we will test whether a continuous spline function using `hour` inside a Generalized Additive Model framework fits the non-linear transition thresholds significantly better than discrete categorical buckets.

Let us look at our grouping now

```{r time-bracket-vs-arrested-ratio}
ratio_by_cat(
  vehstops, arrested, time_bracket,
  "Time bracket", "Arrest Rate (%)", "Arrest ratio by Custom Time Bracket for Cars Stopped in NYC",
  style = "stack", legend_title = "Stop ended in arrest?", target_labels = c("No", "Yes"),
  subtitle = "The night brackets show a slightly higher arrest rates among all"
)
```

What can we say?

- **Key finding**: The ordering of custom brackets by arrest rate precisely mirrors the expected patterns extracted from the continuous diurnal curve, sorting from lowest to highest conversion probability: Morning commute ($1.0\%$), Early morning ($1.4\%$), Midday ($2.6\%$), Afternoon/evening ($3.6\%$), Late night ($4.9\%$), and Night ($5.1\%$). The nighttime blocks show an arrest probability that stands substantially above the global mean baseline.
- **Interpretation**: This step-up distribution confirms that our discrete custom buckets act as highly stable aggregators of the underlying hourly timeline. The progression validates that the relative risk of an arrest given a stop increases fivefold from the morning commute minimum to the night maximum, proving that the temporal circumstance of a vehicle stop exerts a powerful structural influence on the target outcome.
- **Comment**: When comparing these ratios to the two-wheeled vehicle analysis, we note that standard car stops follow the exact same monotonic risk curve as the day progresses, though the values are compressed into a lower range. Even during the peak window in the Night bracket, the absolute majority of interactions ($94.9\%$) still do not result in an arrest. We can argue that, while the relative risk surges significantly during late hours, an encounter ending in an arrest remains a clear minority event across all examined operational blocks.
- **Proposed Action**: We choose `time_bracket` as the primary hour-related categorical predictor in our final arrest classification models. Because the arrest rate scales linearly across these sequential blocks, this custom categorization will allow our framework to absorb the non-monotonic temporal pattern efficiently while avoiding the excessive degrees of freedom associated with the raw $24$-hour variable.

## Driver related features

### Reported Age

```{r age-bracket-vs-target}
ratio_by_cat(
  vehstops, arrested, age_bracket,
  "Reported Age Bracket", "Arrest Rate (%)", "Arrest Ratio by Reported Age Bracket for Cars Stopped in NYC",
  style = "stack", legend_title = "Stop ended in arrest?", target_labels = c("No", "Yes"),
  subtitle = "The 18-34 age bracket shows the highest arrest rates",
  label_vjust = 0.5,
)
```

```{r confirm-age-bracket-vs-target-ratio}
vehstops %>%
  group_by(age_bracket, arrested) %>%
  summarise(count = n(), .groups = "drop_last") %>%
  mutate(percentage = count / sum(count))
```

What can we say?

- **Key finding**: This plot shows a clear demographic trend where the arrest rate generally increases as the reported age bracket gets younger, reaching its absolute peak in the $25–34$ cohort. The lowest arrest rate is found within the $65+$ demographic at approximately $0.66\%$ (specifically $520$ arrests). The highest ratio belongs to the $25–34$ bracket, where the arrest rate reaches $4.57\%$, which also accounts for the absolute highest volume of individuals arrested with $28821$ instances. The $18–24$ group closely follows with a $4.38\%$ arrest rate ($14380$ arrests). Unlike the 2-wheeled vehicle case where the Under $18$ bracket was the distinct peak, for 4-wheeled vehicles, the under $18$ cohort sits at a lower rate of $2.99\%$ ($278$ arrests), while the Unknown category shows a $1.93\%$ arrest rate ($2291$ arrests).
- **Interpretation**: The data demonstrates a non-linear but highly pronounced age effect where young adults (ages $18$ to $34$) face the highest conditional probability of arrest upon being stopped in a 4-wheeled vehicle. This differs notably from the 2-wheeled analysis where minors (Under $18$) exhibited the highest vulnerability to arrest, suggesting that the dynamics of vehicle ownership, licensing compliance, or enforcement behavior shift considerably between vehicle classes.
- **Comment**: When we look qualitatively at these findings, the contrast between the two analyses is striking. For 2-wheeled vehicles, the younger brackets (especially minors) were heavily penalized, whereas for cars, the risk concentrates heavily on young adults in their prime driving years. The sheer volume of stops and arrests in the $25–34$ bracket highlights that this group dominates the absolute impact on the judicial system. The lower relative arrest rate for minors in 4-wheeled vehicles might reflect stricter initial barriers to entry for driving cars compared to scooters or motorcycles, which yields a different subset of interactions with law enforcement.
- **Proposed Action**: We want to include `age_bracket` as a predictor in our classification models instead of just using the raw continuous `age` feature. Given the structural differences in how age interacts with the target variable across vehicle types, we should also explore an interaction term between `age_bracket` and vehicle type or build separate stratified models to preserve these distinct behavioral patterns.

### Reported Sex

```{r sex-vs-target, fig.width=6}
ratio_by_cat(
  vehstops, arrested, sex,
  "Reported sex", "Arrest Rate (%)", "Arrest Ratio by Reported sex for Cars Stopped in NYC",
  style = "stack", legend_title = "Stop ended in arrest?", target_labels = c("No", "Yes"),
  subtitle = "Male drivers are subject to higher arrest rates, followed by Unknown category",
)
```

```{r confirm-sex-vs-target-ratio}
vehstops %>%
  group_by(sex, arrested) %>%
  summarise(count = n(), .groups = "drop_last") %>%
  mutate(percentage = count / sum(count))
```

What can we say?

- **Key finding**: This highlights a clear disparity in arrest rates across reported sex categories for 4-wheeled vehicle stops. Female drivers (`F`) experience the lowest arrest rate at $1.51\%$ ($5966$ arrests), followed by the unknown or unrecorded category (`U`) at $2.06\%$ (with $276$ arrests). Male drivers (`M`) are subject to the highest rate, with $3.70\%$ of their total stops resulting in an arrest. In fact, the total Male driver arrests stand heavily at $64333$, which represents the vast majority of the overall arrest volume.
- **Interpretation**: The ratio of arrest, given a stop, **is more than twice as high for male drivers compared to female** drivers. This variation affects our current analysis by demonstrating that the proportional gap observed in the 2-wheeled analysis persists strongly in the 4-wheeled domain, confirming that reported sex remains a key predictor that needs to be included in our statistical models.
- **Comment**: The higher arrest rate among male drivers across both vehicle types aligns with broader demographic trends in traffic violations, where they are statistically associated with riskier driving behaviors or more severe infractions. Interestingly, the relative arrest rates are generally lower across all categories for 4-wheeled vehicles than for 2-wheeled vehicles (e.g., $3.70\%$ vs $5.7\%$ for males), but the sheer volume of 4-wheeled stops means that the absolute impact on law enforcement resources is driven entirely by cars. The Unknown category continues to show an intermediate rate, which suggests that incomplete data entry might behave differently depending on the specific circumstances of the encounter.
- **Proposed Action**: We will try to understand the importance of this component, because the low overall baseline association with the target and the heavily imbalanced nature of this feature (especially with the dominant volume of male drivers) may become tricky in the modeling phase. We should consider evaluating feature importance metrics or using techniques like downsampling to ensure our models do not over-rely on the dominant demographic class.

### Reported Ethnicity

```{r ethnicity-vs-target, fig.width=6}
ratio_by_cat(
  vehstops, arrested, ethnicity,
  "Reported ethnicity", "Arrest Rate (%)", "Escalation ratio by Reported Ethnicity for Cars Stopped in NYC",
  style = "stack", legend_title = "Stop ended in arrest?", target_labels = c("No", "Yes"),
  subtitle = "Hispanic and Black Motorists have the highest arrest rates and account for most arrests",
)
```

```{r confirm-ethnicity-vs-target-ratio}
vehstops %>%
  group_by(ethnicity, arrested) %>%
  summarise(count = n(), .groups = "drop_last") %>%
  mutate(percentage = count / sum(count))
```

What can we say?

- **Key finding**: Black drivers represent the largest slice with $656109$ total stops and the highest absolute number of arrests at $32954$ ($5.02\%$). Hispanic drivers experience the second highest escalation rate with $26330$ arrests out of $565531$ stops ($4.66\%$). In contrast, White drivers account for $426594$ stops and $4912$ arrests ($1.15\%$). Finally, Asian / Pacific Islander, Other, and Unknown categories maintain the lowest arrest rates, all remaining near or below $1.51\%$.
- **Interpretation**: The conditional probability of a stop resulting in an arrest is more than four times higher for Black and Hispanic drivers compared to White drivers. These rates are calculated from a massive sample size exceeding two million total records, indicating they are highly stable statistical patterns and completely isolated from small sample variance.
- **Comment**: When comparing these results to the 2-wheeled vehicle case, we observe a noticeable structural shift in volume and magnitude, though the relative disparity remains stark. For 2-wheeled vehicles, Hispanic operators faced the absolute highest arrest rate ($7.24\%$), whereas in the 4-wheeled domain, Black drivers experience both the highest total stop volume and the highest escalation rate ($5.02\%$). While the overall baseline arrest rates are lower for cars than for 2-wheelers across all demographics, the proportional gap between minority groups and White drivers has actually widened significantly from roughly a two fold difference to a four fold difference, signaling a distinct operational dynamic in how car stops escalate.
- **Proposed Action**: This severe imbalance indicates that reported ethnicity is a critical factor that must be included in our models. Due to the high disparity and potential proxy correlations with geographic factors, we should assess downstream model fairness using demographic parity metrics to understand how including `ethnicity` impacts predictive equity across different vehicle classes.

------------------------------------------------------------------------

# Final touches

## Re-Ordering Features

We re-order the features for better display

```{r re-order-dataset}
vehstops_ord <- vehstops %>%
  select(
    year, month, weekday,
    hour, time_bracket,
    area, boro,
    sex, ethnicity, 
    age, age_bracket,
    arrested
  )
```

Here is the final version.

```{r summary-final-dataset}
print(dfSummary(vehstops_ord), method="render")
```

## Saving the dataset

```{r save-dataset}
write.csv(vehstops_ord, "data/vehiclestops_4_wheels_final.csv", row.names = FALSE)
```
