---
title: "Predicting Excessive Enforcement Patterns in Police Behavior After Vehicles Stops in New York"
author: "Gianmaria Pizzo"
output:
  html_document:
    fig_caption: yes
    theme: flatly
    highlight: pygments
    code_folding: show
    kee_md: yes
    toc: yes
    toc_depth: 4
    number_sections: no
    toc_float:
      smooth_scroll: no
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

These `.Rmd` files represent the project submission for the course [Statistical Inference and Learning](https://www.unive.it/data/course/451271) by Professor [Cristiano Varin](https://www.unive.it/data/people/5592728) at [Ca' Foscari University of Venice](https://www.unive.it/) for the year 2024/2025.

# Problem Statement

## Goal of the Project

As the title of the project suggests, the goal of the project is to assess **whether** we can say (with enough confidence) that the **police behavior shows excessive enforcement patterns**, after a vehicle stop has been conducted.

In particular, we focus on studying if there exist relationships between violent outcomes and specific groups of people, or areas. This should help to shed some light on the alleged "systemic nature" of the stops.

However, some of the limiting factors for our study, include the small time frame of the data, the absence of information about the reason behind the halt, the self-report nature of some variables and the partial anonymization of the values.

Thus, this analysis will not be limited to performing basic univariate, bivariate or multivariate exploratory analysis. Instead, we first start from formatting the data in a way we can understand the meaning of each variable, and the issues that come with it.

Furthermore, we may also introduce transformations or other means of engineering, until we feel satisfied with the final data. We might add new insights about the effects of time-of-day, day-of-week, precinct to reveal new observations, such as whether racial disparities peak during specific hours or hot-spots.

All of the manipulations, are to be carefully tested for. In fact, we might end up using some automatic variable selection approach to and other means of model selection.

Only then, we can proceed to design the final plot, check for outliers and collinearity, and formulate our main assumptions. We want to prevent complex interactions between terms that might limit the interpretability. This does not mean we are not going to try and evaluate combinations of features (that may also substitute the initial ones) nonetheless. The result of this initial part, aims to increase the reliability of the information, crucial to the next step.

The core of this project is to pragmatically build, analyze and interpret (multiple) predictive models capable of estimating whether a stop escalates into a search, seizure, arrest, or use-of-force based (forecasting outcome) and the significance of the factors used to build such models. This means, we want to differentiate between different target features by context.

Therefore, this requires an extensive training process, including the use of k-fold-cross validation, subset selection and other techniques to promote theoretical and empirical validity.

Finally, as this is mainly a classification problem, we will include all possible classification metrics and perform extensive tests. We hope to quantify both the accuracy and the confidence of our inferences to determine what this data is telling us.

## Previous Research and Novel Approach

Even though there exist some public reports (e.g., [NYCLU - NYPD Vehicle Stops Data](https://www.nyclu.org/data/nypd-vehicle-stops-data)), they only cover descriptive statistics.

As already stated above, our analysis will provide a better insight, through predictive modeling, model comparison (via AIC and cross-validation), and inference on interactions (e.g., race × time or precinct).

We hope this analysis dissipates or pinpoints the police bias clearly, even with limited data, as it might lead to either confirming that some areas and races are more subject to police enforcement than others, or that this has been a product of the prominent propaganda, and the police is "just doing its job".

Last but not least, we do not approach the data with any prior research bias, unlike many journalists.

## Questions we want to ask

Obviously, we could not proceed forward if we did not have some designated questions for our data.

Is it more likely to be the subject of an episode of excessive enforcement behavior...

-   When stopped in a specific area (or borough)?
-   When stopped by a specific command unit?
-   When stopped outside of the 9am-5pm "window"?
-   When stopped during the weekend?
-   If the driver is under 35 years old?
-   If the driver belongs to a specific ethnicity?
-   If the driver identifies with a specific gender?
-   If the vehicles used belongs to a specific category?

Furthermore, we may also formulate questions of the kind

-   Is there a specific sub-category of drivers that is subject of most of these episodes?
-   Is there a specific ethnicity targeted by excessive enforcement behavior in a particular area?
-   Proportionally speaking, since Donald Trump became president (January 20, 2025), has any of the previous aspect changed?

## Goal of this rmarkdown file - TODO

The EDA is the first step to understanding the data and its nature.

This includes analyzing the single features (Univariate Analysis), their interactions with the target (Bivariate Analysis) and their interactions with other features (Multivariate Analysis and Collinearity Check). However, the latter might be hard to compute as it requires the estimation of a regression model that follows certain assumptions and it is usually easier to check this when using linear models.

During my exploration I will look out for possible outliers, correlation and any other statistically relevant problem that might affect my final prediction.

We will start with the environment.

Note that, from now on, I will shift adopt a more **formal or inclusive tone** (e.g., the "editorial we").

------------------------------------------------------------------------

# Environment

The first to do, is to build a reproducible environment for our project.

```{r}
requirements=c( 
  "remotes", "summarytools", "forcats", "scales", "purrr",
  "stringr", "plyr", "dbplyr", "dplyr", "dtplyr", "tidyr", "tidyselect",
  "timeDate", "timechange", "tzdb", "hms", "readr",
  "lubridate", "prettyunits", "RColorBrewer", "viridis", "rcartocolor", "ggbeeswarm", "emmeans",
  "sf", "plotly", "ggplot2", "ggmap", "ggdensity", "UpSetR", "patchwork"
)
```

```{r, include=FALSE}
for (req in requirements){
  if (!require(req, character.only = TRUE)){
      install.packages(req)
  }
}
```

Note that some of the packages may not be used, but to avoid any error related to dependencies, we are going to keep all of these libraries. Now, the loading of all packages should be immediate.

```{r, include=FALSE}
for (req in requirements){
  if (!require(req, character.only = TRUE)){
      library(req)
  }
}
```

```{r plot-helpers, include=FALSE}
plot_binary_bar <- function(DF, FEATURE, OPTION_0, OPTION_1, X_LABEL, Y_LABEL, TITLE,
                            pad = 0.12, text_vjust = -0.35) {

  df_sum <- DF %>%
    mutate(
      .raw  = {{ FEATURE }},
      .feat = dplyr::case_when(
        as.character(.raw) %in% c("1","TRUE","true")  ~ 1L,
        as.character(.raw) %in% c("0","FALSE","false") ~ 0L,
        TRUE ~ NA_integer_
      )
    ) %>%
    filter(!is.na(.feat)) %>%
    count(.feat, name = "n") %>%
    mutate(.lab = factor(.feat, levels = c(0, 1), labels = c(OPTION_0, OPTION_1)))

  ymax <- max(df_sum$n, na.rm = TRUE)

  ggplot(df_sum, aes(x = .lab, y = n, fill = .lab)) +
    geom_col(width = 0.8, show.legend = FALSE) +
    geom_text(aes(label = scales::comma(n)), vjust = text_vjust) +
    scale_y_continuous(limits = c(0, ymax * (1 + pad)),
                       expand = expansion(mult = c(0, 0.02))) +
    coord_cartesian(clip = "off") +
    theme(plot.margin = margin(t = 10, r = 5.5, b = 5.5, l = 5.5)) +
    labs(x = X_LABEL, y = Y_LABEL, title = TITLE)
}

# ---------- Heatmaps ----------
# Heatmap of escalation rate by Y (categorical) × X (numeric)
heat_escalation_yx <- function(DF, Y, X, X_TITLE, Y_TITLE, TITLE,
                               min_n = 0,            # hide cells with < min_n stops
                               drop_na = TRUE,       # drop NA in Y/X
                               x_as_factor = FALSE   # treat X as factor (for non-integer X)
) {
  dat <- DF
  if (drop_na) {
    dat <- dat %>% filter(!is.na({{Y}}), !is.na({{X}}), !is.na(any_escalation))
  } else {
    dat <- dat %>% filter(!is.na(any_escalation))
  }

  # Order Y by overall escalation rate
  y_order <- dat %>%
    group_by({{Y}}) %>%
    summarise(rate = mean(any_escalation == 1, na.rm = TRUE), .groups = "drop") %>%
    arrange(rate) %>%
    pull({{Y}})

  # Grouped rates and counts
  grid <- dat %>%
    group_by({{Y}}, {{X}}) %>%
    summarise(rate = mean(any_escalation == 1, na.rm = TRUE),
              n    = n(), .groups = "drop") %>%
    mutate(
      .y = factor(as.character({{Y}}), levels = as.character(y_order)),
      .x = if (x_as_factor) factor({{X}}) else {{X}},
      rate_plot = ifelse(n >= min_n, rate, NA_real_)
    )

  ggplot(grid, aes(.x, .y, fill = rate_plot)) +
    geom_tile(width = 1, height = 0.9) +
    scale_fill_viridis_c(labels = scales::percent, na.value = "grey90", name = "Rate") +
    labs(x = X_TITLE, y = Y_TITLE, title = TITLE) +
    theme_minimal() +
    theme(panel.grid = element_blank())
}

# Heatmap of escalation rate by Y (categorical) × X (categorical)
heat_escalation_catcat <- function(
  DF, Y, X,
  X_TITLE, Y_TITLE, TITLE,
  min_n = 0,                    # hide cells with < min_n stops
  order_y_by = c("rate","freq","none"),
  order_x_by = c("rate","freq","none"),
  alpha_by_n = FALSE            # TRUE = shade tiles by count as alpha
) {
  order_y_by <- match.arg(order_y_by)
  order_x_by <- match.arg(order_x_by)

  dat <- DF %>%
    filter(!is.na({{Y}}), !is.na({{X}}), !is.na(any_escalation))

  # Stats for ordering
  y_stats <- dat %>%
    group_by({{Y}}) %>%
    summarise(rate = mean(any_escalation == 1, na.rm = TRUE),
              n = n(), .groups = "drop")
  x_stats <- dat %>%
    group_by({{X}}) %>%
    summarise(rate = mean(any_escalation == 1, na.rm = TRUE),
              n = n(), .groups = "drop")

  y_levels <- switch(
    order_y_by,
    rate = y_stats %>% arrange(rate) %>% pull({{Y}}),
    freq = y_stats %>% arrange(n)    %>% pull({{Y}}),
    none = unique(dat %>% pull({{Y}}))
  )
  x_levels <- switch(
    order_x_by,
    rate = x_stats %>% arrange(rate) %>% pull({{X}}),
    freq = x_stats %>% arrange(n)    %>% pull({{X}}),
    none = unique(dat %>% pull({{X}}))
  )

  # Cell rates + counts
  grid <- dat %>%
    group_by({{Y}}, {{X}}) %>%
    summarise(rate = mean(any_escalation == 1, na.rm = TRUE),
              n    = n(), .groups = "drop") %>%
    mutate(
      .y = factor(as.character({{Y}}), levels = as.character(y_levels)),
      .x = factor(as.character({{X}}), levels = as.character(x_levels)),
      rate_plot = ifelse(n >= min_n, rate, NA_real_)
    )

  p <- ggplot(grid, aes(.x, .y)) +
    labs(x = X_TITLE, y = Y_TITLE, title = TITLE) +
    theme_minimal() +
    theme(panel.grid = element_blank()) +
    scale_fill_viridis_c(labels = percent, na.value = "grey90", name = "Rate")

  if (alpha_by_n) {
    p <- p +
      geom_tile(aes(fill = rate_plot, alpha = n), width = 0.95, height = 0.95) +
      scale_alpha_continuous(range = c(0.25, 1), name = "Stops")
  } else {
    p <- p +
      geom_tile(aes(fill = rate_plot), width = 0.95, height = 0.95)
  }

  p
}

# Bivariate % of TARGET==1 across CAT (no split)
ratio_by_cat <- function(
  DF, TARGET, CAT,
  x_title, y_title, title,
  target_labels = c("Not escalated","Escalated"),
  style = c("stack","rate"),
  order_by = c("rate","freq","none"),
  percent_accuracy = 0.1,
  colors = c("#00bfc4", "#f8766d")   # <- blue (0), red (1)
){
  style    <- match.arg(style)
  order_by <- match.arg(order_by)

  dat <- DF %>%
    mutate(
      .t_raw = {{ TARGET }},
      .t     = case_when(
        as.character(.t_raw) %in% c("1","TRUE","true")  ~ 1L,
        as.character(.t_raw) %in% c("0","FALSE","false") ~ 0L,
        TRUE ~ NA_integer_
      ),
      .cat = {{ CAT }}
    ) %>%
    filter(!is.na(.t), !is.na(.cat))

  rates <- dat %>% group_by(.cat) %>%
    summarise(total = n(), rate = mean(.t == 1), .groups = "drop")

  cat_levels <- switch(
    order_by,
    rate = rates %>% arrange(rate)  %>% pull(.cat),
    freq = rates %>% arrange(total) %>% pull(.cat),
    none = unique(dat$.cat)
  )

  if (style == "stack") {
    stack <- dat %>%
      count(.cat, .t, name = "n") %>%
      group_by(.cat) %>%
      mutate(ratio = n / sum(n)) %>%
      ungroup() %>%
      mutate(
        .cat_f = factor(as.character(.cat), levels = as.character(cat_levels)),
        .t_f   = factor(.t, levels = c(0,1), labels = target_labels)
      ) %>%
      left_join(rates, by = ".cat")

    ggplot(stack, aes(x = .cat_f, y = ratio, fill = .t_f)) +
      geom_col(position = "fill") +
      geom_text(
        data = subset(stack, .t_f == target_labels[2]),
        aes(label = scales::percent(rate, accuracy = percent_accuracy)),
        position = position_fill(vjust = 0.5), size = 3
      ) +
      scale_y_continuous(labels = scales::percent) +
      scale_fill_manual(values = setNames(colors, target_labels)) +  # <- here
      labs(x = x_title, y = y_title, fill = NULL, title = title) +
      theme_minimal() +
      theme(axis.text.x = element_text(angle = 45, hjust = 1),
            panel.grid = element_blank())
  } else {
    rates <- rates %>%
      mutate(.cat_f = factor(as.character(.cat), levels = as.character(cat_levels)))

    ggplot(rates, aes(x = .cat_f, y = rate)) +
      geom_col(width = 0.8, fill = colors[2]) +  # bar color for TARGET==1 (red)
      geom_text(aes(label = scales::percent(rate, accuracy = percent_accuracy)),
                vjust = -0.35, size = 3) +
      scale_y_continuous(labels = scales::percent, expand = expansion(mult = c(0, 0.12))) +
      coord_cartesian(clip = "off") +
      labs(x = x_title, y = y_title, title = title,
           subtitle = "Bars show % of target = 1") +
      theme_minimal() +
      theme(axis.text.x = element_text(angle = 45, hjust = 1),
            panel.grid = element_blank())
  }
}


# 100% stacked ratio bars of TARGET by CAT, split by SPLIT (2-level)
ratio_by_cat_split <- function(
  DF, TARGET, CAT, SPLIT,
  x_title, y_title, title,
  target_labels = c("Not escalated", "Escalated"),
  split_labels  = NULL,
  palette = "Paired",
  percent_accuracy = 0.1
){
  dat <- DF %>%
    mutate(
      .target_raw = {{ TARGET }},
      .target_bin = case_when(
        as.character(.target_raw) %in% c("1","TRUE","true")  ~ 1L,
        as.character(.target_raw) %in% c("0","FALSE","false") ~ 0L,
        TRUE ~ NA_integer_
      ),
      .cat   = {{ CAT }},
      .split = {{ SPLIT }}
    ) %>%
    filter(!is.na(.cat), !is.na(.split), !is.na(.target_bin))

  # order categories by overall target rate (low -> high)
  cat_order <- dat %>%
    group_by(.cat) %>%
    summarise(rate = mean(.target_bin == 1), .groups = "drop") %>%
    arrange(rate) %>% pull(.cat)

  # tidy labels
  dat <- dat %>%
    mutate(
      .cat      = factor(as.character(.cat), levels = as.character(cat_order)),
      .target_f = factor(.target_bin, levels = c(0,1), labels = target_labels),
      .split_f  = {
        lv <- unique(as.character(.split))
        if (!is.null(split_labels) && length(split_labels) == length(lv))
          factor(as.character(.split), levels = lv, labels = split_labels)
        else factor(as.character(.split), levels = lv)
      }
    )

  plot_df <- dat %>%
    count(.split_f, .cat, .target_f, name = "n") %>%
    group_by(.split_f, .cat) %>%
    mutate(ratio = n / sum(n)) %>%
    ungroup()

  ggplot(plot_df, aes(x = .cat, y = ratio, fill = .target_f)) +
    geom_col(position = "fill") +
    geom_text(aes(label = percent(ratio, accuracy = percent_accuracy)),
              position = position_fill(vjust = 0.5), size = 3) +
    scale_y_continuous(labels = percent) +
    scale_fill_brewer(palette = palette, name = rlang::as_name(rlang::enquo(TARGET))) +
    labs(x = x_title, y = y_title, title = title) +
    facet_wrap(~ .split_f, nrow = 1) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
          panel.grid = element_blank())
}
```

Let us load the data now.

```{r}
vehstops = read.csv("data/vehiclestops.csv")
vehstops2 = read.csv("data/vehiclestops_clean.csv")
```

```{r}
summary(vehstops)
```


------------------------------------------------------------------------

# Univariate EDA

TODO: Introduction
TODO: Correct the following

In this part we perform a deeper analysis of the target feature, and some of the most important features (according to the domain knowledge in notebook 0) against their distribution in the dataset.

Our goal here is to polish the data, by removing outliers and possible confounders that may introduce bias. Furthermore, we need to understand its distribution and normalizing it if necessary.

Possibly, we can introduce new variables that are transformations or combinations of other ones, and evaluate their contributions later with a bivariate analysis and/or model testing.

Hopefully, this process will reduce the variance.

TODO: Comment everything

## Target feature distribution

```{r target-dist}
plot_binary_bar(
  vehstops, any_escalation, 
  "No", "Yes",
  "Did the stop escalate?", "", "Counts of Stops"
)
```

## Enforcement and Escalation features Distribution

### Core Outcomes

```{r enforcement-dist}
p1 <- plot_binary_bar(
  vehstops, veh_searched, 
  "No", "Yes",
  "Was the vehicle searched?", "", ""
)

p2 <- plot_binary_bar(
  vehstops, veh_seized,
  "No", "Yes",
  "Vehicle Seized?", "", ""
)

p3 <- plot_binary_bar(
  vehstops, force_used,
  "No", "Yes",
  "Did they use force?", "", ""
)

p4 <- plot_binary_bar(
  vehstops, arrested, 
  "No", "Yes",
  "Arrested?", "", ""
)

p5 <- plot_binary_bar(
  vehstops, summoned,
  "No", "Yes",
  "Motorist was summoned?", "", ""
)


row1 <- p1 | p2 | p3
row2 <- p4 | p5 

row1 / plot_spacer() / row2 +
  plot_layout(heights = c(1, 0.08, 1), guides = "collect") + 
  plot_annotation(title = "Counts of stops by outcome - Initial Features") &
  theme(legend.position = "bottom") 
```

### Extra Outcomes

```{r enforcement-extra-dist}
p1 <- plot_binary_bar(
  vehstops, search_without_consent, 
  "No", "Yes",
  "Searched without consent?", "", ""
)

p2 <- plot_binary_bar(
  vehstops, seizure_and_summoning,
  "No", "Yes",
  "Seizure and summoning?", "", ""
)

p3 <- plot_binary_bar(
  vehstops, seizure_and_arrest,
  "No", "Yes",
  "Seizure and arrest?", "", ""
)

p4 <- plot_binary_bar(
  vehstops, force_and_arrest, 
  "No", "Yes",
  "Use of force and arrest?", "", ""
)

row1 <- p1 | p2
row2 <- p3 | p4

row1 / plot_spacer() / row2 +
  plot_layout(heights = c(1, 0.08, 1), guides = "collect") + 
  plot_annotation(title = "Counts of stops by outcome - Engineered Features") &
  theme(legend.position = "bottom") 
```

If we dig into it we can try to see whether our idea of grouping the outcomes into one category demonstrates some relevance with respect of all possible

```{r prevalence-of-enforcement}

setA_cols <- intersect(
  c("veh_searched","search_without_consent","veh_seized",
    "summoned","force_used","arrested", "seizure_and_summoning","seizure_and_arrest","force_and_arrest"),
  names(vehstops)
)

setB_cols <- intersect(
  c("veh_searched", "veh_seized", "summoned", "arrested", "force_used"),
  names(vehstops)
)

setC_cols <- intersect(
  c("summoned", "veh_searched", "arrested"),
  names(vehstops)
)

A <- vehstops %>%
  select(all_of(setA_cols)) %>%
  pivot_longer(everything(), names_to = "feature", values_to = "val") %>%
  group_by(feature) %>%
  summarise(prevalence = mean(val == 1, na.rm = TRUE), .groups = "drop") %>%
  mutate(set = "A) All outcomes")

B <- vehstops %>%
  select(all_of(setB_cols)) %>%
  pivot_longer(everything(), names_to = "feature", values_to = "val") %>%
  group_by(feature) %>%
  summarise(prevalence = mean(val == 1, na.rm = TRUE), .groups = "drop") %>%
  mutate(set = "B) No combos")

C <- vehstops %>%
  select(all_of(setC_cols)) %>%
  pivot_longer(everything(), names_to = "feature", values_to = "val") %>%
  group_by(feature) %>%
  summarise(prevalence = mean(val == 1, na.rm = TRUE), .groups = "drop") %>%
  mutate(set = "C) Top 3 outcomes")

all_prev <- bind_rows(A, B, C) %>%
  mutate(
    feature = str_replace_all(feature, "_", " "),
    set = factor(set, levels = c("A) All outcomes",
                                 "B) No combos",
                                 "C) Top 3 outcomes"))
  )

all_prev <- all_prev %>%
  mutate(
    feature_set = paste(feature, set, sep = "‖")
  ) %>%
  group_by(set) %>%
  mutate(feature_set = fct_reorder(feature_set, prevalence)) %>%
  ungroup()

ggplot(all_prev, aes(prevalence, feature_set)) +
  geom_col(width = 0.8) +
  scale_x_continuous(labels = percent) +
  scale_y_discrete(labels = function(x) sub("‖.*$", "", x)) +
  labs(x = "Prevalence", y = NULL, title = "Prevalence of outcome flags") +
  facet_grid(rows = vars(set), scales = "free_y") +
  theme_minimal()
```

## Command Unit, Geospatial and Area related Features Distribution

### Area

```{r area-dist}
vehstops %>%
  filter(!is.na(area)) %>%
  count(area, name = "n") %>%
  mutate(area = fct_reorder(area, n, .desc = TRUE)) %>% 
  ggplot(aes(x = factor(area), y = n, fill = factor(area))) +
  geom_col(width = 0.8, show.legend = FALSE) +
  geom_text(aes(label = scales::comma(n)), vjust = -0.35, size = 3) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
  coord_cartesian(clip = "off") +
  labs(x = "Area", y = "", title = "Count of Stops") +
  theme_minimal()
```

### Borough

```{r boro-dist}
vehstops %>%
  filter(!is.na(boro)) %>%
  count(boro, name = "n") %>%
  mutate(boro = fct_reorder(boro, n, .desc = TRUE)) %>% 
  ggplot(aes(x = factor(boro), y = n, fill = factor(boro))) +
  geom_col(width = 0.8, show.legend = FALSE) +
  geom_text(aes(label = scales::comma(n)), vjust = -0.35, size = 3) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
  coord_cartesian(clip = "off") +
  labs(x = "Borough", y = "", title = "Count of Stops") +
  theme_minimal() + 
  scale_x_discrete(guide = guide_axis(angle = 30, n.dodge = 1))
```

### Command

```{r command-dist}
vehstops %>%
  filter(!is.na(command)) %>%
  count(command, name = "n") %>%
  mutate(command = fct_reorder(as.factor(command), n, .desc = TRUE)) %>% 
  ggplot(aes(x = factor(command), y = n, fill = factor(command))) +
  geom_col(width = 0.8, show.legend = FALSE) +
  geom_text(aes(label = scales::comma(n)), vjust = -0.5, size = 1) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
  coord_cartesian(clip = "off") +
  labs(x = "Command", y = "", title = "Count of Stops") +
  theme_minimal() + 
  scale_x_discrete(guide = guide_axis(angle = 30, n.dodge = 2))
```

### Latitude and Longitude

```{r stop-density}
nyc_bbox <- list(lat_min = 40.48, lat_max = 40.92, lon_min = -74.27, lon_max = -73.68)
veh_geo_nyc <- vehstops %>%
  filter(!is.na(latitude), !is.na(longitude)) %>%
  filter(
    between(latitude,  nyc_bbox$lat_min, nyc_bbox$lat_max),
    between(longitude, nyc_bbox$lon_min, nyc_bbox$lon_max)
  )

bins <- 80
min_n <- 25

lon_breaks <- seq(nyc_bbox$lon_min, nyc_bbox$lon_max, length.out = bins + 1)
lat_breaks <- seq(nyc_bbox$lat_min, nyc_bbox$lat_max, length.out = bins + 1)

grid_sum <- veh_geo_nyc %>%
  mutate(
    xbin = cut(longitude, lon_breaks, include.lowest = TRUE, right = FALSE),
    ybin = cut(latitude,  lat_breaks, include.lowest = TRUE, right = FALSE)
  ) %>%
  group_by(xbin, ybin) %>%
  summarise(
    n    = n(),
    rate = mean(any_escalation == 1, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    xi = as.integer(xbin),
    yi = as.integer(ybin),
    # bin centers (for geom_tile positioning)
    x = (lon_breaks[xi] + lon_breaks[xi + 1]) / 2,
    y = (lat_breaks[yi] + lat_breaks[yi + 1]) / 2,
    # optionally suppress super-sparse bins in the rate coloring
    rate_plot = ifelse(n >= min_n, rate, NA_real_)
  )

p1 <- ggplot(grid_sum, aes(x, y, fill = n)) +
  geom_tile() +
  coord_equal(xlim = c(nyc_bbox$lon_min, nyc_bbox$lon_max),
              ylim = c(nyc_bbox$lat_min, nyc_bbox$lat_max), expand = FALSE) +
  scale_fill_viridis_c(trans = "log10", name = "Stops") +
  labs(title = "Stop density (2D bins)", x = "Longitude", y = "Latitude")

p1
```

## Time related features Distribution

### Year

```{r year-distribution}
vehstops %>%
  filter(!is.na(year)) %>%
  count(year, name = "n") %>%
  ggplot(aes(x = factor(year), y = n, fill = factor(year))) +
  geom_col(width = 0.8, show.legend = FALSE) +
  geom_text(aes(label = scales::comma(n)), vjust = -0.35, size = 3) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
  coord_cartesian(clip = "off") +
  labs(x = "Year", y = "", title = "Count of Stops") +
  theme_minimal()
```

### Month

Since it is not safe to group by year 2025 only has around 6 months of data), we divide this in 3 parts.

```{r month-distribution}
p1 <- vehstops %>%
  filter(!is.na(year) & year == 2023) %>%
  count(month, name = "n") %>%
  ggplot(aes(x = factor(month), y = n, fill = factor(month))) +
  geom_col(width = 0.8, show.legend = FALSE) +
  geom_text(aes(label = scales::comma(n)), vjust = -0.35, size = 3) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
  coord_cartesian(clip = "off") +
  labs(x = "Months - Year 2023", y = "") +
  theme_minimal()

p2 <- vehstops %>%
  filter(!is.na(year) & year == 2024) %>%
  count(month, name = "n") %>%
  ggplot(aes(x = factor(month), y = n, fill = factor(month))) +
  geom_col(width = 0.8, show.legend = FALSE) +
  geom_text(aes(label = scales::comma(n)), vjust = -0.35, size = 3) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
  coord_cartesian(clip = "off") +
  labs(x = "Months - Year 2024", y = "") +
  theme_minimal()
  
p3 <- vehstops %>%
  filter(!is.na(year) & year == 2025) %>%
  count(month, name = "n") %>%
  ggplot(aes(x = factor(month), y = n, fill = factor(month))) +
  geom_col(width = 0.8, show.legend = FALSE) +
  geom_text(aes(label = scales::comma(n)), vjust = -0.35, size = 3) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
  coord_cartesian(clip = "off") +
  labs(x = "Months - Year 2025", y = "") +
  theme_minimal()

(p1 / p2 / p3) + 
  plot_layout(widths = c(2, 2), guides = "collect") +
  plot_annotation(title = "Counts of Stops by Month") &
  theme(legend.position = "bottom")
```

In fact, now it makes more sense, because the cumulative sum of the first six months of the year of 2025 is higher than the first six months of the previous years! 

### Day and Dow

```{r day-dist}
vehstops %>%
  filter(!is.na(day)) %>%
  count(day, name = "n") %>%
  ggplot(aes(x = factor(day), y = n, fill = factor(day))) +
  geom_col(width = 0.8, show.legend = FALSE) +
  geom_text(aes(label = scales::comma(n)), vjust = -0.35, size = 2) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
  coord_cartesian(clip = "off") +
  labs(x = "Day", y = "", title="Counts of Stops by Day of the Month") +
  theme_minimal()
```

The day of the month does not seem to matter much and we see a decrease on the `31` day because not all months have them.

```{r dow-dis}
vehstops %>%
  filter(!is.na(dow)) %>%
  count(dow, name = "n") %>%
  ggplot(aes(x = factor(dow), y = n, fill = factor(dow))) +
  geom_col(width = 0.8, show.legend = FALSE) +
  geom_text(aes(label = scales::comma(n)), vjust = -0.35, size = 3) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
  coord_cartesian(clip = "off") +
  labs(x = "Day", y = "", title="Counts of Stops by Day of the Week") +
  theme_minimal()
```

This `doy` is hard to account for, and we are going to remove it from the clean version

```{r remove-doy}
vehstops2 <- vehstops2 %>%
  select(-doy)
```

### Hour, Working Hour and Time Bracket

```{r hour-dist}
vehstops %>%
  filter(!is.na(hour)) %>%
  count(hour, name = "n") %>%
  ggplot(aes(x = factor(hour), y = n, fill = factor(hour))) +
  geom_col(width = 0.8, show.legend = FALSE) +
  geom_text(aes(label = scales::comma(n)), vjust = -0.35, size = 2.5) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
  coord_cartesian(clip = "off") +
  labs(x = "Hour", y = "", title = "Count of Stops by Hour of the Day") +
  theme_minimal() 
```

```{r time-bracket-distribution}
vehstops %>%
  filter(!is.na(time_bracket)) %>%
  count(time_bracket, name = "n") %>%
  mutate(time_bracket = fct_reorder(as.factor(time_bracket), n, .desc = TRUE)) %>% 
  ggplot(aes(x = factor(time_bracket), y = n, fill = factor(time_bracket))) +
  geom_col(width = 0.8, show.legend = FALSE) +
  geom_text(aes(label = scales::comma(n)), vjust = -0.35, size = 2.5) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
  coord_cartesian(clip = "off") +
  labs(x = "Time Bracket", y = "", title = "Count of Stops by Time Bracket") +
  theme_minimal() + 
  scale_x_discrete(guide = guide_axis(angle = 30, n.dodge = 1))
```

```{r working-hour-distribution}
plot_binary_bar(
  vehstops, working_hour, 
  "No", "Yes",
  "Was it between 9am and 5pm?", "", "Counts of Stops by 9-5 Window"
)
```


### Minute

```{r minute-distribution}
vehstops %>%
  filter(!is.na(minute)) %>%
  count(minute, name = "n") %>%
  ggplot(aes(x = factor(minute), y = n, fill = factor(minute))) +
  geom_col(width = 0.8, show.legend = FALSE) +
  geom_text(aes(label = scales::comma(n)), vjust = -0.35, size = 1) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
  coord_cartesian(clip = "off") +
  labs(x = "Minute", y = "", title = "Count of Stops by Minute") +
  theme_minimal() + 
  scale_x_discrete(guide = guide_axis(angle = 30, n.dodge = 2))
```

This seems to either require some binning or just drop, but we are going to drop it for the clean version.

```{r drop-minute}
vehstops2 <- vehstops2 %>%
  select(-minute)
```

### Weekend, Holiday and Event Day

```{r weekend-holiday-event-distribution}
p1 <- plot_binary_bar(
  vehstops, weekend, 
  "No", "Yes",
  "During a weekend?", "", "Counts of Stops"
)

p2 <- plot_binary_bar(
  vehstops, holiday,
  "No", "Yes",
  "During an holiday?", "", "Counts of Stops"
)

p3 <- plot_binary_bar(
  vehstops, event_day,
  "No", "Yes",
  "During an event?", "", "Counts of Stops"
)

(p1 + p2 + p3) + 
  plot_layout(ncol = 3, nrow = 1, widths = c(1, 1), guides = "collect") +
  theme(legend.position = "bottom") 
```

None of the two matter much in terms of stops

## Motorist related features

### Reported Age

We have mainly 2 issues here. First, we have many instances of drivers with unknown age

```{r age-bracket-dist}
vehstops %>%
  filter(!is.na(age_bracket)) %>%
  count(age_bracket, name = "n") %>%
  ggplot(aes(x = factor(age_bracket), y = n, fill = factor(age_bracket))) +
  geom_col(width = 0.8, show.legend = FALSE) +
  geom_text(aes(label = scales::comma(n)), vjust = -0.35, size = 3) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +  # top padding
  coord_cartesian(clip = "off") +
  labs(x = "Reported Age Bracket", y = "", title = "Count of Stops by Reported Age Bracket") +
  theme_minimal()
```

But if we forget about them and we plot the distribution

```{r age-distribution}
vehstops %>%
  filter(!is.na(age)) %>%
  count(age) %>%
  ggplot(aes(y = n,x = factor(age), fill = factor(age))) +
  geom_col(width = 0.8, na.rm = TRUE, show.legend = FALSE) +
  labs(x = "Reported Age", y = "", title = "Count of Stops by Reported Age") +
  scale_x_discrete(guide = guide_axis(angle = 45, n.dodge = 2))
```

We can see that our second issue is the presence of very old drivers (90+),

```{r very-old-drivers}
vehstops %>%
  select(year, month, hour, time_bracket, area, age, sex, race, veh_categ, any_escalation) %>%
  filter(age > 90) %>%
  dfSummary()
```

Which may not contribute to much here. So in the second version of the dataset we are going to remove drivers older than 90 (but for the full dataset, we are going to wait).

```{r}
vehstops2 <- vehstops2 %>%
  filter(age <= 90)
```

Thus, 

```{r age-distribution2}
vehstops %>%
  filter(!is.na(age) & age < 90) %>%
  count(age) %>%
  ggplot(aes(y = n,x = factor(age), fill = factor(age))) +
  geom_col(width = 0.8, na.rm = TRUE, show.legend = FALSE) +
  labs(x = "Reported Age", y = "", title = "Count of Stops by Reported Age") +
  scale_x_discrete(guide = guide_axis(angle = 45, n.dodge = 2))
```

If there is evidence of a better binning with respect to the escalation, we might as well re-bin this feature

### Reported Sex

```{r sex-distribution}
vehstops %>%
  filter(!is.na(sex)) %>%
  count(sex, name = "n") %>%
  ggplot(aes(x = factor(sex), y = n, fill = factor(sex))) +
  geom_col(width = 0.8, show.legend = FALSE) +
  geom_text(aes(label = scales::comma(n)), vjust = -0.35, size = 3) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
  coord_cartesian(clip = "off") +
  labs(x = "Reported Sex", y = "", title = "Count of Stops by Reported Sex") +
  theme_minimal()
```

### Reported Race

```{r race-distribution}
vehstops %>%
  filter(!is.na(race)) %>%
  count(race, name = "n") %>%
  ggplot(aes(x = factor(race), y = n, fill = factor(race))) +
  geom_col(width = 0.8, show.legend = FALSE) +
  geom_text(aes(label = scales::comma(n)), vjust = -0.35, size = 3) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
  coord_cartesian(clip = "off") +
  labs(x = "Reported Race", y = "", title = "Count of Stops by Reported Race") +
  theme_minimal() +
  scale_x_discrete(guide = guide_axis(angle = 30, n.dodge = 1))
```

## Vehicle related features

```{r veh-categ-distribution}
vehstops %>%
  filter(!is.na(veh_categ)) %>%
  count(veh_categ, name = "n") %>%
  ggplot(aes(x = factor(veh_categ), y = n, fill = factor(veh_categ))) +
  geom_col(width = 0.8, show.legend = FALSE) +
  geom_text(aes(label = scales::comma(n)), vjust = -0.35, size = 3) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
  coord_cartesian(clip = "off") +
  labs(x = "Vehicle Category", y = "", title = "Count of Stops by Vehicle Category") +
  theme_minimal() +
  scale_x_discrete(guide = guide_axis(angle = 30, n.dodge = 1))
```

```{r veh-categ-clean-distribution}
vehstops %>%
  filter(!is.na(veh_categ_clean)) %>%
  count(veh_categ_clean, name = "n") %>%
  ggplot(aes(x = factor(veh_categ_clean), y = n, fill = factor(veh_categ_clean))) +
  geom_col(width = 0.8, show.legend = FALSE) +
  geom_text(aes(label = scales::comma(n)), vjust = -0.35, size = 3) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
  coord_cartesian(clip = "off") +
  labs(x = "Vehicle Category Clean", y = "", title = "Count of Stops by Vehicle Category Clean") +
  theme_minimal() +
  scale_x_discrete(guide = guide_axis(angle = 30, n.dodge = 1))
```

```{r vehicle-features-distribution}
p1 <- plot_binary_bar(
  vehstops, veh_4_wheels, 
  "2", "4",
  "Count of wheels", "", "Counts of Stops"
)

p2 <- plot_binary_bar(
  vehstops, veh_private,
  "Commercial", "Private",
  "Ownership", "", "Counts of Stops"
)

p3 <- plot_binary_bar(
  vehstops, veh_motorized,
  "No", "Yes",
  "Was it motorized?", "", "Counts of Stops"
)

(p1 + p2 + p3) + 
  plot_layout(ncol = 3, nrow = 1, widths = c(1, 1), guides = "collect") +
  theme(legend.position = "bottom") 
```

------------------------------------------------------------------------

# Bivariate EDA

TODO: Introduction

Here forget about just analyzing the stops counts and we switch towards the main feature to see the proportion.

TODO: Comment everything

## Correlation - TODO

```{r correlation-heatmap}
```

TODO: Correlation Heatmap with all features

TODO: Correlation Heatmap with "main" features

## Collinearity

## Command Unit, Geospatial and Area related Features

### Area

```{r area-vs-target}
ratio_by_cat(
  vehstops, any_escalation, area,
  "Area", "Share of stops", "Escalation ratio by area",
  style = "stack"
)
```


```{r area-counts-stacked}
p1 <- vehstops %>%
  filter(!is.na(area), !is.na(year)) %>%
  count(area, year, name = "n") %>%
  group_by(area) %>% mutate(total = sum(n)) %>% ungroup() %>%
  mutate(area = fct_reorder(area, total),
         year = factor(year)) %>%
  ggplot(aes(x = n, y = area, fill = year)) +
  geom_col(width = 0.8) +
  labs(x = "Stops", y = NULL, fill = "Year",
       title = "Stops by area, split by year") +
  theme_minimal()

p2 <- vehstops %>%
  filter(!is.na(area), !is.na(month)) %>%
  count(area, month, name = "n") %>%
  group_by(area) %>% mutate(total = sum(n)) %>% ungroup() %>%
  mutate(area  = fct_reorder(area, total),
         month = factor(month, levels = 1:12, labels = month.abb)) %>%
  ggplot(aes(x = n, y = area, fill = month)) +
  geom_col(width = 0.8) +
  labs(x = "Stops", y = NULL, fill = "Month",
       title = "Stops by area, split by month") +
  theme_minimal()

p1 / p2
```


### Borough

```{r boro-vs-target}
ratio_by_cat(
  vehstops, any_escalation, boro,
  "Borough", "Share of stops", "Escalation ratio by borough",
  style = "stack"
)
```


```{r boro-counts-stacked}
p1 <- vehstops %>%
  filter(!is.na(boro), !is.na(year)) %>%
  count(boro, year, name = "n") %>%
  group_by(boro) %>% mutate(total = sum(n)) %>% ungroup() %>%
  mutate(boro = fct_reorder(boro, total),
         year = factor(year)) %>%
  ggplot(aes(x = n, y = boro, fill = year)) +
  geom_col(width = 0.8) +
  labs(x = "Stops", y = NULL, fill = "Year",
       title = "Stops by boro, split by year") +
  theme_minimal()

p2 <- vehstops %>%
  filter(!is.na(boro), !is.na(month)) %>%
  count(boro, month, name = "n") %>%
  group_by(boro) %>% mutate(total = sum(n)) %>% ungroup() %>%
  mutate(boro  = fct_reorder(boro, total),
         month = factor(month, levels = 1:12, labels = month.abb)) %>%
  ggplot(aes(x = n, y = boro, fill = month)) +
  geom_col(width = 0.8) +
  labs(x = "Stops", y = NULL, fill = "Month",
       title = "Stops by boro, split by month") +
  theme_minimal()

p1 / p2
```


### Command

```{r command-vs-target}
# FIXME
ratio_by_cat(
  vehstops, any_escalation, command,
  "Command", "Share of stops", "Escalation ratio by command",
  style = "stack"
)
```


```{r command-counts-stacked}
# FIXME
p1 <- vehstops %>%
  filter(!is.na(command), !is.na(year)) %>%
  count(command, year, name = "n") %>%
  group_by(command) %>% mutate(total = sum(n)) %>% ungroup() %>%
  mutate(command = fct_reorder(as.factor(command), total), year = factor(year)) %>%
  ggplot(aes(x = n, y = command, fill = year)) +
  geom_col(width = 0.8) +
  labs(x = "Stops", y = NULL, fill = "Year",
       title = "Stops by Command, split by year") +
  theme_minimal()

p2 <- vehstops %>%
  filter(!is.na(command), !is.na(month)) %>%
  count(command, month, name = "n") %>%
  group_by(command) %>% mutate(total = sum(n)) %>% ungroup() %>%
  mutate(command  = fct_reorder(as.factor(command), total),
         month = factor(month, levels = 1:12, labels = month.abb)) %>%
  ggplot(aes(x = n, y = command, fill = month)) +
  geom_col(width = 0.8) +
  labs(x = "Stops", y = NULL, fill = "Month",
       title = "Stops by Command, split by month") +
  theme_minimal()

p1 / p2
```

TODO: What to do with command? too many labels already grouped in boro and area

## Time related features

### Year

```{r year-vs-target}
ratio_by_cat(
  vehstops, any_escalation, year,
  "Year", "Share of stops", "Escalation ratio by year",
  style = "stack"
)
```

### Month

```{r month-vs-target}
vehstops |>
  filter(!is.na(any_escalation), !is.na(month), !is.na(year)) |>
  group_by(year, month) |>
  summarise(rate = mean(any_escalation), n = n(), .groups = "drop") |>
  ggplot(aes(month, rate, group = 1)) +
  geom_line(linewidth = 0.6) +
  geom_point(size = 0.8, alpha = 0.7) +
  scale_x_continuous(breaks = seq(0, 12, by = 1)) +
  scale_y_continuous(labels = percent_format()) +
  labs(title = "Escalation rate by month", x = "Month", y = "Rate") +
  facet_wrap(~ year, ncol = 1, scales = "free_x") +
  theme_minimal()
```

Let us contextualize how this changes throughout the years

```{r }
vehstops %>%
  mutate(date = ymd(paste(year, month, day, sep = "-"))) %>%
  group_by(year, month) %>%
  summarize(rate = mean(any_escalation == 1, na.rm = TRUE), n = n(), .groups = "drop") %>%
  ggplot(aes(month, rate, group = factor(year), linetype = factor(year))) +
  geom_line(linewidth = 0.6) + geom_point(aes(size = n)) +
  scale_y_continuous(labels = percent) +
  scale_x_continuous(breaks = 1:12) +
  labs(x = "Month", y = "Event rate", linetype = "Year", size = "Stops count",
       title = "Monthly escalation rate by year")
```

```{r}
vehstops %>%
  filter(year %in% c(2023, 2024, 2025)) %>%
  group_by(year, month) %>%
  summarize(stops = n(), .groups = "drop") %>%
  ggplot(aes(month, stops, group = factor(year), linetype = factor(year))) +
  geom_line(linewidth = 0.6) + geom_point(size = 1.5) +
  scale_x_continuous(breaks = 1:12) +
  labs(x = "Month", y = "Stops", linetype = "Year",
       title = "Monthly stops: 2023–2025")
```

### Day, Doy and Dow

```{r doy-vs-target}
vehstops |>
  filter(!is.na(any_escalation), !is.na(doy), !is.na(year)) |>
  group_by(year, doy) |>
  summarise(rate = mean(any_escalation), n = n(), .groups = "drop") |>
  ggplot(aes(doy, rate, group = 1)) +
  geom_line(linewidth = 0.6) +
  geom_point(size = 0.8, alpha = 0.7) +
  scale_x_continuous(breaks = seq(0, 366, by = 30)) +
  scale_y_continuous(labels = percent_format()) +
  labs(title = "Escalation rate by day-of-year", x = "Day of Year", y = "Rate") +
  facet_wrap(~ year, ncol = 1, scales = "free_x") +
  theme_minimal()
```

```{r day-vs-target}
vehstops |>
  filter(!is.na(any_escalation), !is.na(day), !is.na(year)) |>
  group_by(year, day) |>
  summarise(rate = mean(any_escalation), n = n(), .groups = "drop") |>
  ggplot(aes(day, rate, group = 1)) +
  geom_line(linewidth = 0.6) +
  geom_point(size = 0.8, alpha = 0.7) +
  scale_x_continuous(breaks = seq(0, 31, by = 1)) +
  scale_y_continuous(labels = percent_format()) +
  labs(title = "Escalation rate by day-of-month", x = "Day of Month", y = "Rate") +
  facet_wrap(~ year, ncol = 1, scales = "free_x") +
  theme_minimal()
```

```{r dow-vs-target}
vehstops |>
  filter(!is.na(any_escalation), !is.na(dow), !is.na(year)) |>
  group_by(year, dow) |>
  summarise(rate = mean(any_escalation), n = n(), .groups = "drop") |>
  ggplot(aes(dow, rate, group = 1)) +
  geom_line(linewidth = 0.6) +
  geom_point(size = 0.8, alpha = 0.7) +
  scale_x_continuous(breaks = seq(0, 7, by = 1)) +
  scale_y_continuous(labels = percent_format()) +
  labs(title = "Escalation rate by day-of-week", x = "Day of Week", y = "Rate") +
  facet_wrap(~ year, ncol = 1, scales = "free_x") +
  theme_minimal()
```

### Hour, Working Hour and Time Bracket

```{r hour-vs-target}
vehstops |>
  filter(!is.na(any_escalation), !is.na(hour), !is.na(year)) |>
  group_by(year, hour) |>
  summarise(rate = mean(any_escalation), n = n(), .groups = "drop") |>
  ggplot(aes(hour, rate, group = 1)) +
  geom_line(linewidth = 0.6) +
  geom_point(size = 0.8, alpha = 0.7) +
  scale_x_continuous(breaks = seq(0, 23, by = 1)) +
  scale_y_continuous(labels = percent_format()) +
  labs(title = "Escalation rate by Hour", x = "Hour", y = "Rate") +
  facet_wrap(~ year, ncol = 1, scales = "free_x") +
  theme_minimal()
```

```{r}
ratio_by_cat(
  vehstops, any_escalation, time_bracket,
  "Time bracket", "Share of stops", "Escalation ratio by time bracket",
  style = "stack"
)
```

```{r}
ratio_by_cat(
  vehstops, any_escalation, working_hour,
  "Working Hour", "Share of stops", "Escalation ratio by working hour",
  style = "stack"
)
```

### Weekend, Holiday and Event Day

```{r}
vehstops %>%
  filter(!is.na(holiday), !is.na(any_escalation)) %>%
  mutate(
    holiday = factor(holiday, levels = c(0, 1), labels = c("No holiday", "Holiday")),
    escal   = if_else(any_escalation %in% c(1, TRUE), "Escalated", "Not escalated")
  ) %>%
  count(holiday, escal, name = "n") %>%
  group_by(holiday) %>% mutate(total = sum(n)) %>% ungroup() %>%
  ggplot(aes(x = n, y = fct_reorder(holiday, total), fill = escal)) +
  geom_col(width = 0.8) +
  labs(x = "Stops", y = NULL, fill = "Outcome",
       title = "Stops by holiday (escalated vs not)")
```

## Motorist related features

### Reported Age

```{r age-by-target}
vehstops %>%
  filter(!is.na(age), between(age, 16, 100)) %>%
  mutate(bin = ntile(age, 20)) %>%
  group_by(bin) %>%
  summarize(
    age_mid = median(age, na.rm = TRUE),
    rate = mean(any_escalation == 1, na.rm = TRUE),
    n = n(), .groups = "drop"
  ) %>%
  mutate(
    se = sqrt(rate * (1 - rate) / n),
    lo = pmax(rate - 1.96 * se, 0),
    hi = pmin(rate + 1.96 * se, 1)
  ) %>%
  ggplot(aes(age_mid, rate)) +
  geom_point(size = 1.5) +
  geom_line(linewidth = 0.6) +
  geom_ribbon(aes(ymin = lo, ymax = hi), alpha = 0.15) +
  scale_y_continuous(labels = percent) +
  labs(x = "Age (bin median)", y = "Event rate", title = "Escalation rate vs age (deciles)")
```

```{r age-bracket-vs-target}
ratio_by_cat(
  vehstops, any_escalation, age_bracket,
  "Reported Age Bracket", "Share of stops", "Escalation ratio by reported age bracket",
  style = "stack"
)
```

### Reported Sex

```{r sex-vs-target}
ratio_by_cat(
  vehstops, any_escalation, sex,
  "Reported sex", "Share of stops", "Escalation ratio by reported sex",
  style = "stack"
)
```

### Reported Race

```{r race-vs-target}
ratio_by_cat(
  vehstops, any_escalation, race,
  "Reported race", "Share of stops", "Escalation ratio by reported race",
  style = "stack"
)
```

## Vehicle related features vs Target

```{r veh-categ-vs-target}
ratio_by_cat(
  vehstops, any_escalation, veh_categ,
  "Vehicle Category", "Share of stops", "Escalation ratio by vehicle category",
  style = "stack"
)
```

```{r}
vehstops %>%
  filter(!is.na(veh_categ), !is.na(any_escalation)) %>%
  mutate(escal = if_else(any_escalation %in% c(1, TRUE), "Escalated", "Not escalated")) %>%
  count(veh_categ, escal, name = "n") %>%
  group_by(veh_categ) %>% mutate(total = sum(n)) %>% ungroup() %>%
  mutate(veh_categ = fct_reorder(veh_categ, total)) %>%
  ggplot(aes(x = n, y = veh_categ, fill = escal)) +
  geom_col(width = 0.8) +
  labs(x = "Stops", y = NULL, fill = "Outcome",
       title = "Stops by vehicle category (escalated vs not)")
```


```{r veh-categ-clean-vs-target}
ratio_by_cat(
  vehstops, any_escalation, veh_categ_clean,
  "Vehicle Category (Clean)", "Share of stops", "Escalation ratio by vehicle category (clean)",
  style = "stack"
)
```

```{r}
vehstops %>%
  filter(!is.na(veh_categ_clean), !is.na(any_escalation)) %>%
  mutate(escal = if_else(any_escalation %in% c(1, TRUE), "Escalated", "Not escalated")) %>%
  count(veh_categ_clean, escal, name = "n") %>%
  group_by(veh_categ_clean) %>% mutate(total = sum(n)) %>% ungroup() %>%
  mutate(veh_categ_clean = fct_reorder(veh_categ_clean, total)) %>%
  ggplot(aes(x = n, y = veh_categ_clean, fill = escal)) +
  geom_col(width = 0.8) +
  labs(x = "Stops", y = NULL, fill = "Outcome",
       title = "Stops by cleaned vehicle category (escalated vs not)")
```


```{r}
ratio_by_cat(
  vehstops, any_escalation, veh_4_wheels,
  "4 Wheels", "Share of stops", "Escalation ratio by number of wheels",
  style = "stack"
)
```

```{r}
vehstops %>%
  filter(!is.na(veh_4_wheels), !is.na(any_escalation)) %>%
  mutate(
    four_wheels = case_when(veh_4_wheels %in% c(1, TRUE) ~ "Yes",
                            veh_4_wheels %in% c(0, FALSE) ~ "No",
                            TRUE ~ NA_character_),
    escal = if_else(any_escalation %in% c(1, TRUE), "Escalated", "Not escalated")
  ) %>% filter(!is.na(four_wheels)) %>%
  count(four_wheels, escal, name = "n") %>%
  group_by(four_wheels) %>% mutate(total = sum(n)) %>% ungroup() %>%
  mutate(four_wheels = fct_reorder(four_wheels, total)) %>%
  ggplot(aes(x = n, y = four_wheels, fill = escal)) +
  geom_col(width = 0.8) +
  labs(x = "Stops", y = "4 wheels", fill = "Outcome",
       title = "Stops by 4-wheel indicator (escalated vs not)")
```


```{r}
ratio_by_cat(
  vehstops, any_escalation, veh_private,
  "Private?", "Share of stops", "Escalation ratio by ownership",
  style = "stack"
)
```

```{r}
vehstops %>%
  filter(!is.na(veh_private), !is.na(any_escalation)) %>%
  mutate(
    private = case_when(veh_private %in% c(1, TRUE) ~ "Private",
                        veh_private %in% c(0, FALSE) ~ "Non-private",
                        TRUE ~ NA_character_),
    escal   = if_else(any_escalation %in% c(1, TRUE), "Escalated", "Not escalated")
  ) %>% filter(!is.na(private)) %>%
  count(private, escal, name = "n") %>%
  group_by(private) %>% mutate(total = sum(n)) %>% ungroup() %>%
  mutate(private = fct_reorder(private, total)) %>%
  ggplot(aes(x = n, y = private, fill = escal)) +
  geom_col(width = 0.8) +
  labs(x = "Stops", y = NULL, fill = "Outcome",
       title = "Stops by private vs non-private (escalated vs not)")

```


```{r}
ratio_by_cat(
  vehstops, any_escalation, veh_motorized,
  "Motorized?", "Share of stops", "Escalation ratio by motorization",
  style = "stack"
)
```

```{r}
vehstops %>%
  filter(!is.na(veh_motorized), !is.na(any_escalation)) %>%
  mutate(
    motorized = case_when(veh_motorized %in% c(1, TRUE) ~ "Motorized",
                          veh_motorized %in% c(0, FALSE) ~ "Non-motorized",
                          TRUE ~ NA_character_),
    escal     = if_else(any_escalation %in% c(1, TRUE), "Escalated", "Not escalated")
  ) %>% filter(!is.na(motorized)) %>%
  count(motorized, escal, name = "n") %>%
  group_by(motorized) %>% mutate(total = sum(n)) %>% ungroup() %>%
  mutate(motorized = fct_reorder(motorized, total)) %>%
  ggplot(aes(x = n, y = motorized, fill = escal)) +
  geom_col(width = 0.8) +
  labs(x = "Stops", y = NULL, fill = "Outcome",
       title = "Stops by motorization (escalated vs not)")

```

------------------------------------------------------------------------

# Multivariate EDA

Since multivariate analysis is very complex, and we just want to get rid of some extra features, we are going to look at some of the percentages, with respect to escalation, of multiple variables (ex: sex x race)

## Multicollinearity (binary block)

## Two-way event-rate heatmaps (interactions)

```{r two-way-heatmaps}
boro_order <- vehstops %>%
  group_by(boro) %>%
  summarize(rate = mean(any_escalation == 1, na.rm = TRUE), .groups = "drop") %>%
  arrange(rate) %>%
  pull(boro)

vehstops %>%
  filter(!is.na(boro)) %>%
  group_by(boro, hour) %>%
  summarize(rate = mean(any_escalation == 1, na.rm = TRUE), n = n(), .groups = "drop") %>%
  mutate(boro = factor(boro, levels = boro_order)) %>%
  ggplot(aes(hour, boro, fill = rate)) +
  geom_tile(width = 1, height = 1) +
  scale_fill_viridis_c(labels = scales::percent) +
  labs(x = "Hour", y = "Borough", fill = "Rate",
       title = "Escalation rate by borough × hour")

# 2) Sex × Race
vehstops %>%
  filter(!is.na(sex), !is.na(race)) %>%
  group_by(sex, race) %>%
  summarize(rate = mean(any_escalation == 1, na.rm = TRUE), n = n(), .groups = "drop") %>%
  mutate(race = fct_lump_n(race, n = 8, other_level = "Other")) %>%
  ggplot(aes(race, sex, fill = rate)) +
  geom_tile(width = 1, height = 1) +
  scale_fill_viridis_c(labels = scales::percent) +
  labs(x = "Race (lumped)", y = "Sex", fill = "Rate",
       title = "Escalation rate across sex × race")
```

### Space × Time interaction (gridded)

```{r space-time}
vehstops %>%
  filter(!is.na(latitude), !is.na(longitude),
         between(latitude, 40.48, 40.92), between(longitude, -74.27, -73.68)) %>%
  mutate(hour_bin = cut(hour, breaks = c(-1,5,11,17,23), labels = c("Night","Morning","Afternoon","Evening"))) %>%
  group_by(hour_bin) %>%
  sample_n(size = min(n(), 200000), replace = FALSE) %>%  # cap points per facet for speed
  ggplot(aes(longitude, latitude)) +
  stat_summary_2d(aes(z = any_escalation), fun = ~ mean(.x == 1, na.rm = TRUE), bins = 50) +
  facet_wrap(~ hour_bin) +
  coord_equal() +
  scale_fill_viridis_c(labels = percent) +
  labs(x = "Longitude", y = "Latitude", fill = "Rate",
       title = "Escalation rate across space by time-of-day")
```

## Enforcement and Escalation features vs Target

## Command Unit, Geospatial and Area related Features vs Target

Per command unit amount of stops during time (per year, per month, per day)

Per command unit escalation rate during time (per year, per month, per day)

### Area

```{r}
ratio_by_cat_split(
  DF = vehstops,
  TARGET = any_escalation,
  CAT    = area,
  SPLIT  = weekend,
  x_title = "Area",
  y_title = "Share of stops",
  title   = "Escalation ratio by area (split by weekend)",
  target_labels = c("Not escalated","Escalated"),
  split_labels  = c("Weekday","Weekend")
)

```

```{r area-x-hour}
area_order <- vehstops %>%
  group_by(area) %>%
  summarize(rate = mean(any_escalation == 1, na.rm = TRUE), .groups = "drop") %>%
  arrange(rate) %>%
  pull(area)

vehstops %>%
  filter(!is.na(area)) %>%
  group_by(area, hour) %>%
  summarize(rate = mean(any_escalation == 1, na.rm = TRUE), n = n(), .groups = "drop") %>%
  mutate(area = factor(area, levels = area_order)) %>%
  ggplot(aes(hour, area, fill = rate)) +
  geom_tile(width = 1, height = 1) +
  scale_fill_viridis_c(labels = scales::percent) +
  labs(x = "Hour", y = "Area", fill = "Rate",
       title = "Escalation rate by area × hour")
```

```{r area-x-workinghour}
ratio_by_cat_split(
  DF = vehstops,
  TARGET = any_escalation,
  CAT    = area,
  SPLIT  = working_hour,
  x_title = "Area",
  y_title = "Share of stops",
  title   = "Escalation ratio by area (split by working hour)",
  target_labels = c("Not escalated","Escalated"),
  split_labels  = c("Non Working Hour","Working Hour")
)

```

```{r area-x-sex}
ratio_by_cat_split(
  DF = vehstops,
  TARGET = any_escalation,
  CAT    = area,
  SPLIT  = sex,
  x_title = "Area",
  y_title = "Share of stops",
  title   = "Escalation ratio by area (split by reported sex)",
  target_labels = c("Not escalated","Escalated"),
  split_labels  = c("No","Yes")
)
```

```{r}
heat_escalation_catcat(
  vehstops, Y = boro, X = time_bracket,
  X_TITLE = "Time bracket", Y_TITLE = "Borough",
  TITLE = "Escalation rate by borough × time bracket",
  min_n = 30, order_y_by = "rate", order_x_by = "freq", alpha_by_n = TRUE
)
```

```{r}
area_order <- vehstops %>%
  group_by(area) %>%
  summarize(rate = mean(any_escalation == 1, na.rm = TRUE), .groups = "drop") %>%
  arrange(rate) %>%
  pull(area)

vehstops %>%
  filter(!is.na(area)) %>%
  group_by(area, month) %>%
  summarize(rate = mean(any_escalation == 1, na.rm = TRUE), n = n(), .groups = "drop") %>%
  mutate(area = factor(area, levels = area_order)) %>%
  ggplot(aes(month, area, fill = rate)) +
  geom_tile(width = 1, height = 1) +
  scale_fill_viridis_c(labels = scales::percent) +
  labs(x = "Month", y = "Area", fill = "Rate",
       title = "Escalation rate by area × month")
```

Per area amount of stops during time (per year, per month, per day)

Per area escalation rate during time (per year, per month, per day)

```{r area-by-weekend-holiday-event-vs-target}
p1 <- ratio_by_cat_split(
  DF = vehstops,
  TARGET = any_escalation,
  CAT    = area,
  SPLIT  = weekend,
  x_title = "Time bracket",
  y_title = "Share of stops",
  title   = "Escalation ratio by area (split by weekend)",
  target_labels = c("Not escalated","Escalated"),
  split_labels  = c("Weekday","Weekend")
)

p2 <- ratio_by_cat_split(
  DF = vehstops,
  TARGET = any_escalation,
  CAT    = area,
  SPLIT  = holiday,
  x_title = "Time bracket",
  y_title = "Share of stops",
  title   = "Escalation ratio by area (split by holiday)",
  target_labels = c("Not escalated","Escalated"),
  split_labels  = c("Regular Day","Holiday")
)

p3 <- ratio_by_cat_split(
  DF = vehstops,
  TARGET = any_escalation,
  CAT    = area,
  SPLIT  = event_day,
  x_title = "Time bracket",
  y_title = "Share of stops",
  title   = "Escalation ratio by area (split by event_day)",
  target_labels = c("Not escalated","Escalated"),
  split_labels  = c("Regular Day","Event Day")
)

p1
p2
p3
```

### Borough

```{r boro-x-year-escalation}
boro_order <- vehstops %>%
  group_by(boro) %>%
  summarize(rate = mean(any_escalation == 1, na.rm = TRUE), .groups = "drop") %>%
  arrange(rate) %>%
  pull(boro)

vehstops %>%
  filter(!is.na(boro)) %>%
  group_by(boro, year) %>%
  summarize(rate = mean(any_escalation == 1, na.rm = TRUE), n = n(), .groups = "drop") %>%
  mutate(boro = factor(boro, levels = boro_order)) %>%
  ggplot(aes(year, boro, fill = rate)) +
  geom_tile(width = 1, height = 1) +
  scale_fill_viridis_c(labels = scales::percent) +
  labs(x = "Year", y = "Borough", fill = "Rate",
       title = "Escalation rate by borough × year")


```

```{r boro-x-month-escalation}
boro_order <- vehstops %>%
  group_by(boro) %>%
  summarize(rate = mean(any_escalation == 1, na.rm = TRUE), .groups = "drop") %>%
  arrange(rate) %>%
  pull(boro)

vehstops %>%
  filter(!is.na(boro)) %>%
  group_by(boro, month) %>%
  summarize(rate = mean(any_escalation == 1, na.rm = TRUE), n = n(), .groups = "drop") %>%
  mutate(boro = factor(boro, levels = boro_order)) %>%
  ggplot(aes(month, boro, fill = rate)) +
  geom_tile(width = 1, height = 1) +
  scale_fill_viridis_c(labels = scales::percent) +
  labs(x = "Month", y = "Borough", fill = "Rate",
       title = "Escalation rate by borough × month")
```

```{r boro-x-hour-escalation}
boro_order <- vehstops %>%
  group_by(boro) %>%
  summarize(rate = mean(any_escalation == 1, na.rm = TRUE), .groups = "drop") %>%
  arrange(rate) %>%
  pull(boro)

vehstops %>%
  filter(!is.na(boro)) %>%
  group_by(boro, hour) %>%
  summarize(rate = mean(any_escalation == 1, na.rm = TRUE), n = n(), .groups = "drop") %>%
  mutate(boro = factor(boro, levels = boro_order)) %>%
  ggplot(aes(hour, boro, fill = rate)) +
  geom_tile(width = 1, height = 1) +
  scale_fill_viridis_c(labels = scales::percent) +
  labs(x = "Hour", y = "Borough", fill = "Rate",
       title = "Escalation rate by borough × hour")
```

### Command Unit

Per command unit amount of stops during time (per year, per month, per day)

Per command unit escalation rate during time (per year, per month, per day)

```{r}
# Borough × Time bracket (both categorical), hide very sparse cells
p1 <- heat_escalation_catcat(
  vehstops, Y = boro, X = time_bracket,
  X_TITLE = "Time bracket", Y_TITLE = "Borough",
  TITLE = "Escalation rate by borough × time bracket",
  min_n = 30, order_y_by = "rate", order_x_by = "freq", alpha_by_n = TRUE
)

# Cleaned vehicle category × Weekend (0/1 already labeled)
p2 <- heat_escalation_catcat(
  vehstops, Y = race, X = sex,
  "Sex", "Race",
  "Rate by race × sex",
  min_n = 50, order_y_by = "freq", order_x_by = "freq"
)

# Cleaned vehicle category × Weekend (0/1 already labeled)
p3 <- heat_escalation_catcat(
  vehstops, Y = veh_categ_clean, X = weekend,
  "Weekend (0/1)", "Vehicle (cleaned)",
  "Rate by vehicle × weekend",
  min_n = 50, order_y_by = "freq", order_x_by = "none"
)


p1
p2
p3
```

```{r}
heat_escalation_catcat(
  vehstops, Y = race, X = sex,
  "Sex", "Race",
  "Rate by race × sex",
  min_n = 50, order_y_by = "freq", order_x_by = "freq"
)
```

Unknown and Unknown matter somehow!!!

```{r}
# Borough × hour (hide cells with < 50 stops)
p_boro_hour <- heat_escalation_yx(
  DF = vehstops, Y = boro, X = hour,
  X_TITLE = "Hour of day", Y_TITLE = "Borough",
  TITLE = "Escalation rate by borough × hour",
  min_n = 50
)

# Area × day-of-week (X is numeric 1–7)
p_area_dow <- heat_escalation_yx(
  vehstops, Y = area, X = dow,
  "Day of week (1–7)", "Area",
  "Escalation rate by area × DOW", min_n = 25
)

# If X is not integer-spaced (e.g., minute) treat as factor to make equal-width tiles
p_boro_min <- heat_escalation_yx(
  vehstops, Y = boro, X = minute,
  "Minute", "Borough", "Rate by borough × minute",
  x_as_factor = TRUE
)

```

```{r}
p_boro_hour
p_area_dow
p_boro_min
```

## Time related features

### Hour

#### Grouped by Month

```{r month-lines}
# TODO: use? Move to multivariate
vehstops %>%
  filter(year %in% c(2023, 2024, 2025)) %>%
  group_by(year, month) %>%
  summarize(stops = n(), .groups = "drop") %>%
  ggplot(aes(month, stops, group = factor(year), linetype = factor(year))) +
  geom_line(linewidth = 0.6) + geom_point(size = 1.5) +
  scale_x_continuous(breaks = 1:12) +
  labs(x = "Month", y = "Stops", linetype = "Year",
       title = "Monthly stops: 2023–2025")
```

```{r hour-vs-stops-by-month}
vehstops %>%
  filter(month %in% c(1,2,3,4,5,6,7,8,9,10,11,12)) %>%
  group_by(month, hour) %>%
  summarize(stops = n(), .groups = "drop") %>%
  ggplot(aes(hour, stops, group = factor(month), linetype = factor(month))) +
  geom_line(linewidth = 0.6) + geom_point(size = 1.5) +
  scale_x_continuous(breaks = 0:23) +
  labs(x = "Hour", y = "Stops", linetype = "Month",
       title = "Count of Hourly Stops by Month")
```

```{r hour-vs-dow-by-month}
vehstops %>%
  group_by(dow, hour) %>%
  summarize(stops = n(), .groups = "drop") %>%
  ggplot(aes(hour, stops, group = factor(dow), linetype = factor(dow))) +
  geom_line(linewidth = 0.6) + geom_point(size = 1.5) +
  scale_x_continuous(breaks = 0:23) +
  labs(x = "Hour", y = "Stops", linetype = "Day of the Week",
       title = "Count of Hourly Stops by Day of the Week")
```


```{r hour-vs-target-by-year}
vehstops %>%
  mutate(date = ymd(paste(year, month, day, sep = "-"))) %>%
  group_by(month, hour) %>%
  summarize(rate = mean(any_escalation == 1, na.rm = TRUE), n = n(), .groups = "drop") %>%
  ggplot(aes(hour, rate, group = factor(month), linetype = factor(month))) +
  geom_line(linewidth = 0.6) + geom_point(aes(size = n)) +
  scale_y_continuous(labels = percent) +
  scale_x_continuous(breaks = 0:23) +
  labs(x = "Hour", y = "Event rate", linetype = "Hour", size = "Stops count",
       title = "Hourly Escalation Rate by Month")
```

```{r dow-hour-heatmap}
vehstops %>%
  filter(!is.na(dow)) %>%
  group_by(dow, hour) %>%
  summarize(rate = mean(any_escalation == 1, na.rm = TRUE), n = n(), .groups = "drop") %>%
  ggplot(aes(hour, dow, fill = rate)) +
  geom_tile(width = 1, height = 1) +
  scale_fill_viridis_c(labels = percent) +
  scale_x_continuous(breaks = seq(0, 23, 1)) +
  scale_y_continuous(breaks = seq(0, 7, 1)) +
  labs(x = "Hour", y = "Day of week", fill = "Event rate",
       title = "Escalation rate by day and hour")
```

#### Grouped by Year

```{r hour-vs-stops-by-year}
vehstops %>%
  filter(year %in% c(2023, 2024, 2025)) %>%
  group_by(year, hour) %>%
  summarize(stops = n(), .groups = "drop") %>%
  ggplot(aes(hour, stops, group = factor(year), linetype = factor(year))) +
  geom_line(linewidth = 0.6) + geom_point(size = 1.5) +
  scale_x_continuous(breaks = 0:23) +
  labs(x = "Hour", y = "Stops", linetype = "Year",
       title = "Count of Hourly stops by Year")
```

```{r hour-vs-any-escalation-by-year}
vehstops %>%
  mutate(date = ymd(paste(year, month, day, sep = "-"))) %>%
  group_by(year, hour) %>%
  summarize(rate = mean(any_escalation == 1, na.rm = TRUE), n = n(), .groups = "drop") %>%
  ggplot(aes(hour, rate, group = factor(year), linetype = factor(year))) +
  geom_line(linewidth = 0.6) + geom_point(aes(size = n)) +
  scale_y_continuous(labels = percent) +
  scale_x_continuous(breaks = 0:23) +
  labs(x = "Hour", y = "Event rate", linetype = "Year", size = "Stops count",
       title = "Hourly Escalation Rate by Year")
```

#### Grouped by Day

```{r}
vehstops %>%
  filter(!is.na(dow)) %>%
  group_by(dow, hour) %>%
  summarize(rate = mean(any_escalation == 1, na.rm = TRUE), n = n(), .groups = "drop") %>%
  ggplot(aes(hour, dow, fill = rate)) +
  geom_tile(width = 1, height = 1) +
  scale_fill_viridis_c(labels = percent) +
  scale_x_continuous(breaks = seq(0, 23, 1)) +
  scale_y_continuous(breaks = seq(0, 7, 1)) +
  labs(x = "Hour", y = "Day of week", fill = "Event rate",
       title = "Escalation rate by day and hour")
```

### Day

```{r target-rate-over-time, message=FALSE}
vehstop_daily <- vehstops %>%
  mutate(date = ymd(paste(year, month, day, sep = "-"))) %>%
  group_by(date) %>%
  summarize(rate = mean(any_escalation == 1, na.rm = TRUE), n = n(), .groups = "drop")

ggplot(vehstop_daily, aes(date, rate)) +
  geom_line(linewidth = 0.6) +
  geom_point(aes(size = n), alpha = 0.6) +
  geom_smooth(se = FALSE, method = "loess", span = 0.2) +
  scale_y_continuous(labels = percent) +
  labs(x = NULL, y = "Event rate", size = "N/day", title = "Daily escalation rate over time")
```

## Motorist related features

### Reported Age

```{r}
ratio_by_cat_split(
  DF = vehstops,
  TARGET = any_escalation,
  CAT    = age_bracket,
  SPLIT  = weekend,
  x_title = "Reported Age bracket",
  y_title = "Share of stops",
  title   = "Escalation ratio by reported age (split by weekend)",
  target_labels = c("Not escalated","Escalated"),
  split_labels  = c("Weekday","Weekend")
)
```

### Reported Sex

### Reported Race


## Vehicle related features

```{r veh-categ-x-age}
heat_escalation_catcat(
  vehstops, Y = veh_categ, X = age_bracket,
  "Reported Age Bracket", "Vehicle Category",
  "Rate by vehicle category x reported age bracket",
  min_n = 1, order_y_by = "freq", order_x_by = "freq"
)
```

```{r veh-categ-x-sex}
heat_escalation_catcat(
  vehstops, Y = veh_categ, X = sex,
  "Reported Sex", "Vehicle Category",
  "Rate by vehicle category x reported sex",
  min_n = 1, order_y_by = "freq", order_x_by = "freq"
)
```

```{r veh-categ-x-race}
# FIXME
heat_escalation_catcat(
  vehstops, Y = veh_categ, X = race,
  "Reported Race", "Vehicle Category",
  "Rate by vehicle category x reported race",
  min_n = 1, order_y_by = "freq", order_x_by = "freq"
)
```


------------------------------------------------------------------------

# Feature Encoding


---

# Early Feature Selection and Feature Dropping


TODO: 1 check what to keep between area, borough, command and latitude+longitude
TODO: 2 check what to keep between dow or weekday, 
TODO: 3 check whether to keep month or day

Test if month, day matter or just keep hour.

We can safely remove day of the month `dow` since it does not seem to give us a

```{r}
# Just an intuitive feature dropping for later comparison
# vehstops2 <- vehstops2 %>%select( -month, -day, -weekday, -holiday, -event_day, -minute, -latitude, -longitude, -boro, -command)
```

```{r}
```
