---
title: "Predicting Excessive Enforcement Patterns in Police Behavior After Vehicles Stops in New York"
subtitle: "EDA and Feature Engineering - 2 Wheels Vehicles"
author: "Gianmaria Pizzo"
output:
  html_document:
    fig_caption: yes
    theme: flatly
    highlight: pygments
    code_folding: show
    kee_md: yes
    toc: yes
    toc_depth: 4
    number_sections: no
    toc_float:
      smooth_scroll: no
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

These `.Rmd` files represent the project submission for the course [Statistical Inference and Learning](https://www.unive.it/data/course/521944) by Professor [Cristiano Varin](https://www.unive.it/data/people/5592728) at [Ca' Foscari University of Venice](https://www.unive.it/) for the year 2025/2026.

# Problem Statement - TODO

## Goal of this rmarkdown file - TODO

We explore the features in 3 different ways. 

1. First, we separately analyze them and try to grasp their meaning, their distribution and the possible outliers or other problematic aspects. This univariate analysis, although complicated by the presence of many categorical predictors, should give us an initial idea of how the data is distributed.
2. Then, we compare the possible predictors with respect to the target feature, and try to understand if there is any notable relationship between the two of them. Also, we try to see whether there is evidence of correlation, collinearity, or other problematic behaviors they may introduce. 
3. Finally we explore the subsets of features that may be related (ex: geospatial features and command unit, or time of the day and some demographics), to find out whether some of them can be removed or replaced. Plus we try to investigate about multicollinearity and other complications.

The EDA is the first step to understanding the data and its nature.

This includes analyzing the single features (Univariate Analysis), their interactions with the target (Bivariate Analysis) and their interactions with other features (Multivariate Analysis and Collinearity Check). However, the latter might be hard to compute as it requires the estimation of a regression model that follows certain assumptions and it is usually easier to check this when using linear models.

During my exploration I will look out for possible outliers, correlation and any other statistically relevant problem that might affect my final prediction.

We will start with the environment.

Note that, from now on, I will shift adopt a more **formal or inclusive tone** (e.g., the "editorial we").

------------------------------------------------------------------------

# Environment

The first to do, is to build a reproducible environment for our project.

```{r}
requirements=c( 
  "remotes", "summarytools", "forcats", "scales", "purrr",
  "stringr", "plyr", "dbplyr", "dplyr", "dtplyr", "tidyr", "tidyselect",
  "timeDate", "timechange", "tzdb", "hms", "readr",
  "lubridate", "prettyunits", "RColorBrewer", "viridis", "rcartocolor", "ggbeeswarm", "emmeans",
  "sf", "plotly", "ggplot2", "ggmap", "ggdensity", "UpSetR", "patchwork"
)
```


```{r install-requirements, include=FALSE}
for (req in requirements){
  if (!require(req, character.only = TRUE)){
      install.packages(req)
  }
}
```

Note that some of the packages may not be used, but to avoid any error related to dependencies, we are going to keep all of these libraries. Now, the loading of all packages should be immediate.

```{r load-requirements, include=FALSE}
for (req in requirements){
  if (!require(req, character.only = TRUE)){
      library(req)
  }
}
```

```{r remove-vars, include=FALSE}
rm(requirements, req)
```

## Plot Functions 

Here we add the plot functions to make some kinds of visualizations reproducible.

```{r univariate-plot-helpers}
# ==============================================================================
# 1. Plots for Univariate Analysis
# ==============================================================================

#' @title Plot Binary Bar Chart
#' 
#' @description Creates a simple bar chart counting the occurrences of a binary feature.
#' It automatically handles different representations of binary values (1/0, TRUE/FALSE).
#' 
#' @param DF The input dataframe.
#' @param FEATURE The binary column to plot (unquoted).
#' @param OPTION_0 The text label for the "0" or "FALSE" category.
#' @param OPTION_1 The text label for the "1" or "TRUE" category.
#' @param X_LABEL Label for the X axis.
#' @param Y_LABEL Label for the Y axis.
#' @param TITLE Main title of the plot.
#' @param pad Numeric padding for the top of the Y axis to fit text labels.
#' @param text_vjust Vertical adjustment for the text labels on top of bars.
#' 
#' @return A ggplot2 object.
plot_binary_bar <- function(DF, FEATURE, OPTION_0, OPTION_1, X_LABEL, Y_LABEL, TITLE, pad = 0.12, text_vjust = -0.35) {
  # Standardize the binary feature
  df_sum <- DF %>%
    mutate(
      .raw  = {{ FEATURE }},
      # Convert into standard 1L or 0L integers
      .feat = dplyr::case_when(
        as.character(.raw) %in% c("1","TRUE","true")  ~ 1L,
        as.character(.raw) %in% c("0","FALSE","false") ~ 0L,
        TRUE ~ NA_integer_
      )
    ) %>%
    # Remove missing values
    filter(!is.na(.feat)) %>%
    # Count of 0s and 1s
    count(.feat, name = "n") %>%
    # Create factor variable with custom labels for plotting 
    mutate(.lab = factor(.feat, levels = c(0, 1), labels = c(OPTION_0, OPTION_1)))
  
  # Get the maximum bar height
  ymax <- max(df_sum$n, na.rm = TRUE)
  
  # Build the ggplot
  ggplot(df_sum, aes(x = .lab, y = n, fill = .lab)) +
    geom_col(width = 0.8, show.legend = FALSE) +
    # Add count numbers on top of the bars
    geom_text(aes(label = scales::comma(n)), vjust = text_vjust) +
    # Expand the Y axis slightly so the text labels are not cut off
    scale_y_continuous(limits = c(0, ymax * (1 + pad)),
                       expand = expansion(mult = c(0, 0.02))) +
    coord_cartesian(clip = "off") +
    theme(plot.margin = margin(t = 10, r = 5.5, b = 5.5, l = 5.5)) +
    labs(x = X_LABEL, y = Y_LABEL, title = TITLE)
}

# FIXME: Document and comment better
#' @title Plot Univariate Bar Chart (Categorical)
#' 
#' @description Creates a bar chart counting the occurrences of a categorical feature.
#' Allows dynamic ordering (by frequency or original/alphabetical order).
#' 
#' @param DF The input dataframe.
#' @param FEATURE The categorical column to analyze (unquoted).
#' @param X_LABEL Label for the X axis.
#' @param Y_LABEL Label for the Y axis (default is "Count").
#' @param TITLE Main title of the plot.
#' @param top_n Optional integer; if set, keeps only the top N most frequent categories.
#' @param order_by Method to order the X axis: "freq" (descending count) or "none" (alphabetical/original).
#' @param pad Numeric padding for the top of the Y axis to fit text labels.
#' @param text_vjust Vertical adjustment for the text labels on top of bars.
#' 
#' @return A ggplot2 object.
plot_univariate_cat <- function(DF, FEATURE, X_LABEL, Y_LABEL = "Count", TITLE, 
                                top_n = NULL, order_by = c("freq", "none"), 
                                pad = 0.12, text_vjust = -0.35) {
  
  order_by <- match.arg(order_by)
  
  # Step 1: Calcola i conteggi
  df_sum <- DF %>%
    filter(!is.na({{ FEATURE }})) %>%
    count({{ FEATURE }}, name = "n")
  
  # Step 2: Se richiesto 'top_n', mantieni solo i più frequenti
  if (!is.null(top_n)) {
    df_sum <- df_sum %>% slice_max(order_by = n, n = top_n)
  }
  
  # Step 3: Determina l'ordine per l'asse X
  cat_levels <- switch(
    order_by,
    freq = df_sum %>% arrange(desc(n)) %>% pull({{ FEATURE }}),
    none = sort(unique(df_sum %>% pull({{ FEATURE }}))) # Ordine alfabetico / originale
  )
  
  # Applica l'ordine scelto
  df_sum <- df_sum %>%
    mutate(.feat_f = factor(as.character({{ FEATURE }}), levels = as.character(cat_levels)))
  
  # Step 4: Calcola l'altezza massima per il padding dinamico
  ymax <- max(df_sum$n, na.rm = TRUE)

  # Step 5: Costruisci il grafico
  ggplot(df_sum, aes(x = .feat_f, y = n, fill = .feat_f)) +
    geom_col(width = 0.8, show.legend = FALSE) +
    geom_text(aes(label = scales::comma(n)), vjust = text_vjust, size = 3) +
    scale_y_continuous(limits = c(0, ymax * (1 + pad)),
                       expand = expansion(mult = c(0, 0.02))) +
    coord_cartesian(clip = "off") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
          panel.grid.major.x = element_blank(),
          plot.margin = margin(t = 10, r = 5.5, b = 5.5, l = 5.5)) +
    scale_fill_viridis_d(option = "mako", begin = 0.3, end = 0.8) +
    labs(x = X_LABEL, y = Y_LABEL, title = TITLE)
}


#' @title Plot Univariate Histogram
#' 
#' @description Creates a histogram to show the distribution of a continuous 
#' numerical variable.
#' 
#' @param DF The input dataframe.
#' @param FEATURE The numerical column to analyze (unquoted).
#' @param X_LABEL Label for the X axis.
#' @param TITLE Main title of the plot.
#' @param bin_width The width of each bin in the histogram (e.g., 5 for 5-year age groups).
#' 
#' @return A ggplot2 object.
plot_univariate_hist <- function(DF, FEATURE, X_LABEL, TITLE, bin_width = 5) {
  
  ggplot(DF, aes(x = {{ FEATURE }})) +
    # Create the histogram with a specific bin width
    geom_histogram(binwidth = bin_width, fill = "seagreen", color = "white", alpha = 0.8) +
    scale_y_continuous(labels = scales::comma) +
    labs(x = X_LABEL, y = "Frequenza", title = TITLE) +
    theme_minimal() +
    theme(panel.grid.minor = element_blank())
}
```


```{r biviariate-plot-helpers}
# ==============================================================================
# 2. Plots for Bivariate Analysis
# ==============================================================================

# ==============================================================================
# Helper Function: Calculate Cramer's V
# ==============================================================================

#' @title Calculate Cramer's V
#' 
#' @description Calculates the Cramer's V measure of association between two 
#' categorical vectors. Returns a value between 0 (no association) and 1 (perfect association).
#' 
#' @param x First categorical vector.
#' @param y Second categorical vector.
#' 
#' @return A numeric value representing Cramer's V.
cramer_v <- function(x, y) {
  # Crea una tabella di contingenza (conteggi incrociati)
  tbl <- table(x, y)
  
  # Esegue il test Chi-Quadro
  chi2 <- suppressWarnings(chisq.test(tbl, correct = FALSE)$statistic)
  
  # Calcola la dimensione totale e il valore minimo per la formula
  n <- sum(tbl)
  min_dim <- min(nrow(tbl) - 1, ncol(tbl) - 1)
  
  # Evita errori matematici se c'è una sola categoria
  if(min_dim == 0) return(0) 
  
  # Formula di Cramer's V
  v <- sqrt(chi2 / (n * min_dim))
  return(as.numeric(v))
}

# ==============================================================================
# Categorical Correlation Heatmap
# ==============================================================================

# FIXME: Fix doc and comments

#' @title Plot Categorical Correlation Matrix (Cramer's V)
#' 
#' @description Creates a correlation-like heatmap for multiple categorical variables
#' using Cramer's V. Extremely useful for identifying redundant categorical features.
#' 
#' @param DF The input dataframe.
#' @param FEATURES A character vector containing the names of the categorical columns to compare.
#' @param TITLE Main title of the plot.
#' 
#' @return A ggplot2 object.
plot_categorical_correlation <- function(DF, FEATURES, TITLE = "Matrice di Associazione (Cramer's V)") {
  
  # Step 1: Seleziona solo le colonne richieste e rimuove i valori nulli per il test
  df_sub <- DF %>% 
    select(all_of(FEATURES)) %>%
    drop_na()
  
  n_vars <- ncol(df_sub)
  
  # Step 2: Inizializza una matrice vuota riempita di 1 (la diagonale sarà 1)
  mat <- matrix(1, nrow = n_vars, ncol = n_vars)
  rownames(mat) <- colnames(df_sub)
  colnames(mat) <- colnames(df_sub)
  
  # Step 3: Calcola il Cramer's V per ogni possibile coppia di variabili
  for(i in 1:(n_vars - 1)) {
    for(j in (i + 1):n_vars) {
      v <- cramer_v(df_sub[[i]], df_sub[[j]])
      # Riempie la matrice in modo simmetrico
      mat[i, j] <- v
      mat[j, i] <- v
    }
  }
  
  # Step 4: Trasforma la matrice in un formato "lungo" compatibile con ggplot2
  mat_long <- as.data.frame(as.table(mat))
  colnames(mat_long) <- c("Var1", "Var2", "CramerV")
  
  # Step 5: Crea la Heatmap
  ggplot(mat_long, aes(x = Var1, y = Var2, fill = CramerV)) +
    geom_tile(color = "white", width = 1, height = 1) +
    # Aggiunge il valore numerico arrotondato a due decimali dentro i quadrati
    geom_text(aes(label = round(CramerV, 2)), size = 4) +
    # Colori: da grigio chiarissimo (0) a blu scuro (1)
    scale_fill_gradient(low = "whitesmoke", high = "steelblue", limits = c(0, 1), name = "Cramer's V") +
    labs(title = TITLE, x = "", y = "") +
    theme_minimal() +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1, size = 10, face = "bold"),
      axis.text.y = element_text(size = 10, face = "bold"),
      panel.grid = element_blank()
    )
}

# FIXME: Percentage shown in the wrong place.

#' @title Plot Target Ratio by Category
#' 
#' @description Creates a bivariate bar chart showing the relationship between a binary 
#' TARGET and a single categorical variable (CAT). It supports two visual styles: 
#' a 100% stacked bar chart ("stack") or a simple bar chart showing the rate of the target ("rate").
#' 
#' @param DF The input dataframe.
#' @param TARGET The binary column to analyze (unquoted). It will be standardized to 0/1.
#' @param CAT The categorical column for the X axis (unquoted).
#' @param x_title Label for the X axis.
#' @param y_title Label for the Y axis.
#' @param title Main title of the plot.
#' @param target_labels A character vector of length 2 defining the labels for 0 and 1 (e.g., c("Not escalated", "Escalated")).
#' @param style The visual style of the plot: "stack" (100% stacked bars) or "rate" (single bar for target = 1).
#' @param order_by Method to order the X axis categories: by "rate" (ascending target rate), "freq" (ascending frequency), or "none" (alphabetical/original).
#' @param percent_accuracy Accuracy for the percent labels (e.g., 0.1 for one decimal place).
#' @param colors A character vector of length 2 defining the hex color codes for the 0 and 1 categories respectively.
#' 
#' @return A ggplot2 object.
ratio_by_cat <- function(DF, TARGET, CAT, x_title, y_title, title, target_labels = c("Not escalated","Escalated"), style = c("stack","rate"), order_by = c("rate","freq","none"), percent_accuracy = 0.1, colors = c("#00bfc4", "#f8766d")){
  # Validate arguments
  style    <- match.arg(style)
  order_by <- match.arg(order_by)

  # Standardize target and extract the category
  dat <- DF %>%
    mutate(
      .t_raw = {{ TARGET }},
      # Convert into standard 1L or 0L integers
      .t = case_when(
        as.character(.t_raw) %in% c("1","TRUE","true") ~ 1L,
        as.character(.t_raw) %in% c("0","FALSE","false") ~ 0L,
        TRUE ~ NA_integer_
      ),
      .cat = {{ CAT }}
    ) %>%
    # Filter NAs
    filter(!is.na(.t), !is.na(.cat))

  # Compute rates and frequencies
  rates <- dat %>% 
    group_by(.cat) %>%
    summarise(total = n(), rate = mean(.t == 1), .groups = "drop")

  # Determine the ordering for X axis
  cat_levels <- switch(
    order_by,
    rate = rates %>% arrange(rate) %>% pull(.cat),
    freq = rates %>% arrange(total) %>% pull(.cat),
    none = unique(dat$.cat)
  )

  # Build the plot based on the chosen 'style'
  if (style == "stack") {
    # Prepare data for stacked bar chart
    stack <- dat %>%
      count(.cat, .t, name = "n") %>%
      group_by(.cat) %>%
      mutate(ratio = n/sum(n)) %>%
      ungroup() %>%
      mutate(
        # Apply the ordering
        .cat_f = factor(as.character(.cat), levels = as.character(cat_levels)),
        # Apply custom labels
        .t_f = factor(.t, levels = c(0,1), labels = target_labels)
      ) %>%
      # Join with the rates
      left_join(rates, by = ".cat")
    
    # Build the stacked plot
    ggplot(stack, aes(x = .cat_f, y = ratio, fill = .t_f)) +
      # Forces the bars to stack to 100%
      geom_col(position = "fill") +
      # Add exact percentage labels (only for the TARGET == 1 category)
      geom_text(
        data = subset(stack, .t_f == target_labels[2]),
        aes(label = scales::percent(rate, accuracy = percent_accuracy)),
        position = position_fill(vjust = 0.5), 
        size = 3
      ) +
      # Make continuous
      scale_y_continuous(labels = scales::percent) +
      # Map colors
      scale_fill_manual(values = setNames(colors, target_labels)) +
      labs(x = x_title, y = y_title, fill = NULL, title = title) +
      theme_minimal() +
      theme(axis.text.x = element_text(angle = 45, hjust = 1),
            panel.grid = element_blank())
  } else {
    # Prepare for rate bar chart (bars represent % of TARGET == 1)
    rates <- rates %>%
      mutate(
        .cat_f = factor(as.character(.cat),levels = as.character(cat_levels))
      )

    # Build the rate plot
    ggplot(rates, aes(x = .cat_f, y = rate)) +
      # Use second color to represent TARGET == 1
      geom_col(width = 0.8, fill = colors[2]) +
      # Add percentage labels
      geom_text(
        aes(label = scales::percent(rate, accuracy = percent_accuracy)),
        vjust = -0.35, size = 3
      ) +
      # Make continuous and expand
      scale_y_continuous(labels = scales::percent, expand = expansion(mult = c(0, 0.12))) +
      coord_cartesian(clip = "off") +
      labs(x = x_title, y = y_title, title = title,
           subtitle = "Bars show % of target = 1") +
      theme_minimal() +
      theme(axis.text.x = element_text(angle = 45, hjust = 1),
            panel.grid = element_blank())
  }
}

```


```{r multivariate-plot-helpers}
# ==============================================================================
# 3. Plots for Multivariate Analysis
# ==============================================================================

#' @title Heatmap of Target Rate (Categorical Y vs Numeric X)
#' 
#' @description Plots a heatmap showing the rate of a binary TARGET across a categorical
#' Y variable and a numeric (or factor) X variable.
#' 
#' @param DF The input dataframe.
#' @param TARGET The binary column to calculate the rate for (unquoted).
#' @param Y Categorical variable for the Y axis.
#' @param X Numeric or categorical variable for the X axis.
#' @param X_TITLE Label for the X axis.
#' @param Y_TITLE Label for the Y axis.
#' @param TITLE Main title of the plot.
#' @param min_n Minimum number of observations required to show a tile.
#' @param drop_na Boolean; if TRUE, rows with NA in X or Y are removed.
#' @param x_as_factor Boolean; if TRUE, forces the X variable to be treated as a categorical factor.
#' 
#' @return A ggplot2 object.
heat_target_yx <- function(DF, TARGET, Y, X, X_TITLE, Y_TITLE, TITLE, min_n = 0, drop_na = TRUE, x_as_factor = FALSE) {
  # Standardize the binary target
  dat <- DF %>%
    mutate(
      .target_raw = {{ TARGET }},
      .target_bin = dplyr::case_when(
        as.character(.target_raw) %in% c("1", "TRUE", "true")  ~ 1L,
        as.character(.target_raw) %in% c("0", "FALSE", "false") ~ 0L,
        TRUE ~ NA_integer_
      )
    )

  # Filter NAs
  if (drop_na) {
    dat <- dat %>% filter(!is.na({{Y}}), !is.na({{X}}), !is.na(.target_bin))
  } else {
    dat <- dat %>% filter(!is.na(.target_bin))
  }

  # Incrementally sort Y axis based on the mean ratio of the target
  y_order <- dat %>%
    group_by({{Y}}) %>%
    summarise(rate = mean(.target_bin == 1, na.rm = TRUE), .groups = "drop") %>%
    arrange(rate) %>%
    pull({{Y}})

  # Compute the target rate for each combination of Y x X
  grid <- dat %>%
    group_by({{Y}}, {{X}}) %>%
    summarise(
      rate = mean(.target_bin == 1, na.rm = TRUE),
      n = n(), .groups = "drop"
    ) %>%
    mutate(
      .y = factor(as.character({{Y}}), levels = as.character(y_order)),
      .x = if (x_as_factor) factor({{X}}) else {{X}},
      # Hide cells that do not reach the minimum amount of observations
      rate_plot = ifelse(n >= min_n, rate, NA_real_)
    )

  # Build ggplot heatmap
  ggplot(grid, aes(.x, .y, fill = rate_plot)) +
    geom_tile(width = 1, height = 0.9) +
    scale_fill_viridis_c(labels = scales::percent, na.value = "grey90", name = "Rate") +
    labs(x = X_TITLE, y = Y_TITLE, title = TITLE) +
    theme_minimal() +
    theme(panel.grid = element_blank())
}


#' @title Heatmap of Target Rate (Categorical Y vs Categorical X)
#' 
#' @description Plots a heatmap showing the rate of a binary TARGET across two categorical
#' variables (Y and X). Allows dynamic ordering and shading based on sample size.
#' 
#' @param DF The input dataframe.
#' @param TARGET The binary column to calculate the rate for (unquoted).
#' @param Y Categorical variable for the Y axis.
#' @param X Categorical variable for the X axis.
#' @param X_TITLE Label for the X axis.
#' @param Y_TITLE Label for the Y axis.
#' @param TITLE Main title of the plot.
#' @param min_n Minimum number of observations required to show a tile (filters noise).
#' @param order_y_by Method to order the Y axis: by 'rate', 'freq' (frequency), or 'none'.
#' @param order_x_by Method to order the X axis: by 'rate', 'freq' (frequency), or 'none'.
#' @param alpha_by_n Boolean; if TRUE, shades tiles by count using alpha (transparency).
#' 
#' @return A ggplot2 object.
heat_target_catcat <- function(DF, TARGET, Y, X, X_TITLE, Y_TITLE, TITLE, min_n = 0, order_y_by = c("rate","freq","none"), order_x_by = c("rate","freq","none"), alpha_by_n = FALSE ) {
  order_y_by <- match.arg(order_y_by)
  order_x_by <- match.arg(order_x_by)

  # Standardize target
  dat <- DF %>%
    mutate(
      .target_raw = {{ TARGET }},
      .target_bin = dplyr::case_when(
        as.character(.target_raw) %in% c("1", "TRUE", "true")  ~ 1L,
        as.character(.target_raw) %in% c("0", "FALSE", "false") ~ 0L,
        TRUE ~ NA_integer_
      )
    ) %>%
    # Filter NAs
    filter(!is.na({{Y}}), !is.na({{X}}), !is.na(.target_bin))

  # Compute stats to determine the Y axis order
  y_stats <- dat %>%
    group_by({{Y}}) %>%
    summarise(
      rate = mean(.target_bin == 1, na.rm = TRUE),
      n = n(), .groups = "drop"
    )
  
  # Compute stats to determine the X axis order
  x_stats <- dat %>%
    group_by({{X}}) %>%
    summarise(
      rate = mean(.target_bin == 1, na.rm = TRUE), 
      n = n(), .groups = "drop"
    )

  # Apply chosen "order by" to Y
  y_levels <- switch(
    order_y_by,
    rate = y_stats %>% arrange(rate) %>% pull({{Y}}),
    freq = y_stats %>% arrange(n) %>% pull({{Y}}),
    none = unique(dat %>% pull({{Y}}))
  )
  
  # Apply chosen "order by" to X
  x_levels <- switch(
    order_x_by,
    rate = x_stats %>% arrange(rate) %>% pull({{X}}),
    freq = x_stats %>% arrange(n) %>% pull({{X}}),
    none = unique(dat %>% pull({{X}}))
  )

  # Compute the cell's target rate and counts for each combination of Y x X
  grid <- dat %>%
    group_by({{Y}}, {{X}}) %>%
    summarise(
      rate = mean(.target_bin == 1, na.rm = TRUE),
      n = n(), .groups = "drop"
    ) %>%
    mutate(
      .y = factor(as.character({{Y}}), levels = as.character(y_levels)),
      .x = factor(as.character({{X}}), levels = as.character(x_levels)),
      rate_plot = ifelse(n >= min_n, rate, NA_real_)
    )

  # Build the plot
  p <- ggplot(grid, aes(.x, .y)) +
    labs(x = X_TITLE, y = Y_TITLE, title = TITLE) +
    theme_minimal() +
    theme(panel.grid = element_blank()) +
    # Use scales for displaying rates
    scale_fill_viridis_c(labels = scales::percent, na.value = "grey90", name = "Rate")

  if (alpha_by_n) {
    p <- p +
      geom_tile(aes(fill = rate_plot, alpha = n), width = 0.95, height = 0.95) +
      scale_alpha_continuous(range = c(0.25, 1), name = "Stops")
  } else {
    p <- p +
      geom_tile(aes(fill = rate_plot), width = 0.95, height = 0.95)
  }

  p
}


#' @title Stacked Ratio Bars Split by a Third Variable
#' 
#' @description Creates a faceted 100% stacked bar chart. This is highly effective 
#' for visualizing how the rate of a binary TARGET changes across a primary category (CAT), 
#' while simultaneously splitting the data by a third variable (SPLIT) to create side-by-side plots.
#' 
#' @param DF The input dataframe.
#' @param TARGET The binary column to analyze (unquoted). It will be standardized to 0/1.
#' @param CAT The main categorical variable for the X axis (unquoted).
#' @param SPLIT The categorical variable used to split/facet the plots (unquoted).
#' @param x_title Label for the X axis.
#' @param y_title Label for the Y axis.
#' @param title Main title of the entire plot.
#' @param target_labels Character vector of length 2 defining labels for the binary outcomes (e.g., c("Not escalated", "Escalated")).
#' @param split_labels Optional character vector to provide custom titles for the split panels (facets).
#' @param palette A character string specifying an RColorBrewer palette name (e.g., "Paired", "Set1").
#' @param percent_accuracy Accuracy for the percentage labels (e.g., 0.1 for one decimal place).
#' 
#' @return A ggplot2 object.
ratio_by_cat_split <- function(DF, TARGET, CAT, SPLIT, x_title, y_title, title, target_labels = c("Not escalated", "Escalated"), split_labels = NULL, palette = "Paired", percent_accuracy = 0.1){
  # Standardize target
  dat <- DF %>%
    mutate(
      .target_raw = {{ TARGET }},
      # Conversion to 0/1
      .target_bin = case_when(
        as.character(.target_raw) %in% c("1", "TRUE", "true") ~ 1L,
        as.character(.target_raw) %in% c("0", "FALSE", "false") ~ 0L,
        TRUE ~ NA_integer_
      ),
      .cat = {{ CAT }},
      .split = {{ SPLIT }}
    ) %>%
    # Filter NAs
    filter(!is.na(.cat), !is.na(.split), !is.na(.target_bin))

  # Order the main categories based on the overall target rate (from lowest to highest)
  cat_order <- dat %>%
    group_by(.cat) %>%
    summarise(
      rate = mean(.target_bin == 1),
      .groups = "drop"
    ) %>%
    arrange(rate) %>% 
    pull(.cat)

  # Tidy labels by converting character to factors
  dat <- dat %>%
    mutate(
      # Apply the order to X axis
      .cat = factor(as.character(.cat), levels = as.character(cat_order)),
      # Apply custom labels to the target
      .target_f = factor(.target_bin, levels = c(0, 1), labels = target_labels),
      # Format the split variable
      .split_f = {
        lv <- unique(as.character(.split))
        # If custom split labels are provided and the length matches, use them
        if (!is.null(split_labels) && length(split_labels) == length(lv))
          factor(as.character(.split), levels = lv, labels = split_labels)
        else factor(as.character(.split), levels = lv)
      }
    )

  # Compute rates
  plot_df <- dat %>%
    count(.split_f, .cat, .target_f, name = "n") %>%
    group_by(.split_f, .cat) %>%
    mutate(ratio = n/sum(n)) %>%
    ungroup()

  # Build plot
  ggplot(plot_df, aes(x = .cat, y = ratio, fill = .target_f)) +
    # Stacks the bars to exactly 100%
    geom_col(position = "fill") +
    # Add ratio text labels
    geom_text(
      aes(label = percent(ratio, accuracy = percent_accuracy)),
      position = position_fill(vjust = 0.5), 
      size = 3
    ) +
    # Format Y axis as percentages and make continuous
    scale_y_continuous(labels = percent) +
    # Use a clean palette for the target categories
    scale_fill_brewer(palette = palette, name = rlang::as_name(rlang::enquo(TARGET))) +
    labs(x = x_title, y = y_title, title = title) +
    # Split the plot into separate panels based on the .split_f variable
    facet_wrap(~ .split_f, nrow = 1) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
          panel.grid = element_blank())
}
```


```{r}
#' @title Plot Temporal Trend of Target Rate
#' 
#' @description Creates a scatter plot with a LOESS smoothing curve to visualize 
#' how the rate of a target event changes over a continuous temporal variable (e.g., date, day of year).
#' The size of the points represents the volume of stops.
#' 
#' @param DF The input dataframe.
#' @param TARGET The binary column to analyze (unquoted).
#' @param TIME_VAR The continuous time variable for the X axis (e.g., date, doy).
#' @param x_title Label for the X axis.
#' @param y_title Label for the Y axis.
#' @param title Main title of the plot.
plot_temporal_trend <- function(DF, TARGET, TIME_VAR, x_title = "Tempo", y_title = "Tasso", title = "Trend Temporale") {
  
  # Step 1: Standardizzare il target a 0/1
  dat <- DF %>%
    mutate(
      .target_raw = {{ TARGET }},
      .target_bin = case_when(
        as.character(.target_raw) %in% c("1","TRUE","true", "Yes")  ~ 1L,
        as.character(.target_raw) %in% c("0","FALSE","false", "No") ~ 0L,
        TRUE ~ NA_integer_
      ),
      .time = {{ TIME_VAR }}
    ) %>%
    filter(!is.na(.target_bin), !is.na(.time))
  
  # Step 2: Calcolare il tasso e il volume per ogni unità di tempo
  plot_df <- dat %>%
    group_by(.time) %>%
    summarise(
      rate = mean(.target_bin == 1, na.rm = TRUE),
      n = n(),
      .groups = "drop"
    )
  
  # Step 3: Costruire il grafico
  ggplot(plot_df, aes(x = .time, y = rate)) +
    # Punti proporzionali al numero di stop per capire la robustezza del dato
    geom_point(aes(size = n), alpha = 0.5, color = "steelblue") +
    # Linea di tendenza morbida per evidenziare il trend generale ignorando il rumore
    geom_smooth(se = FALSE, method = "loess", span = 0.3, color = "firebrick", size = 1.2) +
    scale_y_continuous(labels = scales::percent) +
    labs(x = x_title, y = y_title, size = "N. di Stop", title = title) +
    theme_minimal() +
    theme(panel.grid.minor = element_blank())
}

#' @title Plot Cyclic Time Heatmap
#' 
#' @description Creates a heatmap specifically designed for cyclic time variables 
#' (like Hour of Day vs Day of Week) to spot temporal hotspots.
#' 
#' @param DF The input dataframe.
#' @param TARGET The binary column to analyze (unquoted).
#' @param HOUR_VAR The variable representing the hour (0-23).
#' @param WEEKDAY_VAR The variable representing the day of the week.
#' @param title Main title of the plot.
plot_cyclic_time_heatmap <- function(DF, TARGET, HOUR_VAR, WEEKDAY_VAR, title = "Hotspots Temporali") {
  
  # Step 1: Standardizzare il target a 0/1
  dat <- DF %>%
    mutate(
      .target_raw = {{ TARGET }},
      .target_bin = case_when(
        as.character(.target_raw) %in% c("1","TRUE","true", "Yes")  ~ 1L,
        as.character(.target_raw) %in% c("0","FALSE","false", "No") ~ 0L,
        TRUE ~ NA_integer_
      ),
      .hour = as.numeric({{ HOUR_VAR }}),
      .day = {{ WEEKDAY_VAR }}
    ) %>%
    filter(!is.na(.target_bin), !is.na(.hour), !is.na(.day))
  
  # Step 2: Raggruppare e calcolare il tasso
  plot_df <- dat %>%
    group_by(.day, .hour) %>%
    summarise(
      rate = mean(.target_bin == 1, na.rm = TRUE),
      n = n(),
      .groups = "drop"
    ) %>%
    # Mostriamo solo incroci con un minimo di stop per evitare rumore statistico
    filter(n >= 10) 
  
  # Step 3: Costruire la mappa di calore temporale
  ggplot(plot_df, aes(x = .hour, y = .day, fill = rate)) +
    geom_tile(color = "white") +
    scale_fill_viridis_c(option = "magma", labels = scales::percent, name = "Tasso") +
    # Impostiamo l'asse X per mostrare tutte le 24 ore
    scale_x_continuous(breaks = 0:23, expand = c(0,0)) +
    labs(x = "Ora del Giorno", y = "", title = title) +
    theme_minimal() +
    theme(
      panel.grid = element_blank(),
      axis.text.y = element_text(face = "bold")
    )
}

#' @title Plot Temporal Trend Split by a Group (Rate or Frequency)
#' 
#' @description Generates a line plot showing how a metric (rate or count) 
#' changes over a generic time unit (X_VAR), split by another variable (Z_VAR).
#' Adapts to any time unit and can display data as overlapping lines or faceted grids.
#' 
#' @param DF The input dataframe.
#' @param X_VAR The generic time variable for the X axis (e.g., month, hour).
#' @param Z_VAR The variable to group/split the plots by (e.g., year, weekend).
#' @param TARGET The binary target variable. Required only if metric = "rate".
#' @param metric "rate" (percentage of TARGET) or "freq" (absolute count of stops).
#' @param style "overlay" (lines superimposed on one plot) or "facet" (separated sub-plots).
#' @param x_title Label for the X axis.
#' @param y_title Label for the Y axis. Default adjusts based on metric.
#' @param title Main title of the plot.
#' 
#' @return A ggplot2 object.
plot_temporal_split <- function(
  DF, X_VAR, Z_VAR, TARGET = NULL, 
  metric = c("rate", "freq"), 
  style = c("overlay", "facet"),
  x_title = "Tempo", y_title = NULL, title = "Trend Temporale"
) {
  
  metric <- match.arg(metric)
  style  <- match.arg(style)
  
  # Step 1: Sicurezza per il Target
  target_quo <- rlang::enquo(TARGET)
  if (metric == "rate" && rlang::quo_is_null(target_quo)) {
    stop("Errore: Devi specificare la colonna TARGET se usi metric = 'rate'.")
  }
  
  # Step 2: Estrazione variabili principali
  dat <- DF %>%
    mutate(
      .x = {{ X_VAR }},
      .z = {{ Z_VAR }}
    ) %>%
    filter(!is.na(.x), !is.na(.z))
  
  # Step 3: Calcolo delle Metriche
  if (metric == "rate") {
    dat <- dat %>%
      mutate(
        .t_raw = !!target_quo,
        .t_bin = case_when(
          as.character(.t_raw) %in% c("1","TRUE","true","Yes")  ~ 1L,
          as.character(.t_raw) %in% c("0","FALSE","false","No") ~ 0L,
          TRUE ~ NA_integer_
        )
      ) %>%
      filter(!is.na(.t_bin))
    
    plot_df <- dat %>%
      group_by(.z, .x) %>%
      summarise(y_val = mean(.t_bin == 1), n = n(), .groups = "drop")
    
    if (is.null(y_title)) y_title <- "Tasso (%)"
    
  } else { # metric == "freq"
    plot_df <- dat %>%
      group_by(.z, .x) %>%
      summarise(y_val = n(), n = n(), .groups = "drop")
    
    if (is.null(y_title)) y_title <- "Numero di Stop"
  }
  
  # Assicuriamoci che l'asse X sia trattato correttamente nel disegno delle linee
  plot_df <- plot_df %>% mutate(.z_f = factor(.z))
  
  # Step 4: Costruzione del Grafico base
  p <- ggplot(plot_df, aes(x = .x, y = y_val))
  
  # Logica dello "Stile"
  if (style == "overlay") {
    # Tutte le linee nello stesso grafico
    p <- p +
      geom_line(aes(color = .z_f, group = .z_f), linewidth = 0.8) +
      geom_point(aes(color = .z_f, size = n), alpha = 0.7) +
      labs(color = rlang::as_name(rlang::enquo(Z_VAR)), size = "Stops Count") +
      scale_color_viridis_d(option = "turbo") # Colori ben distinguibili
      
  } else { 
    # Stile "Facet" (Grafici separati impilati)
    p <- p +
      geom_line(linewidth = 0.6, group = 1, color = "steelblue") +
      geom_point(aes(size = n), alpha = 0.7, color = "steelblue") +
      # free_x permette ad ogni grafico di avere il suo asse se necessario
      facet_wrap(~ .z_f, ncol = 1, scales = "free_x") +
      labs(size = "Stops Count")
  }
  
  # Temi e Asse Y
  p <- p +
    labs(x = x_title, y = y_title, title = title) +
    theme_minimal() +
    theme(
      panel.grid.minor = element_blank(),
      strip.text = element_text(face = "bold", size = 11) # Titoli dei facet in grassetto
    )
  
  # Formattazione dell'Asse Y in base alla metrica scelta
  if (metric == "rate") {
    p <- p + scale_y_continuous(labels = scales::percent)
  } else {
    p <- p + scale_y_continuous(labels = scales::comma)
  }
  
  return(p)
}
```

```{r}
#' @title Plot Overlapping Density
#' @description Compares the distribution of a numeric variable across different categories using density curves.
#' @param DF The input dataframe.
#' @param NUM_VAR The continuous numeric variable (unquoted, e.g., age).
#' @param CAT_VAR The categorical grouping variable (unquoted, e.g., arrested).
#' @param title Main title of the plot.
plot_density_compare <- function(DF, NUM_VAR, CAT_VAR, x_title = "Valore", title = "Confronto Densità") {
  
  dat <- DF %>%
    filter(!is.na({{ NUM_VAR }}), !is.na({{ CAT_VAR }})) %>%
    mutate(.cat = factor({{ CAT_VAR }}))
  
  ggplot(dat, aes(x = {{ NUM_VAR }}, fill = .cat)) +
    # alpha = 0.5 rende i colori trasparenti per vedere le sovrapposizioni
    geom_density(alpha = 0.5, color = "white") +
    scale_fill_brewer(palette = "Set1", name = "Gruppo") +
    labs(x = x_title, y = "Densità (Probabilità)", title = title) +
    theme_minimal() +
    theme(panel.grid.minor = element_blank())
}

#' @title Plot Violin and Boxplot Combination
#' @description Shows the distribution, median, and density of a numeric variable across multiple categories.
#' @param DF The input dataframe.
#' @param CAT_VAR The categorical variable for the X axis (unquoted).
#' @param NUM_VAR The numeric variable for the Y axis (unquoted).
#' @param title Main title of the plot.
plot_violin_boxplot <- function(DF, CAT_VAR, NUM_VAR, x_title = "Categoria", y_title = "Valore Numerico", title = "Distribuzione Numerica per Categoria") {
  
  dat <- DF %>% filter(!is.na({{ CAT_VAR }}), !is.na({{ NUM_VAR }}))
  
  ggplot(dat, aes(x = reorder(factor({{ CAT_VAR }}), {{ NUM_VAR }}, FUN = median, na.rm=TRUE), y = {{ NUM_VAR }}, fill = factor({{ CAT_VAR }}))) +
    # Il violino mostra la densità (più è largo, più dati ci sono)
    geom_violin(alpha = 0.6, trim = FALSE, color = NA) +
    # Il boxplot interno (molto stretto) mostra la mediana e i quartili
    geom_boxplot(width = 0.1, fill = "white", color = "black", alpha = 0.8, outlier.size = 1) +
    scale_fill_viridis_d(option = "turbo") +
    coord_flip() + # Capovolgiamo per leggere meglio le etichette lunghe
    labs(x = x_title, y = y_title, title = title) +
    theme_minimal() +
    theme(legend.position = "none", panel.grid.minor = element_blank())
}

#' @title Plot Spatial Map of Target
#' @description Plots coordinates (Longitude vs Latitude) to create a visual map, coloring points by a target variable.
#' @param DF The input dataframe.
#' @param LON Longitude column (unquoted).
#' @param LAT Latitude column (unquoted).
#' @param TARGET Categorical or binary variable for coloring (unquoted).
#' @param title Main title of the plot.
plot_spatial_map <- function(DF, LON, LAT, TARGET, title = "Mappa Spaziale degli Stop") {
  
  dat <- DF %>% filter(!is.na({{ LON }}), !is.na({{ LAT }}), !is.na({{ TARGET }}))
  
  # Ordiniamo i dati in modo che i target "positivi" (es. arresti) 
  # vengano disegnati per ultimi e non siano nascosti dalla massa di stop normali
  dat <- dat %>% arrange({{ TARGET }})
  
  ggplot(dat, aes(x = {{ LON }}, y = {{ LAT }}, color = factor({{ TARGET }}))) +
    # Usiamo punti molto piccoli e trasparenti per gestire l'overplotting (troppi punti sovrapposti)
    geom_point(size = 0.5, alpha = 0.4) +
    scale_color_brewer(palette = "Set1", name = "Esito") +
    # Manteniamo le proporzioni geografiche corrette (coord_quickmap)
    coord_quickmap() + 
    labs(x = "Longitudine", y = "Latitudine", title = title) +
    theme_minimal() +
    theme(
      axis.text = element_blank(), # Rimuoviamo i numeri degli assi per sembrare una vera mappa
      panel.grid = element_blank()
    )
}
```

---

Let us load the data now, and remove the extra feature `veh_category` because we are **exclusively analyzing 2 wheels vehicles now**.

```{r load-data-drop-veh-categ}
vehstops = read.csv("data/vehiclestops_2_wheels.csv")

vehstops <- vehstops %>%
  select(-veh_categ)
```

So now we have the following dataset, that we present with the usual summary

```{r summary-data}
print(dfSummary(vehstops), method="render")
```

This preview is already interesting.

TODO: Comment

------------------------------------------------------------------------

# Univariate EDA

TODO: Introduction
TODO: Correct the following

In this part we perform a deeper analysis of the target feature, and some of the most important features (according to the domain knowledge in notebook 0) against their distribution in the dataset.

Our goal here is to polish the data, by removing outliers and possible confounders that may introduce bias. Furthermore, we need to understand its distribution and normalizing it if necessary.

Possibly, we can introduce new variables that are transformations or combinations of other ones, and evaluate their contributions later with a bivariate analysis and/or model testing.

Hopefully, this process will reduce the variance.

TODO: Comment everything

## Target

TODO: C

```{r enforcement-dist}
p1 <- plot_binary_bar(
  vehstops, veh_checkpoint, 
  "No", "Yes",
  "", "", "Was it a vehicle checkpoint?"
)

p2 <- plot_binary_bar(
  vehstops, arrested, 
  "No", "Yes",
  "", "", "Was the motorist arrested?"
)

row1 <- p1 | p2

row1 +
plot_layout(heights = c(1, 0.08, 1), guides = "collect") + 
plot_annotation(title = "Counts of stops by target") &
theme(legend.position = "bottom") 
```

## Command Unit, Geospatial and Area-related predictors

### Area

```{r count-stops-by-area}
plot_univariate_cat(
  DF = vehstops, 
  FEATURE = area, 
  X_LABEL = "Area", 
  TITLE = "Distribution of Stops by Area",
  order_by = "freq"
)
```

```{r area-dist}
vehstops %>%
  filter(!is.na(area)) %>%
  count(area, name = "n") %>%
  mutate(area = fct_reorder(area, n, .desc = TRUE)) %>%
  ggplot(aes(x = factor(area), y = n, fill = factor(area))) +
  geom_col(width = 0.8, show.legend = FALSE) +
  geom_text(aes(label = scales::comma(n)), vjust = -0.35, size = 3) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
  coord_cartesian(clip = "off") +
  labs(x = "Area", y = "Count", title = "Distribution of Stops by Area") +
  theme_minimal()
```

### Borough

```{r count-stosps-by-boro}
plot_univariate_cat(
  DF = vehstops, 
  FEATURE = boro, 
  X_LABEL = "Borough", 
  TITLE = "Distribution of Stops by Borough",
)
```

```{r boro-dist}
vehstops %>%
  filter(!is.na(boro)) %>%
  count(boro, name = "n") %>%
  mutate(boro = fct_reorder(boro, n, .desc = TRUE)) %>% 
  ggplot(aes(x = factor(boro), y = n, fill = factor(boro))) +
  geom_col(width = 0.8, show.legend = FALSE) +
  geom_text(aes(label = scales::comma(n)), vjust = -0.35, size = 3) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
  coord_cartesian(clip = "off") +
  labs(x = "Borough", y = "", title = "Count of Stops") +
  theme_minimal() + 
  scale_x_discrete(guide = guide_axis(angle = 30, n.dodge = 1))
```

### Command Unit

```{r count-stopsq-by-co}
plot_univariate_cat(
  DF = vehstops, 
  FEATURE = command, 
  X_LABEL = "Command Unit", 
  TITLE = "Distribution of Stops by Command Unit",
)
```

```{r command-dist}
vehstops %>%
  filter(!is.na(command)) %>%
  count(command, name = "n") %>%
  mutate(command = fct_reorder(as.factor(command), n, .desc = TRUE)) %>% 
  ggplot(aes(x = factor(command), y = n, fill = factor(command))) +
  geom_col(width = 0.8, show.legend = FALSE) +
  geom_text(aes(label = scales::comma(n)), vjust = -0.5, size = 1) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
  coord_cartesian(clip = "off") +
  labs(x = "Command", y = "", title = "Count of Stops") +
  theme_minimal() + 
  scale_x_discrete(guide = guide_axis(angle = 30, n.dodge = 2))
```

### Latitude and Longitude

```{r stop-density}
nyc_bbox <- list(lat_min = 40.48, lat_max = 40.92, lon_min = -74.27, lon_max = -73.68)
veh_geo_nyc <- vehstops %>%
  filter(!is.na(latitude), !is.na(longitude)) %>%
  filter(
    between(latitude,  nyc_bbox$lat_min, nyc_bbox$lat_max),
    between(longitude, nyc_bbox$lon_min, nyc_bbox$lon_max)
  )

bins <- 80
min_n <- 25

lon_breaks <- seq(nyc_bbox$lon_min, nyc_bbox$lon_max, length.out = bins + 1)
lat_breaks <- seq(nyc_bbox$lat_min, nyc_bbox$lat_max, length.out = bins + 1)

grid_sum <- veh_geo_nyc %>%
  mutate(
    xbin = cut(longitude, lon_breaks, include.lowest = TRUE, right = FALSE),
    ybin = cut(latitude,  lat_breaks, include.lowest = TRUE, right = FALSE)
  ) %>%
  group_by(xbin, ybin) %>%
  summarise(
    n    = n(),
    rate = mean(arrested == 1, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    xi = as.integer(xbin),
    yi = as.integer(ybin),
    # bin centers (for geom_tile positioning)
    x = (lon_breaks[xi] + lon_breaks[xi + 1]) / 2,
    y = (lat_breaks[yi] + lat_breaks[yi + 1]) / 2,
    # optionally suppress super-sparse bins in the rate coloring
    rate_plot = ifelse(n >= min_n, rate, NA_real_)
  )

p1 <- ggplot(grid_sum, aes(x, y, fill = n)) +
  geom_tile() +
  coord_equal(xlim = c(nyc_bbox$lon_min, nyc_bbox$lon_max),
              ylim = c(nyc_bbox$lat_min, nyc_bbox$lat_max), expand = FALSE) +
  scale_fill_viridis_c(trans = "log10", name = "Stops") +
  labs(title = "Stop density (2D bins)", x = "Longitude", y = "Latitude")

p1
```


## Time-related predictors

### Year

```{r count-stops-by-boro}
plot_univariate_cat(
  DF = vehstops, 
  FEATURE = year, 
  X_LABEL = "Year", 
  TITLE = "Distribution of Stops by Year",
)
```

```{r year-distribution}
vehstops %>%
  filter(!is.na(year)) %>%
  count(year, name = "n") %>%
  ggplot(aes(x = factor(year), y = n, fill = factor(year))) +
  geom_col(width = 0.8, show.legend = FALSE) +
  geom_text(aes(label = scales::comma(n)), vjust = -0.35, size = 3) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
  coord_cartesian(clip = "off") +
  labs(x = "Year", y = "", title = "Count of Stops") +
  theme_minimal()
```

### Month

Since it is not safe to group by year 2025 only has around 6 months of data), we divide this in 3 parts.

```{r month-distribution}
p1 <- vehstops %>%
  filter(!is.na(year) & year == 2023) %>%
  count(month, name = "n") %>%
  ggplot(aes(x = factor(month), y = n, fill = factor(month))) +
  geom_col(width = 0.8, show.legend = FALSE) +
  geom_text(aes(label = scales::comma(n)), vjust = -0.35, size = 3) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
  coord_cartesian(clip = "off") +
  labs(x = "Months - Year 2023", y = "") +
  theme_minimal()

p2 <- vehstops %>%
  filter(!is.na(year) & year == 2024) %>%
  count(month, name = "n") %>%
  ggplot(aes(x = factor(month), y = n, fill = factor(month))) +
  geom_col(width = 0.8, show.legend = FALSE) +
  geom_text(aes(label = scales::comma(n)), vjust = -0.35, size = 3) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
  coord_cartesian(clip = "off") +
  labs(x = "Months - Year 2024", y = "") +
  theme_minimal()
  
p3 <- vehstops %>%
  filter(!is.na(year) & year == 2025) %>%
  count(month, name = "n") %>%
  ggplot(aes(x = factor(month), y = n, fill = factor(month))) +
  geom_col(width = 0.8, show.legend = FALSE) +
  geom_text(aes(label = scales::comma(n)), vjust = -0.35, size = 3) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
  coord_cartesian(clip = "off") +
  labs(x = "Months - Year 2025", y = "") +
  theme_minimal()

(p1 / p2 / p3) + 
  plot_layout(widths = c(2, 2), guides = "collect") +
  plot_annotation(title = "Counts of Stops by Month") &
  theme(legend.position = "bottom")
```

In fact, now it makes more sense, because the cumulative sum of the first six months of the year of 2025 is higher than the first six months of the previous years! 

### Day and Dow

```{r day-dist}
vehstops %>%
  filter(!is.na(day)) %>%
  count(day, name = "n") %>%
  ggplot(aes(x = factor(day), y = n, fill = factor(day))) +
  geom_col(width = 0.8, show.legend = FALSE) +
  geom_text(aes(label = scales::comma(n)), vjust = -0.35, size = 2) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
  coord_cartesian(clip = "off") +
  labs(x = "Day", y = "", title="Counts of Stops by Day of the Month") +
  theme_minimal()
```

The day of the month does not seem to matter much and we see a decrease on the `31` day because not all months have them.

```{r dow-dis}
vehstops %>%
  filter(!is.na(dow)) %>%
  count(dow, name = "n") %>%
  ggplot(aes(x = factor(dow), y = n, fill = factor(dow))) +
  geom_col(width = 0.8, show.legend = FALSE) +
  geom_text(aes(label = scales::comma(n)), vjust = -0.35, size = 3) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
  coord_cartesian(clip = "off") +
  labs(x = "Day", y = "", title="Counts of Stops by Day of the Week") +
  theme_minimal()
```

This `doy` is hard to account for, and we are going to remove it from the clean version

```{r remove-doy}
# vehstops2 <- vehstops2 %>%
#   select(-doy)
```

### Weekend, Holiday and Event Day

```{r weekend-holiday-event-distribution}
p1 <- plot_binary_bar(
  vehstops, weekend, 
  "No", "Yes",
  "During a weekend?", "", "Counts of Stops"
)

p2 <- plot_binary_bar(
  vehstops, holiday,
  "No", "Yes",
  "During an holiday?", "", "Counts of Stops"
)

p3 <- plot_binary_bar(
  vehstops, event_day,
  "No", "Yes",
  "During an event?", "", "Counts of Stops"
)

(p1 + p2 + p3) + 
  plot_layout(ncol = 3, nrow = 1, widths = c(1, 1), guides = "collect") +
  theme(legend.position = "bottom") 
```

None of the two matter much in terms of stops


### Hour, Working Hour and Time Bracket

```{r}
# Vediamo i 5 Distretti (Borough) con più stop
plot_univariate_cat(
  DF = vehstops, 
  FEATURE = hour, 
  X_LABEL = "hour", 
  TITLE = "Distribuzione degli orari",
)
```

```{r hour-dist}
vehstops %>%
  filter(!is.na(hour)) %>%
  count(hour, name = "n") %>%
  ggplot(aes(x = factor(hour), y = n, fill = factor(hour))) +
  geom_col(width = 0.8, show.legend = FALSE) +
  geom_text(aes(label = scales::comma(n)), vjust = -0.35, size = 2.5) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
  coord_cartesian(clip = "off") +
  labs(x = "Hour", y = "", title = "Count of Stops by Hour of the Day") +
  theme_minimal() 
```

TODO: Check if we can change 0 with 24 for better display

```{r time-bracket-distribution}
vehstops %>%
  filter(!is.na(time_bracket)) %>%
  count(time_bracket, name = "n") %>%
  mutate(time_bracket = fct_reorder(as.factor(time_bracket), n, .desc = TRUE)) %>% 
  ggplot(aes(x = factor(time_bracket), y = n, fill = factor(time_bracket))) +
  geom_col(width = 0.8, show.legend = FALSE) +
  geom_text(aes(label = scales::comma(n)), vjust = -0.35, size = 2.5) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
  coord_cartesian(clip = "off") +
  labs(x = "Time Bracket", y = "", title = "Count of Stops by Time Bracket") +
  theme_minimal() + 
  scale_x_discrete(guide = guide_axis(angle = 30, n.dodge = 1))
```

```{r working-hour-distribution}
plot_binary_bar(
  vehstops, working_hour, 
  "No", "Yes",
  "Was it between 9am and 5pm?", "", "Counts of Stops by 9-5 Window"
)
```


### Minute

```{r minute-distribution}
vehstops %>%
  filter(!is.na(minute)) %>%
  count(minute, name = "n") %>%
  ggplot(aes(x = factor(minute), y = n, fill = factor(minute))) +
  geom_col(width = 0.8, show.legend = FALSE) +
  geom_text(aes(label = scales::comma(n)), vjust = -0.35, size = 1) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
  coord_cartesian(clip = "off") +
  labs(x = "Minute", y = "", title = "Count of Stops by Minute") +
  theme_minimal() + 
  scale_x_discrete(guide = guide_axis(angle = 30, n.dodge = 2))
```

This seems to either require some binning or just drop, but we are going to drop it for the clean version.

```{r drop-minute}
# vehstops2 <- vehstops2 %>%
#   select(-minute)
```


## Motorist related predictors

### Reported Age and Age Bracket

We have mainly 2 issues here. First, we have many instances of drivers with unknown age

```{r age-bracket-dist}
vehstops %>%
  filter(!is.na(age_bracket)) %>%
  count(age_bracket, name = "n") %>%
  ggplot(aes(x = factor(age_bracket), y = n, fill = factor(age_bracket))) +
  geom_col(width = 0.8, show.legend = FALSE) +
  geom_text(aes(label = scales::comma(n)), vjust = -0.35, size = 3) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +  # top padding
  coord_cartesian(clip = "off") +
  labs(x = "Reported Age Bracket", y = "", title = "Count of Stops by Reported Age Bracket") +
  theme_minimal()
```

But if we forget about them and we plot the distribution

```{r age-distribution}
vehstops %>%
  filter(!is.na(age)) %>%
  count(age) %>%
  ggplot(aes(y = n,x = factor(age), fill = factor(age))) +
  geom_col(width = 0.8, na.rm = TRUE, show.legend = FALSE) +
  labs(x = "Reported Age", y = "", title = "Count of Stops by Reported Age") +
  scale_x_discrete(guide = guide_axis(angle = 45, n.dodge = 2))
```

We can see that our second issue is the presence of very old drivers (90+),

```{r very-old-drivers}
vehstops %>%
  select(year, month, hour, time_bracket, area, age, sex, race, arrested) %>%
  filter(age > 90) %>%
  dfSummary()
```

Which may not contribute to much here. So in the second version of the dataset we are going to remove drivers older than 90 (but for the full dataset, we are going to wait).

```{r}
# vehstops2 <- vehstops2 %>%
#   filter(age <= 90)
```

Thus, 

```{r age-distribution2}
vehstops %>%
  filter(!is.na(age) & age < 90) %>%
  count(age) %>%
  ggplot(aes(y = n,x = factor(age), fill = factor(age))) +
  geom_col(width = 0.8, na.rm = TRUE, show.legend = FALSE) +
  labs(x = "Reported Age", y = "", title = "Count of Stops by Reported Age") +
  scale_x_discrete(guide = guide_axis(angle = 45, n.dodge = 2))
```

If there is evidence of a better binning with respect to the escalation, we might as well re-bin this predictor

### Reported Sex

```{r sex-distribution}
vehstops %>%
  filter(!is.na(sex)) %>%
  count(sex, name = "n") %>%
  ggplot(aes(x = factor(sex), y = n, fill = factor(sex))) +
  geom_col(width = 0.8, show.legend = FALSE) +
  geom_text(aes(label = scales::comma(n)), vjust = -0.35, size = 3) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
  coord_cartesian(clip = "off") +
  labs(x = "Reported Sex", y = "", title = "Count of Stops by Reported Sex") +
  theme_minimal()
```

### Reported Race

```{r race-distribution}
vehstops %>%
  filter(!is.na(race)) %>%
  count(race, name = "n") %>%
  ggplot(aes(x = factor(race), y = n, fill = factor(race))) +
  geom_col(width = 0.8, show.legend = FALSE) +
  geom_text(aes(label = scales::comma(n)), vjust = -0.35, size = 3) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
  coord_cartesian(clip = "off") +
  labs(x = "Reported Race", y = "", title = "Count of Stops by Reported Race") +
  theme_minimal() +
  scale_x_discrete(guide = guide_axis(angle = 30, n.dodge = 1))
```

------------------------------------------------------------------------

# Bivariate EDA

TODO: Introduction

Here forget about just analyzing the stops counts and we switch towards the main feature to see the proportion.

TODO: Comment everything

## Correlation - TODO

```{r correlation-heatmap}
colonne_da_verificare <- c("sex", "race", "age_bracket", "boro")

plot_categorical_correlation(
  DF = vehstops, 
  FEATURES = colonne_da_verificare,
  TITLE = "Correlazione tra le Variabili (Cramer's V)"
)
```

TODO: Correlation Heatmap with all features

TODO: Correlation Heatmap with "main" features

## Collinearity

## Command Unit, Geospatial and Area related Features

### Area

```{r area-vs-target}
ratio_by_cat(
  vehstops, arrested, area,
  "Area", "Share of stops", "Escalation ratio by area",
  style = "stack"
)
```


```{r area-counts-stacked}
p1 <- vehstops %>%
  filter(!is.na(area), !is.na(year)) %>%
  count(area, year, name = "n") %>%
  group_by(area) %>% mutate(total = sum(n)) %>% ungroup() %>%
  mutate(area = fct_reorder(area, total),
         year = factor(year)) %>%
  ggplot(aes(x = n, y = area, fill = year)) +
  geom_col(width = 0.8) +
  labs(x = "Stops", y = NULL, fill = "Year",
       title = "Stops by area, split by year") +
  theme_minimal()

p2 <- vehstops %>%
  filter(!is.na(area), !is.na(month)) %>%
  count(area, month, name = "n") %>%
  group_by(area) %>% mutate(total = sum(n)) %>% ungroup() %>%
  mutate(area  = fct_reorder(area, total),
         month = factor(month, levels = 1:12, labels = month.abb)) %>%
  ggplot(aes(x = n, y = area, fill = month)) +
  geom_col(width = 0.8) +
  labs(x = "Stops", y = NULL, fill = "Month",
       title = "Stops by area, split by month") +
  theme_minimal()

p1 / p2
```


### Borough

```{r boro-vs-target}
ratio_by_cat(
  vehstops, arrested, boro,
  "Borough", "Share of stops", "Escalation ratio by borough",
  style = "stack"
)
```


```{r boro-counts-stacked}
p1 <- vehstops %>%
  filter(!is.na(boro), !is.na(year)) %>%
  count(boro, year, name = "n") %>%
  group_by(boro) %>% mutate(total = sum(n)) %>% ungroup() %>%
  mutate(boro = fct_reorder(boro, total),
         year = factor(year)) %>%
  ggplot(aes(x = n, y = boro, fill = year)) +
  geom_col(width = 0.8) +
  labs(x = "Stops", y = NULL, fill = "Year",
       title = "Stops by boro, split by year") +
  theme_minimal()

p2 <- vehstops %>%
  filter(!is.na(boro), !is.na(month)) %>%
  count(boro, month, name = "n") %>%
  group_by(boro) %>% mutate(total = sum(n)) %>% ungroup() %>%
  mutate(boro  = fct_reorder(boro, total),
         month = factor(month, levels = 1:12, labels = month.abb)) %>%
  ggplot(aes(x = n, y = boro, fill = month)) +
  geom_col(width = 0.8) +
  labs(x = "Stops", y = NULL, fill = "Month",
       title = "Stops by boro, split by month") +
  theme_minimal()

p1 / p2
```


### Command

```{r command-vs-target}
# FIXME
ratio_by_cat(
  vehstops, arrested, command,
  "Command", "Share of stops", "Escalation ratio by command",
  style = "stack"
)
```


```{r command-counts-stacked}
# FIXME
p1 <- vehstops %>%
  filter(!is.na(command), !is.na(year)) %>%
  count(command, year, name = "n") %>%
  group_by(command) %>% mutate(total = sum(n)) %>% ungroup() %>%
  mutate(command = fct_reorder(as.factor(command), total), year = factor(year)) %>%
  ggplot(aes(x = n, y = command, fill = year)) +
  geom_col(width = 0.8) +
  labs(x = "Stops", y = NULL, fill = "Year",
       title = "Stops by Command, split by year") +
  theme_minimal()

p2 <- vehstops %>%
  filter(!is.na(command), !is.na(month)) %>%
  count(command, month, name = "n") %>%
  group_by(command) %>% mutate(total = sum(n)) %>% ungroup() %>%
  mutate(command  = fct_reorder(as.factor(command), total),
         month = factor(month, levels = 1:12, labels = month.abb)) %>%
  ggplot(aes(x = n, y = command, fill = month)) +
  geom_col(width = 0.8) +
  labs(x = "Stops", y = NULL, fill = "Month",
       title = "Stops by Command, split by month") +
  theme_minimal()

p1 / p2
```

TODO: What to do with command? too many labels already grouped in boro and area

## Time related features

```{r}
plot_temporal_split(
  DF = vehstops, 
  X_VAR = month, 
  Z_VAR = year, 
  TARGET = arrested, 
  metric = "rate", 
  style = "facet",              # <- Grafici in colonna!
  x_title = "Mese", 
  title = "Escalation rate by month (Faceted)"
)

```

```{r}
plot_temporal_split(
  DF = vehstops, 
  X_VAR = month, 
  Z_VAR = year, 
  TARGET = arrested, 
  metric = "rate", 
  style = "overlay",            # <- Tutti sovrapposti con colori diversi!
  x_title = "Mese", 
  title = "Monthly escalation rate overlaid by year"
)
```



### Year

```{r year-vs-target}
ratio_by_cat(
  vehstops, arrested, year,
  "Year", "Share of stops", "Escalation ratio by year",
  style = "stack"
)
```

### Month

```{r month-vs-target}
vehstops |>
  filter(!is.na(arrested), !is.na(month), !is.na(year)) |>
  group_by(year, month) |>
  summarise(rate = mean(arrested), n = n(), .groups = "drop") |>
  ggplot(aes(month, rate, group = 1)) +
  geom_line(linewidth = 0.6) +
  geom_point(size = 0.8, alpha = 0.7) +
  scale_x_continuous(breaks = seq(0, 12, by = 1)) +
  scale_y_continuous(labels = percent_format()) +
  labs(title = "Escalation rate by month", x = "Month", y = "Rate") +
  facet_wrap(~ year, ncol = 1, scales = "free_x") +
  theme_minimal()
```

Let us contextualize how this changes throughout the years

```{r }
vehstops %>%
  mutate(date = ymd(paste(year, month, day, sep = "-"))) %>%
  group_by(year, month) %>%
  summarize(rate = mean(arrested == 1, na.rm = TRUE), n = n(), .groups = "drop") %>%
  ggplot(aes(month, rate, group = factor(year), linetype = factor(year))) +
  geom_line(linewidth = 0.6) + geom_point(aes(size = n)) +
  scale_y_continuous(labels = percent) +
  scale_x_continuous(breaks = 1:12) +
  labs(x = "Month", y = "Event rate", linetype = "Year", size = "Stops count",
       title = "Monthly escalation rate by year")
```

```{r}
vehstops %>%
  filter(year %in% c(2023, 2024, 2025)) %>%
  group_by(year, month) %>%
  summarize(stops = n(), .groups = "drop") %>%
  ggplot(aes(month, stops, group = factor(year), linetype = factor(year))) +
  geom_line(linewidth = 0.6) + geom_point(size = 1.5) +
  scale_x_continuous(breaks = 1:12) +
  labs(x = "Month", y = "Stops", linetype = "Year",
       title = "Monthly stops: 2023–2025")
```

### Day, Doy and Dow

```{r doy-vs-target}
vehstops |>
  filter(!is.na(arrested), !is.na(doy), !is.na(year)) |>
  group_by(year, doy) |>
  summarise(rate = mean(arrested), n = n(), .groups = "drop") |>
  ggplot(aes(doy, rate, group = 1)) +
  geom_line(linewidth = 0.6) +
  geom_point(size = 0.8, alpha = 0.7) +
  scale_x_continuous(breaks = seq(0, 366, by = 30)) +
  scale_y_continuous(labels = percent_format()) +
  labs(title = "Escalation rate by day-of-year", x = "Day of Year", y = "Rate") +
  facet_wrap(~ year, ncol = 1, scales = "free_x") +
  theme_minimal()
```

```{r day-vs-target}
vehstops |>
  filter(!is.na(arrested), !is.na(day), !is.na(year)) |>
  group_by(year, day) |>
  summarise(rate = mean(arrested), n = n(), .groups = "drop") |>
  ggplot(aes(day, rate, group = 1)) +
  geom_line(linewidth = 0.6) +
  geom_point(size = 0.8, alpha = 0.7) +
  scale_x_continuous(breaks = seq(0, 31, by = 1)) +
  scale_y_continuous(labels = percent_format()) +
  labs(title = "Escalation rate by day-of-month", x = "Day of Month", y = "Rate") +
  facet_wrap(~ year, ncol = 1, scales = "free_x") +
  theme_minimal()
```

```{r dow-vs-target}
vehstops |>
  filter(!is.na(arrested), !is.na(dow), !is.na(year)) |>
  group_by(year, dow) |>
  summarise(rate = mean(arrested), n = n(), .groups = "drop") |>
  ggplot(aes(dow, rate, group = 1)) +
  geom_line(linewidth = 0.6) +
  geom_point(size = 0.8, alpha = 0.7) +
  scale_x_continuous(breaks = seq(0, 7, by = 1)) +
  scale_y_continuous(labels = percent_format()) +
  labs(title = "Escalation rate by day-of-week", x = "Day of Week", y = "Rate") +
  facet_wrap(~ year, ncol = 1, scales = "free_x") +
  theme_minimal()
```

### Hour, Working Hour and Time Bracket

```{r hour-vs-target}
vehstops |>
  filter(!is.na(arrested), !is.na(hour), !is.na(year)) |>
  group_by(year, hour) |>
  summarise(rate = mean(arrested), n = n(), .groups = "drop") |>
  ggplot(aes(hour, rate, group = 1)) +
  geom_line(linewidth = 0.6) +
  geom_point(size = 0.8, alpha = 0.7) +
  scale_x_continuous(breaks = seq(0, 23, by = 1)) +
  scale_y_continuous(labels = percent_format()) +
  labs(title = "Escalation rate by Hour", x = "Hour", y = "Rate") +
  facet_wrap(~ year, ncol = 1, scales = "free_x") +
  theme_minimal()
```

```{r}
ratio_by_cat(
  vehstops, arrested, time_bracket,
  "Time bracket", "Share of stops", "Escalation ratio by time bracket",
  style = "stack"
)
```

```{r}
ratio_by_cat(
  vehstops, arrested, working_hour,
  "Working Hour", "Share of stops", "Escalation ratio by working hour",
  style = "stack"
)
```

### Weekend, Holiday and Event Day

```{r}
vehstops %>%
  filter(!is.na(holiday), !is.na(arrested)) %>%
  mutate(
    holiday = factor(holiday, levels = c(0, 1), labels = c("No holiday", "Holiday")),
    escal   = if_else(arrested %in% c(1, TRUE), "Escalated", "Not escalated")
  ) %>%
  count(holiday, escal, name = "n") %>%
  group_by(holiday) %>% mutate(total = sum(n)) %>% ungroup() %>%
  ggplot(aes(x = n, y = fct_reorder(holiday, total), fill = escal)) +
  geom_col(width = 0.8) +
  labs(x = "Stops", y = NULL, fill = "Outcome",
       title = "Stops by holiday (escalated vs not)")
```

## Motorist related features

### Reported Age

```{r age-by-target}
vehstops %>%
  filter(!is.na(age), between(age, 16, 100)) %>%
  mutate(bin = ntile(age, 20)) %>%
  group_by(bin) %>%
  summarize(
    age_mid = median(age, na.rm = TRUE),
    rate = mean(arrested == 1, na.rm = TRUE),
    n = n(), .groups = "drop"
  ) %>%
  mutate(
    se = sqrt(rate * (1 - rate) / n),
    lo = pmax(rate - 1.96 * se, 0),
    hi = pmin(rate + 1.96 * se, 1)
  ) %>%
  ggplot(aes(age_mid, rate)) +
  geom_point(size = 1.5) +
  geom_line(linewidth = 0.6) +
  geom_ribbon(aes(ymin = lo, ymax = hi), alpha = 0.15) +
  scale_y_continuous(labels = percent) +
  labs(x = "Age (bin median)", y = "Event rate", title = "Escalation rate vs age (deciles)")
```

```{r age-bracket-vs-target}
ratio_by_cat(
  vehstops, arrested, age_bracket,
  "Reported Age Bracket", "Share of stops", "Escalation ratio by reported age bracket",
  style = "stack"
)
```

### Reported Sex

```{r sex-vs-target}
ratio_by_cat(
  vehstops, arrested, sex,
  "Reported sex", "Share of stops", "Escalation ratio by reported sex",
  style = "stack"
)
```

### Reported Race

```{r race-vs-target}
ratio_by_cat(
  vehstops, arrested, race,
  "Reported race", "Share of stops", "Escalation ratio by reported race",
  style = "stack"
)
```

## Vehicle related features vs Target

------------------------------------------------------------------------

# Multivariate EDA

Since multivariate analysis is very complex, and we just want to get rid of some extra features, we are going to look at some of the percentages, with respect to escalation, of multiple variables (ex: sex x race)

## Multicollinearity (binary block)

## Two-way event-rate heatmaps (interactions)

```{r two-way-heatmaps}
boro_order <- vehstops %>%
  group_by(boro) %>%
  summarize(rate = mean(arrested == 1, na.rm = TRUE), .groups = "drop") %>%
  arrange(rate) %>%
  pull(boro)

vehstops %>%
  filter(!is.na(boro)) %>%
  group_by(boro, hour) %>%
  summarize(rate = mean(arrested == 1, na.rm = TRUE), n = n(), .groups = "drop") %>%
  mutate(boro = factor(boro, levels = boro_order)) %>%
  ggplot(aes(hour, boro, fill = rate)) +
  geom_tile(width = 1, height = 1) +
  scale_fill_viridis_c(labels = scales::percent) +
  labs(x = "Hour", y = "Borough", fill = "Rate",
       title = "Escalation rate by borough × hour")

# 2) Sex × Race
vehstops %>%
  filter(!is.na(sex), !is.na(race)) %>%
  group_by(sex, race) %>%
  summarize(rate = mean(arrested == 1, na.rm = TRUE), n = n(), .groups = "drop") %>%
  mutate(race = fct_lump_n(race, n = 8, other_level = "Other")) %>%
  ggplot(aes(race, sex, fill = rate)) +
  geom_tile(width = 1, height = 1) +
  scale_fill_viridis_c(labels = scales::percent) +
  labs(x = "Race (lumped)", y = "Sex", fill = "Rate",
       title = "Escalation rate across sex × race")
```

### Space × Time interaction (gridded)

```{r space-time}
vehstops %>%
  filter(!is.na(latitude), !is.na(longitude),
         between(latitude, 40.48, 40.92), between(longitude, -74.27, -73.68)) %>%
  mutate(hour_bin = cut(hour, breaks = c(-1,5,11,17,23), labels = c("Night","Morning","Afternoon","Evening"))) %>%
  group_by(hour_bin) %>%
  sample_n(size = min(n(), 200000), replace = FALSE) %>%  # cap points per facet for speed
  ggplot(aes(longitude, latitude)) +
  stat_summary_2d(aes(z = arrested), fun = ~ mean(.x == 1, na.rm = TRUE), bins = 50) +
  facet_wrap(~ hour_bin) +
  coord_equal() +
  scale_fill_viridis_c(labels = percent) +
  labs(x = "Longitude", y = "Latitude", fill = "Rate",
       title = "Escalation rate across space by time-of-day")
```

## Enforcement and Escalation features vs Target

## Command Unit, Geospatial and Area related Features vs Target

Per command unit amount of stops during time (per year, per month, per day)

Per command unit escalation rate during time (per year, per month, per day)

### Area

```{r}
ratio_by_cat_split(
  DF = vehstops,
  TARGET = arrested,
  CAT    = area,
  SPLIT  = weekend,
  x_title = "Area",
  y_title = "Share of stops",
  title   = "Escalation ratio by area (split by weekend)",
  target_labels = c("Not escalated","Escalated"),
  split_labels  = c("Weekday","Weekend")
)

```

```{r area-x-hour}
area_order <- vehstops %>%
  group_by(area) %>%
  summarize(rate = mean(arrested == 1, na.rm = TRUE), .groups = "drop") %>%
  arrange(rate) %>%
  pull(area)

vehstops %>%
  filter(!is.na(area)) %>%
  group_by(area, hour) %>%
  summarize(rate = mean(arrested == 1, na.rm = TRUE), n = n(), .groups = "drop") %>%
  mutate(area = factor(area, levels = area_order)) %>%
  ggplot(aes(hour, area, fill = rate)) +
  geom_tile(width = 1, height = 1) +
  scale_fill_viridis_c(labels = scales::percent) +
  labs(x = "Hour", y = "Area", fill = "Rate",
       title = "Escalation rate by area × hour")
```

```{r area-x-workinghour}
ratio_by_cat_split(
  DF = vehstops,
  TARGET = arrested,
  CAT    = area,
  SPLIT  = working_hour,
  x_title = "Area",
  y_title = "Share of stops",
  title   = "Escalation ratio by area (split by working hour)",
  target_labels = c("Not escalated","Escalated"),
  split_labels  = c("Non Working Hour","Working Hour")
)

```

```{r area-x-sex}
ratio_by_cat_split(
  DF = vehstops,
  TARGET = arrested,
  CAT    = area,
  SPLIT  = sex,
  x_title = "Area",
  y_title = "Share of stops",
  title   = "Escalation ratio by area (split by reported sex)",
  target_labels = c("Not escalated","Escalated"),
  split_labels  = c("No","Yes")
)
```

```{r}
# heat_target_catcat(
#   vehstops, Y = boro, X = time_bracket,
#   X_TITLE = "Time bracket", Y_TITLE = "Borough",
#   TITLE = "Escalation rate by borough × time bracket",
#   min_n = 30, order_y_by = "rate", order_x_by = "freq", alpha_by_n = TRUE
# )
```

```{r}
area_order <- vehstops %>%
  group_by(area) %>%
  summarize(rate = mean(arrested == 1, na.rm = TRUE), .groups = "drop") %>%
  arrange(rate) %>%
  pull(area)

vehstops %>%
  filter(!is.na(area)) %>%
  group_by(area, month) %>%
  summarize(rate = mean(arrested == 1, na.rm = TRUE), n = n(), .groups = "drop") %>%
  mutate(area = factor(area, levels = area_order)) %>%
  ggplot(aes(month, area, fill = rate)) +
  geom_tile(width = 1, height = 1) +
  scale_fill_viridis_c(labels = scales::percent) +
  labs(x = "Month", y = "Area", fill = "Rate",
       title = "Escalation rate by area × month")
```

Per area amount of stops during time (per year, per month, per day)

Per area escalation rate during time (per year, per month, per day)

```{r area-by-weekend-holiday-event-vs-target}
p1 <- ratio_by_cat_split(
  DF = vehstops,
  TARGET = arrested,
  CAT    = area,
  SPLIT  = weekend,
  x_title = "Time bracket",
  y_title = "Share of stops",
  title   = "Escalation ratio by area (split by weekend)",
  target_labels = c("Not escalated","Escalated"),
  split_labels  = c("Weekday","Weekend")
)

p2 <- ratio_by_cat_split(
  DF = vehstops,
  TARGET = arrested,
  CAT    = area,
  SPLIT  = holiday,
  x_title = "Time bracket",
  y_title = "Share of stops",
  title   = "Escalation ratio by area (split by holiday)",
  target_labels = c("Not escalated","Escalated"),
  split_labels  = c("Regular Day","Holiday")
)

p3 <- ratio_by_cat_split(
  DF = vehstops,
  TARGET = arrested,
  CAT    = area,
  SPLIT  = event_day,
  x_title = "Time bracket",
  y_title = "Share of stops",
  title   = "Escalation ratio by area (split by event_day)",
  target_labels = c("Not escalated","Escalated"),
  split_labels  = c("Regular Day","Event Day")
)

p1
p2
p3
```

### Borough

```{r boro-x-year-escalation}
boro_order <- vehstops %>%
  group_by(boro) %>%
  summarize(rate = mean(arrested == 1, na.rm = TRUE), .groups = "drop") %>%
  arrange(rate) %>%
  pull(boro)

vehstops %>%
  filter(!is.na(boro)) %>%
  group_by(boro, year) %>%
  summarize(rate = mean(arrested == 1, na.rm = TRUE), n = n(), .groups = "drop") %>%
  mutate(boro = factor(boro, levels = boro_order)) %>%
  ggplot(aes(year, boro, fill = rate)) +
  geom_tile(width = 1, height = 1) +
  scale_fill_viridis_c(labels = scales::percent) +
  labs(x = "Year", y = "Borough", fill = "Rate",
       title = "Escalation rate by borough × year")


```

```{r boro-x-month-escalation}
boro_order <- vehstops %>%
  group_by(boro) %>%
  summarize(rate = mean(arrested == 1, na.rm = TRUE), .groups = "drop") %>%
  arrange(rate) %>%
  pull(boro)

vehstops %>%
  filter(!is.na(boro)) %>%
  group_by(boro, month) %>%
  summarize(rate = mean(arrested == 1, na.rm = TRUE), n = n(), .groups = "drop") %>%
  mutate(boro = factor(boro, levels = boro_order)) %>%
  ggplot(aes(month, boro, fill = rate)) +
  geom_tile(width = 1, height = 1) +
  scale_fill_viridis_c(labels = scales::percent) +
  labs(x = "Month", y = "Borough", fill = "Rate",
       title = "Escalation rate by borough × month")
```

```{r boro-x-hour-escalation}
boro_order <- vehstops %>%
  group_by(boro) %>%
  summarize(rate = mean(arrested == 1, na.rm = TRUE), .groups = "drop") %>%
  arrange(rate) %>%
  pull(boro)

vehstops %>%
  filter(!is.na(boro)) %>%
  group_by(boro, hour) %>%
  summarize(rate = mean(arrested == 1, na.rm = TRUE), n = n(), .groups = "drop") %>%
  mutate(boro = factor(boro, levels = boro_order)) %>%
  ggplot(aes(hour, boro, fill = rate)) +
  geom_tile(width = 1, height = 1) +
  scale_fill_viridis_c(labels = scales::percent) +
  labs(x = "Hour", y = "Borough", fill = "Rate",
       title = "Escalation rate by borough × hour")
```

### Command Unit

Per command unit amount of stops during time (per year, per month, per day)

Per command unit escalation rate during time (per year, per month, per day)

```{r}
# # Borough × Time bracket (both categorical), hide very sparse cells
# p1 <- heat_target_catcat(
#   vehstops, Y = boro, X = time_bracket,
#   X_TITLE = "Time bracket", Y_TITLE = "Borough",
#   TITLE = "Escalation rate by borough × time bracket",
#   min_n = 30, order_y_by = "rate", order_x_by = "freq", alpha_by_n = TRUE
# )
# 
# # Cleaned vehicle category × Weekend (0/1 already labeled)
# p2 <- heat_target_catcat(
#   vehstops, Y = race, X = sex,
#   "Sex", "Race",
#   "Rate by race × sex",
#   min_n = 50, order_y_by = "freq", order_x_by = "freq"
# )
# 
# 
# p1
# p2

```

```{r}
# heat_target_catcat(
#   vehstops, Y = race, X = sex,
#   "Sex", "Race",
#   "Rate by race × sex",
#   min_n = 50, order_y_by = "freq", order_x_by = "freq"
# )
```

Unknown and Unknown matter somehow!!!

```{r}
# # Borough × hour (hide cells with < 50 stops)
# p_boro_hour <- heat_escalation_yx(
#   DF = vehstops, Y = boro, X = hour,
#   X_TITLE = "Hour of day", Y_TITLE = "Borough",
#   TITLE = "Escalation rate by borough × hour",
#   min_n = 50
# )
# 
# # Area × day-of-week (X is numeric 1–7)
# p_area_dow <- heat_escalation_yx(
#   vehstops, Y = area, X = dow,
#   "Day of week (1–7)", "Area",
#   "Escalation rate by area × DOW", min_n = 25
# )
# 
# # If X is not integer-spaced (e.g., minute) treat as factor to make equal-width tiles
# p_boro_min <- heat_escalation_yx(
#   vehstops, Y = boro, X = minute,
#   "Minute", "Borough", "Rate by borough × minute",
#   x_as_factor = TRUE
# )

```

```{r}
p_boro_hour
p_area_dow
p_boro_min
```

## Time related features

### Hour

#### Grouped by Month

```{r month-lines}
# TODO: use? Move to multivariate
vehstops %>%
  filter(year %in% c(2023, 2024, 2025)) %>%
  group_by(year, month) %>%
  summarize(stops = n(), .groups = "drop") %>%
  ggplot(aes(month, stops, group = factor(year), linetype = factor(year))) +
  geom_line(linewidth = 0.6) + geom_point(size = 1.5) +
  scale_x_continuous(breaks = 1:12) +
  labs(x = "Month", y = "Stops", linetype = "Year",
       title = "Monthly stops: 2023–2025")
```

```{r hour-vs-stops-by-month}
vehstops %>%
  filter(month %in% c(1,2,3,4,5,6,7,8,9,10,11,12)) %>%
  group_by(month, hour) %>%
  summarize(stops = n(), .groups = "drop") %>%
  ggplot(aes(hour, stops, group = factor(month), linetype = factor(month))) +
  geom_line(linewidth = 0.6) + geom_point(size = 1.5) +
  scale_x_continuous(breaks = 0:23) +
  labs(x = "Hour", y = "Stops", linetype = "Month",
       title = "Count of Hourly Stops by Month")
```

```{r hour-vs-dow-by-month}
vehstops %>%
  group_by(dow, hour) %>%
  summarize(stops = n(), .groups = "drop") %>%
  ggplot(aes(hour, stops, group = factor(dow), linetype = factor(dow))) +
  geom_line(linewidth = 0.6) + geom_point(size = 1.5) +
  scale_x_continuous(breaks = 0:23) +
  labs(x = "Hour", y = "Stops", linetype = "Day of the Week",
       title = "Count of Hourly Stops by Day of the Week")
```


```{r hour-vs-target-by-year}
vehstops %>%
  mutate(date = ymd(paste(year, month, day, sep = "-"))) %>%
  group_by(month, hour) %>%
  summarize(rate = mean(arrested == 1, na.rm = TRUE), n = n(), .groups = "drop") %>%
  ggplot(aes(hour, rate, group = factor(month), linetype = factor(month))) +
  geom_line(linewidth = 0.6) + geom_point(aes(size = n)) +
  scale_y_continuous(labels = percent) +
  scale_x_continuous(breaks = 0:23) +
  labs(x = "Hour", y = "Event rate", linetype = "Hour", size = "Stops count",
       title = "Hourly Escalation Rate by Month")
```

```{r dow-hour-heatmap}
vehstops %>%
  filter(!is.na(dow)) %>%
  group_by(dow, hour) %>%
  summarize(rate = mean(arrested == 1, na.rm = TRUE), n = n(), .groups = "drop") %>%
  ggplot(aes(hour, dow, fill = rate)) +
  geom_tile(width = 1, height = 1) +
  scale_fill_viridis_c(labels = percent) +
  scale_x_continuous(breaks = seq(0, 23, 1)) +
  scale_y_continuous(breaks = seq(0, 7, 1)) +
  labs(x = "Hour", y = "Day of week", fill = "Event rate",
       title = "Escalation rate by day and hour")
```

#### Grouped by Year

```{r hour-vs-stops-by-year}
vehstops %>%
  filter(year %in% c(2023, 2024, 2025)) %>%
  group_by(year, hour) %>%
  summarize(stops = n(), .groups = "drop") %>%
  ggplot(aes(hour, stops, group = factor(year), linetype = factor(year))) +
  geom_line(linewidth = 0.6) + geom_point(size = 1.5) +
  scale_x_continuous(breaks = 0:23) +
  labs(x = "Hour", y = "Stops", linetype = "Year",
       title = "Count of Hourly stops by Year")
```

```{r hour-vs-any-escalation-by-year}
vehstops %>%
  mutate(date = ymd(paste(year, month, day, sep = "-"))) %>%
  group_by(year, hour) %>%
  summarize(rate = mean(arrested == 1, na.rm = TRUE), n = n(), .groups = "drop") %>%
  ggplot(aes(hour, rate, group = factor(year), linetype = factor(year))) +
  geom_line(linewidth = 0.6) + geom_point(aes(size = n)) +
  scale_y_continuous(labels = percent) +
  scale_x_continuous(breaks = 0:23) +
  labs(x = "Hour", y = "Event rate", linetype = "Year", size = "Stops count",
       title = "Hourly Escalation Rate by Year")
```

#### Grouped by Day

```{r}
vehstops %>%
  filter(!is.na(dow)) %>%
  group_by(dow, hour) %>%
  summarize(rate = mean(arrested == 1, na.rm = TRUE), n = n(), .groups = "drop") %>%
  ggplot(aes(hour, dow, fill = rate)) +
  geom_tile(width = 1, height = 1) +
  scale_fill_viridis_c(labels = percent) +
  scale_x_continuous(breaks = seq(0, 23, 1)) +
  scale_y_continuous(breaks = seq(0, 7, 1)) +
  labs(x = "Hour", y = "Day of week", fill = "Event rate",
       title = "Escalation rate by day and hour")
```

### Day

```{r target-rate-over-time, message=FALSE}
vehstop_daily <- vehstops %>%
  mutate(date = ymd(paste(year, month, day, sep = "-"))) %>%
  group_by(date) %>%
  summarize(rate = mean(arrested == 1, na.rm = TRUE), n = n(), .groups = "drop")

ggplot(vehstop_daily, aes(date, rate)) +
  geom_line(linewidth = 0.6) +
  geom_point(aes(size = n), alpha = 0.6) +
  geom_smooth(se = FALSE, method = "loess", span = 0.2) +
  scale_y_continuous(labels = percent) +
  labs(x = NULL, y = "Event rate", size = "N/day", title = "Daily escalation rate over time")
```

## Motorist related features

### Reported Age

```{r}
ratio_by_cat_split(
  DF = vehstops,
  TARGET = arrested,
  CAT    = age_bracket,
  SPLIT  = weekend,
  x_title = "Reported Age bracket",
  y_title = "Share of stops",
  title   = "Escalation ratio by reported age (split by weekend)",
  target_labels = c("Not escalated","Escalated"),
  split_labels  = c("Weekday","Weekend")
)
```

### Reported Sex

### Reported Race


## Vehicle related features



------------------------------------------------------------------------


