---
title: "Predicting Excessive Enforcement Patterns in Police Behavior After Vehicles Stops in New York"
subtitle: "EDA and Feature Engineering - 4 Wheels Vehicles"
author: "Gianmaria Pizzo"
output:
  html_document:
    fig_caption: yes
    theme: flatly
    highlight: pygments
    code_folding: show
    kee_md: yes
    toc: yes
    toc_depth: 4
    number_sections: no
    toc_float:
      smooth_scroll: no
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

These `.Rmd` files represent the project submission for the course [Statistical Inference and Learning](https://www.unive.it/data/course/521944) by Professor [Cristiano Varin](https://www.unive.it/data/people/5592728) at [Ca' Foscari University of Venice](https://www.unive.it/) for the year 2025/2026.

# Problem Statement

## Goal of this rmarkdown file

This notebook serves as the primary Exploratory Data Analysis (EDA) and (final) Feature Engineering phase for the project. In particular, we focus solely on **2-Wheeled Vehicles** (including e-bikes, motorcycles, and mopeds).

The reason for this is the fact that, these vehicles, have been the subject a of 2.5x increase in stop volume in NYC between 2023 and 2025 (which will be later shown). Furthermore, they show a completely separate behavior with respect to vehicles with four wheels (due to environmental factors like temperature and weather).

The goal is to identify the features that characterize **arrests**.

We approach the dataset through three methodological phases to ensure a robust base for modeling:

1.  **Univariate Analysis**. We begin by analyzing the distributions of individual predictors to understand whether the data shows quality issues, outliers, rare categories, etc...

2.  **Bivariate Analysis**. We evaluate the relationship between potential predictors and our binary target (`arrested`). Given the categorical nature of our data, we will mainly uses **Cramer’s V** as a starting point to measure of effect size. This stage should also point out if predictors have some kind of relation.

Under these premises, we will plot many aspects of the data. Therefore, the way we decided to guide the reader through the Exploratory Data Analysis, is based on four key points that we try to include for any plot:

- **Key finding**: This paragraph describes the main takeaway of the means of visualization employed. This should be a concise description of what we can gather from the plots.
- **Interpretation**: Here we deal with an objective interpretation of our finding, such as understanding how our EDA is affected by . We also detail about how this affects our current framework and
- **Comment**: After a formal description, we deal with a more informal discussion that should highlight the qualitative aspects of the analysis. This paragraph is a fusion of previous considerations that we want the reader to be aware of, and a space for us to reason.
- **Proposed Action**: Finally, we switch to a more pragmatic discussion to understand "what to do next", on the basis of what was previously presented. Without claim of completeness, we will name some best-practices, guidelines or ideas about what is best for our goal. We may also limit ourselves to inform the reader that we decided to take no action at all or post-pone it after further investigation.

Note that we do not always include all four points.

------------------------------------------------------------------------

# Environment

The first to do, is to build a reproducible environment for our project.

```{r}
requirements=c( 
  "remotes", "summarytools", "forcats", "scales", "purrr",
  "stringr", "plyr", "dbplyr", "dplyr", "dtplyr", "tidyr", "tidyselect",
  "timeDate", "timechange", "tzdb", "hms", "readr", "tigris", "lsr", "corrplot",
  "lubridate", "prettyunits", "RColorBrewer", "viridis", "rcartocolor", "ggbeeswarm", "emmeans",
  "sf", "plotly", "ggplot2", "ggmap", "ggdensity", "UpSetR", "patchwork"
)
```

```{r install-requirements, include=FALSE}
for (req in requirements){
  if (!require(req, character.only = TRUE)){
      install.packages(req)
  }
}
```

Note that some of the packages may not be used, but to avoid any error related to dependencies, we are going to keep all of these libraries. Now, the loading of all packages should be immediate.

```{r load-requirements, include=FALSE}
for (req in requirements){
  if (!require(req, character.only = TRUE)){
      library(req, character.only = TRUE)
  }
}
```

```{r remove-vars, include=FALSE}
rm(requirements, req)
```

## Plot Functions

Here we add the plot functions to make some kinds of visualizations reproducible.

### Univariate Plot Functions

```{r univariate-plot-helpers}
# ==============================================================================
# 1. Plots for Univariate Analysis
# ==============================================================================

#' @title Plot Binary Bar Chart
#' 
#' @description Creates a simple bar chart counting the occurrences of a binary feature, 
#' with labels showing both the count and the percentage.
#' 
#' @param DF The input dataframe.
#' @param FEATURE The binary column to plot (unquoted).
#' @param OPTION_0 The text label for the "0" or "FALSE" category.
#' @param OPTION_1 The text label for the "1" or "TRUE" category.
#' @param X_LABEL Label for the X axis.
#' @param Y_LABEL Label for the Y axis.
#' @param TITLE Main title of the plot.
#' @param pad Numeric padding for the top of the Y axis to fit text labels.
#' @param text_vjust Vertical adjustment for the text labels on top of bars.
#' 
#' @return A ggplot2 object.
plot_binary_bar <- function(DF, FEATURE, OPTION_0, OPTION_1, X_LABEL, Y_LABEL, TITLE, pad = 0.15, text_vjust = -0.3) {
  # Standardize the binary feature and calculate percentages
  df_sum <- DF %>%
    mutate(
      .raw  = {{ FEATURE }},
      .feat = dplyr::case_when(
        as.character(.raw) %in% c("1","TRUE","true")  ~ 1L,
        as.character(.raw) %in% c("0","FALSE","false") ~ 0L,
        TRUE ~ NA_integer_
      )
    ) %>%
    filter(!is.na(.feat)) %>%
    count(.feat, name = "n") %>%
    mutate(
      .lab = factor(.feat, levels = c(0, 1), labels = c(OPTION_0, OPTION_1)),
      # Calculate percentage and create combined label
      pct = n / sum(n),
      label_text = paste0(scales::comma(n), "\n(", scales::percent(pct, accuracy = 0.1), ")")
    )
  
  # Get the maximum bar height
  ymax <- max(df_sum$n, na.rm = TRUE)
  
  # Build the ggplot
  ggplot(df_sum, aes(x = .lab, y = n, fill = .lab)) +
    geom_col(width = 0.8, show.legend = FALSE) +
    # Updated to use label_text
    geom_text(aes(label = label_text), vjust = text_vjust, lineheight = 0.8) +
    # Expand the Y axis slightly for the taller multi-line labels
    scale_y_continuous(limits = c(0, ymax * (1 + pad)), expand = expansion(mult = c(0, 0.02))) +
    coord_cartesian(clip = "off") +
    theme_minimal() +
    theme(panel.grid.major.x = element_blank(), plot.margin = margin(t = 20, r = 5.5, b = 5.5, l = 5.5)) +
    scale_fill_viridis_d(option = "mako", begin = 0.3, end = 0.8) +
    labs(x = X_LABEL, y = Y_LABEL, title = TITLE)
}


#' @title Plot Univariate Bar Chart (Categorical)
#' 
#' @description Creates a bar chart counting the occurrences of a categorical feature with 2+ classes.
#' Allows dynamic ordering (by frequency or original/alphabetical order).
#' 
#' @param DF The input dataframe.
#' @param FEATURE The categorical column to analyze (unquoted).
#' @param X_LABEL Label for the X axis.
#' @param Y_LABEL Label for the Y axis (default is "Count").
#' @param TITLE Main title of the plot.
#' @param SUBTITLE Subtitle of the plot.
#' @param top_n Optional integer; if set, keeps only the top N most frequent categories.
#' @param order_by Method to order the X axis: "freq" (descending count) or "none" (alphabetical/original).
#' @param pad Numeric padding for the top of the Y axis to fit text labels.
#' @param text_vjust Vertical adjustment for the text labels on top of bars.
#' 
#' @return A ggplot2 object.
plot_univariate_cat <- function(DF, FEATURE, X_LABEL, Y_LABEL = "Count", TITLE, SUBTITLE, top_n = NULL, order_by = c("freq", "none"), pad = 0.12, text_vjust = -0.35) {
  
  order_by <- match.arg(order_by)
  
  # Compute counts
  df_sum <- DF %>%
    filter(!is.na({{ FEATURE }})) %>%
    count({{ FEATURE }}, name = "n")
  
  # Filter top n results
  if (!is.null(top_n)) {
    df_sum <- df_sum %>% slice_max(order_by = n, n = top_n)
  }
  
  # X axis order (either frequency or std)
  cat_levels <- switch(
    order_by,
    freq = df_sum %>% arrange(desc(n)) %>% pull({{ FEATURE }}),
    none = sort(unique(df_sum %>% pull({{ FEATURE }})))
  )
  
  # Apply sorting
  df_sum <- df_sum %>%
    mutate(.feat_f = factor(as.character({{ FEATURE }}), levels = as.character(cat_levels)))
  
  # Max height for dynamic padding
  ymax <- max(df_sum$n, na.rm = TRUE)

  # Construct plot
  ggplot(df_sum, aes(x = .feat_f, y = n, fill = .feat_f)) +
    geom_col(width = 0.8, show.legend = FALSE) +
    geom_text(aes(label = scales::comma(n)), vjust = text_vjust, size = 3) +
    scale_y_continuous(limits = c(0, ymax * (1 + pad)),
                       expand = expansion(mult = c(0, 0.02))) +
    coord_cartesian(clip = "off") +
    theme_minimal() +
    theme(
      plot.title = element_text(size = 14),
      plot.subtitle = element_text(size = 10, color = "grey50"),
      axis.text.x = element_text(angle = 45, hjust = 1),
      panel.grid.major.x = element_blank(),
      plot.margin = margin(t = 10, r = 5.5, b = 5.5, l = 5.5)) +
    scale_fill_viridis_d(option = "mako", begin = 0.3, end = 0.8) +
    labs(x = X_LABEL, y = Y_LABEL, title = TITLE, subtitle = SUBTITLE)
}
```

### Bivariate Plot Functions

```{r biviariate-plot-helpers}
# ==============================================================================
# 2. Plots for Bivariate Analysis
# ==============================================================================

#' @title Calculate Cramer's V
#' 
#' @description Calculates the Cramer's V measure of association between two 
#' categorical vectors. Returns a value between 0 (no association) and 1 (perfect association).
#' 
#' @param x First categorical vector.
#' @param y Second categorical vector.
#' 
#' @return A numeric value representing Cramer's V.
cramer_v <- function(x, y) {
  # Contingency Table
  tbl <- table(x, y)
  
  # Execute Chi2 Test
  chi2 <- suppressWarnings(chisq.test(tbl, correct = FALSE)$statistic)
  
  # Compute dim and min value
  n <- sum(tbl)
  min_dim <- min(nrow(tbl) - 1, ncol(tbl) - 1)
  
  # Avoid math error for 1 categ
  if(min_dim == 0) return(0) 
  
  # Cramer's formula
  v <- sqrt(chi2 / (n * min_dim))
  return(as.numeric(v))
}

# ==============================================================================
# Categorical "Correlation" Heatmap - Association Matrix based on Cramer's V
# ==============================================================================

#' @title Plot Categorical Correlation Matrix (Cramer's V)
#' 
#' @description Creates a correlation-like heatmap for multiple categorical variables
#' using Cramer's V. Answers "How much do these categorical features overlap with each other?"
#' 
#' @param DF The input dataframe.
#' @param FEATURES A character vector containing the names of the categorical columns to compare.
#' @param TITLE Main title of the plot.
#' 
#' @return A ggplot2 object.
plot_categorical_correlation <- function(DF, FEATURES, TITLE = "Association Matrix for Categorical Predictors (Cramer's V)", text_size = 3) {
  
  # Select columns and drop NAs
  df_sub <- DF %>% 
    select(all_of(FEATURES)) %>%
    drop_na()
  
  n_vars <- ncol(df_sub)
  
  # Initialize matrix and set main diag to 1s
  mat <- matrix(1, nrow = n_vars, ncol = n_vars)
  rownames(mat) <- colnames(df_sub)
  colnames(mat) <- colnames(df_sub)

  # Compute Cramer's V for each possible pair
  for(i in 1:(n_vars - 1)) {
    for(j in (i + 1):n_vars) {
      v <- cramer_v(df_sub[[i]], df_sub[[j]])
      # Symmetric 
      mat[i, j] <- v
      mat[j, i] <- v
    }
  }
  
  # Make matrix compatible with ggplot2
  mat_long <- as.data.frame(as.table(mat))
  colnames(mat_long) <- c("Var1", "Var2", "CramerV")
  
  # Create heatmap
  ggplot(mat_long, aes(x = Var1, y = Var2, fill = CramerV)) +
    geom_tile(color = "white", width = 1, height = 1) +
    # Rounding
    geom_text(aes(label = round(CramerV, 2)), size = text_size) +
    # Setting colors
    scale_fill_gradient(low = "whitesmoke", high = "steelblue", limits = c(0, 1), name = "Cramer's V") +
    labs(title = TITLE, x = "", y = "") +
    theme_minimal() +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1, size = 10),
      axis.text.y = element_text(size = 10),
      panel.grid = element_blank()
    )
}

# ==============================================================================
# Plotting of Targets' Rate Against Categorical Predictors
# ==============================================================================

#' @title Plot Target Ratio by Category
#' 
#' @description Creates a bivariate bar chart showing the relationship between a binary 
#' TARGET and a single categorical variable (CAT). It supports two visual styles: 
#' a 100% stacked bar chart ("stack") or a simple bar chart showing the rate of the target ("rate").
#' 
#' @param DF The input dataframe.
#' @param TARGET The binary column to analyze (unquoted). It will be standardized to 0/1.
#' @param CAT The categorical column for the X axis (unquoted).
#' @param x_title Label for the X axis.
#' @param y_title Label for the Y axis.
#' @param title Main title of the plot.
#' @param target_labels A character vector of length 2 defining the labels for 0 and 1 (e.g., c("Not escalated", "Escalated")).
#' @param style The visual style of the plot: "stack" (100% stacked bars) or "rate" (single bar for target = 1).
#' @param order_by Method to order the X axis categories: by "rate" (ascending target rate), "freq" (ascending frequency), or "none" (alphabetical/original).
#' @param percent_accuracy Accuracy for the percent labels (e.g., 0.1 for one decimal place).
#' @param colors A character vector of length 2 defining the hex color codes for the 0 and 1 categories respectively.
#' 
#' @return A ggplot2 object.
ratio_by_cat <- function(
    DF, TARGET, CAT, x_title, y_title, title, 
    target_labels = c("Not escalated", "Escalated"), 
    style = c("stack", "rate"), order_by = c("rate", "freq", "none"), 
    percent_accuracy = 0.1, palette = "Paired",
    colors = c("grey80", "steelblue"), legend_title = "Legend",
    subtitle = NULL, bar_width = 0.7, aspect_ratio = 1.0, label_vjust = 0.5,
    text_size = 2
) {
  
  # Validate arguments
  style    <- match.arg(style)
  order_by <- match.arg(order_by)
  
  # Standardize target and extract the category
  dat <- DF %>%
    mutate(
      .t_raw = {{ TARGET }},
      .t = case_when(
        as.character(.t_raw) %in% c("1","TRUE","true") ~ 1L,
        as.character(.t_raw) %in% c("0","FALSE","false") ~ 0L,
        TRUE ~ NA_integer_
      ),
      .cat = {{ CAT }}
    ) %>%
    filter(!is.na(.t), !is.na(.cat))
  
  # Compute rates and frequencies
  rates <- dat %>% 
    group_by(.cat) %>%
    summarise(total = n(), rate = mean(.t == 1), .groups = "drop")
  
  # Determine ordering
  cat_levels <- switch(
    order_by,
    rate = rates %>% arrange(rate) %>% pull(.cat),
    freq = rates %>% arrange(total) %>% pull(.cat),
    none = unique(dat$.cat)
  )
  
  # Build the plot based on style
  if (style == "stack") {
    stack <- dat %>%
      count(.cat, .t, name = "n") %>%
      group_by(.cat) %>%
      mutate(ratio = n/sum(n)) %>%
      ungroup() %>%
      mutate(
        .cat_f = factor(as.character(.cat), levels = as.character(cat_levels)),
        .t_f = factor(.t, levels = c(0,1), labels = target_labels)
      )
    
    ggplot(stack, aes(x = .cat_f, y = ratio, fill = .t_f)) +
      geom_col(position = "fill", width = bar_width) +
      geom_text(
        aes(label = scales::percent(ratio, accuracy = percent_accuracy)),
        position = position_fill(vjust = label_vjust), 
        size = text_size
      ) +
      scale_y_continuous(labels = scales::percent) +
      scale_fill_manual(values = setNames(colors, target_labels)) +
      labs(x = x_title, y = y_title, fill = legend_title, title = title, subtitle = subtitle) +
      theme_minimal() +
      theme(
        axis.text.x = element_text(angle = 45, hjust = 1),
        panel.grid = element_blank(),
        plot.subtitle = element_text(color = "grey50"),
        aspect.ratio = aspect_ratio
      )
      
  } else {
    # Rate style
    rates_data <- rates %>%
      mutate(.cat_f = factor(as.character(.cat), levels = as.character(cat_levels)))
    
    ggplot(rates_data, aes(x = .cat_f, y = rate)) +
      geom_col(width = 0.8, fill = colors[2]) +
      geom_text(
        aes(label = scales::percent(rate, accuracy = percent_accuracy)),
        vjust = -0.35, size = text_size
      ) +
      scale_y_continuous(labels = scales::percent, expand = expansion(mult = c(0, 0.12))) +
      coord_cartesian(clip = "off") +
      labs(x = x_title, y = y_title, title = title, subtitle = subtitle) +
      theme_minimal() +
      theme(
        axis.text.x = element_text(angle = 45, hjust = 1),
        panel.grid = element_blank(),
        plot.subtitle = element_text(color = "grey50")
      )
  }
}
```

### Multivariate Plot Functions

```{r multivariate-plot-helpers}
# ==============================================================================
# 3. Plots for Multivariate Analysis
# ==============================================================================

#' @title Stacked Ratio Bars Split by a Third Variable
#' 
#' @description Creates a faceted 100% stacked bar chart. This is highly effective 
#' for visualizing how the rate of a binary TARGET changes across a primary category (CAT), 
#' while simultaneously splitting the data by a third variable (SPLIT) to create side-by-side plots.
#' 
#' @param DF The input dataframe.
#' @param TARGET The binary column to analyze (unquoted). It will be standardized to 0/1.
#' @param CAT The main categorical variable for the X axis (unquoted).
#' @param SPLIT The categorical variable used to split/facet the plots (unquoted).
#' @param x_title Label for the X axis.
#' @param y_title Label for the Y axis.
#' @param title Main title of the entire plot.
#' @param target_labels Character vector of length 2 defining labels for the binary outcomes (e.g., c("Not escalated", "Escalated")).
#' @param split_labels Optional character vector to provide custom titles for the split panels (facets).
#' @param palette A character string specifying an RColorBrewer palette name (e.g., "Paired", "Set1").
#' @param percent_accuracy Accuracy for the percentage labels (e.g., 0.1 for one decimal place).
#' 
#' @return A ggplot2 object.
ratio_by_cat_split <- function(DF, TARGET, CAT, SPLIT, x_title, y_title, title, target_labels = c("Not escalated", "Escalated"), split_labels = NULL, palette = "Paired", percent_accuracy = 0.1){
  # Standardize target
  dat <- DF %>%
    mutate(
      .target_raw = {{ TARGET }},
      # Conversion to 0/1
      .target_bin = case_when(
        as.character(.target_raw) %in% c("1", "TRUE", "true") ~ 1L,
        as.character(.target_raw) %in% c("0", "FALSE", "false") ~ 0L,
        TRUE ~ NA_integer_
      ),
      .cat = {{ CAT }},
      .split = {{ SPLIT }}
    ) %>%
    # Filter NAs
    filter(!is.na(.cat), !is.na(.split), !is.na(.target_bin))

  # Order the main categories based on the overall target rate (from lowest to highest)
  cat_order <- dat %>%
    group_by(.cat) %>%
    summarise(
      rate = mean(.target_bin == 1),
      .groups = "drop"
    ) %>%
    arrange(rate) %>% 
    pull(.cat)

  # Tidy labels by converting character to factors
  dat <- dat %>%
    mutate(
      # Apply the order to X axis
      .cat = factor(as.character(.cat), levels = as.character(cat_order)),
      # Apply custom labels to the target
      .target_f = factor(.target_bin, levels = c(0, 1), labels = target_labels),
      # Format the split variable
      .split_f = {
        lv <- unique(as.character(.split))
        # If custom split labels are provided and the length matches, use them
        if (!is.null(split_labels) && length(split_labels) == length(lv))
          factor(as.character(.split), levels = lv, labels = split_labels)
        else factor(as.character(.split), levels = lv)
      }
    )

  # Compute rates
  plot_df <- dat %>%
    count(.split_f, .cat, .target_f, name = "n") %>%
    group_by(.split_f, .cat) %>%
    mutate(ratio = n/sum(n)) %>%
    ungroup()

  # Build plot
  ggplot(plot_df, aes(x = .cat, y = ratio, fill = .target_f)) +
    # Stacks the bars to exactly 100%
    geom_col(position = "fill") +
    # Add ratio text labels
    geom_text(
      aes(label = percent(ratio, accuracy = percent_accuracy)),
      position = position_fill(vjust = 0.5), 
      size = 3
    ) +
    # Format Y axis as percentages and make continuous
    scale_y_continuous(labels = percent) +
    # Use a clean palette for the target categories
    scale_fill_brewer(palette = palette, name = rlang::as_name(rlang::enquo(TARGET))) +
    labs(x = x_title, y = y_title, title = title) +
    # Split the plot into separate panels based on the .split_f variable
    facet_wrap(~ .split_f, nrow = 1) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
          panel.grid = element_blank())
}
```

------------------------------------------------------------------------

Let us load the data now, and remove the extra feature `veh_category` because we are **exclusively analyzing 2 wheels vehicles now**.

```{r load-data-drop-veh-categ}
vehstops = read.csv("data/vehiclestops_2_wheels.csv")

vehstops <- vehstops %>%
  select(-veh_categ)
```

So now we have the following dataset, that we present with the usual summary

```{r summary-data}
print(dfSummary(vehstops), method="render")
```

This preview is already interesting.

------------------------------------------------------------------------

# Univariate EDA

## Target

```{r enforcement-dist}
p1 <- plot_binary_bar(
  vehstops, arrested, 
  "No", "Yes",
  "Was the motorist arrested?", "Count (Rate %)", ""
)

row1 <- p1

row1 + 
plot_annotation(title = "Distribution of 4-Wheels Vehicles Stops by Target") &
theme(legend.position = "bottom") 
```

```{r, include=FALSE}
rm(p1, row1)
```

What can we say?

- **Key finding**:
- **Interpretation**:
- **Comment**: 
- **Proposed Action**:

## Geospatial predictors

### Area

```{r count-stops-by-area}
plot_univariate_cat(
  DF = vehstops, 
  FEATURE = area, 
  X_LABEL = "Area", 
  TITLE = "Distribution of 4-Wheels Vehicles Stops by Area",
  SUBTITLE = "TODO",
  order_by = "freq"
)
```


What can we say?

- **Key finding**:
- **Interpretation**:
- **Comment**: 
- **Proposed Action**:

### Borough

```{r count-stops-by-boro}
plot_univariate_cat(
  DF = vehstops, 
  FEATURE = boro, 
  X_LABEL = "Borough", 
  TITLE = "Distribution of 4-Wheels Vehicles Stops by Borough",
  SUBTITLE = "TODO"
)
```

What can we say?

- **Key finding**:
- **Interpretation**:
- **Comment**: 
- **Proposed Action**:

### Latitude and Longitude for Stop Density

For what concerns these two predictors, we will use the to plot a spatial heat map of the stops with these values available.

```{r stop-density}
nyc_map <- counties(state = "NY", cb = TRUE) %>%
  filter(NAME %in% c("Bronx", "Kings", "New York", "Queens", "Richmond"))

bins <- 100
lon_breaks <- seq(-74.26, -73.71, length.out = bins + 1)
lat_breaks <- seq(40.48, 40.92, length.out = bins + 1)

grid_sum <- vehstops %>%
  filter(!is.na(longitude) & !is.na(latitude)) %>%
  mutate(
    xbin = cut(longitude, breaks = lon_breaks, include.lowest = TRUE, right = FALSE),
    ybin = cut(latitude, breaks = lat_breaks, include.lowest = TRUE, right = FALSE)
  ) %>%
  mutate(
    x = as.numeric(gsub(".*\\[(.*),.*", "\\1", xbin)) + (lon_breaks[2] - lon_breaks[1])/2,
    y = as.numeric(gsub(".*\\[(.*),.*", "\\1", ybin)) + (lat_breaks[2] - lat_breaks[1])/2
  ) %>%
  group_by(x, y) %>%
  summarise(n = n(), .groups = "drop")

ggplot() +
  geom_sf(data = nyc_map, fill = "white", color = "grey80", size = 0.2) +
  geom_tile(data = grid_sum, aes(x, y, fill = n), alpha = 0.8) +
  scale_fill_viridis_c(trans = "log10", name = "Stops (log10)", option = "mako", direction = -1) +
  coord_sf(xlim = c(-74.26, -73.71), ylim = c(40.48, 40.92)) +
  theme_minimal() +
  theme(panel.grid = element_blank()) +
  labs(title = "Heatmap of 4-Wheels Vehicles Stops Density - 100 Bins", x = "Longitude", y = "Latitude")
```

```{r, include=FALSE}
rm(grid_sum, nyc_map, bins, lat_breaks, lon_breaks)
```

What can we say?

- **Key finding**:
- **Interpretation**:
- **Comment**: 
- **Proposed Action**:

## Time-related predictors

### Year

```{r count-stops-by-year}
vehstops %>%
  filter(!is.na(year)) %>%
  count(year) %>%
  mutate(year = as.numeric(as.character(year))) %>%
  arrange(year) %>%
  mutate(
    prev_n = lag(n), pct_change = (n - prev_n) / prev_n * 100,
    label_text = ifelse(
      is.na(pct_change), 
      comma(n), 
      paste0(comma(n), "\n(", ifelse(pct_change > 0, "+", ""), round(pct_change, 1), "%)")
    )) %>%
  ggplot(aes(x = factor(year), y = n, group = 1)) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_point(color = "steelblue", size = 3) +
  geom_text(aes(label = label_text), vjust = -0.8, nudge_x = -0.1, size = 3.5, lineheight = 0.8) +
  scale_y_continuous(labels = comma, limits = c(0, NA), expand = expansion(mult = c(0, 0.2))) +
  theme_minimal() +
  theme(
    panel.grid.major.x = element_blank(), 
    plot.title = element_text(size = 14),
    plot.subtitle = element_text(size = 10, color = "grey50")
    ) +
  labs(
    title = "Annual Volume of Stops for 4-Wheels Vehicles in NYC",
    subtitle = "TODO",
    x = "Year", y = "Count"
    )
```

What can we say?

- **Key finding**:
- **Interpretation**:
- **Comment**: 
- **Proposed Action**:

### Month

```{r count-stops-by-month}
month_data <- vehstops %>%
  filter(!is.na(month)) %>%
  count(month) %>%
  mutate(month = factor(month, levels = 1:12)) %>%
  arrange(month) %>%
  mutate(
    prev_n = lag(n), pct_change = (n - prev_n) / prev_n * 100,
    label_text = ifelse(
      is.na(pct_change), comma(n),
      paste0(comma(n), "\n(", ifelse(pct_change > 0, "+", ""), round(pct_change, 1), "%)"))
  )

month_data %>%
  ggplot(aes(x = month, y = n, group = 1)) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_point(color = "steelblue", size = 3) +
  geom_text(aes(label = label_text), vjust = -1, nudge_x = -0.1, size = 3, lineheight = 0.8) +
  scale_x_discrete(labels = month.abb) +
  scale_y_continuous(labels = comma, limits = c(0, NA), expand = expansion(mult = c(0, 0.3))) +
  theme_minimal() +
  theme(
    panel.grid.major.x = element_blank(),
    plot.title = element_text(size = 14),
    plot.subtitle = element_text(color = "grey50", size = 10) # Styled as requested
  ) +
  labs(
    title = "Monthly Volume of Stops for 4-Wheels Vehicles in NYC",
    subtitle = "TODO",
    x = "Month",
    y = "Count"
  )
```

```{r, include=FALSE}
rm(month_data)
```

What can we say?

- **Key finding**:
- **Interpretation**:
- **Comment**: 
- **Proposed Action**:

Let us see this with respect to the year

```{r year-monthly-distribution}
plot_data <- vehstops %>%
  filter(!is.na(year) & !is.na(month)) %>%
  count(year, month) %>%
  mutate(month = factor(month, levels = 1:12)) %>%
  group_by(year) %>%
  mutate(
    n_jan = n[month == 1],
    n_max = max(n),
    is_max = n == n_max,
    pct_max = round((n - n_jan) / n_jan * 100, 0),
    pct_dec = round((n - n_max) / n_max * 100, 0),
    
    label_text = case_when(
      month == 1 ~ comma(n), 
      is_max ~ paste0(comma(n), "\n(", ifelse(pct_max > 0, "+", ""), pct_max, "%)"),
      month == 12 ~ paste0(comma(n), "\n(", pct_dec, "%)"), # Dec is usually negative, so '-' is automatic
      TRUE ~ ""
    )
  )

# Faceted plot
plot_data %>%
  ggplot(aes(x = month, y = n, group = 1)) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_point(color = "steelblue", size = 2) +
  geom_text(aes(label = label_text), vjust = -1, hjust = 0.2, size = 2.5, lineheight = 0.8) +
  facet_wrap(~year, nrow = 1) +
  scale_x_discrete(
    labels = c("J", "F", "M", "A", "M", "J", "J", "A", "S", "O", "N", "D"), 
    expand = expansion(mult = c(0.1, 0.1))
    ) +
  scale_y_continuous(labels = comma, limits = c(0, NA), expand = expansion(mult = c(0, 0.4))) +
  coord_cartesian(clip = "off") +
  theme_minimal() +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    plot.title = element_text(size = 14),
    plot.subtitle = element_text(color = "grey50", size = 10),
    strip.text = element_text(size = 11)
  ) +
  labs(
    title = "Monthly Volume of Stops, by Year, for 4-Wheels Vehicles in NYC",
    subtitle = "TODO",
    x = "Month",
    y = "Count"
  )
```

```{r, include=FALSE}
rm(plot_data)
```

What can we say?

- **Key finding**:
- **Interpretation**:
- **Comment**: 
- **Proposed Action**:

### Checking out `hot_season`

Let us proceed with this intuition

```{r add-hot-season}
vehstops <- vehstops %>%
  mutate(hot_season = ifelse(month %in% 4:10, 1, 0))
```

Now, for plotting

```{r year-hot-season-distribution}
plot_data <- vehstops %>%
  filter(!is.na(year) & !is.na(month)) %>%
  mutate(
    is_hot_season = ifelse(month %in% 4:10, "Hot Season", "Off Season"),
    is_hot_season = factor(is_hot_season, levels = c("Hot Season", "Off Season"))
  ) %>%
  count(year, is_hot_season) %>%
  group_by(year) %>%
  mutate(
    year_total = sum(n),
    pct = n / year_total,
    label_text = paste0(comma(n), "\n(", round(pct * 100, 1), "%)")
  )

plot_data %>%
  ggplot(aes(x = is_hot_season, y = n, fill = is_hot_season)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~year, nrow = 1) +
  geom_text(aes(label = label_text), vjust = -0.5, size = 3.5, lineheight = 0.8) +
  scale_y_continuous(labels = comma, expand = expansion(mult = c(0, 0.2))) +
  scale_fill_brewer(palette = "Paired") +
  theme_minimal() +
  theme(
    panel.grid.major.x = element_blank(),
    plot.title = element_text(size = 14),
    plot.subtitle = element_text(color = "grey50", size = 10)
  ) +
  labs(
    title = "Volume of Stops for 4-Wheels Vehicles: Hot Season vs. Off Season",
    subtitle = "TODO",
    x = NULL,
    y = "Count"
  )
```

```{r, include=FALSE}
rm(plot_data)
```

What can we say?

- **Key finding**:
- **Interpretation**:
- **Comment**: 
- **Proposed Action**:

### Dow/Weekday, Day and Doy

We have decided to exclude 'Day of Month' (`day`) and 'Day of Year' (`doy`) from our feature set to mitigate 'calendar artifact' bias.

As we can see below, the variation in month length creates false activity dips for end-of-month dates

```{r day-bias}
vehstops %>%
  filter(!is.na(day)) %>%
  count(day) %>%
  ggplot(aes(x = day, y = n)) +
  geom_line(color = "grey60", alpha = 0.5) +
  geom_smooth(method = "loess", color = "steelblue", se = FALSE) +
  theme_minimal() +
   theme(
    panel.grid.major.x = element_blank(),
    plot.title = element_text(size = 14),
    plot.subtitle = element_text(color = "grey50", size = 10)
  ) +
  labs(title = "Volume of Stops by Day of the Month (1–31) for 4-Wheels Vehicles",
       subtitle = "TODO",
       x = "Day of the Month", y = "Count")
```

Furthermore, `doy` variable suffers from inconsistent data density.

```{r doy-incosistent-3-years-span}
plot_data <- vehstops %>%
  mutate(date = as.Date(paste(year, month, day, sep = "-"))) %>%
  count(date)

plot_data %>%
  ggplot(aes(x = date, y = n)) +
  geom_line(color = "grey70", alpha = 0.6, linewidth = 0.5) +
  geom_smooth(method = "loess", span = 0.2, color = "steelblue", se = FALSE, linewidth = 1.2) +
  scale_x_date(date_breaks = "3 months", date_labels = "%b %Y") +
  theme_minimal() +
  theme(
    panel.grid.minor = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(size = 14),
    plot.subtitle = element_text(color = "grey50", size = 10)
  ) +
  labs(
    title = "Volume of Stops by Day of the Year (1–365) for 4-Wheels Vehicles",
    subtitle = "TODO",
    x = "Period", 
    y = "Daily Stop Count"
  )
```

```{r, include=FALSE}
rm(plot_data)
```

While we observe a consistent seasonal clustering peaking in late summer, our current 3-year dataset captures three seasonal cycles. This is sufficient to identify a recurring pattern, but insufficient to establish it as a long-term trend.

The next plot, demonstrates high daily volatility (indicated by the grey jitter) that obscures the underlying seasonal trend (the blue trend line). Because the daily fluctuation is relative to the overall pattern, including 'Day of Year' as a feature risks overfitting.

```{r doy-inconsistent-density}
vehstops %>%
  filter(!is.na(doy)) %>%
  count(doy) %>%
  ggplot(aes(x = doy, y = n)) +
  geom_line(color = "grey60", alpha = 0.5) +
  geom_smooth(method = "loess", color = "steelblue", se = FALSE) +
  theme_minimal() +
   theme(
    panel.grid.major.x = element_blank(),
    plot.title = element_text(size = 14),
    plot.subtitle = element_text(color = "grey50", size = 10)
  ) +
  labs(title = "Volume of Stops by Day of the Year (1–365) for 4-Wheels Vehicles",
       subtitle = "TODO",
       x = "Day of Year", y = "Count")
```

Finally, these granular temporal markers are prone to overfitting. Excluding them allows the model to focus on more robust predictors, such as seasonality and weekly rhythms.

On the other hand, we can study the "Day of the Week" which usually abstract better.

```{r dow-dis}
vehstops %>%
  filter(!is.na(dow)) %>%
  count(dow) %>%
  mutate(
    dow = factor(dow, levels = 1:7, 
                 labels = c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun")),
    pct_change = (n - lag(n)) / lag(n) * 100,
    label_text = ifelse(is.na(pct_change), 
                        comma(n), 
                        paste0(comma(n), "\n(", ifelse(pct_change > 0, "+", ""), round(pct_change, 1), "%)"))
  ) %>%
  ggplot(aes(x = dow, y = n, group = 1)) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_point(color = "steelblue", size = 3) +
  geom_text(aes(label = label_text), vjust = -1, size = 3, lineheight = 0.8) +
  scale_y_continuous(labels = comma, 
                     limits = c(0, NA), 
                     expand = expansion(mult = c(0, 0.4))) + # Extra room for labels
  theme_minimal() +
  theme(
    panel.grid.major.x = element_blank(),
    plot.title = element_text(size = 14),
    plot.subtitle = element_text(color = "grey50", size = 10)
  ) +
  labs(
    title = "Daily Volume of Stops for 4-Wheels Vehicles",
    subtitle = "TODO",
    x = "Day of the Week", 
    y = "Count"
  )
```

What can we say?

- **Key finding**:
- **Interpretation**:
- **Comment**: 
- **Proposed Action**:

### Hour and Time Bracket

First of all we look at the overall cycle of stops volume

```{r hour-dist}
hours_to_show <- c(0, 4, 8, 12, 16, 19, 20, 23)

full_data <- vehstops %>%
  filter(!is.na(hour)) %>%
  count(hour)

highlight_data <- full_data %>%
  filter(hour %in% hours_to_show) %>%
  arrange(hour) %>%
  mutate(
    pct_change = (n - lag(n)) / lag(n) * 100,
    label_text = ifelse(
      is.na(pct_change), 
      comma(n),
      paste0(comma(n), "\n(", ifelse(pct_change > 0, "+", ""), round(pct_change, 1), "%)"))
  )

ggplot() +
  geom_line(data = full_data, aes(x = hour, y = n), color = "grey80", linewidth = 1) +
  geom_point(data = highlight_data, aes(x = hour, y = n), color = "steelblue", size = 3) +
  geom_text(data = highlight_data, aes(x = hour, y = n, label = label_text), 
            vjust = -0.6, size = 3, lineheight = 0.8) +
  scale_x_continuous(breaks = 0:23) +
  scale_y_continuous(labels = comma, limits = c(0, NA), expand = expansion(mult = c(0, 0.4))) +
  theme_minimal() +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    plot.title = element_text(size = 14),
    plot.subtitle = element_text(color = "grey50", size = 10)
  ) +
  labs(
    title = "Hourly Volume of Stops for 4-Wheels Vehicles",
    subtitle = "TODO",
    x = "Hour of Day (0–23)", 
    y = "Count"
  )
```

```{r, include=FALSE}
rm(hours_to_show, full_data, highlight_data)
```

What can we say?

- **Key finding**:
- **Interpretation**:
- **Comment**: 
- **Proposed Action**:

Before anything else, let us see whether this cycle changed throughout the years.

```{r hour-year-dist}
hours_to_show <- c(0, 4, 8, 12, 16, 20, 23)

full_data <- vehstops %>%
  filter(!is.na(year) & !is.na(hour)) %>%
  count(year, hour)

highlight_data <- full_data %>%
  filter(hour %in% hours_to_show) %>%
  group_by(year) %>%
  arrange(hour) %>%
  mutate(
    pct_change = (n - lag(n)) / lag(n) * 100,
    label_text = ifelse(
      is.na(pct_change), 
      comma(n), 
      paste0(comma(n), "\n(", ifelse(pct_change > 0, "+", ""), round(pct_change, 1), "%)"))
  )

ggplot() +
  geom_line(data = full_data, aes(x = hour, y = n), color = "grey85", linewidth = 0.8) +
  geom_point(data = highlight_data, aes(x = hour, y = n), color = "steelblue", size = 2) +
  geom_text(data = highlight_data, aes(x = hour, y = n, label = label_text), 
            vjust = -0.7, size = 2.5, lineheight = 0.8) +
  facet_wrap(~year, nrow = 1) +
  scale_x_continuous(breaks = c(0, 4, 8, 12, 16, 20, 23)) +
  scale_y_continuous(labels = comma, limits = c(0, NA), expand = expansion(mult = c(0, 0.4))) +
  theme_minimal() +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    plot.title = element_text(size = 14),
    plot.subtitle = element_text(color = "grey50", size = 10),
    strip.text = element_text(size = 11)
  ) +
  labs(
    title = "Hourly Volume of Stops, by Year, for 4-Wheels Vehicles",
    subtitle = "TODO",
    x = "Hour of Day (0–23)", 
    y = "Count"
  )
```

```{r, include=FALSE}
rm(hours_to_show, full_data, highlight_data)
```

What can we say?

- **Key finding**:
- **Interpretation**:
- **Comment**: 
- **Proposed Action**:


Now let us see if the hot season might shift this cycle

```{r hour-hot-season}
plot_data_full <- vehstops %>%
  mutate(
    hot_season = ifelse(month %in% 4:10, 1, 0),
    hot_season_label = factor(hot_season, levels = c(0, 1), 
                              labels = c("Off Season (Nov-Mar)", "Hot Season (Apr-Oct)"))
  ) %>%
  filter(!is.na(hour))

full_data <- plot_data_full %>%
  count(hot_season_label, hour)

hours_to_show <- c(0, 4, 8, 12, 16, 20, 23)

highlight_data <- full_data %>%
  filter(hour %in% hours_to_show) %>%
  group_by(hot_season_label) %>%
  arrange(hour) %>%
  mutate(
    pct_change = (n - lag(n)) / lag(n) * 100,
    label_text = ifelse(is.na(pct_change), 
                        comma(n), 
                        paste0(comma(n), "\n(", ifelse(pct_change > 0, "+", ""), round(pct_change, 1), "%)"))
  )

ggplot() +
  geom_line(data = full_data, aes(x = hour, y = n), color = "grey85", linewidth = 0.8) +
  geom_point(data = highlight_data, aes(x = hour, y = n), color = "steelblue", size = 2) +
  geom_text(data = highlight_data, aes(x = hour, y = n, label = label_text), 
            vjust = -0.7, size = 2.5, lineheight = 0.8) +
  facet_wrap(~hot_season_label, nrow = 1) +
  scale_x_continuous(breaks = c(0, 4, 8, 12, 16, 20, 23)) +
  scale_y_continuous(labels = comma, limits = c(0, NA), expand = expansion(mult = c(0, 0.4))) +
  theme_minimal() +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    plot.title = element_text(size = 14),
    plot.subtitle = element_text(color = "grey50", size = 10),
    strip.text = element_text(size = 12)
  ) +
  labs(
    title = "Hourly Cycle of Enforcement: Hot Season vs. Off Season",
    subtitle = "TODO",
    x = "Hour of Day (0–23)", 
    y = "Count"
  )
```

```{r, include=FALSE}
rm(plot_data_full, hours_to_show, full_data, highlight_data)
```

What can we say?

- **Key finding**:
- **Interpretation**:
- **Comment**: 
- **Proposed Action**:


Let us check the time brackets previously built.

```{r time-bracket-distribution}
plot_univariate_cat(
  DF = vehstops, 
  FEATURE = time_bracket, 
  X_LABEL = "Time Bracket", 
  TITLE = "Distribution of Time Brackets",
  SUBTITLE = "TODO"
)
```

What can we say?

- **Key finding**:
- **Interpretation**:
- **Comment**: 
- **Proposed Action**:

Also, we are going to drop minute as its level of granularity cannot be properly included in the dataset.

## Motorist related predictors

### Reported Age

We have mainly one issue here. First, we have many instances of drivers with unknown age,

```{r age-bracket-dist}
plot_univariate_cat(
  DF = vehstops, 
  FEATURE = age_bracket, 
  X_LABEL = "Age Bracket", 
  TITLE = "Distribution of Stops by Age Bracket",
  SUBTITLE = "TODO"
)
```

What can we say?

- **Key finding**:
- **Interpretation**:
- **Comment**: 
- **Proposed Action**:

And now, let us see how this looks when looking at the full distribution

```{r age-vs-age-bracket}
age_data <- vehstops %>%
  filter(!is.na(age), age >= 0, age < 100) %>%
  mutate(
    age_group = case_when(
      age < 18 ~ "<18",
      age >= 18 & age <= 24 ~ "18–24",
      age >= 25 & age <= 34 ~ "25–34",
      age >= 35 & age <= 44 ~ "35–44",
      age >= 45 & age <= 54 ~ "45–54",
      age >= 55 & age <= 64 ~ "55–64",
      age >= 65 ~ "65+"
    ),
    age_group = factor(age_group, levels = c("<18", "18–24", "25–34", "35–44", "45–54", "55–64", "65+"))
  )

ggplot(age_data, aes(x = age, fill = age_group)) +
  geom_histogram(binwidth = 1, color = "white", linewidth = 0.1) +
  scale_fill_viridis_d() +
  theme_minimal() +
   theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    plot.title = element_text(size = 14),
    plot.subtitle = element_text(color = "grey50", size = 10),
    strip.text = element_text(size = 11)
  ) +
  labs(
    title = "Age Distribution of Drivers by Target Bracket",
    subtitle = "TODO",
    x = "Driver Age", 
    y = "Count",
    fill = "Age Group"
  ) +
  theme(legend.position = "bottom") +
  guides(fill = guide_legend(nrow = 2))
```

```{r, include=FALSE}
rm(age_data)
```

What can we say?

- **Key finding**:
- **Interpretation**:
- **Comment**: 
- **Proposed Action**:

Let us remove the outliers TODO

```{r remove-age-outliers}
# Keep the NAs
# vehstops <- vehstops %>%
#   filter(age < 76 | is.na(age))
```

From now on we will consider drivers of age in the range 16-75.

```{r age-outlier-boxplot-after}
vehstops %>%
  filter(!is.na(age)) %>%
  ggplot(aes(y = age)) +
  geom_boxplot(fill = "steelblue", alpha = 0.6, width = 0.4,
               outlier.size = 1.5, outlier.alpha = 0.4) +
  coord_flip() +
  theme_minimal() +
  theme(
    panel.grid.major.y = element_blank(),
    plot.title = element_text(size = 14),
    plot.subtitle = element_text(color = "grey50", size = 10)
  ) +
  labs(
    title = "Age Distribution after Outlier Removal (76+ excluded)",
    subtitle = "TODO",
    y = "Reported Age", x = NULL
  )
```

What can we say?

- **Key finding**:
- **Interpretation**:
- **Comment**: 
- **Proposed Action**:

### Reported Sex

```{r sex-distribution}
plot_univariate_cat(
  DF = vehstops, 
  FEATURE = sex, 
  X_LABEL = "Reported Sex", 
  TITLE = "Distribution of Stops by Reported Sex (Sorted by Count)",
  SUBTITLE = "TODO",
  order_by = "freq"
)
```

What can we say?

- **Key finding**:
- **Interpretation**:
- **Comment**: 
- **Proposed Action**:

### Reported Ethnicity

```{r ethnicity-distribution}
plot_univariate_cat(
  DF = vehstops, 
  FEATURE = ethnicity, 
  X_LABEL = "Reported Ethnicity", 
  TITLE = "Distribution of Stops by Reported Ethnicity (Sorted by Count)",
  SUBTITLE = "TODO",
  order_by = "freq"
)
```

What can we say?

- **Key finding**:
- **Interpretation**:
- **Comment**: 
- **Proposed Action**:

------------------------------------------------------------------------

# Bivariate EDA

## Association of Binary Response vs. Categorical Predictors

When dealing with a dataset of this size, traditional p-values from chi-squared tests become meaningless. With \~185k observations, virtually any association achieves p \< 0.001 regardless of whether it matters in practice.

To get around this, we use Cramer’s V to measure effect size (association).

Because of how the `table()` function works in R, we can also include integer predictors like `hour` and `age`: they do not take truly continuous values in this dataset, so treating each integer as its own category is valid.

```{r categorical-correlation-heatmap}
# Excluding doy (redundant) and dow (we use weekday instead)
columns <- c(
  "year" , "month", "hot_season", "weekday", 
  "hour", "time_bracket",
  "area", "boro", 
  "sex", "ethnicity", "age", "age_bracket",
  "arrested"
  )

plot_categorical_correlation(
  DF = vehstops, 
  FEATURES = columns,
  TITLE = "Correlation between Categorical Variables (Cramer's V)"
)
```

What can we say?

- **Key finding**:
- **Interpretation**:
- **Comment**: 
- **Proposed Action**:

## Geospatial Features

### Area

```{r area-vs-arrested-ratio}
ratio_by_cat(
  vehstops, arrested, area,
  "Area", "Arrest Rate (%)", "Arrest Ratio by Area",
  style = "stack", legend_title = "Stop ended in arrest?",
  target_labels = c("No", "Yes"),
  subtitle = "TODO",
  label_vjust = 0.5,
  bar_width = 0.6,  
  text_size = 2.5,
  aspect_ratio = 0.9
)
```

Let us look at the numbers too

```{r confirm-area-vs-arrested-ratio}
vehstops %>%
  group_by(area, arrested) %>%
  summarise(count = n(), .groups = "drop_last") %>%
  mutate(percentage = count / sum(count))
```

What can we say?

- **Key finding**:
- **Interpretation**:
- **Comment**: 
- **Proposed Action**:

### Borough

```{r boro-vs-arrested-ratio}
ratio_by_cat(
  vehstops, arrested, boro,
  "Borough", "Arrest Rate (%)", "Arrest Ratio by Borough",
  style = "stack", legend_title = "Stop ended in arrest?",
  target_labels = c("No", "Yes"),
  subtitle = "TODO",
  label_vjust = 0.5,
  bar_width = 0.8,
  text_size = 2.5
)
```

What can we say?

- **Key finding**:
- **Interpretation**:
- **Comment**: 
- **Proposed Action**:

### Latitude and Longitude for Stop Rate

```{r}
nyc_map <- counties(state = "NY", cb = TRUE) %>%
  filter(NAME %in% c("Bronx", "Kings", "New York", "Queens", "Richmond"))

bins <- 100
lon_breaks <- seq(-74.26, -73.71, length.out = bins + 1)
lat_breaks <- seq(40.48, 40.92, length.out = bins + 1)

grid_sum_rate <- vehstops %>%
  filter(!is.na(longitude), !is.na(latitude), !is.na(arrested)) %>%
  mutate(
    xbin = cut(longitude, breaks = lon_breaks, include.lowest = TRUE, right = FALSE),
    ybin = cut(latitude, breaks = lat_breaks, include.lowest = TRUE, right = FALSE)
  ) %>%
  mutate(
    x = as.numeric(gsub(".*\\[(.*),.*", "\\1", xbin)) + (lon_breaks[2] - lon_breaks[1])/2,
    y = as.numeric(gsub(".*\\[(.*),.*", "\\1", ybin)) + (lat_breaks[2] - lat_breaks[1])/2
  ) %>%
  group_by(x, y) %>%
  summarise(
    total_stops = n(),
    arrest_rate = mean(arrested == 1),
    .groups = "drop"
  ) %>%
  filter(total_stops >= 5)

ggplot() +
  geom_sf(data = nyc_map, fill = "white", color = "grey80", linewidth = 0.2) +
  geom_tile(data = grid_sum_rate, aes(x, y, fill = arrest_rate), alpha = 0.8) +
  scale_fill_viridis_c(
    labels = scales::percent, 
    name = "Arrest Rate", 
    option = "inferno",
    direction = -1
  ) +
  coord_sf(xlim = c(-74.26, -73.71), ylim = c(40.48, 40.92)) +
  theme_minimal() +
  theme(
    panel.grid = element_blank(),
    plot.title = element_text(size = 14),
    plot.subtitle = element_text(color = "grey50", size = 10)
  ) +
  labs(
    title = "Spatial Distribution of Arrest Rates - NYC",
    subtitle = "TODO",
    x = "Longitude", 
    y = "Latitude"
  )
```

What can we say?

- **Key finding**:
- **Interpretation**:
- **Comment**: 
- **Proposed Action**:

## Time related features

### Year

```{r year-vs-arrested-ratio}
arrest_trends <- vehstops %>%
  filter(!is.na(year), !is.na(arrested)) %>%
  group_by(year) %>%
  summarise(
    total_stops = n(),
    arrests = sum(arrested == 1),
    rate = arrests / total_stops, 
    .groups = 'drop'
  ) %>%
  arrange(year) %>%
  mutate(
    prev_rate = lag(rate),
    delta = (rate - prev_rate) * 100,
    label_text = ifelse(
      is.na(prev_rate), 
      scales::percent(rate, accuracy = 0.1), 
      paste0(scales::percent(rate, accuracy = 0.1), 
             "\n(", ifelse(delta > 0, "+", ""), round(delta, 1), " pts)")
    )
  )

ggplot(arrest_trends, aes(x = factor(year), y = rate, group = 1)) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_point(color = "steelblue", size = 3) +
  geom_text(aes(label = label_text), vjust = -0.8, size = 3.5, lineheight = 0.8) +
  scale_y_continuous(
    labels = scales::percent, 
    limits = c(0, max(arrest_trends$rate) * 1.3), # Extra room for labels
    expand = expansion(mult = c(0, 0.1))
  ) +
  theme_minimal() +
  theme(
    panel.grid.major.x = element_blank(),
    plot.title = element_text(size = 14),
    plot.subtitle = element_text(size = 10, color = "grey50")
  ) +
  labs(
    title = "Annual Arrest Rate has been Slowly Rising in the last 3 years",
    subtitle = "TODO",
    x = "Year", y = "Arrest Rate (%)"
  )
```

What can we say?

- **Key finding**:
- **Interpretation**:
- **Comment**: 
- **Proposed Action**:

### Month and Hot Season

```{r month-vs-arrested-ratio}
monthly_arrest_data <- vehstops %>%
  filter(!is.na(month), !is.na(arrested)) %>%
  group_by(month) %>%
  summarise(
    total_stops = n(),
    arrests = sum(arrested == 1),
    rate = arrests / total_stops,
    .groups = "drop"
  ) %>%
  mutate(month = factor(month, levels = 1:12)) %>%
  arrange(month) %>%
  mutate(
    prev_rate = lag(rate),
    delta = (rate - prev_rate) * 100,
    label_text = ifelse(
      is.na(prev_rate), 
      scales::percent(rate, accuracy = 0.1), 
      paste0(scales::percent(rate, accuracy = 0.1), 
             "\n(", ifelse(delta > 0, "+", ""), round(delta, 1), " pts)")
    )
  )

monthly_arrest_data %>%
  ggplot(aes(x = month, y = rate, group = 1)) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_point(color = "steelblue", size = 3) +
  geom_text(aes(label = label_text), vjust = -1.2, size = 3, lineheight = 0.8) +
  scale_x_discrete(labels = month.abb) +
  scale_y_continuous(
    labels = scales::percent, 
    limits = c(0, max(monthly_arrest_data$rate) * 1.3), 
    expand = expansion(mult = c(0, 0.1))
  ) +
  theme_minimal() +
  theme(
    panel.grid.major.x = element_blank(),
    plot.title = element_text(size = 14),
    plot.subtitle = element_text(color = "grey50", size = 10)
  ) +
  labs(
    title = "Monthly Arrest Rate Trend",
    subtitle = "TODO",
    x = "Month",
    y = "Arrest Rate (%)"
  )
```

What can we say?

- **Key finding**:
- **Interpretation**:
- **Comment**: 
- **Proposed Action**:

What about `hot_season`?

```{r hot-season-vs-arrested-ratio}
ratio_by_cat(
  vehstops, arrested, hot_season,
  "Hot Season", "Arrest Rate (%)", "Arrest Ratio by Season",
  style = "stack", legend_title = "Stop ended in arrest?",
  target_labels = c("No", "Yes"),
  subtitle = "TODO",
  label_vjust = 0.5,
  text_size = 2.5,
  bar_width = 0.7
)
```

What can we say?

- **Key finding**:
- **Interpretation**:
- **Comment**: 
- **Proposed Action**:

#### Faceted

```{r month-faceted-vs-arrested-ratio}
plot_data <- vehstops %>%
  filter(!is.na(year), !is.na(month), !is.na(arrested)) %>%
  group_by(year, month) %>%
  summarise(
    arrest_rate = mean(arrested == 1), 
    .groups = "drop"
  ) %>%
  mutate(month = factor(month, levels = 1:12)) %>%
  group_by(year) %>%
  mutate(
    rate_jan = arrest_rate[month == 1],
    rate_max = max(arrest_rate),
    is_max = arrest_rate == rate_max,
  
    diff_jan = (arrest_rate - rate_jan) * 100,
    diff_dec = (arrest_rate - rate_max) * 100, 
    
    label_text = case_when(
      month == 1 ~ scales::percent(arrest_rate, accuracy = 0.1), 
      is_max ~ paste0(scales::percent(arrest_rate, accuracy = 0.1), "\n(", ifelse(diff_jan > 0, "+", ""), round(diff_jan, 1), " pts)"),
      month == 12 ~ paste0(scales::percent(arrest_rate, accuracy = 0.1), "\n(", round(diff_dec, 1), " pts)"),
      TRUE ~ ""
    )
  )


plot_data %>%
  ggplot(aes(x = month, y = arrest_rate, group = 1)) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_point(color = "steelblue", size = 2) +
  geom_text(aes(label = label_text), vjust = -1, hjust = 0.2, size = 2.5, lineheight = 0.8) +
  facet_wrap(~year, nrow = 1) +
  scale_x_discrete(
    labels = c("J", "F", "M", "A", "M", "J", "J", "A", "S", "O", "N", "D"), 
    expand = expansion(mult = c(0.1, 0.1))
  ) +
  scale_y_continuous(labels = scales::percent, limits = c(0, NA), expand = expansion(mult = c(0, 0.4))) +
  coord_cartesian(clip = "off") +
  theme_minimal() +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    plot.title = element_text(size = 14),
    plot.subtitle = element_text(color = "grey50", size = 10),
    strip.text = element_text(size = 11)
  ) +
  labs(
    title = "Monthly Arrest Rate, by Year, for 4-Wheels Vehicles in NYC",
    subtitle = "TODO",
    x = "Month",
    y = "Arrest Rate"
  )
```

What can we say?

- **Key finding**:
- **Interpretation**:
- **Comment**: 
- **Proposed Action**:

```{r faceted-hot-season-vs-arrested-ratio}
plot_data <- vehstops %>%
  filter(!is.na(year), !is.na(month), !is.na(arrested)) %>%
  mutate(
    is_hot_season = ifelse(month %in% 4:10, "Hot Season", "Off Season"),
    is_hot_season = factor(is_hot_season, levels = c("Hot Season", "Off Season"))
  ) %>%
  group_by(year, is_hot_season) %>%
  summarise(
    total_stops = n(),
    arrests = sum(arrested == 1),
    arrest_rate = arrests / total_stops,
    .groups = "drop"
  ) %>%
  mutate(
    label_text = paste0(scales::percent(arrest_rate, accuracy = 0.1), "\n(", comma(arrests), " arrests)")
  )

plot_data %>%
  ggplot(aes(x = is_hot_season, y = arrest_rate, fill = is_hot_season)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~year, nrow = 1) +
  geom_text(aes(label = label_text), vjust = -0.5, size = 3, lineheight = 0.8) +
  scale_y_continuous(labels = scales::percent, expand = expansion(mult = c(0, 0.2))) +
  scale_fill_brewer(palette = "Paired") +
  theme_minimal() +
  theme(
    panel.grid.major.x = element_blank(),
    plot.title = element_text(size = 14),
    plot.subtitle = element_text(color = "grey50", size = 10)
  ) +
  labs(
    title = "Arrest Rate for 4-Wheels Vehicles: Hot Season vs. Off Season",
    subtitle = "TODO",
    x = NULL,
    y = "Arrest Rate (%)"
  )
```

What can we say?

- **Key finding**:
- **Interpretation**:
- **Comment**: 
- **Proposed Action**:

### Day, Dow and Weekday

```{r day-vs-arrested-ratio}
daily_arrest_rate <- vehstops %>%
  filter(!is.na(day), !is.na(arrested)) %>%
  group_by(day) %>%
  summarise(
    total_stops = n(),
    arrests = sum(arrested == 1),
    rate = arrests / total_stops,
    .groups = "drop"
  )

daily_arrest_rate %>%
  ggplot(aes(x = day, y = rate)) +
  geom_line(color = "grey60", alpha = 0.5) +
  geom_smooth(method = "loess", color = "steelblue", se = FALSE) +
  scale_y_continuous(labels = scales::percent, limits = c(0, NA)) +
  theme_minimal() + 
  theme(
    panel.grid.major.x = element_blank(),
    plot.title = element_text(size = 14),
    plot.subtitle = element_text(color = "grey50", size = 10)
  ) +
  labs(
    title = "Arrest Rate by Day of the Month (1–31)",
    subtitle = "TODO",
    x = "Day of the Month",
    y = "Arrest Rate (%)"
  )
```

What can we say?

- **Key finding**:
- **Interpretation**:
- **Comment**: 
- **Proposed Action**:

```{r dow-vs-arrested-ratio}
daily_arrest_rate <- vehstops %>%
  filter(!is.na(dow), !is.na(arrested)) %>%
  group_by(dow) %>%
  summarise(
    total_stops = n(),
    arrests = sum(arrested == 1),
    rate = arrests / total_stops,
    .groups = "drop"
  ) %>%
  mutate(
    dow_label = factor(dow, levels = 1:7,
                       labels = c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun")),
    prev_rate = lag(rate),
    delta = (rate - prev_rate) * 100,
    label_text = ifelse(
      is.na(prev_rate),
      scales::percent(rate, accuracy = 0.1),
      paste0(scales::percent(rate, accuracy = 0.1),
             "\n(", ifelse(delta > 0, "+", ""), round(delta, 1), " pts)")
    )
  )

daily_arrest_rate %>%
  ggplot(aes(x = dow_label, y = rate, group = 1)) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_point(color = "steelblue", size = 3) +
  geom_text(aes(label = label_text), vjust = -1.0, size = 3, lineheight = 0.8) +
  scale_y_continuous(
    labels = scales::percent,
    limits = c(0, max(daily_arrest_rate$rate) * 1.3),
    expand = expansion(mult = c(0, 0.1))
  ) +
  theme_minimal() +
  theme(
    panel.grid.major.x = element_blank(),
    plot.title = element_text(size = 14),
    plot.subtitle = element_text(color = "grey50", size = 10)
  ) +
  labs(
    title = "Arrest Rate by Day of the Week for 4-Wheels Vehicles",
    subtitle = "TODO",
    x = "Day of the Week",
    y = "Arrest Rate (%)"
  )
```

What can we say?

- **Key finding**:
- **Interpretation**:
- **Comment**: 
- **Proposed Action**:

### Hour and Time Bracket

```{r hour-vs-arrested-ratio}
hourly_arrest_rate <- vehstops %>%
  filter(!is.na(hour), !is.na(arrested)) %>%
  group_by(hour) %>%
  summarise(
    total_stops = n(),
    arrests = sum(arrested == 1),
    rate = arrests / total_stops,
    .groups = "drop"
  )

hourly_arrest_rate %>%
  ggplot(aes(x = hour, y = rate)) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_point(color = "steelblue", size = 2) +
  scale_x_continuous(breaks = 0:23) +
  scale_y_continuous(labels = scales::percent, limits = c(0, NA),
                     expand = expansion(mult = c(0, 0.15))) +
  theme_minimal() +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    plot.title = element_text(size = 14),
    plot.subtitle = element_text(color = "grey50", size = 10)
  ) +
  labs(
    title = "Arrest Rate by Hour of Day for 4-Wheels Vehicles",
    subtitle = "TODO",
    x = "Hour of Day (0–23)",
    y = "Arrest Rate (%)"
  )
```

What can we say?

- **Key finding**:
- **Interpretation**:
- **Comment**: 
- **Proposed Action**:

Let us look at our grouping now

```{r time-bracket-vs-arrested-ratio}
ratio_by_cat(
  vehstops, arrested, time_bracket,
  "Time bracket", "Arrest Rate (%)", "Arrest ratio by Custom Time Bracket",
  style = "stack", legend_title = "Stop ended in arrest?", target_labels = c("No", "Yes"),
  subtitle = "TODO"
)
```

What can we say?

- **Key finding**:
- **Interpretation**:
- **Comment**: 
- **Proposed Action**:

## Motorist related features

### Reported Age

```{r age-bracket-vs-target}
ratio_by_cat(
  vehstops, arrested, age_bracket,
  "Reported Age Bracket", "Arrest Rate (%)", "Arrest Ratio by Reported Age Bracket",
  style = "stack", legend_title = "Stop ended in arrest?", target_labels = c("No", "Yes"),
  subtitle = "TODO",
  label_vjust = 0.5
)
```

What can we say? TODO

- **Key finding**: 
- **Interpretation**:
- **Comment**:
- **Proposed Action**: We want to include `age_bracket` as a predictor in our models instead of just using `age`.

In fact,

### Reported Sex

```{r sex-vs-target, fig.width=6}
ratio_by_cat(
  vehstops, arrested, sex,
  "Reported sex", "Arrest Rate (%)", "Arrest Ratio by Reported sex",
  style = "stack", legend_title = "Stop ended in arrest?", target_labels = c("No", "Yes"),
  subtitle = "TODO",
)
```

What can we say? TODO

- **Key finding**: 
- **Interpretation**: 
- **Comment**:
- **Proposed Action**:

### Reported Ethnicity

```{r ethnicity-vs-target, fig.width=6}
ratio_by_cat(
  vehstops, arrested, ethnicity,
  "Reported ethnicity", "Share of stops", "Escalation ratio by Reported Ethnicity",
  style = "stack", legend_title = "Stop ended in arrest?", target_labels = c("No", "Yes"),
  subtitle = "TODO",
)
```

What can we say? TODO

- **Key finding**:
- **Interpretation**:
- **Comment**: 
- **Proposed Action**:

---

# Recap

TODO

------------------------------------------------------------------------

# Final touches

## Re-Ordering Features

We re-order the features for better display

```{r re-order-dataset}
vehstops_ord <- vehstops %>%
  select(
    year, month, hot_season, weekday,
    hour, time_bracket,
    area, boro,
    sex, ethnicity, 
    age, age_bracket,
    arrested
  )
```

Here is the final version.

```{r summary-final-dataset}
print(dfSummary(vehstops_ord), method="render")
```

## Saving the dataset

```{r save-dataset}
write.csv(vehstops_ord, "data/vehiclestops_4_wheels_final.csv", row.names = FALSE)
```
