---
title: "Predicting Excessive Enforcement Patterns in Police Behavior After Vehicles Stops in New York"
subtitle: "Model Selection and Evaluation - GAM 2-Wheels"
author: "Gianmaria Pizzo"
output:
  html_document:
    fig_caption: yes
    theme: flatly
    highlight: pygments
    code_folding: show
    toc: yes
    toc_depth: 4
    number_sections: no
    toc_float:
      smooth_scroll: no
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

These `.Rmd` files represent the project submission for the course [Statistical Inference and Learning](https://www.unive.it/data/course/521944) by Professor [Cristiano Varin](https://www.unive.it/data/people/5592728) at [Ca' Foscari University of Venice](https://www.unive.it/) for the year 2025/2026.

------------------------------------------------------------------------

# 0) Environment

## 0.1) Packages

```{r requirements}
requirements <- c(
  "remotes", "summarytools", "forcats", "scales", "purrr", "broom",
  "stringr", "dbplyr", "dplyr", "dtplyr", "tidyr", "tidyselect",
  "timeDate", "timechange", "tzdb", "hms", "readr", "ggcorrplot",
  "lubridate", "prettyunits", "RColorBrewer", "viridis", "rcartocolor",
  "sf", "ggplot2", "patchwork", "pROC", "kableExtra", "mgcv", "glmnet"
)
```

```{r install-requirements, include=FALSE}
for (req in requirements){
  if (!require(req, character.only = TRUE)){
      install.packages(req)
  }
}
```

Note that some of the packages may not be used, but to avoid any error related to dependencies, we are going to keep all of these libraries. Now, the loading of all packages should be immediate.

```{r load-requirements, include=FALSE}
for (req in requirements){
  if (!require(req, character.only = TRUE)){
      library(req, character.only = TRUE)
  }
}
```

```{r remove-vars, include=FALSE}
rm(requirements, req)
```

## 0.2) Functions

Here we define the helper functions used throughout model selection and evaluation, kept in one place for reuse across every model fitted below.

### 0.2.1) Data-Split and Cross-Validation Helpers

```{r fn-check-prop}
# ==============================================================================
# 1. Train / Test Split Diagnostics
# ==============================================================================

#' @title Check Train/Test Level Proportions
#'
#' @description Tabulates one variable's level proportions separately in
#' `train` and `test` and reports the discrepancy in percentage points. Used
#' to verify that the stratified split preserved the marginal distribution of
#' key covariates.
#'
#' @param var The column name (present in both `train` and `test`) to tabulate.
#' @param label Human-readable label for the variable, stored in the output.
#'
#' @return A tibble with columns variable, level, train, test, diff_ppt.
check_prop <- function(var, label) {
  tr <- prop.table(table(train[[var]]))
  te <- prop.table(table(test[[var]]))
  tibble(
    level = names(tr),
    train = as.numeric(tr),
    test = as.numeric(te),
    diff_ppt = round((as.numeric(te) - as.numeric(tr)) * 100, 2)
  ) %>%
    mutate(variable = label) %>%
    relocate(variable)
}
```

```{r fn-cv-auc}
# ==============================================================================
# 2. Cross-Validated AUC
# ==============================================================================

#' @title Cross-Validated AUC for a Logistic Regression Formula
#'
#' @description Refits a binomial GLM on each of the k training folds and
#' scores it on the held-out fold, returning the mean AUC across folds. Folds
#' are processed in parallel since refitting a GLM on ~1.7M rows ten times is
#' the dominant cost of the model-comparison sections below.
#'
#' @param formula Model formula passed to glm(family = binomial).
#' @param data Training data.frame; must contain the response `arrested`.
#' @param folds_vec Integer vector, same length as nrow(data), with values
#' 1..k giving each row's fold assignment.
#'
#' @return A numeric scalar, the mean out-of-fold AUC (NA folds are dropped).
cv_auc <- function(formula, data, folds_vec) {
  k <- max(folds_vec)

  auc_vals <- parallel::mclapply(seq_len(k), function(i) {
    tr <- data[folds_vec != i, ]
    te <- data[folds_vec == i, ]

    # A fold can fail to fit (e.g. rank deficiency in a rare stratum) — treat
    # as a missing fold rather than aborting the whole CV run
    fit <- tryCatch(
      glm(formula, family = binomial, data = tr),
      error = function(e) NULL
    )
    if (is.null(fit)) return(NA_real_)

    prob <- tryCatch(
      predict(fit, newdata = te, type = "response"),
      error = function(e) rep(NA_real_, nrow(te))
    )

    tryCatch(
      as.numeric(pROC::auc(te$arrested, prob, quiet = TRUE)),
      error = function(e) NA_real_
    )
  }, mc.cores = parallel::detectCores() - 1)

  mean(unlist(auc_vals), na.rm = TRUE)
}
```

```{r fn-nb-cv-auc}
#' @title Cross-Validated AUC for Naive Bayes at a Given Laplace Constant
#'
#' @description Refits e1071::naiveBayes on each training fold with smoothing
#' constant `k` and scores it on the held-out fold, returning the mean AUC.
#' Mirrors cv_auc() but calls naiveBayes() instead of glm() and reads the
#' positive-class probability from the second column of the raw posterior.
#'
#' @param k Numeric Laplace smoothing constant passed to naiveBayes(laplace = k).
#' @param data Training data.frame; must contain the response `arrested`.
#' @param folds_vec Integer vector, same length as nrow(data), with values
#' 1..k giving each row's fold assignment.
#' @param formula Model formula passed to naiveBayes().
#'
#' @return A numeric scalar, the mean out-of-fold AUC (NA folds are dropped).
nb_cv_auc <- function(k, data, folds_vec, formula) {
  n_folds  <- max(folds_vec)
  auc_vals <- numeric(n_folds)
  for (i in seq_len(n_folds)) {
    tr  <- data[folds_vec != i, ]
    te  <- data[folds_vec == i, ]
    fit <- e1071::naiveBayes(formula, data = tr, laplace = k)

    # Posterior: columns named after class levels; column 2 = P(arrested=1)
    prob <- predict(fit, newdata = te, type = "raw")[, 2]
    auc_vals[i] <- tryCatch(
      as.numeric(pROC::auc(te$arrested, prob, quiet = TRUE)),
      error = function(e) NA_real_
    )
  }
  mean(auc_vals, na.rm = TRUE)
}
```

### 0.2.2) Threshold and Confusion-Matrix Helpers

```{r fn-eval-helpers}
# ==============================================================================
# 3. Threshold Selection and Confusion-Matrix Helpers
# ==============================================================================

#' @title Find the Youden-Optimal Classification Threshold
#'
#' @description Finds the probability threshold on an ROC curve that
#' maximises Youden's J statistic (sensitivity + specificity − 1), used
#' throughout as the single decision threshold for turning predicted
#' probabilities into classes.
#'
#' @param roc_obj A pROC "roc" object, as returned by pROC::roc().
#'
#' @return A numeric scalar, the optimal threshold on the probability scale.
best_thresh <- function(roc_obj) {
  pROC::coords(roc_obj, "best", ret = "threshold", best.method = "youden",
               quiet = TRUE)$threshold
}


#' @title Tally a Confusion Matrix
#'
#' @description Cross-tabulates actual vs predicted binary class labels.
#'
#' @param actual Integer/numeric vector of 0/1 true labels.
#' @param predicted Integer/numeric vector of 0/1 predicted labels.
#'
#' @return A tibble with columns Actual, Predicted, n (one row per non-empty cell).
conf_tib <- function(actual, predicted) {
  tibble(
    Actual    = factor(actual,    levels = c(0, 1)),
    Predicted = factor(predicted, levels = c(0, 1))
  ) %>% count(Actual, Predicted)
}


#' @title Plot a Confusion-Matrix Heatmap
#'
#' @description Renders conf_tib() output as a 2x2 tile plot with cell counts.
#'
#' @param actual Integer/numeric vector of 0/1 true labels.
#' @param predicted Integer/numeric vector of 0/1 predicted labels.
#' @param title Main title of the plot.
#'
#' @return A ggplot2 object.
plot_cm <- function(actual, predicted, title = "Confusion Matrix") {
  conf_tib(actual, predicted) %>%
    ggplot(aes(Predicted, Actual, fill = n)) +
    geom_tile(colour = "white") +
    geom_text(aes(label = scales::comma(n)), size = 5.5, fontface = "bold") +
    scale_fill_gradient(low = "#e8f4f8", high = "#2c7bb6", guide = "none") +
    scale_x_discrete(labels = c("0" = "No arrest", "1" = "Arrest")) +
    scale_y_discrete(labels = c("0" = "No arrest", "1" = "Arrest")) +
    labs(title = title, x = "Predicted", y = "Actual") +
    theme_minimal(base_size = 12) +
    theme(panel.grid = element_blank())
}


#' @title Compute Threshold-Dependent Classification Metrics
#'
#' @description Computes sensitivity, specificity, balanced accuracy, PPV,
#' and NPV for a fixed set of hard class predictions against the true labels.
#'
#' @param actual Integer/numeric vector of 0/1 true labels.
#' @param predicted Integer/numeric vector of 0/1 predicted labels.
#' @param pos Value in `actual`/`predicted` denoting the positive class.
#'
#' @return A named list with sensitivity, specificity, balanced_acc, ppv, npv.
class_metrics <- function(actual, predicted, pos = 1) {
  tp <- sum(actual == pos & predicted == pos)
  tn <- sum(actual != pos & predicted != pos)
  fp <- sum(actual != pos & predicted == pos)
  fn <- sum(actual == pos & predicted != pos)
  sens <- tp / (tp + fn)
  spec <- tn / (tn + fp)
  list(
    sensitivity  = sens,
    specificity  = spec,
    balanced_acc = (sens + spec) / 2,
    ppv          = tp / (tp + fp),
    npv          = tn / (tn + fn)
  )
}


# ==============================================================================
# 4. Probability-Based Evaluation Metrics
# ==============================================================================

#' @title Compute the Full Evaluation-Metrics Tibble for One Model
#'
#' @description Given predicted probabilities and true labels, computes
#' threshold-free metrics (AUC, PR-AUC, log-loss, Brier score) plus
#' threshold-dependent metrics at the Youden-optimal cut-off (sensitivity,
#' specificity, precision, F1, MCC, balanced accuracy). Returns one row per
#' call, so results from multiple models can be row-bound into a comparison
#' table.
#'
#' @param actual Integer/numeric vector of 0/1 true labels.
#' @param prob Numeric vector of predicted P(arrested = 1), same length as `actual`.
#' @param model_name Label stored in the Model column.
#'
#' @return A one-row tibble of rounded metrics (see column names in the body).
compute_metrics <- function(actual, prob, model_name = "Model") {
  valid  <- !is.na(actual) & !is.na(prob)
  actual <- actual[valid]; prob <- prob[valid]
  roc_obj <- pROC::roc(actual, prob, quiet = TRUE)
  auc_val <- as.numeric(pROC::auc(roc_obj))
  thr     <- best_thresh(roc_obj)
  pred_cl <- as.integer(prob >= thr)

  tp <- sum(actual == 1 & pred_cl == 1); tn <- sum(actual == 0 & pred_cl == 0)
  fp <- sum(actual == 0 & pred_cl == 1); fn <- sum(actual == 1 & pred_cl == 0)
  sens <- if ((tp+fn) == 0) NA_real_ else tp / (tp+fn)
  spec <- if ((tn+fp) == 0) NA_real_ else tn / (tn+fp)
  prec <- if ((tp+fp) == 0) NA_real_ else tp / (tp+fp)
  f1   <- if (is.na(prec) || is.na(sens) || (prec+sens) == 0) NA_real_ else
            2 * prec * sens / (prec + sens)

  # Matthews correlation coefficient — guard the denominator against any
  # degenerate (all-one-class) confusion matrix, which would divide by zero
  mcc_den <- sqrt(as.numeric(tp+fp) * as.numeric(tp+fn) *
                  as.numeric(tn+fp) * as.numeric(tn+fn))
  mcc_val <- if (mcc_den == 0) NA_real_ else (tp*tn - fp*fn) / mcc_den

  # eps floors probabilities away from 0/1 so log() never evaluates to -Inf
  eps     <- 1e-15
  logloss <- -mean(actual * log(pmax(prob, eps)) + (1-actual) * log(pmax(1-prob, eps)))
  brier   <- mean((prob - actual)^2)

  # PR-AUC via trapezoid rule over pROC coords (pROC has no built-in PR-AUC)
  pr_raw <- pROC::coords(roc_obj, "all",
                          ret = c("sensitivity", "precision"),
                          transpose = FALSE, drop = FALSE)
  names(pr_raw)[names(pr_raw) == "sensitivity"] <- "recall"
  pr_raw <- pr_raw %>%
    filter(!is.nan(precision), !is.na(precision), !is.nan(recall)) %>%
    arrange(recall)
  pr_auc_val <- if (nrow(pr_raw) >= 2)
    sum(diff(pr_raw$recall) *
          (head(pr_raw$precision, -1) + tail(pr_raw$precision, -1)) / 2)
  else NA_real_

  tibble(
    Model       = model_name,
    AUC         = round(auc_val, 4),
    PR_AUC      = round(pr_auc_val, 4),
    Log_Loss    = round(logloss, 4),
    Brier       = round(brier, 4),
    Threshold   = round(thr, 4),
    Sensitivity = round(sens, 3),
    Specificity = round(spec, 3),
    Precision   = round(prec, 3),
    F1          = round(f1, 3),
    MCC         = round(mcc_val, 3),
    Bal_Acc     = round(if (!is.na(sens) && !is.na(spec)) (sens+spec)/2
                        else NA_real_, 3)
  )
}
```

### 0.2.3) Model Evaluation Plot Functions

```{r fn-plot-roc-single}
# ==============================================================================
# 5. Plots for Single-Model Evaluation
# ==============================================================================

#' @title Plot the ROC Curve for a Single Model
#'
#' @description Plots sensitivity vs specificity across all thresholds, with
#' the AUC annotated on the panel.
#'
#' @param actual Integer/numeric vector of 0/1 true labels.
#' @param prob Numeric vector of predicted P(arrested = 1), same length as `actual`.
#' @param title Main title of the plot.
#'
#' @return A ggplot2 object.
plot_roc_single <- function(actual, prob, title = "ROC Curve") {
  roc_obj <- pROC::roc(actual, prob, quiet = TRUE)
  auc_lbl <- paste0("AUC = ", round(as.numeric(pROC::auc(roc_obj)), 4))
  pROC::ggroc(roc_obj, colour = "#2c7bb6", linewidth = 0.9) +
    geom_abline(slope = 1, intercept = 1, linetype = "dashed", colour = "grey60") +
    annotate("text", x = 0.25, y = 0.06, label = auc_lbl,
             colour = "#2c7bb6", size = 4) +
    labs(x = "Specificity", y = "Sensitivity", title = title) +
    theme_minimal(base_size = 11) + theme(panel.grid.minor = element_blank())
}
```

```{r fn-plot-pr-single}
#' @title Plot the Precision-Recall Curve for a Single Model
#'
#' @description Plots precision vs recall across all thresholds against the
#' no-skill baseline (the class prevalence), with PR-AUC annotated. More
#' informative than the ROC curve under the severe class imbalance here.
#'
#' @param actual Integer/numeric vector of 0/1 true labels.
#' @param prob Numeric vector of predicted P(arrested = 1), same length as `actual`.
#' @param title Main title of the plot.
#'
#' @return A ggplot2 object.
plot_pr_single <- function(actual, prob, title = "Precision-Recall Curve") {
  roc_obj <- pROC::roc(actual, prob, quiet = TRUE)
  pr_raw  <- pROC::coords(roc_obj, "all",
                           ret = c("sensitivity", "precision"),
                           transpose = FALSE, drop = FALSE)
  names(pr_raw)[names(pr_raw) == "sensitivity"] <- "recall"
  pr_df <- pr_raw %>%
    filter(!is.nan(precision), !is.na(precision)) %>% arrange(recall)

  # PR-AUC via trapezoid rule, same approach as in compute_metrics()
  pr_auc_lbl <- if (nrow(pr_df) >= 2) {
    v <- sum(diff(pr_df$recall) *
               (head(pr_df$precision,-1) + tail(pr_df$precision,-1)) / 2)
    paste0("PR-AUC = ", round(v, 4))
  } else "PR-AUC = NA"
  baseline <- mean(actual)
  ggplot(pr_df, aes(x = recall, y = precision)) +
    geom_line(colour = "#d73027", linewidth = 0.9) +
    geom_hline(yintercept = baseline, linetype = "dashed", colour = "grey55") +
    annotate("text", x = 0.02, y = baseline + 0.015,
             label = paste0("No-skill = ", round(baseline, 3)),
             colour = "grey40", size = 3.2, hjust = 0) +
    annotate("text", x = 0.55, y = 0.90, label = pr_auc_lbl,
             colour = "#d73027", size = 4) +
    scale_x_continuous(limits = c(0, 1)) + scale_y_continuous(limits = c(0, 1)) +
    labs(x = "Recall", y = "Precision", title = title) +
    theme_minimal(base_size = 11) + theme(panel.grid.minor = element_blank())
}
```

```{r fn-plot-calibration-single}
#' @title Plot the Calibration (Reliability) Curve for a Single Model
#'
#' @description Bins predictions into `n_bins` equal-frequency groups and
#' plots mean predicted probability against observed positive rate per bin,
#' against the perfect-calibration diagonal.
#'
#' @param actual Integer/numeric vector of 0/1 true labels.
#' @param prob Numeric vector of predicted P(arrested = 1), same length as `actual`.
#' @param title Main title of the plot.
#' @param n_bins Number of quantile bins of `prob`.
#'
#' @return A ggplot2 object.
plot_calibration_single <- function(actual, prob, title = "Calibration Plot",
                                    n_bins = 10) {
  # Quantile breaks (not equal-width) so each bin has a comparable sample size
  # despite the predicted probabilities being heavily right-skewed
  brks  <- unique(quantile(prob, seq(0, 1, length.out = n_bins + 1), na.rm = TRUE))
  cal_df <- tibble(actual = actual, prob = prob) %>%
    mutate(bin = cut(prob, breaks = brks, include.lowest = TRUE)) %>%
    filter(!is.na(bin)) %>%
    group_by(bin) %>%
    summarise(mean_pred = mean(prob), mean_obs = mean(actual),
              n = n(), .groups = "drop")
  ggplot(cal_df, aes(x = mean_pred, y = mean_obs)) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey55") +
    geom_line(colour = "#2c7bb6", linewidth = 0.8) +
    geom_point(aes(size = n), colour = "#2c7bb6", alpha = 0.85) +
    scale_size_continuous(name = "Count", range = c(2, 9)) +
    scale_x_continuous(labels = scales::percent_format(accuracy = 0.1)) +
    scale_y_continuous(labels = scales::percent_format(accuracy = 0.1)) +
    labs(x = "Mean predicted probability", y = "Observed positive rate",
         title = title,
         caption = "Size ∝ bin count.  Dashed = perfect calibration.") +
    theme_minimal(base_size = 11) +
    theme(legend.position = "bottom", panel.grid.minor = element_blank())
}
```

```{r fn-plot-threshold-single}
#' @title Plot the Threshold-Sweep Metrics for a Single Model
#'
#' @description Plots sensitivity, specificity, precision, and F1 as
#' functions of the classification threshold, with the Youden-optimal
#' threshold marked.
#'
#' @param actual Integer/numeric vector of 0/1 true labels.
#' @param prob Numeric vector of predicted P(arrested = 1), same length as `actual`.
#' @param title Main title of the plot.
#'
#' @return A ggplot2 object.
plot_threshold_single <- function(actual, prob, title = "Threshold Analysis") {
  roc_obj  <- pROC::roc(actual, prob, quiet = TRUE)
  thr_data <- pROC::coords(roc_obj, "all",
                            ret = c("threshold","sensitivity","specificity","precision"),
                            transpose = FALSE, drop = FALSE) %>%
    filter(is.finite(threshold), threshold >= 0, threshold <= 1) %>%
    mutate(f1 = ifelse(!is.nan(precision) & (precision + sensitivity) > 0,
                       2 * precision * sensitivity / (precision + sensitivity), 0)) %>%
    pivot_longer(c(sensitivity, specificity, precision, f1),
                 names_to = "Metric", values_to = "Value") %>%
    mutate(Metric = recode(Metric,
      sensitivity = "Sensitivity", specificity = "Specificity",
      precision   = "Precision",   f1          = "F1 Score"))
  thr_opt <- best_thresh(roc_obj)
  ggplot(thr_data, aes(x = threshold, y = Value, colour = Metric)) +
    geom_line(linewidth = 0.75, na.rm = TRUE) +
    geom_vline(xintercept = thr_opt, linetype = "dashed",
               colour = "grey35", linewidth = 0.55) +
    annotate("text", x = thr_opt, y = 0.04,
             label = paste0("Youden\n", round(thr_opt, 3)),
             hjust = -0.08, size = 3, colour = "grey30") +
    scale_colour_viridis_d(option = "turbo", begin = 0.05, end = 0.92) +
    scale_y_continuous(limits = c(0, 1)) +
    labs(x = "Threshold", y = "Metric value", colour = NULL, title = title,
         caption = "Dashed: Youden's J optimal threshold") +
    theme_minimal(base_size = 11) +
    theme(legend.position = "bottom", panel.grid.minor = element_blank())
}
```

```{r fn-plot-eval-quad}
#' @title Plot the 2x2 Evaluation Panel for a Single Model
#'
#' @description Composes plot_roc_single(), plot_pr_single(),
#' plot_calibration_single(), and plot_threshold_single() into one
#' (ROC | PR) / (Calibration | Threshold) patchwork layout.
#'
#' @param actual Integer/numeric vector of 0/1 true labels.
#' @param prob Numeric vector of predicted P(arrested = 1), same length as `actual`.
#' @param model_name Prefixed to each sub-plot's title.
#'
#' @return A patchwork object (composed ggplot2 object).
eval_quad_plot <- function(actual, prob, model_name) {
  p1 <- plot_roc_single(actual, prob,   paste(model_name, "— ROC"))
  p2 <- plot_pr_single(actual, prob,    paste(model_name, "— Precision-Recall"))
  p3 <- plot_calibration_single(actual, prob, paste(model_name, "— Calibration"))
  p4 <- plot_threshold_single(actual, prob, paste(model_name, "— Threshold Analysis"))
  (p1 | p2) / (p3 | p4)
}
```

```{r fn-plot-nb-feature}
#' @title Plot a Naive Bayes Conditional-Probability Bar Chart
#'
#' @description Plots P(feature level | class) from one entry of a fitted
#' naiveBayes model's frequency tables, grouped by class, so the strongest
#' discriminators under the conditional-independence assumption are visible.
#'
#' @param feature_name Name of a predictor in `mod$tables`.
#' @param mod A fitted e1071::naiveBayes model object.
#'
#' @return A ggplot2 object.
plot_nb_feature <- function(feature_name, mod) {
  tbl <- mod$tables[[feature_name]]
  as.data.frame(tbl) %>%
    rename(level = 2, class = 1, prob = Freq) %>%
    mutate(class = factor(class, labels = c("Not arrested", "Arrested"))) %>%
    ggplot(aes(x = level, y = prob, fill = class)) +
    geom_col(position = position_dodge(width = 0.75), width = 0.65) +
    scale_fill_viridis_d(option = "mako", begin = 0.3, end = 0.8) +
    scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
    labs(x = NULL, y = "P(level | class)", title = feature_name, fill = NULL) +
    theme_minimal(base_size = 9) +
    theme(
      axis.text.x    = element_text(angle = 30, hjust = 1),
      legend.position = "bottom",
      panel.grid.minor = element_blank()
    )
}
```

### 0.2.4) Model-Summary Helpers

```{r fn-build-modal-row}
# ==============================================================================
# 6. Marginal-Effect and Coefficient Helpers
# ==============================================================================

#' @title Build a One-Row Data Frame of Modal/Median Covariate Values
#'
#' @description For each variable in `vars`, takes the most frequent level
#' (factor/character) or the median (numeric) from `data`. Used to hold
#' non-focal predictors fixed at a representative value when computing
#' marginal-effect predictions for an interaction term.
#'
#' @param data Data.frame to summarise (typically `train`).
#' @param vars Character vector of column names in `data`.
#'
#' @return A one-row data.frame with one column per entry in `vars`.
build_modal_row <- function(data, vars) {
  row_list <- lapply(vars, function(v) {
    x <- data[[v]]
    if (is.factor(x)) {
      factor(names(sort(table(x), decreasing = TRUE))[1], levels = levels(x))
    } else if (is.numeric(x)) {
      median(x, na.rm = TRUE)
    } else {
      names(sort(table(x), decreasing = TRUE))[1]
    }
  })
  as.data.frame(setNames(row_list, vars))
}
```

```{r fn-extract-coefs}
#' @title Extract Non-Zero LASSO Coefficients at a Given Lambda
#'
#' @description Extracts the coefficients selected by a cv.glmnet fit at a
#' named lambda (e.g. "lambda.min" or "lambda.1se"), drops the zeroed-out
#' coefficients, and adds the exponentiated odds ratio.
#'
#' @param cv_fit A fitted glmnet::cv.glmnet object.
#' @param s_name Lambda selector passed to coef(cv_fit, s = s_name), typically
#' "lambda.min" or "lambda.1se".
#'
#' @return A tibble with columns feature, estimate (log-odds), OR (odds ratio),
#' sorted by descending |estimate|.
extract_coefs <- function(cv_fit, s_name) {
  cf <- coef(cv_fit, s = s_name)
  tibble(
    feature  = rownames(cf)[-1],
    estimate = as.numeric(cf[-1])
  ) %>%
    filter(estimate != 0) %>%
    arrange(desc(abs(estimate))) %>%
    mutate(OR = round(exp(estimate), 3), estimate = round(estimate, 4))
}
```

------------------------------------------------------------------------

# 1) Data Loading and Preparation

```{r load-data}
vehstop <- read.csv("data/vehiclestops_2_wheels_final.csv")
```

## 1.1) Factor conversion and Reference Levels

We convert all categorical predictors to factors and set interpretable reference levels before any modelling step, ensuring consistency between all downstream model fits and cross-validation folds.

```{r factor-prep}
# Ordered weekday levels (ISO order)
wday_levels <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday")

# Ordered time bracket levels
tbkt_levels <- c(
  "Late night (0–3)",
  "Early morning (4–7)",
  "Morning commute (8–11)",
  "Midday (12–15)",
  "Afternoon/evening (16–19)",
  "Night (20–23)"
)

# Ordered age-bracket levels
abkt_levels <- c("Under 18", "18–24", "25–34", "35–44", "45–54", "55–64", "65+", "Unknown")

vehstop <- vehstop %>%
  mutate(
    year = factor(year),
    month = factor(month, levels = 1:12),
    weekday = factor(weekday, levels = wday_levels),
    hour = factor(hour, levels = 0:23),
    time_bracket = factor(time_bracket, levels = tbkt_levels),
    area = factor(area),
    boro = factor(boro),
    sex = factor(sex),
    ethnicity = factor(ethnicity),
    age_bracket = factor(age_bracket, levels = abkt_levels)
  )

# Reference levels chosen for interpretable contrasts
vehstop$sex <- relevel(vehstop$sex, ref = "M")
vehstop$ethnicity <- relevel(vehstop$ethnicity, ref = "White")
vehstop$weekday <- relevel(vehstop$weekday, ref = "Monday")
vehstop$time_bracket <- relevel(vehstop$time_bracket, ref = "Morning commute (8–11)")
vehstop$age_bracket <- relevel(vehstop$age_bracket, ref = "25–34")

rm(wday_levels, tbkt_levels, abkt_levels)
```

## 1.2) Removing NAs from `age`

We had to exclude `age` as a predictor from the other models, since we had missing values that we labeled as `Unknown` in `age_bracket`. Here instead we want to include it so we will drop the NAs and use a smoothing curve.

```{r drop-age-na}
# TODO: Drop `age` NAs
vehstop <- vehstop %>% 
  dplyr::filter(!is.na(age))
```

## 1.3) Stratified Train / Test Split

For the train/test split, a **stratified sampling approach** is used to ensure that the training and test sets have comparable distributions of the outcome variable (`arrested`) and relevant contextual variables.

This is particularly important because the target variable is imbalanced, with arrested cases representing only a small proportion of all observations. Rather than stratifying only on `arrested`, the code constructs a **joint stratification variable** from `arrested`, `area`, `sex`, `ethnicity`, and `hour` using `interaction()`. Thus, each observation is assigned to a stratum corresponding to a particular combination of these variables.

Formally, the strata can be represented as

$$S_i = (\text{arrested}_i,\text{area}_i,\text{sex}_i,\text{ethnicity}_i,\text{hour}_i),$$

where $S_i$ denotes the stratum to which observation $i$ belongs. The objective is to sample approximately 80% of observations within each stratum for the training set and retain the remaining 20% for testing:

$$P(i \in \mathcal{D}_{train} \mid S_i=s) \approx 0.8,
\qquad
P(i \in \mathcal{D}_{test} \mid S_i=s) \approx 0.2.$$

```{r stratified-split}
set.seed(2026)

# Joint stratum key; drop = TRUE removes empty combinations
vehstop$.strat <- interaction(
  vehstop$arrested, vehstop$area, vehstop$sex, vehstop$ethnicity, vehstop$hour,
  drop = TRUE
)

train_rows <- vehstop %>%
  dplyr::mutate(.row = row_number()) %>%
  dplyr::group_by(.strat) %>%
  dplyr::slice_sample(prop = 0.8) %>%
  dplyr::pull(.row)

train <- vehstop[ train_rows, ] %>% dplyr::select(-.strat)
test <- vehstop[-train_rows, ] %>% dplyr::select(-.strat)
vehstop <- vehstop %>% dplyr::select(-.strat)
```

In the code, `group_by(.strat) %>% slice_sample(prop = 0.8)` performs this within-stratum sampling, while the observations not selected for training form the test set. This approach aims to preserve not only the overall prevalence of the outcome,

$$P_{\text{train}}(\text{arrested}=1) \approx P_{\text{test}}(\text{arrested}=1),$$

but also the composition of the data across geographic, demographic, and temporal characteristics. This is desirable because differences in these characteristics could otherwise introduce an unintended distributional difference between the training and test sets. The temporary `.strat` variable is removed after the split because it is used solely to define the sampling groups and is not intended to be used as a predictor in the subsequent modelling.

Finally, the verification code compares the empirical marginal distributions of `arrested`, `area`, `sex`, and `ethnicity` in the two datasets. For a variable $X$ and level $k$, the proportion in each dataset is calculated as

$$\hat{P}_{train}(X=k)
=
\frac{\sum_{i \in \mathcal{D}_{train}} \mathbb{I}(X_i=k)}
{|\mathcal{D}_{train}|},$$

and analogously for the test set. The difference is then reported in **percentage points**:

$$\Delta_k =
100\left[
\hat{P}_{test}(X=k)
-
\hat{P}_{train}(X=k)
\right].$$

Small values of $\Delta_k$ indicate that the corresponding characteristic is similarly represented in the two datasets. The final summary additionally compares the number of observations and the proportion of arrested cases in the training and test sets, providing a direct check that the intended approximately 80/20 split and the class distribution have been maintained.

```{r split-verification}
bind_rows(
  check_prop("arrested", "arrested"),
  check_prop("area", "area"),
  check_prop("sex", "sex"),
  check_prop("ethnicity", "ethnicity")
) %>%
  kbl(
    digits = 4,
    caption = "Proportion of each level in train vs test (diff in percentage points)",
    col.names = c("Variable", "Level", "Train", "Test", "Δ (ppt)")
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE) %>%
  collapse_rows(columns = 1, valign = "top")
```

```{r split-summary}
cat(sprintf(
  "Training set : %d rows  |  Arrested: %.2f%%\nTest set     : %d rows  |  Arrested: %.2f%%\n",
  nrow(train), 100 * mean(train$arrested),
  nrow(test),  100 * mean(test$arrested)
))
```

The maximum absolute discrepancy between the two sets is on the order of fractions of a percentage point, confirming that the stratification preserved the joint distribution of the target and the key covariates.

## 1.4) Numerical Conversion

In order to be usable, `hour` and `age` must be integers, so we just add two newly converted versions of the same predictors.

```{r num-conv-train-test}
train <- train %>%
  mutate(
    hour_num = as.numeric(as.character(hour)),
    age_num = as.numeric(as.character(age))
  )

test <- test %>%
  mutate(
    hour_num = as.numeric(as.character(hour)),
    age_num = as.numeric(as.character(age))
  )
```

Finally, we remove unused data.

```{r remove-vehstop-train-test}
# Remove vehstop, train, test for memory purposes
rm(vehstop, train_rows)
```

------------------------------------------------------------------------

# 6) GAM

A Generalized Additive Model (GAM) extends a generalized linear model by allowing the effect of selected predictors to be represented through **flexible smooth functions** rather than assuming a strictly linear relationship.

For binary classification, GAMs can be formulated as an extension of logistic regression:

$$Y = \beta_0 + f_1(X_1) + f_2(X_2) + \cdots + f_p(X_p) + \epsilon$$

where $Y \in \left\{ 0, 1 \right\}$ and each $f_j(\cdot)$ is a smooth function estimated from the data, commonly using penalized regression splines.

**Why did we choose this model?** This flexibility is useful when predictors such as age, time of day, or other continuous variables may have non-linear relationships with the probability of arrest.

## 6.1) Assumptions and Setup

### 6.1.2) New Formulation — Adding Splines for `hour` and `age`

We formulate the model as follows:

```{r f-gam-model}
f_gam <- arrested ~ year + month + weekday + s(hour_num, bs = 'cc', k = 10) + boro + sex + ethnicity + s(age_num, bs = 'cr', k = 6)
```

The specification `s(hour_num, bs = "cc", k = 10)` uses a **cyclic cubic regression spline**, which is appropriate for hour of day because the beginning and end of the 24-hour cycle are connected. In fact, the EDA showed a non-linear enforcement cycle across the day that a linear slope or fixed brackets cannot fully capture.

Similarly we use `s(age_num, bs = "cr", k = 6)`, a cubic regression spline with a relatively modest maximum basis dimension. The reason for this is that the EDA revealed a non-monotone age-arrest relationship that neither a linear slope nor an arbitrary bracket optimally represents.

All other predictors (`year`, `month`, `weekday`, `sex`, `ethnicity`, `boro`) remain factors.

## 6.2) Model Fit

We fit with `mgcv::gam` using `family = binomial` and REML smoothness selection.

```{r fit-gam}
mod_gam <- mgcv::gam(f_gam, family = binomial, data = train, method = "REML")
```

## 6.3) Specific Analysis and Interpretation

```{r summary-gam}
summary(mod_gam)
```

The model summary indicates that there are nonlinear effects for `hour_num` and `age_num` (we will see later).

The model explains approximately $9.61\%$ ($R^{2} = 0.0391$) of the variation of the post-stop arrests in NY during the period 2023-2025. These values indicate that the **included variables explain only a limited proportion of the overall variation in arrest outcomes**. This is not unexpected for a rare binary outcome such as `arrested`, where many determinants of an arrest may not be represented by these predictor

### 6.3.1) Coefficients

#### 6.3.1.1) General Interpretation

The fitted GAM models the conditional probability of arrest using a binomial distribution with a logit link:

$$\log\left(\frac{P(\text{arrested}=1)}{1-P(\text{arrested}=1)}\right)
=
\beta_0 + \mathbf{X}\boldsymbol{\beta}
+ f_1(\text{hour})
+ f_2(\text{age}),$$

All parametric coefficients are expressed on the **log-odds scale**.

Exponentiating a coefficient gives the corresponding **odds ratio (OR)** $OR=e^\beta$. An $OR>1$ indicates higher odds of arrest relative to the reference category, while $OR<1$ indicates lower odds, holding all other variables constant.

#### 6.3.1.2) Specific Interpretation — Coefficients

The **intercept** is $-2.962$, corresponding to baseline odds of $e^{-2.962}\approx0.052$, or a probability of approximately $4.9\%$.

- However, the intercept should not be interpreted as the unconditional arrest rate. It refers to an **observation belonging to all categorical reference groups while the centered smooth effects are equal to zero**.

For the remaining coefficients, interpretations are conditional on all other predictors in the model.

Let us see the significant (at $5\%$) *time-related* coefficients first.

| Effect | Estimate | OR | Interpretation |
|------------------|-----------------:|-----------------:|-------------------|
| `year2024` | 0.135 | 1.144 | Approximately **14% higher odds** of arrest than 2023 |
| `year2025` | 0.209 | 1.232 | Approximately **23% higher odds** than 2023 |
| `month4` | 0.146 | 1.157 | April has approximately **16% higher odds** than January |
| `month5` | 0.143 | 1.153 | May has approximately **15% higher odds** than January |
| `month6` | 0.137 | 1.147 | June has approximately **15% higher odds** than January |
| `month7` | 0.210 | 1.234 | July has approximately **23% higher odds** than January |
| `month9` | 0.231 | 1.260 | September has approximately **26% higher odds** than January |
| `month10` | 0.146 | 1.157 | October has approximately **16% higher odds** than January |
| `weekdayTuesday` | 0.123 | 1.131 | Tuesday has approximately **13% higher odds** than Monday |
| `weekdayFriday` | -0.129 | 0.879 | Friday has approximately **12% lower odds** than Monday |
| `weekdaySunday` | -0.132 | 0.876 | Sunday has approximately **12% lower odds** than Monday |

As it is easy to see, there are other month and weekday coefficients but they are not "statistically different from their corresponding reference categories at the conventional $5\%$ level".

In particular,

- February, March, August, November and December do not differ significantly from January;

- Wednesday and Thursday similarly do not differ significantly from Monday, while Saturday shows weak evidence ($p=0.082$).

The coefficients from `boro` are all statistically significant at the same level:

| Borough category  | Estimate |    OR | Relative odds |
|-------------------|---------:|------:|--------------:|
| Brooklyn (North)  |   -0.878 | 0.416 | **58% lower** |
| Brooklyn (South)  |   -0.539 | 0.583 | **42% lower** |
| Manhattan (North) |   -1.120 | 0.326 | **67% lower** |
| Manhattan (South) |   -1.073 | 0.342 | **66% lower** |
| Queens (North)    |   -0.980 | 0.375 | **63% lower** |
| Queens (South)    |   -0.706 | 0.493 | **51% lower** |
| Staten Island     |   -0.474 | 0.623 | **38% lower** |

Plus, each coefficient has significantly lower conditional odds of arrest than the borough reference category (`Bronx`). These effects are **some of the strongest parametric associations** in the model.

For what regards `sex`'s coefficients

- We have a great association for `sexF`, $e^{-1.174}=0.309$. This means, the related observations have approximately **69% lower odds of arrest** than the reference sex category (`sexM`), fixing the remaining predictors;
- However, the coefficient for `sexU` is small and non-significant ($p=0.816$). In other words, we see no evidence of a difference from the reference category.

`ethnicity` also shows substantial conditional differences:

| Ethnicity                | Estimate |    OR |  Relative odds |
|--------------------------|---------:|------:|---------------:|
| Asian / Pacific Islander |   -0.675 | 0.509 |  **49% lower** |
| Black                    |    0.447 | 1.563 | **56% higher** |
| Hispanic                 |    0.642 | 1.900 | **90% higher** |
| Other                    |   -2.569 | 0.077 |  **92% lower** |
| Unknown                  |   -1.055 | 0.348 |  **65% lower** |

These comparisons are all relative to `White`.

For example (holding the other variables constant) the estimated odds of arrest for observations categorized as `Hispanic` are around **1.90 times** those of `White`. Instead, `Asian / Pacific Islander` have approximately half the odds.

*We move the discussion of smooth effects to the next section*.

### 6.3.2) Smooth Term and Plots

```{r gam-smooth-plots, fig.width=10, fig.height=4}
par(mfrow = c(1, 2), mar = c(4, 4, 2.5, 1))
plot(mod_gam, shade = TRUE, shade.col = "lightblue", seWithMean = TRUE,
     pages = 0, scale = 0)
par(mfrow = c(1, 1))
```

These plots now allow a much more specific interpretation of the two smooth terms than the GAM summary alone.

#### 6.3.2.1) General Interpretation

First of all we focus on **EDF** or Estimated Degrees of Freedom. They tell us **how complex the fitted smooth actually is after penalization and gives us indication about the linearity**.

- $EDF \approx 1$ would indicate an approximately linear effect or relationship.
- $EDF >> 1$ instead indicates evidence of a **non-linear relationship**.

Then, we have a value for $\chi^{2}$. This tests **whether the smooth term** **contributes to explaining the variance in the response after fixing the other terms in the model**.

- The null hypothesis is $H_{0} : f_{j}(x) = 0$ meaning that the predictor has no meaningful contribution to the linear predictor once the other variables are accounted for.

Finally, the **p-value** indicates how strong is the evidence that the smooth terms **are associated with the conditional probability of arrest (or whether the smooth contributes to the model)**, after fixing the other predictors.

For what concerns the plot, the **vertical axis** is the estimated **partial effect on the log-odds of arrest** (after fixing the remaining predictors and smooth terms). Instead, the horizontal axis follows the numerical range of values for the term.

The smooths are centered around zero, so:

- Regions where $f(x)>0$ indicates higher conditional log-odds of arrest relative to the model's average smooth effect (baseline), while
- $f(x)<0$ indicates lower conditional log-odds.

The magnitude can also be converted into a multiplicative odds effect using $e^{f(x)}$. The **dashed curves** represent uncertainty around the estimated smooth.

#### 6.3.2.2) Focus: Smooth effect of hour

The smooth term for `hour` is statistically significant:

$$\text{edf}=7.52,\qquad \chi^2=1164,\qquad p<2\times10^{-16}.$$

This provides very strong evidence that **time of day is associated with arrest probability after controlling for the other predictors**. In other words, the hourly pattern matters.

Furthermore, its **effect is strongly non-linear** ($\text{edf}=7.52 >> 1,\ p<2\times10^{-16}$).

Since the significance test is not enough to indicate which particular hours have the highest or lowest arrest probabilities, we look again at the picture.

The estimated arrest propensity (or "conditional log-odds of arrest relative to the model's average smooth effect")

1.  Is **relatively high** during the **late-night/very early morning**. In fact, we can see that from around 8 PM to 3 AM we have $> 0.5$, with a peak at **1-2 AM**;
2.  Then, it decreases during the early morning from approximately **4 AM onward** and finds its minimum around **7–8 AM**;
3.  After approximately 8 AM, the smooth rises rapidly and after around 12 PM it gradually increases through the afternoon and evening;
4.  By the **late evening (approximately 8 PM - 11 PM)**, it is clearly elevated.

Thus, the model provides further evidence towards the hypothesis we made during the EDA.

The fitted smooth $f_{\text{hour}}$ covers around $2.3$ log-odds units between its minimum and maximum. This corresponds to $\approx 10 \times$ difference/change in estimated arrest odds between the corresponding hours, conditional on the other variables in the model.

#### 6.3.2.3) Focus: Smooth effect of age

Age also has a highly significant smooth effect:

$$\text{edf}=4.66,\qquad
\chi^2=108,\qquad
p<2\times10^{-16}.$$

Thus, there is **strong evidence that arrest probability varies with age** after fixing the remaining predictors.

Again, $\text{edf}=4.66,\ p<2\times10^{-16}$ indicates that the relationship is meaningfully **non-linear** and statistically significant (although its variation is smaller than `hour`).

The curve indicates **relatively elevated conditional arrest propensity** among the **youngest drivers**. Then, we observe a local minimum at approximately **28–30 years**. From roughly **30-50 years** the age effect is close to zero. Finally, the curve rises to a local maximum at **60–62 years**, before a big decline for **65–70 years** (which have wider confidence intervals indicating uncertainty, consistent with the number of observations).

## 6.4) Evaluation on Test Set

```{r gam-eval}
prob_gam  <- predict(mod_gam, newdata = test, type = "response")
roc_gam   <- pROC::roc(test$arrested, prob_gam, quiet = TRUE)
auc_gam   <- as.numeric(pROC::auc(roc_gam))
thr_gam   <- best_thresh(roc_gam)
class_gam <- as.integer(prob_gam >= thr_gam)
met_gam   <- class_metrics(test$arrested, class_gam)

cat(sprintf(
  "AUC: %.4f  |  Threshold (Youden): %.4f\nSensitivity: %.3f  |  Specificity: %.3f  |  Balanced accuracy: %.3f\n",
  auc_gam, thr_gam, met_gam$sensitivity, met_gam$specificity, met_gam$balanced_acc
))
```

The Youden threshold is **0.0554**, way below the 0.5 cutoff.

Moreover, the **AUC of 0.6897** indicates **moderate discriminatory ability**: the GAM separates arrest from non-arrest cases a bit better than chance.

### 6.4.1) Confusion Matrix

```{r gam-cm-plot, fig.width=4, fig.height=3.5}
plot_cm(test$arrested, class_gam, title = "GAM — Confusion Matrix (Youden threshold)")
```

The confusion matrix shows

- **1,496 true positives**;
- **816 false negatives**;
- **22,008 true negatives**;
- And **12,729 false positives**.

Now for the derived metrics:

| Metric | Value | Interpretation |
|-------------------------------|-----------------:|-----------------------|
| Accuracy | 0.634 | About 63.4% of all observations are classified correctly |
| Sensitivity / Recall | 0.647 | The model correctly identifies about 64.7% of actual arrests |
| Specificity | 0.634 | About 63.4% of actual non-arrests are correctly classified |
| Precision | 0.105 | Only about 10.5% of predicted arrests are actual arrests |
| F1 | 0.181 | The low value reflects the difficult trade-off between recall and precision for the minority class |
| Balanced Accuracy | 0.640 | Averaging sensitivity and specificity gives a more appropriate summary than ordinary accuracy in this imbalanced setting |

### 6.4.2) Performance Analysis

This was partly covered in the previous section, so here we will be focusing on the metrics that we have yet to mention.

```{r gam-metrics-table}
compute_metrics(test$arrested, prob_gam, "GAM") %>%
  kbl(
    caption   = "GAM — full evaluation metrics (test set, Youden threshold)",
    col.names = c("Model","AUC","PR-AUC","Log-Loss","Brier",
                  "Threshold","Sensitivity","Specificity","Precision","F1","MCC","Bal. Acc.")
  ) %>%
  kable_styling(bootstrap_options = c("striped","condensed"), full_width = TRUE)
```

```{r gam-eval-plots, fig.width=10, fig.height=8}
eval_quad_plot(test$arrested, prob_gam, "GAM")
```

#### 6.4.2.1) Overall performance

| Metric | Value | Interpretation |
|----------------|---------------:|----------------------------------------|
| **AUC** | 0.6897 | Moderate discrimination between arrests and non-arrests |
| **PR-AUC** | 0.1247 | Above the no-skill baseline of $\approx 0.062$, but still low in absolute terms |
| **Log-Loss** | 0.2235 | Predicted probabilities contain useful information, though improvement over a prevalence-only model is modest |
| **Brier** | 0.0570 | Reasonably accurate probability estimates and slightly better than the prevalence-only baseline |
| Threshold | 0.0554 | Low cutoff selected by Youden's criterion, appropriate for the rare positive class |
| **MCC** | 0.140 | Weak positive association between predicted and observed classes |
| Balanced Accuracy | 0.640 | Moderate performance when both classes are weighted equally |

The main conclusion is that the GAM captures **meaningful but limited predictive signal**. It manages to discriminate arrests from non-arrests better than chance and produces reasonably calibrated probabilities, but its **ability to make precise positive classifications is still weak**.

#### 6.4.2.2) AUC

The test-set AUC is $\approx 0.690$.

This means that if we randomly select (i) one observation where an arrest occurred, and (ii) one observation where an arrest did not occur, the model assigns a higher predicted probability to the arrest observation about 69% of the time.

The ROC curve **remains consistently above the main diagonal**, indicating that the model contains useful ranking information. However, an AUC below $0.70$ also shows that there is still a great deal of overlap between the predicted probabilities of the two classes.

#### 6.4.2.3) PR-AUC

This is particularly important given our imbalanced outcome.

Our positive class is rare, so ROC-AUC can sometimes give an overly optimistic impression of performance. **Precision-Recall AUC focuses directly on the positive class**, considering the trade-off between precision and recall.

Here, we obtain:

$$\text{PR-AUC}=0.1247$$

The baseline PR-AUC is approximately equal to the positive-class prevalence:

$$\text{PR-AUC}_{\text{baseline}}\approx0.062$$

*But how good is* $0.1247$? It is roughly **twice the baseline**, indicating that the **GAM does provide useful information for ranking observations according to arrest risk**. However, the precision is still very low.

#### 6.4.2.4) Log-Loss

> Log-loss evaluates the **quality of the predicted probabilities**, rather than simply stating whether the final classification is correct.

This measure cares about **how confident the model is**. In our case:

$$\text{Log-Loss}=0.2235$$

TODO: Rephrase the following.

In order to give some more context, a simple model assigning to every observation the test-set arrest prevalence of approximately $6.2\%$ would obtain a log-loss of about $0.234$.

The GAM **improves on this simple baseline, although very little**. Its probability estimates contain useful information, but a considerable amount of uncertainty in the outcome remains unexplained.

#### 6.4.2.5) Brier Score

The Brier score is useful here because it evaluates whether the predicted probabilities are close to the observed outcomes, rather than merely whether the model ranks observations correctly.

> Formally, the Brier score measures the squared error of the predicted probabilities (the lower, the better).

For the GAM,

$$\text{Brier}=0.0570$$

A baseline model would obtain approximately

$$0.0585$$

so the GAM performs just slightly better.

Together with the calibration plot, this suggests that the predicted probabilities are **reasonably calibrated overall**: **groups assigned higher arrest probabilities tend to show correspondingly higher observed arrest rates**.

#### 6.4.2.6) MCC

The **Matthews Correlation Coefficient (MCC)** is particularly useful for imbalanced classification because it incorporates all four components of the confusion matrix:

$$MCC =
\frac{TP\cdot TN-FP\cdot FN}
{\sqrt{(TP+FP)(TP+FN)(TN+FP)(TN+FN)}}
\in [-1,1]$$

where:

- $1$ = perfect prediction
- $0$ = no association between predictions and outcomes
- $-1$ = completely incorrect prediction

Our result is:

$$MCC=0.140$$

Thus, there is a **positive but weak association between predictions and actual outcomes** (consistent with CM). Simply put:

- The model is doing meaningfully better than random classification;
- But that its classification performance is very limited.

#### 6.4.2.7) Thresholds

The classification threshold selected using **Youden's index** is

$$t^*=0.0554$$

The threshold analysis also shows the expected trade-off: lowering the threshold increases sensitivity but reduces specificity, while increasing it improves specificity at the cost of missing more arrests.

#### 6.4.2.8) Final Remarks

Let us put the metrics together:

- **Discrimination**, $\text{AUC} \approx 0.690$: the GAM has moderate ability to rank arrests above non-arrests;
- **Rare-class performance**, $\text{PR-AUC} \approx 0.125$: clearly better than the $\approx0.062$ no-skill baseline, but positive-class identification remains difficult;
- **Probability predictions**, $\text{Brier} \approx 0.057,\; \text{LogLoss} \approx 0.224$: probabilities are reasonably calibrated and improve modestly over a prevalence-only model;
- **Classification**, $\text{MCC} \approx 0.140$: binary predictions show only a weak association with the true outcomes;
- **Class balance**, $\text{ACC}_{\text{Balanced}} \approx 0.640$: sensitivity and specificity are reasonably balanced, but overall classification performance remains moderate.

Now for the final conclusions:

- The GAM captures a **real but limited predictive signal** in the available demographic, temporal, and geographical variables.
- Its strongest result is its ability to **rank observations by relative arrest risk**, rather than to produce highly accurate binary classifications.
- The low positive-class prevalence strongly affects precision: obtaining a sensitivity of $0.647$ requires accepting a substantial number of false positives.
- The probability estimates appear more useful than the thresholded classifications, particularly given the reasonably good calibration.
- Consequently, the GAM is best interpreted as a model of **relative arrest propensity and estimated probability**, rather than as a highly accurate classifier of individual arrests.
