---
title: "Predicting Excessive Enforcement Patterns in Police Behavior After Vehicles Stops in New York"
subtitle: "Model Selection and Evaluation - GAM 4-Wheels"
author: "Gianmaria Pizzo"
output:
  html_document:
    fig_caption: yes
    theme: flatly
    highlight: pygments
    code_folding: show
    toc: yes
    toc_depth: 4
    number_sections: no
    toc_float:
      smooth_scroll: no
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

These `.Rmd` files represent the project submission for the course [Statistical Inference and Learning](https://www.unive.it/data/course/521944) by Professor [Cristiano Varin](https://www.unive.it/data/people/5592728) at [Ca' Foscari University of Venice](https://www.unive.it/) for the year 2025/2026.

------------------------------------------------------------------------

# 0) Environment

## 0.1) Packages

```{r requirements}
requirements <- c(
  "scales", "dplyr", "tidyr", "viridis",
  "ggplot2", "patchwork", "pROC", "kableExtra", "mgcv"
)
```

```{r install-requirements, include=FALSE}
for (req in requirements){
  if (!require(req, character.only = TRUE)){
      install.packages(req)
  }
}
```

The list only contains packages actually used in this notebook. Now, the loading of all packages should be immediate.

```{r load-requirements, include=FALSE}
for (req in requirements){
  if (!require(req, character.only = TRUE)){
      library(req, character.only = TRUE)
  }
}
```

```{r remove-vars, include=FALSE}
rm(requirements, req)
```

## 0.2) Functions

Here we define the helper functions used throughout model selection and evaluation, kept in one place for reuse across every model fitted below.

### 0.2.1) Data-Split Diagnostics

```{r fn-check-prop}
# ==============================================================================
# 1. Train / Test Split Diagnostics
# ==============================================================================

#' @title Check Train/Test Level Proportions
#'
#' @description Tabulates one variable's level proportions separately in
#' `train` and `test` and reports the discrepancy in percentage points. Used
#' to verify that the stratified split preserved the marginal distribution of
#' key covariates.
#'
#' @param var The column name (present in both `train` and `test`) to tabulate.
#' @param label Human-readable label for the variable, stored in the output.
#'
#' @return A tibble with columns variable, level, train, test, diff_ppt.
check_prop <- function(var, label) {
  # Compute the relative frequency of each level in the training set
  tr <- prop.table(table(train[[var]]))
  # Compute the relative frequency of each level in the test set
  te <- prop.table(table(test[[var]]))
  # Create a summary table comparing train and test proportions
  tibble(
    # Variable levels
    level = names(tr),
    # Proportion in training set
    train = as.numeric(tr),
    # Proportion in test set
    test = as.numeric(te),
    # Difference in % between test and training proportions
    diff_ppt = round((as.numeric(te) - as.numeric(tr)) * 100, 2)
  ) %>%
    # Add descriptive variable name
    mutate(variable = label) %>% 
    # Move variable name to first column
    relocate(variable)
}
```

### 0.2.2) Threshold and Confusion-Matrix Helpers

```{r fn-eval-helpers}
# ==============================================================================
# 2. Threshold Selection and Confusion-Matrix Helpers
# ==============================================================================

#' @title Find the Youden-Optimal Classification Threshold
#'
#' @description Finds the probability threshold on an ROC curve that
#' maximises Youden's J statistic (sensitivity + specificity − 1), used
#' throughout as the single decision threshold for turning predicted
#' probabilities into classes.
#'
#' @param roc_obj A pROC "roc" object, as returned by pROC::roc().
#'
#' @return A numeric scalar, the optimal threshold on the probability scale.
best_thresh <- function(roc_obj) {
  pROC::coords(
    roc_obj, "best", ret = "threshold", 
    best.method = "youden", quiet = TRUE
  )$threshold
}


#' @title Tally a Confusion Matrix
#'
#' @description Cross-tabulates actual vs predicted binary class labels.
#'
#' @param actual Integer/numeric vector of 0/1 true labels.
#' @param predicted Integer/numeric vector of 0/1 predicted labels.
#'
#' @return A tibble with columns Actual, Predicted, n (one row per non-empty cell).
conf_tib <- function(actual, predicted) {
  tibble(
    Actual = factor(actual, levels = c(0, 1)),
    Predicted = factor(predicted, levels = c(0, 1))
  ) %>% 
    count(Actual, Predicted)
}


#' @title Plot a Confusion-Matrix Heatmap
#'
#' @description Renders conf_tib() output as a 2x2 tile plot with cell counts.
#'
#' @param actual Integer/numeric vector of 0/1 true labels.
#' @param predicted Integer/numeric vector of 0/1 predicted labels.
#' @param title Main title of the plot.
#'
#' @return A ggplot2 object.
plot_cm <- function(actual, predicted, title = "Confusion Matrix") {
  conf_tib(actual, predicted) %>%
    ggplot(aes(Predicted, Actual, fill = n)) +
    geom_tile(colour = "white") +
    geom_text(aes(label = scales::comma(n)), size = 5.5, fontface = "bold") +
    scale_fill_gradient(low = "#e8f4f8", high = "#2c7bb6", guide = "none") +
    scale_x_discrete(labels = c("0" = "No arrest", "1" = "Arrest")) +
    scale_y_discrete(labels = c("0" = "No arrest", "1" = "Arrest")) +
    labs(title = title, x = "Predicted", y = "Actual") +
    theme_minimal(base_size = 12) +
    theme(panel.grid = element_blank())
}


#' @title Compute Threshold-Dependent Classification Metrics
#'
#' @description Computes sensitivity, specificity, balanced accuracy, PPV,
#' and NPV for a fixed set of hard class predictions against the true labels.
#'
#' @param actual Integer/numeric vector of 0/1 true labels.
#' @param predicted Integer/numeric vector of 0/1 predicted labels.
#' @param pos Value in `actual`/`predicted` denoting the positive class.
#'
#' @return A named list with sensitivity, specificity, balanced_acc, ppv, npv.
class_metrics <- function(actual, predicted, pos = 1) {
  tp <- as.numeric(sum(actual == pos & predicted == pos))
  tn <- as.numeric(sum(actual != pos & predicted != pos))
  fp <- as.numeric(sum(actual != pos & predicted == pos))
  fn <- as.numeric(sum(actual == pos & predicted != pos))
  sens <- tp / (tp + fn)
  spec <- tn / (tn + fp)
  list(
    sensitivity = sens,
    specificity = spec,
    balanced_acc = (sens + spec) / 2,
    ppv = tp / (tp + fp),
    npv = tn / (tn + fn)
  )
}


# ==============================================================================
# 3. Probability-Based Evaluation Metrics
# ==============================================================================

#' @title Compute the Full Evaluation-Metrics Tibble for One Model
#'
#' @description Given predicted probabilities and true labels, computes
#' threshold-free metrics (AUC, PR-AUC, log-loss, Brier score) plus
#' threshold-dependent metrics at the Youden-optimal cut-off (sensitivity,
#' specificity, precision, F1, MCC, balanced accuracy). Returns one row per
#' call, so results from multiple models can be row-bound into a comparison
#' table.
#'
#' @param actual Integer/numeric vector of 0/1 true labels.
#' @param prob Numeric vector of predicted P(arrested = 1), same length as `actual`.
#' @param thr Classification threshold selected externally (on the training set),
#' applied here rather than recomputed from `actual`/`prob` to avoid leaking test labels.
#' @param model_name Label stored in the Model column.
#'
#' @return A one-row tibble of rounded metrics (see column names in the body).
compute_metrics <- function(actual, prob, thr, model_name = "Model") {
  valid <- !is.na(actual) & !is.na(prob)
  actual <- actual[valid]
  prob <- prob[valid]

  roc_obj <- pROC::roc(actual, prob, quiet = TRUE)
  auc_val <- as.numeric(pROC::auc(roc_obj))
  pred_cl <- as.integer(prob >= thr)
  
  # Confusion Matrix Metrics
  tp <- as.numeric(sum(actual == 1 & pred_cl == 1))
  tn <- as.numeric(sum(actual == 0 & pred_cl == 0))
  fp <- as.numeric(sum(actual == 0 & pred_cl == 1))
  fn <- as.numeric(sum(actual == 1 & pred_cl == 0))

  # Confusion Matrix Derived Metrics
  sens <- (
    if ((tp+fn) == 0) NA_real_ 
    else tp / (tp+fn)
  )
  spec <- (
    if ((tn+fp) == 0) NA_real_ 
    else tn / (tn+fp)
  )
  prec <- (
    if ((tp+fp) == 0) NA_real_ 
    else tp / (tp+fp)
  )
  f1 <- (
    if (is.na(prec) || is.na(sens) || (prec+sens) == 0) NA_real_ 
    else 2 * prec * sens / (prec + sens)
  )

  # MCC
  mcc_den <- sqrt(
    as.numeric(tp+fp) * as.numeric(tp+fn) *
    as.numeric(tn+fp) * as.numeric(tn+fn)
  )
  mcc_val <- (
    if (mcc_den == 0) NA_real_ 
    else (tp*tn - fp*fn) / mcc_den
  )

  # Logloss
  eps <- 1e-15
  logloss <- -mean(
    actual * log(pmax(prob, eps)) + 
    (1-actual) * log(pmax(1-prob, eps))
  )
  
  # Brier Score
  brier <- mean((prob - actual)^2)

  # PR-AUC via trapezoid rule over pROC coords (pROC has no built-in PR-AUC)
  pr_raw <- pROC::coords(
    roc_obj, "all", ret = c("sensitivity", "precision"),
    transpose = FALSE, drop = FALSE
  )
  
  names(pr_raw)[names(pr_raw) == "sensitivity"] <- "recall"

  pr_raw <- pr_raw %>%
    filter(!is.nan(precision), !is.na(precision), !is.nan(recall)) %>%
    arrange(recall)

  pr_auc_val <- if (nrow(pr_raw) >= 2)
    sum(diff(pr_raw$recall) * (head(pr_raw$precision, -1) + tail(pr_raw$precision, -1)) / 2)
  else NA_real_

  # Print Metrics
  tibble(
    Model = model_name,
    AUC = round(auc_val, 4),
    PR_AUC = round(pr_auc_val, 4),
    Log_Loss = round(logloss, 4),
    Brier = round(brier, 4),
    Threshold = round(thr, 4),
    Sensitivity = round(sens, 3),
    Specificity = round(spec, 3),
    Precision = round(prec, 3),
    F1 = round(f1, 3),
    MCC = round(mcc_val, 3),
    Bal_Acc = round(
      if (!is.na(sens) && !is.na(spec)) (sens+spec)/2
      else NA_real_, 3
    )
  )
}
```

### 0.2.3) Model Evaluation Plot Functions

```{r fn-plot-roc-single}
# ==============================================================================
# 4. Plots for Single-Model Evaluation
# ==============================================================================

#' @title Plot the ROC Curve for a Single Model
#'
#' @description Plots sensitivity vs specificity across all thresholds, with
#' the AUC annotated on the panel.
#'
#' @param actual Integer/numeric vector of 0/1 true labels.
#' @param prob Numeric vector of predicted P(arrested = 1), same length as `actual`.
#' @param title Main title of the plot.
#'
#' @return A ggplot2 object.
plot_roc_single <- function(actual, prob, title = "ROC Curve") {
  roc_obj <- pROC::roc(actual, prob, quiet = TRUE)
  auc_lbl <- paste0("AUC = ", round(as.numeric(pROC::auc(roc_obj)), 4))
  pROC::ggroc(roc_obj, colour = "#2c7bb6", linewidth = 0.9) +
    geom_abline(slope = 1, intercept = 1, linetype = "dashed", colour = "grey60") +
    annotate("text", x = 0.25, y = 0.06, label = auc_lbl, colour = "#2c7bb6", size = 4) +
    labs(x = "Specificity", y = "Sensitivity", title = title) +
    theme_minimal(base_size = 11) + theme(panel.grid.minor = element_blank())
}
```

```{r fn-plot-pr-single}
#' @title Plot the Precision-Recall Curve for a Single Model
#'
#' @description Plots precision vs recall across all thresholds against the
#' no-skill baseline (the class prevalence), with PR-AUC annotated. More
#' informative than the ROC curve under the severe class imbalance here.
#'
#' @param actual Integer/numeric vector of 0/1 true labels.
#' @param prob Numeric vector of predicted P(arrested = 1), same length as `actual`.
#' @param title Main title of the plot.
#'
#' @return A ggplot2 object.
plot_pr_single <- function(actual, prob, title = "Precision-Recall Curve") {
  roc_obj <- pROC::roc(actual, prob, quiet = TRUE)
  pr_raw  <- pROC::coords(roc_obj, "all",
                           ret = c("sensitivity", "precision"),
                           transpose = FALSE, drop = FALSE)
  names(pr_raw)[names(pr_raw) == "sensitivity"] <- "recall"
  pr_df <- pr_raw %>%
    filter(!is.nan(precision), !is.na(precision)) %>% arrange(recall)

  # PR-AUC via trapezoid rule, same approach as in compute_metrics()
  pr_auc_lbl <- if (nrow(pr_df) >= 2) {
    v <- sum(diff(pr_df$recall) *
               (head(pr_df$precision,-1) + tail(pr_df$precision,-1)) / 2)
    paste0("PR-AUC = ", round(v, 4))
  } else "PR-AUC = NA"
  baseline <- mean(actual)
  ggplot(pr_df, aes(x = recall, y = precision)) +
    geom_line(colour = "#d73027", linewidth = 0.9) +
    geom_hline(yintercept = baseline, linetype = "dashed", colour = "grey55") +
    annotate("text", x = 0.02, y = baseline + 0.015,
             label = paste0("No-skill = ", round(baseline, 3)),
             colour = "grey40", size = 3.2, hjust = 0) +
    annotate("text", x = 0.55, y = 0.90, label = pr_auc_lbl,
             colour = "#d73027", size = 4) +
    scale_x_continuous(limits = c(0, 1)) + scale_y_continuous(limits = c(0, 1)) +
    labs(x = "Recall", y = "Precision", title = title) +
    theme_minimal(base_size = 11) + theme(panel.grid.minor = element_blank())
}
```

```{r fn-plot-calibration-single}
#' @title Plot the Calibration (Reliability) Curve for a Single Model
#'
#' @description Bins predictions into `n_bins` equal-frequency groups and
#' plots mean predicted probability against observed positive rate per bin,
#' against the perfect-calibration diagonal.
#'
#' @param actual Integer/numeric vector of 0/1 true labels.
#' @param prob Numeric vector of predicted P(arrested = 1), same length as `actual`.
#' @param title Main title of the plot.
#' @param n_bins Number of quantile bins of `prob`.
#'
#' @return A ggplot2 object.
plot_calibration_single <- function(actual, prob, title = "Calibration Plot",
                                    n_bins = 10) {
  # Quantile breaks (not equal-width) so each bin has a comparable sample size
  # despite the predicted probabilities being heavily right-skewed
  brks  <- unique(quantile(prob, seq(0, 1, length.out = n_bins + 1), na.rm = TRUE))
  cal_df <- tibble(actual = actual, prob = prob) %>%
    mutate(bin = cut(prob, breaks = brks, include.lowest = TRUE)) %>%
    filter(!is.na(bin)) %>%
    group_by(bin) %>%
    summarise(mean_pred = mean(prob), mean_obs = mean(actual),
              n = n(), .groups = "drop")
  ggplot(cal_df, aes(x = mean_pred, y = mean_obs)) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey55") +
    geom_line(colour = "#2c7bb6", linewidth = 0.8) +
    geom_point(aes(size = n), colour = "#2c7bb6", alpha = 0.85) +
    scale_size_continuous(name = "Count", range = c(2, 9)) +
    scale_x_continuous(labels = scales::percent_format(accuracy = 0.1)) +
    scale_y_continuous(labels = scales::percent_format(accuracy = 0.1)) +
    labs(x = "Mean predicted probability", y = "Observed positive rate",
         title = title,
         caption = "Size ∝ bin count.  Dashed = perfect calibration.") +
    theme_minimal(base_size = 11) +
    theme(legend.position = "bottom", panel.grid.minor = element_blank())
}
```

```{r fn-plot-threshold-single}
#' @title Plot the Threshold-Sweep Metrics for a Single Model
#'
#' @description Plots sensitivity, specificity, precision, and F1 as
#' functions of the classification threshold, with the Youden-optimal
#' threshold marked.
#'
#' @param actual Integer/numeric vector of 0/1 true labels.
#' @param prob Numeric vector of predicted P(arrested = 1), same length as `actual`.
#' @param thr Classification threshold selected externally (on the training set).
#' @param title Main title of the plot.
#'
#' @return A ggplot2 object.
plot_threshold_single <- function(actual, prob, thr, title = "Threshold Analysis") {
  roc_obj  <- pROC::roc(actual, prob, quiet = TRUE)
  thr_data <- pROC::coords(roc_obj, "all",
                            ret = c("threshold","sensitivity","specificity","precision"),
                            transpose = FALSE, drop = FALSE) %>%
    filter(is.finite(threshold), threshold >= 0, threshold <= 1) %>%
    mutate(f1 = ifelse(!is.nan(precision) & (precision + sensitivity) > 0,
                       2 * precision * sensitivity / (precision + sensitivity), 0)) %>%
    pivot_longer(c(sensitivity, specificity, precision, f1),
                 names_to = "Metric", values_to = "Value") %>%
    mutate(Metric = recode(Metric,
      sensitivity = "Sensitivity", specificity = "Specificity",
      precision   = "Precision",   f1          = "F1 Score"))
  ggplot(thr_data, aes(x = threshold, y = Value, colour = Metric)) +
    geom_line(linewidth = 0.75, na.rm = TRUE) +
    geom_vline(xintercept = thr, linetype = "dashed",
               colour = "grey35", linewidth = 0.55) +
    annotate("text", x = thr, y = 0.04,
             label = paste0("Train-selected\n", round(thr, 3)),
             hjust = -0.08, size = 3, colour = "grey30") +
    scale_colour_viridis_d(option = "turbo", begin = 0.05, end = 0.92) +
    scale_y_continuous(limits = c(0, 1)) +
    labs(x = "Threshold", y = "Metric value", colour = NULL, title = title,
         caption = "Dashed: threshold selected on the training set") +
    theme_minimal(base_size = 11) +
    theme(legend.position = "bottom", panel.grid.minor = element_blank())
}
```

```{r fn-plot-eval-quad}
#' @title Plot the 2x2 Evaluation Panel for a Single Model
#'
#' @description Composes plot_roc_single(), plot_pr_single(),
#' plot_calibration_single(), and plot_threshold_single() into one
#' (ROC | PR) / (Calibration | Threshold) patchwork layout.
#'
#' @param actual Integer/numeric vector of 0/1 true labels.
#' @param prob Numeric vector of predicted P(arrested = 1), same length as `actual`.
#' @param thr Classification threshold selected externally (on the training set).
#' @param model_name Prefixed to each sub-plot's title.
#'
#' @return A patchwork object (composed ggplot2 object).
eval_quad_plot <- function(actual, prob, thr, model_name) {
  p1 <- plot_roc_single(actual, prob,   paste(model_name, "— ROC"))
  p2 <- plot_pr_single(actual, prob,    paste(model_name, "— Precision-Recall"))
  p3 <- plot_calibration_single(actual, prob, paste(model_name, "— Calibration"))
  p4 <- plot_threshold_single(actual, prob, thr, paste(model_name, "— Threshold Analysis"))
  (p1 | p2) / (p3 | p4)
}
```

------------------------------------------------------------------------

# 1) Data Loading and Preparation

```{r load-data}
vehstop <- read.csv("data/vehiclestops_4_wheels_final.csv")
```

## 1.1) Factor conversion and Reference Levels

We convert all categorical predictors to factors and set interpretable reference levels before any modelling step, ensuring consistency between all downstream model fits and cross-validation folds.

```{r factor-prep}
# Ordered weekday levels (ISO order)
wday_levels <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday")

# Ordered time bracket levels
tbkt_levels <- c(
  "Late night (0–3)",
  "Early morning (4–7)",
  "Morning commute (8–11)",
  "Midday (12–15)",
  "Afternoon/evening (16–19)",
  "Night (20–23)"
)

# Ordered age-bracket levels
abkt_levels <- c("Under 18", "18–24", "25–34", "35–44", "45–54", "55–64", "65+", "Unknown")

vehstop <- vehstop %>%
  mutate(
    year = factor(year),
    month = factor(month, levels = 1:12),
    weekday = factor(weekday, levels = wday_levels),
    hour = factor(hour, levels = 0:23),
    time_bracket = factor(time_bracket, levels = tbkt_levels),
    area = factor(area),
    boro = factor(boro),
    sex = factor(sex),
    ethnicity = factor(ethnicity),
    age_bracket = factor(age_bracket, levels = abkt_levels)
  )

# Reference levels chosen for interpretable contrasts
vehstop$sex <- relevel(vehstop$sex, ref = "M")
vehstop$ethnicity <- relevel(vehstop$ethnicity, ref = "White")
vehstop$weekday <- relevel(vehstop$weekday, ref = "Monday")
vehstop$time_bracket <- relevel(vehstop$time_bracket, ref = "Morning commute (8–11)")
vehstop$age_bracket <- relevel(vehstop$age_bracket, ref = "25–34")

rm(wday_levels, tbkt_levels, abkt_levels)
```

## 1.2) Removing NAs from `age`

We had to exclude `age` as a predictor from the other models, since we had missing values that we labeled as `Unknown` in `age_bracket`. Here instead we want to include it so we will drop the NAs and use a smoothing curve.

```{r drop-age-na}
vehstop <- vehstop %>% 
  dplyr::filter(!is.na(age))
```

> Note: This makes GAM not directly comparable with the other models because the dataset changes with respect to this filtering. Therefore, we will not just say "GAM performs better than Logistic/NB/LASSO" but we will take into account the interesting aspects that emerge.

## 1.3) Stratified Train / Test Split

For the train/test split, a **stratified sampling approach** is used to ensure that the training and test sets have comparable distributions of the outcome variable (`arrested`) and relevant contextual variables.

This is particularly important because the target variable is imbalanced, with arrested cases representing only a small proportion of all observations. Rather than stratifying only on `arrested`, the code constructs a **joint stratification variable** from `arrested`, `area`, `sex`, `ethnicity`, and `hour` using `interaction()`. Thus, each observation is assigned to a stratum corresponding to a particular combination of these variables.

Formally, the strata can be represented as

$$S_i = (\text{arrested}_i,\text{area}_i,\text{sex}_i,\text{ethnicity}_i,\text{hour}_i),$$

where $S_i$ denotes the stratum to which observation $i$ belongs. The objective is to sample approximately 80% of observations within each stratum for the training set and retain the remaining 20% for testing:

$$P(i \in \mathcal{D}_{train} \mid S_i=s) \approx 0.8,
\qquad
P(i \in \mathcal{D}_{test} \mid S_i=s) \approx 0.2.$$

```{r stratified-split}
set.seed(2026)

# Joint stratum key; drop = TRUE removes empty combinations
vehstop$.strat <- interaction(
  vehstop$arrested, vehstop$area, vehstop$sex, vehstop$ethnicity, vehstop$hour,
  drop = TRUE
)

train_rows <- vehstop %>%
  dplyr::mutate(.row = row_number()) %>%
  dplyr::group_by(.strat) %>%
  dplyr::slice_sample(prop = 0.8) %>%
  dplyr::pull(.row)

train <- vehstop[ train_rows, ] %>% dplyr::select(-.strat)
test <- vehstop[-train_rows, ] %>% dplyr::select(-.strat)
vehstop <- vehstop %>% dplyr::select(-.strat)
```

In the code, `group_by(.strat) %>% slice_sample(prop = 0.8)` performs this within-stratum sampling, while the observations not selected for training form the test set. This approach aims to preserve not only the overall prevalence of the outcome,

$$P_{\text{train}}(\text{arrested}=1) \approx P_{\text{test}}(\text{arrested}=1),$$

but also the composition of the data across geographic, demographic, and temporal characteristics. This is desirable because differences in these characteristics could otherwise introduce an unintended distributional difference between the training and test sets. The temporary `.strat` variable is removed after the split because it is used solely to define the sampling groups and is not intended to be used as a predictor in the subsequent modelling.

Finally, the verification code compares the empirical marginal distributions of `arrested`, `area`, `sex`, and `ethnicity` in the two datasets. For a variable $X$ and level $k$, the proportion in each dataset is calculated as

$$\hat{P}_{train}(X=k)
=
\frac{\sum_{i \in \mathcal{D}_{train}} \mathbb{I}(X_i=k)}
{|\mathcal{D}_{train}|},$$

and analogously for the test set. The difference is then reported in **percentage points**:

$$\Delta_k =
100\left[
\hat{P}_{test}(X=k)
-
\hat{P}_{train}(X=k)
\right].$$

Small values of $\Delta_k$ indicate that the corresponding characteristic is similarly represented in the two datasets. The final summary additionally compares the number of observations and the proportion of arrested cases in the training and test sets, providing a direct check that the intended approximately 80/20 split and the class distribution have been maintained.

```{r split-verification}
bind_rows(
  check_prop("arrested", "arrested"),
  check_prop("area", "area"),
  check_prop("sex", "sex"),
  check_prop("ethnicity", "ethnicity")
) %>%
  kbl(
    digits = 4,
    caption = "Proportion of each level in train vs test (diff in percentage points)",
    col.names = c("Variable", "Level", "Train", "Test", "Δ (ppt)")
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE) %>%
  collapse_rows(columns = 1, valign = "top")
```

```{r split-summary}
cat(sprintf(
  "Training set : %d rows  |  Arrested: %.2f%%\nTest set     : %d rows  |  Arrested: %.2f%%\n",
  nrow(train), 100 * mean(train$arrested),
  nrow(test),  100 * mean(test$arrested)
))
```

The maximum absolute discrepancy between the two sets is on the order of fractions of a percentage point, confirming that the stratification preserved the joint distribution of the target and the key covariates.

## 1.4) Numerical Conversion

In order to be usable, `hour` and `age` must be integers, so we just add two newly converted versions of the same predictors.

```{r num-conv-train-test}
train <- train %>%
  mutate(
    hour_num = as.numeric(as.character(hour)),
    age_num = as.numeric(as.character(age))
  )

test <- test %>%
  mutate(
    hour_num = as.numeric(as.character(hour)),
    age_num = as.numeric(as.character(age))
  )
```

Finally, we remove unused data.

```{r remove-vehstop-train-test}
# Remove vehstop, train, test for memory purposes
rm(vehstop, train_rows)
```

------------------------------------------------------------------------

# 6) GAM

A Generalized Additive Model (GAM) extends a generalized linear model by allowing the effect of selected predictors to be represented through **flexible smooth functions** rather than assuming a strictly linear relationship.

For binary classification, GAMs can be formulated as an extension of logistic regression:

$$\log \left( \frac{P(\texttt{arrested} = 1 \mid X)}{1-P(\texttt{arrested} = 1 \mid X)} \right) = \beta_{0} + \mathbf{X}\boldsymbol{\beta} + f_{1}(\text{hour}) + f_{2}(\text{age})$$

where $\texttt{arrested} = Y \in \left\{ 0, 1 \right\}$ and each $f_j(\cdot)$ is a smooth function estimated from the data, commonly using penalized regression splines.

**Why did we choose this model?** This flexibility is useful when predictors such as age, time of day, or other continuous variables may have non-linear relationships with the probability of arrest.

## 6.1) Assumptions and Setup

### 6.1.2) New Formulation — Adding Splines for `hour` and `age`

We formulate the model as follows:

```{r f-gam-model}
f_gam <- arrested ~ year + month + weekday + s(hour_num, bs = 'cc', k = 10) + boro + sex + ethnicity + s(age_num, bs = 'cr', k = 6)
```

The specification `s(hour_num, bs = "cc", k = 10)` uses a **cyclic cubic regression spline**, which is appropriate for hour of day because the beginning and end of the 24-hour cycle are connected. In fact, the EDA showed a non-linear enforcement cycle across the day that a linear slope or fixed brackets cannot fully capture.

Similarly we use `s(age_num, bs = "cr", k = 6)`, a cubic regression spline with a relatively modest maximum basis dimension. The reason for this is that the EDA revealed a non-monotone age-arrest relationship that neither a linear slope nor an arbitrary bracket optimally represents.

All other predictors (`year`, `month`, `weekday`, `sex`, `ethnicity`, `boro`) remain factors.

## 6.2) Model Fit

We fit with `mgcv::gam` using `family = binomial` and REML smoothness selection.

```{r fit-gam}
mod_gam <- mgcv::gam(f_gam, family = binomial, data = train, method = "REML")
```

## 6.3) Specific Analysis and Interpretation

```{r summary-gam}
summary(mod_gam)
```

The model explains approximately **9.31% of the null deviance** ($R^{2}_{adj} = 0.0268$) in post-stop arrests in New York during the period 2023–2025. These values indicate that the **included variables explain only a limited proportion of the overall arrest outcome**. This is not unexpected for a rare binary outcome such as `arrested`, where many determinants of an arrest may not be represented by the available predictors.

### 6.3.1) Coefficients

#### 6.3.1.1) General Interpretation

The fitted GAM models the conditional probability of arrest using a binomial distribution with a logit link.

All parametric coefficients are expressed on the **log-odds scale**.

Exponentiating a coefficient gives the corresponding **odds ratio (OR)**,

$$OR=e^\beta$$

An $OR>1$ indicates higher odds of arrest relative to the reference category, while $OR<1$ indicates lower odds, holding all other variables constant.

#### 6.3.1.2) Specific Interpretation — Coefficients

The **intercept** is $-4.177$, corresponding to baseline odds of $e^{-4.177} \approx 0.0153$ or a probability of approximately **1.51%**.

- However, the intercept should not be interpreted as the unconditional arrest rate. It refers to an **observation belonging to all categorical reference groups while the centered smooth effects are equal to zero**.

For the remaining coefficients, interpretations are conditional on all other predictors in the model.

Let us first consider the significant (at $5%$) *time-related* coefficients

| Effect | Estimate | OR | Interpretation |
|-----------|----------:|----------:|---------------------------------------|
| `year2024` | 0.176 | 1.193 | Approximately **19% higher odds** of arrest than 2023 |
| `year2025` | 0.162 | 1.176 | Approximately **18% higher odds** than 2023 |
| `month2` | -0.108 | 0.898 | February has approximately **10% lower odds** than January |
| `month3` | -0.079 | 0.924 | March has approximately **8% lower odds** than January |
| `month4` | -0.131 | 0.878 | April has approximately **12% lower odds** than January |
| `month5` | -0.192 | 0.825 | May has approximately **18% lower odds** than January |
| `month6` | -0.168 | 0.845 | June has approximately **16% lower odds** than January |
| `month7` | -0.220 | 0.803 | July has approximately **20% lower odds** than January |
| `month8` | -0.212 | 0.809 | August has approximately **19% lower odds** than January |
| `month9` | -0.168 | 0.845 | September has approximately **16% lower odds** than January |
| `month10` | -0.135 | 0.874 | October has approximately **13% lower odds** than January |
| `month11` | -0.133 | 0.875 | November has approximately **13% lower odds** than January |
| `weekdayTuesday` | 0.079 | 1.082 | Tuesday has approximately **8% higher odds** than Monday |
| `weekdayWednesday` | 0.092 | 1.097 | Wednesday has approximately **10% higher odds** than Monday |
| `weekdayThursday` | 0.055 | 1.057 | Thursday has approximately **6% higher odds** than Monday |
| `weekdayFriday` | 0.054 | 1.056 | Friday has approximately **6% higher odds** than Monday |
| `weekdaySaturday` | 0.056 | 1.058 | Saturday has approximately **6% higher odds** than Monday |

Most time-related coefficients are statistically significant in this larger sample.

In particular,

- both 2024 and 2025 have significantly higher conditional odds of arrest than 2023;
- all months from February to November have significantly **lower** conditional odds than January;
- December does not differ significantly from January ($p=0.621$);
- Tuesday through Saturday have significantly higher odds than Monday;
- Sunday does not differ significantly from Monday ($p=0.368$).

The coefficients from `boro` are all statistically significant at the same level:

| Borough category  | Estimate |    OR | Relative odds |
|-------------------|---------:|------:|--------------:|
| Brooklyn (North)  |   -0.246 | 0.782 | **22% lower** |
| Brooklyn (South)  |   -0.077 | 0.926 |  **7% lower** |
| Manhattan (North) |   -0.126 | 0.881 | **12% lower** |
| Manhattan (South) |   -0.630 | 0.532 | **47% lower** |
| Queens (North)    |   -0.586 | 0.556 | **44% lower** |
| Queens (South)    |   -0.376 | 0.687 | **31% lower** |
| Staten Island     |   -0.113 | 0.893 | **11% lower** |

Each borough category therefore has significantly lower conditional odds of arrest than the reference category, **Bronx**, after fixing the remaining predictors.

The largest geographical differences are observed for `Manhattan (South)` and `Queens (North)`, whose estimated odds are approximately **47% and 44% lower**, respectively, than those in the Bronx.

For what regards `sex`:

- `sexF` has a coefficient of $-0.908$, giving $e^{-0.908} \approx 0.403$. Thus, observations labeled as `sexF` have approximately **60% lower odds of arrest** than the reference category (`sexM`), fixing the remaining predictors.
- The coefficient for `sexU` is small and non-significant ($p=0.373$), so there is no statistical evidence of a difference from `sexM`.

`ethnicity` also shows substantial conditional differences:

| Ethnicity                | Estimate |    OR |   Relative odds |
|--------------------------|---------:|------:|----------------:|
| Asian / Pacific Islander |    0.138 | 1.148 |  **15% higher** |
| Black                    |    1.209 | 3.349 | **235% higher** |
| Hispanic                 |    1.159 | 3.185 | **219% higher** |
| Other                    |   -0.453 | 0.636 |   **36% lower** |
| Unknown                  |   -0.466 | 0.628 |   **37% lower** |

These comparisons are all relative to `White`.

For example, holding all other predictors constant, the estimated odds of arrest for observations categorized as `Black` are approximately **3.35 times** those for `White`, while the odds for `Hispanic` observations are approximately **3.19 times** those for `White`.

`Asian / Pacific Islander` also shows statistically significant higher odds, at approximately **15% higher odds** than `White`.

On the other hand, observations categorized as `Other` and `Unknown` have approximately **36–37% lower conditional odds** than the reference category.

*We move the discussion of the smooth effects for `hour_num` and `age_num` to the next section.*

### 6.3.2) Smooth Term and Plots

```{r gam-smooth-plots, fig.width=10, fig.height=4}
par(mfrow = c(1, 2), mar = c(4, 4, 2.5, 1))
plot(mod_gam, shade = TRUE, shade.col = "lightblue", seWithMean = TRUE,
     pages = 0, scale = 0)
par(mfrow = c(1, 1))
```

These plots now allow a much more specific interpretation of the two smooth terms than the GAM summary alone.

#### 6.3.2.1) General Interpretation

First of all we focus on **EDF** or Estimated Degrees of Freedom. They tell us **how complex the fitted smooth actually is after penalization and gives us indication about the linearity**.

- $EDF \approx 1$ would indicate an approximately linear effect or relationship.
- $EDF >> 1$ instead indicates evidence of a **non-linear relationship**.

Then, we have a value for $\chi^{2}$. This tests **whether the smooth term** **contributes to explaining the variance in the response after fixing the other terms in the model**.

- The null hypothesis is $H_{0} : f_{j}(x) = 0$ meaning that the predictor has no meaningful contribution to the linear predictor once the other variables are accounted for.

Finally, the **p-value** indicates how strong is the evidence that the smooth terms **are associated with the conditional probability of arrest (or whether the smooth contributes to the model)**, after fixing the other predictors.

For what concerns the plot, the **vertical axis** is the estimated **partial effect on the log-odds of arrest** (after fixing the remaining predictors and smooth terms). Instead, the horizontal axis follows the numerical range of values for the term.

The smooths are centered around zero, so:

- Regions where $f(x)>0$ indicates higher conditional log-odds of arrest relative to the model's average smooth effect (baseline), while
- $f(x)<0$ indicates lower conditional log-odds.

The magnitude can also be converted into a multiplicative odds effect using $e^{f(x)}$. The **dashed curves** represent uncertainty around the estimated smooth.

#### 6.3.2.2) Focus: Smooth effect of hour

The smooth term for `hour` is highly statistically significant:

$$\text{edf}=7.925,\qquad
\chi^2=6499,\qquad
p<2\times10^{-16}$$

This provides very strong evidence that **time of day is associated with the conditional probability of arrest after controlling for the other predictors**.

Furthermore, the relationship is clearly **non-linear**, since $\text{edf}=7.925 >> 1$.

The estimated arrest propensity follows exactly the same pattern of the 2-wheels analysis. This provides further evidence for the temporal pattern already observed during the EDA.

The fitted smooth ranges approximately from $-1.3$ at its minimum to around $+0.6$ at its maximum, corresponding to a difference of about **1.9 log-odds units**. This means that the estimated odds of arrest can differ by roughly $e^{1.9} \approx 6.7 \times$ between the highest and lowest portions of the daily curve\*\*, fixing the remaining variables in the model.

#### 6.3.2.3) Focus: Smooth effect of age

Age also has a highly significant smooth effect:

$$\text{edf}=4.878,\qquad
\chi^2=4057,\qquad
p<2\times10^{-16}$$

Thus, there is very strong evidence that **the conditional probability of arrest varies with age**, after fixing the remaining predictors.

Again, $\text{edf}=4.878 >> 1$ indicates that this relationship is **non-linear**.

The shape of the smooth differs substantially from the one observed in the previous, smaller dataset. The estimated age effect:

1.  Is slightly negative for the **youngest drivers**, below age 20;
2.  Then increases rapidly, and (becoming positive) it reaches its maximum at **27–30 years**;
3.  After approximately age 30, there is a gradual decline and it crosses the zero line at **38–40 years**;
4.  From approximately **40 years**, the conditional effect becomes increasingly negative;
5.  The decline becomes particularly pronounced after **50–55 years**, with the lowest estimated values observed among the oldest drivers.

Therefore, the model suggests that the **highest conditional arrest propensity occurs around the late twenties**, after which the estimated effect decreases progressively with age.

The fitted smooth moves from approximately $+0.35$ (max) to around $-1.8$. This represents a difference of approximately **2 log-odds units**, which corresponds to an odds ratio of roughly $e^{2} \approx 7.4$. Thus, the model suggests a big change in conditional arrest odds between the age range with the highest estimated effect and the oldest observations.

## 6.4) Evaluation on Test Set

```{r gam-eval}
# Select threshold on training data
prob_gam_train <- predict(mod_gam, newdata = train, type = "response")
roc_gam_train <- pROC::roc(train$arrested, prob_gam_train, quiet = TRUE)
thr_gam <- best_thresh(roc_gam_train)

prob_gam <- predict(mod_gam, newdata = test, type = "response")
roc_gam <- pROC::roc(test$arrested, prob_gam, quiet = TRUE)
auc_gam <- as.numeric(pROC::auc(roc_gam))
class_gam <- as.integer(prob_gam >= thr_gam)
met_gam <- class_metrics(test$arrested, class_gam)
```

### 6.4.1) Confusion Matrix

```{r gam-cm-plot, fig.width=4, fig.height=3.5}
plot_cm(test$arrested, class_gam, title = "GAM — Confusion Matrix (Youden threshold, train-selected)")

cat(sprintf(
  "AUC: %.4f  |  Threshold (Youden, train-selected): %.4f\nSensitivity: %.3f  |  Specificity: %.3f  |  Balanced accuracy: %.3f\n",
  auc_gam, thr_gam, met_gam$sensitivity, met_gam$specificity, met_gam$balanced_acc
))
```

The confusion matrix shows:

- 10,755 true positives;
- 3,507 false negatives;
- 240,490 true negatives;
- And 152,017 false positives.

The model therefore identifies a relatively large proportion of the actual arrests, but this comes at the cost of a very large number of false-positive arrest predictions.

The Youden threshold is 0.0300, which is again far below the conventional 0.5 cutoff. This is consistent with the strong class imbalance in the test set, where only approximately 3.51% of observations correspond to arrests. This value is selected by maximising Youden's $J$ on the training-set ROC curve, then frozen and applied unchanged to the test set below, so no test-set information enters the cutoff choice.

Moreover, the AUC of 0.7347 indicates moderate discriminatory ability: the GAM separates arrest from non-arrest cases a bit better than chance.

Now for the derived metrics:

| Metric | Value | Interpretation |
|-----------|----------:|---------------------------------------------------|
| Accuracy | 0.618 | About 61.8% of all observations are classified correctly |
| Sensitivity / Recall | 0.754 | The model correctly identifies about 75.4% of actual arrests |
| Specificity | 0.613 | About 61.3% of actual non-arrests are correctly classified |
| Precision | 0.066 | Only about 6.6% of predicted arrests are actual arrests |
| F1 | 0.122 | The low value reflects the strong imbalance and the trade-off between relatively high recall and very low precision |
| Balanced Accuracy | 0.683 | Averaging sensitivity and specificity gives a more appropriate overall measure than ordinary accuracy in this imbalanced setting |

### 6.4.2) Performance Analysis

This was partly covered in the previous section, so here we will be focusing on the metrics that we have yet to mention.

```{r gam-metrics-table}
compute_metrics(test$arrested, prob_gam, thr_gam, "GAM") %>%
  kbl(
    caption   = "GAM — full evaluation metrics (test set; Youden threshold selected on training set)",
    col.names = c("Model","AUC","PR-AUC","Log-Loss","Brier",
                  "Threshold","Sensitivity","Specificity","Precision","F1","MCC","Bal. Acc.")
  ) %>%
  kable_styling(bootstrap_options = c("striped","condensed"), full_width = TRUE)
```

```{r gam-eval-plots, fig.width=10, fig.height=8}
eval_quad_plot(test$arrested, prob_gam, thr_gam, "GAM")
```

#### 6.4.2.1) Overall performance

| Metric | Value | Interpretation |
|-----------------|----------------:|--------------------------------------|
| **AUC** | 0.7347 | Moderate discrimination between arrests and non-arrests |
| **PR-AUC** | 0.0790 | More than twice the no-skill baseline of $\approx 0.035$, although still low in absolute terms |
| **Log-Loss** | 0.1400 | Probability predictions improve on a prevalence-only model, although substantial uncertainty remains |
| **Brier** | 0.0330 | Slightly better than the prevalence-only baseline and consistent with generally good calibration |
| Threshold | 0.0300 | Low cutoff, selected on the training set via Youden's criterion, reflecting the rare positive class |
| **MCC** | $\approx 0.138$ | Weak positive association between predicted and observed classes |
| Balanced Accuracy | 0.683 | Moderate performance when sensitivity and specificity are weighted equally |

The main conclusion is that the GAM captures a **meaningful but still limited predictive signal**. Compared with the previous analysis, discrimination is stronger, with an AUC above $0.73$, and the model detects a substantial proportion of arrests. However, the severe class imbalance continues to produce **very low precision** when probabilities are converted into binary classifications.

Note: `MCC` is reported as `NA` in the metrics table, but it can be computed directly from the confusion matrix and is approximately $0.138$.

#### 6.4.2.2) AUC

The test-set AUC is $\text{AUC} = 0.7347$

This means that if we randomly select (i) one observation where an arrest occurred and (ii) one observation where an arrest did not occur, the model assigns a higher predicted probability to the arrest observation approximately **73.5% of the time**.

The ROC curve remains clearly above the main diagonal across the range of thresholds. Thus, the model contains useful ranking information and shows **moderate discriminatory ability**.

Unfortunately, an AUC of approximately $0.735$ is still far from perfect discrimination, meaning that considerable overlap remains between the predicted probabilities of arrests and non-arrests.

#### 6.4.2.3) PR-AUC

This is particularly important given our imbalanced outcome.

Our positive class is rare, so ROC-AUC can sometimes give an overly optimistic impression of performance. **Precision-Recall AUC focuses directly on the positive class**, considering the trade-off between precision and recall.

The test-set prevalence of `arrested == 1` is approximately $3.51\%$ so the no-skill PR-AUC is approximately $\text{PR-AUC}_{\text{baseline}}\approx0.035$

Here, we obtain:

$$\text{PR-AUC}=0.079$$

*But how good is it?* It is more than **twice the no-skill baseline**, indicating that the model provides useful information for ranking observations according to arrest propensity. However, the absolute value remains low (precision of $0.066$).

Thus, the PR curve shows us that **ranking performance is meaningful, but precise identification of the rare positive class remains difficult**.

#### 6.4.2.4) Log-Loss

> Log-loss evaluates the **quality of the predicted probabilities**, rather than simply stating whether the final classification is correct.

This measure cares about **how confident the model is**. In our case:

$$\text{Log-Loss}=0.140$$

To provide a baseline comparison, a model assigning the test-set arrest prevalence of approximately $3.51\%$ to every observation would obtain a log-loss of approximately $0.152$.

The GAM therefore improves on this simple prevalence-only model. The improvement is important but not large: the predictor information improves the estimated probabilities, while a great amount of uncertainty about individual arrest outcomes remains unexplained.

#### 6.4.2.5) Brier Score

The Brier score is useful here because it evaluates whether the predicted probabilities are close to the observed outcomes, rather than merely whether the model ranks observations correctly.

> Formally, the Brier score measures the squared error of the predicted probabilities (the lower, the better).

For the GAM,

$$\text{Brier}=0.033$$

A prevalence-only model would obtain approximately

$$0.0338$$

so, the GAM performs only slightly better than the baseline.

Together with the calibration plot, this suggests that the predicted probabilities are **reasonably calibrated overall**: **groups assigned higher arrest probabilities tend to show correspondingly higher observed arrest rates**.

#### 6.4.2.6) MCC

The **Matthews Correlation Coefficient (MCC)** is particularly useful for imbalanced classification because it incorporates all four components of the confusion matrix:

$$MCC =
\frac{TP\cdot TN-FP\cdot FN}
{\sqrt{(TP+FP)(TP+FN)(TN+FP)(TN+FN)}}
\in [-1,1]$$

where:

- $1$ = perfect prediction
- $0$ = no association between predictions and outcomes
- $-1$ = completely incorrect prediction

Our result is:

$$MCC\approx0.138$$

Thus, there is a **positive but weak association between the predicted classes and the actual outcomes** (consistent with CM). Simply put:

This is consistent with the other classification metrics:

- The model is doing meaningfully better than random classification;
- But that its classification performance is very limited.

#### 6.4.2.7) Thresholds

The classification threshold selected using **Youden's index** is

$$t^*=0.0300$$

This threshold is selected once by maximising Youden's $J$ on the training-set ROC curve, then frozen and applied unchanged to the test set, so no test-set information is used in choosing the cutoff.

The threshold analysis also shows the expected trade-off: lowering the threshold increases sensitivity but reduces specificity, while increasing it improves specificity at the cost of missing more arrests.

#### 6.4.2.8) Final Remarks

Let us put the metrics together:

- **Discrimination**, $\text{AUC}\approx0.735$: the GAM has moderate ability to rank arrests above non-arrests;
- **Rare-class performance**, $\text{PR-AUC}\approx0.079$: more than twice the $\approx0.035$ no-skill baseline, although positive-class identification remains difficult;
- **Probability predictions**, $\text{Brier}\approx0.033,\, \text{LogLoss}\approx0.140$: probabilities are generally well calibrated and improve modestly over a prevalence-only model;
- **Classification**, $\text{MCC}\approx0.138$: binary predictions show only a weak positive association with the true outcomes;
- **Class balance**, $\text{ACC}_{\text{Balanced}}\approx0.683$: the model obtains a reasonably balanced sensitivity-specificity trade-off;
- **Sensitivity**, $0.754$: approximately three quarters of actual arrests are detected, although this comes at the cost of very low precision ($0.066$).

Now for the final conclusions:

- The GAM captures a **meaningful predictive signal** in the available demographic, temporal, and geographical variables.
- Its strongest result is its ability to **rank observations according to relative arrest propensity**, with an AUC of approximately $0.735$.
- The PR-AUC is substantially above the prevalence baseline, confirming that the model contains useful information about the rare positive class, even though precision remains low.
- The severe class imbalance strongly affects classification: obtaining a sensitivity of $0.754$ requires accepting a large number of false-positive arrest predictions.
- The predicted probabilities appear **more informative than the final binary classifications**, particularly given the relatively good calibration.
- Consequently, the GAM is best interpreted as a model for **estimating and ranking conditional arrest propensity**, rather than as a highly accurate classifier of individual arrests.

## 6.5) GAM — Smoothness Diagnostics (gam.check)

The `mgcv::gam.check()` produces four residual plots (QQplot, residuals vs fitted, histogram of residuals, response vs fitted values) and tests whether the basis dimension $k$ is sufficient for each smooth term.

Note that for a binomial GAM, these diagnostic plots need to be interpreted differently from the familiar residual diagnostics of gaussian linear regression. In particular, deviance residuals are not expected to be normally distributed because the response is binary.

```{r gam-check}
par(mfrow = c(2, 2))
mgcv::gam.check(mod_gam, pch = 19, cex = 0.3)
```

Overall, the four plots do not reveal a clear diagnostic failure. Much of their unusual appearance is an expected consequence of fitting a **binomial GAM to a strongly imbalanced binary response**. 

### 3.5.1) QQplot

The stepped/clumped structure is a consequence of the `0/1` response and the imbalance. The upper positive residuals correspond mainly to observations with `arrested = 1` that received relatively low predicted probabilities.

The Q-Q plot indicates a strongly non-Gaussian residual distribution, but this is **expected in this setting and is not by itself evidence of model misspecification**.

### 3.5.2) Residuals vs. linear predictor

The residuals form two clearly separated bands. This is also typical of a binary model: 

- The lower band corresponds to observations with `arrested = 0`,
- The upper band corresponds to observations with `arrested = 1`.

Positive observations tend to have relatively large positive residuals because arrests are rare and the fitted probabilities are generally low. On the other hand non-arrests mostly have small negative residuals.

There is no irregular cloud or unexpected structure beyond that induced by the binary response. In other words, the pattern reflects the **rare-event nature of the outcome rather than a heteroscedasticity problem**.

### 3.5.3) Histogram of residuals

The residual distribution is highly asymmetric:

- Most residuals are concentrated just below zero, and they correspond to the majority of correctly predicted non-arrest observations.
- A much smaller group is found at positive residual values. These correspond to actual arrests for which the model assigned relatively low probabilities.

This shape is consistent with the strong class imbalance.

### 3.5.4) Response vs. fitted values

Because the response is binary, observations necessarily appear on two horizontal lines at $Y=0$ and $Y=1$.

- Most fitted probabilities are relatively small, which is consistent with the nature of the response.
- There is considerable overlap in predicted probabilities between arrests and non-arrests. In particular, many actual arrests still receive relatively low fitted probabilities. 

This is consistent with the model's previously observed **moderate discrimination rather than strong class separation**.
