---
title: "Predicting Excessive Enforcement Patterns in Police Behavior After Vehicles Stops in New York"
subtitle: "Model Selection and Evaluation - GAM 2-Wheels"
author: "Gianmaria Pizzo"
output:
  html_document:
    fig_caption: yes
    theme: flatly
    highlight: pygments
    code_folding: show
    toc: yes
    toc_depth: 4
    number_sections: no
    toc_float:
      smooth_scroll: no
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

These `.Rmd` files represent the project submission for the course [Statistical Inference and Learning](https://www.unive.it/data/course/521944) by Professor [Cristiano Varin](https://www.unive.it/data/people/5592728) at [Ca' Foscari University of Venice](https://www.unive.it/) for the year 2025/2026.

---

# 0) Environment

## 0.1) Packages

```{r requirements}
requirements <- c(
  "remotes", "summarytools", "forcats", "scales", "purrr", "broom",
  "stringr", "dbplyr", "dplyr", "dtplyr", "tidyr", "tidyselect",
  "timeDate", "timechange", "tzdb", "hms", "readr", "ggcorrplot",
  "lubridate", "prettyunits", "RColorBrewer", "viridis", "rcartocolor",
  "sf", "ggplot2", "patchwork", "pROC", "kableExtra", "mgcv", "glmnet"
)
```

```{r install-requirements, include=FALSE}
for (req in requirements){
  if (!require(req, character.only = TRUE)){
      install.packages(req)
  }
}
```

Note that some of the packages may not be used, but to avoid any error related to dependencies, we are going to keep all of these libraries. Now, the loading of all packages should be immediate.

```{r load-requirements, include=FALSE}
for (req in requirements){
  if (!require(req, character.only = TRUE)){
      library(req, character.only = TRUE)
  }
}
```

```{r remove-vars, include=FALSE}
rm(requirements, req)
```

## 0.2) Functions

Here we define the helper functions used throughout model selection and evaluation, kept in one place for reuse across every model fitted below.

### 0.2.1) Data-Split and Cross-Validation Helpers

```{r fn-check-prop}
# ==============================================================================
# 1. Train / Test Split Diagnostics
# ==============================================================================

#' @title Check Train/Test Level Proportions
#'
#' @description Tabulates one variable's level proportions separately in
#' `train` and `test` and reports the discrepancy in percentage points. Used
#' to verify that the stratified split preserved the marginal distribution of
#' key covariates.
#'
#' @param var The column name (present in both `train` and `test`) to tabulate.
#' @param label Human-readable label for the variable, stored in the output.
#'
#' @return A tibble with columns variable, level, train, test, diff_ppt.
check_prop <- function(var, label) {
  tr <- prop.table(table(train[[var]]))
  te <- prop.table(table(test[[var]]))
  tibble(
    level = names(tr),
    train = as.numeric(tr),
    test = as.numeric(te),
    diff_ppt = round((as.numeric(te) - as.numeric(tr)) * 100, 2)
  ) %>%
    mutate(variable = label) %>%
    relocate(variable)
}
```

```{r fn-cv-auc}
# ==============================================================================
# 2. Cross-Validated AUC
# ==============================================================================

#' @title Cross-Validated AUC for a Logistic Regression Formula
#'
#' @description Refits a binomial GLM on each of the k training folds and
#' scores it on the held-out fold, returning the mean AUC across folds. Folds
#' are processed in parallel since refitting a GLM on ~1.7M rows ten times is
#' the dominant cost of the model-comparison sections below.
#'
#' @param formula Model formula passed to glm(family = binomial).
#' @param data Training data.frame; must contain the response `arrested`.
#' @param folds_vec Integer vector, same length as nrow(data), with values
#' 1..k giving each row's fold assignment.
#'
#' @return A numeric scalar, the mean out-of-fold AUC (NA folds are dropped).
cv_auc <- function(formula, data, folds_vec) {
  k <- max(folds_vec)

  auc_vals <- parallel::mclapply(seq_len(k), function(i) {
    tr <- data[folds_vec != i, ]
    te <- data[folds_vec == i, ]

    # A fold can fail to fit (e.g. rank deficiency in a rare stratum) — treat
    # as a missing fold rather than aborting the whole CV run
    fit <- tryCatch(
      glm(formula, family = binomial, data = tr),
      error = function(e) NULL
    )
    if (is.null(fit)) return(NA_real_)

    prob <- tryCatch(
      predict(fit, newdata = te, type = "response"),
      error = function(e) rep(NA_real_, nrow(te))
    )

    tryCatch(
      as.numeric(pROC::auc(te$arrested, prob, quiet = TRUE)),
      error = function(e) NA_real_
    )
  }, mc.cores = parallel::detectCores() - 1)

  mean(unlist(auc_vals), na.rm = TRUE)
}
```

```{r fn-nb-cv-auc}
#' @title Cross-Validated AUC for Naive Bayes at a Given Laplace Constant
#'
#' @description Refits e1071::naiveBayes on each training fold with smoothing
#' constant `k` and scores it on the held-out fold, returning the mean AUC.
#' Mirrors cv_auc() but calls naiveBayes() instead of glm() and reads the
#' positive-class probability from the second column of the raw posterior.
#'
#' @param k Numeric Laplace smoothing constant passed to naiveBayes(laplace = k).
#' @param data Training data.frame; must contain the response `arrested`.
#' @param folds_vec Integer vector, same length as nrow(data), with values
#' 1..k giving each row's fold assignment.
#' @param formula Model formula passed to naiveBayes().
#'
#' @return A numeric scalar, the mean out-of-fold AUC (NA folds are dropped).
nb_cv_auc <- function(k, data, folds_vec, formula) {
  n_folds  <- max(folds_vec)
  auc_vals <- numeric(n_folds)
  for (i in seq_len(n_folds)) {
    tr  <- data[folds_vec != i, ]
    te  <- data[folds_vec == i, ]
    fit <- e1071::naiveBayes(formula, data = tr, laplace = k)

    # Posterior: columns named after class levels; column 2 = P(arrested=1)
    prob <- predict(fit, newdata = te, type = "raw")[, 2]
    auc_vals[i] <- tryCatch(
      as.numeric(pROC::auc(te$arrested, prob, quiet = TRUE)),
      error = function(e) NA_real_
    )
  }
  mean(auc_vals, na.rm = TRUE)
}
```

### 0.2.2) Threshold and Confusion-Matrix Helpers

```{r fn-eval-helpers}
# ==============================================================================
# 3. Threshold Selection and Confusion-Matrix Helpers
# ==============================================================================

#' @title Find the Youden-Optimal Classification Threshold
#'
#' @description Finds the probability threshold on an ROC curve that
#' maximises Youden's J statistic (sensitivity + specificity − 1), used
#' throughout as the single decision threshold for turning predicted
#' probabilities into classes.
#'
#' @param roc_obj A pROC "roc" object, as returned by pROC::roc().
#'
#' @return A numeric scalar, the optimal threshold on the probability scale.
best_thresh <- function(roc_obj) {
  pROC::coords(roc_obj, "best", ret = "threshold", best.method = "youden",
               quiet = TRUE)$threshold
}


#' @title Tally a Confusion Matrix
#'
#' @description Cross-tabulates actual vs predicted binary class labels.
#'
#' @param actual Integer/numeric vector of 0/1 true labels.
#' @param predicted Integer/numeric vector of 0/1 predicted labels.
#'
#' @return A tibble with columns Actual, Predicted, n (one row per non-empty cell).
conf_tib <- function(actual, predicted) {
  tibble(
    Actual    = factor(actual,    levels = c(0, 1)),
    Predicted = factor(predicted, levels = c(0, 1))
  ) %>% count(Actual, Predicted)
}


#' @title Plot a Confusion-Matrix Heatmap
#'
#' @description Renders conf_tib() output as a 2x2 tile plot with cell counts.
#'
#' @param actual Integer/numeric vector of 0/1 true labels.
#' @param predicted Integer/numeric vector of 0/1 predicted labels.
#' @param title Main title of the plot.
#'
#' @return A ggplot2 object.
plot_cm <- function(actual, predicted, title = "Confusion Matrix") {
  conf_tib(actual, predicted) %>%
    ggplot(aes(Predicted, Actual, fill = n)) +
    geom_tile(colour = "white") +
    geom_text(aes(label = scales::comma(n)), size = 5.5, fontface = "bold") +
    scale_fill_gradient(low = "#e8f4f8", high = "#2c7bb6", guide = "none") +
    scale_x_discrete(labels = c("0" = "No arrest", "1" = "Arrest")) +
    scale_y_discrete(labels = c("0" = "No arrest", "1" = "Arrest")) +
    labs(title = title, x = "Predicted", y = "Actual") +
    theme_minimal(base_size = 12) +
    theme(panel.grid = element_blank())
}


#' @title Compute Threshold-Dependent Classification Metrics
#'
#' @description Computes sensitivity, specificity, balanced accuracy, PPV,
#' and NPV for a fixed set of hard class predictions against the true labels.
#'
#' @param actual Integer/numeric vector of 0/1 true labels.
#' @param predicted Integer/numeric vector of 0/1 predicted labels.
#' @param pos Value in `actual`/`predicted` denoting the positive class.
#'
#' @return A named list with sensitivity, specificity, balanced_acc, ppv, npv.
class_metrics <- function(actual, predicted, pos = 1) {
  tp <- sum(actual == pos & predicted == pos)
  tn <- sum(actual != pos & predicted != pos)
  fp <- sum(actual != pos & predicted == pos)
  fn <- sum(actual == pos & predicted != pos)
  sens <- tp / (tp + fn)
  spec <- tn / (tn + fp)
  list(
    sensitivity  = sens,
    specificity  = spec,
    balanced_acc = (sens + spec) / 2,
    ppv          = tp / (tp + fp),
    npv          = tn / (tn + fn)
  )
}


# ==============================================================================
# 4. Probability-Based Evaluation Metrics
# ==============================================================================

#' @title Compute the Full Evaluation-Metrics Tibble for One Model
#'
#' @description Given predicted probabilities and true labels, computes
#' threshold-free metrics (AUC, PR-AUC, log-loss, Brier score) plus
#' threshold-dependent metrics at the Youden-optimal cut-off (sensitivity,
#' specificity, precision, F1, MCC, balanced accuracy). Returns one row per
#' call, so results from multiple models can be row-bound into a comparison
#' table.
#'
#' @param actual Integer/numeric vector of 0/1 true labels.
#' @param prob Numeric vector of predicted P(arrested = 1), same length as `actual`.
#' @param model_name Label stored in the Model column.
#'
#' @return A one-row tibble of rounded metrics (see column names in the body).
compute_metrics <- function(actual, prob, model_name = "Model") {
  valid  <- !is.na(actual) & !is.na(prob)
  actual <- actual[valid]; prob <- prob[valid]
  roc_obj <- pROC::roc(actual, prob, quiet = TRUE)
  auc_val <- as.numeric(pROC::auc(roc_obj))
  thr     <- best_thresh(roc_obj)
  pred_cl <- as.integer(prob >= thr)

  tp <- sum(actual == 1 & pred_cl == 1); tn <- sum(actual == 0 & pred_cl == 0)
  fp <- sum(actual == 0 & pred_cl == 1); fn <- sum(actual == 1 & pred_cl == 0)
  sens <- if ((tp+fn) == 0) NA_real_ else tp / (tp+fn)
  spec <- if ((tn+fp) == 0) NA_real_ else tn / (tn+fp)
  prec <- if ((tp+fp) == 0) NA_real_ else tp / (tp+fp)
  f1   <- if (is.na(prec) || is.na(sens) || (prec+sens) == 0) NA_real_ else
            2 * prec * sens / (prec + sens)

  # Matthews correlation coefficient — guard the denominator against any
  # degenerate (all-one-class) confusion matrix, which would divide by zero
  mcc_den <- sqrt(as.numeric(tp+fp) * as.numeric(tp+fn) *
                  as.numeric(tn+fp) * as.numeric(tn+fn))
  mcc_val <- if (mcc_den == 0) NA_real_ else (tp*tn - fp*fn) / mcc_den

  # eps floors probabilities away from 0/1 so log() never evaluates to -Inf
  eps     <- 1e-15
  logloss <- -mean(actual * log(pmax(prob, eps)) + (1-actual) * log(pmax(1-prob, eps)))
  brier   <- mean((prob - actual)^2)

  # PR-AUC via trapezoid rule over pROC coords (pROC has no built-in PR-AUC)
  pr_raw <- pROC::coords(roc_obj, "all",
                          ret = c("sensitivity", "precision"),
                          transpose = FALSE, drop = FALSE)
  names(pr_raw)[names(pr_raw) == "sensitivity"] <- "recall"
  pr_raw <- pr_raw %>%
    filter(!is.nan(precision), !is.na(precision), !is.nan(recall)) %>%
    arrange(recall)
  pr_auc_val <- if (nrow(pr_raw) >= 2)
    sum(diff(pr_raw$recall) *
          (head(pr_raw$precision, -1) + tail(pr_raw$precision, -1)) / 2)
  else NA_real_

  tibble(
    Model       = model_name,
    AUC         = round(auc_val, 4),
    PR_AUC      = round(pr_auc_val, 4),
    Log_Loss    = round(logloss, 4),
    Brier       = round(brier, 4),
    Threshold   = round(thr, 4),
    Sensitivity = round(sens, 3),
    Specificity = round(spec, 3),
    Precision   = round(prec, 3),
    F1          = round(f1, 3),
    MCC         = round(mcc_val, 3),
    Bal_Acc     = round(if (!is.na(sens) && !is.na(spec)) (sens+spec)/2
                        else NA_real_, 3)
  )
}
```

### 0.2.3) Model Evaluation Plot Functions

```{r fn-plot-roc-single}
# ==============================================================================
# 5. Plots for Single-Model Evaluation
# ==============================================================================

#' @title Plot the ROC Curve for a Single Model
#'
#' @description Plots sensitivity vs specificity across all thresholds, with
#' the AUC annotated on the panel.
#'
#' @param actual Integer/numeric vector of 0/1 true labels.
#' @param prob Numeric vector of predicted P(arrested = 1), same length as `actual`.
#' @param title Main title of the plot.
#'
#' @return A ggplot2 object.
plot_roc_single <- function(actual, prob, title = "ROC Curve") {
  roc_obj <- pROC::roc(actual, prob, quiet = TRUE)
  auc_lbl <- paste0("AUC = ", round(as.numeric(pROC::auc(roc_obj)), 4))
  pROC::ggroc(roc_obj, colour = "#2c7bb6", linewidth = 0.9) +
    geom_abline(slope = 1, intercept = 1, linetype = "dashed", colour = "grey60") +
    annotate("text", x = 0.25, y = 0.06, label = auc_lbl,
             colour = "#2c7bb6", size = 4) +
    labs(x = "Specificity", y = "Sensitivity", title = title) +
    theme_minimal(base_size = 11) + theme(panel.grid.minor = element_blank())
}
```

```{r fn-plot-pr-single}
#' @title Plot the Precision-Recall Curve for a Single Model
#'
#' @description Plots precision vs recall across all thresholds against the
#' no-skill baseline (the class prevalence), with PR-AUC annotated. More
#' informative than the ROC curve under the severe class imbalance here.
#'
#' @param actual Integer/numeric vector of 0/1 true labels.
#' @param prob Numeric vector of predicted P(arrested = 1), same length as `actual`.
#' @param title Main title of the plot.
#'
#' @return A ggplot2 object.
plot_pr_single <- function(actual, prob, title = "Precision-Recall Curve") {
  roc_obj <- pROC::roc(actual, prob, quiet = TRUE)
  pr_raw  <- pROC::coords(roc_obj, "all",
                           ret = c("sensitivity", "precision"),
                           transpose = FALSE, drop = FALSE)
  names(pr_raw)[names(pr_raw) == "sensitivity"] <- "recall"
  pr_df <- pr_raw %>%
    filter(!is.nan(precision), !is.na(precision)) %>% arrange(recall)

  # PR-AUC via trapezoid rule, same approach as in compute_metrics()
  pr_auc_lbl <- if (nrow(pr_df) >= 2) {
    v <- sum(diff(pr_df$recall) *
               (head(pr_df$precision,-1) + tail(pr_df$precision,-1)) / 2)
    paste0("PR-AUC = ", round(v, 4))
  } else "PR-AUC = NA"
  baseline <- mean(actual)
  ggplot(pr_df, aes(x = recall, y = precision)) +
    geom_line(colour = "#d73027", linewidth = 0.9) +
    geom_hline(yintercept = baseline, linetype = "dashed", colour = "grey55") +
    annotate("text", x = 0.02, y = baseline + 0.015,
             label = paste0("No-skill = ", round(baseline, 3)),
             colour = "grey40", size = 3.2, hjust = 0) +
    annotate("text", x = 0.55, y = 0.90, label = pr_auc_lbl,
             colour = "#d73027", size = 4) +
    scale_x_continuous(limits = c(0, 1)) + scale_y_continuous(limits = c(0, 1)) +
    labs(x = "Recall", y = "Precision", title = title) +
    theme_minimal(base_size = 11) + theme(panel.grid.minor = element_blank())
}
```

```{r fn-plot-calibration-single}
#' @title Plot the Calibration (Reliability) Curve for a Single Model
#'
#' @description Bins predictions into `n_bins` equal-frequency groups and
#' plots mean predicted probability against observed positive rate per bin,
#' against the perfect-calibration diagonal.
#'
#' @param actual Integer/numeric vector of 0/1 true labels.
#' @param prob Numeric vector of predicted P(arrested = 1), same length as `actual`.
#' @param title Main title of the plot.
#' @param n_bins Number of quantile bins of `prob`.
#'
#' @return A ggplot2 object.
plot_calibration_single <- function(actual, prob, title = "Calibration Plot",
                                    n_bins = 10) {
  # Quantile breaks (not equal-width) so each bin has a comparable sample size
  # despite the predicted probabilities being heavily right-skewed
  brks  <- unique(quantile(prob, seq(0, 1, length.out = n_bins + 1), na.rm = TRUE))
  cal_df <- tibble(actual = actual, prob = prob) %>%
    mutate(bin = cut(prob, breaks = brks, include.lowest = TRUE)) %>%
    filter(!is.na(bin)) %>%
    group_by(bin) %>%
    summarise(mean_pred = mean(prob), mean_obs = mean(actual),
              n = n(), .groups = "drop")
  ggplot(cal_df, aes(x = mean_pred, y = mean_obs)) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey55") +
    geom_line(colour = "#2c7bb6", linewidth = 0.8) +
    geom_point(aes(size = n), colour = "#2c7bb6", alpha = 0.85) +
    scale_size_continuous(name = "Count", range = c(2, 9)) +
    scale_x_continuous(labels = scales::percent_format(accuracy = 0.1)) +
    scale_y_continuous(labels = scales::percent_format(accuracy = 0.1)) +
    labs(x = "Mean predicted probability", y = "Observed positive rate",
         title = title,
         caption = "Size ∝ bin count.  Dashed = perfect calibration.") +
    theme_minimal(base_size = 11) +
    theme(legend.position = "bottom", panel.grid.minor = element_blank())
}
```

```{r fn-plot-threshold-single}
#' @title Plot the Threshold-Sweep Metrics for a Single Model
#'
#' @description Plots sensitivity, specificity, precision, and F1 as
#' functions of the classification threshold, with the Youden-optimal
#' threshold marked.
#'
#' @param actual Integer/numeric vector of 0/1 true labels.
#' @param prob Numeric vector of predicted P(arrested = 1), same length as `actual`.
#' @param title Main title of the plot.
#'
#' @return A ggplot2 object.
plot_threshold_single <- function(actual, prob, title = "Threshold Analysis") {
  roc_obj  <- pROC::roc(actual, prob, quiet = TRUE)
  thr_data <- pROC::coords(roc_obj, "all",
                            ret = c("threshold","sensitivity","specificity","precision"),
                            transpose = FALSE, drop = FALSE) %>%
    filter(is.finite(threshold), threshold >= 0, threshold <= 1) %>%
    mutate(f1 = ifelse(!is.nan(precision) & (precision + sensitivity) > 0,
                       2 * precision * sensitivity / (precision + sensitivity), 0)) %>%
    pivot_longer(c(sensitivity, specificity, precision, f1),
                 names_to = "Metric", values_to = "Value") %>%
    mutate(Metric = recode(Metric,
      sensitivity = "Sensitivity", specificity = "Specificity",
      precision   = "Precision",   f1          = "F1 Score"))
  thr_opt <- best_thresh(roc_obj)
  ggplot(thr_data, aes(x = threshold, y = Value, colour = Metric)) +
    geom_line(linewidth = 0.75, na.rm = TRUE) +
    geom_vline(xintercept = thr_opt, linetype = "dashed",
               colour = "grey35", linewidth = 0.55) +
    annotate("text", x = thr_opt, y = 0.04,
             label = paste0("Youden\n", round(thr_opt, 3)),
             hjust = -0.08, size = 3, colour = "grey30") +
    scale_colour_viridis_d(option = "turbo", begin = 0.05, end = 0.92) +
    scale_y_continuous(limits = c(0, 1)) +
    labs(x = "Threshold", y = "Metric value", colour = NULL, title = title,
         caption = "Dashed: Youden's J optimal threshold") +
    theme_minimal(base_size = 11) +
    theme(legend.position = "bottom", panel.grid.minor = element_blank())
}
```

```{r fn-plot-eval-quad}
#' @title Plot the 2x2 Evaluation Panel for a Single Model
#'
#' @description Composes plot_roc_single(), plot_pr_single(),
#' plot_calibration_single(), and plot_threshold_single() into one
#' (ROC | PR) / (Calibration | Threshold) patchwork layout.
#'
#' @param actual Integer/numeric vector of 0/1 true labels.
#' @param prob Numeric vector of predicted P(arrested = 1), same length as `actual`.
#' @param model_name Prefixed to each sub-plot's title.
#'
#' @return A patchwork object (composed ggplot2 object).
eval_quad_plot <- function(actual, prob, model_name) {
  p1 <- plot_roc_single(actual, prob,   paste(model_name, "— ROC"))
  p2 <- plot_pr_single(actual, prob,    paste(model_name, "— Precision-Recall"))
  p3 <- plot_calibration_single(actual, prob, paste(model_name, "— Calibration"))
  p4 <- plot_threshold_single(actual, prob, paste(model_name, "— Threshold Analysis"))
  (p1 | p2) / (p3 | p4)
}
```

```{r fn-plot-nb-feature}
#' @title Plot a Naive Bayes Conditional-Probability Bar Chart
#'
#' @description Plots P(feature level | class) from one entry of a fitted
#' naiveBayes model's frequency tables, grouped by class, so the strongest
#' discriminators under the conditional-independence assumption are visible.
#'
#' @param feature_name Name of a predictor in `mod$tables`.
#' @param mod A fitted e1071::naiveBayes model object.
#'
#' @return A ggplot2 object.
plot_nb_feature <- function(feature_name, mod) {
  tbl <- mod$tables[[feature_name]]
  as.data.frame(tbl) %>%
    rename(level = 2, class = 1, prob = Freq) %>%
    mutate(class = factor(class, labels = c("Not arrested", "Arrested"))) %>%
    ggplot(aes(x = level, y = prob, fill = class)) +
    geom_col(position = position_dodge(width = 0.75), width = 0.65) +
    scale_fill_viridis_d(option = "mako", begin = 0.3, end = 0.8) +
    scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
    labs(x = NULL, y = "P(level | class)", title = feature_name, fill = NULL) +
    theme_minimal(base_size = 9) +
    theme(
      axis.text.x    = element_text(angle = 30, hjust = 1),
      legend.position = "bottom",
      panel.grid.minor = element_blank()
    )
}
```

### 0.2.4) Model-Summary Helpers

```{r fn-build-modal-row}
# ==============================================================================
# 6. Marginal-Effect and Coefficient Helpers
# ==============================================================================

#' @title Build a One-Row Data Frame of Modal/Median Covariate Values
#'
#' @description For each variable in `vars`, takes the most frequent level
#' (factor/character) or the median (numeric) from `data`. Used to hold
#' non-focal predictors fixed at a representative value when computing
#' marginal-effect predictions for an interaction term.
#'
#' @param data Data.frame to summarise (typically `train`).
#' @param vars Character vector of column names in `data`.
#'
#' @return A one-row data.frame with one column per entry in `vars`.
build_modal_row <- function(data, vars) {
  row_list <- lapply(vars, function(v) {
    x <- data[[v]]
    if (is.factor(x)) {
      factor(names(sort(table(x), decreasing = TRUE))[1], levels = levels(x))
    } else if (is.numeric(x)) {
      median(x, na.rm = TRUE)
    } else {
      names(sort(table(x), decreasing = TRUE))[1]
    }
  })
  as.data.frame(setNames(row_list, vars))
}
```

```{r fn-extract-coefs}
#' @title Extract Non-Zero LASSO Coefficients at a Given Lambda
#'
#' @description Extracts the coefficients selected by a cv.glmnet fit at a
#' named lambda (e.g. "lambda.min" or "lambda.1se"), drops the zeroed-out
#' coefficients, and adds the exponentiated odds ratio.
#'
#' @param cv_fit A fitted glmnet::cv.glmnet object.
#' @param s_name Lambda selector passed to coef(cv_fit, s = s_name), typically
#' "lambda.min" or "lambda.1se".
#'
#' @return A tibble with columns feature, estimate (log-odds), OR (odds ratio),
#' sorted by descending |estimate|.
extract_coefs <- function(cv_fit, s_name) {
  cf <- coef(cv_fit, s = s_name)
  tibble(
    feature  = rownames(cf)[-1],
    estimate = as.numeric(cf[-1])
  ) %>%
    filter(estimate != 0) %>%
    arrange(desc(abs(estimate))) %>%
    mutate(OR = round(exp(estimate), 3), estimate = round(estimate, 4))
}
```

---

# 1) Data Loading and Preparation

```{r load-data}
vehstop <- read.csv("data/vehiclestops_2_wheels_final.csv")
```

## 1.1) Factor conversion and Reference Levels

We convert all categorical predictors to factors and set interpretable reference levels before any modelling step, ensuring consistency between all downstream model fits and cross-validation folds.

```{r factor-prep}
# Ordered weekday levels (ISO order)
wday_levels <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday")

# Ordered time bracket levels
tbkt_levels <- c(
  "Late night (0–3)",
  "Early morning (4–7)",
  "Morning commute (8–11)",
  "Midday (12–15)",
  "Afternoon/evening (16–19)",
  "Night (20–23)"
)

# Ordered age-bracket levels
abkt_levels <- c("Under 18", "18–24", "25–34", "35–44", "45–54", "55–64", "65+", "Unknown")

vehstop <- vehstop %>%
  mutate(
    year = factor(year),
    month = factor(month, levels = 1:12),
    weekday = factor(weekday, levels = wday_levels),
    hour = factor(hour, levels = 0:23),
    time_bracket = factor(time_bracket, levels = tbkt_levels),
    area = factor(area),
    boro = factor(boro),
    sex = factor(sex),
    ethnicity = factor(ethnicity),
    age_bracket = factor(age_bracket, levels = abkt_levels)
  )

# Reference levels chosen for interpretable contrasts
vehstop$sex <- relevel(vehstop$sex, ref = "M")
vehstop$ethnicity <- relevel(vehstop$ethnicity, ref = "White")
vehstop$weekday <- relevel(vehstop$weekday, ref = "Monday")
vehstop$time_bracket <- relevel(vehstop$time_bracket, ref = "Morning commute (8–11)")
vehstop$age_bracket <- relevel(vehstop$age_bracket, ref = "25–34")

rm(wday_levels, tbkt_levels, abkt_levels)
```

## 1.2) Removing NAs from `age`

We had to exclude `age` as a predictor from the other models, since we had missing values that we labeled as `Unknown` in `age_bracket`. Here instead we want to include it so we will drop the NAs and use a smoothing curve.

```{r drop-age-na}
# TODO: Drop `age` NAs
vehstop <- vehstop %>% 
  dplyr::filter(!is.na(age))
```

## 1.3) Stratified Train / Test Split

For the train/test split, a **stratified sampling approach** is used to ensure that the training and test sets have comparable distributions of the outcome variable (`arrested`) and relevant contextual variables. 

This is particularly important because the target variable is imbalanced, with arrested cases representing only a small proportion of all observations. Rather than stratifying only on `arrested`, the code constructs a **joint stratification variable** from `arrested`, `area`, `sex`, `ethnicity`, and `hour` using `interaction()`. Thus, each observation is assigned to a stratum corresponding to a particular combination of these variables. 

Formally, the strata can be represented as

$$
S_i = (\text{arrested}_i,\text{area}_i,\text{sex}_i,\text{ethnicity}_i,\text{hour}_i),
$$

where $S_i$ denotes the stratum to which observation $i$ belongs. The objective is to sample approximately 80% of observations within each stratum for the training set and retain the remaining 20% for testing:

$$
P(i \in \mathcal{D}_{train} \mid S_i=s) \approx 0.8,
\qquad
P(i \in \mathcal{D}_{test} \mid S_i=s) \approx 0.2.
$$

```{r stratified-split}
set.seed(2026)

# Joint stratum key; drop = TRUE removes empty combinations
vehstop$.strat <- interaction(
  vehstop$arrested, vehstop$area, vehstop$sex, vehstop$ethnicity, vehstop$hour,
  drop = TRUE
)

train_rows <- vehstop %>%
  dplyr::mutate(.row = row_number()) %>%
  dplyr::group_by(.strat) %>%
  dplyr::slice_sample(prop = 0.8) %>%
  dplyr::pull(.row)

train <- vehstop[ train_rows, ] %>% dplyr::select(-.strat)
test <- vehstop[-train_rows, ] %>% dplyr::select(-.strat)
vehstop <- vehstop %>% dplyr::select(-.strat)
```

In the code, `group_by(.strat) %>% slice_sample(prop = 0.8)` performs this within-stratum sampling, while the observations not selected for training form the test set. This approach aims to preserve not only the overall prevalence of the outcome,

$$
P_{\text{train}}(\text{arrested}=1) \approx P_{\text{test}}(\text{arrested}=1),
$$

but also the composition of the data across geographic, demographic, and temporal characteristics. This is desirable because differences in these characteristics could otherwise introduce an unintended distributional difference between the training and test sets. The temporary `.strat` variable is removed after the split because it is used solely to define the sampling groups and is not intended to be used as a predictor in the subsequent modelling. 

Finally, the verification code compares the empirical marginal distributions of `arrested`, `area`, `sex`, and `ethnicity` in the two datasets. For a variable $X$ and level $k$, the proportion in each dataset is calculated as

$$
\hat{P}_{train}(X=k)
=
\frac{\sum_{i \in \mathcal{D}_{train}} \mathbb{I}(X_i=k)}
{|\mathcal{D}_{train}|},
$$

and analogously for the test set. The difference is then reported in **percentage points**:

$$
\Delta_k =
100\left[
\hat{P}_{test}(X=k)
-
\hat{P}_{train}(X=k)
\right].
$$

Small values of $\Delta_k$ indicate that the corresponding characteristic is similarly represented in the two datasets. The final summary additionally compares the number of observations and the proportion of arrested cases in the training and test sets, providing a direct check that the intended approximately 80/20 split and the class distribution have been maintained.

```{r split-verification}
bind_rows(
  check_prop("arrested", "arrested"),
  check_prop("area", "area"),
  check_prop("sex", "sex"),
  check_prop("ethnicity", "ethnicity")
) %>%
  kbl(
    digits = 4,
    caption = "Proportion of each level in train vs test (diff in percentage points)",
    col.names = c("Variable", "Level", "Train", "Test", "Δ (ppt)")
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE) %>%
  collapse_rows(columns = 1, valign = "top")
```

```{r split-summary}
cat(sprintf(
  "Training set : %d rows  |  Arrested: %.2f%%\nTest set     : %d rows  |  Arrested: %.2f%%\n",
  nrow(train), 100 * mean(train$arrested),
  nrow(test),  100 * mean(test$arrested)
))
```

The maximum absolute discrepancy between the two sets is on the order of fractions of a percentage point, confirming that the stratification preserved the joint distribution of the target and the key covariates.

## 1.4) Numerical Conversion

```{r num-conv-train-test}
train <- train %>%
  mutate(
    hour_num = as.numeric(as.character(hour)),
    age_num = as.numeric(as.character(age))
  )

test <- test %>%
  mutate(
    hour_num = as.numeric(as.character(hour)),
    age_num = as.numeric(as.character(age))
  )
```

```{r remove-vehstop-train-test}
# Remove vehstop, train, test for memory purposes
rm(vehstop, train_rows)
```

---

# 6) GAM

**Why did we choose this model?** TODO

**Modus Operandi:** TODO

## 6.1) Assumptions and Setup

A Generalized Additive Model (GAM) extends logistic regression by allowing selected continuous predictors to enter through smooth functions rather than linear terms.

### 6.1.2) New Formulation — Adding Splines for `hour` and `age`

TODO: introduce why we add splines for these metrics, just discussive

Two continuous predictors receive data-driven smooth terms:
-   `s(hour_num, bs = "cc", k = 10)`: a **cyclic cubic spline** for hour-of-day (hours 23 and 0 are adjacent on the clock). The EDA showed a non-linear enforcement cycle across the day that a linear slope or fixed brackets cannot fully capture.
-   `s(age_num, bs = "cr", k = 6)`: a **cubic regression spline** for driver age. The EDA found a non-monotone age–arrest relationship (elevated under-18 and over-65 rates) that neither a linear slope nor an arbitrary bracket optimally represents.

All other predictors (`year`, `month`, `weekday`, `sex`, `ethnicity`, `boro`) remain parametric factors, matching the selected predictor forms from section ii.

```{r f-gam-model}
# TODO: Move fit_gam here
f_gam <- arrested ~ year + month + weekday + s(hour_num, bs = 'cc', k = 10) + boro + sex + ethnicity + s(age_num, bs = 'cr', k = 6)
```

## 6.2) Model Fit

We fit with `mgcv::gam` using `family = binomial` and REML smoothness selection.

```{r fit-gam}
mod_gam <- mgcv::gam(f_gam, family = binomial, data = train, method = "REML")
```

TODO: Summary and briefly comment

**What does this say about the training?** TODO

## 6.3) Specific Analysis and Interpretation

### 6.3.1) Coefficients

```{r summary-gam}
summary(mod_gam)
```

### 6.3.2) Smooth Term and Plots

Estimated smooth effects on the log-odds scale, centred at zero. Shaded bands are 2 SE pointwise confidence intervals. Effective degrees of freedom (edf) reported in the summary — values above 1 indicate non-linearity.

```{r gam-smooth-plots, fig.width=10, fig.height=4}
par(mfrow = c(1, 2), mar = c(4, 4, 2.5, 1))
plot(mod_gam, shade = TRUE, shade.col = "lightblue", seWithMean = TRUE,
     pages = 0, scale = 0)
par(mfrow = c(1, 1))
```

TODO: Comment

## 6.4) Evaluation on Test Set

```{r gam-eval}
prob_gam  <- predict(mod_gam, newdata = test, type = "response")
roc_gam   <- pROC::roc(test$arrested, prob_gam, quiet = TRUE)
auc_gam   <- as.numeric(pROC::auc(roc_gam))
thr_gam   <- best_thresh(roc_gam)
class_gam <- as.integer(prob_gam >= thr_gam)
met_gam   <- class_metrics(test$arrested, class_gam)

cat(sprintf(
  "AUC: %.4f  |  Threshold (Youden): %.4f\nSensitivity: %.3f  |  Specificity: %.3f  |  Balanced accuracy: %.3f\n",
  auc_gam, thr_gam, met_gam$sensitivity, met_gam$specificity, met_gam$balanced_acc
))
```

### 6.4.1) Confusion Matrix

```{r gam-cm-plot, fig.width=4, fig.height=3.5}
plot_cm(test$arrested, class_gam, title = "GAM — Confusion Matrix (Youden threshold)")
```

Let us start from the derived metrics:

| Metric               	| Value 	| Interpretation 	|
|----------------------	|-------	|----------------	|
| Accuracy             	|       	|                	|
| Sensitivity / Recall 	|       	|                	|
| Specificity          	|       	|                	|
| Precision            	|       	|                	|
| F1                   	|       	|                	|
| Balanced Accuracy    	|       	|                	|

TODO: Create as many "####" subsections for the interpretation as we need, for the derived metrics like LASSO

Conclusion: TODO

Questions we can answer: TODO

### 6.4.2) Performance Analysis

This was partly covered in the previous section, so here we will be focusing on the metrics that we have yet to mention.

```{r gam-metrics-table}
compute_metrics(test$arrested, prob_gam, "GAM") %>%
  kbl(
    caption   = "GAM — full evaluation metrics (test set, Youden threshold)",
    col.names = c("Model","AUC","PR-AUC","Log-Loss","Brier",
                  "Threshold","Sensitivity","Specificity","Precision","F1","MCC","Bal. Acc.")
  ) %>%
  kable_styling(bootstrap_options = c("striped","condensed"), full_width = TRUE)
```

```{r gam-eval-plots, fig.width=10, fig.height=8}
eval_quad_plot(test$arrested, prob_gam, "GAM")
```

#### 6.4.2.1) Overall performance

| Metric            	| Value 	| Interpretation 	|
|-------------------	|-------	|----------------	|
| **AUC**           	|       	|                	|
| **PR-AUC**        	|       	|                	|
| **Log-Loss**      	|       	|                	|
| **Brier**         	|       	|                	|
| Threshold         	|       	|                	|
| **MCC**           	|       	|                	|
| Balanced Accuracy 	|       	|                	|

The main conclusion is TODO

#### 6.4.2.2) AUC

The test-set AUC is $\approx x$. 

TODO

#### 6.4.2.3) PR-AUC

This is particularly important given our imbalanced outcome.

Our positive class is rare, so ROC-AUC can sometimes give an overly optimistic impression of performance. **Precision-Recall AUC focuses directly on the positive class**, considering the trade-off between precision and recall.

Here, we obtain:

TODO

*But how good is x?* TODO

#### 6.4.2.4) Log-Loss

> Log-loss evaluates the **quality of the predicted probabilities**, rather than simply stating whether the final classification is correct.

This measure cares about **how confident the model is**. For example, predicting $p=0.51$ for an actual arrest is not treated the same as predicting $p=0.99$. In fact, a highly confident incorrect prediction receives a strong penalty.

In our case:

TODO

#### 6.4.2.5) Brier Score

The Brier score is useful here because it evaluates whether the predicted probabilities are close to the observed outcomes, rather than merely whether the model ranks observations correctly.

> Formally, the Brier score measures the squared error of the predicted probabilities (the lower, the better).

TODO

#### 6.4.2.6) MCC

The **Matthews Correlation Coefficient (MCC)** is particularly useful for imbalanced classification because it incorporates all four components of the confusion matrix:

$$MCC = \frac{TP\cdot TN-FP\cdot FN}{\sqrt{(TP+FP)(TP+FN)(TN+FP)(TN+FN)}} \in [-1, 1]$$
Where, 

- $1$ = perfect prediction
- $0$ = no better than random association
- $-1$ = completely incorrect prediction

Our results are:

TODO

#### 6.4.2.7) Thresholds

TODO

#### 6.4.2.8) Final Remarks

Let us up the metrics together:

- **Discrimination**, $\text{AUC} \approx x$: COMMENT.
- **Rare-class performance**, $\text{PR} \text{-} \text{AUC} \approx x$: COMMENT.
- **Probability predictions**, $\text{Brier} \approx x, \,\, \text{LogLoss} \approx x$: COMMENT.
- **Classification**, $\text{MCC} \approx x$: COMMENT.
- **Class balance** $\text{ACC}_{\text{Balanced}} \approx x$: COMMENT.


Now for the final conclusions:

- TODO

## 6.5) Diagnostics — Smoothness Diagnostics (`gam.check`)

`mgcv::gam.check()` produces four residual plots (Q-Q, residuals vs fitted, histogram of residuals, response vs fitted values) and tests whether the basis dimension $k$ is sufficient for each smooth term. A significant $p$-value in the $k$-index test indicates the smooth may be under-smoothed — the effective degrees of freedom are hitting the basis limit and a higher $k$ should be tried.

```{r gam-check, fig.width=10, fig.height=8, message=FALSE, warning=FALSE}
par(mfrow = c(2, 2))
mgcv::gam.check(mod_gam, pch = 19, cex = 0.3)
par(mfrow = c(1, 1))
```

```{r gam-smooth-plots2, fig.width=10, fig.height=4, message=FALSE, warning=FALSE}
par(mfrow = c(1, 2))
plot(
  mod_gam,
  shade     = TRUE,
  shade.col = "lightblue",
  seWithMean = TRUE,
  scale     = 0,
  pages     = 1,
  main      = "GAM — Estimated smooth terms with 95% CI"
)
par(mfrow = c(1, 1))
```

The Q-Q plot assesses normality of deviance residuals — for logistic GAMs, heavy tails or systematic curvature indicate model misspecification. The residuals vs fitted plot should show no trend; a funnel or arch suggests an omitted non-linear term or interaction. The smooth-term plots (second figure) show the estimated partial effect of `hour_num` and `age` on the log-odds of arrest: a flat line within the confidence band would indicate the smooth contributes no signal, while a pronounced curve validates the non-linear specification over a simpler linear term.
