---
title: "Predicting Excessive Enforcement Patterns in Police Behavior After Vehicles Stops in New York"
subtitle: "Model Selection and Evaluation - LASSO 4-Wheels"
author: "Gianmaria Pizzo"
output:
  html_document:
    fig_caption: yes
    theme: flatly
    highlight: pygments
    code_folding: show
    toc: yes
    toc_depth: 4
    number_sections: no
    toc_float:
      smooth_scroll: no
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

These `.Rmd` files represent the project submission for the course [Statistical Inference and Learning](https://www.unive.it/data/course/521944) by Professor [Cristiano Varin](https://www.unive.it/data/people/5592728) at [Ca' Foscari University of Venice](https://www.unive.it/) for the year 2025/2026.

---

TODO:

- [ ] Remove extra functions
- [ ] Remove extra packages
- [ ] Modify for 4 wheels
- [ ] Comment Code
- [ ] Format Code

---

# 0) Environment

## 0.1) Packages

```{r requirements}
requirements <- c(
  "remotes", "summarytools", "forcats", "scales", "purrr", "broom",
  "stringr", "dbplyr", "dplyr", "dtplyr", "tidyr", "tidyselect",
  "timeDate", "timechange", "tzdb", "hms", "readr",
  "lubridate", "prettyunits", "RColorBrewer", "viridis", "rcartocolor",
  "sf", "ggplot2", "patchwork", "pROC", "kableExtra", "mgcv", "glmnet"
)
```

```{r install-requirements, include=FALSE}
for (req in requirements){
  if (!require(req, character.only = TRUE)){
      install.packages(req)
  }
}
```

Note that some of the packages may not be used, but to avoid any error related to dependencies, we are going to keep all of these libraries. Now, the loading of all packages should be immediate.

```{r load-requirements, include=FALSE}
for (req in requirements){
  if (!require(req, character.only = TRUE)){
      library(req, character.only = TRUE)
  }
}
```

```{r remove-vars, include=FALSE}
rm(requirements, req)
```

## 0.2) Functions

Here we define the helper functions used throughout model selection and evaluation, kept in one place for reuse across every model fitted below.

### 0.2.1) Data-Split and Cross-Validation Helpers

```{r fn-check-prop}
# ==============================================================================
# 1. Train / Test Split Diagnostics
# ==============================================================================

#' @title Check Train/Test Level Proportions
#'
#' @description Tabulates one variable's level proportions separately in
#' `train` and `test` and reports the discrepancy in percentage points. Used
#' to verify that the stratified split preserved the marginal distribution of
#' key covariates.
#'
#' @param var The column name (present in both `train` and `test`) to tabulate.
#' @param label Human-readable label for the variable, stored in the output.
#'
#' @return A tibble with columns variable, level, train, test, diff_ppt.
check_prop <- function(var, label) {
  tr <- prop.table(table(train[[var]]))
  te <- prop.table(table(test[[var]]))
  tibble(
    level = names(tr),
    train = as.numeric(tr),
    test = as.numeric(te),
    diff_ppt = round((as.numeric(te) - as.numeric(tr)) * 100, 2)
  ) %>%
    mutate(variable = label) %>%
    relocate(variable)
}
```

```{r fn-cv-auc}
# ==============================================================================
# 2. Cross-Validated AUC
# ==============================================================================

#' @title Cross-Validated AUC for a Logistic Regression Formula
#'
#' @description Refits a binomial GLM on each of the k training folds and
#' scores it on the held-out fold, returning the mean AUC across folds. Folds
#' are processed in parallel since refitting a GLM on ~1.7M rows ten times is
#' the dominant cost of the model-comparison sections below.
#'
#' @param formula Model formula passed to glm(family = binomial).
#' @param data Training data.frame; must contain the response `arrested`.
#' @param folds_vec Integer vector, same length as nrow(data), with values
#' 1..k giving each row's fold assignment.
#'
#' @return A numeric scalar, the mean out-of-fold AUC (NA folds are dropped).
cv_auc <- function(formula, data, folds_vec) {
  k <- max(folds_vec)

  auc_vals <- parallel::mclapply(seq_len(k), function(i) {
    tr <- data[folds_vec != i, ]
    te <- data[folds_vec == i, ]

    # A fold can fail to fit (e.g. rank deficiency in a rare stratum) — treat
    # as a missing fold rather than aborting the whole CV run
    fit <- tryCatch(
      glm(formula, family = binomial, data = tr),
      error = function(e) NULL
    )
    if (is.null(fit)) return(NA_real_)

    prob <- tryCatch(
      predict(fit, newdata = te, type = "response"),
      error = function(e) rep(NA_real_, nrow(te))
    )

    tryCatch(
      as.numeric(pROC::auc(te$arrested, prob, quiet = TRUE)),
      error = function(e) NA_real_
    )
  }, mc.cores = parallel::detectCores() - 1)

  mean(unlist(auc_vals), na.rm = TRUE)
}
```

```{r fn-nb-cv-auc}
#' @title Cross-Validated AUC for Naive Bayes at a Given Laplace Constant
#'
#' @description Refits e1071::naiveBayes on each training fold with smoothing
#' constant `k` and scores it on the held-out fold, returning the mean AUC.
#' Mirrors cv_auc() but calls naiveBayes() instead of glm() and reads the
#' positive-class probability from the second column of the raw posterior.
#'
#' @param k Numeric Laplace smoothing constant passed to naiveBayes(laplace = k).
#' @param data Training data.frame; must contain the response `arrested`.
#' @param folds_vec Integer vector, same length as nrow(data), with values
#' 1..k giving each row's fold assignment.
#' @param formula Model formula passed to naiveBayes().
#'
#' @return A numeric scalar, the mean out-of-fold AUC (NA folds are dropped).
nb_cv_auc <- function(k, data, folds_vec, formula) {
  n_folds  <- max(folds_vec)
  auc_vals <- numeric(n_folds)
  for (i in seq_len(n_folds)) {
    tr  <- data[folds_vec != i, ]
    te  <- data[folds_vec == i, ]
    fit <- e1071::naiveBayes(formula, data = tr, laplace = k)

    # Posterior: columns named after class levels; column 2 = P(arrested=1)
    prob <- predict(fit, newdata = te, type = "raw")[, 2]
    auc_vals[i] <- tryCatch(
      as.numeric(pROC::auc(te$arrested, prob, quiet = TRUE)),
      error = function(e) NA_real_
    )
  }
  mean(auc_vals, na.rm = TRUE)
}
```

### 0.2.2) Threshold and Confusion-Matrix Helpers

```{r fn-eval-helpers}
# ==============================================================================
# 3. Threshold Selection and Confusion-Matrix Helpers
# ==============================================================================

#' @title Find the Youden-Optimal Classification Threshold
#'
#' @description Finds the probability threshold on an ROC curve that
#' maximises Youden's J statistic (sensitivity + specificity − 1), used
#' throughout as the single decision threshold for turning predicted
#' probabilities into classes.
#'
#' @param roc_obj A pROC "roc" object, as returned by pROC::roc().
#'
#' @return A numeric scalar, the optimal threshold on the probability scale.
best_thresh <- function(roc_obj) {
  pROC::coords(roc_obj, "best", ret = "threshold", best.method = "youden",
               quiet = TRUE)$threshold
}


#' @title Tally a Confusion Matrix
#'
#' @description Cross-tabulates actual vs predicted binary class labels.
#'
#' @param actual Integer/numeric vector of 0/1 true labels.
#' @param predicted Integer/numeric vector of 0/1 predicted labels.
#'
#' @return A tibble with columns Actual, Predicted, n (one row per non-empty cell).
conf_tib <- function(actual, predicted) {
  tibble(
    Actual    = factor(actual,    levels = c(0, 1)),
    Predicted = factor(predicted, levels = c(0, 1))
  ) %>% count(Actual, Predicted)
}


#' @title Plot a Confusion-Matrix Heatmap
#'
#' @description Renders conf_tib() output as a 2x2 tile plot with cell counts.
#'
#' @param actual Integer/numeric vector of 0/1 true labels.
#' @param predicted Integer/numeric vector of 0/1 predicted labels.
#' @param title Main title of the plot.
#'
#' @return A ggplot2 object.
plot_cm <- function(actual, predicted, title = "Confusion Matrix") {
  conf_tib(actual, predicted) %>%
    ggplot(aes(Predicted, Actual, fill = n)) +
    geom_tile(colour = "white") +
    geom_text(aes(label = scales::comma(n)), size = 5.5, fontface = "bold") +
    scale_fill_gradient(low = "#e8f4f8", high = "#2c7bb6", guide = "none") +
    scale_x_discrete(labels = c("0" = "No arrest", "1" = "Arrest")) +
    scale_y_discrete(labels = c("0" = "No arrest", "1" = "Arrest")) +
    labs(title = title, x = "Predicted", y = "Actual") +
    theme_minimal(base_size = 12) +
    theme(panel.grid = element_blank())
}


#' @title Compute Threshold-Dependent Classification Metrics
#'
#' @description Computes sensitivity, specificity, balanced accuracy, PPV,
#' and NPV for a fixed set of hard class predictions against the true labels.
#'
#' @param actual Integer/numeric vector of 0/1 true labels.
#' @param predicted Integer/numeric vector of 0/1 predicted labels.
#' @param pos Value in `actual`/`predicted` denoting the positive class.
#'
#' @return A named list with sensitivity, specificity, balanced_acc, ppv, npv.
class_metrics <- function(actual, predicted, pos = 1) {
  tp <- sum(actual == pos & predicted == pos)
  tn <- sum(actual != pos & predicted != pos)
  fp <- sum(actual != pos & predicted == pos)
  fn <- sum(actual == pos & predicted != pos)
  sens <- tp / (tp + fn)
  spec <- tn / (tn + fp)
  list(
    sensitivity  = sens,
    specificity  = spec,
    balanced_acc = (sens + spec) / 2,
    ppv          = tp / (tp + fp),
    npv          = tn / (tn + fn)
  )
}


# ==============================================================================
# 4. Probability-Based Evaluation Metrics
# ==============================================================================

#' @title Compute the Full Evaluation-Metrics Tibble for One Model
#'
#' @description Given predicted probabilities and true labels, computes
#' threshold-free metrics (AUC, PR-AUC, log-loss, Brier score) plus
#' threshold-dependent metrics at the Youden-optimal cut-off (sensitivity,
#' specificity, precision, F1, MCC, balanced accuracy). Returns one row per
#' call, so results from multiple models can be row-bound into a comparison
#' table.
#'
#' @param actual Integer/numeric vector of 0/1 true labels.
#' @param prob Numeric vector of predicted P(arrested = 1), same length as `actual`.
#' @param model_name Label stored in the Model column.
#'
#' @return A one-row tibble of rounded metrics (see column names in the body).
compute_metrics <- function(actual, prob, model_name = "Model") {
  valid  <- !is.na(actual) & !is.na(prob)
  actual <- actual[valid]; prob <- prob[valid]
  roc_obj <- pROC::roc(actual, prob, quiet = TRUE)
  auc_val <- as.numeric(pROC::auc(roc_obj))
  thr     <- best_thresh(roc_obj)
  pred_cl <- as.integer(prob >= thr)

  tp <- sum(actual == 1 & pred_cl == 1); tn <- sum(actual == 0 & pred_cl == 0)
  fp <- sum(actual == 0 & pred_cl == 1); fn <- sum(actual == 1 & pred_cl == 0)
  sens <- if ((tp+fn) == 0) NA_real_ else tp / (tp+fn)
  spec <- if ((tn+fp) == 0) NA_real_ else tn / (tn+fp)
  prec <- if ((tp+fp) == 0) NA_real_ else tp / (tp+fp)
  f1   <- if (is.na(prec) || is.na(sens) || (prec+sens) == 0) NA_real_ else
            2 * prec * sens / (prec + sens)

  # Matthews correlation coefficient — guard the denominator against any
  # degenerate (all-one-class) confusion matrix, which would divide by zero
  mcc_den <- sqrt(as.numeric(tp+fp) * as.numeric(tp+fn) *
                  as.numeric(tn+fp) * as.numeric(tn+fn))
  mcc_val <- if (mcc_den == 0) NA_real_ else (tp*tn - fp*fn) / mcc_den

  # eps floors probabilities away from 0/1 so log() never evaluates to -Inf
  eps     <- 1e-15
  logloss <- -mean(actual * log(pmax(prob, eps)) + (1-actual) * log(pmax(1-prob, eps)))
  brier   <- mean((prob - actual)^2)

  # PR-AUC via trapezoid rule over pROC coords (pROC has no built-in PR-AUC)
  pr_raw <- pROC::coords(roc_obj, "all",
                          ret = c("sensitivity", "precision"),
                          transpose = FALSE, drop = FALSE)
  names(pr_raw)[names(pr_raw) == "sensitivity"] <- "recall"
  pr_raw <- pr_raw %>%
    filter(!is.nan(precision), !is.na(precision), !is.nan(recall)) %>%
    arrange(recall)
  pr_auc_val <- if (nrow(pr_raw) >= 2)
    sum(diff(pr_raw$recall) *
          (head(pr_raw$precision, -1) + tail(pr_raw$precision, -1)) / 2)
  else NA_real_

  tibble(
    Model       = model_name,
    AUC         = round(auc_val, 4),
    PR_AUC      = round(pr_auc_val, 4),
    Log_Loss    = round(logloss, 4),
    Brier       = round(brier, 4),
    Threshold   = round(thr, 4),
    Sensitivity = round(sens, 3),
    Specificity = round(spec, 3),
    Precision   = round(prec, 3),
    F1          = round(f1, 3),
    MCC         = round(mcc_val, 3),
    Bal_Acc     = round(if (!is.na(sens) && !is.na(spec)) (sens+spec)/2
                        else NA_real_, 3)
  )
}
```

### 0.2.3) Model Evaluation Plot Functions

```{r fn-plot-roc-single}
# ==============================================================================
# 5. Plots for Single-Model Evaluation
# ==============================================================================

#' @title Plot the ROC Curve for a Single Model
#'
#' @description Plots sensitivity vs specificity across all thresholds, with
#' the AUC annotated on the panel.
#'
#' @param actual Integer/numeric vector of 0/1 true labels.
#' @param prob Numeric vector of predicted P(arrested = 1), same length as `actual`.
#' @param title Main title of the plot.
#'
#' @return A ggplot2 object.
plot_roc_single <- function(actual, prob, title = "ROC Curve") {
  roc_obj <- pROC::roc(actual, prob, quiet = TRUE)
  auc_lbl <- paste0("AUC = ", round(as.numeric(pROC::auc(roc_obj)), 4))
  pROC::ggroc(roc_obj, colour = "#2c7bb6", linewidth = 0.9) +
    geom_abline(slope = 1, intercept = 1, linetype = "dashed", colour = "grey60") +
    annotate("text", x = 0.25, y = 0.06, label = auc_lbl,
             colour = "#2c7bb6", size = 4) +
    labs(x = "Specificity", y = "Sensitivity", title = title) +
    theme_minimal(base_size = 11) + theme(panel.grid.minor = element_blank())
}
```

```{r fn-plot-pr-single}
#' @title Plot the Precision-Recall Curve for a Single Model
#'
#' @description Plots precision vs recall across all thresholds against the
#' no-skill baseline (the class prevalence), with PR-AUC annotated. More
#' informative than the ROC curve under the severe class imbalance here.
#'
#' @param actual Integer/numeric vector of 0/1 true labels.
#' @param prob Numeric vector of predicted P(arrested = 1), same length as `actual`.
#' @param title Main title of the plot.
#'
#' @return A ggplot2 object.
plot_pr_single <- function(actual, prob, title = "Precision-Recall Curve") {
  roc_obj <- pROC::roc(actual, prob, quiet = TRUE)
  pr_raw  <- pROC::coords(roc_obj, "all",
                           ret = c("sensitivity", "precision"),
                           transpose = FALSE, drop = FALSE)
  names(pr_raw)[names(pr_raw) == "sensitivity"] <- "recall"
  pr_df <- pr_raw %>%
    filter(!is.nan(precision), !is.na(precision)) %>% arrange(recall)

  # PR-AUC via trapezoid rule, same approach as in compute_metrics()
  pr_auc_lbl <- if (nrow(pr_df) >= 2) {
    v <- sum(diff(pr_df$recall) *
               (head(pr_df$precision,-1) + tail(pr_df$precision,-1)) / 2)
    paste0("PR-AUC = ", round(v, 4))
  } else "PR-AUC = NA"
  baseline <- mean(actual)
  ggplot(pr_df, aes(x = recall, y = precision)) +
    geom_line(colour = "#d73027", linewidth = 0.9) +
    geom_hline(yintercept = baseline, linetype = "dashed", colour = "grey55") +
    annotate("text", x = 0.02, y = baseline + 0.015,
             label = paste0("No-skill = ", round(baseline, 3)),
             colour = "grey40", size = 3.2, hjust = 0) +
    annotate("text", x = 0.55, y = 0.90, label = pr_auc_lbl,
             colour = "#d73027", size = 4) +
    scale_x_continuous(limits = c(0, 1)) + scale_y_continuous(limits = c(0, 1)) +
    labs(x = "Recall", y = "Precision", title = title) +
    theme_minimal(base_size = 11) + theme(panel.grid.minor = element_blank())
}
```

```{r fn-plot-calibration-single}
#' @title Plot the Calibration (Reliability) Curve for a Single Model
#'
#' @description Bins predictions into `n_bins` equal-frequency groups and
#' plots mean predicted probability against observed positive rate per bin,
#' against the perfect-calibration diagonal.
#'
#' @param actual Integer/numeric vector of 0/1 true labels.
#' @param prob Numeric vector of predicted P(arrested = 1), same length as `actual`.
#' @param title Main title of the plot.
#' @param n_bins Number of quantile bins of `prob`.
#'
#' @return A ggplot2 object.
plot_calibration_single <- function(actual, prob, title = "Calibration Plot",
                                    n_bins = 10) {
  # Quantile breaks (not equal-width) so each bin has a comparable sample size
  # despite the predicted probabilities being heavily right-skewed
  brks  <- unique(quantile(prob, seq(0, 1, length.out = n_bins + 1), na.rm = TRUE))
  cal_df <- tibble(actual = actual, prob = prob) %>%
    mutate(bin = cut(prob, breaks = brks, include.lowest = TRUE)) %>%
    filter(!is.na(bin)) %>%
    group_by(bin) %>%
    summarise(mean_pred = mean(prob), mean_obs = mean(actual),
              n = n(), .groups = "drop")
  ggplot(cal_df, aes(x = mean_pred, y = mean_obs)) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey55") +
    geom_line(colour = "#2c7bb6", linewidth = 0.8) +
    geom_point(aes(size = n), colour = "#2c7bb6", alpha = 0.85) +
    scale_size_continuous(name = "Count", range = c(2, 9)) +
    scale_x_continuous(labels = scales::percent_format(accuracy = 0.1)) +
    scale_y_continuous(labels = scales::percent_format(accuracy = 0.1)) +
    labs(x = "Mean predicted probability", y = "Observed positive rate",
         title = title,
         caption = "Size ∝ bin count.  Dashed = perfect calibration.") +
    theme_minimal(base_size = 11) +
    theme(legend.position = "bottom", panel.grid.minor = element_blank())
}
```

```{r fn-plot-threshold-single}
#' @title Plot the Threshold-Sweep Metrics for a Single Model
#'
#' @description Plots sensitivity, specificity, precision, and F1 as
#' functions of the classification threshold, with the Youden-optimal
#' threshold marked.
#'
#' @param actual Integer/numeric vector of 0/1 true labels.
#' @param prob Numeric vector of predicted P(arrested = 1), same length as `actual`.
#' @param title Main title of the plot.
#'
#' @return A ggplot2 object.
plot_threshold_single <- function(actual, prob, title = "Threshold Analysis") {
  roc_obj  <- pROC::roc(actual, prob, quiet = TRUE)
  thr_data <- pROC::coords(roc_obj, "all",
                            ret = c("threshold","sensitivity","specificity","precision"),
                            transpose = FALSE, drop = FALSE) %>%
    filter(is.finite(threshold), threshold >= 0, threshold <= 1) %>%
    mutate(f1 = ifelse(!is.nan(precision) & (precision + sensitivity) > 0,
                       2 * precision * sensitivity / (precision + sensitivity), 0)) %>%
    pivot_longer(c(sensitivity, specificity, precision, f1),
                 names_to = "Metric", values_to = "Value") %>%
    mutate(Metric = recode(Metric,
      sensitivity = "Sensitivity", specificity = "Specificity",
      precision   = "Precision",   f1          = "F1 Score"))
  thr_opt <- best_thresh(roc_obj)
  ggplot(thr_data, aes(x = threshold, y = Value, colour = Metric)) +
    geom_line(linewidth = 0.75, na.rm = TRUE) +
    geom_vline(xintercept = thr_opt, linetype = "dashed",
               colour = "grey35", linewidth = 0.55) +
    annotate("text", x = thr_opt, y = 0.04,
             label = paste0("Youden\n", round(thr_opt, 3)),
             hjust = -0.08, size = 3, colour = "grey30") +
    scale_colour_viridis_d(option = "turbo", begin = 0.05, end = 0.92) +
    scale_y_continuous(limits = c(0, 1)) +
    labs(x = "Threshold", y = "Metric value", colour = NULL, title = title,
         caption = "Dashed: Youden's J optimal threshold") +
    theme_minimal(base_size = 11) +
    theme(legend.position = "bottom", panel.grid.minor = element_blank())
}
```

```{r fn-plot-eval-quad}
#' @title Plot the 2x2 Evaluation Panel for a Single Model
#'
#' @description Composes plot_roc_single(), plot_pr_single(),
#' plot_calibration_single(), and plot_threshold_single() into one
#' (ROC | PR) / (Calibration | Threshold) patchwork layout.
#'
#' @param actual Integer/numeric vector of 0/1 true labels.
#' @param prob Numeric vector of predicted P(arrested = 1), same length as `actual`.
#' @param model_name Prefixed to each sub-plot's title.
#'
#' @return A patchwork object (composed ggplot2 object).
eval_quad_plot <- function(actual, prob, model_name) {
  p1 <- plot_roc_single(actual, prob,   paste(model_name, "— ROC"))
  p2 <- plot_pr_single(actual, prob,    paste(model_name, "— Precision-Recall"))
  p3 <- plot_calibration_single(actual, prob, paste(model_name, "— Calibration"))
  p4 <- plot_threshold_single(actual, prob, paste(model_name, "— Threshold Analysis"))
  (p1 | p2) / (p3 | p4)
}
```

```{r fn-plot-nb-feature}
#' @title Plot a Naive Bayes Conditional-Probability Bar Chart
#'
#' @description Plots P(feature level | class) from one entry of a fitted
#' naiveBayes model's frequency tables, grouped by class, so the strongest
#' discriminators under the conditional-independence assumption are visible.
#'
#' @param feature_name Name of a predictor in `mod$tables`.
#' @param mod A fitted e1071::naiveBayes model object.
#'
#' @return A ggplot2 object.
plot_nb_feature <- function(feature_name, mod) {
  tbl <- mod$tables[[feature_name]]
  as.data.frame(tbl) %>%
    rename(level = 2, class = 1, prob = Freq) %>%
    mutate(class = factor(class, labels = c("Not arrested", "Arrested"))) %>%
    ggplot(aes(x = level, y = prob, fill = class)) +
    geom_col(position = position_dodge(width = 0.75), width = 0.65) +
    scale_fill_viridis_d(option = "mako", begin = 0.3, end = 0.8) +
    scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
    labs(x = NULL, y = "P(level | class)", title = feature_name, fill = NULL) +
    theme_minimal(base_size = 9) +
    theme(
      axis.text.x    = element_text(angle = 30, hjust = 1),
      legend.position = "bottom",
      panel.grid.minor = element_blank()
    )
}
```

### 0.2.4) Model-Summary Helpers

```{r fn-build-modal-row}
# ==============================================================================
# 6. Marginal-Effect and Coefficient Helpers
# ==============================================================================

#' @title Build a One-Row Data Frame of Modal/Median Covariate Values
#'
#' @description For each variable in `vars`, takes the most frequent level
#' (factor/character) or the median (numeric) from `data`. Used to hold
#' non-focal predictors fixed at a representative value when computing
#' marginal-effect predictions for an interaction term.
#'
#' @param data Data.frame to summarise (typically `train`).
#' @param vars Character vector of column names in `data`.
#'
#' @return A one-row data.frame with one column per entry in `vars`.
build_modal_row <- function(data, vars) {
  row_list <- lapply(vars, function(v) {
    x <- data[[v]]
    if (is.factor(x)) {
      factor(names(sort(table(x), decreasing = TRUE))[1], levels = levels(x))
    } else if (is.numeric(x)) {
      median(x, na.rm = TRUE)
    } else {
      names(sort(table(x), decreasing = TRUE))[1]
    }
  })
  as.data.frame(setNames(row_list, vars))
}
```

```{r fn-extract-coefs}
#' @title Extract Non-Zero LASSO Coefficients at a Given Lambda
#'
#' @description Extracts the coefficients selected by a cv.glmnet fit at a
#' named lambda (e.g. "lambda.min" or "lambda.1se"), drops the zeroed-out
#' coefficients, and adds the exponentiated odds ratio.
#'
#' @param cv_fit A fitted glmnet::cv.glmnet object.
#' @param s_name Lambda selector passed to coef(cv_fit, s = s_name), typically
#' "lambda.min" or "lambda.1se".
#'
#' @return A tibble with columns feature, estimate (log-odds), OR (odds ratio),
#' sorted by descending |estimate|.
extract_coefs <- function(cv_fit, s_name) {
  cf <- coef(cv_fit, s = s_name)
  tibble(
    feature  = rownames(cf)[-1],
    estimate = as.numeric(cf[-1])
  ) %>%
    filter(estimate != 0) %>%
    arrange(desc(abs(estimate))) %>%
    mutate(OR = round(exp(estimate), 3), estimate = round(estimate, 4))
}
```

---

# 1) Data Loading and Preparation

```{r load-data}
vehstop <- read.csv("data/vehiclestops_4_wheels_final.csv")
```

## 1.1) Factor conversion and Reference Levels

We convert all categorical predictors to factors and set interpretable reference levels before any modelling step, ensuring consistency between all downstream model fits and cross-validation folds.

```{r factor-prep}
# Ordered weekday levels (ISO order)
wday_levels <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday")

# Ordered time bracket levels
tbkt_levels <- c(
  "Late night (0–3)",
  "Early morning (4–7)",
  "Morning commute (8–11)",
  "Midday (12–15)",
  "Afternoon/evening (16–19)",
  "Night (20–23)"
)

# Ordered age-bracket levels
abkt_levels <- c("Under 18", "18–24", "25–34", "35–44", "45–54", "55–64", "65+", "Unknown")

vehstop <- vehstop %>%
  mutate(
    year = factor(year),
    month = factor(month, levels = 1:12),
    weekday = factor(weekday, levels = wday_levels),
    hour = factor(hour, levels = 0:23),
    time_bracket = factor(time_bracket, levels = tbkt_levels),
    area = factor(area),
    boro = factor(boro),
    sex = factor(sex),
    ethnicity = factor(ethnicity),
    age_bracket = factor(age_bracket, levels = abkt_levels)
  )

# Reference levels chosen for interpretable contrasts
vehstop$sex <- relevel(vehstop$sex, ref = "M")
vehstop$ethnicity <- relevel(vehstop$ethnicity, ref = "White")
vehstop$weekday <- relevel(vehstop$weekday, ref = "Monday")
vehstop$time_bracket <- relevel(vehstop$time_bracket, ref = "Morning commute (8–11)")
vehstop$age_bracket <- relevel(vehstop$age_bracket, ref = "25–34")

rm(wday_levels, tbkt_levels, abkt_levels)
```


## 1.2) Stratified Train / Test Split

For the train/test split, a **stratified sampling approach** is used to ensure that the training and test sets have comparable distributions of the outcome variable (`arrested`) and relevant contextual variables. 

This is particularly important because the target variable is imbalanced, with arrested cases representing only a small proportion of all observations. Rather than stratifying only on `arrested`, the code constructs a **joint stratification variable** from `arrested`, `area`, `sex`, `ethnicity`, and `hour` using `interaction()`. Thus, each observation is assigned to a stratum corresponding to a particular combination of these variables. 

Formally, the strata can be represented as

$$
S_i = (\text{arrested}_i,\text{area}_i,\text{sex}_i,\text{ethnicity}_i,\text{hour}_i),
$$

where $S_i$ denotes the stratum to which observation $i$ belongs. The objective is to sample approximately 80% of observations within each stratum for the training set and retain the remaining 20% for testing:

$$
P(i \in \mathcal{D}_{train} \mid S_i=s) \approx 0.8,
\qquad
P(i \in \mathcal{D}_{test} \mid S_i=s) \approx 0.2.
$$

```{r stratified-split}
set.seed(2026)

# Joint stratum key; drop = TRUE removes empty combinations
vehstop$.strat <- interaction(
  vehstop$arrested, vehstop$area, vehstop$sex, vehstop$ethnicity, vehstop$hour,
  drop = TRUE
)

train_rows <- vehstop %>%
  dplyr::mutate(.row = row_number()) %>%
  dplyr::group_by(.strat) %>%
  dplyr::slice_sample(prop = 0.8) %>%
  dplyr::pull(.row)

train <- vehstop[ train_rows, ] %>% dplyr::select(-.strat)
test  <- vehstop[-train_rows, ] %>% dplyr::select(-.strat)
vehstop <- vehstop %>% dplyr::select(-.strat)
```

In the code, `group_by(.strat) %>% slice_sample(prop = 0.8)` performs this within-stratum sampling, while the observations not selected for training form the test set. This approach aims to preserve not only the overall prevalence of the outcome,

$$
P_{\text{train}}(\text{arrested}=1) \approx P_{\text{test}}(\text{arrested}=1),
$$

but also the composition of the data across geographic, demographic, and temporal characteristics. This is desirable because differences in these characteristics could otherwise introduce an unintended distributional difference between the training and test sets. The temporary `.strat` variable is removed after the split because it is used solely to define the sampling groups and is not intended to be used as a predictor in the subsequent modelling. 

Finally, the verification code compares the empirical marginal distributions of `arrested`, `area`, `sex`, and `ethnicity` in the two datasets. For a variable $X$ and level $k$, the proportion in each dataset is calculated as

$$
\hat{P}_{train}(X=k)
=
\frac{\sum_{i \in \mathcal{D}_{train}}\mathbb{I}(X_i=k)}
{|\mathcal{D}_{train}|},
$$

and analogously for the test set. The difference is then reported in **percentage points**:

$$
\Delta_k =
100\left[
\hat{P}_{test}(X=k)
-
\hat{P}_{train}(X=k)
\right].
$$

Small values of $\Delta_k$ indicate that the corresponding characteristic is similarly represented in the two datasets. The final summary additionally compares the number of observations and the proportion of arrested cases in the training and test sets, providing a direct check that the intended approximately 80/20 split and the class distribution have been maintained.

```{r split-verification}
bind_rows(
  check_prop("arrested", "arrested"),
  check_prop("area", "area"),
  check_prop("sex", "sex"),
  check_prop("ethnicity", "ethnicity")
) %>%
  kbl(
    digits = 4,
    caption = "Proportion of each level in train vs test (diff in percentage points)",
    col.names = c("Variable", "Level", "Train", "Test", "Δ (ppt)")
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE) %>%
  collapse_rows(columns = 1, valign = "top")
```

```{r split-summary}
cat(sprintf(
  "Training set : %d rows  |  Arrested: %.2f%%\nTest set     : %d rows  |  Arrested: %.2f%%\n",
  nrow(train), 100 * mean(train$arrested),
  nrow(test),  100 * mean(test$arrested)
))
```

The maximum absolute discrepancy between the two sets is on the order of fractions of a percentage point, confirming that the stratification preserved the joint distribution of the target and the key covariates.

```{r remove-vehstop}
# Remove vehstop for memory purposes
rm(vehstop)
```

---

# 5) LASSO

**Why did we choose this model?** Given the size of the dataset and the rarity of the target, a penalised logistic regression is a trivial choice: 

- It scales linearly in $n$, 
- It handles many correlated dummy-encoded predictors, and
- It performs feature selection (derived from the $\ell_1$ penalty). 

**Modus Operandi:** In this section we fit a LASSO logistic regression on `arrested` using the reduced predictor set (see below), then we compare the model obtained at $\lambda_{\min}$ against the more parsimonious one at $\lambda_{\text{1se}}$.

## 5.1) Assumptions and Setup

**LASSO (Least Absolute Shrinkage and Selection Operator)** is a regularization method that extends logistic regression by adding a penalty on the absolute size of the model coefficients. 

In binary classification, it is particularly useful when there are many potentially relevant predictors and we want to identify a smaller set of variables associated with the probability of the positive outcome.

For a binary target \(Y \in \{0,1\}\), LASSO logistic regression models

$$ P(Y=1\mid X)=\frac{1}{1+\exp(-\eta)}, $$

where

$$ \eta=\beta_0+\beta_1X_1+\cdots+\beta_pX_p. $$

Instead of simply maximizing the logistic likelihood, LASSO minimizes the penalized negative log-likelihood:

$$ \hat{\beta} = \arg\min_{\beta} \left\{ -\ell(\beta) + \lambda\sum_{j=1}^{p}|\beta_j| \right\}. $$

Here, $\lambda \geq 0$ controls the strength of regularization.

> In other words, this model penalises the sum of absolute coefficient values ($\ell_1$ penalty), which shrinks small effects to exactly zero and performs implicit variable selection.

We had to exclude `age` as a predictor, since we had missing values that we labeled as `Unknown` in `age_bracket`. Therefore, this is the final formulation of the model

TODO: Write in latex the formulation

```{r final-model-formula}
f_final <- arrested ~ year + month + weekday + hour + boro + sex + ethnicity + age_bracket
```

### 5.1.1) Model Matrix

Unlike `glm()`, `glmnet()` does not accept a formula and factor predictors directly: it requires a numeric design matrix, so every categorical predictor must be dummy-encoded first via `model.matrix()` (see also section 1.1).

Then, the intercept column produced by `model.matrix()` is dropped (`[, -1]`) because `glmnet` fits its own unpenalised intercept internally. 

Standardisation of the columns before penalisation is handled automatically by `glmnet` (`standardize = TRUE` is the default), so no manual scaling is needed here. 

```{r lasso-matrix}
mm_train <- model.frame(f_final, data = train, na.action = na.omit)
X_train <- model.matrix(f_final, data = mm_train)[, -1]
y_train <- mm_train$arrested

mm_test <- model.frame(f_final, data = test, na.action = na.omit)
X_test <- model.matrix(f_final, data = mm_test)[, -1]
y_test <- mm_test$arrested

cat("Training matrix:", nrow(X_train), "×", ncol(X_train), "\n")
cat("Test matrix :", nrow(X_test),  "×", ncol(X_test),  "\n")
```

## 5.2) Hyperparameter Tuning and Model Fit — Cross-Validated Lambda

The regularization parameter is typically selected using cross-validation. With an imbalanced binary outcome, it is important that the cross-validation folds preserve approximately the same proportion of positive cases. Unfortunately, we did not have enough time to create a more precise CV function to take into account proportions in between folds.

We optimise $\lambda$ by maximising cross-validated AUC (`type.measure = "auc"`), which is the appropriate metric for an imbalanced binary outcome. In fact, for a $3-5\%$ positive class, ROC-AUC is generally preferable to classification error, because the latter can be dominated by the majority class.

```{r fit-lasso-cv, cache=TRUE}
set.seed(2026)
cv_lasso <- cv.glmnet(
  X_train, y_train,
  family = "binomial",
  alpha = 1,
  nfolds = 5,
  type.measure = "auc"
)

cat(sprintf(
  "λ.min  = %.6f  (CV AUC = %.4f)\nλ.1se  = %.6f  (CV AUC = %.4f)\n",
  cv_lasso$lambda.min, max(cv_lasso$cvm),
  cv_lasso$lambda.1se,
  cv_lasso$cvm[cv_lasso$lambda == cv_lasso$lambda.1se]
))
```

TODO: Comment (brief, we will se in detail later)

```{r lasso-cv-plot, fig.width=7, fig.height=4}
# Replicate the cv.glmnet plot in ggplot2
cv_df <- tibble(
  log_lambda = log(cv_lasso$lambda),
  cvm = cv_lasso$cvm,
  cvup = cv_lasso$cvup,
  cvlo = cv_lasso$cvlo,
  nzero = cv_lasso$nzero
)

ggplot(cv_df, aes(x = log_lambda, y = cvm)) +
  geom_ribbon(aes(ymin = cvlo, ymax = cvup), alpha = 0.15, fill = "#2c7bb6") +
  geom_line(colour = "#2c7bb6") +
  geom_point(size = 0.8, colour = "#2c7bb6") +
  geom_vline(xintercept = log(cv_lasso$lambda.min), linetype = "dashed",
             colour = "#d73027", linewidth = 0.7) +
  geom_vline(xintercept = log(cv_lasso$lambda.1se), linetype = "dashed",
             colour = "#fdae61", linewidth = 0.7) +
  annotate("text", x = log(cv_lasso$lambda.min), y = min(cv_df$cvlo),
           label = "λ.min", colour = "#d73027", hjust = -0.1, size = 3.5) +
  annotate("text", x = log(cv_lasso$lambda.1se), y = min(cv_df$cvlo),
           label = "λ.1se", colour = "#b36200", hjust = -0.1, size = 3.5) +
  labs(
    x     = "log(λ)",
    y     = "5-fold CV AUC",
    title = "LASSO — Cross-validated AUC vs regularisation strength"
  ) +
  theme_minimal(base_size = 11) +
  theme(panel.grid.minor = element_blank())
```

TODO: Comment

## 5.3) Specific Analysis and Interpretation

### 5.3.1) Lambda and Interpretation

TODO: Show the lambdas and do a comment like this:

EXAMPLE:

```txt
Lasso Logistic Regression — Cross-Validation for Lambda Selection (Interpretation)

    λ_min = 0.00150778 (x = -log(λ) = 6.4971)
        CV binomial deviance = 0.7564 (SE = 0.0148)
        Model complexity: 16 non-zero coefficients
        Interpretation: this is the best CV performance (lowest deviance), but keeps more predictors.
    λ_1se = 0.0140616 (x = -log(λ) = 4.2643)
        CV binomial deviance = 0.7698 (SE = 0.0145)
        Model complexity: 8 non-zero coefficients
        Interpretation: this is the 1-SE rule choice: slightly worse deviance but a much simpler model.

What this tells us

    The deviance improvement from λ_1se to λ_min is 0.7698 − 0.7564 = 0.0134, which is small.
    In exchange, λ_1se reduces the model from 16 → 8 predictors (about 50% fewer).

Practical decision

    If the goal is maximum predictive performance, choose λ_min.
    If the goal is interpretability + stability with almost the same performance, choose λ_1se
```

### 5.3.2) Coefficient Path

TODO: Introduce what this plot means, what it represents

```{r lasso-path-plot, fig.width=9, fig.height=5}
# Tidy the full coefficient path
coef_path <- as.matrix(coef(cv_lasso$glmnet.fit))   # (p+1) × n_lambda
lambdas   <- cv_lasso$glmnet.fit$lambda

path_df <- as.data.frame(t(coef_path[-1, ])) %>%   # drop intercept row
  mutate(log_lambda = log(lambdas)) %>%
  pivot_longer(-log_lambda, names_to = "feature", values_to = "coef") %>%
  filter(coef != 0)

# Only label non-zero at lambda.min for clarity
active_min <- names(which(coef(cv_lasso, s = "lambda.min")[-1, 1] != 0))

ggplot(path_df, aes(x = log_lambda, y = coef, group = feature,
                    colour = feature %in% active_min)) +
  geom_line(linewidth = 0.45, alpha = 0.8) +
  geom_vline(xintercept = log(cv_lasso$lambda.min), linetype = "dashed",
             colour = "#d73027", linewidth = 0.7) +
  geom_vline(xintercept = log(cv_lasso$lambda.1se), linetype = "dashed",
             colour = "#fdae61", linewidth = 0.7) +
  scale_colour_manual(values = c("TRUE" = "#2c7bb6", "FALSE" = "grey70"),
                      guide  = "none") +
  labs(
    x       = "log(λ)",
    y       = "Coefficient",
    title   = "LASSO — Coefficient paths",
    caption = "Blue paths: active at λ.min  |  Dashed lines: λ.min (red) and λ.1se (orange)"
  ) +
  theme_minimal(base_size = 11) +
  theme(panel.grid.minor = element_blank())
```

TODO: Comment

### 5.3.2) Selected Non-Zero Coefficients

```{r lasso-coef-tables}
coef_min <- extract_coefs(cv_lasso, "lambda.min")
coef_1se <- extract_coefs(cv_lasso, "lambda.1se")

cat("Non-zero coefficients at λ.min:", nrow(coef_min), "\n")
cat("Non-zero coefficients at λ.1se:", nrow(coef_1se), "\n")
```

#### 5.3.2.1) Non-Zero Coefficients ats λ.min

```{r lasso-coef-min-table}
coef_min %>%
  kbl(
    caption   = "LASSO coefficients at λ.min (sorted by |estimate|)",
    col.names = c("Feature", "Log-odds", "Odds Ratio")
  ) %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE)
```

Path plot of the lasso estimates as a function of $\lambda$ on the logarithm scale

```{r lasso-coef-min-plot, fig.width=8, fig.height=0.35 * nrow(coef_min) + 1.5}
coef_min %>%
  mutate(
    direction = ifelse(estimate > 0, "Increases odds", "Decreases odds"),
    feature   = stringr::str_wrap(feature, 30)
  ) %>%
  ggplot(aes(x = estimate, y = reorder(feature, estimate), fill = direction)) +
  geom_col(width = 0.7) +
  geom_vline(xintercept = 0, linewidth = 0.4) +
  scale_fill_manual(values = c("Increases odds" = "#d73027", "Decreases odds" = "#2c7bb6")) +
  labs(
    x     = "LASSO coefficient (log-odds scale)",
    y     = NULL,
    fill  = NULL,
    title = "LASSO — Non-zero coefficients at λ.min"
  ) +
  theme_minimal(base_size = 10) +
  theme(legend.position = "bottom", panel.grid.minor = element_blank())
```

TODO: Comment

#### 5.3.2.2) Non-Zero Coefficients ats λ.1se

TODO:

```{r lasso-coef-1se-table}
coef_1se %>%
  kbl(
    caption   = "LASSO coefficients at λ.1se (sorted by |estimate|)",
    col.names = c("Feature", "Log-odds", "Odds Ratio")
  ) %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE)
```

Path plot of the lasso estimates as a function of $\lambda$ on the logarithm scale

```{r lasso-coef-1se-plot, fig.width=8, fig.height=0.35 * nrow(coef_min) + 1.5}
coef_1se %>%
  mutate(
    direction = ifelse(estimate > 0, "Increases odds", "Decreases odds"),
    feature   = stringr::str_wrap(feature, 30)
  ) %>%
  ggplot(aes(x = estimate, y = reorder(feature, estimate), fill = direction)) +
  geom_col(width = 0.7) +
  geom_vline(xintercept = 0, linewidth = 0.4) +
  scale_fill_manual(values = c("Increases odds" = "#d73027", "Decreases odds" = "#2c7bb6")) +
  labs(
    x     = "LASSO coefficient (log-odds scale)",
    y     = NULL,
    fill  = NULL,
    title = "LASSO — Non-zero coefficients at λ.1se"
  ) +
  theme_minimal(base_size = 10) +
  theme(legend.position = "bottom", panel.grid.minor = element_blank())
```

TODO: Comment

## 5.4) Evaluation on Test Set

TODO: Intro of metrics used and why

```{r lasso-eval}
# TODO: Move the
prob_lasso_min <- as.numeric(predict(cv_lasso, newx = X_test, s = "lambda.min", type = "response"))
prob_lasso_1se <- as.numeric(predict(cv_lasso, newx = X_test, s = "lambda.1se", type = "response"))

roc_lasso_min <- pROC::roc(y_test, prob_lasso_min, quiet = TRUE)
roc_lasso_1se <- pROC::roc(y_test, prob_lasso_1se, quiet = TRUE)
auc_lasso_min <- as.numeric(pROC::auc(roc_lasso_min))
auc_lasso_1se <- as.numeric(pROC::auc(roc_lasso_1se))

thr_lasso_min  <- best_thresh(roc_lasso_min)
class_lasso_min <- as.integer(prob_lasso_min >= thr_lasso_min)
met_lasso_min   <- class_metrics(y_test, class_lasso_min)

cat(sprintf(
  "λ.min — AUC: %.4f  |  Threshold: %.4f  |  Sens: %.3f  |  Spec: %.3f  |  Bal.Acc: %.3f\n",
  auc_lasso_min, thr_lasso_min,
  met_lasso_min$sensitivity, met_lasso_min$specificity, met_lasso_min$balanced_acc
))

thr_lasso_1se   <- best_thresh(roc_lasso_1se)
class_lasso_1se <- as.integer(prob_lasso_1se >= thr_lasso_1se)
met_lasso_1se   <- class_metrics(y_test, class_lasso_1se)

cat(sprintf(
  "λ.1se — AUC: %.4f  |  Threshold: %.4f  |  Sens: %.3f  |  Spec: %.3f  |  Bal.Acc: %.3f\n",
  auc_lasso_1se, thr_lasso_1se,
  met_lasso_1se$sensitivity, met_lasso_1se$specificity, met_lasso_1se$balanced_acc
))
```

TODO: Comment briefly (we see this again down there)

### 5.4.1) Confusion Matrix and FN/FP Inspection

```{r lasso-cm-plots, fig.width=8, fig.height=3.5}
p_cm_min <- plot_cm(y_test, class_lasso_min, "LASSO (λ.min) — Confusion Matrix")
p_cm_1se <- plot_cm(y_test, class_lasso_1se, "LASSO (λ.1se) — Confusion Matrix")

p_cm_min | p_cm_1se
```

TODO: Comment CM

Raw counts of misclassified stops are not directly comparable across groups, since groups differ both in size and in their true prevalence of `arrested`. 

We therefore recover the original (non dummy-encoded) covariates for the test rows via `mm_test`'s row names, split the misclassifications into false positives (predicted arrest, no arrest occurred) and false negatives (missed an actual arrest), and summarise them as **rates conditional on the true class**: the miss rate $\text{FN}/(\text{FN}+\text{TP})$ within each group, and the false-alarm rate $\text{FP}/(\text{FP}+\text{TN})$. This is the same logic behind equal-opportunity/predictive-equality style fairness diagnostics, and is directly relevant to the "excessive enforcement" framing of the project: a group with a much higher false-alarm rate is one the model is comparatively more prone to over-flag, a group with a much higher miss rate is one whose arrests the model comparatively under-detects.

```{r lasso-fpfn-build}
vars_focus <- c("hour", "boro", "sex", "ethnicity", "age_bracket")

# mm_test keeps the original row names of `test`, so this recovers the
# untransformed covariates for every row scored in X_test/y_test
test_ids <- rownames(mm_test)
test_meta <- test[test_ids, vars_focus]

fp_min <- test_meta[y_test == 0 & class_lasso_min == 1, ]
fn_min <- test_meta[y_test == 1 & class_lasso_min == 0, ]
fp_1se <- test_meta[y_test == 0 & class_lasso_1se == 1, ]
fn_1se <- test_meta[y_test == 1 & class_lasso_1se == 0, ]

cat(sprintf(
  "λ.min — FP: %d  |  FN: %d\nλ.1se — FP: %d  |  FN: %d\n",
  nrow(fp_min), nrow(fn_min), nrow(fp_1se), nrow(fn_1se)
))
```

```{r lasso-fpfn-rates, fig.width=13, fig.height=7}
#' @title Group-wise Miss Rate and False-Alarm Rate
#'
#' @description For one grouping variable, computes the miss rate
#' (FN / (FN+TP)) and false-alarm rate (FP / (FP+TN)) per level, given a set
#' of true labels and hard predicted classes.
#'
#' @param group Factor/character vector of group levels, same length as actual.
#' @param actual Integer/numeric vector of 0/1 true labels.
#' @param predicted Integer/numeric vector of 0/1 predicted labels.
#'
#' @return A tibble with columns group, n, miss_rate, false_alarm.
error_rate_by_group <- function(group, actual, predicted) {
  tibble(group = group, actual = actual, predicted = predicted) %>%
    group_by(group) %>%
    summarise(
      n           = n(),
      miss_rate   = sum(actual == 1 & predicted == 0) / sum(actual == 1),
      false_alarm = sum(actual == 0 & predicted == 1) / sum(actual == 0),
      .groups = "drop"
    )
}

error_rates <- purrr::map_dfr(vars_focus, function(v) {
  bind_rows(
    error_rate_by_group(test_meta[[v]], y_test, class_lasso_min) %>% mutate(model = "λ.min"),
    error_rate_by_group(test_meta[[v]], y_test, class_lasso_1se) %>% mutate(model = "λ.1se")
  ) %>% mutate(variable = v)
}) %>% relocate(variable, model, group)

error_rates %>%
  pivot_longer(c(miss_rate, false_alarm), names_to = "metric", values_to = "rate") %>%
  mutate(metric = recode(metric, miss_rate = "Miss rate (FN / actual arrests)",
                          false_alarm = "False-alarm rate (FP / actual non-arrests)")) %>%
  ggplot(aes(x = group, y = rate, fill = model)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.6) +
  facet_grid(metric ~ variable, scales = "free_x", space = "free_x") +
  scale_fill_manual(values = c("λ.min" = "#2c7bb6", "λ.1se" = "#fdae61")) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  labs(x = NULL, y = NULL, fill = NULL,
       title = "LASSO — Miss rate and false-alarm rate by group") +
  theme_minimal(base_size = 9) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "bottom", panel.grid.minor = element_blank())
```

A bar noticeably taller than its neighbours within a facet flags a group the model treats asymmetrically relative to its own true-label base rate, not merely a group with more errors in absolute terms. As with the rest of this analysis, `arrested` is self-reported and observational, so these gaps describe how the fitted model behaves on this test split rather than any causal claim about enforcement itself.

TODO: Comment

### 5.4.2) Performance Analysis

```{r lasso-metrics-table}
bind_rows(
  compute_metrics(y_test, prob_lasso_min, "LASSO λ.min"),
  compute_metrics(y_test, prob_lasso_1se, "LASSO λ.1se")
) %>%
  kbl(
    caption = "LASSO — full evaluation metrics (test set, Youden threshold)",
    col.names = c(
      "Model","AUC","PR-AUC","Log-Loss","Brier", "Threshold","Sensitivity",
      "Specificity","Precision","F1","MCC","Bal. Acc."
    )
  ) %>%
  kable_styling(
    bootstrap_options = c("striped","condensed"),
    full_width = TRUE
  )
```

```{r lasso-eval-plots-min, fig.width=10, fig.height=8}
eval_quad_plot(y_test, prob_lasso_min, "LASSO (λ.min)")
```

```{r lasso-eval-plots-1se, fig.width=10, fig.height=8}
eval_quad_plot(y_test, prob_lasso_1se, "LASSO (λ.1se)")
```

TODO: Comment and Talk about the performances

## 5.5) Diagnostics — Coefficient Paths and CV Deviance vs $\lambda$

LASSO's $\ell_1$ penalty induces sparsity: as $\lambda$ grows, predictors are progressively zeroed out. The coefficient path reveals which predictors survive at $\lambda_{\min}$ and which are eliminated at the more parsimonious $\lambda_{\text{1se}}$.

```{r lasso-diag, fig.width=12, fig.height=5, message=FALSE, warning=FALSE}
lasso_fit <- cv_lasso$glmnet.fit
coef_mat_l <- as.matrix(coef(lasso_fit))[-1, ]
log_lam_l <- log(lasso_fit$lambda)

coef_path_l <- as.data.frame(t(coef_mat_l)) %>%
  mutate(log_lambda = log_lam_l) %>%
  tidyr::pivot_longer(-log_lambda, names_to = "predictor", values_to = "coef")

p_lasso_path <- ggplot(
  coef_path_l, aes(x = log_lambda, y = coef, group = predictor)
  ) +
  geom_line(alpha = 0.35, linewidth = 0.5, colour = "#d73027") +
  geom_vline(
    xintercept = log(cv_lasso$lambda.min), 
    linetype = "dashed", colour = "firebrick", linewidth = 0.7
  ) +
  geom_vline(
    xintercept = log(cv_lasso$lambda.1se), 
    linetype = "dotted", colour = "darkorange", linewidth = 0.7
  ) +
  labs(
    x = expression(log(lambda)), y = "Coefficient", 
    title = "LASSO — Coefficient paths (sparsity)",
    subtitle = "Red dashed = λ.min · Orange dotted = λ.1se"
  ) +
  theme_minimal(base_size = 11) + 
  theme(panel.grid.minor = element_blank())

cv_df_l <- data.frame(
  log_lambda = log(cv_lasso$lambda),
  cvm = cv_lasso$cvm,
  cvup = cv_lasso$cvup,
  cvlo = cv_lasso$cvlo
)

p_lasso_cv <- ggplot(
  cv_df_l, aes(x = log_lambda, y = cvm)
  ) +
  geom_ribbon(
    aes(ymin = cvlo, ymax = cvup),
    alpha = 0.15, fill = "#d73027"
  ) +
  geom_line(colour = "#d73027", linewidth = 0.9) +
  geom_vline(
    xintercept = log(cv_lasso$lambda.min),
    linetype = "dashed", colour = "firebrick", linewidth = 0.7
  ) +
  geom_vline(
    xintercept = log(cv_lasso$lambda.1se),
    linetype = "dotted", colour = "darkorange", linewidth = 0.7
  ) +
  labs(
    x = expression(log(lambda)), y = "CV Binomial Deviance",
    title = "LASSO — Cross-validated deviance",
    subtitle = "Ribbon = ±1 SE"
  ) +
  theme_minimal(base_size = 11) +
  theme(panel.grid.minor = element_blank())

p_lasso_path + p_lasso_cv
```

The point where paths first leave zero marks each predictor's entry into the model. Predictors that survive to $\lambda_{\text{1se}}$ are the most robustly informative signal carriers. A wide flat region in the CV curve indicates that the model is relatively insensitive to $\lambda$ in that range — the bias–variance trade-off is shallow — while a sharp minimum suggests strong sensitivity.

TODO: Comment
