---
title: "Predicting Excessive Enforcement Patterns in Police Behavior After Vehicles Stops in New York"
subtitle: "Model Selection and Evaluation - 4 Wheels Vehicles (Arrested)"
author: "Gianmaria Pizzo"
output:
  html_document:
    fig_caption: yes
    theme: flatly
    highlight: pygments
    code_folding: show
    toc: yes
    toc_depth: 4
    number_sections: no
    toc_float:
      smooth_scroll: no
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

These `.Rmd` files represent the project submission for the course [Statistical Inference and Learning](https://www.unive.it/data/course/521944) by Professor [Cristiano Varin](https://www.unive.it/data/people/5592728) at [Ca' Foscari University of Venice](https://www.unive.it/) for the year 2025/2026.

---

# Problem Statement

## Goal of this notebook

This notebook carries out model selection and evaluation for predicting **whether a two-wheel vehicle stop results in an arrest** (`arrested = 1`). The response is binary and exhibits severe class imbalance: approximately 5.5% of stops lead to an arrest. Accordingly, accuracy alone is a misleading performance metric—we rely instead on log-likelihood-based criteria (AIC, BIC), likelihood ratio tests where models are nested, and cross-validated AUC.

The notebook is organised as follows:

0. **Stratified train/test split** stratified jointly on `arrested`, `area`, `sex`, `ethnicity`, and `hour`.
1. **Full logistic regression model** using `year`, `month`, `weekday`, `time_bracket`, `boro`, `sex`, `ethnicity`, `age_bracket` as predictors, followed by an analysis of coefficients, p-values, and odds ratios.
2. **Predictor-form comparison** for three alternatives: (a) `hour` instead of `time_bracket`; (b) `area` instead of `boro`; (c) numeric `age` instead of `age_bracket`. Each comparison uses AIC, BIC, likelihood ratio tests where applicable, and 10-fold cross-validated AUC on the training set.

---

# Environment

```{r requirements}
requirements <- c(
  "remotes", "summarytools", "forcats", "scales", "purrr", "broom",
  "stringr", "plyr", "dbplyr", "dplyr", "dtplyr", "tidyr", "tidyselect",
  "timeDate", "timechange", "tzdb", "hms", "readr",
  "lubridate", "prettyunits", "RColorBrewer", "viridis", "rcartocolor",
  "sf", "ggplot2", "patchwork", "pROC", "kableExtra", "mgcv"
)
```

```{r install-requirements, include=FALSE}
for (req in requirements){
  if (!require(req, character.only = TRUE)){
      install.packages(req)
  }
}
```

Note that some of the packages may not be used, but to avoid any error related to dependencies, we are going to keep all of these libraries. Now, the loading of all packages should be immediate.

```{r load-requirements, include=FALSE}
for (req in requirements){
  if (!require(req, character.only = TRUE)){
      library(req, character.only = TRUE)
  }
}
```

```{r remove-vars, include=FALSE}
rm(requirements, req)
```

## Plot Functions

```{r}
check_prop <- function(var, label) {
  tr <- prop.table(table(train[[var]]))
  te <- prop.table(table(test[[var]]))
  tibble(
    level    = names(tr),
    train    = as.numeric(tr),
    test     = as.numeric(te),
    diff_ppt = round((as.numeric(te) - as.numeric(tr)) * 100, 2)
  ) %>%
    mutate(variable = label) %>%
    relocate(variable)
}
```

```{r}
cv_auc <- function(formula, data, folds_vec) {
  k <- max(folds_vec)

  auc_vals <- parallel::mclapply(seq_len(k), function(i) {
    tr  <- data[folds_vec != i, ]
    te  <- data[folds_vec == i, ]

    fit <- tryCatch(
      glm(formula, family = binomial, data = tr),
      error = function(e) NULL
    )
    if (is.null(fit)) return(NA_real_)

    prob <- tryCatch(
      predict(fit, newdata = te, type = "response"),
      error = function(e) rep(NA_real_, nrow(te))
    )

    tryCatch(
      as.numeric(pROC::auc(te$arrested, prob, quiet = TRUE)),
      error = function(e) NA_real_
    )
  }, mc.cores = parallel::detectCores() - 1)

  mean(unlist(auc_vals), na.rm = TRUE)
}

# Shared fold assignment — same folds used for every model
set.seed(2026)
cv_folds <- sample(rep(1:10, length.out = nrow(train)))
```

```{r}
# ── Threshold and confusion-matrix helpers ───────────────────────────────────

best_thresh <- function(roc_obj) {
  pROC::coords(roc_obj, "best", ret = "threshold", best.method = "youden",
               quiet = TRUE)$threshold
}

conf_tib <- function(actual, predicted) {
  tibble(
    Actual    = factor(actual,    levels = c(0, 1)),
    Predicted = factor(predicted, levels = c(0, 1))
  ) %>% count(Actual, Predicted)
}

plot_cm <- function(actual, predicted, title = "Confusion Matrix") {
  conf_tib(actual, predicted) %>%
    ggplot(aes(Predicted, Actual, fill = n)) +
    geom_tile(colour = "white") +
    geom_text(aes(label = scales::comma(n)), size = 5.5, fontface = "bold") +
    scale_fill_gradient(low = "#e8f4f8", high = "#2c7bb6", guide = "none") +
    scale_x_discrete(labels = c("0" = "No arrest", "1" = "Arrest")) +
    scale_y_discrete(labels = c("0" = "No arrest", "1" = "Arrest")) +
    labs(title = title, x = "Predicted", y = "Actual") +
    theme_minimal(base_size = 12) +
    theme(panel.grid = element_blank())
}

class_metrics <- function(actual, predicted, pos = 1) {
  tp <- sum(actual == pos & predicted == pos)
  tn <- sum(actual != pos & predicted != pos)
  fp <- sum(actual != pos & predicted == pos)
  fn <- sum(actual == pos & predicted != pos)
  sens <- tp / (tp + fn)
  spec <- tn / (tn + fp)
  list(
    sensitivity  = sens,
    specificity  = spec,
    balanced_acc = (sens + spec) / 2,
    ppv          = tp / (tp + fp),
    npv          = tn / (tn + fn)
  )
}

# ── Probability-based metrics and evaluation plots ───────────────────────────

# Full metrics tibble (one row per model)
compute_metrics <- function(actual, prob, model_name = "Model") {
  valid  <- !is.na(actual) & !is.na(prob)
  actual <- actual[valid]; prob <- prob[valid]
  roc_obj <- pROC::roc(actual, prob, quiet = TRUE)
  auc_val <- as.numeric(pROC::auc(roc_obj))
  thr     <- best_thresh(roc_obj)
  pred_cl <- as.integer(prob >= thr)

  tp <- sum(actual == 1 & pred_cl == 1); tn <- sum(actual == 0 & pred_cl == 0)
  fp <- sum(actual == 0 & pred_cl == 1); fn <- sum(actual == 1 & pred_cl == 0)
  sens <- if ((tp+fn) == 0) NA_real_ else tp / (tp+fn)
  spec <- if ((tn+fp) == 0) NA_real_ else tn / (tn+fp)
  prec <- if ((tp+fp) == 0) NA_real_ else tp / (tp+fp)
  f1   <- if (is.na(prec) || is.na(sens) || (prec+sens) == 0) NA_real_ else
            2 * prec * sens / (prec + sens)
  mcc_den <- sqrt(as.numeric(tp+fp) * as.numeric(tp+fn) *
                  as.numeric(tn+fp) * as.numeric(tn+fn))
  mcc_val <- if (mcc_den == 0) NA_real_ else (tp*tn - fp*fn) / mcc_den

  eps     <- 1e-15
  logloss <- -mean(actual * log(pmax(prob, eps)) + (1-actual) * log(pmax(1-prob, eps)))
  brier   <- mean((prob - actual)^2)

  # PR-AUC via trapezoid rule over pROC coords
  pr_raw <- pROC::coords(roc_obj, "all",
                          ret = c("sensitivity", "precision"),
                          transpose = FALSE, drop = FALSE)
  names(pr_raw)[names(pr_raw) == "sensitivity"] <- "recall"
  pr_raw <- pr_raw %>%
    filter(!is.nan(precision), !is.na(precision), !is.nan(recall)) %>%
    arrange(recall)
  pr_auc_val <- if (nrow(pr_raw) >= 2)
    sum(diff(pr_raw$recall) *
          (head(pr_raw$precision, -1) + tail(pr_raw$precision, -1)) / 2)
  else NA_real_

  tibble(
    Model       = model_name,
    AUC         = round(auc_val, 4),
    PR_AUC      = round(pr_auc_val, 4),
    Log_Loss    = round(logloss, 4),
    Brier       = round(brier, 4),
    Threshold   = round(thr, 4),
    Sensitivity = round(sens, 3),
    Specificity = round(spec, 3),
    Precision   = round(prec, 3),
    F1          = round(f1, 3),
    MCC         = round(mcc_val, 3),
    Bal_Acc     = round(if (!is.na(sens) && !is.na(spec)) (sens+spec)/2
                        else NA_real_, 3)
  )
}

# ROC curve (single model)
plot_roc_single <- function(actual, prob, title = "ROC Curve") {
  roc_obj <- pROC::roc(actual, prob, quiet = TRUE)
  auc_lbl <- paste0("AUC = ", round(as.numeric(pROC::auc(roc_obj)), 4))
  pROC::ggroc(roc_obj, colour = "#2c7bb6", linewidth = 0.9) +
    geom_abline(slope = 1, intercept = 1, linetype = "dashed", colour = "grey60") +
    annotate("text", x = 0.25, y = 0.06, label = auc_lbl,
             colour = "#2c7bb6", size = 4) +
    labs(x = "Specificity", y = "Sensitivity", title = title) +
    theme_minimal(base_size = 11) + theme(panel.grid.minor = element_blank())
}

# Precision-Recall curve (single model)
plot_pr_single <- function(actual, prob, title = "Precision-Recall Curve") {
  roc_obj <- pROC::roc(actual, prob, quiet = TRUE)
  pr_raw  <- pROC::coords(roc_obj, "all",
                           ret = c("sensitivity", "precision"),
                           transpose = FALSE, drop = FALSE)
  names(pr_raw)[names(pr_raw) == "sensitivity"] <- "recall"
  pr_df <- pr_raw %>%
    filter(!is.nan(precision), !is.na(precision)) %>% arrange(recall)
  pr_auc_lbl <- if (nrow(pr_df) >= 2) {
    v <- sum(diff(pr_df$recall) *
               (head(pr_df$precision,-1) + tail(pr_df$precision,-1)) / 2)
    paste0("PR-AUC = ", round(v, 4))
  } else "PR-AUC = NA"
  baseline <- mean(actual)
  ggplot(pr_df, aes(x = recall, y = precision)) +
    geom_line(colour = "#d73027", linewidth = 0.9) +
    geom_hline(yintercept = baseline, linetype = "dashed", colour = "grey55") +
    annotate("text", x = 0.02, y = baseline + 0.015,
             label = paste0("No-skill = ", round(baseline, 3)),
             colour = "grey40", size = 3.2, hjust = 0) +
    annotate("text", x = 0.55, y = 0.90, label = pr_auc_lbl,
             colour = "#d73027", size = 4) +
    scale_x_continuous(limits = c(0, 1)) + scale_y_continuous(limits = c(0, 1)) +
    labs(x = "Recall", y = "Precision", title = title) +
    theme_minimal(base_size = 11) + theme(panel.grid.minor = element_blank())
}

# Calibration (reliability) plot
plot_calibration_single <- function(actual, prob, title = "Calibration Plot",
                                    n_bins = 10) {
  brks  <- unique(quantile(prob, seq(0, 1, length.out = n_bins + 1), na.rm = TRUE))
  cal_df <- tibble(actual = actual, prob = prob) %>%
    mutate(bin = cut(prob, breaks = brks, include.lowest = TRUE)) %>%
    filter(!is.na(bin)) %>%
    group_by(bin) %>%
    summarise(mean_pred = mean(prob), mean_obs = mean(actual),
              n = n(), .groups = "drop")
  ggplot(cal_df, aes(x = mean_pred, y = mean_obs)) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey55") +
    geom_line(colour = "#2c7bb6", linewidth = 0.8) +
    geom_point(aes(size = n), colour = "#2c7bb6", alpha = 0.85) +
    scale_size_continuous(name = "Count", range = c(2, 9)) +
    scale_x_continuous(labels = scales::percent_format(accuracy = 0.1)) +
    scale_y_continuous(labels = scales::percent_format(accuracy = 0.1)) +
    labs(x = "Mean predicted probability", y = "Observed positive rate",
         title = title,
         caption = "Size ∝ bin count.  Dashed = perfect calibration.") +
    theme_minimal(base_size = 11) +
    theme(legend.position = "bottom", panel.grid.minor = element_blank())
}

# Threshold analysis: sensitivity, specificity, precision, F1 vs threshold
plot_threshold_single <- function(actual, prob, title = "Threshold Analysis") {
  roc_obj  <- pROC::roc(actual, prob, quiet = TRUE)
  thr_data <- pROC::coords(roc_obj, "all",
                            ret = c("threshold","sensitivity","specificity","precision"),
                            transpose = FALSE, drop = FALSE) %>%
    filter(is.finite(threshold), threshold >= 0, threshold <= 1) %>%
    mutate(f1 = ifelse(!is.nan(precision) & (precision + sensitivity) > 0,
                       2 * precision * sensitivity / (precision + sensitivity), 0)) %>%
    pivot_longer(c(sensitivity, specificity, precision, f1),
                 names_to = "Metric", values_to = "Value") %>%
    mutate(Metric = recode(Metric,
      sensitivity = "Sensitivity", specificity = "Specificity",
      precision   = "Precision",   f1          = "F1 Score"))
  thr_opt <- best_thresh(roc_obj)
  ggplot(thr_data, aes(x = threshold, y = Value, colour = Metric)) +
    geom_line(linewidth = 0.75, na.rm = TRUE) +
    geom_vline(xintercept = thr_opt, linetype = "dashed",
               colour = "grey35", linewidth = 0.55) +
    annotate("text", x = thr_opt, y = 0.04,
             label = paste0("Youden\n", round(thr_opt, 3)),
             hjust = -0.08, size = 3, colour = "grey30") +
    scale_colour_viridis_d(option = "turbo", begin = 0.05, end = 0.92) +
    scale_y_continuous(limits = c(0, 1)) +
    labs(x = "Threshold", y = "Metric value", colour = NULL, title = title,
         caption = "Dashed: Youden's J optimal threshold") +
    theme_minimal(base_size = 11) +
    theme(legend.position = "bottom", panel.grid.minor = element_blank())
}

# 2×2 evaluation panel (ROC | PR) / (Calibration | Threshold)
eval_quad_plot <- function(actual, prob, model_name) {
  p1 <- plot_roc_single(actual, prob,   paste(model_name, "— ROC"))
  p2 <- plot_pr_single(actual, prob,    paste(model_name, "— Precision-Recall"))
  p3 <- plot_calibration_single(actual, prob, paste(model_name, "— Calibration"))
  p4 <- plot_threshold_single(actual, prob, paste(model_name, "— Threshold Analysis"))
  (p1 | p2) / (p3 | p4)
}
```

```{r}
# One-row data frame of modal / median values for a set of variable names
build_modal_row <- function(data, vars) {
  row_list <- lapply(vars, function(v) {
    x <- data[[v]]
    if (is.factor(x)) {
      factor(names(sort(table(x), decreasing = TRUE))[1], levels = levels(x))
    } else if (is.numeric(x)) {
      median(x, na.rm = TRUE)
    } else {
      names(sort(table(x), decreasing = TRUE))[1]
    }
  })
  as.data.frame(setNames(row_list, vars))
}
```

---

# Data Loading and Preparation

```{r load-data}
vehstop <- read.csv("data/vehiclestops_4_wheels_final.csv")
```

We convert all categorical predictors to factors and set interpretable reference levels before any modelling step, ensuring consistency between all downstream model fits and cross-validation folds.

```{r factor-prep}
# Ordered weekday levels (ISO order)
wday_levels <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday")

# Ordered time bracket levels
tbkt_levels <- c(
  "Late night (0–3)",
  "Early morning (4–7)",
  "Morning commute (8–11)",
  "Midday (12–15)",
  "Afternoon/evening (16–19)",
  "Night (20–23)"
)

# Ordered age-bracket levels
abkt_levels <- c("Under 18", "18–24", "25–34", "35–44", "45–54", "55–64", "65+", "Unknown")

vehstop <- vehstop %>%
  mutate(
    year         = factor(year),
    month        = factor(month, levels = 1:12),
    weekday      = factor(weekday, levels = wday_levels),
    hour         = factor(hour, levels = 0:23),
    time_bracket = factor(time_bracket, levels = tbkt_levels),
    area         = factor(area),
    boro         = factor(boro),
    sex          = factor(sex),
    ethnicity    = factor(ethnicity),
    age_bracket  = factor(age_bracket, levels = abkt_levels)
  )

# Reference levels chosen for interpretable contrasts
vehstop$sex          <- relevel(vehstop$sex,          ref = "M")
vehstop$ethnicity    <- relevel(vehstop$ethnicity,    ref = "White")
vehstop$weekday      <- relevel(vehstop$weekday,      ref = "Monday")
vehstop$time_bracket <- relevel(vehstop$time_bracket, ref = "Morning commute (8–11)")
vehstop$age_bracket  <- relevel(vehstop$age_bracket,  ref = "25–34")

rm(wday_levels, tbkt_levels, abkt_levels)
```

---

# 0 · Stratified Train / Test Split

We reserve 20% of observations as a hold-out test set. A simple random split preserves the overall `arrested` proportion but can distort the joint distribution of the target with key covariates in the minority class. We therefore stratify **jointly** on `arrested`, `area`, `sex`, `ethnicity`, and `hour`: within every non-empty combination of those five variables we sample 80% for training and 20% for testing.

```{r stratified-split}
set.seed(2026)

# Joint stratum key; drop = TRUE removes empty combinations
vehstop$.strat <- interaction(
  vehstop$arrested, vehstop$area, vehstop$sex, vehstop$ethnicity, vehstop$hour,
  drop = TRUE
)

train_rows <- vehstop %>%
  mutate(.row = row_number()) %>%
  group_by(.strat) %>%
  slice_sample(prop = 0.8) %>%
  pull(.row)

train <- vehstop[ train_rows, ] %>% select(-.strat)
test  <- vehstop[-train_rows, ] %>% select(-.strat)
vehstop <- vehstop %>% select(-.strat)
```

```{r split-verification}
check_prop <- function(var, label) {
  tr <- prop.table(table(train[[var]]))
  te <- prop.table(table(test[[var]]))
  tibble(
    level    = names(tr),
    train    = as.numeric(tr),
    test     = as.numeric(te),
    diff_ppt = round((as.numeric(te) - as.numeric(tr)) * 100, 2)
  ) %>%
    mutate(variable = label) %>%
    relocate(variable)
}

bind_rows(
  check_prop("arrested",  "arrested"),
  check_prop("area",      "area"),
  check_prop("sex",       "sex"),
  check_prop("ethnicity", "ethnicity")
) %>%
  kbl(
    digits    = 4,
    caption   = "Proportion of each level in train vs test (diff in percentage points)",
    col.names = c("Variable", "Level", "Train", "Test", "Δ (ppt)")
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE) %>%
  collapse_rows(columns = 1, valign = "top")
```

```{r split-summary}
cat(sprintf(
  "Training set : %d rows  |  Arrested: %.2f%%\nTest set     : %d rows  |  Arrested: %.2f%%\n",
  nrow(train), 100 * mean(train$arrested),
  nrow(test),  100 * mean(test$arrested)
))
```

The maximum absolute discrepancy between the two sets is on the order of fractions of a percentage point, confirming that the stratification preserved the joint distribution of the target and the key covariates.

---

# i · Full Logistic Regression Model

## Model Specification

We fit a logistic regression model on the training data. All predictors enter as unordered factors, so each level is estimated against the chosen reference category. No interactions are included at this stage—they will be considered in later modelling steps once the main effects are understood.

```{r fit-full-model}
f_base <- arrested ~ year + month + weekday + time_bracket + boro + sex + ethnicity + age_bracket

mod_base <- glm(f_base, family = binomial, data = train)
```

## Coefficient Summary

```{r model-summary}
summary(mod_base)
```

The model has `r length(coef(mod_base))` estimated parameters on `r nrow(train)` training observations (`r round(100 * mean(train$arrested), 1)`% positive). The null deviance is `r round(mod_base$null.deviance, 1)` and the residual deviance is `r round(mod_base$deviance, 1)`, a reduction of `r round(mod_base$null.deviance - mod_base$deviance, 1)` on `r mod_base$df.null - mod_base$df.residual` degrees of freedom.

## Odds Ratios and Significance

We exponentiate the coefficients to obtain odds ratios (OR) and compute 95% profile-likelihood confidence intervals. An OR above 1 indicates increased odds of arrest relative to the reference level; below 1 indicates decreased odds.

```{r odds-ratios-table}
predictors <- c("year", "month", "weekday", "time_bracket", "boro", "sex", "ethnicity", "age_bracket")

or_tbl <- tidy(mod_base, conf.int = FALSE, exponentiate = FALSE) %>%
  filter(term != "(Intercept)") %>%
  mutate(
    conf.low  = exp(estimate - qnorm(0.975) * std.error),
    conf.high = exp(estimate + qnorm(0.975) * std.error),
    estimate  = exp(estimate),
    group = purrr::map_chr(term, function(t) {
      g <- predictors[purrr::map_lgl(predictors, ~ startsWith(t, .x))]
      if (length(g) == 0) "Other" else g[which.max(nchar(g))]
    }),
    level = purrr::map2_chr(term, group, ~ sub(paste0("^", .y), "", .x)),
    sig = case_when(
      p.value < 0.001 ~ "***",
      p.value < 0.01  ~ "**",
      p.value < 0.05  ~ "*",
      p.value < 0.1   ~ ".",
      TRUE            ~ ""
    )
  ) %>%
  select(group, level, estimate, conf.low, conf.high, p.value, sig)

or_tbl %>%
  kbl(
    digits    = 3,
    caption   = "Odds ratios with 95% CI and significance (reference levels: year=2023, month=Jan, weekday=Mon, time_bracket=Morning commute, boro=Bronx, sex=M, ethnicity=White, age_bracket=25-34)",
    col.names = c("Group", "Level vs Reference", "OR", "CI 2.5%", "CI 97.5%", "p-value", "")
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE) %>%
  collapse_rows(columns = 1, valign = "top")
```

## Odds Ratio Plot

The forest plot below displays estimated odds ratios with 95% confidence intervals, faceted by predictor. Points are coloured by significance at the 5% level. The vertical dashed line marks OR = 1 (no effect). Note the logarithmic x-axis.

```{r or-plot, fig.width=11, fig.height=10}
or_plot_data <- or_tbl %>%
  mutate(
    is_sig = p.value < 0.05,
    level  = stringr::str_wrap(level, width = 28),
    group  = factor(group, levels = c(
      "year", "month", "weekday", "time_bracket",
      "boro", "sex", "ethnicity", "age_bracket"
    ))
  )

ggplot(or_plot_data, aes(x = estimate, y = reorder(level, estimate), colour = is_sig)) +
  geom_vline(xintercept = 1, linetype = "dashed", colour = "grey55", linewidth = 0.5) +
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high), height = 0.25, linewidth = 0.45) +
  geom_point(size = 2) +
  scale_x_log10(labels = scales::number_format(accuracy = 0.01)) +
  scale_colour_manual(
    values = c("TRUE" = "#2c7bb6", "FALSE" = "#d73027"),
    labels = c("TRUE" = "p < 0.05", "FALSE" = "p ≥ 0.05")
  ) +
  facet_wrap(~ group, scales = "free_y", ncol = 2) +
  labs(
    x       = "Odds Ratio (log scale)",
    y       = NULL,
    colour  = "Significance",
    title   = "Full Logistic Model — Odds Ratios with 95% CI",
    caption = paste0(
      "Reference levels: year = 2023, month = 1, weekday = Monday,\n",
      "time_bracket = Morning commute, boro = Bronx,\n",
      "sex = M, ethnicity = White, age_bracket = 25–34"
    )
  ) +
  theme_minimal(base_size = 11) +
  theme(
    strip.text       = element_text(face = "bold"),
    legend.position  = "bottom",
    panel.grid.minor = element_blank()
  )
```

## Key Findings

-   **Year**: Arrest probability shifted across years; interpret coefficients relative to 2023 to detect a temporal trend in enforcement intensity.
-   **Month**: Seasonal variation is present. Summer months (April–October, the "hot season" identified in the EDA) tend toward higher enforcement activity.
-   **Weekday**: There is variation by day of the week, with some days showing significantly different arrest odds relative to Monday. Weekday effects partially overlap with the time-of-day signal.
-   **Time bracket**: Late-night and early-morning stops carry higher arrest odds than the morning-commute reference period, consistent with the EDA finding that enforcement character changes after hours.
-   **Borough**: The Bronx (reference) and Staten Island show markedly higher arrest odds than the Manhattan sub-boroughs. The sub-borough granularity of `boro` is what we evaluate in comparison (b).
-   **Sex**: Female stops are estimated at lower arrest odds than male stops. The "Unknown" sex coefficient should be interpreted cautiously given the small cell count.
-   **Ethnicity**: Stops of Black and Hispanic motorists are associated with higher arrest odds relative to White motorists, consistent with the EDA bivariate analysis. These are observational associations and should not be read as causal.
-   **Age bracket**: Younger riders (especially Under 18) show elevated arrest odds; odds decrease with age beyond the 25–34 reference. The monotonicity assumption is tested in comparison (c).

---

# ii · Predictor-Form Comparison

## Strategy

The base model uses `time_bracket` (6-level factor, 5 df), `boro` (8-level factor, 7 df), and `age_bracket` (7-level factor, 6 df). We compare each against a finer or coarser alternative:

| Comparison | Base form | Alternative | Nesting | Tests |
|---|---|---|---|---|
| (a) | `time_bracket` (5 df) | `hour` as factor (23 df) | Base nested in alternative | AIC, BIC, LRT |
| (b) | `boro` (7 df) | `area` (4 df) | Alternative nested in base | AIC, BIC, LRT |
| (c) | `age_bracket` (6 df) | `age` numeric (1 df) | Not nested | AIC, BIC, CV AUC |

For (a) and (b) we use the **likelihood ratio test**: `time_bracket` constrains all hours within a bin to share one coefficient (18 extra restrictions); `area` constrains boroughs within a geographic group to share one coefficient (3 extra restrictions). For (c) the LRT is not applicable—a linear numeric term and a step-function factor are not nested in either direction.

**Cross-validated AUC** uses the same 10 stratified folds on the training set for all models.

```{r cv-auc-function}
cv_auc <- function(formula, data, folds_vec) {
  k <- max(folds_vec)

  auc_vals <- parallel::mclapply(seq_len(k), function(i) {
    tr  <- data[folds_vec != i, ]
    te  <- data[folds_vec == i, ]

    fit <- tryCatch(
      glm(formula, family = binomial, data = tr),
      error = function(e) NULL
    )
    if (is.null(fit)) return(NA_real_)

    prob <- tryCatch(
      predict(fit, newdata = te, type = "response"),
      error = function(e) rep(NA_real_, nrow(te))
    )

    tryCatch(
      as.numeric(pROC::auc(te$arrested, prob, quiet = TRUE)),
      error = function(e) NA_real_
    )
  }, mc.cores = parallel::detectCores() - 1)

  mean(unlist(auc_vals), na.rm = TRUE)
}

# Shared fold assignment — same folds used for every model
set.seed(2026)
cv_folds <- sample(rep(1:10, length.out = nrow(train)))
```

---

## (a) `time_bracket` vs `hour`

`hour` encodes the exact hour of stop (0–23) as a 24-level factor (23 df). `time_bracket` groups those hours into 6 bins (5 df). The `time_bracket` model is **nested** inside the `hour` model: it imposes the constraint that all hours within a bin have equal log-odds, a restriction with 23 − 5 = 18 degrees of freedom.

```{r fit-model-hour, cache=TRUE}
f_hour   <- arrested ~ year + month + weekday + hour + boro + sex + ethnicity + age_bracket
mod_hour <- glm(f_hour, family = binomial, data = train)
```

### AIC and BIC

```{r compare-a-aic}
aic_a <- AIC(mod_base, mod_hour)
bic_a <- BIC(mod_base, mod_hour)

tibble(
  Model     = c("Base (time_bracket, 5 df)", "Alternative (hour, 23 df)"),
  df        = aic_a$df,
  AIC       = round(aic_a$AIC, 1),
  BIC       = round(bic_a$BIC, 1),
  delta_AIC = round(aic_a$AIC - min(aic_a$AIC), 1),
  delta_BIC = round(bic_a$BIC - min(bic_a$BIC), 1)
) %>%
  kbl(
    caption   = "Comparison (a): time_bracket vs hour",
    col.names = c("Model", "# Param", "AIC", "BIC", "Δ AIC", "Δ BIC")
  ) %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE)
```

### Likelihood Ratio Test

```{r lrt-a}
lrt_a <- anova(mod_base, mod_hour, test = "Chisq")
lrt_a
```

```{r lrt-a-interp}
chi_a <- lrt_a[["Deviance"]][2]
df_a  <- lrt_a[["Df"]][2]
p_a   <- lrt_a[["Pr(>Chi)"]][2]
cat(sprintf("LRT (a): χ²(%d) = %.2f,  p = %.4g\n", df_a, chi_a, p_a))
```

This likelihood ratio test (LRT) compares two nested logistic regression models:

* **Model 1:** Uses **time_bracket** (a categorical grouping of hours).
* **Model 2:** Uses **hour** (individual hours) instead of **time_bracket**.

The hypotheses are:

* **H₀:** `time_bracket` explains the data as well as `hour` (the simpler model is sufficient).
* **H₁:** Using `hour` provides a significantly better fit.

#### Results

* Likelihood ratio statistic: **χ²(18) = 1506.36**
* Degrees of freedom: **18**
* **p = 2.05 × 10⁻³⁰⁹** (reported by R as `< 2.2e-16` in the ANOVA table because of printing limits)

#### Interpretation

The LRT is highly significant (**χ²(18) = 1506.36, p < .001**). Therefore, you reject the null hypothesis and conclude that replacing **time_bracket** with the more detailed **hour** variable significantly improves model fit.

The residual deviance decreases from:

* **448,316** (Model 1)
* to **446,809** (Model 2)

a reduction of **1,506.36**, which is far larger than would be expected by chance.

#### Practical interpretation

Statistically, **hour** captures meaningful variation in arrest probability that is lost when hours are grouped into broader time brackets. This suggests that the relationship between time of day and arrests is more nuanced than the categories in `time_bracket` can represent.

However, because your sample size is extremely large (over **1.7 million observations**), even relatively small improvements in fit can produce extremely small p-values. Therefore, it is also useful to assess whether the improvement is practically meaningful by comparing measures such as:

* **AIC** (lower is better),
* **BIC** (if model parsimony is important),
* pseudo-(R^2),
* or predictive performance (e.g., cross-validated AUC or log loss).

#### Example write-up (APA style)

> A likelihood ratio test was conducted to compare a logistic regression model including **time_bracket** with a model including **hour**. The model with **hour** provided a significantly better fit to the data, **χ²**(18) = **1506.36**, *p* < .001, indicating that modeling the specific hour of the day explains arrest outcomes better than using broader time-of-day categories.

This is an appropriate conclusion if your goal is to justify using `hour` instead of `time_bracket` in the final model.


### CV AUC

```{r cv-auc-a, cache=TRUE}
auc_base_a <- cv_auc(f_base, train, cv_folds)
auc_hour_a <- cv_auc(f_hour, train, cv_folds)

tibble(
  Model   = c("Base (time_bracket)", "Alternative (hour)"),
  CV_AUC  = round(c(auc_base_a, auc_hour_a), 4)
) %>%
  kbl(caption = "10-fold CV AUC on training set — comparison (a)") %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE)
```

**Decision (a):** A significant LRT with a large chi-square statistic indicates that within-bracket hour-level variation is real—the bins conceal genuine non-linearity in the time-of-day effect. If the AIC of the `hour` model is meaningfully lower but BIC is higher (penalising the 18 extra parameters), and the CV AUC difference is small (< 0.002), the `time_bracket` parameterisation is preferred on parsimony grounds. If instead the CV AUC gain is substantial, `hour` better captures the enforcement cycle.

---

## (b) `boro` vs `area`

`area` (5 levels, 4 df) merges the sub-borough splits: both Brooklyn sub-boroughs become "Brooklyn", both Manhattan sub-boroughs become "Manhattan", both Queens sub-boroughs become "Queens". The `area` model is **nested** inside the `boro` model: it constrains three pairs of sub-borough coefficients to be equal, a restriction with 7 − 4 = 3 degrees of freedom.

```{r fit-model-area, cache=TRUE}
f_area   <- arrested ~ year + month + weekday + time_bracket + area + sex + ethnicity + age_bracket
mod_area <- glm(f_area, family = binomial, data = train)
```

### AIC and BIC

```{r compare-b-aic}
aic_b <- AIC(mod_area, mod_base)
bic_b <- BIC(mod_area, mod_base)

tibble(
  Model     = c("Alternative (area, 4 df)", "Base (boro, 7 df)"),
  df        = aic_b$df,
  AIC       = round(aic_b$AIC, 1),
  BIC       = round(bic_b$BIC, 1),
  delta_AIC = round(aic_b$AIC - min(aic_b$AIC), 1),
  delta_BIC = round(bic_b$BIC - min(bic_b$BIC), 1)
) %>%
  kbl(
    caption   = "Comparison (b): area vs boro",
    col.names = c("Model", "# Param", "AIC", "BIC", "Δ AIC", "Δ BIC")
  ) %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE)
```

### Likelihood Ratio Test

```{r lrt-b}
lrt_b <- anova(mod_area, mod_base, test = "Chisq")
lrt_b
```

```{r lrt-b-interp}
chi_b <- lrt_b[["Deviance"]][2]
df_b  <- lrt_b[["Df"]][2]
p_b   <- lrt_b[["Pr(>Chi)"]][2]
cat(sprintf("LRT (b): χ²(%d) = %.2f,  p = %.4g\n", df_b, chi_b, p_b))
```

### CV AUC

```{r cv-auc-b, cache=TRUE}
auc_area_b <- cv_auc(f_area, train, cv_folds)
auc_base_b <- cv_auc(f_base, train, cv_folds)

tibble(
  Model   = c("Alternative (area)", "Base (boro)"),
  CV_AUC  = round(c(auc_area_b, auc_base_b), 4)
) %>%
  kbl(caption = "10-fold CV AUC on training set — comparison (b)") %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE)
```

**Decision (b):** A significant LRT (on only 3 df) means that at least one pair of sub-boroughs has genuinely different arrest odds and the aggregation to `area` loses predictive information. Conversely, a non-significant result—and the EDA heatmaps suggested within-borough heterogeneity is moderate—supports retaining `area` for parsimony and reduced variance in the very sparse Staten Island stratum.

---

## (c) `age_bracket` vs numeric `age`

Numeric `age` enters as a single continuous term (1 df), assuming log-odds change linearly with age. `age_bracket` uses a step function over seven brackets (6 df). The EDA showed that arrest rates peak at younger ages and then decline, but not always monotonically. These two parameterisations are **not nested**: a linear function cannot reproduce a step function and vice versa. No LRT is applicable; we rely on AIC, BIC, and CV AUC.

```{r fit-model-age, cache=TRUE}
f_age   <- arrested ~ year + month + weekday + time_bracket + boro + sex + ethnicity + age
mod_age <- glm(f_age, family = binomial, data = train)
```

### AIC and BIC

```{r compare-c-aic}
aic_c <- AIC(mod_base, mod_age)
bic_c <- BIC(mod_base, mod_age)

tibble(
  Model     = c("Base (age_bracket, 6 df)", "Alternative (age numeric, 1 df)"),
  df        = aic_c$df,
  AIC       = round(aic_c$AIC, 1),
  BIC       = round(bic_c$BIC, 1),
  delta_AIC = round(aic_c$AIC - min(aic_c$AIC), 1),
  delta_BIC = round(bic_c$BIC - min(bic_c$BIC), 1)
) %>%
  kbl(
    caption   = "Comparison (c): age_bracket vs numeric age",
    col.names = c("Model", "# Param", "AIC", "BIC", "Δ AIC", "Δ BIC")
  ) %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE)
```

### CV AUC

```{r cv-auc-c, cache=TRUE}
auc_base_c <- cv_auc(f_base, train, cv_folds)
auc_age_c  <- cv_auc(f_age,  train, cv_folds)

tibble(
  Model   = c("Base (age_bracket)", "Alternative (age numeric)"),
  CV_AUC  = round(c(auc_base_c, auc_age_c), 4)
) %>%
  kbl(caption = "10-fold CV AUC on training set — comparison (c)") %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE)
```

**Decision (c):** If `age_bracket` achieves lower AIC and higher CV AUC, the non-linear bracket structure captures genuine non-monotonicity in the age effect that a single linear slope misses. If `age` (1 df) achieves lower BIC with a similar CV AUC, the linearity assumption is adequate and the simpler parameterisation is preferred. Given the EDA evidence of a U-shaped age–arrest relationship (elevated Under 18 and 65+ rates), we expect the bracket model to perform better.

---

## Full Comparison Summary

```{r summary-table, cache=TRUE}
# Collect all AUC values (shared folds)
auc_summary <- c(auc_base_a, auc_hour_a, auc_area_b, auc_age_c)

models_list <- list(mod_base, mod_hour, mod_area, mod_age)

summary_tbl <- tibble(
  Model      = c(
    "Base  (time_bracket + boro + age_bracket)",
    "(a)   hour + boro + age_bracket",
    "(b)   time_bracket + area + age_bracket",
    "(c)   time_bracket + boro + age (numeric)"
  ),
  Parameters = sapply(models_list, function(m) length(coef(m))),
  Deviance   = sapply(models_list, function(m) round(deviance(m), 1)),
  AIC        = sapply(models_list, function(m) round(AIC(m), 1)),
  BIC        = sapply(models_list, function(m) round(BIC(m), 1)),
  CV_AUC     = round(auc_summary, 4)
) %>%
  mutate(
    delta_AIC = round(AIC - min(AIC), 1),
    delta_BIC = round(BIC - min(BIC), 1)
  )

summary_tbl %>%
  kbl(
    caption   = "All model alternatives compared — lower AIC/BIC and higher CV AUC preferred",
    col.names = c("Model", "# Param", "Deviance", "AIC", "BIC", "CV AUC", "Δ AIC", "Δ BIC")
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = TRUE) %>%
  row_spec(which.min(summary_tbl$AIC), bold = TRUE, color = "white", background = "#2c7bb6")
```

```{r lrt-summary}
tibble(
  Comparison = c("(a) time_bracket → hour (18 df freed)", "(b) boro → area (3 df constrained)"),
  `χ²` = round(c(chi_a, chi_b), 2),
  df      = c(df_a, df_b),
  p_value = signif(c(p_a, p_b), 3)
) %>%
  kbl(caption = "Likelihood ratio tests for nested comparisons") %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE)
```

### AIC / BIC profile

```{r aic-bic-plot, fig.width=9, fig.height=4}
summary_tbl %>%
  select(Model, AIC, BIC) %>%
  pivot_longer(c(AIC, BIC), names_to = "Criterion", values_to = "Value") %>%
  mutate(Model = stringr::str_wrap(Model, 35)) %>%
  ggplot(aes(x = reorder(Model, Value), y = Value, fill = Criterion)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.6) +
  coord_flip() +
  scale_fill_viridis_d(option = "mako", begin = 0.35, end = 0.75) +
  labs(
    x     = NULL,
    y     = "Information criterion (lower is better)",
    title = "AIC and BIC across predictor-form alternatives",
    fill  = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom", panel.grid.minor = element_blank())
```

### CV AUC profile

```{r auc-plot, fig.width=8, fig.height=3.5}
summary_tbl %>%
  mutate(Model = stringr::str_wrap(Model, 35)) %>%
  ggplot(aes(x = reorder(Model, CV_AUC), y = CV_AUC)) +
  geom_col(fill = "#2c7bb6", width = 0.6) +
  geom_text(aes(label = sprintf("%.4f", CV_AUC)), hjust = -0.1, size = 3.5) +
  coord_flip(ylim = c(
    min(summary_tbl$CV_AUC) - 0.005,
    max(summary_tbl$CV_AUC) + 0.012
  )) +
  labs(
    x     = NULL,
    y     = "10-fold CV AUC (higher is better)",
    title = "Cross-validated AUC across predictor-form alternatives"
  ) +
  theme_minimal(base_size = 11) +
  theme(panel.grid.minor = element_blank())
```

---

## Selected Predictor Forms

Based on the three comparisons, the preferred predictor forms feeding into all subsequent models are:

| Axis | Chosen form | Rationale |
|---|---|---|
| Time of day | `time_bracket` or `hour` | Prefer `time_bracket` unless LRT (a) is significant **and** CV AUC gain > 0.002 |
| Geography | `boro` or `area` | Prefer `area` if LRT (b) is non-significant; prefer `boro` otherwise |
| Age | `age_bracket` or `age` | Prefer whichever achieves lower AIC; expect bracket to win given the EDA |

The winning form is determined programmatically from the AIC values already computed above and stored in `f_final`, which is used consistently by all subsequent models.

```{r final-formula}
use_time <- if (AIC(mod_hour) <= AIC(mod_base)) "hour"         else "time_bracket"
use_geo  <- if (AIC(mod_base) <= AIC(mod_area)) "boro"         else "area"
use_age  <- if (AIC(mod_base) <= AIC(mod_age))  "age_bracket"  else "age"

cat("Time predictor :", use_time, "\n")
cat("Geo predictor  :", use_geo,  "\n")
cat("Age predictor  :", use_age,  "\n")

f_final <- reformulate(
  c("year", "month", "weekday", use_time, use_geo, "sex", "ethnicity", use_age),
  response = "arrested"
)
```

```{r eval-helpers}
# ── Threshold and confusion-matrix helpers ───────────────────────────────────

best_thresh <- function(roc_obj) {
  pROC::coords(roc_obj, "best", ret = "threshold", best.method = "youden",
               quiet = TRUE)$threshold
}

conf_tib <- function(actual, predicted) {
  tibble(
    Actual    = factor(actual,    levels = c(0, 1)),
    Predicted = factor(predicted, levels = c(0, 1))
  ) %>% count(Actual, Predicted)
}

plot_cm <- function(actual, predicted, title = "Confusion Matrix") {
  conf_tib(actual, predicted) %>%
    ggplot(aes(Predicted, Actual, fill = n)) +
    geom_tile(colour = "white") +
    geom_text(aes(label = scales::comma(n)), size = 5.5, fontface = "bold") +
    scale_fill_gradient(low = "#e8f4f8", high = "#2c7bb6", guide = "none") +
    scale_x_discrete(labels = c("0" = "No arrest", "1" = "Arrest")) +
    scale_y_discrete(labels = c("0" = "No arrest", "1" = "Arrest")) +
    labs(title = title, x = "Predicted", y = "Actual") +
    theme_minimal(base_size = 12) +
    theme(panel.grid = element_blank())
}

class_metrics <- function(actual, predicted, pos = 1) {
  tp <- sum(actual == pos & predicted == pos)
  tn <- sum(actual != pos & predicted != pos)
  fp <- sum(actual != pos & predicted == pos)
  fn <- sum(actual == pos & predicted != pos)
  sens <- tp / (tp + fn)
  spec <- tn / (tn + fp)
  list(
    sensitivity  = sens,
    specificity  = spec,
    balanced_acc = (sens + spec) / 2,
    ppv          = tp / (tp + fp),
    npv          = tn / (tn + fn)
  )
}

# ── Probability-based metrics and evaluation plots ───────────────────────────

# Full metrics tibble (one row per model)
compute_metrics <- function(actual, prob, model_name = "Model") {
  valid  <- !is.na(actual) & !is.na(prob)
  actual <- actual[valid]; prob <- prob[valid]
  roc_obj <- pROC::roc(actual, prob, quiet = TRUE)
  auc_val <- as.numeric(pROC::auc(roc_obj))
  thr     <- best_thresh(roc_obj)
  pred_cl <- as.integer(prob >= thr)

  tp <- sum(actual == 1 & pred_cl == 1); tn <- sum(actual == 0 & pred_cl == 0)
  fp <- sum(actual == 0 & pred_cl == 1); fn <- sum(actual == 1 & pred_cl == 0)
  sens <- if ((tp+fn) == 0) NA_real_ else tp / (tp+fn)
  spec <- if ((tn+fp) == 0) NA_real_ else tn / (tn+fp)
  prec <- if ((tp+fp) == 0) NA_real_ else tp / (tp+fp)
  f1   <- if (is.na(prec) || is.na(sens) || (prec+sens) == 0) NA_real_ else
            2 * prec * sens / (prec + sens)
  mcc_den <- sqrt(as.numeric(tp+fp) * as.numeric(tp+fn) *
                  as.numeric(tn+fp) * as.numeric(tn+fn))
  mcc_val <- if (mcc_den == 0) NA_real_ else (tp*tn - fp*fn) / mcc_den

  eps     <- 1e-15
  logloss <- -mean(actual * log(pmax(prob, eps)) + (1-actual) * log(pmax(1-prob, eps)))
  brier   <- mean((prob - actual)^2)

  # PR-AUC via trapezoid rule over pROC coords
  pr_raw <- pROC::coords(roc_obj, "all",
                          ret = c("sensitivity", "precision"),
                          transpose = FALSE, drop = FALSE)
  names(pr_raw)[names(pr_raw) == "sensitivity"] <- "recall"
  pr_raw <- pr_raw %>%
    filter(!is.nan(precision), !is.na(precision), !is.nan(recall)) %>%
    arrange(recall)
  pr_auc_val <- if (nrow(pr_raw) >= 2)
    sum(diff(pr_raw$recall) *
          (head(pr_raw$precision, -1) + tail(pr_raw$precision, -1)) / 2)
  else NA_real_

  tibble(
    Model       = model_name,
    AUC         = round(auc_val, 4),
    PR_AUC      = round(pr_auc_val, 4),
    Log_Loss    = round(logloss, 4),
    Brier       = round(brier, 4),
    Threshold   = round(thr, 4),
    Sensitivity = round(sens, 3),
    Specificity = round(spec, 3),
    Precision   = round(prec, 3),
    F1          = round(f1, 3),
    MCC         = round(mcc_val, 3),
    Bal_Acc     = round(if (!is.na(sens) && !is.na(spec)) (sens+spec)/2
                        else NA_real_, 3)
  )
}

# ROC curve (single model)
plot_roc_single <- function(actual, prob, title = "ROC Curve") {
  roc_obj <- pROC::roc(actual, prob, quiet = TRUE)
  auc_lbl <- paste0("AUC = ", round(as.numeric(pROC::auc(roc_obj)), 4))
  pROC::ggroc(roc_obj, colour = "#2c7bb6", linewidth = 0.9) +
    geom_abline(slope = 1, intercept = 1, linetype = "dashed", colour = "grey60") +
    annotate("text", x = 0.25, y = 0.06, label = auc_lbl,
             colour = "#2c7bb6", size = 4) +
    labs(x = "Specificity", y = "Sensitivity", title = title) +
    theme_minimal(base_size = 11) + theme(panel.grid.minor = element_blank())
}

# Precision-Recall curve (single model)
plot_pr_single <- function(actual, prob, title = "Precision-Recall Curve") {
  roc_obj <- pROC::roc(actual, prob, quiet = TRUE)
  pr_raw  <- pROC::coords(roc_obj, "all",
                           ret = c("sensitivity", "precision"),
                           transpose = FALSE, drop = FALSE)
  names(pr_raw)[names(pr_raw) == "sensitivity"] <- "recall"
  pr_df <- pr_raw %>%
    filter(!is.nan(precision), !is.na(precision)) %>% arrange(recall)
  pr_auc_lbl <- if (nrow(pr_df) >= 2) {
    v <- sum(diff(pr_df$recall) *
               (head(pr_df$precision,-1) + tail(pr_df$precision,-1)) / 2)
    paste0("PR-AUC = ", round(v, 4))
  } else "PR-AUC = NA"
  baseline <- mean(actual)
  ggplot(pr_df, aes(x = recall, y = precision)) +
    geom_line(colour = "#d73027", linewidth = 0.9) +
    geom_hline(yintercept = baseline, linetype = "dashed", colour = "grey55") +
    annotate("text", x = 0.02, y = baseline + 0.015,
             label = paste0("No-skill = ", round(baseline, 3)),
             colour = "grey40", size = 3.2, hjust = 0) +
    annotate("text", x = 0.55, y = 0.90, label = pr_auc_lbl,
             colour = "#d73027", size = 4) +
    scale_x_continuous(limits = c(0, 1)) + scale_y_continuous(limits = c(0, 1)) +
    labs(x = "Recall", y = "Precision", title = title) +
    theme_minimal(base_size = 11) + theme(panel.grid.minor = element_blank())
}

# Calibration (reliability) plot
plot_calibration_single <- function(actual, prob, title = "Calibration Plot",
                                    n_bins = 10) {
  brks  <- unique(quantile(prob, seq(0, 1, length.out = n_bins + 1), na.rm = TRUE))
  cal_df <- tibble(actual = actual, prob = prob) %>%
    mutate(bin = cut(prob, breaks = brks, include.lowest = TRUE)) %>%
    filter(!is.na(bin)) %>%
    group_by(bin) %>%
    summarise(mean_pred = mean(prob), mean_obs = mean(actual),
              n = n(), .groups = "drop")
  ggplot(cal_df, aes(x = mean_pred, y = mean_obs)) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey55") +
    geom_line(colour = "#2c7bb6", linewidth = 0.8) +
    geom_point(aes(size = n), colour = "#2c7bb6", alpha = 0.85) +
    scale_size_continuous(name = "Count", range = c(2, 9)) +
    scale_x_continuous(labels = scales::percent_format(accuracy = 0.1)) +
    scale_y_continuous(labels = scales::percent_format(accuracy = 0.1)) +
    labs(x = "Mean predicted probability", y = "Observed positive rate",
         title = title,
         caption = "Size ∝ bin count.  Dashed = perfect calibration.") +
    theme_minimal(base_size = 11) +
    theme(legend.position = "bottom", panel.grid.minor = element_blank())
}

# Threshold analysis: sensitivity, specificity, precision, F1 vs threshold
plot_threshold_single <- function(actual, prob, title = "Threshold Analysis") {
  roc_obj  <- pROC::roc(actual, prob, quiet = TRUE)
  thr_data <- pROC::coords(roc_obj, "all",
                            ret = c("threshold","sensitivity","specificity","precision"),
                            transpose = FALSE, drop = FALSE) %>%
    filter(is.finite(threshold), threshold >= 0, threshold <= 1) %>%
    mutate(f1 = ifelse(!is.nan(precision) & (precision + sensitivity) > 0,
                       2 * precision * sensitivity / (precision + sensitivity), 0)) %>%
    pivot_longer(c(sensitivity, specificity, precision, f1),
                 names_to = "Metric", values_to = "Value") %>%
    mutate(Metric = recode(Metric,
      sensitivity = "Sensitivity", specificity = "Specificity",
      precision   = "Precision",   f1          = "F1 Score"))
  thr_opt <- best_thresh(roc_obj)
  ggplot(thr_data, aes(x = threshold, y = Value, colour = Metric)) +
    geom_line(linewidth = 0.75, na.rm = TRUE) +
    geom_vline(xintercept = thr_opt, linetype = "dashed",
               colour = "grey35", linewidth = 0.55) +
    annotate("text", x = thr_opt, y = 0.04,
             label = paste0("Youden\n", round(thr_opt, 3)),
             hjust = -0.08, size = 3, colour = "grey30") +
    scale_colour_viridis_d(option = "turbo", begin = 0.05, end = 0.92) +
    scale_y_continuous(limits = c(0, 1)) +
    labs(x = "Threshold", y = "Metric value", colour = NULL, title = title,
         caption = "Dashed: Youden's J optimal threshold") +
    theme_minimal(base_size = 11) +
    theme(legend.position = "bottom", panel.grid.minor = element_blank())
}

# 2×2 evaluation panel (ROC | PR) / (Calibration | Threshold)
eval_quad_plot <- function(actual, prob, model_name) {
  p1 <- plot_roc_single(actual, prob,   paste(model_name, "— ROC"))
  p2 <- plot_pr_single(actual, prob,    paste(model_name, "— Precision-Recall"))
  p3 <- plot_calibration_single(actual, prob, paste(model_name, "— Calibration"))
  p4 <- plot_threshold_single(actual, prob, paste(model_name, "— Threshold Analysis"))
  (p1 | p2) / (p3 | p4)
}
```

---

# ii-b · Best Logistic Regression — Evaluation

We fit the best logistic model (with the predictor forms selected in section ii) on the training set and evaluate it on the hold-out test set. All subsequent models use the same split and the same `f_final` formula.

```{r logit-best-fit}
mod_logit_best <- glm(f_final, family = binomial, data = train)
prob_logit     <- predict(mod_logit_best, newdata = test, type = "response")
roc_logit      <- pROC::roc(test$arrested, prob_logit, quiet = TRUE)
auc_logit      <- as.numeric(pROC::auc(roc_logit))
thr_logit      <- best_thresh(roc_logit)
class_logit    <- as.integer(prob_logit >= thr_logit)
met_logit      <- class_metrics(test$arrested, class_logit)
```

```{r logit-metrics-table}
compute_metrics(test$arrested, prob_logit, "Logistic Regression") %>%
  kbl(
    caption   = "Logistic Regression — full evaluation metrics (test set, Youden threshold)",
    col.names = c("Model","AUC","PR-AUC","Log-Loss","Brier",
                  "Threshold","Sensitivity","Specificity","Precision","F1","MCC","Bal. Acc.")
  ) %>%
  kable_styling(bootstrap_options = c("striped","condensed"), full_width = TRUE)
```

```{r logit-cm-plot, fig.width=4, fig.height=3.5}
plot_cm(test$arrested, class_logit,
        "Logistic Regression — Confusion Matrix (Youden threshold)")
```

```{r logit-eval-plots, fig.width=10, fig.height=8}
eval_quad_plot(test$arrested, prob_logit, "Logistic Regression")
```

### Notes

-   **ROC curve**: Measures overall discrimination regardless of threshold. AUC = P(score of random positive > score of random negative).
-   **PR curve**: More informative than ROC under class imbalance. A model with no skill scores PR-AUC ≈ prevalence (≈ 5.5%). PR-AUC above that baseline indicates genuine discrimination.
-   **Calibration**: A well-calibrated model has predicted probabilities that match observed proportions. Curves bowing above the diagonal indicate under-confidence; below the diagonal indicates over-confidence.
-   **Threshold analysis**: Shows how sensitivity, specificity, precision, and F1 change as the decision threshold varies. The Youden threshold maximises sensitivity + specificity − 1.

---

# ii-c · Interactions and Marginal Effects

## Motivation

The additive logistic model fitted in section ii-b (`mod_logit_best`) assumes that every predictor's contribution to the log-odds is independent of all other predictors. The EDA and the main-effects model suggest this may be too restrictive:

-   **`ethnicity:sex`** — Both ethnicity and sex have significant main effects on arrest odds. If the male–female gap in arrests is systematically wider for some ethnic groups than others, the additive model misrepresents the data; the interaction term captures that differential.
-   **`ethnicity:` + geographic variable** — Ethnic composition and police enforcement intensity co-vary by neighbourhood. If the ethnic disparity in arrest odds is structurally larger in some areas than others, the additive model distributes the geographic and demographic effects independently when in reality they overlap.

We fit one model for each interaction, compare each against `mod_logit_best` via likelihood ratio test and BIC, then visualise the interaction on the probability scale through marginal effect plots.

## Model Specifications

```{r inter-formulas}
f_inter1 <- update(f_final, ~ . + ethnicity:sex)
f_inter2 <- update(f_final, as.formula(paste0("~ . + ethnicity:", use_geo)))

cat("f_inter1 :", deparse(f_inter1), "\n\n")
cat("f_inter2 :", deparse(f_inter2), "\n")
```

```{r inter-fit}
mod_logit_inter1 <- glm(f_inter1, family = binomial, data = train)
mod_logit_inter2 <- glm(f_inter2, family = binomial, data = train)
```

---

## Model 1 — `ethnicity:sex`

### Interaction Odds Ratios

The interaction OR for the cell `ethnicity_X:sexF` answers: *relative to White males, is the female odds ratio for ethnic group X different from the female odds ratio for White motorists?* An OR > 1 means the sex gap is wider for that group; OR < 1 means it is narrower.

```{r inter1-or-table}
or_inter1 <- tidy(mod_logit_inter1, conf.int = FALSE, exponentiate = FALSE) %>%
  filter(grepl(":", term)) %>%
  mutate(
    conf.low  = exp(estimate - qnorm(0.975) * std.error),
    conf.high = exp(estimate + qnorm(0.975) * std.error),
    estimate  = exp(estimate),
    sig = case_when(
      p.value < 0.001 ~ "***",
      p.value < 0.01  ~ "**",
      p.value < 0.05  ~ "*",
      p.value < 0.1   ~ ".",
      TRUE            ~ ""
    )
  ) %>%
  select(term, estimate, conf.low, conf.high, p.value, sig)

or_inter1 %>%
  kbl(
    digits    = 3,
    caption   = "Model 1 (ethnicity:sex) — Interaction odds ratios with 95% CI",
    col.names = c("Term", "OR", "CI 2.5%", "CI 97.5%", "p-value", "")
  ) %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE)
```

### LRT and BIC vs `mod_logit_best`

```{r inter1-lrt}
lrt_inter1 <- anova(mod_logit_best, mod_logit_inter1, test = "Chisq")
chi_i1 <- lrt_inter1[["Deviance"]][2]
df_i1  <- lrt_inter1[["Df"]][2]
p_i1   <- lrt_inter1[["Pr(>Chi)"]][2]
bic_i1_base  <- BIC(mod_logit_best)
bic_i1_inter <- BIC(mod_logit_inter1)

cat(sprintf(
  "LRT (ethnicity:sex): χ²(%d) = %.2f,  p = %.4g\n",
  df_i1, chi_i1, p_i1
))
cat(sprintf(
  "BIC — mod_logit_best: %.1f  |  mod_logit_inter1: %.1f  |  ΔBIC = %.1f\n",
  bic_i1_base, bic_i1_inter, bic_i1_inter - bic_i1_base
))
```

A significant LRT ($p < 0.05$) indicates the sex effect on arrest probability is not uniform across ethnic groups. ΔBIC > 0 means the extra parameters are not justified on parsimony grounds even if the LRT rejects; ΔBIC < 0 means the interaction is both statistically and informationally beneficial.

---

## Model 2 — `ethnicity:` + geographic variable

### Interaction Odds Ratios

Each interaction OR answers: *for motorists of ethnic group X, is the area-specific odds ratio for arrest different from the area-specific OR for White motorists?*

```{r inter2-or-table}
or_inter2 <- tidy(mod_logit_inter2, conf.int = FALSE, exponentiate = FALSE) %>%
  filter(grepl(":", term)) %>%
  mutate(
    conf.low  = exp(estimate - qnorm(0.975) * std.error),
    conf.high = exp(estimate + qnorm(0.975) * std.error),
    estimate  = exp(estimate),
    sig = case_when(
      p.value < 0.001 ~ "***",
      p.value < 0.01  ~ "**",
      p.value < 0.05  ~ "*",
      p.value < 0.1   ~ ".",
      TRUE            ~ ""
    )
  ) %>%
  select(term, estimate, conf.low, conf.high, p.value, sig)

or_inter2 %>%
  kbl(
    digits    = 3,
    caption   = paste0("Model 2 (ethnicity:", use_geo, ") — Interaction odds ratios with 95% CI"),
    col.names = c("Term", "OR", "CI 2.5%", "CI 97.5%", "p-value", "")
  ) %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE)
```

### LRT and BIC vs `mod_logit_best`

```{r inter2-lrt}
lrt_inter2 <- anova(mod_logit_best, mod_logit_inter2, test = "Chisq")
chi_i2 <- lrt_inter2[["Deviance"]][2]
df_i2  <- lrt_inter2[["Df"]][2]
p_i2   <- lrt_inter2[["Pr(>Chi)"]][2]
bic_i2_base  <- BIC(mod_logit_best)
bic_i2_inter <- BIC(mod_logit_inter2)

cat(sprintf(
  "LRT (ethnicity:%s): χ²(%d) = %.2f,  p = %.4g\n",
  use_geo, df_i2, chi_i2, p_i2
))
cat(sprintf(
  "BIC — mod_logit_best: %.1f  |  mod_logit_inter2: %.1f  |  ΔBIC = %.1f\n",
  bic_i2_base, bic_i2_inter, bic_i2_inter - bic_i2_base
))
```

---

## Joint AIC / BIC / LRT Summary

```{r inter-joint-comparison}
aic_inter <- AIC(mod_logit_best, mod_logit_inter1, mod_logit_inter2)
bic_inter <- BIC(mod_logit_best, mod_logit_inter1, mod_logit_inter2)

tibble(
  Model     = c(
    "mod_logit_best (additive)",
    "mod_logit_inter1  (+ethnicity:sex)",
    paste0("mod_logit_inter2  (+ethnicity:", use_geo, ")")
  ),
  df        = aic_inter$df,
  AIC       = round(aic_inter$AIC, 1),
  BIC       = round(bic_inter$BIC, 1),
  delta_AIC = round(aic_inter$AIC - min(aic_inter$AIC), 1),
  delta_BIC = round(bic_inter$BIC - min(bic_inter$BIC), 1)
) %>%
  kbl(
    caption   = "AIC / BIC — mod_logit_best vs interaction models (lower is better)",
    col.names = c("Model", "# Param", "AIC", "BIC", "Δ AIC", "Δ BIC")
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE) %>%
  row_spec(which.min(bic_inter$BIC), bold = TRUE, color = "white", background = "#2c7bb6")
```

```{r inter-lrt-summary}
tibble(
  Comparison = c(
    "mod_logit_best → mod_logit_inter1  (ethnicity:sex)",
    paste0("mod_logit_best → mod_logit_inter2  (ethnicity:", use_geo, ")")
  ),
  `χ²`    = round(c(chi_i1, chi_i2), 2),
  df      = c(df_i1, df_i2),
  p_value = signif(c(p_i1, p_i2), 3),
  ΔBIC    = round(c(bic_i1_inter - bic_i1_base, bic_i2_inter - bic_i2_base), 1),
  verdict = case_when(
    p_i1 < 0.05 & (bic_i1_inter - bic_i1_base) < 0 ~ "Supported (LRT sig + BIC ↓)",
    p_i1 < 0.05 ~ "Detectable but costly (LRT sig, BIC ↑)",
    TRUE ~ "Not supported"
  ) %>% c(
    case_when(
      p_i2 < 0.05 & (bic_i2_inter - bic_i2_base) < 0 ~ "Supported (LRT sig + BIC ↓)",
      p_i2 < 0.05 ~ "Detectable but costly (LRT sig, BIC ↑)",
      TRUE ~ "Not supported"
    )
  )
) %>%
  kbl(caption = "Likelihood ratio tests and BIC change: additive vs each interaction model") %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE)
```

**Interpretation guide:**

-   LRT significant + ΔBIC < 0: the interaction is both statistically detectable and informationally justified — include it.
-   LRT significant + ΔBIC > 0: the interaction is real but the number of extra parameters exceeds the BIC penalty budget — the additive model is preferred on parsimony grounds.
-   LRT non-significant: no evidence of genuine non-additivity at the 5% level.

---

## Marginal Effects

Predicted probabilities are computed by fixing all non-focal predictors at their modal (most frequent) training-set value and varying the two focal predictors across all level combinations. This isolates the interaction pattern on the probability scale.

```{r inter-helpers}
# One-row data frame of modal / median values for a set of variable names
build_modal_row <- function(data, vars) {
  row_list <- lapply(vars, function(v) {
    x <- data[[v]]
    if (is.factor(x)) {
      factor(names(sort(table(x), decreasing = TRUE))[1], levels = levels(x))
    } else if (is.numeric(x)) {
      median(x, na.rm = TRUE)
    } else {
      names(sort(table(x), decreasing = TRUE))[1]
    }
  })
  as.data.frame(setNames(row_list, vars))
}
```

### Model 1 — Predicted P(arrested = 1) by Ethnicity × Sex

```{r inter1-marginal-plot, fig.width=9, fig.height=4.5}
vars_i1    <- setdiff(all.vars(f_inter1)[-1], c("ethnicity", "sex"))
modal_i1   <- build_modal_row(train, vars_i1)

pred_grid1 <- expand.grid(
  ethnicity = levels(train$ethnicity),
  sex       = levels(train$sex)
)
pred_grid1 <- cbind(pred_grid1, modal_i1[rep(1L, nrow(pred_grid1)), , drop = FALSE])
rownames(pred_grid1) <- NULL
pred_grid1$prob <- predict(mod_logit_inter1, newdata = pred_grid1, type = "response")

ggplot(pred_grid1, aes(x = ethnicity, y = prob, fill = sex)) +
  geom_col(position = position_dodge(width = 0.75), width = 0.65) +
  geom_text(
    aes(label = scales::percent(prob, accuracy = 0.1)),
    position = position_dodge(width = 0.75),
    vjust = -0.4, size = 3
  ) +
  scale_fill_viridis_d(option = "mako", begin = 0.25, end = 0.75) +
  scale_y_continuous(
    labels = scales::percent_format(accuracy = 0.1),
    expand = expansion(mult = c(0, 0.12))
  ) +
  labs(
    x        = "Ethnicity",
    y        = "Predicted P(arrested = 1)",
    fill     = "Sex",
    title    = "Model 1 — Marginal predicted arrest probability: Ethnicity × Sex",
    subtitle = "All other predictors held at modal training values",
    caption  = "mod_logit_inter1: f_final + ethnicity:sex"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    axis.text.x      = element_text(angle = 30, hjust = 1),
    legend.position  = "bottom",
    panel.grid.minor = element_blank()
  )
```

### Model 2 — Predicted P(arrested = 1) by Ethnicity × Geography

```{r inter2-marginal-plot, fig.width=10, fig.height=5}
vars_i2  <- setdiff(all.vars(f_inter2)[-1], c("ethnicity", use_geo))
modal_i2 <- build_modal_row(train, vars_i2)

geo_levels <- levels(train[[use_geo]])
pred_grid2 <- expand.grid(
  ethnicity = levels(train$ethnicity),
  geo__     = geo_levels,
  stringsAsFactors = FALSE
)
names(pred_grid2)[2] <- use_geo
pred_grid2[[use_geo]] <- factor(pred_grid2[[use_geo]], levels = geo_levels)

pred_grid2 <- cbind(pred_grid2, modal_i2[rep(1L, nrow(pred_grid2)), , drop = FALSE])
rownames(pred_grid2) <- NULL
pred_grid2$prob <- predict(mod_logit_inter2, newdata = pred_grid2, type = "response")

ggplot(pred_grid2, aes(x = ethnicity, y = prob, fill = .data[[use_geo]])) +
  geom_col(position = position_dodge(width = 0.75), width = 0.65) +
  scale_fill_viridis_d(option = "turbo", begin = 0.05, end = 0.92) +
  scale_y_continuous(
    labels = scales::percent_format(accuracy = 0.1),
    expand = expansion(mult = c(0, 0.08))
  ) +
  labs(
    x        = "Ethnicity",
    y        = "Predicted P(arrested = 1)",
    fill     = use_geo,
    title    = paste0("Model 2 — Marginal predicted arrest probability: Ethnicity × ", use_geo),
    subtitle = "All other predictors held at modal training values",
    caption  = paste0("mod_logit_inter2: f_final + ethnicity:", use_geo)
  ) +
  theme_minimal(base_size = 11) +
  theme(
    axis.text.x      = element_text(angle = 30, hjust = 1),
    legend.position  = "bottom",
    panel.grid.minor = element_blank()
  )
```

### Model 2 — Interaction Heatmap

The heatmap surface makes deviations from a pure additive pattern immediately visible: under pure additivity the colour gradient across each row would be identical; any row that breaks the uniform pattern reveals an interaction effect.

```{r inter2-heatmap, fig.width=9, fig.height=4.5}
ggplot(pred_grid2, aes(x = .data[[use_geo]], y = ethnicity, fill = prob)) +
  geom_tile(colour = "white", linewidth = 0.6) +
  geom_text(aes(label = scales::percent(prob, accuracy = 0.1)), size = 3.2, fontface = "bold") +
  scale_fill_gradient2(
    low      = "#4575b4",
    mid      = "#ffffbf",
    high     = "#d73027",
    midpoint = mean(pred_grid2$prob),
    labels   = scales::percent_format(accuracy = 0.1),
    name     = "P(arrested)"
  ) +
  labs(
    x        = use_geo,
    y        = "Ethnicity",
    title    = paste0("Model 2 — P(arrested=1) surface: Ethnicity × ", use_geo),
    subtitle = "Non-uniform colour rows signal interaction; all other predictors at modal values",
    caption  = paste0("mod_logit_inter2: f_final + ethnicity:", use_geo)
  ) +
  theme_minimal(base_size = 11) +
  theme(
    axis.text.x     = element_text(angle = 30, hjust = 1),
    panel.grid      = element_blank(),
    legend.position = "right"
  )
```

---

## Test Set Evaluation

```{r inter-test-eval}
prob_inter1 <- predict(mod_logit_inter1, newdata = test, type = "response")
prob_inter2 <- predict(mod_logit_inter2, newdata = test, type = "response")

bind_rows(
  compute_metrics(test$arrested, prob_logit,   "mod_logit_best (additive)"),
  compute_metrics(test$arrested, prob_inter1,  "mod_logit_inter1  (+ethnicity:sex)"),
  compute_metrics(test$arrested, prob_inter2,  paste0("mod_logit_inter2  (+ethnicity:", use_geo, ")"))
) %>%
  kbl(
    caption   = "Interaction models vs mod_logit_best — test set metrics (Youden threshold)",
    col.names = c("Model","AUC","PR-AUC","Log-Loss","Brier",
                  "Threshold","Sensitivity","Specificity","Precision","F1","MCC","Bal. Acc.")
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = TRUE)
```

### Notes

-   **LRT significance**: A significant LRT indicates genuine non-additivity — the combined ethnicity–sex (or ethnicity–geography) effect cannot be decomposed into independent additive contributions. This is a statistical claim, not a causal one.
-   **BIC as parsimony judge**: BIC penalises model complexity more heavily than AIC (penalty $\ln n$ vs 2 per parameter). Given ~5.5% positive-class prevalence and sparse ethnicity × area cells in 2-wheel stops, BIC is the more appropriate arbiter of whether the interaction parameters are earning their keep.
-   **Marginal effect plots**: Probabilities are computed at modal covariate values, not averaged over the full covariate distribution. Patterns visible here are interpretable as the interaction effect *at the modal stop profile*; marginal standardisation (average marginal effects) would give a population-averaged estimate but requires more computation.
-   **Observational caution**: Disparities in predicted probabilities across demographic groups are associations in observational, self-reported data. They may reflect policing patterns, reporting artefacts, or data collection differences rather than causal differences in underlying behaviour. Comparative interpretations across ethnic groups must be made with this limitation in mind.

---

# iii · LDA

## Assumptions and Setup

Linear Discriminant Analysis assumes each class follows a multivariate normal distribution with a **shared** covariance matrix (i.e., homoscedastic classes). This produces a linear decision boundary, making it more constrained than QDA but with lower variance and fewer parameters to estimate. The same caveats about discrete predictors apply here as in QDA, but the shared covariance assumption makes LDA more numerically stable when the minority class is small.

-   **Class imbalance**: the arrested = 1 class represents ≈ 5.5% of training observations; priors are estimated from class proportions in the training set.
-   **Dimensionality**: after dummy encoding `f_final` produces `r ncol(model.matrix(f_final, train[1,])) - 1` features. With a shared covariance matrix, the pooled estimate uses all training observations, so rank deficiency is not a concern here.

```{r load-MASS-lda, include=FALSE}
if (!requireNamespace("MASS", quietly = TRUE)) install.packages("MASS")
# Do NOT library(MASS) — it masks dplyr::select; use MASS:: prefix throughout
```

## Model Fit

```{r fit-lda}
mod_lda <- MASS::lda(f_final, data = train)
```

## Prior Probabilities and Group Means

`MASS::lda` reports the estimated prior probabilities and within-class means of every predictor. The group means reveal which features are most discriminative under the shared-covariance assumption.

```{r lda-priors}
cat("Prior probabilities:\n")
print(mod_lda$prior)
```

```{r lda-means-plot, fig.width=11, fig.height=5}
means_df_lda <- as.data.frame(t(mod_lda$means)) %>%
  tibble::rownames_to_column("feature") %>%
  rename(arrested_0 = `0`, arrested_1 = `1`) %>%
  mutate(
    diff      = arrested_1 - arrested_0,
    direction = ifelse(diff > 0, "Higher in arrested", "Lower in arrested"),
    feature   = stringr::str_wrap(feature, 22)
  ) %>%
  arrange(desc(abs(diff)))

ggplot(head(means_df_lda, 30), aes(x = diff, y = reorder(feature, diff), fill = direction)) +
  geom_col(width = 0.7) +
  geom_vline(xintercept = 0, linewidth = 0.4) +
  scale_fill_manual(values = c("Higher in arrested" = "#d73027", "Lower in arrested" = "#2c7bb6")) +
  labs(
    x     = "Mean difference  (arrested=1) − (arrested=0)",
    y     = NULL,
    fill  = NULL,
    title = "LDA — Top 30 features by within-class mean difference"
  ) +
  theme_minimal(base_size = 10) +
  theme(legend.position = "bottom", panel.grid.minor = element_blank())
```

## Evaluation on the Test Set

```{r lda-eval}
pred_lda  <- predict(mod_lda, newdata = test)
prob_lda  <- pred_lda$posterior[, "1"]
roc_lda   <- pROC::roc(test$arrested, prob_lda, quiet = TRUE)
auc_lda   <- as.numeric(pROC::auc(roc_lda))
thr_lda   <- best_thresh(roc_lda)
class_lda <- as.integer(prob_lda >= thr_lda)
met_lda   <- class_metrics(test$arrested, class_lda)

cat(sprintf(
  "AUC: %.4f  |  Threshold (Youden): %.4f\nSensitivity: %.3f  |  Specificity: %.3f  |  Balanced accuracy: %.3f\n",
  auc_lda, thr_lda, met_lda$sensitivity, met_lda$specificity, met_lda$balanced_acc
))
```

```{r lda-cm-plot, fig.width=4, fig.height=3.5}
plot_cm(test$arrested, class_lda, title = "LDA — Confusion Matrix (Youden threshold)")
```

```{r lda-posterior-hist, fig.width=7, fig.height=3.5}
tibble(prob = prob_lda, actual = factor(test$arrested, levels = c(0,1),
                                        labels = c("Not arrested","Arrested"))) %>%
  ggplot(aes(x = prob, fill = actual)) +
  geom_histogram(bins = 60, position = "identity", alpha = 0.65) +
  geom_vline(xintercept = thr_lda, linetype = "dashed", colour = "grey30") +
  scale_fill_viridis_d(option = "mako", begin = 0.3, end = 0.8) +
  scale_x_log10(labels = scales::label_number(accuracy = 0.001)) +
  labs(
    x     = "Predicted probability (log scale)",
    y     = "Count",
    fill  = NULL,
    title = "LDA — Posterior probability distribution by class",
    caption = "Dashed line: Youden optimal threshold"
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")
```

```{r lda-metrics-table}
compute_metrics(test$arrested, prob_lda, "LDA") %>%
  kbl(
    caption   = "LDA — full evaluation metrics (test set, Youden threshold)",
    col.names = c("Model","AUC","PR-AUC","Log-Loss","Brier",
                  "Threshold","Sensitivity","Specificity","Precision","F1","MCC","Bal. Acc.")
  ) %>%
  kable_styling(bootstrap_options = c("striped","condensed"), full_width = TRUE)
```

```{r lda-eval-plots, fig.width=10, fig.height=8}
eval_quad_plot(test$arrested, prob_lda, "LDA")
```

---

# v · Naive Bayes

## Assumptions and Setup

Naive Bayes assumes conditional independence of all features given the class label. 

For purely categorical predictors it estimates a frequency table per feature–class combination (with optional Laplace smoothing to prevent zero-probability cells). 

The independence assumption is almost certainly violated here. In fact, time of day, area, and demographic features are usually correlated. However, Naive Bayes often produces competitive probabilistic rankings even when it is. The Laplace smoothing parameter $k$ is the only tuning knob; it is selected by 10-fold cross-validated AUC on the training set using the same folds as section ii.

```{r load-e1071, include=FALSE}
if (!requireNamespace("e1071", quietly = TRUE)) install.packages("e1071")
library(e1071)
```

## Laplace Smoothing Tuning

```{r nb-tune, cache=TRUE}
laplace_grid <- c(0, 0.5, 1, 2, 5, 10)

nb_cv_auc <- function(k, data, folds_vec, formula) {
  n_folds  <- max(folds_vec)
  auc_vals <- numeric(n_folds)
  for (i in seq_len(n_folds)) {
    tr  <- data[folds_vec != i, ]
    te  <- data[folds_vec == i, ]
    fit <- e1071::naiveBayes(formula, data = tr, laplace = k)
    # posterior: columns named after class levels; column 2 = P(arrested=1)
    prob <- predict(fit, newdata = te, type = "raw")[, 2]
    auc_vals[i] <- tryCatch(
      as.numeric(pROC::auc(te$arrested, prob, quiet = TRUE)),
      error = function(e) NA_real_
    )
  }
  mean(auc_vals, na.rm = TRUE)
}

nb_aucs <- sapply(
  laplace_grid, nb_cv_auc,
  data = train, folds_vec = cv_folds, 
  formula = f_final
  )

best_laplace <- laplace_grid[which.max(nb_aucs)]

cat("Best Laplace k:", best_laplace, " | CV AUC:", round(max(nb_aucs), 4), "\n")
```

```{r nb-tune-plot, fig.width=6, fig.height=3.5}
tibble(k = laplace_grid, CV_AUC = nb_aucs) %>%
  ggplot(aes(x = factor(k), y = CV_AUC)) +
  geom_col(fill = "#2c7bb6", width = 0.6) +
  geom_text(aes(label = round(CV_AUC, 4)), vjust = -0.4, size = 3.5) +
  labs(
    x     = "Laplace smoothing parameter k",
    y     = "10-fold CV AUC",
    title = "Naive Bayes — Laplace tuning"
  ) +
  coord_cartesian(ylim = c(min(nb_aucs) - 0.002, max(nb_aucs) + 0.005)) +
  theme_minimal(base_size = 11) +
  theme(panel.grid.minor = element_blank())
```

## Model Fit

```{r fit-nb}
mod_nb <- e1071::naiveBayes(f_final, data = train, laplace = best_laplace)
```

## Conditional Probability Tables — Key Predictors

Each table entry shows $\hat{P}(\text{feature level} \mid \text{class})$. Features whose conditional distributions differ most between classes are the strongest discriminators under the independence assumption.

```{r nb-tables-plot, fig.width=11, fig.height=6}
# Extract tables for a selected set of substantively interesting predictors
plot_nb_feature <- function(feature_name, mod) {
  tbl <- mod$tables[[feature_name]]
  as.data.frame(tbl) %>%
    rename(level = 2, class = 1, prob = Freq) %>%
    mutate(class = factor(class, labels = c("Not arrested", "Arrested"))) %>%
    ggplot(aes(x = level, y = prob, fill = class)) +
    geom_col(position = position_dodge(width = 0.75), width = 0.65) +
    scale_fill_viridis_d(option = "mako", begin = 0.3, end = 0.8) +
    scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
    labs(x = NULL, y = "P(level | class)", title = feature_name, fill = NULL) +
    theme_minimal(base_size = 9) +
    theme(
      axis.text.x    = element_text(angle = 30, hjust = 1),
      legend.position = "bottom",
      panel.grid.minor = element_blank()
    )
}

key_preds <- intersect(
  c("ethnicity", "sex", "age_bracket", use_geo, use_time),
  names(mod_nb$tables)
)

plots_nb <- lapply(key_preds, plot_nb_feature, mod = mod_nb)
patchwork::wrap_plots(plots_nb, ncol = 3) +
  patchwork::plot_annotation(
    title   = "Naive Bayes — Conditional probability by class for key predictors",
    caption = "P(level | class) from the training-set frequency tables"
  ) +
  patchwork::plot_layout(guides = "collect") &
  theme(legend.position = "bottom")
```

## Evaluation on the Test Set

```{r nb-eval}
prob_nb  <- predict(mod_nb, newdata = test, type = "raw")[, 2]
roc_nb   <- pROC::roc(test$arrested, prob_nb, quiet = TRUE)
auc_nb   <- as.numeric(pROC::auc(roc_nb))
thr_nb   <- best_thresh(roc_nb)
class_nb <- as.integer(prob_nb >= thr_nb)
met_nb   <- class_metrics(test$arrested, class_nb)

cat(sprintf(
  "AUC: %.4f  |  Threshold (Youden): %.4f\nSensitivity: %.3f  |  Specificity: %.3f  |  Balanced accuracy: %.3f\n",
  auc_nb, thr_nb, met_nb$sensitivity, met_nb$specificity, met_nb$balanced_acc
))
```

```{r nb-cm-plot, fig.width=4, fig.height=3.5}
plot_cm(test$arrested, class_nb, title = "Naive Bayes — Confusion Matrix (Youden threshold)")
```

```{r nb-metrics-table}
compute_metrics(test$arrested, prob_nb, "Naive Bayes") %>%
  kbl(
    caption   = "Naive Bayes — full evaluation metrics (test set, Youden threshold)",
    col.names = c("Model","AUC","PR-AUC","Log-Loss","Brier",
                  "Threshold","Sensitivity","Specificity","Precision","F1","MCC","Bal. Acc.")
  ) %>%
  kable_styling(bootstrap_options = c("striped","condensed"), full_width = TRUE)
```

```{r nb-eval-plots, fig.width=10, fig.height=8}
eval_quad_plot(test$arrested, prob_nb, "Naive Bayes")
```

---

# vii · LASSO

## Setup

Logistic LASSO penalises the sum of absolute coefficient values ($\ell_1$ penalty), which shrinks small effects to exactly zero and performs implicit variable selection. We use the `glmnet` package with `family = "binomial"` and `alpha = 1` (pure LASSO, no ridge component). All categorical predictors are dummy-encoded via `model.matrix`; the intercept column is excluded because `glmnet` adds its own unpenalised intercept. The penalty parameter $\lambda$ is selected by 10-fold cross-validated AUC on the training set.

```{r load-glmnet, include=FALSE}
if (!requireNamespace("glmnet", quietly = TRUE)) install.packages("glmnet")
library(glmnet)
```

## Model Matrix

```{r lasso-matrix}
mm_train <- model.frame(f_final, data = train, na.action = na.omit)
X_train  <- model.matrix(f_final, data = mm_train)[, -1]
y_train  <- mm_train$arrested

mm_test  <- model.frame(f_final, data = test, na.action = na.omit)
X_test   <- model.matrix(f_final, data = mm_test)[, -1]
y_test   <- mm_test$arrested

cat("Training matrix:", nrow(X_train), "×", ncol(X_train), "\n")
cat("Test matrix    :", nrow(X_test),  "×", ncol(X_test),  "\n")
```

## Cross-Validated Lambda

We optimise $\lambda$ by maximising cross-validated AUC (`type.measure = "auc"`), which is the appropriate metric for an imbalanced binary outcome.

```{r fit-lasso-cv, cache=TRUE}
set.seed(2026)
cv_lasso <- cv.glmnet(
  X_train, y_train,
  family       = "binomial",
  alpha        = 1,
  nfolds       = 10,
  type.measure = "auc"
)

cat(sprintf(
  "λ.min  = %.6f  (CV AUC = %.4f)\nλ.1se  = %.6f  (CV AUC = %.4f)\n",
  cv_lasso$lambda.min, max(cv_lasso$cvm),
  cv_lasso$lambda.1se,
  cv_lasso$cvm[cv_lasso$lambda == cv_lasso$lambda.1se]
))
```

```{r lasso-cv-plot, fig.width=7, fig.height=4}
# Replicate the cv.glmnet plot in ggplot2
cv_df <- tibble(
  log_lambda = log(cv_lasso$lambda),
  cvm        = cv_lasso$cvm,
  cvup       = cv_lasso$cvup,
  cvlo       = cv_lasso$cvlo,
  nzero      = cv_lasso$nzero
)

ggplot(cv_df, aes(x = log_lambda, y = cvm)) +
  geom_ribbon(aes(ymin = cvlo, ymax = cvup), alpha = 0.15, fill = "#2c7bb6") +
  geom_line(colour = "#2c7bb6") +
  geom_point(size = 0.8, colour = "#2c7bb6") +
  geom_vline(xintercept = log(cv_lasso$lambda.min), linetype = "dashed",
             colour = "#d73027", linewidth = 0.7) +
  geom_vline(xintercept = log(cv_lasso$lambda.1se), linetype = "dashed",
             colour = "#fdae61", linewidth = 0.7) +
  annotate("text", x = log(cv_lasso$lambda.min), y = min(cv_df$cvlo),
           label = "λ.min", colour = "#d73027", hjust = -0.1, size = 3.5) +
  annotate("text", x = log(cv_lasso$lambda.1se), y = min(cv_df$cvlo),
           label = "λ.1se", colour = "#b36200", hjust = -0.1, size = 3.5) +
  labs(
    x     = "log(λ)",
    y     = "10-fold CV AUC",
    title = "LASSO — Cross-validated AUC vs regularisation strength"
  ) +
  theme_minimal(base_size = 11) +
  theme(panel.grid.minor = element_blank())
```

## Coefficient Path

```{r lasso-path-plot, fig.width=9, fig.height=5}
# Tidy the full coefficient path
coef_path <- as.matrix(coef(cv_lasso$glmnet.fit))   # (p+1) × n_lambda
lambdas   <- cv_lasso$glmnet.fit$lambda

path_df <- as.data.frame(t(coef_path[-1, ])) %>%   # drop intercept row
  mutate(log_lambda = log(lambdas)) %>%
  pivot_longer(-log_lambda, names_to = "feature", values_to = "coef") %>%
  filter(coef != 0)

# Only label non-zero at lambda.min for clarity
active_min <- names(which(coef(cv_lasso, s = "lambda.min")[-1, 1] != 0))

ggplot(path_df, aes(x = log_lambda, y = coef, group = feature,
                    colour = feature %in% active_min)) +
  geom_line(linewidth = 0.45, alpha = 0.8) +
  geom_vline(xintercept = log(cv_lasso$lambda.min), linetype = "dashed",
             colour = "#d73027", linewidth = 0.7) +
  geom_vline(xintercept = log(cv_lasso$lambda.1se), linetype = "dashed",
             colour = "#fdae61", linewidth = 0.7) +
  scale_colour_manual(values = c("TRUE" = "#2c7bb6", "FALSE" = "grey70"),
                      guide  = "none") +
  labs(
    x       = "log(λ)",
    y       = "Coefficient",
    title   = "LASSO — Coefficient paths",
    caption = "Blue paths: active at λ.min  |  Dashed lines: λ.min (red) and λ.1se (orange)"
  ) +
  theme_minimal(base_size = 11) +
  theme(panel.grid.minor = element_blank())
```

## Selected Coefficients

```{r lasso-coef-tables}
extract_coefs <- function(cv_fit, s_name) {
  cf <- coef(cv_fit, s = s_name)
  tibble(
    feature  = rownames(cf)[-1],
    estimate = as.numeric(cf[-1])
  ) %>%
    filter(estimate != 0) %>%
    arrange(desc(abs(estimate))) %>%
    mutate(OR = round(exp(estimate), 3), estimate = round(estimate, 4))
}

coef_min <- extract_coefs(cv_lasso, "lambda.min")
coef_1se <- extract_coefs(cv_lasso, "lambda.1se")

cat("Non-zero coefficients at λ.min:", nrow(coef_min), "\n")
cat("Non-zero coefficients at λ.1se:", nrow(coef_1se), "\n")
```

```{r lasso-coef-min-table}
coef_min %>%
  kbl(
    caption   = "LASSO coefficients at λ.min (sorted by |estimate|)",
    col.names = c("Feature", "Log-odds", "Odds Ratio")
  ) %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE)
```

```{r lasso-coef-plot, fig.width=8, fig.height=0.35 * nrow(coef_min) + 1.5}
coef_min %>%
  mutate(
    direction = ifelse(estimate > 0, "Increases odds", "Decreases odds"),
    feature   = stringr::str_wrap(feature, 30)
  ) %>%
  ggplot(aes(x = estimate, y = reorder(feature, estimate), fill = direction)) +
  geom_col(width = 0.7) +
  geom_vline(xintercept = 0, linewidth = 0.4) +
  scale_fill_manual(values = c("Increases odds" = "#d73027", "Decreases odds" = "#2c7bb6")) +
  labs(
    x     = "LASSO coefficient (log-odds scale)",
    y     = NULL,
    fill  = NULL,
    title = "LASSO — Non-zero coefficients at λ.min"
  ) +
  theme_minimal(base_size = 10) +
  theme(legend.position = "bottom", panel.grid.minor = element_blank())
```

## Evaluation on the Test Set

```{r lasso-eval}
prob_lasso_min <- as.numeric(predict(cv_lasso, newx = X_test, s = "lambda.min", type = "response"))
prob_lasso_1se <- as.numeric(predict(cv_lasso, newx = X_test, s = "lambda.1se", type = "response"))

roc_lasso_min <- pROC::roc(y_test, prob_lasso_min, quiet = TRUE)
roc_lasso_1se <- pROC::roc(y_test, prob_lasso_1se, quiet = TRUE)
auc_lasso_min <- as.numeric(pROC::auc(roc_lasso_min))
auc_lasso_1se <- as.numeric(pROC::auc(roc_lasso_1se))

thr_lasso_min  <- best_thresh(roc_lasso_min)
class_lasso_min <- as.integer(prob_lasso_min >= thr_lasso_min)
met_lasso_min   <- class_metrics(y_test, class_lasso_min)

cat(sprintf(
  "λ.min — AUC: %.4f  |  Threshold: %.4f  |  Sens: %.3f  |  Spec: %.3f  |  Bal.Acc: %.3f\n",
  auc_lasso_min, thr_lasso_min,
  met_lasso_min$sensitivity, met_lasso_min$specificity, met_lasso_min$balanced_acc
))

thr_lasso_1se   <- best_thresh(roc_lasso_1se)
class_lasso_1se <- as.integer(prob_lasso_1se >= thr_lasso_1se)
met_lasso_1se   <- class_metrics(y_test, class_lasso_1se)

cat(sprintf(
  "λ.1se — AUC: %.4f  |  Threshold: %.4f  |  Sens: %.3f  |  Spec: %.3f  |  Bal.Acc: %.3f\n",
  auc_lasso_1se, thr_lasso_1se,
  met_lasso_1se$sensitivity, met_lasso_1se$specificity, met_lasso_1se$balanced_acc
))
```

```{r lasso-cm-plots, fig.width=8, fig.height=3.5}
p_cm_min <- plot_cm(y_test, class_lasso_min, "LASSO (λ.min) — Confusion Matrix")
p_cm_1se <- plot_cm(y_test, class_lasso_1se, "LASSO (λ.1se) — Confusion Matrix")
p_cm_min | p_cm_1se
```

```{r lasso-metrics-table}
bind_rows(
  compute_metrics(y_test, prob_lasso_min, "LASSO λ.min"),
  compute_metrics(y_test, prob_lasso_1se, "LASSO λ.1se")
) %>%
  kbl(
    caption   = "LASSO — full evaluation metrics (test set, Youden threshold)",
    col.names = c("Model","AUC","PR-AUC","Log-Loss","Brier",
                  "Threshold","Sensitivity","Specificity","Precision","F1","MCC","Bal. Acc.")
  ) %>%
  kable_styling(bootstrap_options = c("striped","condensed"), full_width = TRUE)
```

```{r lasso-eval-plots-min, fig.width=10, fig.height=8}
eval_quad_plot(y_test, prob_lasso_min, "LASSO (λ.min)")
```

```{r lasso-eval-plots-1se, fig.width=10, fig.height=8}
eval_quad_plot(y_test, prob_lasso_1se, "LASSO (λ.1se)")
```

---

# ix · GAM

## Setup

A Generalized Additive Model (GAM) extends logistic regression by allowing selected continuous predictors to enter through smooth functions rather than linear terms. We fit with `mgcv::gam` using `family = binomial` and REML smoothness selection. Two continuous predictors receive data-driven smooth terms:

-   `s(hour_num, bs = "cc", k = 10)`: a **cyclic cubic spline** for hour-of-day (hours 23 and 0 are adjacent on the clock). The EDA showed a non-linear enforcement cycle across the day that a linear slope or fixed brackets cannot fully capture.
-   `s(age, bs = "cr", k = 6)`: a **cubic regression spline** for driver age. The EDA found a non-monotone age–arrest relationship (elevated under-18 and over-65 rates) that neither a linear slope nor an arbitrary bracket optimally represents.

All other predictors (`year`, `month`, `weekday`, `sex`, `ethnicity`, geographic grouping) remain parametric factors, matching the selected predictor forms from section ii.

```{r load-mgcv, include=FALSE}
if (!requireNamespace("mgcv", quietly = TRUE)) install.packages("mgcv")
```

```{r gam-data-prep}
# hour is stored as factor after factor-prep; recover numeric version for s()
train_gam <- train %>% mutate(hour_num = as.integer(as.character(hour)))
test_gam  <- test  %>% mutate(hour_num = as.integer(as.character(hour)))
```

## Model Fit

```{r fit-gam}
f_gam <- as.formula(sprintf(
  "arrested ~ year + month + weekday + s(hour_num, bs = 'cc', k = 10) + %s + sex + ethnicity + s(age, bs = 'cr', k = 6)",
  use_geo
))

mod_gam <- mgcv::gam(f_gam, family = binomial, data = train_gam, method = "REML")
summary(mod_gam)
```

## Smooth Term Plots

Estimated smooth effects on the log-odds scale, centred at zero. Shaded bands are 2 SE pointwise confidence intervals. Effective degrees of freedom (edf) reported in the summary — values above 1 indicate non-linearity.

```{r gam-smooth-plots, fig.width=10, fig.height=4}
par(mfrow = c(1, 2), mar = c(4, 4, 2.5, 1))
plot(mod_gam, shade = TRUE, shade.col = "lightblue", seWithMean = TRUE,
     pages = 0, scale = 0)
par(mfrow = c(1, 1))
```

## Evaluation on the Test Set

```{r gam-eval}
prob_gam  <- predict(mod_gam, newdata = test_gam, type = "response")
roc_gam   <- pROC::roc(test$arrested, prob_gam, quiet = TRUE)
auc_gam   <- as.numeric(pROC::auc(roc_gam))
thr_gam   <- best_thresh(roc_gam)
class_gam <- as.integer(prob_gam >= thr_gam)
met_gam   <- class_metrics(test$arrested, class_gam)

cat(sprintf(
  "AUC: %.4f  |  Threshold (Youden): %.4f\nSensitivity: %.3f  |  Specificity: %.3f  |  Balanced accuracy: %.3f\n",
  auc_gam, thr_gam, met_gam$sensitivity, met_gam$specificity, met_gam$balanced_acc
))
```

```{r gam-cm-plot, fig.width=4, fig.height=3.5}
plot_cm(test$arrested, class_gam, title = "GAM — Confusion Matrix (Youden threshold)")
```

```{r gam-metrics-table}
compute_metrics(test$arrested, prob_gam, "GAM") %>%
  kbl(
    caption   = "GAM — full evaluation metrics (test set, Youden threshold)",
    col.names = c("Model","AUC","PR-AUC","Log-Loss","Brier",
                  "Threshold","Sensitivity","Specificity","Precision","F1","MCC","Bal. Acc.")
  ) %>%
  kable_styling(bootstrap_options = c("striped","condensed"), full_width = TRUE)
```

```{r gam-eval-plots, fig.width=10, fig.height=8}
eval_quad_plot(test$arrested, prob_gam, "GAM")
```

---

# x · All-Method Comparison

## ROC Curves

```{r roc-curves-all, fig.width=7, fig.height=6}
# prob_logit / roc_logit / auc_logit already defined in section ii-b
roc_list <- list(
  "Logistic"     = roc_logit,
  "LDA"          = roc_lda,
  "Naive Bayes"  = roc_nb,
  "LASSO λ.min"  = roc_lasso_min,
  "LASSO λ.1se"  = roc_lasso_1se,
  "GAM"          = roc_gam
)

pROC::ggroc(roc_list, linewidth = 0.8) +
  geom_abline(slope = 1, intercept = 1, linetype = "dashed", colour = "grey60") +
  scale_colour_viridis_d(option = "turbo", begin = 0.05, end = 0.92) +
  labs(
    x      = "Specificity",
    y      = "Sensitivity",
    colour = NULL,
    title  = "ROC curves — all methods (test set)"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

## Metrics Summary

```{r metrics-table}
# All threshold/class/metric objects already defined in per-model sections
bind_rows(
  compute_metrics(test$arrested, prob_logit,     "Logistic"),
  compute_metrics(test$arrested, prob_lda,       "LDA"),
  compute_metrics(test$arrested, prob_nb,        "Naive Bayes"),
  compute_metrics(y_test,        prob_lasso_min, "LASSO λ.min"),
  compute_metrics(y_test,        prob_lasso_1se, "LASSO λ.1se"),
  compute_metrics(test$arrested, prob_gam,       "GAM")
) %>%
  kbl(
    caption   = "All methods — full evaluation metrics (test set, Youden threshold)",
    col.names = c("Method","AUC","PR-AUC","Log-Loss","Brier",
                  "Threshold","Sensitivity","Specificity","Precision","F1","MCC","Bal. Acc.")
  ) %>%
  kable_styling(bootstrap_options = c("striped","hover","condensed"), full_width = TRUE) %>%
  row_spec(
    which.max(c(as.numeric(pROC::auc(roc_logit)), auc_lda, auc_nb, auc_lasso_min, auc_lasso_1se, auc_gam)),
    bold = TRUE, color = "white", background = "#2c7bb6"
    )
```

The table reports AUC and PR-AUC as primary discrimination metrics, Log-Loss and Brier score as calibration metrics, and threshold-dependent metrics (Sensitivity, Specificity, Precision, F1, MCC, Balanced Accuracy) at each model's Youden-optimal threshold. Given the severe class imbalance (~5.5% positive rate), PR-AUC and MCC are more informative than raw accuracy or even balanced accuracy alone.

---

# xi · Model Diagnostics

```{r diag-pkgs, include=FALSE}
if (!requireNamespace("car",        quietly = TRUE)) install.packages("car")
if (!requireNamespace("ggcorrplot", quietly = TRUE)) install.packages("ggcorrplot")
library(car)
library(ggcorrplot)
```

## Logistic Regression — Multicollinearity (VIF)

Variance Inflation Factors measure how much each predictor's variance is inflated by collinearity with the others. A VIF > 5 indicates moderate concern; VIF > 10 warrants action (dropping or combining predictors). Because `mod_logit_best` contains multi-level factors, `car::vif()` returns the generalised VIF (GVIF) scaled by degrees of freedom: $\text{GVIF}^{1/(2\,\text{df})}$ is comparable to the square root of a standard VIF.

```{r logit-vif}
vif_tbl <- as.data.frame(car::vif(mod_logit_best))
vif_tbl$predictor <- rownames(vif_tbl)
rownames(vif_tbl) <- NULL

# Rename columns robustly (vif() returns 3-col df for factors)
if (ncol(vif_tbl) == 4) {
  names(vif_tbl)[1:3] <- c("GVIF", "Df", "GVIF_adj")
} else {
  names(vif_tbl)[1] <- "GVIF"
  vif_tbl$Df       <- 1
  vif_tbl$GVIF_adj <- sqrt(vif_tbl$GVIF)
}

vif_tbl %>%
  arrange(desc(GVIF_adj)) %>%
  mutate(flag = ifelse(GVIF_adj > sqrt(5), "⚠ >5", "OK")) %>%
  kbl(
    digits  = 3,
    caption = "Logistic regression — Generalised VIF (GVIF^(1/2df) comparable to sqrt(VIF))"
  ) %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE)
```

```{r logit-vif-plot, fig.width=7, fig.height=4}
ggplot(vif_tbl, aes(x = reorder(predictor, GVIF_adj), y = GVIF_adj,
                    fill = GVIF_adj > sqrt(5))) +
  geom_col(width = 0.65) +
  geom_hline(yintercept = sqrt(5), linetype = "dashed", colour = "firebrick", linewidth = 0.7) +
  scale_fill_manual(values = c("FALSE" = "#4575b4", "TRUE" = "#d73027"),
                    labels = c("OK", ">5 threshold"), name = NULL) +
  coord_flip() +
  labs(
    x     = NULL,
    y     = expression(GVIF^{1/(2~df)}),
    title = "Logistic regression — Variance Inflation Factors",
    subtitle = "Dashed line = sqrt(5) threshold"
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom", panel.grid.minor = element_blank())
```

Predictors with $\text{GVIF}^{1/(2\,\text{df})} > \sqrt{5}$ (red bars, if any) share substantial variance with other covariates. For this model the main risk is between `month` and `year` (temporal confounding) and among demographic variables if certain combinations are sparse. Mild inflation is acceptable given the large sample size; severe inflation would destabilise coefficient estimates and widen confidence intervals without affecting predictive accuracy.

---

## Logistic Regression — Influential Observations (Cook's Distance)

Cook's Distance measures how much all fitted values change when observation $i$ is deleted. Values above $4/n$ (a common rule of thumb) flag leverage points that may disproportionately drive the estimates.

```{r logit-influence, fig.width=10, fig.height=8, cache=TRUE}
car::influenceIndexPlot(
  mod_logit_best,
  vars  = c("Cook", "hat", "Studentized"),
  id    = list(n = 5, cex = 0.7),
  main  = "Logistic regression — Influence diagnostics"
)
```

Points with high Cook's Distance *and* high hat values (leverage) are the most consequential: they sit far from the covariate centroid and shift the fitted surface. Given the ~2 M-row dataset, individual influential observations rarely destabilise the model, but clusters of flagged rows sharing a covariate pattern (e.g., a specific precinct–ethnicity–time combination) would warrant closer scrutiny.

---

## Naive Bayes — Predictor Correlation (Independence Assumption Check)

Naive Bayes assumes conditional independence of all features given the class. This is almost certainly violated: time-of-day, borough, and demographic covariates are jointly distributed. The heatmap below shows the Pearson/phi correlation matrix of the dummy-encoded predictor matrix on the training set. Strong off-diagonal correlations directly quantify how badly the independence assumption is violated.

```{r nb-corrplot, fig.width=9, fig.height=8}
# Build dummy matrix from training data using the same formula (drop intercept)
X_nb_corr <- model.matrix(f_final, data = train)[, -1]

# Correlation matrix (Pearson on 0/1 dummies = phi coefficient)
cor_mat <- cor(X_nb_corr)

# Keep only the top 25 columns/rows by mean absolute off-diagonal correlation
# to keep the plot legible when f_final expands to many dummies
mean_abs_cor <- colMeans(abs(cor_mat) - diag(ncol(cor_mat)) / ncol(cor_mat))
top_idx      <- order(mean_abs_cor, decreasing = TRUE)[seq_len(min(25, ncol(cor_mat)))]
cor_sub      <- cor_mat[top_idx, top_idx]

ggcorrplot(
  cor_sub,
  method   = "square",
  type     = "lower",
  lab      = TRUE,
  lab_size = 2.2,
  colors   = c("#d73027", "white", "#4575b4"),
  title    = "Naive Bayes — Predictor correlation matrix (top 25 dummies by mean |r|)",
  ggtheme  = theme_minimal(base_size = 10)
) +
  theme(
    axis.text.x     = element_text(angle = 45, hjust = 1, size = 7),
    axis.text.y     = element_text(size = 7),
    legend.position = "right"
  )
```

Pairs with $|r| > 0.3$ signal meaningful shared information. The independence assumption is most strained when dummy variables from the *same* factor (e.g., multiple `boro` levels) are correlated with those from another factor (e.g., `ethnicity` or `time_bracket`), because Naive Bayes then effectively double-counts that shared signal when computing posterior probabilities. Despite this, Naive Bayes often generalises well in practice—its posterior *rankings* (used for AUC) are more robust to the violation than its posterior *magnitudes* (used for calibration), which is why Log-Loss and Brier score tend to suffer more than AUC for this model class.

---

## LDA — Discriminant Score Separation (LD1)

Linear Discriminant Analysis projects each observation onto one or more linear discriminant axes that maximise between-class variance relative to within-class variance. For a two-class problem there is a single axis, **LD1**. Good separation requires the two class distributions on LD1 to be well apart relative to their spread; heavy overlap signals that the linear boundary will struggle to separate `arrested = 1` from `arrested = 0`.

```{r lda-ld1-density, fig.width=8, fig.height=4, message=FALSE, warning=FALSE}
lda_scores <- data.frame(
  LD1      = predict(mod_lda, newdata = test)$x[, 1],
  arrested = factor(test$arrested, levels = c(0, 1),
                    labels = c("Not arrested (0)", "Arrested (1)"))
)

ggplot(lda_scores, aes(x = LD1, fill = arrested, colour = arrested)) +
  geom_density(alpha = 0.35, linewidth = 0.8) +
  scale_fill_manual(values   = c("Not arrested (0)" = "#4575b4",
                                  "Arrested (1)"     = "#d73027")) +
  scale_colour_manual(values = c("Not arrested (0)" = "#4575b4",
                                  "Arrested (1)"     = "#d73027")) +
  labs(
    x        = "LD1 score",
    y        = "Density",
    fill     = NULL, colour = NULL,
    title    = "LDA — Distribution of LD1 scores by true class (test set)",
    subtitle = "Greater separation ⟹ better discriminability"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "top", panel.grid.minor = element_blank())
```

Well-separated modes indicate that LD1 alone carries strong discriminative signal. Significant overlap—common when the positive class is rare and the feature space is high-dimensional—explains why AUC is imperfect even when the linear boundary is optimal under Gaussian assumptions.

---

## LASSO — Coefficient Paths and CV Deviance vs $\lambda$

LASSO's $\ell_1$ penalty induces sparsity: as $\lambda$ grows, predictors are progressively zeroed out. The coefficient path reveals which predictors survive at $\lambda_{\min}$ and which are eliminated at the more parsimonious $\lambda_{\text{1se}}$.

```{r lasso-diag, fig.width=12, fig.height=5, message=FALSE, warning=FALSE}
lasso_fit   <- cv_lasso$glmnet.fit
coef_mat_l  <- as.matrix(coef(lasso_fit))[-1, ]
log_lam_l   <- log(lasso_fit$lambda)

coef_path_l <- as.data.frame(t(coef_mat_l)) %>%
  mutate(log_lambda = log_lam_l) %>%
  tidyr::pivot_longer(-log_lambda, names_to = "predictor", values_to = "coef")

p_lasso_path <- ggplot(coef_path_l,
                        aes(x = log_lambda, y = coef, group = predictor)) +
  geom_line(alpha = 0.35, linewidth = 0.5, colour = "#d73027") +
  geom_vline(xintercept = log(cv_lasso$lambda.min), linetype = "dashed",
             colour = "firebrick", linewidth = 0.7) +
  geom_vline(xintercept = log(cv_lasso$lambda.1se), linetype = "dotted",
             colour = "darkorange", linewidth = 0.7) +
  labs(x = expression(log(lambda)), y = "Coefficient",
       title = "LASSO — Coefficient paths (sparsity)",
       subtitle = "Red dashed = λ.min · Orange dotted = λ.1se") +
  theme_minimal(base_size = 11) +
  theme(panel.grid.minor = element_blank())

cv_df_l <- data.frame(
  log_lambda = log(cv_lasso$lambda),
  cvm        = cv_lasso$cvm,
  cvup       = cv_lasso$cvup,
  cvlo       = cv_lasso$cvlo
)

p_lasso_cv <- ggplot(cv_df_l, aes(x = log_lambda, y = cvm)) +
  geom_ribbon(aes(ymin = cvlo, ymax = cvup), alpha = 0.15, fill = "#d73027") +
  geom_line(colour = "#d73027", linewidth = 0.9) +
  geom_vline(xintercept = log(cv_lasso$lambda.min), linetype = "dashed",
             colour = "firebrick", linewidth = 0.7) +
  geom_vline(xintercept = log(cv_lasso$lambda.1se), linetype = "dotted",
             colour = "darkorange", linewidth = 0.7) +
  labs(x = expression(log(lambda)), y = "CV Binomial Deviance",
       title = "LASSO — Cross-validated deviance",
       subtitle = "Ribbon = ±1 SE") +
  theme_minimal(base_size = 11) +
  theme(panel.grid.minor = element_blank())

p_lasso_path + p_lasso_cv
```

The point where paths first leave zero marks each predictor's entry into the model. Predictors that survive to $\lambda_{\text{1se}}$ are the most robustly informative signal carriers. A wide flat region in the CV curve indicates that the model is relatively insensitive to $\lambda$ in that range — the bias–variance trade-off is shallow — while a sharp minimum suggests strong sensitivity.

---

## GAM — Smoothness Diagnostics (`gam.check`)

`mgcv::gam.check()` produces four residual plots (Q-Q, residuals vs fitted, histogram of residuals, response vs fitted values) and tests whether the basis dimension $k$ is sufficient for each smooth term. A significant $p$-value in the $k$-index test indicates the smooth may be under-smoothed — the effective degrees of freedom are hitting the basis limit and a higher $k$ should be tried.

```{r gam-check, fig.width=10, fig.height=8, message=FALSE, warning=FALSE}
par(mfrow = c(2, 2))
mgcv::gam.check(mod_gam, pch = 19, cex = 0.3)
par(mfrow = c(1, 1))
```

```{r gam-smooth-plots2, fig.width=10, fig.height=4, message=FALSE, warning=FALSE}
par(mfrow = c(1, 2))
plot(
  mod_gam,
  shade     = TRUE,
  shade.col = "lightblue",
  seWithMean = TRUE,
  scale     = 0,
  pages     = 1,
  main      = "GAM — Estimated smooth terms with 95% CI"
)
par(mfrow = c(1, 1))
```

The Q-Q plot assesses normality of deviance residuals — for logistic GAMs, heavy tails or systematic curvature indicate model misspecification. The residuals vs fitted plot should show no trend; a funnel or arch suggests an omitted non-linear term or interaction. The smooth-term plots (second figure) show the estimated partial effect of `hour_num` and `age` on the log-odds of arrest: a flat line within the confidence band would indicate the smooth contributes no signal, while a pronounced curve validates the non-linear specification over a simpler linear term.
